### 2024-04-05 Latex/main.tex
Add Latex source files and pdf

\documentclass[11pt]{article}

%%PACKAGES
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage[affil-it]{authblk}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

%% SHORTCUTS
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\erf}{erf}
\newcommand {\R}{\mathbb{R}}
\newcommand {\E}{\mathbb{E}}
\newcommand {\prob}{\mathbb{P}}
\newcommand {\N}{\mathbb{N}}
\newcommand {\Z}{\mathbb{Z}}
\newcommand {\1}{\mathbb{1}}

%TITLE
\title{
 \textbf{Analysis of Greenland narwhals's behavioural responses to ship noise and airguns exposure using varying coefficients correlated velocity models}}

\author{Adeline Samson%
\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}

\author{Alexandre Delporte%
\thanks{\texttt{alexandre.delporte@univ-grenoble-alpes.fr}} }
\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}
\author{Susanne Ditlevsen%
\thanks{\texttt{susanne@math.ku.dk}}}
\affil{Department of Statistics, University of Copenhagen, Denmark}

\date{Dated: \today}

\addbibresource{biblio.bib}

\begin{document}

\maketitle

\newpage

\section{Introduction}

%context and literature review, objective of the paper, new contributions
% ARCTIC , narwhals etc
Assessment of human disturbance on animal behaviour has become a major concern in the ecology community. Both instantaneous and long-term changes in animal behaviour have been studied by taking advantage of the large amount of wildlife data now available. One of the commonly accepted method to analyse such behavioural changes due to human activities is to perform "controlled exposure experiments". Animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, sound exposure level, temperature, pressure, heart beat, stroke rate...  
Several studies involving beaked whales \parencite{tyack_beaked_2011}, common sole larvaes \parencite{bolle_common_2012}, narwhals \parencite{heide-jorgensen_behavioral_2021} and many other species have already been done. Marine mammals are particularly challenging to track since GPS data can only be retrieved when they emerge at the surface of the sea and a satellite is near enough. Moreover, some species can dive at great depth, so that the equipment must keep working in such extreme environment \parencite{cioffi_trade-offs_nodate}. In this paper, a data-set of east Greenland narwhals's movement under controlled exposure experiments will be analysed. \\

As a consequence of climate change and decrease in sea ice coverage, the Greenland ecosystem is likely to be more and more exposed to anthropogenic activity in the next decades \cite{heide-jorgensen_behavioral_2021}. Shipping activities as well as sonar and seismic surveys are a few instances of new disturbances that could alter animals's habits.  Hence, there is an urge for more knowledge about the impact this could have on endemic species such as the narwhals. This marine mammal is one of the most sensitive to sound exposure as it uses echolocation to orient itself beneath the water \cite{tervo_narwhals_2021}. Moreover, recent studies showed that the population of narwahls in Southeast Greenland is already in decline \cite{garde_biological_2022}. The effects of sound and ship exposure on its diving and foraging behaviour have already been the subject of several studies \cite{heide-jorgensen_behavioral_2021,ngo_understanding_2019}. However, their horizontal movement has for now been analyzed with discrete time methods such as Markov chains  \cite{heide-jorgensen_behavioral_2021}. A covariate representing the level of sound received by the narwhals was used to model the transition probabilities of the Markov Chain, and it was concluded that narwhals effectively tried to avoid the ship by moving fast toward the shore.\textcolor{red}{add more detailed explanations}. Here, we use a continuous time model to analyse the GPS positions and try to corroborate this analysis. The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\

We propose to use a stochastic differential equation with varying coefficient to model the GPS tracks of the individuals. The continuous time correlated random walk (also known as integrated Ornstein-Uhlenbeck, or correlated velocity model) that has been introduced in \cite{johnson_continuous_2008} is one of the most common model one can find in the literature to analyse such data. It has been  widely used in ecology to model 2D animal movement \cite{gurarie_correlated_2017,albertsen_generalizing_2019} \textcolor{red}{add more references}, and even 3D movement \cite{gurarie_estimating_2011}.

 Following \cite{gurarie_correlated_2017}, this model can be written
\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-\frac{1}{\tau} (v(t) -\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: CVM equation}
\end{equation}
$v(t)$ is the horizontal 2D velocity at time $t$ and $x(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, and $\mu$ is the long term velocity. This equation is a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. \\

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process. However, this phenomenon is not taken into account in model~\ref{eq: CVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recent papers including constraints are \cite{hanks_reflected_2017} and \cite{russell_spatially_2018}.
 In \cite{hanks_reflected_2017}, a reflected version of the CVM is considered to study sea lion telemetry data.
For the simulation, points ending up in land are projected on the coastline.  Estimation of the SDE parameters relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. This requirement is met by linearly interpolating between the discrete time observations of the process.
In \cite{russell_spatially_2018}, a repulsive potential function is used to constrain ants movement within a box.
Usually, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For the narwhals, it was concluded from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.

To add a rotational component, we propose to use the rotational version of the CVM \cite{gurarie_correlated_2017,albertsen_generalizing_2019}.
\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-A(v(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega \\
	\omega & \frac{1}{\tau}
\end{pmatrix}$.
The new parameter $\omega$ is an angular velocity that controls how fast the velocity vector should rotate. Notice that if $\omega=0$ this is simply equation \ref{eq: CVM equation}.

The novelty of this paper consists in estimating the parameters $\tau$ and $\omega$ as smooth functions of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This is a way to make the drift in the velocity equation dependent on the location process $x$. \\
Such a model can be estimated in the framework of the \texttt{R} package \texttt{smoothSDE}. To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuous_2008}. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.

On top of this, the parameter $\nu$ will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
The effect of sound exposure is assessed both in the constrained and unconstrained frameworks.

To summarize, the main contributions of this paper are
\begin{itemize}
	\item Derivation of a state-space model for the RACVM defined in \cite{gurarie_correlated_2017}, and addition of this method in the framework of \texttt{smoothSDE}, to enable the use of smooth parameters $\nu$,$\tau$,$\omega$ and $\mu$, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a Constrained Rotational Correlated Velocity Model (CRCVM) for simulation study. Deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the Distance to the ship.
\end{itemize}

Note that simpler SDE models as Brownian motion or Ornstein-Uhlenbeck with varying coefficient already proved to be convenient at assessing behavioural responses by examining deviation from a baseline model that represents the animal behaviour under normal conditions. For instance, it was used in \cite{michelot_continuous-time_2022} to estimate avoidance behaviours of beaked whales after exposure to military sonar systems.

Eventually, while a lot of results are known about the existence and properties of constrained diffusions, their application in the context of animal motion remains relatively underdeveloped \cite{brillinger_simulating_2003}. This text aims to contribute to bridging this gap.
\newpage

\section{Movement data of Greenland endemic narwhals}
%describe how the data was obtained, the geography, type of whales, year, exposure to boats etc..
%trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
The dataset analysed in this paper has already been the subject of several studies, focusing mainly on the diving behaviour of the narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021,ngo_understanding_2019}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters. The Scoresby Sound fjord system
is known to be the summer residence for an isolated population of narwhals.
Here we recall briefly how the data has been retrieved. For more details about the study area and the tagging of the animals, the reader may refer to \cite{heide-jorgensen_behavioral_2021}. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every 80 s,
and the GPS navigation system recorded the
location of every shot. 
Data includes narwhals's latitude and longitude, distance of the narwhals relative to the ship in metres, GPS time, distance to the shore in metres. GPS positions are known only at specific times when the narwhals get close to the sea surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.\textcolor{red}{add histograms of time steps}\\

Experiments were divided into unexposed periods - for which the narwhals are not in line of sight with the ship - trial periods - when the narwhals are exposed to the ship and airguns are shot - and intertrial periods - when the narwhals are exposed to the ship but airguns are not shot. These periods are indicated by a categorical variable $T_{ship}$ in the dataset. Figure~\ref{fig: trials and intertrials distributions} shows how these periods are distributed among the $6$ narwhals that were tracked.

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.4]{images/data_exploration/trials.png}
    \caption{Trial and Intertrial periods for each narwhal}
    \label{fig: trials and intertrials distributions}
\end{figure}

For each narwhal, the entire track was separated into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period during exposure, after the narwhal has been in line of sight with the ship for the first time. We kept only GPS measures 24 
hours after tagging, to avoid any tagging effects on the narwhals's movement, as recommended in \cite{heide-jorgensen_behavioral_2021}.
Measures that resulted in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
Overall, $4192$ GPS positions were kept for the analysis and projected to UTM coordinates zone 26. The splitting between data before and after exposure resulted in $935$ measures before exposure and $3257$ measures after exposure. Table~\ref{table: data distribution} shows how the data is distributed among the different narwhals. Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and during exposure.

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
		\hline
		A1 & 258 & 576\\
		\hline
		A2  & 40 & 515 \\
		\hline
		A3 & 222 & 680 \\
		\hline
		A4 & 291 & 642  \\
		\hline
		A5 & 48 & 419\\
		\hline
		A6 & 76 & 425 \\
		\hline
		\textbf{Total} & \textbf{935} & \textbf{3257} \\
		\hline
	\end{tabular}
	\caption{Distribution of the data among the 6 individuals}
	\label{table: data distribution}
\end{table}

\begin{figure}[H]
    \centering
\begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[scale=0.38]{images/data_exploration/all_tracksBE.png}
    \caption{Tracks before exposure experiments}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[scale=0.38]{images/data_exploration/all_tracksAE.png}
    \caption{Tracks during exposure experiments}
\end{subfigure}
\caption{Movement data of East Greenland narwhals}
\label{fig: tracks before and after exposure}
\end{figure}

The land geometry for this specific region in South-East Greenland was gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
For each GPS measurement time $t_j$, the nearest point $p(t_j)$ on the shoreline was computed and the distance to the shore was determined as the distance to this point. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This was primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points were ignored. Then, the exposure to shore covariate was defined with the following formula
\begin{equation}
\left\{
\begin{array}{ll}
	 E_{shore}=0 & \mbox{if } D_{shore}>D_0 \\
	E_{shore}=\frac{1}{D_{shore}} & \mbox{if } D_{shore} < D_0
\end{array}
\right.
\label{eq: exp shore definition}
\end{equation}
where the threshold $D_0$ was chosen equal to $0.5$ km.
The angle between the narwhal's heading and the shoreline was computed as the angle $\theta$ between each step vector $x(t_j+\Delta_j)-x(t_j)$ and the vector $n(t_j)=x(t_j)-p(t_j)$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,\frac{-\pi}{2}\right[$ indicates movement toward the shore. 
The distance from the ship was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}. The covariates $E_{shore}$ and $\Theta$ will only be used in the constrained SDE models.

\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		%boundary
		\draw[gray, thick] (0,0) -- (0,4);
		%position
		\draw[red,fill=gray] (2,2) circle (.3ex);
		%nearest shore point
		\draw[red,fill=red] (0,2) circle (.3ex);
		%velocity
		\draw[->,gray] (4,0) -- (2,2);
		\draw[->,gray,dashed,thick] (2,2) coordinate (O) -- (0,4) coordinate (A);
		%normal vector
		\draw[->,red,dashed] (O) -- (4,2) coordinate (B);
		%trigonometric circle
		\draw[gray,dashed] (2,2) circle (2cm);
		%angle 
		\pic[draw=orange,->,angle eccentricity=1.2,angle radius=0.5cm] {angle=B--O--A};

		%nodes
		\node[left] at (0,0) {$\partial \mathcal{D}$};
		\node[left] at (0,2) {$p(t_j)$};
		\node[left] at (1.2,2.8) {$\vec{v}(t_j)$};
		\node[below,red] at (3,2) {$\vec{n}(t_j)$};

		\node[above,orange] at (2.3,2.3) {$\theta(t_j)$};
	\end{tikzpicture}
	\caption{Example of nearest shore point and angle $\theta$}
	\label{fig: illustration nearest shore points}
\end{figure}
\begin{figure}[H]
    \centering
    \begin{subfigure}{0.48\textwidth}
    \centering
     \includegraphics[scale=0.45]{images/data_exploration/Dship_histo.png}
    \caption{Histograms of the distance to the ship \\ for each narwhal}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[scale=0.45]{images/data_exploration/Dshore_histo.png}
    \caption{Histograms of the distance to the shore \\ for each narwhal}
    \end{subfigure}
    \caption{Distance to ship and distance to shore covariates}
    \label{fig: distance to shore and distance to ship histograms}
\end{figure}

The exposure to ship covariate was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D_{ship}$ is not known when the narwhal is not in line of sight with the boat, so that exposures are set to $0$ in this case. The closer the narwhal is to the ship, the greater the exposure. This is a proxy for sound exposure levels received by the narwhals. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.All the covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{images/data_exploration/realExp_through_time}
	\caption{}
	\label{fig: realexpthroughtime}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}{|c|c|p{8cm}|c|}
\hline
Covariate & Unit & Description & Domain \\
\hline
$D_{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
\hline
$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1\}$\\
\hline
$E_{ship}(t)=\frac{1}{D_{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
\hline
$E_{ship,I}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=1}$& $km^{-1}$ & intertrial exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
\hline
 $E_{ship,T}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=2}$ & $km^{-1}$ & trial exposure level of the narwhal to the ship disturbance at time $t$. & $\R^+$\\
 \hline
 $D_{shore}(t)$ & km & distance between the narwhal and the nearest point in the shore at time $t$ & $\R^+$ \\
 \hline
 $E_{shore}(t)=\frac{1}{D_{shore}(t)} \1_{D_{shore}(t)<D_0}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
 \hline
 $\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical step vector at time $t$  & $[-\pi,\pi]$ \\
 \hline 
\end{tabular}
\caption{Summary of the covariates}
\label{tab: covariates}
\end{table}

\section{Diffusion models}

\subsection{Rotational correlated velocity model (RCVM)}
\label{section: RCVM}

The more general equation for a rotational correlated velocity model is \cite{albertsen_generalizing_2019}:

\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-A(v(t)-\mu)dt+\Sigma dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$.
Here, $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity. This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2018}.
There is an explicit form for the gaussian transition density of the velocity process \cite{albertsen_generalizing_2019}, where the covariance matrix is expressed using a Kronecker sum.

It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$.
This diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$ to give biological meaning to the parameters \cite{gurarie_correlated_2017}.These simplifications produce the equation~\ref{eq: RACVM equation}.

When parameters $\tau$, $\mu$ and $\nu$ are constant over time, both the velocity and the position processes can be solved explicitly.
The transition density for the velocity process is such that 
\begin{equation}v(t+\Delta) \vert v(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a rotation matrix of angle $-\omega \Delta$.
Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.

%stationary distribution, distribution of the norm, distribution of the phase
The stationary distribution of the velocity process is $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$. This result can be used to get the long term distribution of the velocity norm and the velocity phase angle. After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$. In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$. Hence, the order of magnitude of the parameter $\nu$ can be deduced from the data by computing the overall mean empirical velocity norm. The distribution of the phase angle $\Phi$ of the velocity has been derived in \cite{pawula_phase_1982}.
When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$. Hence, taking a look at the histograms of the phase angle of the data can help to assess whether a non zero mean velocity is necessary in the model or not.

Then, by integrating the velocity equation over time one gets the transition density of the location process.
\begin{equation}
	x(t+\Delta) \vert x(t)=x, v(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the covariance matrix $Q(\Delta)$ has an explicit form.

\[
	Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
Note that the position process is not markovian, only the whole process $(x(t), v(t))^\top$ is markovian.
Remark that both components of the location and velocity processes are independent.
One can check that for $\omega=0$, the formulas match the one of the CTCRW model \cite{johnson_continuous_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase. Derivation of the formulas given in this paragraph is detailed in appendix~\ref{section: detailed RACVM computations}.\\

%samples with non zero and zero mu

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/tau5_nu1_omega0mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
	\caption{(a)-(b) Two samples of a RCVM with respective parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h  and $\tau=5 h$, $\nu=1$ km/h,$\omega=0$ rad/s, $\mu=(0,0)$ km/h. The black cross indicates the starting point of the process. (c) - (d) Histograms of the empirical velocities obtained from the samples, along with the theoretical probability density function. (e) - (f) Histograms of the empirical velocity phase angles obtained from the samples, along with the theoretical probability density function. }
	\label{fig: sample RACVM}
\end{figure}

\subsection{Constrained rotational correlated velocity model}
\label{section: CRCVM}

To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
Denote $\mathcal{D} \subset \R^2$ the domain of the process $x$.
\begin{equation} \left\{
	\begin{array}{l}
		dv(t)=-A(x(t),v(t))v(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
		dx(t)=v(t) dt
	\end{array}
	\right.
	\label{eq: CRCVM equation}
\end{equation}

Here, $A(x(t),v(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(D_{shore}(t),\Theta(t)) \\
	\omega(D_{shore}(t),\Theta(t)) & \frac{1}{\tau}
\end{pmatrix}$
with $D_{shore}(t)=d(x(t),\mathcal{D})$,
 $\Theta(t)=\widehat{(x(t)-p(t),v(t))}$ and $p(t)$ is the projection of $x(t)$ on $\partial \mathcal{D}$.

 Solving explicitly such a model is out of reach due to the non linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
 For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, then we can use the rough approximations
 \[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
 In this case, the formulas of section~\ref{section: RCVM} are still valid.
 The question still hanging is : what are the conditions on the function $\omega(D_{shore},\Theta)$ to ensure that the process is constrained ?
 We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $x$ (see section \ref{section: simulation study}).
 An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation. For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$. Moreover, a rotation should be considered only for some specific values of the angle $\Theta$, more precisely $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$, that is when the narwhal is heading in the direction of the shore.\\

 Note that this general model is very flexible, and different species-specific behaviours could be described. For the narwhals, a tendency to move along the shore could be considered by specifying a function $\omega$ that is non zero even when $\Theta \in \left[\frac{\pi}{2}-\varepsilon,\frac{\pi}{2}\right] \cup \left[-\frac{\pi}{2},-\frac{\pi}{2}+\varepsilon\right]$, so that when the animal would be on the verge of leaving the shore, its velocity would be tweaked to align with the shoreline. It is also possible  to specify $\tau$ similarly as a function of $D_{shore}$ and $\Theta$ to model specific behaviour near the shore (e.g higher persistence along the shoreline).\\

\textcolor{red}{
	conditions on the shore (regularity) and time steps\\
 	discuss conditions on omega and tau, existence of solution etc
 }

\subsection{Ship exposure effect}
\label{subsection: ship exposure effect}

%WHY TENSOR SPLINES : because covariates THeta and Eshore are not on the same scale, and identity link function
% FOR TAU: this makes sense to use additive splines since the link function is log
To assess if the exposure to the ship has a significative effect on the narwhals movement, we first fit a baseline model only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the disturbance. Then, we fit a model on the data during exposure and check whether the estimated smooth parameters deviate significatively from the baseline values. We present here the more general baseline and response models. For practical work, they shall be simplified to make estimations more reliable.\\

For the baseline model, only the effect of the shore is included, as discussed in section~\ref{section: CRCVM}. To do this, we propose to use the shore exposure covariate $E_{shore}$, defined as the inverse of the distance to the shore, and the angle $\Theta$.
Moreover, although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance. This is taken into account by using a mixed effect model for the RCVM parameters.
For each narwhal $i \in \{1, \cdots, 6\}$, write $t_{exp}^{(i)}$ the first time when it gets in line of sight with the ship (start of exposure). Then for $t \leq t_{exp}^{(i)}$, that is before exposure to the ship, the narwhals motion processes $v^{(i)}(t)=\begin{pmatrix} v^{(i)}_1(t) & v^{(i)}_2(t) \end{pmatrix}^\top$ and $x^{(i)}(t)=\begin{pmatrix} x^{(i)}_1(t) & x^{(i)}_2(t) \end{pmatrix}^\top$ are supposed to solve equation~\ref{eq: CRCVM equation}. Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters. The parameter $\omega$ is expressed as a smooth function of $\Theta$ and $D_{shore}$ through splines. Since both covariates are not on the same scale (one is in $rad$ while the other is in $km$), tensor splines were judged more appropriate than thin plate regression splines.

\begin{equation} 
	\left\{
	\begin{array}{l}

			\log(\tau^{(i)}(t))=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
			\log(\nu^{(i)}(t))=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
			\omega^{(i)}(t)=\omega_{0,pre}+\sum_{k=1}^{q_E \times q_{\Theta}} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t),\Theta^{(i)}(t)) + b_{\omega,pre}^{(i)}
	\end{array}
	\right.
	\label{eq: baseline model}
\end{equation}

$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are unknown intercepts and $b^{(i)}_{pre}=\begin{pmatrix} b_{1,pre}^{(i)} & b_{2,pre}^{(i)} & b_{3,pre}^{(i)} \end{pmatrix}^\top \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{pre}}\right)\right)$
where $\lambda_{pre}=\begin{pmatrix} \frac{1}{\sigma_{\tau,pre}^2} & \frac{1}{\sigma_{\nu,pre}^2} & \frac{1}{\sigma_{\omega,pre}^2} \end{pmatrix}^\top$ and the variances $\sigma_{\tau,pre}^2$, $\sigma_{\nu,pre}^2$, $\sigma_{\omega,pre}^2$ are unknown.

The $\psi_k$ in equation~\ref{eq: baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
\textcolor{red}{Add plot of the spline basis functions.}\\
The numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$. It is needed to choose them before hand. One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower. \\
Based on the approximation that the covariates are constant on each time step $[t_j^{(i)},t_{j+1}^{(i)}]$, the parameters are piecewise constant.

\begin{equation} 
	\forall j \in \{1,\cdots, n_i\}
	 \left\{
	\begin{array}{l}

		\log(\tau^{(i)}_j)=\log(\tau_{pre})+b_{1,pre}^{(i)} \\
		\log(\nu^{(i)}_j)=\log(\nu_{pre})+b_{2,pre}^{(i)}  \\
		\omega^{(i)}_j=\omega_{0,pre}+\sum_{k=1}^{q_E \times q_{\Theta}} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{3,pre}^{(i)}
	\end{array}
	\right.
	\label{eq: approximate baseline model}
\end{equation}

In matrix form equation~\ref{eq: approximate baseline model} can be written as
\begin{equation}
	\rho_{pre}= X_{pre} \beta_{pre}
	+Z_{pre}b_{pre}
\end{equation}

$\rho_{pre}$ is the vector of the parameter values on each time step, that has size $3n$ \[\rho_{pre}=\begin{pmatrix}
\log(\tau^{(1)}) & \cdots & \log(\tau^{(6)}) & \log(\nu^{(1)}) &
\cdots & \log(\nu^{(6)}) & \omega^{(1)}  &\cdots & \omega^{(6)}
\end{pmatrix}^\top\]
with $	\log(\tau^{(i)})=\begin{pmatrix} \log(\tau^{(i)}_1) & \cdots \log(\tau^{(i)}_{n_i})\end{pmatrix}$ and similarly for the $\log(\nu^{(i)})$ and $\omega^{(i)}$.
$X_{pre}$ and $Z_{pre}$ are respectively the fixed and random effects model matrices with sizes $3n \times (q_E\times q_{\Theta}+3)$ and $3n \times 18$.

$\beta_{pre}=\begin{pmatrix}
	\log(\tau_{pre}) &
	\log(\nu_{pre})  &
	\omega_{0,pre} &
	\omega_{1,pre} &
	\cdots & 
	\omega_{q_E q_{\Theta},pre}
\end{pmatrix}^\top$ is the vector of coefficients to estimate for the pre-exposure model and $b_{pre}=\begin{pmatrix}  b_{\tau,pre}^{(1)} &\cdots& b_{\tau,pre}^{(6)} &b_{\nu,pre}^{(1)}& \cdots & b_{\nu,pre}^{(6)} & b_{\omega,pre}^{(1)} & \cdots & b_{\omega,pre}^{(6)} \end{pmatrix}$ is the vector of random effects.\\

Then, the response model is designed to assess a deviation from the baseline model due to exposure to the ship. We illustrate this by addding a dependency on the covariate $E_{ship}$ in the parameters $\tau$ and $\nu$. 
For $i \in \{1, \cdots, 6\}$ and $t \geq t_{exp}^{(i)}$

\begin{equation} \left\{
	\begin{array}{l}

		\log(\tau^{(i)}(t))=\log(\tau_{0,post}) +\sum_{k=1}^{q_{\tau}} \log(\tau_{k,post})\phi_{k}(E_{ship}(t))+b^{(i)}_{\tau,post} \\
		\log(\nu^{(i)}(t))=\log(\nu_{0,post}) +  \sum_{k=1}^{q_{\nu}} \log(\nu_{k,post}) \xi_{k}(E_{ship}(t)) +b^{(i)}_{\nu,post}  \\
		\omega^{(i)}(t)=\omega_{0,post}+\sum_{k=1}^{q_E\times q_{\Theta}} \omega_{k,post} \psi_k(D_{shore}(t),\Theta(t)) + b_{\omega,post}^{(i)}
	\end{array}
	\right.
	\label{eq: response model}
\end{equation}

Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are unknown intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$.
Under the approximation that parameters are piecewise constant, we can write the mixed effect models in matrix form with the same notations as above.
\begin{equation}
	\rho_{post}
	=X_{post} \beta_{post}
	+Z_{post} b_{post}
\end{equation}

Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations will be discussed in section~\ref{section: inference}.\\

\subsection{Inference}
\label{section: inference}

\subsubsection{State space model for the rotational CVM}

As discussed in \cite{ditlevsen_hypoelliptic_2018}, estimation for such hypoelliptic models can be challenging. The main difficulties come from
\begin{itemize}
	 \item irregular times of observations : we only get position when the marine mammal reaches the sea surface.
	 \item GPS measurement errors, with gaussian or student distribution.
	 \item partial observations : two processes are involved in equation~\ref{eq: RACVM equation}, however only one of them can be observed, that is the position.
\end{itemize}

Several method already exist to estimate constant parameters $\tau$, $\nu$ and $\mu$.
If the parameters are constant, it is possible to use a method of moments based on the distributional results about the velocity process \cite{gurarie_correlated_2017}, but it is only feasible when measurement errors are negligible. More efficiently, maximum likelihood with Kalman filter allows to estimate from observations with errors \cite{johnson_continuous_2008}.
If the parameters are allowed to depend on time through external covariates, then \texttt{smoothSDE} package, which is based on \cite{johnson_continuous_2008} but exploit the capabilities of both the \texttt{mgcv} \cite{wood_generalized_2017} and \texttt{TMB} \cite{kristensen_tmb_2016} packages, is a convenient tool for estimation. 
%Discuss the improvement/differences compared to Gurarie and Albertsen, Michelot etc

Consider equation~\ref{eq: RACVM equation} with parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
If we suppose these parameters to be constant on each time step $[t_j,t_{j+1}]$ it can be solved explicitly. Denote $\Delta_j=t_{j+1}-t_j$, $v_j=v(t_j)$, $x_j=x(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=A(t_j)$. Then applying Itô's formula gives

\begin{equation}
	\forall j \in \{1, \cdots, n\} \quad 
	\left\{ \begin{array}{l}
		v_{j+1}=\mu_j +\exp(-A_j\Delta_j)(v_j-\mu_j)+\zeta_j \\
		x_{j+1}=x_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(v_j-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RCVM autoregressive formula}
\end{equation}

where $\zeta_j$ and $\xi_j$ are centered normal random variables.\\
Denote $\alpha_j=\begin{pmatrix} x_{j,1}  & x_{j,2} & v_{j,1} & v_{j,2}\end{pmatrix}^\top$ the hidden state and $y_j$ the observed GPS position at time $t_j$. Measurement errors are supposed to be gaussian in order to be able to use the basic Kalman filter equations. The state space equations are

\[
\left\{
\begin{array}{l}
	y_j=Z\alpha_j+\varepsilon_j \\
	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
\end{array}
\right.\]
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
\]

$\varepsilon_j \sim \mathcal{N}(0,\sigma_{obs}^2 I_2)$ and the vector $\eta_j=\begin{pmatrix} \xi_{j,1} & \xi_{j,2} & \zeta_{j,1} & \zeta_{j,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero. Its covariance matrix $Q_j$ can be computed explicitly.
\begin{equation}
	Q_j=\begin{pmatrix}
		M_j & L_j \\
		L_j & N_j
	\end{pmatrix}
	\label{eq: state space covariance}
\end{equation}
where $M_j$, $N_j$and $L_j$ are block matrices of size $2 \times 2$ such that 
\[M_j=\left( \Delta-2 \frac{\omega_j \sin(\omega_j \Delta_j)-\frac{1}{\tau_j} \cos(\omega_j \Delta_j)}{\frac{1}{\tau_j^2}+\omega_j^2 } \exp\left( -\frac{\Delta_j}{\tau_j} \right) +\frac{\tau_j}{2} \left( \frac{\omega_j^2-\frac{3}{\tau_j^2}}{\frac{1}{\tau_j^2}+\omega_j^2}-\exp\left( -\frac{2\Delta_j}{\tau_j}\right)\right) \right)I_2\]

\[N_j=\frac{2\nu_j^2}{\pi}\left(1-e^{-\frac{2 \Delta_j}{\tau_j}}\right)I_2\]

\[L_j=\begin{pmatrix} l_1 & l_2 \\
	l_2 & l_1\end{pmatrix}\]
where 
\[l_1=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times \left( 1+\exp\left( -\frac{2\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right) \cos(\omega_j \Delta_j)\right)\]
\[l_2=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times\left( \exp\left( -\frac{\Delta_j}{\tau_j}\right) \sin(\omega_j \Delta_j)-\frac{\omega_j \tau_j}{2} \left(1-\exp\left( -2 \frac{\Delta_j}{\tau_j}\right) \right)\right)\]

One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model.
We coded these formulas in the \texttt{smoothSDE} package to allow fast maximum likelihood inference of smooth parameters $\tau$,$\nu$ and $\omega$ by making the most of the power of the R package \texttt{TMB}.

\subsubsection{Maximum likelihood estimation}
The formulation of a state space model now allows for maximum likelihood estimation using Kalman filter.
We detail the maximum likelihood estimation for the response model in section~\ref{subsection: ship exposure effect}. \\

The spline coefficients in equation~\ref{eq: response model} will be treated as random effects to include a penalization in the log-likelihood. 
Introduce smoothing parameters $\lambda_{\tau}$,$\lambda_{\nu}$, $\lambda_{\omega,E}$ and $\lambda_{\omega,\Theta}$. The vectors of spline regression coefficients $\log(\tau_{post})$, $\log(\nu_{post})$, $\omega_{post}$,  conditioned on the smoothing parameter are supposed to follow independent multivariate normal distributions with means zero and respective precision matrices $\lambda_{\tau}S_{\tau}$, $\lambda_{\nu}S_{\nu}$ and $\lambda_{\omega,E}S_{\omega,E}+\lambda_{\omega,\Theta}S_{\omega,\Theta}$. $S_{\tau}$, $S_{\nu}$, $S_{\omega,E}$ and $S_{\omega,\Theta}$ are penalization matrices of known coefficients such that 
$\log(\tau_{post})^\top S_{\tau}\log(\tau_{post})$, $\log(\nu_{post})^\top S_{\nu}\log(\nu_{post})$, and $\omega_{post}^\top (S_{\omega,E}+S_{\omega,\Theta})\omega_{post}$ measure the wiggliness of the fitted splines \cite{michelot_varying-coefficient_2021}. Thus, the chosen prior favours smooth functions over wiggly ones. Now, the mixed effect model equation can be rewritten as $\rho_{post}=X_{post}\beta_{post}+Z_{post} b_{post}$ where the fixed effect coefficients are changed to $\beta_{post}=\begin{pmatrix} \log(\tau_{0,post}) & \log(\nu_{0,post}) & \omega_{0,post} \end{pmatrix} $ and the random effects to 
\[b_{post}=\begin{pmatrix} \log(\tau_{post}) & b_{\tau,post} & \log(\nu_{post}) & b_{2,post} & \omega_{post} & b_{\omega,post}\end{pmatrix}\]
Note that the actual random effects $b_{\tau,post}$, $b_{\nu,post}$,$b_{\omega,post}$ also enter the same framework with a "penalization" matrix which is identity and smoothing parameters that are the inverse of the variances to be estimated.
All these smoothing parameters are gathered in a vector $\log(\lambda_{post})$ that is estimated from the data.
\\

Denote $y^{(i)}$,  $i \in \{1, \cdots, 6\}$, the observed track for the $i$-th narwhal after exposure, which consists in $n_i$ GPS positions.

The complete likelihood of the data and the random effects  is

\begin{align*}
	L_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs}))&= 
	\prod_{i=1}^{6} p\left(\pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post} \vert \log(\lambda_{post})\right) \\
	&\times p\left(y^{(i)} \vert \pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs})\right)
	\label{eq: random effects likelihood}
\end{align*}

The terms $p\left(b^{(i)}_{post},  
\log(\tau_{post}),\log(\nu_{post}),\omega_{post}\vert \log(\lambda_{post})\right)$ can be computed since $b^{(i)}_{post}$, $\log(\tau_{post})$ and  $\log(\nu_{post})$ are independent and their distributions are known multivariate gaussians.\\

For each individual $i \in \{1,\cdots,6\}$, the distribution of the hidden state $\alpha^{(i)}_j=\begin{pmatrix} x^{(i)}_{j,1} & x^{(i)}_{j,2} & v^{(i)}_{j,1} & v^{(i)}_{j,2}\end{pmatrix}^\top$ conditioned on previous observations and random effects, i.e $\alpha_j^{(i)} \vert (y^{(i)}_{1:j-1},b_{post}^{(i)},\log(\tau_{post}),\log(\nu_{post}),\omega_{post})$, is $\mathcal{N}(\alpha_j^{(i)-}, R_j^{(i)-})$ where the matrices $\alpha_j^{(i)-}$ and $R_j^{(i)-}$ are computed by the Kalman filter with parameter values according to the conditioned values of the random effects and the equation~\ref{eq: response model}.
Hence
\begin{equation} y_j^{(i)} \vert (y_{1:j-1}^{(i)}, b_{pre}^{(i)}) \sim \mathcal{N} \left( Z \alpha_j^{(i)-},Z R_j^{(i)-} Z^\top+\sigma_{obs}^2 I_2 \right)
	\label{eq: ditribution of observations conditioned on random effects}
\end{equation}
and the density $p\left(y^{(i)} \vert \pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post},\beta_{post},\log(\lambda),\log(\sigma_{obs})\right)$ is deduced from this result.

Let $\mathcal{L}_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)= - \log( L_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs})))$ be the joint negative log-likelihood of the data and the random effects. 
Then the likelihood of the data is the integral 
\[
L(\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)=\int_{\mathbb{R}^{N}} \exp(-\mathcal{L}_c(\pmb{b_{post}},\theta_{pre},\log(\lambda_{pre}),\log(\sigma_{obs});y) )d\pmb{b_{post}}
\]
where in this specific case, $N=18+q_{\tau}+q_{\nu}+q_{E}\times q_{\Theta}$.

The package \texttt{smooothSDE} relies on the package \texttt{TMB} to optimize this likelihood and get estimations of the coefficients.
Laplace's approximation consists in finding a minimum of the function $\mathcal{L}_c$ with respect to $b_{post}$ and then approximating the integral over this parameter by the integral of a gaussian curve.
This approximation is then optimized over $\beta_{post}$, $\log(\lambda_{post})$ and $\log(\sigma_{obs})$ using gradient methods and auto-differentiation \cite{kristensen_tmb_2016}. More precisely, optimization is  performed by using the usual \texttt{optim} R function, and plugging in the gradient function calculated by \texttt{TMB} via automatic differentiation. The BFGS method (Broyden–Fletcher–Goldfarb–Shanno algorithm) is generally used for non linear unconstrained problems. If we set constraints on the SDE parameter (maximum values of $\tau$ and $\nu$) based on biological expertise for instance, then Limited Memory BFGS with box constraints method should be preferred. In this study, BFGS method was used.\\
Results of the optimization are estimated parameters intercepts $\hat{\beta}_{post}$, estimated smooothing parameters $\log(\hat{\lambda}_{post})$, estimated modes $\hat{b}_{post}^{(i)}$ of the random effects, and possibly estimated log of the standard deviation $\log(\hat{\sigma}_{obs})$ for the measurement error.
Uncertainties on the estimates are derived from the inverse of the hessian matrix of the negative log-likelihood at estimated values and the delta method \cite{kristensen_tmb_2016}.

\textcolor{red}{Talk about the consistency and asymptotic properties of such estimators}

\section{Simulation study}
\label{section: simulation study}

%samples and estimation
%discussion about tensor splines vs additive splines

\subsection{Simulation of constrained motion}
In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.
For the process to be constrained, the velocity should rotate whenever the position is close to the boundary of the domain. For the velocity to rotate, it is necessary to have a non zero value of $\omega$. Moreover, the closer we are to the shore, the faster the velocity should rotate. 
Eventually, at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. In this case, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
First define a weight based on the distance to the shore.
\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Then define how $\omega$ should depend on the angle $\Theta$ and the distance to the shore $D_{shore}$ based on this weight :

\[
\omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\frac{\pi}{2}))+\tanh(\lambda(\theta-\frac{\pi}{2})))\right)\times c(D_{shore})
\]
for some $\omega_0 \in \R$, $\lambda>0$

\textcolor{red}{Add more explanation about why we choose these formulas}

Figure~\ref{fig: trueomega} shows the smooth function obtained for the parameter $\omega$ and  values different values of $D_0$, $\lambda$ and $\kappa$.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/true_omega_rect.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/true_omega_fjords.png}
		\caption{}
	\end{subfigure}
	\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
		(a) $D_0=0.3$ km, $\lambda_{\omega}=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min. 
		(b)  $D_0=0.3$ km, $\lambda_{\omega}=2$, $\kappa=0.2$ and $\omega_0=\frac{\pi}{2}$ rad/min.}
	\label{fig: trueomega}
\end{figure}

$3$ days long trajectories were sampled in a rectangular domain with sides of length $8$ km from the CRCVM with the smooth function for $\omega$ displayed in figure~\ref{fig: trueomega}(a) and other parameters kept constant ($\nu = 4$ km/h, $\tau=1$ h). For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary. Time steps were chosen equal to $1$ min for high frequency and subsampled to $5$ minutes for low frequency. Gaussian measurement errors with low standard deviation $\sigma_{obs}=3$ m and large standard deviations $\sigma_{obs}=30$ m were added to provide the actual observations. Of these $100$ trajectories, none reach the boundary for $\sigma_{obs}=3$ m and only one for $\sigma_{obs}=30$m. However, too small values of $D_0$ and $\omega_0$, as well as too large measurement errors may produce samples that reach the boundary of the domain. Figure~\ref{fig: samples_rect_CRCVM} shows some of the 100 samples obtained in the rectangular domain. \\

10 trajectories were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints. The parameter $\omega$ was defined as the smooth function displayed in~\ref{fig: trueomega}(b). Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps were also kept constant, equal to $1$ min for high frequency and subsampled to $5$ min for low frequency. None of the 10 trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_ne_omega.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_e_omega.png}
		\caption{}
	\end{subfigure}
	\caption{Samples of the CRCVM. x and y axis are in metres and initial positions are indicated by a red cross (a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: samples_rect_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_ne_omega.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_e_omega.png}
		\caption{}
	\end{subfigure}
	\caption{Samples of the CRCVM. x and y axis are in metres and initial positions are indicated by a red cross. (a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: samples_fjords_CRCVM}
\end{figure}

\begin{comment}

To assess whether the CRCVM can describe a tendecy to move along the boundary, $\tau$ was then also expressed as a smooth function of $D_{shore}$ and $\Theta$.
\begin{align*}	
	\tau(\theta,D_{shore})&=(1-c(D_{shore}))\times \tau_0 +c(D_{shore}) \times   \\
	&+\left(\Delta_0+(\tau_0-\Delta_0)\exp\left(\frac{-\theta^2}{\pi^2}\right)+
	(\tau_1-\Delta_0)\exp\left(-\lambda_{\tau}\left(\theta-\frac{\pi}{2}\right)^2\right)+(\tau_2-\Delta_0)\exp\left(-\lambda_{\tau}\left(\theta+\frac{\pi}{2}\right)^2\right)\right)
\end{align*}
for some $0<\Delta_0<\tau_0<\tau_1,\tau_2$ and $\lambda_{\tau}>0$.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{"images/simulation study/true_tau"}
	\caption{$\tau$ (unit: $h$) as a smoth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$}
	\label{fig:truetau}
\end{figure}

\end{comment}

\subsection{Estimation with tensor splines}

The samples from previous section were used to estimate the function $\omega$, both in the rectangular domain and in Scoresby Sound fjords system. Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom was set to $5$ for each covariate. Hence, $25$ coefficients (including the intercept) were to be estimated. The other parameters of the diffusion were supposed to be known, as well as the measurement error. Underestimation of the smooth parameter $\omega$ is observed close to the shore as $\Theta$ approaches $\pm \pi$ (movement toward the shore). This is easily seen in figures \ref{fig: marginal_estimates_rect_CRCVM}(a)-(b) and and \ref{fig: marginal_estimates_fjords_CRCVM}(a)-(b).

As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging. THta is why slight differences in figures ~\ref{fig: marginal_estimates_rect_CRCVM}(a)-(b).

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in the rectangular domain. (a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: estimate_rect_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_far.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_far.png}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in the rectangular domain. The red curve represents the true value of $\omega$. (a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m. (c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
\label{fig: marginal_estimates_rect_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in Scoresby Sound fjords system. (a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: estimate_fjords_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
		\caption{}
	\end{subfigure}

	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in Scoresby Sound fjords system. The red curve represents the true value of $\omega$. (a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m. (c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: marginal_estimates_fjords_CRCVM}

\end{figure}

\section{Estimation of ship exposure effect on real narwhal data}

A baseline model is fitted on the narwhals's track before exposure. It is supposed to capture the characteristics of the movement under normal conditions. Deviation from the baseline is then assessed by fitting a model with covariates $E$ on the tracks during exposure.

\subsection{Estimation without constraints}

First, a simple CVM (see equation \ref{eq: CVM equation}) was used to assess any behavioural response of the narwhals to the ship exposure.

\subsubsection{Baseline estimations}
Due to the low variability of the estimates of $\tau$ for each individual, it was chosen to only estimate one intercept value for this parameter. It was estimated to $\hat{\tau}_{pre}=0.90 $ km/h with $95 \%$ of its values comprised between $0.85$ and $1.1$ h.
In comparison, harbour seal in Alaska were proved to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \cite{johnson_continuous_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \cite{gurarie_correlated_2017}.
The intercept value for $\nu$ was estimated to be $\hat{\nu}_{pre}=4.19$ km/h with $95\%$ CI $3.79-4.63$.
The estimates for the long term velocity $\mu$ show that all the narwhals have a tendency to move toward the south ($\mu_2<0$) with more individual variability for the east-west direction $\mu_1$. However, confidence intervals are large and it might be possible to simply force $\mu$ to $(0,0)$.

\begin{figure}[H]
    \centering

    \begin{subfigure}{0.49\textwidth}
    \includegraphics[scale=0.3]{images/unconstrained_models/baseline/re_baseline2_mu1.png}
    \caption{Estimated $\mu_1$ (km/h) for the baseline model}
    \label{fig: re_baseline1_mu1}
    \end{subfigure}
     \begin{subfigure}{0.49\textwidth}
    \includegraphics[scale=0.3]{images/unconstrained_models/baseline/re_baseline2_mu2.png}
    \caption{Estimated $\mu_2$ (km/h) for the baseline model}
    \label{fig: re_baseline1_mu2}
    \end{subfigure}
  	\begin{subfigure}{0.49\textwidth}
	\includegraphics[scale=0.3]{images/unconstrained_models/baseline/re_baseline2_nu.png}
	\caption{Estimated $\nu$ (km/h)for the baseline model}
	\label{fig: re_baseline1_tau}
	\end{subfigure}
    \caption{Estimated parameters for each individual in the baseline model : the y axis are the estimated parameters along with $95 \%$ confidence interval and the x-axis represents the narwhals's ID.}
    \label{fig: constraints free baseline estimates}
\end{figure}

The estimated values for $\nu$ are all close to $4$ km/h with slight variations among individuals. This value should not be come as a surprise since the mean value of the horizontal velocity norm in the data is $4.4$ km/h, and it is known  that once the velocity process has reached stationarity (typically after $3\tau$), its norm follows a Rice distribution, whose mean is close to $\nu$ when $\vert\vert\mu\vert \vert$ is 
close to $0$.\\

The response model after exposure will be compared to this baseline model to assess any external effect on the narwhals's movement. Examining deviations from this baseline only makes sense when the model is reliable and describes well the narwhals's movement under normal conditions.
To check the goodness of fit of the baseline model, two statistics are computed from the observed tracks of each narwhal : mean velocity norm and the mean velocity phase. For the $i$-th narwhal, the mean velocity norm and phase are approximated as
\[\tilde{v}^{(i)}=\frac{1}{n_i-1}\sum_{j=1}^{n_i-1}  \tilde{v}^{(i)}_j \quad \tilde{\phi}^{(i)}=\frac{1}{n_i-1} \sum_{j=1}^{n_i-1} \widehat{(v_j^{(i)},e)} \mbox{ where } e=(1,0)\]
where $\tilde{v}^{(i)}_j=\frac{y_{j+1}^{(i)}-y^{(i)}_j}{\Delta^{(i)}_j}$, $n_i$ is the number of measurements, $y^{(i)}_j$ is the GPS position of the narwhal $i$ at time $t_j^{(i)}$ and $\Delta^{(i)}_j=t^{(i)}_{j+1}-t^{(i)}_j$.\\
Then the intercepts and modes of the random effects of the fitted baseline model are sampled 500 times from the posterior distribution using the joint covariance matrix produced by the optimization of the likelihood. Draws of the parameters of the SDE are deduced from the sampled coefficients and one trajectory is simulated for each narwhal according to the sampled parameters. For each simulated trajectory, mean of the norm and the phase of the velocity are estimated. This produces an histogram for each statistic (norm and phase) and each narwhal. If sampled tracks have enough similarities with the observed tracks, then the observed statistics should appear near the middle of the histograms. This checking procedure was introduced in \cite{michelot_continuous-time_2022}. Figures~\ref{fig: baseline velocity norm posterior check} and~\ref{fig: baseline velocity phase posterior check} show respectively the histograms of the mean velocity norm and the mean velocity phase.

 \begin{figure}[H]
    \centering
    \includegraphics[scale=0.6]{images/unconstrained_models/baseline/check_mean_vnorm_baseline2.png}
    \caption{Histogram of the mean velocity computed from trajectories of the fitted model for each narwhal, along with observed values in red.}
    \label{fig: baseline velocity norm posterior check}
\end{figure}

 \begin{figure}[H]
    \centering
    \includegraphics[scale=0.6]{images/unconstrained_models/baseline/check_mean_phase_baseline2.png}
    \caption{Histogram of the phase angle of the velocity computed from trajectories of the fitted model for each narwhal, along with observed values in red.}
    \label{fig: baseline velocity phase posterior check}
\end{figure}

The fitted baseline model seems to capture well the mean velocity norm and phase of the observed data for the majority of the narwhals. However, attention should be drawn on the phase angle histogram for the narwhal A6, which reveals a mismatch between the model and the observed phase angle. Narwhal A6 is likely to be heading in the south direction (mean phase angle close to $-\frac{\pi}{2})$.

\subsubsection{Response estimations}

Once the baseline model is deemed satisfactory, a response model is fitted on the tracks during exposure. We computed the marginal and conditional AIC values with different number of knotrs for the spline estimation of the parameters $\tau$ and $\nu$. The marginal AIC is defined in \cite{wood_generalized_2017} as $- 2L + 2k$ where L is the maximum marginal log-likelihood (of fixed effects), and $k$ is the number of degrees of freedom of the fixed effect component of the model. The conditional AIC also defined in \cite{wood_generalized_2017} is $- 2L' + 2k'$ where $L'$ is the maximum joint log-likelihood (of fixed and random effects), and $k'$ is the number of effective degrees of freedom of the model. Note that  the effective degree of freedom is not properly a number of degrees of freedom, but rather a measure of the flexibility or complexity of the model %detail how AIC is computed (wood); based on the trace of the Hessian matrix of the negative log likelihood.
%Intuition for the effective degreed of freedom
%Tendency to favour more complex model when the true model is not in the set of models tested

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.45]{images/unconstrained_models/response/plotAIC3.png}
	\caption{}
	\label{fig: plotAIC3}
\end{figure}

After $q=7$, the decrease in the marginal AIC curve seems to be lower than it was before. This suggests to choose $q-1=6$ knots for the estimation of the parameters with cubic spline. Note that choosing $q$ a bit lower or higher does not change dramatically the estimated smooth functions.
Figure~\ref{fig: fe_response3_tau_ExpShip} shows the estimated fixed effects of the covariate $E$ for the parameter $\tau$. Figure~\ref{fig: fe_response3_nu_ExpShip} shows the same plot but for the smooth parameter $\nu$. The variable on the x-axis is the distance to the shore, that is the inverse of $E$.

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.45]{images/unconstrained_models/response/fe_response3_tau_ExpShip.png}
    \caption{Estimation of the fixed effects for $\tau$ with 95\% pointwise confidence interval. The red curve is the estimated function, the dotted line is the intercept estimated in the baseline with its 95\% confidence interval in light grey.}
    \label{fig: fe_response3_tau_ExpShip}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.45]{images/unconstrained_models/response/fe_response3_nu_ExpShip.png}
	\caption{Estimation of the fixed effects for $\nu$ with 95\% pointwise confidence interval. The red curve is the estimated function, the dotted line is the intercept estimated in the baseline with its 95\% confidence interval in light grey.}
	\label{fig: fe_response3_nu_ExpShip}
\end{figure}

Deviations from undisturbed behaviour coincide with intervals on the x-axis where the confidence intervals of the baseline estimate and the response estimate are disjoint. 
This happens for the parameter $\nu$ when the narwhals are highly exposed to the ship, more precisely when they are at a distance to the ship lower than $8$ km. In this case, $\nu$ is significantly higher than the baseline value, proving that narwhals tend to move faster when highly exposed to the ship noise. Again, this should not come as a surprise since, for instance, the mean experimental velocity norm when the narwhals are less than $5$ km away from the ship is $5.7$ km/h, which is more than $1$ km/h higher than the mean experimental velocity norm before exposure. 
This also shows that the random component in the velocity equation has more importance when the narwhals are highly exposed to the ship, and can be interpreted as more variability in the narwhals horizontal velocity.
Regarding the persistence parameter, it is significantly lower than average in two cases. First, when the narwhals are very close to the ship (distance $<3.5$ km), it can be up to third times less persistent than usual. Then, up to $25$ km away from the ship, no significant deviation from the baseline value can be ascertained. However, more than $25$ km away from the ship, it can still be detected that the narwhals persistence is a bit lower than the baseline value (around $0.6$ instead of $0.9$). This decrease in persistence reveals that the animals gets generally agitated and have a tendency to change their heading more frequently when the military ship is sailing in the fjords.\\

Narwhals position relative to the ship were considered :
\begin{itemize}
	\item very close when the distance to the ship was lower than the $5\%$. quantile, which is $4.04$ km.
	\item close when the distance to the ship was lower than the $25\%$ quantile, which is $9.35$ km.
	\item medium when the distance to the ship was between the $25\%$ and $75\%$ quantile, that is between $9.35$ and $25.80$ km.
	\item far otherwise.
\end{itemize}
\begin{table}[H]
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline
	Position relative to the coast & mean $\hat{\tau}$  & mean $\hat{\nu}$ & $\frac{\hat{\tau}-\hat{\tau}_{pre}}{\tau_{pre}}$ & $\frac{\hat{\nu}-\hat{\nu}_{pre}}{\hat{\nu}_{pre}}$   \\
	\hline
	Very close & 0.50 & 6.1 & $-44 \%$ & $+45 \%$ \\
	\hline
	Close & 0.88 & 5.3 & not significaant & $+26\%$ \\
	\hline
	Medium&  1.01 &  4.6 &  not significant & not significant \\
	\hline
	Far &  0.60 & 4.3 & $-33\%$ & not significant \\
	\hline
\end{tabular}
\caption{Comparison of response and baseline values of the parameters for different levels of exposure to the ship}
\label{table: baseline vs response parameters comparison}
\end{table}

\textcolor{red}{ADD POSTERIOR PREDICTIVE CHECKS}

\subsection{Estimation with constraints}

\subsubsection{Baseline estimations}

The effect of the shore was included directly in the baseline model through bivariate splines of the Distance to the shore and the angle $\Theta$.
The mean velocity $\mu$ was fixed to $(0,0)$ and $\omega$ was estimated as smooth functions of $D_{shore}$ and $\Theta$.

\textcolor{red}{More details about tensor splines estimations can be found in chapter 4 of \cite{wood_generalized_2017}.}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{images/constrained_models/omega_estimations}
	\caption{Estimation of smooth parameter $\omega$ with tensor splines}
	\label{fig:omegaestimations}
\end{figure}

\subsubsection{Response estimations}

\section{Discussion and further work}

\newpage

\printbibliography

\newpage

\appendix

\section{Detailed computations of the integrated rotational Ornstein-Uhlenbeck model}
\label{section: detailed RACVM computations}

\subsection{Velocity process}

\subsubsection{Solution to the SDE}

To solve this equation, the method is similar to the 1-dimensional case.

Consider $X(t)=\exp(At)v(t)$ where $\exp(At)=\sum_{n=0}^{+\infty} \frac{A^n t^n}{n!}$ is a $2 \times 2$ matrix.\\
Thus $X(t)=g(t,v_t)$ where 
\[\begin{array}{ccccc}
	g & : & \mathbb{R}^+ \times \mathbb{R}^2 & \longrightarrow & \mathbb{R} \\
	& & (t,x_1,x_2) & \longmapsto & e^{A t}
	\begin{pmatrix} 
		x_1 \\
		x_2 \end{pmatrix}
\end{array}\]

The multidimensional Itô formula (theorem 4.2.1 in \cite{oksendal}) gives
\[dX(t)=A\exp(At) v(t) dt +\exp(At) dv(t)\]
Replacing $dv(t)$ by its expression 
\begin{align*}
	dX(t)&=A\exp(At) v(t) dt +\exp(At)\left(-A(v(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \right)\\
	&=A\exp(At) \mu dt+\frac{2\nu}{\sqrt{\pi \tau}} \exp(At)dW(t)
\end{align*}
Integrate from $0$ to $t$
\[\exp(At)v(t)-v(0)=(\exp(At)-I_2) \mu +\frac{2\nu}{\sqrt{\pi \tau}} \int_{0}^{t} \exp(As) dW(s)\]
Hence
\begin{equation}
	v(t)=\exp(-At) v(0)+ (I_2-\exp(-At))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)
	\label{eq: RACVM solution}
\end{equation}
In the sequel, we shall use the notation 
\[\zeta(0,t) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)\]

Replacing $\exp(At)$ by its expression gives more explicitly
\begin{equation}
	v_1(t)=e^{-\frac{t}{\tau}} \left(\cos(\omega t)v_1(0)+\sin(\omega t)v_2(0)\right)+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_1-e^{-\frac{t}{\tau}} \sin(\omega t) \mu_2 +\zeta_1(0,t)
	\label{eq: first random component RACVM}
\end{equation}

\begin{equation}
	v_2(t)=e^{-\frac{t}{\tau}} \left(-\sin(\omega t)v_1(0)+\cos(\omega t)v_2(0)\right)+e^{-\frac{t}{\tau}} \sin(\omega t) \mu_1+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_2 +\zeta_2(0,t)
	\label{eq: second random component RACVM}
\end{equation}
where \begin{align*}
	\zeta_1(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s) \\
	\zeta_2(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_1(s) +\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_2(s)
\end{align*}
\subsubsection{Autoregressive equation}

For $t \geq 0$ and $\Delta >0$, 
\begin{align*}
	v(t+\Delta)&=\exp(-A(t+\Delta))v(0)+(I_2-\exp(-A(t+\Delta))) \mu +\frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} \exp(A(s-(t+\Delta)))dW(s) \\
	&= \exp(-A\Delta) \bigg( \exp(-At) v(0) +(\exp(A\Delta)-\exp(At))\mu \\
	&+\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t \exp(A(s-t)) dW(s) + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-t))dW(s) \bigg)
\end{align*}
However 
\[\exp(-At)v(0)++\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t \exp(A(s-t)) dW(s)=v(t)-(I_2-\exp(-At))\mu\]
Hence

\begin{align*}
	v(t+\Delta)=\exp(-A\Delta) \bigg( (\exp(A\Delta)-\exp(-At)) \mu +v(t)-(I_2-\exp(-At))\mu + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-t))dW(s) \bigg)
\end{align*}

In the end,

\[v(t+\Delta)=\mu +\exp(-A\Delta)(v(t)-\mu) + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s) \]

Let $\zeta(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)$. It is a multivariate gaussian with mean zero.

The variances are given by Itô's isometry.

\begin{align*}
	Var(\zeta_1(t,t+\Delta))&= \frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) \cos^2(\omega (s-(t+\Delta))) ds \\ + &\frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) \sin^2(\omega (s-(t+\Delta))) ds \\
	&= \frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) ds \\
	&= \frac{2\nu^2}{\pi}(1-e^{-\frac{2 \Delta}{\tau}})
\end{align*}

Similarly, 
\[Var(\zeta_2(t,t+\Delta))=\frac{2\nu^2}{\pi}(1-e^{-\frac{2 \Delta}{\tau}})\]

And the terms in $Cov(\zeta_1(t,t+\Delta),\zeta_2(t,t+\Delta))$ cancel so that 
\[
Cov(\zeta_1(t,t+\Delta),\zeta_2(t,t+\Delta))=0
\]

Consequently, 
\[v(t+\Delta) \vert v(t) \sim \mathcal{N}\left( \mu + \exp(-A\Delta)(v(t)-\mu), \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) \]

Intuitively, the next velocity has a mean that is the sum of the long term mean velocity $\mu$ and the deviation of the previous velocity from this long term mean rotated by an angle $\omega \Delta$.
\subsubsection{Stationary distribution}

We have seen that 
\[v(t)=\exp(-At) v(0)+(I_2-\exp(-At))\mu + \zeta(0,t)\]
where $\zeta(0,t) \sim \mathcal{N} \left(0, \frac{2\nu^2}{\pi} (1-e^{-2\frac{t}{\tau}}) I_2 \right) $ so that 
\[v(t) \underset{t \rightarrow +\infty}{\overset{\mathcal{D}}{\longrightarrow}}\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2 \right) \]

If $v(0) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ then for all $t \geq 0$, $v(t) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$

\subsubsection{Distribution of the velocity norm and phase}

suppose that $t\geq \frac{3}{\beta}$ so that the velocity has reached its stationary distribution
\begin{align*}
	&v_1(t) \sim \mathcal{N}\left(\mu_1, \frac{2\nu^2}{\pi}\right) \\
	&v_2(t) \sim \mathcal{N}\left(\mu_2, \frac{2\nu^2}{\pi}\right)
\end{align*}

\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		% Draw axes
		\draw[->] (-1,0) -- (5,0) node[right] {$x$};
		\draw[->] (0,-1) -- (0,5) node[above] {$y$};

		% Draw point at (3,4) with circle
		\fill (3,4) circle (2pt);
		\draw[blue,dashed] (3,4) circle (1.5);

		%Draw vectors
		\draw[->,dotted,red] (-0,0) -- (3.5,3.5) node[right] {$v(t)$};
		\draw[->,dotted,red] (0,0) -- (2.6,4.2) node[above] {$v(t')$};
		% Add node with text $\mu$
		\node at (3,4) [above right] {$\mu$};

		% Add node beside the circle with text $\frac{\sigma^2}{2\beta}$
		\node[blue] at (4.5,3) [anchor=west] {$\frac{\sigma^2}{2\beta}$};
	\end{tikzpicture}
\end{figure}
Thus, $ \vert \vert v(t) \vert \vert$ follows a \href{https://en.wikipedia.org/wiki/Rice_distribution}{Rice distribution} with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$. Its probability density function is thus defined by
\[\forall x \geq 0, \quad f(x)= \frac{\pi}{2\nu^2}x \exp\left(-\frac{\pi (x^2+ \vert \vert \mu \vert \vert^2)}{4\nu^2}\right) I_{0}\left(   \frac{\pi}{2\nu^2}\times \vert \vert \mu \vert \vert x \right)\]
where $I_0$ is a modified Besssel function of the first kind defined such that 
\[\forall y  \in \R, \quad I_0(y)=\sum_{m=0}^{+\infty} \frac{1}{m!^2}\left(\frac{y}{2}\right)^{2m}\]
Its moments can be expressed with
\[\E\left(\vert \vert v(t) \vert \vert\right)=\sqrt{\frac{2}{\pi}}\nu \times \sqrt{\frac{\pi}{2}} L_{\frac{1}{2}} \left( -\frac{\beta}{\sigma^2} \vert \vert \mu \vert \vert ^2 \right)\]

Here, $L_{\frac{1}{2}}(x)=e^{\frac{x}{2}} \left( (1-x) I_0\left( -\frac{x}{2}\right)-xI_1\left( -\frac{x}{2}\right)\right)$ with $I_1(y)=\sum_{m=0}^{+\infty} \frac{1}{m!(m+1)!}\left(\frac{y}{2}\right)^{2m+1}$.\\

In the case $\mu=(0,0)$, the mean velocity norm is simply 
\[\E(\vert \vert v(t) \vert \vert)=\nu L_{\frac{1}{2}}(0)=\nu I_{0}(0)=\nu\]

Then, in the plane, $v(t)$ can be thought of as a complex random variable $v(t)=v_1(t)+iv_2(t)$ that has complex normal distribution with mean $\mu_1+i\mu_2$, variance $\E\left((v(t)-\E(v(t)))\overline{(v(t)-\E(v(t)))}\right)=\frac{4\nu^
	2}{\pi}$ and pseudo-variance $\E\left((v(t)-\E(v(t)))^2\right)=0$. Thus, $v(t)$ has a circular symmetric normal distribution with mean $\mu_1+i\mu_2=\vert \vert \mu \vert \vert e^{i\theta}$ where $\theta=\arctan(\mu_2,\mu_1) \in [-\pi, \pi]$ and variance $\frac{\sigma^2}{\beta}$.
Denote $\Phi=\arctan(v_2(t),v_1(t)) \in [-\pi, \pi]$ the phase angle of $v(t)$.
According to \cite{letzepis_vonmises_2015,bennett_solving_1956}, the distribution of $\Phi$ is 
\[\forall \phi \in [-\pi,\pi], \quad f_{\Phi}(\phi)=\frac{e^{-\frac{\beta \vert \vert \mu \vert \vert^2}{2\sigma^2}}}{2\pi} +\frac{\sqrt{\beta} \vert \vert \mu \vert \vert}{2\sigma\sqrt{2\pi}} \cos(\phi-\theta)e^{-\frac{\beta\vert \vert \mu \vert \vert^2 \sin^2(\phi-\theta)}{2\sigma^2}} \times \left(1+\erf\left(\frac{\beta^{\frac{1}{4}}\vert \vert \mu \vert \vert \cos(\phi-\theta)}{\sqrt{2\sigma}}\right)\right)\]
where $\erf$ is the Gauss error function defined by 
\[\forall z \in \R, \quad \erf(z)=\frac{2}{\sqrt{\pi}} \int_0^z e^{-t^2} dt\]
When $\mu=0$, this simplifies to a uniform distribution over $[-\pi,\pi]$.\\
%plot of the function for non zero mu
\subsubsection{Velocity autocovariance function}

In this section we suppose that $v(0)$ has the stationary distribution.The autocovariance function is the expected dot product of the velocity vectors over different lags.

\[\mathbb{E}(\langle v(t), v(t+\Delta) \rangle =Cov(v_1(t)v_1(t+\Delta))+Cov(v_2(t),v_2(t+\Delta))+\mathbb{E}(v_1(t)) \mathbb{E}(v_1(t+\Delta))+\mathbb{E}(v_2(t)) \mathbb{E}(v_2(t+\Delta))\]

However,
\begin{align*}
	Cov(v_1(t)v_1(t+\Delta))&=e^{-\frac{2t+\Delta}{\tau}} (\cos(\omega t)\cos(\omega (t+\Delta))  Var(v_1(0))+\sin(\omega t)\sin(\omega (t+\Delta))  Var(v_2(0)))\\
	&+Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta)) \\
	&= 2\frac{\nu^2}{\tau} \cos(\omega \Delta)  e^{-\frac{2t+\Delta}{\tau}} +Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))
\end{align*}

Then compute
\begin{align*}
	&Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))=\\
	&Cov \bigg( \frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s), \\
	& \frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \cos(\omega(s-(t+\Delta))) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \sin(\omega(s-(t+\Delta))) dW_2(s)\bigg) \\
	&=
	\frac{4 \nu^2}{\pi \tau} Cov \left( \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s), \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \cos(\omega(s-(t\Delta))) dW_1(s) \right) \\
	&+ \frac{4 \nu^2}{\pi \tau} Cov \left( \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s), \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \sin(\omega(s-(t\Delta))) dW_2(s) \right) \\
	&= \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \cos(\omega(s-t)) \cos(\omega(s-(t+\Delta))) ds \\
	& + \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \sin(\omega(s-t)) \sin(\omega(s-(t+\Delta))) ds \\
	&= \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \cos(\omega \Delta) ds
\end{align*}
Therefore we get
\[
Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}} (1-e^{-2\frac{t}{\tau}})
\]
And eventually, 
\[ Cov(v_1(t)v_1(t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}\]
Similarly one gets
\[ Cov(v_2(t)v_2(t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}\]

Therefore, the velocity autocovariance function is

\begin{equation}
	\mathbb{E}(\langle v(t), v(t+\Delta) \rangle )=\frac{4\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}+\vert \vert \mu \vert \vert^2
	\label{eq: RACVM autocovariance function}
\end{equation}

\subsection{Location process} \label{RACVM location process}
For $t \geq 0$ and $\Delta >0$,

\begin{align*}
	x(t+\Delta)&=x(t)+\int_t^{t+\Delta} v(s) ds \mbox{ where } v(s)=\mu+\exp(-A(s-t))(v(t)-\mu)+\zeta(t,s) \\
	&= x(t)+\mu \Delta+\int_t^{t+\Delta} \exp(-A(s-t))(v(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
	&= x(t)+\mu \Delta +\int_0^{\Delta} \exp(-As)(v(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
	&=x(t)+\mu \Delta + (A^{-1}(v(t)-\mu)-A^{-1}\exp(-A\Delta)(v(t)-\mu))+\int_t^{t+\Delta} \zeta(t,s) ds \\
\end{align*}
Thus, 
\begin{equation}
	x(t+\Delta)=x(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v(t)-\mu) +\xi(t,t+\Delta) \\
	\mbox{ where } \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds
\end{equation}
Note that when $\Delta \rightarrow 0$, the deterministic part tends to $x(t)+\Delta v(t)$ as one would expect frome the Euler Maruyama scheme.
It is known that $\xi$ is normally distributed with mean zero. We still have to compute its covariance matrix.

\begin{align*}
	\int_t^{t+\Delta} \zeta(t,s) ds &=\int_t^{t+÷\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
	&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
\end{align*}
From this formula we deduce an expression for both components $\xi_1(t,t+\Delta)$ and $\xi_2(t,t+\Delta)$ of the gaussian vector.
\begin{align*}
	\begin{pmatrix}
		\xi_1(t,t+\Delta) \\ \xi_2(t,t+\Delta) 
	\end{pmatrix}
	&=\frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \frac{1}{\frac{1}{\tau^2}+\omega^2} 
	\begin{pmatrix}
		\frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau}
	\end{pmatrix} \times \\
	& \left[ \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}-\exp\left( -\frac{t+\Delta -u}{\tau}\right) 
	\begin{pmatrix} \cos(\omega(t+\Delta-u)) & \sin(\omega(t+\Delta-u)) \\
		-\sin(\omega(t+\Delta-u)) & \cos(\omega(t+\delta-u)) 
	\end{pmatrix} \right]
	\begin{pmatrix}
		dW_1(u) \\ dW_2(u)
	\end{pmatrix}\\
	&= \frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \int_t ^{t+\Delta} \begin{pmatrix} \frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau} \end{pmatrix} \times \\
	&\begin{pmatrix}  1-\exp\left(-\frac{(
			t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) & -\exp\left( -\frac{t+\Delta-u}{\tau}]\right) \sin(\omega(t+\Delta-u)) \\
		\exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u)) & 1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\end{pmatrix}
	\times \begin{pmatrix} 
		dW_1(u) \\ dW_2(u)
	\end{pmatrix} \\
	&= \frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \int_t ^{t+\Delta} \begin{pmatrix} \frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau} \end{pmatrix} \times \\
	& \begin{pmatrix} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right) dW_1(u)-\exp\left(  -\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u))dW_2(u) \\
		\exp\left(  -\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u))dW_1(u)+\left(1- \exp\left( -\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_2(u)
	\end{pmatrix}
\end{align*}

Eventually, 
\begin{align*}
	\xi_1(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \frac{1}{\tau} \times \Bigg( \textcolor{red}{\int_t^{t+\Delta} \left( 1-\exp\left(-\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_1(u)}\\
	&-\textcolor{blue}{\int_t^{t+\Delta} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))dW_2(u)}\Bigg) \\
	&+\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times \omega \Bigg( \textcolor{red}{\int_t^{t+\Delta} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))dW_1(u)}+ \\
	& \textcolor{blue}{\int_t^{t+\Delta} \left( 1-\exp\left(-\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_2(u)}\Bigg)
\end{align*}

\begin{align*}
	\xi_1(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times  \Bigg( \textcolor{red}{\int_t^{t+\Delta} \bigg( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right)}  \\
	& \textcolor{red}{+\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\bigg)dW_1(u)} \\
	&\textcolor{blue}{+\int_t^{t+\Delta} \bigg( \omega \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta_u))\right)} \\
	&\textcolor{blue}{-\frac{1}{\tau} \exp\left(-\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u)) \bigg) dW_2(u)} \Bigg)
\end{align*}

Similarly, 
\begin{align*}
	\xi_2(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times \Bigg( \textcolor{red}{\int_t^{t+\Delta} \bigg( \frac{1}{\tau} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))} \\
	&\textcolor{red}{-\omega \left( 1-\exp \left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) \bigg)dW_1(u)} \\
	&  \textcolor{blue}{+\int_t^{t+\Delta} \bigg( \omega \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta_u))} \\
	&\textcolor{blue}{+\frac{1}{\tau} \left(1-\exp\left(-\frac{t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) \right) \bigg) dW_2(u)} \Bigg)
\end{align*}

Denote 
\begin{align*}
	& \textcolor{red}{X_1=\int_t^{t+\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) +\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\right)dW_1(u)} \\ 
	&\textcolor{red}{Y_1=\int_t^{t+\Delta} \left( \frac{1}{\tau} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u)) -\omega \left( 1-\exp \left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) \right)dW_1(u)} \\
	&\textcolor{blue}{X_2=\int_t^{t+\Delta} \left( \omega \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right)-\frac{1}{\tau} \exp\left(-\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u)) \right) dW_2(u)} \\
	&\textcolor{blue}{Y_2=\int_t^{t+\Delta} \left( \omega \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))+\frac{1}{\tau} \left(1-\exp\left(-\frac{t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) \right) \right) dW_2(u)} 
\end{align*}
so that both $X_1$,$Y_1$ and $X_2$
,$Y_2$ are independent and 
\[\xi_1(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times (X_1+Y_1) \quad \xi_2(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times (X_2+Y_2)\]
\[Var(\xi_1(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} (Var(X_1)+Var(Y_1)) \quad Var(\xi_1(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} (Var(X_2)+Var(Y_2)) \]

Itô's isometry yields
\begin{align*}
	Var(X_1)&=\int_t^{t+\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) +\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\right)^2du \\
	&= \int_0^{\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{u}{\tau}\right) \cos(\omega u)\right) +\omega \exp \left( -\frac{u}{\tau}\right) \sin(\omega u)\right)^2du 
\end{align*}
Similarly, 
\[
Var(Y_1)=\int_0^{\Delta} \left( \frac{1}{\tau} \exp\left( -\frac{u}{\tau}\right) \sin(\omega u) -\omega \left( 1-\exp \left( -\frac{u}{\tau}\right) \cos(\omega u)\right) \right)^2 du
\]
Hence,
\begin{align*}
	Var(\xi_1(t,t+\Delta))&=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} \Bigg( \int_0^{\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{u}{\tau}\right) \cos(\omega u)\right) +\omega \exp \left( -\frac{u}{\tau}\right) \sin(\omega u)\right)^2+ \\
	& \left( \frac{1}{\tau} \exp\left( -\frac{u}{\tau}\right) \sin(\omega u) -\omega \left( 1-\exp \left( -\frac{u}{\tau}\right) \cos(\omega u)\right) \right)^2 du\Bigg)
\end{align*}

It is possible to see that the formula is exactly the same for $\xi_2(t,t+\Delta)$.

The time integral can be calculated explicitly to obtain
\begin{align*}
	Var(\xi_1(t,t+\Delta))&=Var(\xi_2(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
	& \left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
\end{align*}
It is easy to check that when $\omega=0$, the formula match the one for the basic correlated velocity model the hidden state. \\
With similar calculations, it can be shown that
\[Cov(\xi_1(t,t+\Delta), \xi_2(t,t+\Delta)=0\]

\subsection{State space model, simulation and parameter estimation}
WIth the same notation as before, for $j \in \{1,\cdots,,n\}$, denote $\alpha_j=\begin{pmatrix} x_{1j} & x_{2j} & v_{1j}-\mu_1 & v_{2j}-\mu_2 \end{pmatrix}^\top$.
To formulate a state space model, we shall need the covariances between the velocity and the locations.
Calculations similar to  section \ref{RACVM location process} give for all $t \geq 0$, $\Delta >0$
\begin{align*}
	Cov(\xi_1(t,t+\Delta),\zeta_1(t,t+\Delta))&=Cov(\xi_2(t,t+\Delta)\zeta_2(t,t+\Delta))=\frac{4\nu^2}{2 \pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
	& \left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right)
\end{align*}

\begin{align*}
	Cov(\xi_1(t,t+\Delta),\zeta_2(t,t+\Delta))&=Cov(\xi_2(t,t+\Delta),\zeta_1(t,t+\Delta))=\frac{4\nu^2}{2 \pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
	&\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)
\end{align*}

Therefore, the recursive equations for the state space are
\begin{align*}
	\forall j \in \{1, \cdots, n\}&, \quad 
	\left\{ \begin{array}{l}
		y_j=x_j+\varepsilon_j \\
		v_{j+1}=\mu +\exp(-A\Delta_j)(v_j-\mu)+\zeta_j \\
		x_{j+1}=x_j+\mu \Delta_j+A^{-1}(I_2-\exp(-A\Delta_j))(v_j-\mu)+\xi_j
	\end{array}\right. \\
	&\Leftrightarrow \left\{ \begin{array}{l}
		y_j=Z\alpha_j+\varepsilon_j \\
		\alpha_{j+1}=T_j \alpha_j+B_j \mu + \eta_j
	\end{array}\right.
\end{align*}
where $Z=\begin{pmatrix} I_2 & 0_2 \end{pmatrix}$, $T_j=\begin{pmatrix} I_2 & A^{-1}(I_2-\exp(-A\Delta_j)) \\
	0_2 & \exp(-A\Delta_j) \end{pmatrix}$, $B_j=\begin{pmatrix}
	\Delta_j I_2 \\
	0_2 \end{pmatrix}$ and $\eta_j=\begin{pmatrix} \xi_j \\ \zeta_j\end{pmatrix}=\begin{pmatrix} \xi(t_j,t_j+\Delta_j) \\ \zeta(t_j,t_{j+\Delta_j})\end{pmatrix} \sim \mathcal{N}(0,Q_j)$ with 
\[Q_j=\begin{pmatrix} 
	Var(\xi_{1,j} & 0 & Cov(\xi_{1,j},\zeta_{1,j}) & Cov(\xi_{1,j},\zeta_{2,j})) \\
	\cdots & Var(\xi_{2,j}) & Cov(\xi_{2,j}, \zeta_{1,j}) & Cov(\xi_{2,j},\zeta_{2,j}) \\
	\cdots & \cdots & Var(\zeta_{1,j}) & 0 \\
	\cdots & \cdots & \cdots & Var(\zeta_{2,j}) 
\end{pmatrix}\]

\subsection{Kalman filter equations and likelihood}

Now, the conditional law of $\alpha_i$ knowing $y_{1:i}$ is $\mathcal{N}(\hat{\alpha}_i,R_i)$ and the conditional law of $\alpha_i$ knowing $y_{1:i-1}$ is $\mathcal{N}(\hat{\alpha}_i^-,R_i^-)$ where the expectation and covariance matrices are computed as follows.

\paragraph{Kalman filter equations}
\subparagraph{initialization:}
\[\hat{\alpha}_0=\mathbb{E}(\alpha_0)=\begin{pmatrix} \mathbb{E}(x_{1,0}) \\ \mathbb{E}(x_{2,0})  \\ \mathbb{E}(v_{1,0})-\mu_1 \\ \mathbb{E}(v_{2,0})-\mu_2\end{pmatrix} \quad R_0=\begin{pmatrix} Var(x_{1,0}) & 0 & 0 & 0 \\ 0 & Var(x_{2,0})  & 0 \\ 0 & 0 & Var(v_{1,0}) & 0 \\ 0 & 0 & 0 & Var(v_{2,0}) \end{pmatrix}\]
\subparagraph{prediction:}
For $j \in \{2,\cdots,N\}$,
\[\hat{\alpha}_j^-=\tilde{T}_{j-1} \hat{\alpha}_{j-1}+ \tilde{B}_{j-1} \mu\]
\[ R_j^{-} = \tilde{T}_{j-1} R_{j-1} \tilde{T}_{j-1}^\top + \tilde{Q}_{j-1}\]
\subparagraph{correction:}
\[K_j=R_j^- \tilde{Z}^\top \left( \tilde{Z} R_j^- \tilde{Z}^\top +H_j\right)^{-1} \]
\[\hat{\alpha}_j=\hat{\alpha_j}^{-}+K_j \left( y_j-\tilde{Z} \hat{\alpha}_j^-\right)\]
\[R_j=\left(I_4-K_j \tilde{Z} \right) R_j^-\]
where $I_4$ is the $4 \times 4$ identity matrix. \\

From these equations we can deduce the likelihood of an observation $y^*=\begin{pmatrix} y_1^* &\cdots& y_n^* \end{pmatrix}$.\\
Since $y_j=\tilde{Z}\alpha_j+\varepsilon_j$,  $\alpha_j \vert y_{1:j-1} \sim \mathcal{N}(\hat{\alpha}_j^-,R_j^-)$, and $\varepsilon_j \sim \mathcal{N}(0,H_j)$ is independent from the observations, we have
\[y_j \vert y_{1:j-1} \sim \mathcal{N}\left(\tilde{Z} \hat{\alpha}_i^-,\tilde{Z} R_i^- \tilde{Z}^\top+H_i \right)\]

Thus, if one writes $m_i=\tilde{Z} \hat{\alpha}_i^-$ and $\Gamma_i=\tilde{Z} R_i^- \tilde{Z}^\top +H_i$ then 
\[\mathcal{L}(\sigma, \beta, \mu \vert y^*) =\mathbb{P}(y_0=y_0^*) \prod_{i=2}^n \mathbb{P}(y_i=y_i^* \vert y_{1:i-1}=y_{1:i-1}^*)\]
If we suppose that the initial observation is fixed (not random), then
\[\mathcal{L}(\sigma, \beta, \mu \vert y^*) = \prod_{i=2}^n \frac{1}{\sqrt{(2\pi)^2 \vert \det(\mu_i) \vert}} \exp \left( -\frac{1}{2} (y_i^*-m_i)^\top \Gamma_i^{-1} (y_i^*-m_i) \right) \]
Hence, the log-likelihood is 
\begin{align*}
    \log \left(\mathcal{L}(\sigma,\beta,\mu \vert y^*)\right)&=\sum_{i=2}^n -\log(\sqrt{(2\pi)^2 \vert \det(\Gamma_i) \vert}) -\frac{1}{2} (y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i)\\
    &=\sum_{i=2}^n -\frac{1}{2}\log\left((2\pi)^2 \vert \det(\Gamma_i) \vert\right) -\frac{1}{2} (y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i) \\
    &=-(n-1)\log(2\pi)-\frac{1}{2} \sum_{i=2}^n \log(\vert \det(\Gamma_i) \vert) +(y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i)
\end{align*}
Consequently, the negative log-likelihood that we need to minimize to obtain parameter estimates is 
\[
-\log \left(\mathcal{L}(\sigma,\beta,\mu \vert y^*)\right)=(n-1)\log(2\pi)+\frac{1}{2} \sum_{i=2}^n \log(\vert \det(\mu_i) \vert) +(y^*_i-m_i)^\top \mu_i^{-1} (y^*_i-m_i)
\]

The estimated parameters are 
\[\hat{\theta}=(\hat{\sigma},\hat{\beta},\hat{\mu})=\underset{\sigma,\beta,\mu}{\mathrm{argmin}} \left( \sum_{i=2}^n \log(\vert \det(\Gamma_i) \vert) +(y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i) \right) \]

\section{Precisions about the statistical models}

In the baseline mixed effect model, the explicit model matrices are

\[X_{pre}= \begin{bNiceArray}{ccc|ccc}[margin]
	1 & 0 & 0 & \Block{3-3}<\Large>{0} & &  \\
	\vdots & \vdots & \vdots & & &  \\
	1 & 0 & 0 & & & \\
	\hline
	0 & 1 & 0 &  \Block{3-3}<\Large>{0} & & \\
	\vdots & \vdots & \vdots & & &  \\
	0 & 1 & 0 & & & \\
	\hline
	0 & 0 & 1 &  \Block{3-3}<\Large>{M_{pre}} & & \\
	\vdots & \vdots & \vdots & & &  \\
	0 & 0 & 1 & & & \\
\end{bNiceArray} \quad 
Z_{pre} = \begin{pmatrix}
	J_{1,n_1} \\
	\vdots \\
	J_{6,n_6} \\
	\vdots \\
	J_{13,n_1} \\
	\vdots \\
	J_{18,n_6}
\end{pmatrix} 
\]
where $M_{pre}$ is the matrix with elements  $\psi_k(E^{(i)}_{shore}(t^{(i)}_j))$ on column $k$ and row associated to observation $j$ for narwhal $i$, and $J_{l,m}$ is the matrix with $m$ rows and $18$ columns that has zero everywhere except on column $l$, where there are only ones.

\section{More simulation study}

One track for each of the $6$ narwhals is simulated with known spline fixed effects of the exposure and known random effects. The fixed effects coefficients are estimated from the samples. Figure~\ref{fig: check_fe_response4_sample} shows a sample of the model described in section $3.2$ with known coefficients. This sample is then used to estimate the exposure effects on $\tau$ and $\nu$. True and estimated smooth functions are displayed in figures~\ref{fig: check_fe_response4_sample} and~\ref{fig: check_fe_response4_nu_Xtilde}.

\end{document}

### 2024-04-05 R/preprocessing/data_preprocessing.Rmd
Add R code for data analysis, plots and sde fitting

---
title: "Preprocessing and exploration of the narwhal data for the analysis with smoothSDE"
author: "Alexandre Delporte"
date: "2023-03-01"
output: 
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: sentence
---

# Set up

## Set up Rmarkdown options

```{r options}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, eval = TRUE)
```

## Import necessary packages

```{r import packages,message=FALSE}

#for tex code
library(tinytex)

#for the plots
library(ggplot2)
library(plotly)

#to manage dataframes 
library(dplyr)

#to get polygons for the shore
library(jsonlite)
library(sf)

#to convert to UTM coordinates
library(sp)

#to get stamenmap
library(ggmap)

#to get animation of the narwhals trajectories
library(gganimate)
library(transformr)
library(gifski)

#to show progress bars
library(progress)

#to get function na.approx to replace NA values by interpolation
library(zoo)

```

## Choose directory where the plots should be saved

```{r create plots folder}

plots_folder <-"plots"

if (!(dir.exists(plots_folder))) {
  dir.create(plots_folder)
}

```

# Read the raw data

## Read

```{r read raw data, cache=TRUE}
# Set the path to the directory containing the data
par_dir=dirname(dirname(getwd())) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#get narwhal data
data_file="db_narwhal_2017_2018_with_gps.txt"
rawData <- read.table(file.path(narwhal_data_path,data_file),
                      header = TRUE, sep = "\t",dec = ".")

```

## Summary

```{r summary raw data}
print(summary(rawData))
```

```{=tex}
Here is a brief description of the raw dataframe.
We have $4132760$ observations and $32$ variables.
The columns are
\begin{itemize}
\item \texttt{rawData\$Posixct} : The date and hour in characters
\item \texttt{rawData\$Depth} : The narwhal depth in metres. It goes from $0$ to $-1044$ metres. The GPS positions given by \texttt{rawData\$Lat} and 
\texttt{rawData\$Long} should be known only when depth is reasonably low (close to 0). 
\item \texttt{rawData\$ODBA} : overall dynamic body acceleration, sometimes used as a proxy for energy expenditure, expressed in $m/s^{-2}$ or in $g$ ($1 g=9.8 m s^{-2}$). It is computed from accelerometer data (see "Moving towards acceleration for estimates of
activity-specific metabolic rate in free-living animals: the case of the cormorant)" wILSON 2006 for details.
\item \texttt{rawData\$VeDBA} : vector dynamic body acceleration
\item \texttt{rawData\$Acou\_qua} : 
\item \texttt{rawData\$GPS\_time} : character for date and time when signal was emitted by the GPS : $54$ seconds ahead compared to \texttt{rawData\$Posixct}
\item \texttt{rawData\$Lat} : latitude in $[-90,90]$ degrees in the World Geodetic System 1984. It is either a latitude measure by the GPS or linearly interpolated using previous and next latitude.
\item \texttt{rawData\$DistanceShore} : distance to the nearest shore point, in metres.
\item \texttt{rawData\$Long] : }  longitude in $[-180,180]$ degrees in the World Geodetic System 1984. It is either a latitude measure by the GPS or linearly interpolated using previous and next latitude.
\item \texttt{rawData\$LOS} : binary variable : 1 if the narwhal is in Line of sight from the ship, 0 otherwise.
\item \texttt{rawData\$Seismik} : binary variable
\item \texttt{rawData\$Type} : categorical variable with values "trial" when the narwhal is exposed to the ship and the ship is shooting airguns,
"intertrial" when the narwhal is exposed to the ship without airguns being shot.
\item \texttt{rawData\$TrialNo} : categorical variable with values "T1", ..., "T8", "I1", ..., "I9" where the letter stands for the type 
(intertrial  or trial) and the number for the number of the trial.
\item \texttt{rawData\$Exposure} : categorical variable with levels from 1 to 8.
\item \texttt{rawData\$Year} : year when the data was collected
\item \texttt{rawData\$Click} : binary variable, indicating whether the narwhal is emitting click sounds or not
\item \texttt{rawData\$Call} : binary variable, indicating whether the narwhal is emitting call sounds or not
\item \texttt{rawData\$Buzz} : binary variable, indicating whether the narwhal is buzzing
"The closer the narwhals get to their food, the faster they click, until the noise becomes a buzz". Calls are emitted for communication with
other individuals.
\item \texttt{rawData\$Dist\_Ship} : Distance to the military ship in metres
\item \texttt{rawData\$Area} : categorical variable with "IF" "R"  "O"  "H"  "OG" (outer Gasefjord) "G"  "Vi" "F"(Fonfjord)  "IG" matching the area in Scoresby Sound :

\item \texttt{rawData\$GunSize} : categorical variable with values "B" (big) and "S" (small). 
Big airguns were used only in 2018, and small airgun in 2017 (see Jorgensen 2021)
\item \texttt{rawData\$temperatureCorrected} : 
\item \texttt{rawData\$Dist\_to\_CPA} : Distance in km to the closest point of approach (CPA), that is the point where the narwhals and the ship would be the 
closest if they kept moving linearly (see "CPA calculation method based on AIS position prediction", Yan)
\item \texttt{rawData\$Time\_to\_CPA} : time needed to reach the CPA (I don't get why some values are negative)
\item \texttt{rawData\$Direction\_cat} : categorical variable with values "increasing" or "decreasing".
\item \texttt{rawData\$NoOfStrokes\_20sec} : number of strokes per 20 seconds
\item \texttt{rawData\$Divecat} : categorical variable with values "S", "M", "D" (for surface, medium, deep ?).
\item \texttt{rawData\$Target\_depth} : target depth in metres
\item \texttt{rawData\$Dive\_phase} : categorical variable with values "surface","Descent","Bottom","Ascent","0".
\item \texttt{rawData\$GPS\_time\_chr} : Character for the date and time : this column is identical to column rawData\$GPS\_time. Can be deleted
\item \texttt{rawData\$pos\_org} : binary variable with value $0$ if latitude and longitude have been interpolated, $1$ if it is an actual GPS measure.
\end{itemize}
```

# First preprocessing steps

```{=tex}
Two controlled exposure experiments were performed in 2017 and 2018.
The main differences are that only small guns were used in 2017 while only big guns were used in 2018, and the tagged whales used a much larger part of Scoresby Sound in 2018.

First preprocessing steps include : 
\begin{itemize}
\item keeping only 2018 data - adding a time since tagging column 
\item converting distance to ship and distance to shore to km - adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.
\item Renaming the "Ind" column to "ID" and converting it to factors.
\item Converting column "Type" in factor with levels trial and intertrial 
\item Converting column "Posixct" to POSIXct format for date and time
\end{itemize}

The first measurement time is set to $1$ second and then measures are interpolated every second (see Jorgensen et al.).
Thus, the time $T_k$ of the $k$-th measurement is $$T_{k}=\frac{k}{3600}\mbox{ h }$$ As proposed in Jorgensen et al., we also add a column with a covariate $E$ defined as $$E(t)=\frac{1}{\mbox{distance to ship}(t)}$$ where distance to ship has been converted in km.

\texttt{ExpShip} is a proxy for sound exposure level (SEL).
```

```{r preprocess data 1, cache=TRUE}

#keep only 2018 data
data=rawData[rawData$Year==2018,]

#convert Distance Shore and Dist_ship to km
data$Dist_Ship=data$Dist_Ship/1000
data$DistanceShore=data$DistanceShore/1000

#add time since tagging column

TimeFunction <- function(n, dt = 1){
  ## n: Number of observations. Positive integer
  ## dt: Time step between observations in seconds
  Time <- (1:n)*(dt/3600) ## Time in hours since tagging
}

nrows=length(data[,1])
data$time <- rep(0,nrows)
for(ind in unique(data$Ind)){
  indicator <- (data$Ind == ind)
  data$time[indicator] <- TimeFunction(n = sum(indicator))
}

## Defining exposures

ExposureFunction <- function(X){
  ## X: Distance to ship in meters measured each second. Supposed to be positive

  ## Exposure variable
  E <- 1/X ## Inverse of distance to ship in kilometers
  E[is.na(E)] <- 0 ## If not in line of sight, exposure is zero
  E
}

## Ship Exposure variable based on 1/distance in km
data$ExpShip <- ExposureFunction(X = data$Dist_Ship)

#change name of the IDs column to ID
colnames(data)[1]="ID"
data$ID=factor(data$ID)

#convert Type to factors
data$Type <- factor(data$Type, levels = c("intertrial", "trial"))

#convert Posixct to POSIXct class for time and dates
data$Posixct=as.POSIXct(data$Posixct)
```

# Data exploration after first preprocessing

## Trials and intertrials periods

```{=tex}
During the experiment, two types of exposure can be distinguished
\begin{itemize}
\item[1] Trials : Narwhals are in line of sight with the military ship and the ship is shooting airguns in the sea.
\item[2] Intertrials: Narwhals are in line of sight with the military ship but it has stopped shooting airguns.
\end{itemize}
See Jorgensen et al. for more details.
```

```{r plot trials, cache=TRUE}

#plot with all the narwhals rescaled so that the experiments start at the same time 
trials1=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
  xlab("Number of days after tagging")+ylab("Exposure")+
  geom_tile(lwd = 1,linetype = 1)+ 
  scale_fill_manual(values=c("blue","red"),na.value="white",
                    name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

#plot with all the narwhals at the real experiments time
trials2=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
  xlab("Day")+ylab("Exposure")+
  geom_tile(lwd = 1.5,linetype = 1)+ 
   scale_fill_manual(values=c("blue","red"), na.value="white",
                     name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

ggsave(path=plots_folder,filename="trials_rescaled.png",plot=trials1,width=10,height=5)
ggsave(path=plots_folder,filename="trials.png",plot=trials2,width=10,height=5)
```

## Ship exposure through time and depending on trial type

```{r define ExpShipI and ExpShipT, cache=TRUE}

data$ExpShipI=data$ExpShip
data$ExpShipT=data$ExpShip

data$ExpShipI[data$Type=="trial"]=0
data$ExpShipT[data$Type=="intertrial"]=0

```

It appears that the exposure is not so smooth with respect to time, meaning that a narwhal can be highly exposed at some time and be completely out of sight from the airguns a few seconds later.
This phenomenon appears when the values in DistShip suddenly change to NA.

Moreover, exposure times are relatively short compare to the overall duration of the measurement period, so that most of the time, ExpShip is zero but short peaks appear when trials or intertrials are happening.
We would need to zoom on these peaks to assess their smoothness.

```{r ggplot ship exposure, cache=TRUE}
#plot exposure
pExpAll=ggplot(data,aes(time,ExpShip,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

#plot exposure durinh intertrial
pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

#plot exposure during trial
pExpTr=ggplot(data,aes(time,ExpShipT,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

ggsave(path=plots_folder,filename="exposure_through_time.png",
       pExpAll,width=10,height=5)
ggsave(path=plots_folder,filename="intertrial_exp_through_time.png",
       pExpInt,width=10,height=5)
ggsave(path=plots_folder,filename="trial_exp_through_time.png",
       pExpTr,width=10,height=5)

```

```{r display pExpAll, fig.cap="Global exposure to ship through time"}
pExpAll
```

```{r display pExpTr, fig.cap="Trial exposure to ship through time"}
pExpTr
```

```{r display pExpInt, fig.cap="Intertrial exposure to ship through time"}
pExpInt
```

## True exposure per individual

```{=tex}
\textbf{\texttt{Lat} and \texttt{Long} columns have been interpolated every second, and distance to ship has been computed from these interpolations.
The true measured distance to ship are in the rows for which the column \texttt{pos\_org}  has value 1.}
```

```{r plot exposure per individual,cache=TRUE}

for (id in unique(data$ID)) {
  plot_exp=ggplot(data[data$ID==id,],aes(time,ExpShip))+geom_line()+
  geom_point(data=data[(data$pos_org==1)&(data$ID==id),],
             aes(time,ExpShip),shape=3,colour="red",size=1)

  ggsave(path=plots_folder,
         filename=paste(id,"ship_exposure_through_time.png",sep="_"),
         plot_exp,width=10,height=5)
}
```

```{r example of individual exposure with red points for measured values, fig.cap=""}
ggplotly(plot_exp)
```

# Second preprocesssing step

```{=tex}
\begin{itemize}
\item  Drop missing latitude and longitude
\item Keep GPS positions indicated by value 1 in \texttt{pos\_org} column
\item  Project to UTM zone 26. The narwhal positions are in latitude and longitude.To analyze the track and visualize it better we project it on a plane in UTM coordinates.
The positions of the narwhals are in the UTM zone 26 north (in Greenland).
\item Add a column "speed" with empirical horizontal speed obtained from the GPS positions (not interpolated)
\end{itemize}

```

```{r preprocess data 2, cache=TRUE}

dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]
dataprep=dataprep[dataprep$pos_org==1,]
nrows=length(dataprep[,1])

#project to UTM (Universal Transverse Mercator coordinate system)

llcoord=SpatialPoints(dataprep[,c("Long","Lat")],
                      proj4string=CRS("+proj=longlat +datum=WGS84"))
utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84"))

#add UTM locations in km to data frame
dataprep$x=attr(utmcoord,"coords")[,1]
dataprep$y=attr(utmcoord,"coords")[,2]

#compute empirical horizontal speed
n=length(dataprep$time)
dataprep$horizontal_speed=rep(NA,n)

rows_index=c()
for (id in unique(dataprep$ID)){

  #filter data with specific id
  sub_ind=(dataprep$ID==id)
  sub_data=dataprep[sub_ind,]
  n_sub=length(sub_data$time)

  #time steps
  dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

  #step lengths
  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])/1000
  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])/1000

  speed=c(NA,sqrt((dx/dtimes)^2+(dy/dtimes)^2))

  dataprep[sub_ind,"horizontal_speed"]=speed
}

```

# Data exploration after second preprocessing

## measurement periods for each narwhal
```{r measurement periods}
#print measurement periods for each narwhal

for (id in unique(dataprep$ID)) {
  dataid=dataprep[dataprep$ID==id,]
  print(paste(id,"measurement period : ",min(as.POSIXct(dataid$Posixct)),"to",
              max(as.POSIXct(dataid$Posixct))))
}
```

## Explore horizontal speed

```{r summary horizontal speed}
summary(dataprep$horizontal_speed)
```

```{=tex}
The mean speed is about 4.5 km/h.
Usually, computing the empirical horizontal speed between two positions amounts to consider that the narwhal moved in a straight line between
the two points. But this is obviously not true, especially for long time steps, because the narwhal make U-turns for instance, and more importantly, it dives to find food. Therefore the empirical horizontal speed always underestimates the true horizontal speed.
Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram
```

```{r histogram horizontal speed,warnings=FALSE}
speed_histo=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")

speed_histo
ggsave(path=plots_folder,"speed_histo.png", plot = speed_histo)
```

## Filter horizontal speed

```{r speed higher than 20km/h }
dataprep[dataprep$horizontal_speed>20,c("ID","time","horizontal_speed")]
```

```{=tex}
There are only two horizontal speeds (both for Kyrri track) higher than $20$ km/h. We decide to keep this threshold of 20 km/h to filter the rows.
```

```{r remove rows with speed > 20km/h}
rows_index=which(dataprep$horizontal_speed>20)
dataprep=dataprep[-rows_index,]
```

```{r check filtered dataprep}
#check that now all speeds are inferior to 20 km/h
dataprep[dataprep$horizontal_speed>20,]
```

# More data exploration

## Time steps distribution

```{r compute time steps,warnings=FALSE}

dtimes=data.frame("delta"=rep(NA,nrow(dataprep)),"ID"=rep(NA,nrow(dataprep)))

for (id in unique(dataprep$ID)) {
  ind=dataprep$ID==id
  data_id=dataprep[ind,]
  nrows=length(data_id[,1])
  delta_id=data_id[2:nrows,"time"]-data_id[1:(nrows-1),"time"]
  delta_id=c(NA,delta_id)
  dtimes[ind,"delta"]=delta_id
  dtimes[ind,"ID"]=rep(id,nrows)
}

```

```{r summary time steps, cache=FALSE, echo=TRUE}
print(summary(dtimes$delta*60))
```

```{r}
times_histo=ggplot(dtimes, aes(x =60*delta,col=ID)) +
  geom_histogram(fill="white",bins=100)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Time step (min) ")

times_histo
ggsave(path=plots_folder,"time_steps_histo.png", plot = times_histo)
```

```{r time steps >2h, cache=FALSE, echo=TRUE}

cat("Percentage of time steps > 2h :",length(delta[delta>2])/length(delta)*100)
```

```{r display hist , fig.cap="Histogram of time steps",warnings=FALSE}
all_times_histo=ggplot(data=dtimes,aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")
all_times_histo
ggsave(path=plots_folder,"all_time_steps_histo.png", plot = all_times_histo)
```

```{=tex}
The median of the time steps between consecutive measurement is about $5$ minutes.
Only two time steps exceed $4$h.
```

## Plot all GPS tracks (without interpolation) on a map

```{r milne land map, cache=TRUE}
height <- max(dataprep$Lat) - min(dataprep$Lat)
width <- max(dataprep$Long) - min(dataprep$Long)
borders <- c(bottom  = min(dataprep$Lat), 
                 top     = max(dataprep$Lat),
                 left    = min(dataprep$Long),
                 right   = max(dataprep$Long))

register_stadiamaps(key ="a3062dd7-0f3d-44ea-a6ac-e11c9eeb94b4")
milne_land <- get_stadiamap(borders, zoom = 10, maptype = c("stamen_terrain"))
milne_land_map=ggmap(milne_land)

```

```{r trajectories map, cache=TRUE}

#static map
map=milne_land_map + 
  geom_point(dataprep[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataprep[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataprep[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

ggsave(path=plots_folder,filename="all_narwhals_tracks.png", 
       plot = map,width=10,height=5)

#animated map with gganimate
animate_map=milne_land_map + 
  geom_line(dataprep[,c("ID","Long","Lat","time")], 
            mapping=aes(Long, Lat, col = ID),size= 1,alpha=0.5) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+transition_reveal(time)

animation=animate(animate_map,renderer = gifski_renderer())
anim_save(path=plots_folder,filename="all_narwhals_tracks.gif", 
          animation = animation,width=10,height=5)
```

```{r display map , fig.cap="All narwhal tracks"}
map
```

# Coastline and land data

## Extract coastline and land polygons

```{r extract coastline and land,cache=TRUE}

# Set the path to the directory containing coastline data
par_dir=dirname(dirname(getwd())) #parent directory 
greenland_data_path <- file.path(par_dir,"Data", "Greenland")

#get the coastline geometry from the geojson file
coastline_utm<-st_read(file.path(greenland_data_path,"filtered_coastline_utm.geojson"))
st_crs(coastline_utm)="+init=EPSG:32626"

#get land polygons
land_utm<-st_read(file.path(greenland_data_path,"filtered_land_utm.shp"))

```

## Plot coastline and land

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill=NA,col="brown")+
  coord_sf(datum=st_crs(32626))
plot_coastline=ggplot()+geom_sf(data=coastline_utm$geometry,fill=NA,col="blue")+
  coord_sf(datum=st_crs(32626))
```

```{r display land , fig.cap="Land polygons"}
plot_land
```

```{r display coastline , fig.cap="Coastline polygons and linestrings"}
plot_coastline
```

## Compute nearest point on land and coastline for each GPS position

```{r compute nearest shore points,cache=TRUE}

nearest_points=function(positions,coastline) {
  #positions: dataframe of positions in UTM coordinates in metres
  #coastline : geometry of the coastline projected in the same UTM zone as positions in metres

  # Create an empty vector to store the nearest points
  nearest_points <- matrix(nrow = nrow(positions), ncol = 2)
  min_distances<-matrix(nrow = nrow(positions), ncol = 1)

   #progress bar
  pb <- progress_bar$new(format = "[:bar] :percent",total = nrow(positions), clear = FALSE, width = 60)

  # Iterate over each position
  for (i in 1:nrow(positions)) {

    pb$tick()

    # Extract the position coordinates
    coords <- st_point(c(positions$x[i], positions$y[i]))

    # Initialize variables to store the minimum distance and nearest point
    min_distance <- Inf
    nearest_point <- NULL

    # Iterate over each polygon in coastline
    for (j in 1:length(coastline$geometry)) {

      # Get the current polygon
      polygon <- coastline$geometry[[j]]

      distance <- st_distance(coords,polygon)

      candidate=st_nearest_points(coords,polygon)

      coordinates=st_coordinates(candidate)[2,c("X","Y")]

      if (distance < min_distance) {
        min_distance=distance
        nearest_point=coordinates
      }
    }

    # Store the nearest point as an 'sf' point
    nearest_points[i, ] <- nearest_point
    min_distances[i,]<-min_distance
  }
  return (list("nearest_points"=nearest_points,"min_distances"=min_distances))
}

narwhal_positions=dataprep[,c("x","y")]

#nearest points in coastline and land

result=system.time({

res_land=nearest_points(positions=narwhal_positions,coastline=land_utm)
res_coastline=nearest_points(positions=narwhal_positions,coastline=coastline_utm)
})

```

```{r computation time}
cat("Elapsed time:", result[3]/60, "minutes\n")
```

## Compare computed distances to shore

```{r compare distances to shore}
plot(res_land$min_distances/1000-dataprep$DistanceShore)
plot(res_coastline$min_distances/1000-dataprep$DistanceShore)
```
```{r summary distances to shore}
summary(res_land$min_distances)
summary(res_coastline$min_distances)
summary(dataprep$DistanceShore*1000)
```
Coastline distances are a bit closer to original distances in the dataset.
Anyway, we shall keep distance computed from the land in the sequel, since the land polygons will be the reference domain
for constrained movement of the narwhals.

```{r histograms distances to shore}
histo1=ggplot(dataprep, aes(x = DistanceShore*1000)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (m)")

df=data.frame(DistanceShore=res_coastline$min_distances)
histo2=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (m)")

df=data.frame(DistanceShore=res_land$min_distances)
histo3=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (m)")
```

```{r display histograms}
histo1
histo2
histo3
```

```{=tex}
There are pretty big differences between distances calculates with the coastline data and the land data.

For the rest of the analysis, we'll keep the bounded domain defined by the polygons in \texttt{land\_utm} sf object as a reference.
Doing that involves having some points on land, that is points with Distance to shore equal to 0.
This should be essentially due to the error in the land polygons than to the error in GPS measurements.
```

## Points on land

For coastline, distance is 0 only if the position of the narwhal is exactly on the coastline or on a small island. A proper "land domain" is not defined in this case, but only the boundary of the domain (the coastline).

```{r percentages of points on land}
cat("Percentage of points on land with the original distance to shore : \n",
    length(dataprep[dataprep$DistanceShore==0,"time"])/length(dataprep$time)*100,"\n")

cat("Percentage of points on land with distance from land : \n",
    length(which(res_land$min_distances/1000==0))/length(dataprep$time)*100,"\n")

cat("Percentage of points on land with distance from coastline : \n",
    length(which(res_coastline$min_distances/1000==0))/length(dataprep$time)*100)
```

## Check points on land on a map

Some points that appear on land on the map don't have distance to shore 0. This may be due to the map inaccuracy.
If we compute the distance to shore based on the map, as done in res_land, we get more points on the land.

```{r check original points on land}
check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(check_plot1)
```

```{r inspect points on land}
dataprep[dataprep$DistanceShore==0,c("Depth","time")]
```

```{r check new points on land}
df=dataprep
df$DistanceShore=res_land$min_distances

check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(check_plot2)
```

```{r percentage points on land per ID}

#percentage of points on land per ID
table(df[df$DistanceShore==0,"ID"])/table(dataprep$ID)*100

```

## Points too close to land

We expect the narwhals to stay at a certain distance from land, say 50 meters.
Let's check the distribution of points near land to choose a minimum value for the distance to shore below which we should drop the points.

```{r}

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<100 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (m)")
```

```{r}
histo
```

This seems to be pretty uniformly distributed, so let's choose arbitrarily a threshold of 50 metres.

# Third preprocessing step

##  Add distance and coordinates of the nearest shore point in the dataframe

```{r add shore points to the df}

#keep points calculated from land polygons
nearest_points=res_land$nearest_points
min_distances=res_land$min_distances

#add columns px and py with coordinates of the nearest point on the shore line
dataprep$px=nearest_points[,1]
dataprep$py=nearest_points[,2]

# add columns with difference between narwhal position and nearest point on the shoreline
dataprep$nx=dataprep$x-dataprep$px
dataprep$ny=dataprep$y-dataprep$py

#replace distance to shore with distances calculated from land polygons (in km)
dataprep$DistanceShore=min_distances/1000

```

# Add shore exposure covariate

```{r}
#add ExpShore covariate based on 1/distance in km
dataprep$ExpShore <- ExposureFunction(X = dataprep$DistanceShore)

```

# Convert all UTM coordinates in km

```{r convert to km}

dataprep$px=dataprep$px/1000
dataprep$py=dataprep$py/1000
dataprep$nx=dataprep$nx/1000
dataprep$ny=dataprep$ny/1000

dataprep$x=dataprep$x/1000
dataprep$y=dataprep$y/1000

```

# Some quantities of interest : empirical velocities and deviation angles

## Compute deviation angles related to the shore

```{r empirical velocities and deviation angles, cache=TRUE}

signed_angle <- function(u, v) {
    #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
    # as in 
    # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
    u <- matrix(u, ncol = 2)
    v <- matrix(v, ncol = 2)
    if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
                                  rows")
    result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
    ind1 <- which(result > pi)
    ind2 <- which(result <= -pi)
    result[ind1] <- result[ind1] - 2*pi
    result[ind2] <- result[ind2] + 2*pi
    return(result) 
}

#observed deviation angles
n=length(dataprep$time)

dataprep$AngleShore=rep(NA,n)
dataprep$AngleVelocities=rep(NA,n)
dataprep$AngleNormal=rep(NA,n)
dataprep$VelocityPhase=rep(NA,n)

for (id in unique(dataprep$ID)) {

  #filter data with specific id
  sub_ind=(dataprep$ID==id)
  sub_data=dataprep[sub_ind,]
  n_sub=length(sub_data$time)

  #time steps
  dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

  #step lengths
  dx=sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"]
  dy=sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"]

  #matrix of empirical velocity
  vexp_df=cbind(dx/dtimes,dy/dtimes)

  #matrix of normal vectors
  normal=as.matrix(sub_data[2:n_sub,c("nx","ny")])

  #angle between velocity and normal vector
  theta_coast=signed_angle(vexp_df,normal)

  #deviation angle between consecutive velocities
  theta_vel=signed_angle(vexp_df[1:(n_sub-2),],vexp_df[2:(n_sub-1),])

  #velocity phase angle
  vel_phase=signed_angle(matrix(rep(c(1,0),each=n_sub-2),ncol=2),vexp_df[1:(n_sub-2),])

  #deviation from trajectory parallel to coastline
  phi=ifelse(theta_coast > 0, theta_coast - pi/2, theta_coast + pi/2)

  #adjust lengths by adding NA when necessary
  theta_vel=c(NA,theta_vel,NA)
  phi=c(NA,phi)
  theta=c(NA,theta_coast)
  vel_phase=c(NA,vel_phase,NA)

  dataprep[sub_ind,"AngleShore"]=phi
  dataprep[sub_ind,"AngleVelocities"]=theta_vel
  dataprep[sub_ind,"AngleNormal"]=-theta
  dataprep[sub_ind,"VelocityPhase"]=vel_phase
}

```

## Plot empirical velocities and deviation angles through time

```{=tex}
$9 \%$ of the points are on land.
If the points are too close to then land (let's say distance to shore < 50m), the covariates AngleShore and Angle Normal are likely to be irrelevant.
```

```{r plot empirical velocities and deviation angles through time}

#highlight regions when the narwhals are exposed to the ship
#get starting times of experiment
start_times <- dataprep %>%
  filter((Type == "intertrial" | Type == "trial") & lag(is.na(Type))) %>% 
  select(time,ID)%>% rename(start_time = time)
#get ending times of experiments
end_times <- dataprep %>%
  filter((Type == "intertrial" | Type == "trial") & lead(is.na(Type))) %>% 
  select(time)%>% rename(end_time = time)
#dataframe with time period when the narwhals are exposed
regions=cbind(start_times,end_times)

#plot with highlight on the periods when the narwhals are exposed
plot_speed=ggplot()+ geom_line(data=dataprep,aes(time,horizontal_speed,col=ID))+ 
  geom_rect(data = regions, 
            aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),
            alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical velocity")

plot_AngleShore=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleShore,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")

plot_AngleVelocities=ggplot()+geom_line(data=dataprep,aes(time,AngleVelocities,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation between consecutive velocities")

plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleNormal,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Angle between empirical velocity and normal vector")

ggsave(path=plots_folder,"plot_horizontal_speed.png", plot = plot_speed)
ggsave(path=plots_folder,"plot_AngleShore.png", plot = plot_AngleShore)
ggsave(path=plots_folder,"plot_AngleVelocities.png", plot = plot_AngleVelocities)
ggsave(path=plots_folder,"plot_AngleNormal.png", plot = plot_AngleNormal)

```

```{r display plot_vexp , fig.cap="",warning=FALSE}
plot_speed
```

```{r display plot_AngleShore , fig.cap="",warning=FALSE}
plot_AngleShore
plot_AngleVelocities
```

```{r display plot_AngleVelocities , fig.cap="",warning=FALSE}
plot_AngleVelocities
```

## Histograms deviation angles

```{r histogram deviation angles}

AngleShore_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleShore,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")

AngleVelocities_histo=ggplot(dataprep, aes(x = AngleVelocities,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation between consecutive velocities (rad) ")

AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleNormal,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Angle between empirical velocity and normal vector (rad) ")

VelocityPhase_histo=ggplot(dataprep, aes(x = VelocityPhase,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Velocity phase angele (rad) ")

ggsave(path=plots_folder,"AngleShore_histo.png", plot = AngleShore_histo)
ggsave(path=plots_folder,"AngleVelocities.png", plot = AngleVelocities_histo)
ggsave(path=plots_folder,"AngleNormal.png", plot = AngleNormal_histo)
ggsave(path=plots_folder,"VelocityPhase.png", plot = VelocityPhase_histo)
```

```{r display AngleShore_histo, fig.cap="",warning=FALSE}
AngleShore_histo
```

```{r display AngleVelocities_histo, fig.cap="",warning=FALSE}
AngleVelocities_histo
```

```{r, display AngleNormal_histo,warning=FALSE}
AngleNormal_histo
```

```{r, display VelocityPhase_histo,warning=FALSE}
VelocityPhase_histo
```

## Empirical velocity and deviation angle depending on ExpShip

```{r plot empirical velocity and deviation angle depending on ExpSHip }

plot_AngleVelocitiesExp=ggplot()+geom_point(data=dataprep,aes(ExpShip,AngleVelocities,col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Exposure to ship (km)")+ylab("Empirical deviation between consecutive velocities")

ggsave(path=plots_folder,"plot_AngleVelocities_ExpShip.png", plot = plot_AngleVelocitiesExp)

plot_speedExp=ggplot()+geom_point(data=dataprep,aes(ExpShip,horizontal_speed,col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Exposure to ship (km)")+ylab("Empirical velocity")

ggsave(path=plots_folder,"plot_speed_ExpShip.png", plot = plot_speedExp)
```

```{r display plot_AngleVelocitiesExp, fig.cap=""}
plot_AngleVelocitiesExp
```

```{r display plot_vexpExp, fig.cap=""}
plot_speedExp
```

No clear pattern can be identified from these plots.

## Empirical velocity and deviation angles depending on Distance to shore

```{r plot empirical velocity and deviation angle depending on DistanceShore }

plot_AngleShoreDistance=ggplot()+geom_point(data=dataprep,aes(DistanceShore,AngleShore,col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation from coastline")

ggsave(path=plots_folder,"plot_AngleShore_DistanceShore.png", plot = plot_AngleShoreDistance)

plot_AngleVelocitiesDistance=ggplot()+geom_point(data=dataprep,aes(DistanceShore,AngleVelocities,col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation between consecutive velocities")

ggsave(path=plots_folder,"plot_AngleVelocities_DistanceShore.png", plot = plot_AngleVelocitiesDistance)

plot_speedDistance=ggplot()+geom_point(data=dataprep,aes(DistanceShore,horizontal_speed,col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical velocity")

ggsave(path=plots_folder,"plot_speed_DistanceShore.png", plot = plot_speedDistance)
```

```{r display plot_AngleShoreDistance, fig.cap=""}
plot_AngleShoreDistance
```

```{r display plot_AngleVelocitiesDistance, fig.cap=""}
plot_AngleVelocitiesDistance
```

```{r display plot_vexpDistance, fig.cap=""}
plot_speedDistance
```

## Deviation angles and velocity as a smooth function of DistanceShore

```{=tex}
We are using \texttt{geom\_smooth} function to get a "smooth" version of the previous plots. The smooth approximation is computed with \texttt{gam} package and cubic spline basis functions.
```

```{r deviation angles and velocity as a smooth function of DistanceShore}

# Plot smoothed AngleSHore
plot_AngleShoreDistanceSmoothed <-  ggplot()+ geom_smooth(data=dataprep,aes(DistanceShore,abs(AngleShore),col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation from coastline")

# Plot smoothed AngleVelocities
plot_AngleVelocitiesDistanceSmoothed <-  ggplot()+ geom_smooth(data=dataprep,aes(DistanceShore,abs(AngleVelocities),col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation between consecutive velocities")

#Plot smoothed speed
plot_speedDistanceSmoothed <- ggplot() +geom_smooth(data=dataprep,aes(DistanceShore,horizontal_speed,col=ID))+
  facet_grid(ID ~ ., labeller = as_labeller(c("Asgeir" = "A1", "Frederik" = "A2", "Helge18" = "A3", "Kyrri" = "A4", "Nemo" = "A5", "Siggi" = "A6"))) +
  scale_color_viridis_d(guide = "none") +
  theme(legend.position = "none") +
  xlab("Distance to shore (km)") +
  ylab("Empirical velocity")
```

```{r display plot_AngleShoreDistanceSmoothed, fig.cap=""}
plot_AngleShoreDistanceSmoothed
```

```{r display plot_AngleVelocitiesDistanceSmoothed, fig.cap=""}
plot_AngleVelocitiesDistanceSmoothed
```

```{r display plot_vexpDistanceSmoothed, fig.cap=""}
plot_speedDistanceSmoothed
```

```{=tex}
The increase in deviation angle from coastline for small values of Distance to shore might indicate a rotation of the narwhals at a certain threshold. However, it may also be due to the points on land ($2 \%$ of the data), that is the points for which \texttt{DistanceShore} is $0$. This can be checked by removing these points from the plot.
```

## Same plots without points on land

```{r plot empirical velocities and deviation angles without points on land}
# Plot smoothed AngleSHore
plot_AngleShoreDistanceSmoothed <-  ggplot()+
  geom_smooth(data=dataprep[dataprep$DistanceShore>0,],aes(DistanceShore,abs(AngleShore),col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="none")+xlab("Distance to shore (km)")+
  ylab("Empirical deviation from coastline")

# Plot smoothed AngleVelocities
plot_AngleVelocitiesDistanceSmoothed <-  ggplot()+
  geom_smooth(data=dataprep[dataprep$DistanceShore>0,],aes(DistanceShore,abs(AngleVelocities),col=ID))+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+
  ylab("Empirical deviation between consecutive velocities")

#Plot smoothed vexp
plot_speedDistanceSmoothed <- ggplot() +geom_smooth(data=dataprep[dataprep$DistanceShore>0,],aes(DistanceShore,horizontal_speed,col=ID))+
  facet_grid(ID ~ ., labeller = as_labeller(c("Asgeir" = "A1", "Frederik" = "A2", "Helge18" = "A3", "Kyrri" = "A4", "Nemo" = "A5", "Siggi" = "A6"))) +
  scale_color_viridis_d(guide = "none") +
  theme(legend.position = "none") +
  xlab("Distance to shore (km)") +
  ylab("Empirical velocity")

```

```{r display plot_AngleShoreDistanceSmoothed without points on land, fig.cap=""}
plot_AngleShoreDistanceSmoothed
```

```{r display plot_AngleVelocitiesDistanceSmoothed without points on land, fig.cap=""}
plot_AngleVelocitiesDistanceSmoothed
```

```{r display plot_vexpDistanceSmoothed without points on land, fig.cap=""}
plot_speedDistanceSmoothed
```

```{=tex}
It turns out that the effect was due to points on land. 
We can now observe a slight decrease in the angle of deviation from the coastline, which may indicate that narwhals tend to move parallel to the coast when close enough.
However, no effects of Distance to shore is observed on the empirical deviation between consecutive velocities.
This may be due to the averaging that is made to get those plots. The narwhal only turns for a few positions and then keep its heading paralell to the shore.
```

# Visualize individual tracks

## Plot individual tracks on the map

```{r}

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
 mapID=milne_land_map + 
  geom_point(sub_data[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = time),size= 1,alpha=0.5) +
  scale_color_viridis_c()

  ggsave(path=plots_folder,paste(id,"map_track.png",sep="_"), 
         plot =mapID,width=10,height=5)

}

```

## Plot each individual track along with nearest shore points

The UTM coordinates for the coastlines are in meters, so that we need to multiply by 1000 when plotting the nearest shore points.

```{r individual tracks}

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(1000*px,1000*py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(1000*x,1000*y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(1000*x,1000*y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(1000*x,1000*y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

  ggsave(path=plots_folder,paste(id,"track.png",sep="_"), 
         plot =id_track)

}

```

```{r display one track, fig.cap="Example of individual track with computed nearest shore points"}
id_track
```

## Plot distance to shore histogram for each narwhal

```{r distance shore histogram}
Dshore_histo=ggplot(dataprep, aes(x = DistanceShore)) +
  geom_histogram(fill="white",color="black",bins=50)+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Distance to shore (km)")

ggsave(path=plots_folder,"Dshore_histo.png", plot = Dshore_histo)
```

```{r display Dshore_histo, fig.cap="Histograms of distance to shore"}
Dshore_histo
```
## Plot distance to ship histogram for each narwhal

We also plot the density of the Dist_Ship variable to understand better its distribution.

```{r distance ship histogram}

Dship_histo=ggplot(dataprep, aes(x = Dist_Ship)) +
  geom_histogram(fill="white",col="black",bins=30)+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Distance to ship (km)")

ggsave(path=plots_folder,"Dship_histo.png", plot = Dship_histo)

```

```{r display Dship_histo, fig.cap="Histograms of distance to ship"}
Dship_histo
```
# Further details

## Plot ship exposure without interpolated points

```{r plot real exposure}
pExpShip=ggplot(dataprep,aes(time,ExpShip,col=ID))+geom_line(color="black")+
  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")

ggsave(path=plots_folder,"realExpShip_through_time.png", plot=pExpShip)
```

```{r display pExp, fig.cap="Measured exposure to ship through time"}
pExpShip
```
## Plot shore exposure

```{r}
pExpShore=ggplot(dataprep[dataprep$DistanceShore>0.05,],aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")

ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExp)
```

```{r}
pExpShore
```

# Tracks before and after exposure to ship

## Split between tracks before and after exposure

```{r split tracks}

tagtime=24

dataBE1=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > maxtime) & (time>tagtime)))
}

dataBE2=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > maxtime) & (time>tagtime/2)))
}

dataAE=data.frame()
for (id in unique(dataprep$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataAE =rbind(dataAE,subset(narwhalData, (time > maxtime) & (time>tagtime)))
}

```

```{r}
#number of data points for each individual
for (id in unique(data$ID)) {
  narwhalDataBE=dataBE1[dataBE1$ID==id,]
  narwhalDataAE=dataAE[dataAE$ID==id,]
  cat(id," : ", length(narwhalDataBE$time)," GPS measurement before exposure and ", length(narwhalDataAE$time)," after exposure.\n",sep="")
}
```

## Plot data before and after exposure

```{r plots before and after exposure}
pBE1=milne_land_map + geom_point(dataBE1, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pBE2=milne_land_map + geom_point(dataBE2, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pAE=milne_land_map + geom_point(dataAE, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

ggsave(path=plots_folder,"tracksBE1.png",pBE1,width=10,height=5)
ggsave(path=plots_folder,"tracksBE2.png",pBE2,width=10,height=5)
ggsave(path=plots_folder,"tracksAE.png",pAE,width=10,height=5)
```

```{r display pBE1, fig.cap="Tracks before exposure, 24h after tagging"}
pBE1
```

```{r display pBE2, fig.cap="Tracks before exposure, 12h after tagging"}
pBE2
```

```{r display pAE, fig.cap="Tracks after exposure"}
pAE
```

# Save preprocessed data frames as csv files for further smoothSDE analysis

```{r save preprocessed data}

write.csv(dataprep,file.path(narwhal_data_path,"allData.csv"))
write.csv(dataBE1,file.path(narwhal_data_path,"DataBE1.csv"))
write.csv(dataBE2,file.path(narwhal_data_path,"DataBE2.csv"))
write.csv(dataAE,file.path(narwhal_data_path,"DataAE.csv"))
```

### 2024-04-05 R/simulation_study/simulation_study_fjords.Rmd
Add R code for data analysis, plots and sde fitting

---
title: "simulation study in fjords"
output: html_document
date: "2024-03-13"
---

# Import libraries and data

```{r import libraries}
#seed for reproducibility 
set.seed(1)

#for tex code
library(tinytex)

#for tex expression in plot legends
library(latex2exp)

#for pipeline operator
library(dplyr)

#plots
library(ggplot2)
library(plotly)
library(htmlwidgets)

#data managing
library(tidyr)

#progress bar
library(progress)

#for land polygons
library(sf)

#SDE fit
library(smoothSDE)

#auxiliary function to simulate from RACVM with specified smooth function for omega
source("CVM_functions.R")
```

# Get land polygons

```{r get land polygons}
#land polygons

# Set the path to the directory containing the data
par_dir=dirname(dirname(getwd())) #parent directory 
greenland_data_path <- file.path(par_dir,"Data", "Greenland")

#get the coastline geometry from the geojson file
land<-st_read(file.path(greenland_data_path,"filtered_land_utm.shp"))
```

# Sample initial positions uniformly

```{r sample initial positions}

#number of tracks
n_samples=10

#initial velocity and location
v0=c(0,0)

#generate random initial points
x0=matrix(rep(NA,n_samples*2),ncol=2)
colnames(x0)=c("x1","x2")

i=1
while (i<=n_samples) {
  #choose location uniformly in the map
  x=c(runif(1,min=430,max=500),runif(1,min=7760,max=7900))
  p=nearest_shore_point(st_point(x),land)
  Dshore=(x[1]-p[1])^2+(x[2]-p[2])^2
  #keep it as initial position if it is at least 50 metres away from the shore
  if (Dshore>0.5) {
    x0[i,]=x
    i=i+1
  }
}
```

## Fix time steps for observations

```{r fix time steps}
#observation times
Tmax=3*24 #max time in hour

hf=1/60 #1 minute time step
lf=5/60 #5 minute time step

#high frequency time steps
times_hf=seq(0,Tmax,by=hf)
n_hf=length(times_hf)-1

#low frequency time step
times_lf=seq(0,Tmax,by=lf)
n_lf=length(times_lf)-1
```

## Set observation error

```{r set observation error}
#no measurement error (we cannot put 0 because only the log of this quantity is used)
sigma_obs_0=0.003
H0_lf=array(rep(sigma_obs_0^2*diag(2),n_lf),dim=c(2,2,n_lf))
H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
#GPS measurement error 
sigma_obs_1=0.03
H1_lf=array(rep(sigma_obs_1^2*diag(2),n_lf),dim=c(2,2,n_lf))
H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf),dim=c(2,2,n_hf))

```

## Fix smooth parameters

```{r define smooth parameters}

ftau=function(Dshore,theta,D0=0.3,delta0=0.2,tau0=1,tau1=2,tau2=2,lambda=4,kappa=0.1) {
  coeff=exp(-kappa*(Dshore/D0)^2)
  tau=tau0*(1-coeff)+(delta0+(tau0-delta0)*exp(-theta^2/pi^2)+(tau1-delta0)*exp(-lambda*(theta-pi/2)^2)+(tau2-delta0)*exp(-lambda*(theta+pi/2)^2))*coeff
  return (tau)
}

#mixture of independent gaussian
ftau_gauss=function(Dshore,theta,D0=0.3,tau0=1,tau1=3) {
  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/10))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/10))^2-(Dshore/D0)^2)
  return (tau)
}

#constant tau
ftau_constant=function(Dshore,theta,tau=1) {
  return (tau)
}

fomega=function(Dshore,theta,D0=0.3,omega0=60*pi/2,lambda=2,kappa=0.2) {
    coeff=exp(-kappa*(Dshore/D0)^2)
    omega=omega0/2*(tanh(lambda*(theta+pi/2-atanh(-0.9)/lambda))+tanh(lambda*(theta-(pi/2-atanh(-0.9)/lambda))))*coeff
    return(omega)
}

nu=4

```

Plot smooth functions.
```{r plot smooth parameters }

Dshore=seq(from=0.1,to=2,length.out=30)
theta=seq(from=-pi,to=pi,length.out=30)
grid<- expand.grid(Dshore,theta)
colnames(grid)=c("DistanceShore","theta")
true_par=grid
true_par$tau=ftau_gauss(grid$DistanceShore,grid$theta)
true_par$omega=fomega(grid$DistanceShore,grid$theta)

plot_true_tau <- plot_ly(true_par,type = "mesh3d",
                         x = ~theta,y = ~DistanceShore,z=~tau,intensity=~tau,
                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                layout(title=paste("True tau"),scene=list(xaxis = list(title =  "Theta",showgrid = F),
                                                                 yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "tau")))

plot_true_omega <- plot_ly(true_par,type = "mesh3d",
                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
                                                                 yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_true_tau
plot_true_omega
```

## Estimation when omega only is smooth

### Generate samples : high frequency and no measurement error

```{r high frequency samples no error (hf ne)}

#initialize data frame
data_fjords_hf_ne=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)

for (i in 1:n_samples) {

  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=land,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  data_fjords_hf_ne=rbind(data_fjords_hf_ne,cbind(data_sim,data_shore))
}

data_fjords_hf_ne=data_fjords_hf_ne[-1,]

```

## percentage of samples that reached land

```{r percentage of hf ne samples that reached land}
count=0
for (id in unique(data_fjords_hf_ne$ID)){
  sub_data=data_fjords_hf_ne[data_fjords_hf_ne$ID==id,]
  if (nrow(sub_data) < n_hf) {
    print(id)
    count=count+1
  }
}
cat(count/n_samples*100,"percent of the samples reached land")
```

### Plot samples

```{r plot hf ne samples}

plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(y1,y2),size=0.1)+
  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(p1,p2),size=0.01,col="red",shape=5)+
  geom_path(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(y1,y2),linewidth=0.1)+
  geom_point(data = data_fjords_hf_ne[data_fjords_hf_ne$ID==5,] %>% filter(!duplicated(ID)),
             aes(x = y1, y = y2), shape = 3, size = 4,col = "blue")+
  xlab("x") + ylab("y")

ggsave(filename="samples_fjords_hf_ne_omega.png",plot=plot_samples_fjords_hf_ne)
ggplotly(plot_samples_fjords_hf_ne)
```

### Generate samples : low frequency and no measurement error

```{r subsample hf ne samples to low frequency (lf ne)}
data_fjords_lf_ne=data_fjords_hf_ne[seq(1,nrow(data_fjords_hf_ne),by=5),]
```

### Plot samples

```{r plot lf ne samples}

plot_samples_fjords_lf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*p1,1000*p2),size=0.01,col="red",shape=5)+
  geom_path(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = data_fjords_lf_ne[data_fjords_lf_ne$ID==6,] %>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "blue")+
  xlab("x") + ylab("y")

ggsave(filename="samples_fjords_lf_ne_omega.png",plot=plot_samples_fjords_lf_ne)
ggplotly(plot_samples_fjords_lf_ne)
```

### Generate samples : high frequency and measurement error

```{r generate high frequency samples with error (hf e)}

#initialize data frame
data_fjords_hf_e=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)

for (i in 1:n_samples) {

  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=land,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  data_fjords_hf_e=rbind(data_fjords_hf_e,cbind(data_sim,data_shore))
}

data_fjords_hf_e=data_fjords_hf_e[-1,]

```

```{r percentage of hf e samples that reached land}
count=0
for (id in unique(data_fjords_hf_e$ID)){
  sub_data=data_fjords_hf_e[data_fjords_hf_e$ID==id,]
  if (nrow(sub_data) < n_hf) {
    print(id)
    count=count+1
  }
}
cat(count/n_samples*100,"percent of the samples reached land")
```
### Plot samples

```{r plot hf e samples}

plot_samples_fjords_hf_e=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_fjords_hf_e[data_fjords_hf_e$ID==8,],mapping=aes(y1,y2),size=0.1)+
   geom_path(data=data_fjords_hf_e[data_fjords_hf_e$ID==8,],mapping=aes(y1,y2),linewidth=0.1)+
  geom_point(data = data_fjords_hf_e[data_fjords_hf_e$ID==8,] %>% filter(!duplicated(ID)),
             aes(x = y1, y = y2 ), shape = 3, size = 4,col = "red")+
    xlab("x") + ylab("y")

ggsave(filename="samples_fjords_hf_e_omega.png",plot=plot_samples_fjords_hf_e)
ggplotly(plot_samples_fjords_hf_e)
```

### Generate samples : low frequency and measurement error

```{r subsamble hf e samples to low frequency (lf e)}

data_fjords_lf_e=data_fjords_hf_e[seq(1,nrow(data_fjords_hf_e),by=5),]

```

### Plot samples

```{r plot lf e samples}

plot_samples_fjords_lf_e=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_fjords_lf_e[data_fjords_lf_e$ID==8,],mapping=aes(1000*y1,1000*y2),size=0.1)+
   geom_path(data=data_fjords_lf_e[data_fjords_lf_e$ID==8,],mapping=aes(1000*y1,1000*y2),linewidth=0.1)+
  geom_point(data = data_fjords_lf_e[data_fjords_lf_e$ID==8,] %>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4,col = "red")+
  xlab("x") + ylab("y")

ggsave(filename="samples_fjords_lf_e_omega.png",plot=plot_samples_fjords_lf_e)
ggplotly(plot_samples_fjords_lf_e)
```

### Compute distance to shore and empirical deviation angles

```{r compute covariates}

signed_angle <- function(u, v) {
    #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
    # as in 
    # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
    u <- matrix(u, ncol = 2)
    v <- matrix(v, ncol = 2)
    if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
                                  rows")
    result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
    ind1 <- which(result > pi)
    ind2 <- which(result <= -pi)
    result[ind1] <- result[ind1] - 2*pi
    result[ind2] <- result[ind2] + 2*pi
    return(result) 
}

add_covs=function(data) {
  #data has at least columns "time","y1" "y2", "p1","p2"
  #this function returns a dataframe with three mùore columns : "theta", "DistanceShore" and "ExpShore"

  new_data=data
  #observed deviation angles
  n=length(new_data$time)

  new_data$theta=rep(NA,n)

  for (id in unique(new_data$ID)) {

    #filter data with specific id
    sub_ind=(new_data$ID==id)
    sub_data=new_data[sub_ind,]
    n_sub=length(sub_data$time)

    #time steps
    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

    #step lengths
    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]

    #matrix of empirical velocity
    vexp_df=cbind(dx/dtimes,dy/dtimes)

    #matrix of normal vectors
    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])

    #angle between velocity and normal vector
    theta_coast=signed_angle(normal,vexp_df)

    #adjust lengths 
    theta_coast=c(theta_coast,1)

    new_data[sub_ind,"theta"]=theta_coast
  }

  #distance to shore
  Dshore=sqrt((new_data$y1-new_data$p1)^2+(new_data$y2-new_data$p2)^2)
  new_data$DistanceShore=Dshore
  new_data$ExpShore=1/Dshore

  return(new_data)
}

data_fjords_hf_ne=add_covs(data_fjords_hf_ne)
data_fjords_lf_ne=add_covs(data_fjords_lf_ne)

data_fjords_hf_e=add_covs(data_fjords_hf_e)
data_fjords_lf_e=add_covs(data_fjords_lf_e)
```

### Estimate from high frequency data without error : ExpShore and tensor splines ti

```{r estimate from hf e samples with cov ExpShore}

#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_fjords_hf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[data_fjords_hf_ne$ID==6,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H0_hf))
crcvm_fjords_hf_ne1$fit(method="BFGS")
```

```{r get plots from hf e estimates with Expshore}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
res=crcvm_fjords_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
res$plot_omega_theta_ExpShore
```

### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.

```{r compare hf e estimates with Expshore to true smooth parameters for fixed values of ExpShore}

#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/1,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))

sde_par_close <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(1,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 1 km")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 3 km")

plot_omega_close
plot_omega_far

ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_close.png",plot=plot_omega_close)
ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_far.png",plot=plot_omega_far)

```

### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.

```{r compare hf e estimates with Expshore to true smooth parameters for fixed values of Theta}

#plot spline estimates of the parameter omega close and far to the shore

predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))

sde_par_away <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_away)
sde_CI_away <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_away)
sde_par_toward<- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_toward)
sde_CI_toward <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_toward)

# Data frame for point estimates
sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])

# Data frame for CIs
sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
                          lowomega = sde_CI_away["omega", "low",],
                          upomega = sde_CI_away["omega", "upp",])
sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
                          lowomega = sde_CI_toward["omega", "low",],
                          upomega = sde_CI_toward["omega", "upp",])

omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))

plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))

plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))

plot_omega_away
plot_omega_toward

ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_away.png",plot=plot_omega_away)
ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_toward.png",plot=plot_omega_toward)

```

### Estimation from low frequency data without error

```{r estimate from lf ne samples with cov ExpShore}

#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_fjords_lf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("log_sigmaobs0"=log(0.003)))
crcvm_fjords_lf_ne1$fit(method="BFGS")
```

```{r get plots from lf ne estimates with Expshore}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
res=crcvm_fjords_lf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_lf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
res$plot_omega_theta_DistanceShore
```

### Estimate from high frequency data with error : ExpShore and tensor splines te

```{r estimate from hf e samples with cov ExpShore}

#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_fjords_hf_e1 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==8,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_hf))
crcvm_fjords_hf_e1$fit(method="BFGS")
```

```{r get plots from hf e estimates with Expshore}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
res=crcvm_fjords_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
res$plot_omega_theta_ExpShore
```

### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))

sde_par_close <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 0.5 km")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 3 km")

plot_omega_close
plot_omega_far

ggsave(filename="crcvm_fjords_hf_e1/crcvm_fjords_hf_e1_omega_close.png",plot=plot_omega_close)
ggsave(filename="crcvm_fjords_hf_e1/crcvm_fjords_hf_e1_omega_far.png",plot=plot_omega_far)

```

### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))

sde_par_away <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_away)
sde_CI_away <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_away)
sde_par_toward<- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_toward)
sde_CI_toward <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_toward)

# Data frame for point estimates
sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])

# Data frame for CIs
sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
                          lowomega = sde_CI_away["omega", "low",],
                          upomega = sde_CI_away["omega", "upp",])
sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
                          lowomega = sde_CI_toward["omega", "low",],
                          upomega = sde_CI_toward["omega", "upp",])

omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))

plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))

plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))

plot_omega_away
plot_omega_toward

ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_away.png",plot=plot_omega_away)
ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_toward.png",plot=plot_omega_toward)

```

### Estimate from low frequency data with error : ExpShore and tensor splines ti

```{r}

#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_fjords_lf_e1 <- SDE$new(formulas = formulas,data = data_fjords_lf_e[data_fjords_lf_e$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_lf))
crcvm_fjords_lf_e1$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.2)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
crcvm_fjords_lf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_lf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)

```

### 2024-04-05 R/simulation_study/simulation_study_rect.Rmd
Add R code for data analysis, plots and sde fitting

---
title: "Simulation study in rectangular domain"
output: html_notebook
---

# Import libraries and data

```{r}
#seed for reproducibility 
set.seed(1)

#for tex code
library(tinytex)

#for tex expression in plot legends
library(latex2exp)

#for pipeline operator
library(dplyr)

#plots
library(ggplot2)
library(plotly)
library(htmlwidgets)

#data managing
library(tidyr)

#progress bar
library(progress)

#for land polygons
library(sf)

#SDE fit
library(smoothSDE)

#some auxiliary functions
source(file.path("/home","delporta","Documents","Recherche","Codes","CVM","CVM_functions.R"))
```

# Create the rectangle

We create a square with sides of length 8km.
```{r}

#rectangular domain 
large_rectangle_coords <- matrix(c(0, 0, 0, 10000, 10000, 10000, 10000, 0, 0, 0), ncol = 2, byrow = TRUE)
small_rectangle_coords <- matrix(c(1000,1000, 1000, 9000, 9000, 9000, 9000, 1000, 1000, 1000), ncol = 2, byrow = TRUE)

# Create an sf object representing the rectangles
large_rectangle <- st_sfc(st_polygon(list(large_rectangle_coords)))
large_rectangle<-st_sf(geometry = large_rectangle)

small_rectangle <- st_sfc(st_polygon(list(small_rectangle_coords)))
small_rectangle<-st_sf(geometry = small_rectangle)

#change its crs
domain<- st_sym_difference(large_rectangle, small_rectangle)

```

```{r}

plot_domain=ggplot()+geom_sf(data=domain$geometry,fill="blue")
plot_domain
```

# Sample initial positions uniformly in the domain

```{r}

#number of tracks
n_samples=100

#initial velocity and location
v0=c(0,0)

#generate random initial points
x0=matrix(rep(NA,n_samples*2),ncol=2)
colnames(x0)=c("x1","x2")

for (i in 1:n_samples) {
  #choose location uniformly in the domain
  x=c(runif(1,min=1.5,max=8.5),runif(1,min=1.5,max=8.5))
  x0[i,]=x
  }
```

# Fix time steps for observations

```{r}
#observation times
Tmax=3*24 #max time in hour

hf=1/60 #1 minute time step
lf=5/60 #5 minute time step

#high frequency time steps
times_hf=seq(0,Tmax,by=hf)
n_hf=length(times_hf)-1

#low frequency time step
times_lf=seq(0,Tmax,by=lf)
n_lf=length(times_lf)-1
```

# Set observation error

```{r}
#no measurement error (we cannot put 0 because only the log of this quantity is used)
sigma_obs_0=0.003
H0_lf=array(rep(sigma_obs_0^2*diag(2),n_lf),dim=c(2,2,n_lf))
H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
#GPS measurement error 
sigma_obs_1=0.03
H1_lf=array(rep(sigma_obs_1^2*diag(2),n_lf),dim=c(2,2,n_lf))
H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf),dim=c(2,2,n_hf))

```

# Smooth parameters of the CRCVM

Define smooth functions for tau and omega.

```{r}

#smooth parameters

ftau=function(Dshore,theta,D0=0.3,delta0=0.2,tau0=1,tau1=2,tau2=2,lambda=4,kappa=0.1) {
  coeff=exp(-kappa*(Dshore/D0)^2)
  tau=tau0*(1-coeff)+(delta0+(tau0-delta0)*exp(-theta^2/pi^2)+(tau1-delta0)*exp(-lambda*(theta-pi/2)^2)+(tau2-delta0)*exp(-lambda*(theta+pi/2)^2))*coeff
  return (tau)
}

#mixture of independent gaussian
ftau_gauss=function(Dshore,theta,D0=0.5,tau0=1,tau1=1.5) {
  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/10))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/10))^2-(Dshore/D0)^2)
  return (tau)
}

#constant tau
ftau_constant=function(Dshore,theta,tau=1) {
  return (tau)
}

fomega=function(Dshore,theta,D0=0.3,omega0=60*pi,lambda=1.5,kappa=0.5) {
    coeff=exp(-kappa*(Dshore/D0)^2)
    omega=omega0/2*(tanh(lambda*(theta+pi/2-atanh(-0.9)/lambda))+tanh(lambda*(theta-(pi/2-atanh(-0.9)/lambda))))*coeff
    return(omega)
}

nu=4
```

Plot smooth functions.
```{r}

Dshore=seq(from=0.3,to=2,length.out=30)
theta=seq(from=-pi,to=pi,length.out=30)
grid<- expand.grid(Dshore,theta)
colnames(grid)=c("DistanceShore","theta")
true_par=grid
true_par$tau=ftau_gauss(grid$DistanceShore,grid$theta)
true_par$omega=fomega(grid$DistanceShore,grid$theta)

plot_true_tau <- plot_ly(true_par,type = "mesh3d",
                         x = ~theta,y = ~DistanceShore,z=~tau,intensity=~tau,
                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                layout(title=paste("True tau"),scene=list(xaxis = list(title = "Theta",showgrid = F),
                                                                 yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "tau")))

plot_true_omega <- plot_ly(true_par,type = "mesh3d",
                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
                                                                 yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_true_tau
plot_true_omega
```

# Simulation and estimation when only omega is smooth (other parameters are fixed)

## Generate samples : no measurement errors and high frequency

```{r}

#initialize data frame
data_rect_hf_ne=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)

for (i in 1:n_samples) {

  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=domain,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  data_rect_hf_ne=rbind(data_rect_hf_ne,cbind(data_sim,data_shore))
}

#for the first value we cannot compute the angle theta
data_rect_hf_ne=data_rect_hf_ne[-1,]

```

## percentage of samples that reached land

```{r}
count=0
for (id in unique(data_rect_hf_ne$ID)){
  sub_data=data_rect_hf_ne[data_rect_hf_ne$ID==id,]
  if (nrow(sub_data) < n_hf) {
    print(id)
    count=count+1
  }
}
cat(count,"percent of the samples reached land")
```

## plot one sample

```{r}

plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_path(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = data_rect_hf_ne[data_rect_hf_ne$ID==1,]%>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
  xlab("x") + ylab("y")

ggsave(filename="samples_rect_hf_ne_omega.png",plot=plot_samples_rect_hf_ne)
ggplotly(plot_samples_rect_hf_ne)
```

## Subsample : no measurement errors and low frequency

```{r}

data_rect_lf_ne=data_rect_hf_ne[seq(1,n_hf,by=lf/hf),]

```

## plot subsamples

```{r}

plot_samples_rect_lf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_rect_lf_ne[data_rect_lf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_path(data=data_rect_lf_ne[data_rect_lf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = data_rect_lf_ne[data_rect_lf_ne$ID==1,]%>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
    xlab("x") + ylab("y")

ggsave(filename="samples_rect_lf_ne_omega.png",plot=plot_samples_rect_lf_ne)
ggplotly(plot_samples_rect_lf_ne)
```

## Generate samples : 30m measurement errors and high frequency

```{r}

#initialize data frame
data_rect_hf_e=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)

for (i in 1:n_samples) {

  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=domain,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  data_rect_hf_e=rbind(data_rect_hf_e,cbind(data_sim,data_shore))
}

#for the first value we cannot compute the angle theta
data_rect_hf_e=data_rect_hf_e[-1,]

```

## percentage of samples that reached land

```{r}
count=0
for (id in unique(data_rect_hf_ne$ID)){
  sub_data=data_rect_hf_e[data_rect_hf_e$ID==id,]
  if (nrow(sub_data) < n_hf) {
    print(id)
    count=count+1
  }
}
cat(count,"percent of the samples reached land")
```
## plot samples

```{r}

plot_samples_rect_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = data_rect_hf_e[data_rect_hf_e$ID==1,]%>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
    xlab("x") + ylab("y")

ggsave(filename="samples_rect_hf_e_omega.png",plot=plot_samples_rect_hf_e)
ggplotly(plot_samples_rect_hf_e)
```

## Subsample : measurement errors and low frequency

```{r}

data_rect_lf_e=data_rect_hf_e[seq(1,n_hf,by=lf/hf),]

```

## plot subsamples

```{r}

plot_samples_rect_lf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=data_rect_lf_e[data_rect_lf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = data_rect_hf_e[data_rect_hf_e$ID==1,]%>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
    xlab("x") + ylab("y")

ggsave(filename="samples_rect_lf_e_omega.png",plot=plot_samples_rect_lf_e)
ggplotly(plot_samples_rect_lf_e)
```
## Compute distance to shore and empirical deviation angles

```{r}

signed_angle <- function(u, v) {
    #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
    # as in 
    # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
    u <- matrix(u, ncol = 2)
    v <- matrix(v, ncol = 2)
    if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
                                  rows")
    result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
    ind1 <- which(result > pi)
    ind2 <- which(result <= -pi)
    result[ind1] <- result[ind1] - 2*pi
    result[ind2] <- result[ind2] + 2*pi
    return(result) 
}

add_covs=function(data) {
  #data has at least columns "time","y1" "y2", "p1","p2"
  #this function modifies data to add columns "theta", "DistanceShore" and "ExpShore"

  new_data=data
  #observed deviation angles
  n=length(new_data$time)

  new_data$theta=rep(NA,n)

  for (id in unique(new_data$ID)) {

    #filter data with specific id
    sub_ind=(new_data$ID==id)
    sub_data=new_data[sub_ind,]
    n_sub=length(sub_data$time)

    #time steps
    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

    #step lengths
    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]

    #matrix of empirical velocity
    vexp_df=cbind(dx/dtimes,dy/dtimes)

    #matrix of normal vectors
    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])

    #angle between velocity and normal vector
    theta_coast=signed_angle(normal,vexp_df)

    #adjust lengths 
    theta_coast=c(theta_coast,1)

    new_data[sub_ind,"theta"]=theta_coast
  }

  #distance to shore
  Dshore=sqrt((new_data$y1-new_data$p1)^2+(new_data$y2-new_data$p2)^2)
  new_data$DistanceShore=Dshore
  new_data$ExpShore=1/Dshore

  return(new_data)
}

data_rect_hf_ne=add_covs(data_rect_hf_ne)
data_rect_lf_ne=add_covs(data_rect_lf_ne)

data_rect_hf_e=add_covs(data_rect_hf_e)
data_rect_lf_e=add_covs(data_rect_lf_e)
```

## Estimate from high frequency data without error : ExpShore and tensor splines te

### Fit sde
```{r}

#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,ExpShore,k=c(10,5),bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1 & data_rect_hf_ne$time<2*24,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H0_hf))
crcvm_rect_hf_ne1$fit(method="BFGS")
```

### Plot smooth parameter omega in 3D

```{r}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
res=crcvm_rect_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
res$plot_omega_theta_ExpShore
```

### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))

sde_par_close <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 0.5 km")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 3 km")

plot_omega_close
plot_omega_far

ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_close.png",plot=plot_omega_close)
ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_far.png",plot=plot_omega_far)

```

### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))

sde_par_away <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_away)
sde_CI_away <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_away)
sde_par_toward<- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_toward)
sde_CI_toward <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_toward)

# Data frame for point estimates
sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])

# Data frame for CIs
sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
                          lowomega = sde_CI_away["omega", "low",],
                          upomega = sde_CI_away["omega", "upp",])
sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
                          lowomega = sde_CI_toward["omega", "low",],
                          upomega = sde_CI_toward["omega", "upp",])

omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))

plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))

plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))

plot_omega_away
plot_omega_toward

ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_away.png",plot=plot_omega_away)
ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_toward.png",plot=plot_omega_toward)

```

### Simulate from fitted model

```{r}

fExpShore=function(z,v,p) {

  #normal vector
  normal=c(z[1]-p[1],z[2]-p[2])

  #distance to shore
  Dshore=sqrt(normal[1]^2+normal[2]^2)
  if (Dshore>0.5){
    return (0)
  }
  else if (Dshore<0.01) {
    return (100)
  }
  else {
    return (1/Dshore)
  }
}

fangle=function(z,v,p) {

  #normal vector
  normal=c(z[1]-p[1],z[2]-p[2])

  #return angle
  return (signed_angle(normal,v))
}

atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
sim_data=crcvm_rect_hf_e1$simulate(z0=x0[1,],data=crcvm_rect_hf_e1$data(),atw=atw,land=domain)

```

```{r}

plot_sim=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
  geom_point(data=sim_data,mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_path(data=sim_data,mapping=aes(1000*y1,1000*y2),size=0.1)+
  geom_point(data = sim_data%>% filter(!duplicated(ID)),
             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
    xlab("x") + ylab("y")

ggplotly(plot_sim)
```

### S

## Estimate from high frequency data without error: DistanceShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H0_hf))
crcvm_rect_hf_ne2$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"DistanceShore"=0.3)
xmax=list("theta"=pi,"DistanceShore"=2)
res=crcvm_rect_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",npost=1000,level=0.95,
                          xmin=xmin,xmax=xmax)
res$plot_omega_theta_DistanceShore
```

Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.

```{r}
#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),DistanceShore=rep(0.3,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),DistanceShore=rep(2,100))

sde_par_close <- crcvm_rect_hf_ne2$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_rect_hf_ne2$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_rect_hf_ne2$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_rect_hf_ne2$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.3,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  xlab("Theta")+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  xlab("Theta")+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")

plot_omega_close
plot_omega_far
```

## Estimate from low frequency data without error : ExpShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_lf_ne1 <- SDE$new(formulas = formulas,data = data_rect_lf_ne[data_rect_lf_ne$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),other_data=list(H=H0_lf))
crcvm_rect_lf_ne1$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.1)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
crcvm_rect_lf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)

```

```{r}
#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.3,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))

sde_par_close <- crcvm_rect_lf_ne1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_rect_lf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_rect_lf_ne1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_rect_lf_ne1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.3,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  xlab("Theta")+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  xlab("Theta")+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")

plot_omega_close
plot_omega_far
```

## Estimate from low frequency data without error : DistanceShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_lf_ne2 <- SDE$new(formulas = formulas,data = data_rect_lf_ne[data_rect_lf_ne$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H0_lf))
crcvm_rect_lf_ne2$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"DistanceShore"=0.1)
xmax=list("theta"=pi,"DistanceShore"=2)
crcvm_rect_lf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_ne2",npost=1000,level=0.95,
                          xmin=xmin,xmax=xmax)

```

## Estimate from high frequency data with error : ExpShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_hf_e1 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_hf))
crcvm_rect_hf_e1$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
res=crcvm_rect_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
res$plot_omega_theta_ExpShore
```

### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))

sde_par_close <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 0.5 km")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
  ylab("omega")+ggtitle("Distance to shore = 3 km")

plot_omega_close
plot_omega_far

ggsave(filename="crcvm_rect_hf_e1/crcvm_rect_hf_e1_omega_close.png",plot=plot_omega_close)
ggsave(filename="crcvm_rect_hf_e1/crcvm_rect_hf_e1_omega_far.png",plot=plot_omega_far)
```

### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))

sde_par_away <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_away)
sde_CI_away <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_away)
sde_par_toward<- crcvm_rect_hf_e1$par(t = "all",new_data=predict_toward)
sde_CI_toward <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_toward)

# Data frame for point estimates
sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])

# Data frame for CIs
sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
                          lowomega = sde_CI_away["omega", "low",],
                          upomega = sde_CI_away["omega", "upp",])
sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
                          lowomega = sde_CI_toward["omega", "low",],
                          upomega = sde_CI_toward["omega", "upp",])

omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))

plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))

plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = 2\pi/3)'))

plot_omega_away
plot_omega_toward
```
## Estimate from high frequency data with error : DistanceShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(DistanceShore,k=5,bs="cs")+ti(theta,DistanceShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_hf_e2 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_hf))
crcvm_rect_hf_e2$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"DistanceShore"=0.3)
xmax=list("theta"=pi,"DistanceShore"=2)
crcvm_rect_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",npost=1000,level=0.95,
                          xmin=xmin,xmax=xmax)

```

## Estimate from low frequency data with error : ExpShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_lf_e1 <- SDE$new(formulas = formulas,data = data_rect_lf_e[data_rect_lf_e$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_lf))
crcvm_rect_lf_e1$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"ExpShore"=1/2)
xmax=list("theta"=pi,"ExpShore"=1/0.3)
xlabel=list("ExpShore"="Distance to shore")
link=list("ExpShore"=(\(x) 1/x))
crcvm_rect_lf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)

```

```{r}

#plot spline estimates of the parameter omega close and far to the shore

predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))

sde_par_close <- crcvm_rect_lf_e1$par(t = "all",new_data=predict_close)
sde_CI_close <- crcvm_rect_lf_e1$CI_simultaneous(t = "all",new_data=predict_close)
sde_par_far <- crcvm_rect_lf_e1$par(t = "all",new_data=predict_far)
sde_CI_far <- crcvm_rect_lf_e1$CI_simultaneous(t = "all",new_data=predict_far)

# Data frame for point estimates
sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])

# Data frame for CIs
sde_ci_df_close<- data.frame(theta=predict_close$theta,
                          lowomega = sde_CI_close["omega", "low",],
                          upomega = sde_CI_close["omega", "upp",])
sde_ci_df_far<- data.frame(theta=predict_far$theta,
                          lowomega = sde_CI_far["omega", "low",],
                          upomega = sde_CI_far["omega", "upp",])

omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))

plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+ xlab("Omega")

plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+ xlab("Omega")

plot_omega_close
plot_omega_far
```

## Estimate from low frequency data with error : DistanceShore and tensor splines ti

```{r}
#estimation with tensor splines
formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(DistanceShore,k=5,bs="cs")+ti(theta,DistanceShore,k=5,bs="cs"))
par0 <- c(0,0,1,4,0)

crcvm_rect_lf_e2 <- SDE$new(formulas = formulas,data = data_rect_lf_e[data_rect_lf_e$ID==1,],type = "RACVM1",
                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
                      other_data=list("H"=H1_lf))
crcvm_rect_lf_e2$fit(method="BFGS")
```

```{r}
xmin=list("theta"=-pi,"DistanceShore"=0.3)
xmax=list("theta"=pi,"DistanceShore"=2)
crcvm_rect_lf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_e2",npost=1000,level=0.95,
                          xmin=xmin,xmax=xmax)

```

# ESTIMATION WHEN TAU AND OMEGA ARE SMOOTH

### 2024-04-05 R/preprocessing/data_preprocessing.Rmd
Preprocessing updates

@@
 
-```{r import packages,message=FALSE}
+```{r libraries,message=FALSE}
 
@@
 
-```{r read raw data, cache=TRUE}
+```{r read raw data}
 # Set the path to the directory containing the data
@@
 
-```{r preprocess data 1, cache=TRUE}
+```{r preprocess data 1}
 
@@
 
-```{r plot trials, cache=TRUE}
+```{r plot trials}
 
@@
 
-```{r define ExpShipI and ExpShipT, cache=TRUE}
+```{r define ExpShipI and ExpShipT}
 
@@
 
-```{r ggplot ship exposure, cache=TRUE}
+```{r ggplot ship exposure}
 #plot exposure
@@
 
-```{r plot exposure per individual,cache=TRUE}
+```{r plot exposure per individual}
 
@@
 
-```{r preprocess data 2, cache=TRUE}
+```{r preprocess data 2}
 
@@
 
-```{r}
+```{r time steps histogram}
 times_histo=ggplot(dtimes, aes(x =60*delta,col=ID)) +
@@
 
-```{r milne land map, cache=TRUE}
+```{r milne land map}
 height <- max(dataprep$Lat) - min(dataprep$Lat)
@@
 
-```{r trajectories map, cache=TRUE}
+```{r trajectories map}
 
@@
 
-```{r extract coastline and land,cache=TRUE}
+```{r extract coastline and land}
 
@@
 #get the coastline geometry from the geojson file
-coastline_utm<-st_read(file.path(greenland_data_path,"filtered_coastline_utm.geojson"))
+coastline_utm<-st_read(file.path(greenland_data_path,"coastline_utm.geojson"))
 st_crs(coastline_utm)="+init=EPSG:32626"
@@
 #get land polygons
-land_utm<-st_read(file.path(greenland_data_path,"filtered_land_utm.shp"))
+land_utm<-st_read(file.path(greenland_data_path,"land_utm.shp"))
 
@@
 
-```{r compute nearest shore points,cache=TRUE}
+```{r compute nearest shore points}
 
@@
 ```
+
 ```{r summary distances to shore}
@@
 
-```{r}
+```{r distribution of points close to the shore}
 
@@
   geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (m)")
-```
-
-```{r}
+
 histo
@@
 
-```{r empirical velocities and deviation angles, cache=TRUE}
+```{r empirical velocities and deviation angles}
 

### 2024-04-06 R/preprocessing/data_preprocessing.Rmd
Update preprocessing

@@
 
-```{r read raw data}
+```{r read raw data,cache=TRUE}
 # Set the path to the directory containing the data
@@
 
-```{r trajectories map}
+```{r trajectories map,cache=TRUE}
 
@@
 
-```{r compute nearest shore points}
+```{r compute nearest shore points,cache=TRUE}
 

### 2024-04-11 R/simulation_study/simulation_study_fjords.Rmd
update simulation study

@@
 
-## Fix time steps for observations
+# Fix time steps for observations
 
@@
 
-## Set observation error
+# Set observation error
 
@@
 
-## Fix smooth parameters
+# Fix smooth parameters
 
@@
 
-## Estimation when omega only is smooth
+# Estimation when omega only is smooth
 
-### Generate samples : high frequency and no measurement error
+## Generate samples
+
+### High frequency and no measurement error
 
@@
 
-## percentage of samples that reached land
+### percentage of samples that reached land
 
@@
 
-### Generate samples : low frequency and no measurement error
+### Low frequency and no measurement error
 
@@
 
-### Generate samples : high frequency and measurement error
+### High frequency and measurement error
 
@@
 ```
+
 ### Plot samples
@@
 
-### Generate samples : low frequency and measurement error
+### Low frequency and measurement error
 
@@
 
-### Compute distance to shore and empirical deviation angles
+## Compute distance to shore and empirical deviation angles
 
@@
 
-### Estimate from high frequency data without error : ExpShore and tensor splines ti
+## Estimations
+
+### High frequency data without error : ExpShore and tensor splines ti
 
@@
 
-### Estimation from low frequency data without error
+### Low frequency data without error
 
@@
 
-### Estimate from high frequency data with error : ExpShore and tensor splines te
+### High frequency data with error : ExpShore and tensor splines te
 
@@
 
-### Estimate from low frequency data with error : ExpShore and tensor splines ti
+### Low frequency data with error : ExpShore and tensor splines ti
 

### 2024-04-11 R/simulation_study/simulation_study_rect.Rmd
update simulation study

@@
 title: "Simulation study in rectangular domain"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
+  html_notebook: default
+  pdf_document: default
 ---
@@
 
+#used in function split_term to get covariates names from the formulas
+library(stringr)
+
+#to fit smooth function with tensor splines
+library(mgcv)
+
 #some auxiliary functions
-source(file.path("/home","delporta","Documents","Recherche","Codes","CVM","CVM_functions.R"))
+source("CVM_functions.R")
 ```
 
-# Create the rectangle
+# SIMULATION
+
+## Create the rectangle
 
@@
 #rectangular domain 
-large_rectangle_coords <- matrix(c(0, 0, 0, 10000, 10000, 10000, 10000, 0, 0, 0), ncol = 2, byrow = TRUE)
-small_rectangle_coords <- matrix(c(1000,1000, 1000, 9000, 9000, 9000, 9000, 1000, 1000, 1000), ncol = 2, byrow = TRUE)
+large_rectangle_coords <- matrix(c(0, 0, 0, 10, 10, 10, 10, 0, 0, 0), ncol = 2, byrow = TRUE)
+small_rectangle_coords <- matrix(c(1,1, 1, 9, 9, 9, 9, 1, 1, 1), ncol = 2, byrow = TRUE)
 
@@
 
-# Sample initial positions uniformly in the domain
+## Sample initial positions uniformly in the domain
 
@@
 #number of tracks
-n_samples=100
+n_samples=1
 
@@
 
-# Fix time steps for observations
+## Fix time steps for observations
 
@@
 hf=1/60 #1 minute time step
-lf=5/60 #5 minute time step
 
@@
 
-#low frequency time step
-times_lf=seq(0,Tmax,by=lf)
-n_lf=length(times_lf)-1
 ```
 
-# Set observation error
+## Set observation error
 
@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs_0=0.003
-H0_lf=array(rep(sigma_obs_0^2*diag(2),n_lf),dim=c(2,2,n_lf))
+sigma_obs_0=0.005
 H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
@@
 sigma_obs_1=0.03
-H1_lf=array(rep(sigma_obs_1^2*diag(2),n_lf),dim=c(2,2,n_lf))
 H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf),dim=c(2,2,n_hf))
@@
 
-# Smooth parameters of the CRCVM
+## Smooth parameters of the CRCVM
 
-Define smooth functions for tau and omega.
+### Define smooth functions for tau and omega.
 
@@
 
-ftau=function(Dshore,theta,D0=0.3,delta0=0.2,tau0=1,tau1=2,tau2=2,lambda=4,kappa=0.1) {
+ftau=function(cov_data,D0=0.3,delta0=0.2,tau0=1,tau1=2,tau2=2,lambda=4,kappa=0.1) {
+  Dshore=cov_data$DistanceShore
+  theta=cov_data$theta
   coeff=exp(-kappa*(Dshore/D0)^2)
@@
 #mixture of independent gaussian
-ftau_gauss=function(Dshore,theta,D0=0.5,tau0=1,tau1=1.5) {
+ftau_gauss=function(cov_data,D0=0.5,tau0=1,tau1=1.5) {
+  Dshore=cov_data$DistanceShore
+  theta=cov_data$theta
   tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/10))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/10))^2-(Dshore/D0)^2)
@@
 #constant tau
-ftau_constant=function(Dshore,theta,tau=1) {
+ftau_constant=function(cov_data,tau=1) {
   return (tau)
@@
 
-fomega=function(Dshore,theta,D0=0.3,omega0=60*pi,lambda=1.5,kappa=0.5) {
+fomega=function(cov_data,D0=0.3,omega0=60*pi,lambda=1.5,kappa=0.5) {
+    Dshore=cov_data$DistanceShore
+    theta=cov_data$theta
+    if (is.null(Dshore)){
+      Dshore=1/cov_data$ExpShore
+    }
     coeff=exp(-kappa*(Dshore/D0)^2)
@@
 
-Plot smooth functions.
+### Plot smooth functions.
+
 ```{r}
@@
 theta=seq(from=-pi,to=pi,length.out=30)
-grid<- expand.grid(Dshore,theta)
+grid<- as.data.frame(expand.grid(Dshore,theta))
 colnames(grid)=c("DistanceShore","theta")
 true_par=grid
-true_par$tau=ftau_gauss(grid$DistanceShore,grid$theta)
-true_par$omega=fomega(grid$DistanceShore,grid$theta)
+true_par$tau=ftau_constant(grid)
+true_par$omega=fomega(grid)
 
@@
 
-# Simulation and estimation when only omega is smooth (other parameters are fixed)
-
-## Generate samples : no measurement errors and high frequency
+### Approximate smooth functions with tensor splines of DistanceShore
 
 ```{r}
+sp_df=5
+#number of points
+n <- 500
+
+#sample theta and DistanceShore points in the domain
+theta <- runif(n,-pi,pi);
+DistanceShore <- runif(n,0.2,2);
+samples=data.frame(theta=theta,DistanceShore=DistanceShore)
+
+#define grid of values
+theta_v <- seq(-pi,pi,length=30)
+Dshore_v<- seq(0.2,2,length=30)
+pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
+
+# true values of the function over this grid
+truth <- matrix(fomega(pr),30,30)
+
+#points on the surface perturbed by gaussian noise
+f <- fomega(samples)
+y <- f+0*rnorm(n)
+
+#plot true function
+persp(theta_v,Dshore_v,truth);
+title("truth")
+
+#fit with bivariate splines te
+m1 <- gam(y~te(theta,DistanceShore,k=sp_df))
+
+#visualize
+vis.gam(m1);title("tensor product")
+
+df=data.frame("x"=theta,"z"=DistanceShore,"y"=y)
+fig <- plot_ly(df,type="scatter3d", x = ~x, y = ~z, z = ~y,mode="markers",marker=list(size=4))
+fig
+```
+
+### Approximate smooth functions with tensor splines of ExpShore
+
+```{r}
+ExpShore=1/DistanceShore
+
+#fit with bivariate splines te
+m2 <- gam(y~te(theta,ExpShore,k=sp_df))
+
+#visualize
+vis.gam(m2);title("tensor product")
+
+```
+### Get estimated spline coefficients
+
+```{r}
+sp_coeff_Dshore=m1$coefficients
+sp_coeff_ExpShore=m2$coefficients
+```
+
+## Simulation with measurement errors and high frequency
+
+### Generate sample
+```{r, cache=TRUE,results='hide'}
 
@@
 }
-cat(count,"percent of the samples reached land")
+cat(count/n_samples*100,"percent of the samples reached land")
 ```
 
-## plot one sample
+### plot one sample
 
@@
 plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_path(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
+  geom_point(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(y1,y2),size=0.1)+
   geom_point(data = data_rect_hf_ne[data_rect_hf_ne$ID==1,]%>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
   xlab("x") + ylab("y")
@@
 
-## Subsample : no measurement errors and low frequency
+## Simulation with high measurement error and high frequency
 
-```{r}
+### Generate samples
 
-data_rect_lf_ne=data_rect_hf_ne[seq(1,n_hf,by=lf/hf),]
-
-```
-
-## plot subsamples
-
-```{r}
-
-plot_samples_rect_lf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_lf_ne[data_rect_lf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_path(data=data_rect_lf_ne[data_rect_lf_ne$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_point(data = data_rect_lf_ne[data_rect_lf_ne$ID==1,]%>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
-    xlab("x") + ylab("y")
-
-ggsave(filename="samples_rect_lf_ne_omega.png",plot=plot_samples_rect_lf_ne)
-ggplotly(plot_samples_rect_lf_ne)
-```
-
-## Generate samples : 30m measurement errors and high frequency
-
-```{r}
+```{r, cache=TRUE,results='hide'}
 
@@
 }
-cat(count,"percent of the samples reached land")
+cat(count/n_samples*100,"percent of the samples reached land")
 ```
-## plot samples
+
+### plot one sample
 
@@
 plot_samples_rect_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
+  geom_point(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(y1,y2),size=0.1)+
   geom_point(data = data_rect_hf_e[data_rect_hf_e$ID==1,]%>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
     xlab("x") + ylab("y")
@@
 
-## Subsample : measurement errors and low frequency
+# ESTIMATION
 
-```{r}
-
-data_rect_lf_e=data_rect_hf_e[seq(1,n_hf,by=lf/hf),]
-
-```
-
-## plot subsamples
-
-```{r}
-
-plot_samples_rect_lf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_lf_e[data_rect_lf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_point(data = data_rect_hf_e[data_rect_hf_e$ID==1,]%>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
-    xlab("x") + ylab("y")
-
-ggsave(filename="samples_rect_lf_e_omega.png",plot=plot_samples_rect_lf_e)
-ggplotly(plot_samples_rect_lf_e)
-```
 ## Compute distance to shore and empirical deviation angles
@@
 data_rect_hf_ne=add_covs(data_rect_hf_ne)
-data_rect_lf_ne=add_covs(data_rect_lf_ne)
 
 data_rect_hf_e=add_covs(data_rect_hf_e)
-data_rect_lf_e=add_covs(data_rect_lf_e)
 ```
 
-## Estimate from high frequency data without error : ExpShore and tensor splines te
+## Estimate from high frequency data without low error
 
-### Fit sde
-```{r}
+### Use ExpShore covariate and tensor splines te
+
+#### Fit sde
+
+```{r,cache=TRUE}
 
@@
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=c(10,5),bs="cs"))
+                 nu=~1,omega=~te(theta,ExpShore,k=sp_df,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1 & data_rect_hf_ne$time<2*24,],type = "RACVM1",
+crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 
-### Plot smooth parameter omega in 3D
+#### Plot estimated smooth parameter omega in 3D
 
@@
 link=list("ExpShore"=(\(x) 1/x))
-res=crcvm_rect_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-res$plot_omega_theta_ExpShore
+true_smooths=list("omega"=fomega)
+res=crcvm_rect_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_ExpShore
 ```
 
-### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.
+#### Violin plot of the estimated spline coefficients
 
@@
 
-#plot spline estimates of the parameter omega close and far to the shore
+true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
+rownames(true_coeff)=NULL
 
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))
+p=crcvm_rect_hf_ne1$plot_violin_coeff(model_name="crcvm_rect_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-sde_par_close <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_far)
-
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 0.5 km")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 3 km")
-
-plot_omega_close
-plot_omega_far
-
-ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_close.png",plot=plot_omega_close)
-ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_far.png",plot=plot_omega_far)
-
+ggplotly(p)
 ```
 
-### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.
+#### Simulate from fitted model
 
-```{r}
-
-#plot spline estimates of the parameter omega close and far to the shore
-
-predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
-predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))
-
-sde_par_away <- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_away)
-sde_CI_away <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_away)
-sde_par_toward<- crcvm_rect_hf_ne1$par(t = "all",new_data=predict_toward)
-sde_CI_toward <- crcvm_rect_hf_ne1$CI_simultaneous(t = "all",new_data=predict_toward)
-
-# Data frame for point estimates
-sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
-sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
-                          lowomega = sde_CI_away["omega", "low",],
-                          upomega = sde_CI_away["omega", "upp",])
-sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
-                          lowomega = sde_CI_toward["omega", "low",],
-                          upomega = sde_CI_toward["omega", "upp",])
-
-omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
-                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))
-
-plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
-  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
-  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))
-
-plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
-  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
-  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))
-
-plot_omega_away
-plot_omega_toward
-
-ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_away.png",plot=plot_omega_away)
-ggsave(filename="crcvm_rect_hf_ne1/crcvm_rect_hf_ne1_omega_toward.png",plot=plot_omega_toward)
-
-```
-
-### Simulate from fitted model
-
-```{r}
+```{r,cache=TRUE}
 
@@
 atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
-sim_data=crcvm_rect_hf_e1$simulate(z0=x0[1,],data=crcvm_rect_hf_e1$data(),atw=atw,land=domain)
+sim_data=crcvm_rect_hf_ne1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne1$data(),atw=atw,land=domain)
 
 ```
+
+#### Plot simulation from fitted model
 
@@
 plot_sim=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=sim_data,mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_path(data=sim_data,mapping=aes(1000*y1,1000*y2),size=0.1)+
+  geom_point(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=sim_data,mapping=aes(y1,y2),size=0.1)+
   geom_point(data = sim_data%>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "red")+
+             aes(x = y1, y = y2 ), shape = 3, size = 4, col = "red")+
     xlab("x") + ylab("y")
 
+ggsave("sim_crcvm_rect_hf_ne1.png",plot=plot_sim,width=10,height=5,path="crcvm_rect_hf_ne1")
 ggplotly(plot_sim)
@@
 
-### S
+###  UseDistanceShore covariate and tensor splines ti
 
-## Estimate from high frequency data without error: DistanceShore and tensor splines ti
+#### Fit sde
 
-```{r}
+```{r,cache=TRUE}
 #estimation with tensor splines
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=5,bs="cs"))
+                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_rect_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM1",
+crcvm_rect_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 
+#### Plots estimated smooth in 3D
+
 ```{r}
@@
 xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_rect_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",npost=1000,level=0.95,
-                          xmin=xmin,xmax=xmax)
-res$plot_omega_theta_DistanceShore
+res=crcvm_rect_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_DistanceShore
 ```
 
-Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.
+#### Violin plot of the estimated spline coefficients
 
 ```{r}
-#plot spline estimates of the parameter omega close and far to the shore
 
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),DistanceShore=rep(0.3,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),DistanceShore=rep(2,100))
+true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
+rownames(true_coeff)=NULL
 
-sde_par_close <- crcvm_rect_hf_ne2$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_rect_hf_ne2$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_rect_hf_ne2$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_rect_hf_ne2$CI_simultaneous(t = "all",new_data=predict_far)
+p=crcvm_rect_hf_ne2$plot_violin_coeff(model_name="crcvm_rect_hf_ne2",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.3,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  xlab("Theta")+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  xlab("Theta")+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")
-
-plot_omega_close
-plot_omega_far
+ggplotly(p)
 ```
 
-## Estimate from low frequency data without error : ExpShore and tensor splines ti
+## Estimate from high frequency data with error :
 
-```{r}
+### Use ExpShore and tensor splines te
+
+#### Fit sde
+
+```{r, cache=TRUE}
 #estimation with tensor splines
@@
 
-crcvm_rect_lf_ne1 <- SDE$new(formulas = formulas,data = data_rect_lf_ne[data_rect_lf_ne$ID==1,],type = "RACVM1",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),other_data=list(H=H0_lf))
-crcvm_rect_lf_ne1$fit(method="BFGS")
-```
-
-```{r}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.1)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-crcvm_rect_lf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-
-```
-
-```{r}
-#plot spline estimates of the parameter omega close and far to the shore
-
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.3,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))
-
-sde_par_close <- crcvm_rect_lf_ne1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_rect_lf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_rect_lf_ne1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_rect_lf_ne1$CI_simultaneous(t = "all",new_data=predict_far)
-
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.3,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  xlab("Theta")+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  xlab("Theta")+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")
-
-plot_omega_close
-plot_omega_far
-```
-
-## Estimate from low frequency data without error : DistanceShore and tensor splines ti
-
-```{r}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_rect_lf_ne2 <- SDE$new(formulas = formulas,data = data_rect_lf_ne[data_rect_lf_ne$ID==1,],type = "RACVM1",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H0_lf))
-crcvm_rect_lf_ne2$fit(method="BFGS")
-```
-
-```{r}
-xmin=list("theta"=-pi,"DistanceShore"=0.1)
-xmax=list("theta"=pi,"DistanceShore"=2)
-crcvm_rect_lf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_ne2",npost=1000,level=0.95,
-                          xmin=xmin,xmax=xmax)
-
-```
-
-## Estimate from high frequency data with error : ExpShore and tensor splines ti
-
-```{r}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_rect_hf_e1 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM1",
+crcvm_rect_hf_e1 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 ```
+
+#### Plot estimated smooth in 3D
 
@@
 link=list("ExpShore"=(\(x) 1/x))
-res=crcvm_rect_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-res$plot_omega_theta_ExpShore
+true_smooths=list("omega"=fomega)
+res=crcvm_rect_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_ExpShore
 ```
 
-### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.
+#### Violin plot of the estimated spline coefficients
 
@@
 
-#plot spline estimates of the parameter omega close and far to the shore
+true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
+rownames(true_coeff)=NULL
 
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))
+p=crcvm_rect_hf_e1$plot_violin_coeff(model_name="crcvm_rect_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-sde_par_close <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_far)
-
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 0.5 km")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 3 km")
-
-plot_omega_close
-plot_omega_far
-
-ggsave(filename="crcvm_rect_hf_e1/crcvm_rect_hf_e1_omega_close.png",plot=plot_omega_close)
-ggsave(filename="crcvm_rect_hf_e1/crcvm_rect_hf_e1_omega_far.png",plot=plot_omega_far)
+ggplotly(p)
 ```
 
-### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.
+### Use DistanceShore and tensor splines te
 
-```{r}
+#### Fit sde
 
-#plot spline estimates of the parameter omega close and far to the shore
-
-predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
-predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))
-
-sde_par_away <- crcvm_rect_hf_e1$par(t = "all",new_data=predict_away)
-sde_CI_away <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_away)
-sde_par_toward<- crcvm_rect_hf_e1$par(t = "all",new_data=predict_toward)
-sde_CI_toward <- crcvm_rect_hf_e1$CI_simultaneous(t = "all",new_data=predict_toward)
-
-# Data frame for point estimates
-sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
-sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
-                          lowomega = sde_CI_away["omega", "low",],
-                          upomega = sde_CI_away["omega", "upp",])
-sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
-                          lowomega = sde_CI_toward["omega", "low",],
-                          upomega = sde_CI_toward["omega", "upp",])
-
-omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
-                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))
-
-plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
-  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
-  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))
-
-plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
-  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
-  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = 2\pi/3)'))
-
-plot_omega_away
-plot_omega_toward
-```
-## Estimate from high frequency data with error : DistanceShore and tensor splines ti
-
-```{r}
+```{r,cache=TRUE}
 #estimation with tensor splines
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(DistanceShore,k=5,bs="cs")+ti(theta,DistanceShore,k=5,bs="cs"))
+                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_rect_hf_e2 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM1",
+crcvm_rect_hf_e2 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 
-```{r}
-xmin=list("theta"=-pi,"DistanceShore"=0.3)
-xmax=list("theta"=pi,"DistanceShore"=2)
-crcvm_rect_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",npost=1000,level=0.95,
-                          xmin=xmin,xmax=xmax)
-
-```
-
-## Estimate from low frequency data with error : ExpShore and tensor splines ti
-
-```{r}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_rect_lf_e1 <- SDE$new(formulas = formulas,data = data_rect_lf_e[data_rect_lf_e$ID==1,],type = "RACVM1",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_lf))
-crcvm_rect_lf_e1$fit(method="BFGS")
-```
-
-```{r}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.3)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-crcvm_rect_lf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-
-```
-
-```{r}
-
-#plot spline estimates of the parameter omega close and far to the shore
-
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/2,100))
-
-sde_par_close <- crcvm_rect_lf_e1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_rect_lf_e1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_rect_lf_e1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_rect_lf_e1$CI_simultaneous(t = "all",new_data=predict_far)
-
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(2,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+ xlab("Omega")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+ xlab("Omega")
-
-plot_omega_close
-plot_omega_far
-```
-
-## Estimate from low frequency data with error : DistanceShore and tensor splines ti
-
-```{r}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(DistanceShore,k=5,bs="cs")+ti(theta,DistanceShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_rect_lf_e2 <- SDE$new(formulas = formulas,data = data_rect_lf_e[data_rect_lf_e$ID==1,],type = "RACVM1",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_lf))
-crcvm_rect_lf_e2$fit(method="BFGS")
-```
+#### Plot estimated smooth in 3D
 
@@
 xmax=list("theta"=pi,"DistanceShore"=2)
-crcvm_rect_lf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_lf_e2",npost=1000,level=0.95,
-                          xmin=xmin,xmax=xmax)
-
+res=crcvm_rect_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_DistanceShore
 ```
 
-# ESTIMATION WHEN TAU AND OMEGA ARE SMOOTH
+#### Violin plot of the estimated spline coefficients
+
+```{r}
+
+true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
+rownames(true_coeff)=NULL
+
+p=crcvm_rect_hf_e2$plot_violin_coeff(model_name="crcvm_rect_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
+
+ggplotly(p)
+```

### 2024-04-13 Latex/main.tex
changes in tex file and mixed effects analysis

@@
 
-\author{Adeline Samson%
-\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}
-
 \author{Alexandre Delporte%
@@
 
+\author{Adeline Samson%
+	\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}
+	\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}
+
 \date{Dated: \today}
@@
 
+% Underwater noise in Arctic : lower than in other environments, anthropogenic nois : vessel traffic, ice breakers, drilling, sonars, seismic airguns for oil and gas exploration. (Halliday 2020)
+% vessel traffic likely to increase due to the melting of the ice cap
+% Largest impact on underwater noise due to seismic airguns (Halliday 2020)
+%Marine mammals sound exposure effects : behavioral and physical impacts (stranding events), more difficult to assess behavioral effect (southall 2007, 2021)
+%Sound threshold, regulation, research recommendations : behavioral response
+%Behavioral response study, control exposure experiments (examples: pile driving, seismic airguns, sonar)
+
+% Studies about a lot of species : narwhals, toothed whale,  mid frequency mammals
+% Greenland narwhals studies
+% effects on dive, vocalization and horizontal movement
+% data, diffusion models 
+% spatial constraints
 %context and literature review, objective of the paper, new contributions
-% ARCTIC , narwhals etc
+
+Anthropogenic underwater noise in the Arctic environment causes disturbances in the endemic wildlife.
+More precisely, marine mammals, whose vital functions highly depend on sound perception for communication and orientation, are most likely to react to such disturbances.
+The Arctic Council report CITATION emphasized the need for more knowledge about the impact of underwater noise on Arctic marine mammals.
+The main sources of noise in the region are vessel traffic, ice breakers, drilling, sonars and seismic airguns used for oil and gas exploration.
+Among those sources, an increase in vessel traffic is anticipated due to the melting of the ice cap, and seismic airguns were reported to have the largest impact on underwater noise.
+Two types of consequences can be considered : injury and behavior disturbance.
+While physical harm is straightforward to assess and mitigate by determining species-specific received sound level thresholds, no canonical method is expected to be found to examine behavior disturbances, due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions.
+Research recommendations from  CITATION include a particular focus on measurement of behavioral reactions to various type of sound exposure.
+Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
+Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises CITATION, belugas, sperm whales, harbour porpoises, narwhals and many other marine mammals
+A severity scale was defined in CITATION to help assessing which behavioral responses are more likely to alter a population's capacity to survive.
+Changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations  were ranked between 4 and 7 on a scale out of 9 (9 being stranding events that can directly lead to the death of the animal) depending on the magnitude of the changes and their duration.
+
+So far, the studies about narwhals behavioral responses to sound exposure proved that they exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns.
+Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source.
+The data used in this paper is the 2018 data used in \cite{heide-jorgensen_behavioral_2021} CITATIONS.
+Horizontal movement has for now been analyzed with discrete time methods such as Markov chains .
+A covariate representing the level of sound received by the narwhals was judged appropriate to model the transition probabilities of the Markov Chain, and this analysis led to the conclusions above-mentioned.
+Here, we decide to make use of a continuous time model to analyse the GPS positions and try to corroborate this analysis.
+The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\
+% severity scale for behavioral response : behavior that can alter the population capacity to survive (reproduce, eat etc)
+ %%Studies about narwhals : diving, avoidance etc
+ %% Southeast Greenland narwhals population in decline
+  \\ \\
+
 Assessment of human disturbance on animal behaviour has become a major concern in the ecology community. 

### 2024-04-15 Latex/main.tex
Commit Latex and simulation study changes

--- 2024-04-05 Latex/main.tex
@@
 \usepackage{tikz}
+\usepackage{yhmath}
 \usetikzlibrary{quotes,angles}
@@
 
+%CITATIONS APPEARANCE
+\DeclareCiteCommand{\cite}[\mkbibbold]
+{\usebibmacro{prenote}}
+{\usebibmacro{citeindex}%
+	\printtext[bibhyperref]{\usebibmacro{cite}}}
+{\multicitedelim}
+{\usebibmacro{postnote}}
+
+\DeclareFieldFormat{boldcite}{\mkbibbold{#1}}
+
+\DeclareCiteCommand{\cite}[\mkbibbold]
+{\usebibmacro{prenote}}
+{\printtext[brackets]{\usebibmacro{cite}}}
+{\multicitedelim}
+{\usebibmacro{postnote}}
+
 %TITLE
@@
  \textbf{Analysis of Greenland narwhals's behavioural responses to ship noise and airguns exposure using varying coefficients correlated velocity models}}
-
-\author{Adeline Samson%
-\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}
 
@@
 
+% Underwater noise in Arctic : lower than in other environments, anthropogenic nois : vessel traffic, ice breakers, drilling, sonars, seismic airguns for oil and gas exploration. (Halliday 2020)
+% vessel traffic likely to increase due to the melting of the ice cap
+% Largest impact on underwater noise due to seismic airguns (Halliday 2020)
+%Marine mammals sound exposure effects : behavioral and physical impacts (stranding events), more difficult to assess behavioral effect (southall 2007, 2021)
+%Sound threshold, regulation, research recommendations : behavioral response
+%Behavioral response study, control exposure experiments (examples: pile driving, seismic airguns, sonar)
+
+% Studies about a lot of species : narwhals, toothed whale,  mid frequency mammals
+% Greenland narwhals studies
+% effects on dive, vocalization and horizontal movement
+% data, diffusion models 
+% spatial constraints
 %context and literature review, objective of the paper, new contributions
-% ARCTIC , narwhals etc
-Assessment of human disturbance on animal behaviour has become a major concern in the ecology community.
-Both instantaneous and long-term changes in animal behaviour have been studied by taking advantage of the large amount of wildlife data now available.
-One of the commonly accepted method to analyse such behavioural changes due to human activities is to perform "controlled exposure experiments".
-Animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, sound exposure level, temperature, pressure, heart beat, stroke rate...
-Several studies involving beaked whales \parencite{tyack_beaked_2011}, common sole larvaes \parencite{bolle_common_2012}, narwhals \parencite{heide-jorgensen_behavioral_2021} and many other species have already been done.
-Marine mammals are particularly challenging to track since GPS data can only be retrieved when they emerge at the surface of the sea and a satellite is near enough.
-Moreover, some species can dive at great depth, so that the equipment must keep working in such extreme environment \parencite{cioffi_trade-offs_nodate}.
-In this paper, a data-set of east Greenland narwhals's movement under controlled exposure experiments will be analysed.
+ %% Southeast Greenland narwhals population in decline
+
+Anthropogenic underwater noise in the Arctic environment causes disturbances in the endemic wildlife.
+More precisely, marine mammals, whose vital functions highly depend on sound perception for communication and orientation, are likely to react to such disturbances.
+The Arctic Council report \cite{halliday_underwater_2020} emphasized the need for more knowledge about the impact of underwater noise on Arctic marine mammals.
+The main sources of noise in the region are vessel traffic, ice breakers, drilling, sonars and seismic airguns used for oil and gas exploration.
+Among those sources, an increase in vessel traffic is anticipated due to the melting of the ice cap, and seismic airguns are reported to have the largest impact on underwater noise.
+Two types of consequences can be considered : injury and behavior disturbance.
+While physical harm is straightforward to assess and mitigate by determining species-specific received sound level thresholds, no canonical method is expected to be found to examine behavior disturbances, due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions.\\
+Research recommendations include a particular focus on measurement of behavioral reactions to various type of sound exposure \cite{southall_marine_2008,southall_marine_2019}.
+Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
+Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises \cite{tyack_beaked_2011,cioffi_trade-offs_2022}, belugas \cite{martin_exposure_2023}, sperm whales \cite{madsen_quantitative_2006}, blue whales \cite{friedlaender_preymediated_2016}, narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021} and many other marine mammals.
 \\
-
-As a consequence of climate change and decrease in sea ice coverage, the Greenland ecosystem is likely to be more and more exposed to anthropogenic activity in the next decades \cite{heide-jorgensen_behavioral_2021}.
-Shipping activities as well as sonar and seismic surveys are a few instances of new disturbances that could alter animals's habits.
-Hence, there is an urge for more knowledge about the impact this could have on endemic species such as the narwhals.
-This marine mammal is one of the most sensitive to sound exposure as it uses echolocation to orient itself beneath the water \cite{tervo_narwhals_2021}.
-Moreover, recent studies showed that the population of narwahls in Southeast Greenland is already in decline \cite{garde_biological_2022}.
-The effects of sound and ship exposure on its diving and foraging behaviour have already been the subject of several studies \cite{heide-jorgensen_behavioral_2021,ngo_understanding_2019}.
-However, their horizontal movement has for now been analyzed with discrete time methods such as Markov chains  \cite{heide-jorgensen_behavioral_2021}.
-A covariate representing the level of sound received by the narwhals was used to model the transition probabilities of the Markov Chain, and it was concluded that narwhals effectively tried to avoid the ship by moving fast toward the shore.\textcolor{red}{add more detailed explanations}.
-Here, we use a continuous time model to analyse the GPS positions and try to corroborate this analysis.
+A severity scale was defined in \cite{southall_marine_2008} to help assessing which behavioral responses are more likely to alter a population's capacity to survive, reproduce or forage.
+Changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations  were ranked between 4 and 7 on a scale out of 9 (9 being stranding events that can directly lead to the death of the animal) depending on the magnitude of the changes and their duration.\\
+
+So far, the studies about narwhals behavioral responses to sound exposure proved that they exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}.
+Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source \cite{tervo_narwhals_2021}.
+These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}.
+Resulting GPS positions of the narwhals have been analyzed with discrete time methods, typically relying on Markov chains.
+Since GPS positions are retrieved irregularly in time, each narwhal track was linearly interpolated to obtain a position every second, and the positions were then classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second.
+An exposure covariate defined as the inverse of the distance to the sound source was used to model the transition probabilities of the Markov Chain, and it was found that the probabilities of being in states "Close to the shore" and "Moving toward the shore" increased with exposure to the sound, while probability of being in state "Far from the shore" decreased.
+In this paper, we adopt a different approach to describe horizontal movement and make use of a continuous time model to analyse the GPS positions.
 The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\
 
-We propose to use a stochastic differential equation with varying coefficient to model the GPS tracks of the individuals.
-The continuous time correlated random walk (also known as integrated Ornstein-Uhlenbeck, or correlated velocity model) that has been introduced in \cite{johnson_continuous_2008} is one of the most common model one can find in the literature to analyse such data.
-It has been  widely used in ecology to model 2D animal movement \cite{gurarie_correlated_2017,albertsen_generalizing_2019} \textcolor{red}{add more references}, and even 3D movement \cite{gurarie_estimating_2011}.
-
- Following \cite{gurarie_correlated_2017}, this model can be written
+A stochastic differential equation with time-varying coefficient is employed to describe the GPS tracks of the individuals \cite{michelot_varying-coefficient_2021}.
+One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}.
+It has been used extensively to analyze animal movement in two or even three dimensions \cite{johnson_continuoustime_2008,gurarie_correlated_2017,alt_correlation_1990,gurarie_estimating_2011,albertsen_generalizing_2018}, and can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
+The following formulation of the equations is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
 \begin{equation} \left\{
 	\begin{array}{l}
-		dv(t)=-\frac{1}{\tau} (v(t) -\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
-		dx(t)=v(t) dt
-	\end{array}
-	\right.
-	\label{eq: CVM equation}
-\end{equation}
-$v(t)$ is the horizontal 2D velocity at time $t$ and $x(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
-The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, and $\mu$ is the long term velocity.
-This equation is a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
-\\
-
-In general, the motion of marine mammals is subject to spatial constraints induced by the coastline.
-This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process.
-However, this phenomenon is not taken into account in model~\ref{eq: CVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
-Recent papers including constraints are \cite{hanks_reflected_2017} and \cite{russell_spatially_2018}.
- In \cite{hanks_reflected_2017}, a reflected version of the CVM is considered to study sea lion telemetry data.
-For the simulation, points ending up in land are projected on the coastline.
-Estimation of the SDE parameters relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data.
-This requirement is met by linearly interpolating between the discrete time observations of the process.
-In \cite{russell_spatially_2018}, a repulsive potential function is used to constrain ants movement within a box.
-Usually, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}.
-For the narwhals, it was concluded from the data that they have a proclivity to move along the shoreline.
-Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.
-
-To add a rotational component, we propose to use the rotational version of the CVM \cite{gurarie_correlated_2017,albertsen_generalizing_2019}.
-\begin{equation} \left\{
-	\begin{array}{l}
-		dv(t)=-A(v(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
-		dx(t)=v(t) dt
+		dX(t)=V(t)dt \\
+		dV(t)=-\begin{pmatrix} 
+			\frac{1}{\tau} & -\omega \\
+			\omega & \frac{1}{\tau}
+		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
 	\end{array}
@@
 \end{equation}
-where $A=\begin{pmatrix} 
-	\frac{1}{\tau} & -\omega \\
-	\omega & \frac{1}{\tau}
-\end{pmatrix}$.
-The new parameter $\omega$ is an angular velocity that controls how fast the velocity vector should rotate. Notice that if $\omega=0$ this is simply equation \ref{eq: CVM equation}.
-
-The novelty of this paper consists in estimating the parameters $\tau$ and $\omega$ as smooth functions of the distance to the boundary and the angle between the animal's direction and the boundary normal vector.
-This is a way to make the drift in the velocity equation dependent on the location process $x$.
-\\
-Such a model can be estimated in the framework of the \texttt{R} package \texttt{smoothSDE}.
-To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuous_2008}.
-Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.
-
-On top of this, the parameter $\nu$ will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
-The effect of sound exposure is assessed both in the constrained and unconstrained frameworks.
+$V(t)$ is the horizontal 2D velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
+The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
+Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
+Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.\\
+
+In general, the motion of marine mammals is subject to spatial constraints induced by the coastline.
+This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process.
+However, this phenomenon is not taken into account in model~\ref{eq: RACVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
+Recent research has been conducted to include constraints in these type of models.
+A reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
+Simulation of constrained paths was achieved by projecting on the coastline the  points that eventually reached land.
+Estimation of the SDE parameters relied on computationally intensive MCMC algorithm (Metroplis-Hastings), and required high frequency data.
+This requirement was met by linearly interpolating between the discrete time observations of the process.
+Another idea is to add a repulsive potential function in the drift.
+This has been done to constrain ants movement within a box \cite{russell_spatially_2018}, but requires very simple boundaries to get an explicit expression of the potential function.\\
+Moreover, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}.
+For narwhals, it was observed from the data that they have a proclivity to move along the shoreline.
+Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.
+The novelty of this paper consists in estimating the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector, therefore making the drift in the velocity equation dependent on the location process $x$ and the boundary of the domain.
+Note that in this framework, the special case $\omega=0$ would be the unconstrained correlated velocity model of \cite{johnson_continuoustime_2008}.
+\\ %More biblio about estimation methods
+The angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE}.
+To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
+Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\
+
+Sound exposure will be considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}.
+This effect will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
+This choice of covariate is motivated by previous work in \cite{heide-jorgensen_behavioral_2021}.
+A response model will be compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
+The effect of sound exposure is assessed both in the constrained and unconstrained ($\omega=0$) frameworks.
 
@@
 \begin{itemize}
-	\item Derivation of a state-space model for the RACVM defined in \cite{gurarie_correlated_2017}, and addition of this method in the framework of \texttt{smoothSDE}, to enable the use of smooth parameters $\nu$,$\tau$,$\omega$ and $\mu$, and estimation from noisy observations irregularly spaced in time.
+	\item Derivation of a state-space model for the so called "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}) \cite{gurarie_correlated_2017}, and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
 	\item Definition of a Constrained Rotational Correlated Velocity Model (CRCVM) for simulation study.
 Deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
-	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the Distance to the ship.
+	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
 \end{itemize}
 
-Note that simpler SDE models as Brownian motion or Ornstein-Uhlenbeck with varying coefficient already proved to be convenient at assessing behavioural responses by examining deviation from a baseline model that represents the animal behaviour under normal conditions.
-For instance, it was used in \cite{michelot_continuous-time_2022} to estimate avoidance behaviours of beaked whales after exposure to military sonar systems.
-
-Eventually, while a lot of results are known about the existence and properties of constrained diffusions, their application in the context of animal motion remains relatively underdeveloped \cite{brillinger_simulating_2003}.
-This text aims to contribute to bridging this gap.
-\newpage
+Section 2 gives an overview of the narwhal data available for the analysis.
+Then in section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models.
+Section 4 is dedicated to a comprehensive simulation study, whereas in the last section, the results of the application to the narwhal data are examined.
 
@@
 %trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
-The dataset analysed in this paper has already been the subject of several studies, focusing mainly on the diving behaviour of the narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021,ngo_understanding_2019}.
-Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
+As already mentioned, the dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}.
+Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
 The Scoresby Sound fjord system
@@
 location of every shot. 
-Data includes narwhals's latitude and longitude, distance of the narwhals relative to the ship in metres, GPS time, distance to the shore in metres.
+Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres.
 GPS positions are known only at specific times when the narwhals get close to the sea surface.
-The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.\textcolor{red}{add histograms of time steps}\\
+The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.
+Figure \ref{fig:alltimestepshisto} shows the distribution of time steps in the data.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.5]{images/data_exploration/all_time_steps_histo}
+	\caption{Histogram of time step between consecutive GPS positions}
+	\label{fig:alltimestepshisto}
+\end{figure}
 
@@
 The land geometry for this specific region in South-East Greenland was gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
-For each GPS measurement time $t_j$, the nearest point $p(t_j)$ on the shoreline was computed and the distance to the shore was determined as the distance to this point.
+Supposing there are $n$ observations, for $j \in \{1,\cdots,n\}$, the GPS position obtained at time $t_j$ is denoted $y(t_j)$, and the nearest point on the shoreline is denoted $p(t_j)$.
+The distance to the shore, denoted $D_{shore}(t_j)$ in the sequel, is determined as the distance to this point.
 The resulting values range from $0$ to $7.5$ km.
@@
 \begin{array}{ll}
-	 E_{shore}=0 & \mbox{if } D_{shore}>D_0 \\
-	E_{shore}=\frac{1}{D_{shore}} & \mbox{if } D_{shore} < D_0
+	 E_{shore}(t_j)=0 & \mbox{if } D_{shore}(t_j)>D_0 \\
+	E_{shore}(t_j)=\frac{1}{D_{shore}(t_j)} & \mbox{if } D_{shore}(t_j) \leq D_0
 \end{array}
@@
 where the threshold $D_0$ was chosen equal to $0.5$ km.
-The angle between the narwhal's heading and the shoreline was computed as the angle $\theta$ between each step vector $x(t_j+\Delta_j)-x(t_j)$ and the vector $n(t_j)=x(t_j)-p(t_j)$.
+The angle between the narwhal's heading and the shoreline is estimated as the angle $\Theta(t_j)$ between the observed empirical velocity $\hat{v}(t_j)=\frac{y(t_j+\Delta_j)-y(t_j)}{\Delta_j}$, where $\Delta_j=t_{j+1}-t_j$, and the vector $n(t_j)=y(t_j)-p(t_j)$.
 This is illustrated in figure~\ref{fig: illustration nearest shore points}.
-A value $\Theta=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,\frac{-\pi}{2}\right[$ indicates movement toward the shore.
-The distance from the ship was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
+A value $\Theta(t_j)=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta(t_j) \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.
+The distance from the ship, denoted $D_{ship}(t_j)$, was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
 Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}.
-The covariates $E_{shore}$ and $\Theta$ will only be used in the constrained SDE models.
+The covariates $E_{shore}$ and $\Theta$ will only be used for the smooth angular velocity parameter $\omega$ in the constrained SDE models.
 
@@
 		\node[left] at (0,2) {$p(t_j)$};
-		\node[left] at (1.2,2.8) {$\vec{v}(t_j)$};
+		\node[left] at (1.2,2.8) {$\hat{v}(t_j)$};
+		\node[below] at (1.8,2) {$y(t_j)$};
 		\node[below,red] at (3,2) {$\vec{n}(t_j)$};
 
-		\node[above,orange] at (2.3,2.3) {$\theta(t_j)$};
+		\node[above,orange] at (2.3,2.3) {$\Theta(t_j)$};
 	\end{tikzpicture}
-	\caption{Example of nearest shore point and angle $\theta$}
+	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain.}
 	\label{fig: illustration nearest shore points}
@@
 
-The exposure to ship covariate was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
-$D_{ship}$ is not known when the narwhal is not in line of sight with the boat, so that exposures are set to $0$ in this case.
-The closer the narwhal is to the ship, the greater the exposure.
-This is a proxy for sound exposure levels received by the narwhals.
-Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.All the covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
+The exposure to ship covariate, denoted $E_{ship}$, was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
+$D_{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
+This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source.
+This covariate is meant to be a proxy for sound exposure levels received by the narwhals.
+Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure.
+Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.
+All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
 
@@
 
-\subsection{Rotational correlated velocity model (RCVM)}
+\subsection{Standard Rotational correlated velocity model (RCVM)}
 \label{section: RCVM}
 
-The more general equation for a rotational correlated velocity model is \cite{albertsen_generalizing_2019}:
+A general formulation for a rotational correlated velocity model is \cite{albertsen_generalizing_2018}:
 
@@
 	\begin{array}{l}
-		dv(t)=-A(v(t)-\mu)dt+\Sigma dW(t) \\
-		dx(t)=v(t) dt
+		dX(t)=V(t) dt \\
+		dV(t)=-A(V(t)-\mu)dt+\Sigma dW(t)
+
 	\end{array}
@@
 	\omega & \frac{1}{\tau_2}
-\end{pmatrix}$.
-Here, $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
-This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2018}.
-There is an explicit form for the gaussian transition density of the velocity process \cite{albertsen_generalizing_2019}, where the covariance matrix is expressed using a Kronecker sum.
-
-It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$.
-This diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$ to give biological meaning to the parameters \cite{gurarie_correlated_2017}.These simplifications produce the equation~\ref{eq: RACVM equation}.
+\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
+This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2019}.
+There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
+
+It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$ \cite{johnson_continuoustime_2008}.
+The diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$, where $\nu$ is homogeneous to a velocity.
+This allows easier biological interpretation  of the parameters in the equation \cite{gurarie_correlated_2017}.
+These simplifications produce the equation~\ref{eq: RACVM equation}.
 
@@
 The transition density for the velocity process is such that 
-\begin{equation}v(t+\Delta) \vert v(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
+\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
 \end{equation}
@@
 In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$.
-Hence, the order of magnitude of the parameter $\nu$ can be deduced from the data by computing the overall mean empirical velocity norm.
-The distribution of the phase angle $\Phi$ of the velocity has been derived in \cite{pawula_phase_1982}.
+Hence, the order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm.
+The distribution of the phase angle $\Phi$ of the velocity has been derived in \cite{pawula_distribution_1982}.
 When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$.
@@
 
-Then, by integrating the velocity equation over time one gets the transition density of the location process.
+By integrating the velocity equation over time one gets the transition density of the location process.
 \begin{equation}
-	x(t+\Delta) \vert x(t)=x, v(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
+	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
 \end{equation}
@@
 \]
-Note that the position process is not markovian, only the whole process $(x(t), v(t))^\top$ is markovian.
-Remark that both components of the location and velocity processes are independent.
-One can check that for $\omega=0$, the formulas match the one of the CTCRW model \cite{johnson_continuous_2008}.
+The position process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
+Both components of the location and velocity processes are independent.
+and one can check that for $\omega=0$, the formulas match the ones of the  CTCRW model \cite{johnson_continuoustime_2008}.
 Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase.
@@
 This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
-Denote $\mathcal{D} \subset \R^2$ the domain of the process $x$.
+Denote $\mathcal{D} \subset \R^2$ the domain of the process $X$.
 \begin{equation} \left\{
 	\begin{array}{l}
-		dv(t)=-A(x(t),v(t))v(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \\
-		dx(t)=v(t) dt
+		dX(t)=V(t) dt \\
+		dV(t)=-A(X(t),V(t))v(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)
+
 	\end{array}
@@
 
-Here, $A(x(t),v(t))=\begin{pmatrix} 
+Here, $A(X(t),V(t))=\begin{pmatrix} 
 	\frac{1}{\tau} & -\omega(D_{shore}(t),\Theta(t)) \\
@@
 \end{pmatrix}$
-with $D_{shore}(t)=d(x(t),\mathcal{D})$,
- $\Theta(t)=\widehat{(x(t)-p(t),v(t))}$ and $p(t)$ is the projection of $x(t)$ on $\partial \mathcal{D}$.
-
- Solving explicitly such a model is out of reach due to the non linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
- For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, then we can use the rough approximations
+with $D_{shore}(t)=d(X(t),\mathcal{D})$,
+ $\Theta(t)=\widehat{(X(t)-p(t),V(t))}$ and $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$.
+
+ Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
+ For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
  \[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
@@
  The question still hanging is : what are the conditions on the function $\omega(D_{shore},\Theta)$ to ensure that the process is constrained ?
- We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $x$ (see section \ref{section: simulation study}).
+ We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $x$.
  An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation.
 For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$.
-Moreover, a rotation should be considered only for some specific values of the angle $\Theta$, more precisely $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$, that is when the narwhal is heading in the direction of the shore.\\
-
- Note that this general model is very flexible, and different species-specific behaviours could be described.
+Moreover, a rotation needs to be considered only for some specific values of the angle $\Theta$, more precisely $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$, that is when the narwhal is heading in the direction of the shore.
+ The closer we are to the shore, the faster the velocity should rotate. 
+ whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
+Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
+ A weight based on the distance to the shore can be defined as follows :
+ \[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
+
+ Based on this weight, a relationship between $\omega$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :
+
+ \[
+ \omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\frac{\pi}{2}))+\tanh(\lambda(\theta-\frac{\pi}{2})))\right)\times c(D_{shore})
+ \]
+ for some $\omega_0 \in \R$, $\lambda>0$
+
+ This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
+Any smooth function that meets the minimum requirements could be used instead.
+The general model is very flexible, and different species-specific behaviours could be described.
 For the narwhals, a tendency to move along the shore could be considered by specifying a function $\omega$ that is non zero even when $\Theta \in \left[\frac{\pi}{2}-\varepsilon,\frac{\pi}{2}\right] \cup \left[-\frac{\pi}{2},-\frac{\pi}{2}+\varepsilon\right]$, so that when the animal would be on the verge of leaving the shore, its velocity would be tweaked to align with the shoreline.
 It is also possible  to specify $\tau$ similarly as a function of $D_{shore}$ and $\Theta$ to model specific behaviour near the shore (e.g higher persistence along the shoreline).\\
-
-\textcolor{red}{
-	conditions on the shore (regularity) and time steps\\
- 	discuss conditions on omega and tau, existence of solution etc
- }
-
-\subsection{Ship exposure effect}
+ Figure~\ref{fig: trueomega} shows the smooth function obtained for the parameter $\omega$ and different values of $D_0$, $\lambda$ and $\kappa$.
+
+ \begin{figure}[H]
+ 	\centering
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.3]{images/simulation study/true_omega_rect.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.3]{images/simulation study/true_omega_fjords.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
+ 		(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min. 
+ 		(b)  $D_0=0.3$ km, $\lambda=2$, $\kappa=0.2$ and $\omega_0=\frac{\pi}{2}$ rad/min.}
+ 	\label{fig: trueomega}
+ \end{figure}
+
+%conditions on the shore (regularity) and time steps
+%discuss conditions on omega and tau, existence of solution etc
+
+\subsection{Mixed effet model for ship exposure and shore effects}
 \label{subsection: ship exposure effect}
@@
 % FOR TAU: this makes sense to use additive splines since the link function is log
-To assess if the exposure to the ship has a significative effect on the narwhals movement, we first fit a baseline model only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the disturbance.
-Then, we fit a model on the data during exposure and check whether the estimated smooth parameters deviate significatively from the baseline values.
-We present here the more general baseline and response models.
-For practical work, they shall be simplified to make estimations more reliable.\\
-
-For the baseline model, only the effect of the shore is included, as discussed in section~\ref{section: CRCVM}.
-To do this, we propose to use the shore exposure covariate $E_{shore}$, defined as the inverse of the distance to the shore, and the angle $\Theta$.
-Moreover, although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance.
+To assess if the exposure to the ship has a significative effect on the narwhals movement, a baseline model is first fitted only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the disturbance.
+Then, a response model is fitted on the data during exposure and check whether the estimated smooth parameters deviate significatively from the baseline values.
+We present here the baseline and response models.\\
+
+For the baseline model, only the effect of the shore is included through the shore exposure covariate $E_{shore}$ and the angle covariate $\Theta$.\\
+Although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance.
 This is taken into account by using a mixed effect model for the RCVM parameters.
-For each narwhal $i \in \{1, \cdots, 6\}$, write $t_{exp}^{(i)}$ the first time when it gets in line of sight with the ship (start of exposure).
-Then for $t \leq t_{exp}^{(i)}$, that is before exposure to the ship, the narwhals motion processes $v^{(i)}(t)=\begin{pmatrix} v^{(i)}_1(t) & v^{(i)}_2(t) \end{pmatrix}^\top$ and $x^{(i)}(t)=\begin{pmatrix} x^{(i)}_1(t) & x^{(i)}_2(t) \end{pmatrix}^\top$ are supposed to solve equation~\ref{eq: CRCVM equation}.
+For each narwhal $i \in \{1, \cdots, N\}$, write $t_{exp}^{(i)}$ the first time when it gets in line of sight with the ship (start of exposure).
+Then for $t \leq t_{exp}^{(i)}$, that is before exposure to the ship, the narwhals motion processes $V^{(i)}(t)=\begin{pmatrix} V^{(i)}_1(t) & V^{(i)}_2(t) \end{pmatrix}^\top$ and $X^{(i)}(t)=\begin{pmatrix} X^{(i)}_1(t) & X^{(i)}_2(t) \end{pmatrix}^\top$ are supposed to solve equation~\ref{eq: CRCVM equation}.
 Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters.
-The parameter $\omega$ is expressed as a smooth function of $\Theta$ and $D_{shore}$ through splines.
-Since both covariates are not on the same scale (one is in $rad$ while the other is in $km$), tensor splines were judged more appropriate than thin plate regression splines.
+The angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $D_{shore}$ through splines.
+Since both covariates are not on the same scale (one is in $rad$ while the other is in $km$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\
+
+Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau^{(i)}$, $\nu^{(i)}$  and $\omega^{(i)}$ the parameters of the diffusion for this narwhal $i$ and  $\tau^{(i)}_j$, $\nu_j^{(i)}$, $\omega_j^{(i)}$ their values at time $t_j^{(i)}$ for $j \in \{1, \cdots,n_{i,pre}\}$,  the mixed effect model is written
 
 \begin{equation} 
+	\forall j \in \{1,\cdots, n_{i,pre}\},
 	\left\{
@@
 
-			\log(\tau^{(i)}(t))=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
-			\log(\nu^{(i)}(t))=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
-			\omega^{(i)}(t)=\omega_{0,pre}+\sum_{k=1}^{q_E \times q_{\Theta}} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t),\Theta^{(i)}(t)) + b_{\omega,pre}^{(i)}
-	\end{array}
-	\right.
-	\label{eq: baseline model}
-\end{equation}
-
-$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are unknown intercepts and $b^{(i)}_{pre}=\begin{pmatrix} b_{1,pre}^{(i)} & b_{2,pre}^{(i)} & b_{3,pre}^{(i)} \end{pmatrix}^\top \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{pre}}\right)\right)$
-where $\lambda_{pre}=\begin{pmatrix} \frac{1}{\sigma_{\tau,pre}^2} & \frac{1}{\sigma_{\nu,pre}^2} & \frac{1}{\sigma_{\omega,pre}^2} \end{pmatrix}^\top$ and the variances $\sigma_{\tau,pre}^2$, $\sigma_{\nu,pre}^2$, $\sigma_{\omega,pre}^2$ are unknown.
-
-The $\psi_k$ in equation~\ref{eq: baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
-\textcolor{red}{Add plot of the spline basis functions.}\\
-The numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$.
-It is needed to choose them before hand.
-One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.
-\\
-Based on the approximation that the covariates are constant on each time step $[t_j^{(i)},t_{j+1}^{(i)}]$, the parameters are piecewise constant.
-
-\begin{equation} 
-	\forall j \in \{1,\cdots, n_i\}
-	 \left\{
-	\begin{array}{l}
-
-		\log(\tau^{(i)}_j)=\log(\tau_{pre})+b_{1,pre}^{(i)} \\
-		\log(\nu^{(i)}_j)=\log(\nu_{pre})+b_{2,pre}^{(i)}  \\
-		\omega^{(i)}_j=\omega_{0,pre}+\sum_{k=1}^{q_E \times q_{\Theta}} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{3,pre}^{(i)}
+		\log(\tau^{(i)}_j)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
+		\log(\nu^{(i)}_j)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
+		\omega^{(i)}_j=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{\omega,pre}^{(i)}
 	\end{array}
@@
 
-In matrix form equation~\ref{eq: approximate baseline model} can be written as
+$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{pre}=\begin{pmatrix} b_{\tau,pre}^{(i)} & b_{\nu,pre}^{(i)} & b_{\omega,pre}^{(i)} \end{pmatrix}^\top \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{pre}}\right)\right)$
+are the individual parameters, where $\lambda_{pre}=\begin{pmatrix} \frac{1}{\sigma_{\tau,pre}^2} & \frac{1}{\sigma_{\nu,pre}^2} & \frac{1}{\sigma_{\omega,pre}^2} \end{pmatrix}^\top$ with variances $\sigma_{\tau,pre}^2$, $\sigma_{\nu,pre}^2$, $\sigma_{\omega,pre}^2$.
+
+The $\psi_k$ in equation~\ref{eq: approximate baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
+%Add plot of the spline basis functions \\
+$L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$.
+It is needed to choose them before hand.
+One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.
+Equation \ref{eq: approximate baseline model} is based on the approximation that the covariates are constant on each time step $[t_j^{(i)},t_{j+1}^{(i)}]$, so that the parameters are piecewise constant, which allows to form a state-space model for estimation as described in section \ref{section: inference}.\\
+In matrix form, equation~\ref{eq: approximate baseline model} can be written as
 \begin{equation}
@@
 
-$\rho_{pre}$ is the vector of the parameter values on each time step, that has size $3n$ \[\rho_{pre}=\begin{pmatrix}
-\log(\tau^{(1)}) & \cdots & \log(\tau^{(6)}) & \log(\nu^{(1)}) &
-\cdots & \log(\nu^{(6)}) & \omega^{(1)}  &\cdots & \omega^{(6)}
+where $\rho_{pre}$ is the vector of the parameter values on each time step, that has size $3n=$ number of observations $\times$ number of parameters  \[\rho_{pre}=\begin{pmatrix}
+\log(\tau^{(1)}) & \cdots & \log(\tau^{(N)}) & \log(\nu^{(1)}) &
+\cdots & \log(\nu^{(N)}) & \omega^{(1)}  &\cdots & \omega^{(N)}
 \end{pmatrix}^\top\]
 with $	\log(\tau^{(i)})=\begin{pmatrix} \log(\tau^{(i)}_1) & \cdots \log(\tau^{(i)}_{n_i})\end{pmatrix}$ and similarly for the $\log(\nu^{(i)})$ and $\omega^{(i)}$.
-$X_{pre}$ and $Z_{pre}$ are respectively the fixed and random effects model matrices with sizes $3n \times (q_E\times q_{\Theta}+3)$ and $3n \times 18$.
+$X_{pre}$ and $Z_{pre}$ are respectively the fixed and random effects model matrices with sizes $3n \times (L+3)$ and $3n \times 3N$.
+Expressions of these matrices are given in Appendix \ref{section: precisions about statistical models}.
 
@@
 	\omega_{q_E q_{\Theta},pre}
-\end{pmatrix}^\top$ is the vector of coefficients to estimate for the pre-exposure model and $b_{pre}=\begin{pmatrix}  b_{\tau,pre}^{(1)} &\cdots& b_{\tau,pre}^{(6)} &b_{\nu,pre}^{(1)}& \cdots & b_{\nu,pre}^{(6)} & b_{\omega,pre}^{(1)} & \cdots & b_{\omega,pre}^{(6)} \end{pmatrix}$ is the vector of random effects.\\
+\end{pmatrix}^\top$ is the vector of coefficients to estimate for the pre-exposure model and $b_{pre}=\begin{pmatrix}  b_{\tau,pre}^{(1)} &\cdots& b_{\tau,pre}^{(N)} &b_{\nu,pre}^{(1)}& \cdots & b_{\nu,pre}^{(N)} & b_{\omega,pre}^{(1)} & \cdots & b_{\omega,pre}^{(N)} \end{pmatrix}$ is the vector of random effects.\\
 
@@
 We illustrate this by addding a dependency on the covariate $E_{ship}$ in the parameters $\tau$ and $\nu$.
-For $i \in \{1, \cdots, 6\}$ and $t \geq t_{exp}^{(i)}$
-
-\begin{equation} \left\{
-	\begin{array}{l}
-
-		\log(\tau^{(i)}(t))=\log(\tau_{0,post}) +\sum_{k=1}^{q_{\tau}} \log(\tau_{k,post})\phi_{k}(E_{ship}(t))+b^{(i)}_{\tau,post} \\
-		\log(\nu^{(i)}(t))=\log(\nu_{0,post}) +  \sum_{k=1}^{q_{\nu}} \log(\nu_{k,post}) \xi_{k}(E_{ship}(t)) +b^{(i)}_{\nu,post}  \\
-		\omega^{(i)}(t)=\omega_{0,post}+\sum_{k=1}^{q_E\times q_{\Theta}} \omega_{k,post} \psi_k(D_{shore}(t),\Theta(t)) + b_{\omega,post}^{(i)}
+For $i \in \{1, \cdots, N\}$, denote $n_{i,post}$ the number of post-exposure observations for the narwhal $i$.
+
+\begin{equation} 
+	\forall j \in \{1,\cdots, n_{i,post}\}, \quad 
+	\left\{\begin{array}{l}
+
+		\log(\tau^{(i)}_j)=\log(\tau_{0,post}) +\sum_{k=1}^{q_{\tau}} \log(\tau_{k,post})\phi_{k}(E_{ship}(t))+b^{(i)}_{\tau,post} \\
+		\log(\nu^{(i)}_j)=\log(\nu_{0,post}) +  \sum_{k=1}^{q_{\nu}} \log(\nu_{k,post}) \xi_{k}(E_{ship}(t)) +b^{(i)}_{\nu,post}  \\
+		\omega^{(i)}_j=\omega_{0,post}+\sum_{k=1}^{q_E\times q_{\Theta}} \omega_{k,post} \psi_k(D_{shore}(t),\Theta(t)) + b_{\omega,post}^{(i)}
 	\end{array}
@@
 
-Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are unknown intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$.
-Under the approximation that parameters are piecewise constant, we can write the mixed effect models in matrix form with the same notations as above.
+Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are unknown intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$.$q_{\tau}$ and $q_{\nu}$ are the spline degrees of freedom for the exposure to the ship covariate.
+We can write the mixed effect models in matrix form with the same notations as above.
 \begin{equation}
@@
 
-Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations will be discussed in section~\ref{section: inference}.\\
+Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations are discussed in section~\ref{section: inference}.\\
 
@@
 
-As discussed in \cite{ditlevsen_hypoelliptic_2018}, estimation for such hypoelliptic models can be challenging. The main difficulties come from
+As discussed in \cite{ditlevsen_hypoelliptic_2019}, estimation for such hypoelliptic models is challenging. The main difficulties come from
 \begin{itemize}
+	\item degenerate noise structure that prevents the use of standard approximation schemes as Euler-Maruyama.
 	 \item irregular times of observations : we only get position when the marine mammal reaches the sea surface.
@@
 Several method already exist to estimate constant parameters $\tau$, $\nu$ and $\mu$.
-If the parameters are constant, it is possible to use a method of moments based on the distributional results about the velocity process \cite{gurarie_correlated_2017}, but it is only feasible when measurement errors are negligible.
-More efficiently, maximum likelihood with Kalman filter allows to estimate from observations with errors \cite{johnson_continuous_2008}.
-If the parameters are allowed to depend on time through external covariates, then \texttt{smoothSDE} package, which is based on \cite{johnson_continuous_2008} but exploit the capabilities of both the \texttt{mgcv} \cite{wood_generalized_2017} and \texttt{TMB} \cite{kristensen_tmb_2016} packages, is a convenient tool for estimation.
+If the parameters are constant, it is possible to use a method of moments based on the distributional results about the velocity process \cite{gurarie_correlated_2017}, but it is only feasible with high frequency data and negligible measurement errors.
+Maximum likelihood with Kalman filter allows to estimate more efficiently from observations with errors \cite{johnson_continuoustime_2008}.
+If the parameters are allowed to depend on time through external covariates, then \texttt{smoothSDE} package, which is based on \cite{johnson_continuoustime_2008} but exploit the capabilities of both the \texttt{mgcv} package \cite{wood_generalized_2017} to specify the time dependency through covariates, and the \texttt{TMB} \cite{kristensen_tmb_2016} package to integrate over the random effect and optimize the likelihodd, is a convenient tool for estimation.
 %Discuss the improvement/differences compared to Gurarie and Albertsen, Michelot etc
@@
 If we suppose these parameters to be constant on each time step $[t_j,t_{j+1}]$ it can be solved explicitly.
-Denote $\Delta_j=t_{j+1}-t_j$, $v_j=v(t_j)$, $x_j=x(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=A(t_j)$.
-Then applying Itô's formula gives
+Denote $\Delta_j=t_{j+1}-t_j$, $v_j=v(t_j)$, $x_j=x(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
+	\frac{1}{\tau_j} & -\omega_j \\
+	\omega_j & \frac{1}{\tau_j}
+\end{pmatrix}$. Then applying Itô's formula gives
 
@@
 $S_{\tau}$, $S_{\nu}$, $S_{\omega,E}$ and $S_{\omega,\Theta}$ are penalization matrices of known coefficients such that
-$\log(\tau_{post})^\top S_{\tau}\log(\tau_{post})$, $\log(\nu_{post})^\top S_{\nu}\log(\nu_{post})$, and $\omega_{post}^\top (S_{\omega,E}+S_{\omega,\Theta})\omega_{post}$ measure the wiggliness of the fitted splines \cite{michelot_varying-coefficient_2021}.
+$\log(\tau_{post})^\top S_{\tau}\log(\tau_{post})$, $\log(\nu_{post})^\top S_{\nu}\log(\nu_{post})$, and $\omega_{post}^\top (S_{\omega,E}+S_{\omega,\Theta})\omega_{post}$ measure the wiggliness of the fitted splines \cite{michelot_varying-coefficient_2021,wood_generalized_2017}.
 Thus, the chosen prior favours smooth functions over wiggly ones.
 Now, the mixed effect model equation can be rewritten as $\rho_{post}=X_{post}\beta_{post}+Z_{post} b_{post}$ where the fixed effect coefficients are changed to $\beta_{post}=\begin{pmatrix} \log(\tau_{0,post}) & \log(\nu_{0,post}) & \omega_{0,post} \end{pmatrix} $ and the random effects to
-\[b_{post}=\begin{pmatrix} \log(\tau_{post}) & b_{\tau,post} & \log(\nu_{post}) & b_{2,post} & \omega_{post} & b_{\omega,post}\end{pmatrix}\]
-Note that the actual random effects $b_{\tau,post}$, $b_{\nu,post}$,$b_{\omega,post}$ also enter the same framework with a "penalization" matrix which is identity and smoothing parameters that are the inverse of the variances to be estimated.
+\[b_{post}=\begin{pmatrix} \log(\tau_{post}) & b_{\tau,post} & \log(\nu_{post}) & b_{\nu,post} & \omega_{post} & b_{\omega,post}\end{pmatrix}\]
+The actual random effects $b_{\tau,post}$, $b_{\nu,post}$,$b_{\omega,post}$ also enter the same framework with a "penalization" matrix which is identity and smoothing parameters that are the inverse of the variances to be estimated.
 All these smoothing parameters are gathered in a vector $\log(\lambda_{post})$ that is estimated from the data.
@@
 \[
-L(\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)=\int_{\mathbb{R}^{N}} \exp(-\mathcal{L}_c(\pmb{b_{post}},\theta_{pre},\log(\lambda_{pre}),\log(\sigma_{obs});y) )d\pmb{b_{post}}
+L(\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)=\int_{\mathbb{R}^{d}} \exp(-\mathcal{L}_c(\pmb{b_{post}},\theta_{pre},\log(\lambda_{pre}),\log(\sigma_{obs});y) )d\pmb{b_{post}}
 \]
-where in this specific case, $N=18+q_{\tau}+q_{\nu}+q_{E}\times q_{\Theta}$.
+where in this specific case, $d=3N+q_{\tau}+q_{\nu}+L$.
 
@@
 
-\textcolor{red}{Talk about the consistency and asymptotic properties of such estimators}
+%Talk about the consistency and asymptotic properties of such estimators
 
@@
 \subsection{Simulation of constrained motion}
-In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.
-For the process to be constrained, the velocity should rotate whenever the position is close to the boundary of the domain.
-For the velocity to rotate, it is necessary to have a non zero value of $\omega$.
-Moreover, the closer we are to the shore, the faster the velocity should rotate.
-Eventually, at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
-In this case, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
-First define a weight based on the distance to the shore.
-\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
-
-Then define how $\omega$ should depend on the angle $\Theta$ and the distance to the shore $D_{shore}$ based on this weight :
-
-\[
-\omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\frac{\pi}{2}))+\tanh(\lambda(\theta-\frac{\pi}{2})))\right)\times c(D_{shore})
-\]
-for some $\omega_0 \in \R$, $\lambda>0$
-
-\textcolor{red}{Add more explanation about why we choose these formulas}
-
-Figure~\ref{fig: trueomega} shows the smooth function obtained for the parameter $\omega$ and  values different values of $D_0$, $\lambda$ and $\kappa$.
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/true_omega_rect.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/true_omega_fjords.png}
-		\caption{}
-	\end{subfigure}
-	\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-		(a) $D_0=0.3$ km, $\lambda_{\omega}=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min. 
-		(b)  $D_0=0.3$ km, $\lambda_{\omega}=2$, $\kappa=0.2$ and $\omega_0=\frac{\pi}{2}$ rad/min.}
-	\label{fig: trueomega}
-\end{figure}
+In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.\\
 
@@
 Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom was set to $5$ for each covariate.
+The marginal spline degree of freedom was set to $5$ for each covariate, which corresponds to $3$ knots in the splines.
 Hence, $25$ coefficients (including the intercept) were to be estimated.
@@
 The estimates for the long term velocity $\mu$ show that all the narwhals have a tendency to move toward the south ($\mu_2<0$) with more individual variability for the east-west direction $\mu_1$.
-However, confidence intervals are large and it might be possible to simply force $\mu$ to $(0,0)$.
+However, confidence intervals are large and  forcin $\mu$ to $(0,0)$ is also a reasonable choice.
 
@@
 
-To solve this equation, the method is similar to the 1-dimensional case.
-
-Consider $X(t)=\exp(At)v(t)$ where $\exp(At)=\sum_{n=0}^{+\infty} \frac{A^n t^n}{n!}$ is a $2 \times 2$ matrix.\\
-Thus $X(t)=g(t,v_t)$ where 
-\[\begin{array}{ccccc}
-	g & : & \mathbb{R}^+ \times \mathbb{R}^2 & \longrightarrow & \mathbb{R} \\
-	& & (t,x_1,x_2) & \longmapsto & e^{A t}
-	\begin{pmatrix} 
-		x_1 \\
-		x_2 \end{pmatrix}
-\end{array}\]
-
-The multidimensional Itô formula (theorem 4.2.1 in \cite{oksendal}) gives
-\[dX(t)=A\exp(At) v(t) dt +\exp(At) dv(t)\]
-Replacing $dv(t)$ by its expression 
-\begin{align*}
-	dX(t)&=A\exp(At) v(t) dt +\exp(At)\left(-A(v(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) \right)\\
-	&=A\exp(At) \mu dt+\frac{2\nu}{\sqrt{\pi \tau}} \exp(At)dW(t)
-\end{align*}
-Integrate from $0$ to $t$
-\[\exp(At)v(t)-v(0)=(\exp(At)-I_2) \mu +\frac{2\nu}{\sqrt{\pi \tau}} \int_{0}^{t} \exp(As) dW(s)\]
-Hence
 \begin{equation}
@@
 For $t \geq 0$ and $\Delta >0$,
-\begin{align*}
-	v(t+\Delta)&=\exp(-A(t+\Delta))v(0)+(I_2-\exp(-A(t+\Delta))) \mu +\frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} \exp(A(s-(t+\Delta)))dW(s) \\
-	&= \exp(-A\Delta) \bigg( \exp(-At) v(0) +(\exp(A\Delta)-\exp(At))\mu \\
-	&+\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t \exp(A(s-t)) dW(s) + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-t))dW(s) \bigg)
-\end{align*}
-However 
-\[\exp(-At)v(0)++\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t \exp(A(s-t)) dW(s)=v(t)-(I_2-\exp(-At))\mu\]
-Hence
-
-\begin{align*}
-	v(t+\Delta)=\exp(-A\Delta) \bigg( (\exp(A\Delta)-\exp(-At)) \mu +v(t)-(I_2-\exp(-At))\mu + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-t))dW(s) \bigg)
-\end{align*}
-
-In the end,
-
-\[v(t+\Delta)=\mu +\exp(-A\Delta)(v(t)-\mu) + \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s) \]
-
-Let $\zeta(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)$. It is a multivariate gaussian with mean zero.
-
-The variances are given by Itô's isometry.
-
-\begin{align*}
-	Var(\zeta_1(t,t+\Delta))&= \frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) \cos^2(\omega (s-(t+\Delta))) ds \\ + &\frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) \sin^2(\omega (s-(t+\Delta))) ds \\
-	&= \frac{4\nu^2}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \exp \left( 2 \frac{s-(t+\Delta)}{\tau} \right) ds \\
-	&= \frac{2\nu^2}{\pi}(1-e^{-\frac{2 \Delta}{\tau}})
-\end{align*}
-
-Similarly, 
-\[Var(\zeta_2(t,t+\Delta))=\frac{2\nu^2}{\pi}(1-e^{-\frac{2 \Delta}{\tau}})\]
-
-And the terms in $Cov(\zeta_1(t,t+\Delta),\zeta_2(t,t+\Delta))$ cancel so that 
-\[
-Cov(\zeta_1(t,t+\Delta),\zeta_2(t,t+\Delta))=0
-\]
-
-Consequently, 
+
 \[v(t+\Delta) \vert v(t) \sim \mathcal{N}\left( \mu + \exp(-A\Delta)(v(t)-\mu), \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) \]
@@
 
-We have seen that 
-\[v(t)=\exp(-At) v(0)+(I_2-\exp(-At))\mu + \zeta(0,t)\]
-where $\zeta(0,t) \sim \mathcal{N} \left(0, \frac{2\nu^2}{\pi} (1-e^{-2\frac{t}{\tau}}) I_2 \right) $ so that 
-\[v(t) \underset{t \rightarrow +\infty}{\overset{\mathcal{D}}{\longrightarrow}}\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2 \right) \]
-
 If $v(0) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ then for all $t \geq 0$, $v(t) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$
@@
 
-suppose that $t\geq \frac{3}{\beta}$ so that the velocity has reached its stationary distribution
+Suppose that $t\geq \frac{3}{\beta}$ so that the velocity has reached its stationary distribution
 \begin{align*}
@@
 %plot of the function for non zero mu
+
 \subsubsection{Velocity autocovariance function}
 
-In this section we suppose that $v(0)$ has the stationary distribution.The autocovariance function is the expected dot product of the velocity vectors over different lags.
+Suppose that $v(0)$ has the stationary distribution.The autocovariance function is the expected dot product of the velocity vectors over different lags.
 
 \[\mathbb{E}(\langle v(t), v(t+\Delta) \rangle =Cov(v_1(t)v_1(t+\Delta))+Cov(v_2(t),v_2(t+\Delta))+\mathbb{E}(v_1(t)) \mathbb{E}(v_1(t+\Delta))+\mathbb{E}(v_2(t)) \mathbb{E}(v_2(t+\Delta))\]
-
-However,
-\begin{align*}
-	Cov(v_1(t)v_1(t+\Delta))&=e^{-\frac{2t+\Delta}{\tau}} (\cos(\omega t)\cos(\omega (t+\Delta))  Var(v_1(0))+\sin(\omega t)\sin(\omega (t+\Delta))  Var(v_2(0)))\\
-	&+Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta)) \\
-	&= 2\frac{\nu^2}{\tau} \cos(\omega \Delta)  e^{-\frac{2t+\Delta}{\tau}} +Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))
-\end{align*}
-
-Then compute
-\begin{align*}
-	&Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))=\\
-	&Cov \bigg( \frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s), \\
-	& \frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \cos(\omega(s-(t+\Delta))) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \sin(\omega(s-(t+\Delta))) dW_2(s)\bigg) \\
-	&=
-	\frac{4 \nu^2}{\pi \tau} Cov \left( \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s), \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \cos(\omega(s-(t\Delta))) dW_1(s) \right) \\
-	&+ \frac{4 \nu^2}{\pi \tau} Cov \left( \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s), \int_0^{t+\Delta} e^{\frac{s-(t+\Delta)}{\tau}} \sin(\omega(s-(t\Delta))) dW_2(s) \right) \\
-	&= \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \cos(\omega(s-t)) \cos(\omega(s-(t+\Delta))) ds \\
-	& + \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \sin(\omega(s-t)) \sin(\omega(s-(t+\Delta))) ds \\
-	&= \frac{4\nu^2}{\pi \tau} \int_0^t e^{-\frac{t}{\tau}} e^{2 \frac{s-t}{\tau}} \cos(\omega \Delta) ds
-\end{align*}
-Therefore we get
-\[
-Cov(\zeta_1(0,t),\zeta_1(0,t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}} (1-e^{-2\frac{t}{\tau}})
-\]
-And eventually, 
-\[ Cov(v_1(t)v_1(t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}\]
-Similarly one gets
-\[ Cov(v_2(t)v_2(t+\Delta))=\frac{2\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}\]
 
@@
 \section{Precisions about the statistical models}
-
+\label{section: precisions about statistical models}
 In the baseline mixed effect model, the explicit model matrices are
@@
 
-\section{More simulation study}
-
-One track for each of the $6$ narwhals is simulated with known spline fixed effects of the exposure and known random effects.
-The fixed effects coefficients are estimated from the samples.
-Figure~\ref{fig: check_fe_response4_sample} shows a sample of the model described in section $3.2$ with known coefficients.
-This sample is then used to estimate the exposure effects on $\tau$ and $\nu$.
-True and estimated smooth functions are displayed in figures~\ref{fig: check_fe_response4_sample} and~\ref{fig: check_fe_response4_nu_Xtilde}.
-
 \end{document}

### 2024-04-15 R/simulation_study/simulation_study_fjords.Rmd
Commit Latex and simulation study changes

@@
 ---
-title: "simulation study in fjords"
-output: html_document
+title: "Simulation study in fjords"
+output:
+  html_document:
+    df_print: paged
+  html_notebook: default
+  pdf_document: default
 date: "2024-03-13"
@@
 
+#to fit smooth function with tensor splines
+library(mgcv)
+
 #auxiliary function to simulate from RACVM with specified smooth function for omega
@@
 #get the coastline geometry from the geojson file
-land<-st_read(file.path(greenland_data_path,"filtered_land_utm.shp"))
+land<-st_read(file.path(greenland_data_path,"land_utm.shp"))
 ```
 
-# Sample initial positions uniformly
+# SIMULATION
+
+## Sample initial positions uniformly
 
@@
 #number of tracks
-n_samples=10
+n_samples=5
 
@@
 sigma_obs_0=0.003
-H0_lf=array(rep(sigma_obs_0^2*diag(2),n_lf),dim=c(2,2,n_lf))
 H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
@@
 
-# Fix smooth parameters
+## Smooth parameters
+
+### Define smooth functions for omega and tau
 
@@
 #mixture of independent gaussian
-ftau_gauss=function(Dshore,theta,D0=0.3,tau0=1,tau1=3) {
+ftau_gauss=function(cov_data,D0=0.3,tau0=1,tau1=3) {
+  Dshore=cov_data$DistanceShore
+  theta=cov_data$theta
   tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/10))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/10))^2-(Dshore/D0)^2)
@@
 
-fomega=function(Dshore,theta,D0=0.3,omega0=60*pi/2,lambda=2,kappa=0.2) {
+fomega=function(cov_data,D0=0.3,omega0=60*pi/2,lambda=2,kappa=0.2) {
+    Dshore=cov_data$DistanceShore
+    theta=cov_data$theta
+    if (is.null(Dshore)){
+      Dshore=1/cov_data$ExpShore
+    }
     coeff=exp(-kappa*(Dshore/D0)^2)
@@
 
-Plot smooth functions.
+### Plot smooth functions.
+
 ```{r plot smooth parameters }
@@
 
-# Estimation when omega only is smooth
+### Approximate smooth functions with tensor splines of DistanceShore
 
-## Generate samples
+```{r}
+sp_df=5
+#number of points
+n <- 500
 
-### High frequency and no measurement error
+#sample theta and DistanceShore points in the domain
+theta <- runif(n,-pi,pi);
+DistanceShore <- runif(n,0.2,2);
+samples=data.frame(theta=theta,DistanceShore=DistanceShore)
+
+#define grid of values
+theta_v <- seq(-pi,pi,length=30)
+Dshore_v<- seq(0.2,2,length=30)
+pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
+
+# true values of the function over this grid
+truth <- matrix(fomega(pr),30,30)
+
+#points on the surface perturbed by gaussian noise
+f <- fomega(samples)
+y <- f+0*rnorm(n)
+
+#plot true function
+persp(theta_v,Dshore_v,truth);
+title("truth")
+
+#fit with bivariate splines te
+m1 <- gam(y~te(theta,DistanceShore,k=sp_df))
+
+#visualize
+vis.gam(m1);title("tensor product")
+
+df=data.frame("x"=theta,"z"=DistanceShore,"y"=y)
+fig <- plot_ly(df,type="scatter3d", x = ~x, y = ~z, z = ~y,mode="markers",marker=list(size=4))
+fig
+```
+
+### Approximate smooth functions with tensor splines of ExpShore
+
+```{r}
+ExpShore=1/DistanceShore
+
+#fit with bivariate splines te
+m2 <- gam(y~te(theta,ExpShore,k=sp_df))
+
+#visualize
+vis.gam(m2);title("tensor product")
+
+```
+
+### Get estimated spline coefficients
+
+```{r}
+sp_coeff_Dshore=m1$coefficients
+sp_coeff_ExpShore=m2$coefficients
+```
+
+## Simulation with measurement errors and high frequency
+
+### Generate samples
 
@@
 plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(y1,y2),size=0.1)+
-  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(p1,p2),size=0.01,col="red",shape=5)+
-  geom_path(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==5,],mapping=aes(y1,y2),linewidth=0.1)+
-  geom_point(data = data_fjords_hf_ne[data_fjords_hf_ne$ID==5,] %>% filter(!duplicated(ID)),
+  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(y1,y2),size=0.1)+
+  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(p1,p2),size=0.01,col="red",shape=5)+
+  geom_path(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(y1,y2),linewidth=0.1)+
+  geom_point(data = data_fjords_hf_ne[data_fjords_hf_ne$ID==2,] %>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4,col = "blue")+
@@
 
-### Low frequency and no measurement error
+## sImulate with high frequency and measurement error
 
-```{r subsample hf ne samples to low frequency (lf ne)}
-data_fjords_lf_ne=data_fjords_hf_ne[seq(1,nrow(data_fjords_hf_ne),by=5),]
-```
-
-### Plot samples
-
-```{r plot lf ne samples}
-
-plot_samples_fjords_lf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_point(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*p1,1000*p2),size=0.01,col="red",shape=5)+
-  geom_path(data=data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-  geom_point(data = data_fjords_lf_ne[data_fjords_lf_ne$ID==6,] %>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4, col = "blue")+
-  xlab("x") + ylab("y")
-
-ggsave(filename="samples_fjords_lf_ne_omega.png",plot=plot_samples_fjords_lf_ne)
-ggplotly(plot_samples_fjords_lf_ne)
-```
-
-### High frequency and measurement error
+### Generate samples
 
@@
 
+### Percentage of samples that reached land
 ```{r percentage of hf e samples that reached land}
@@
 
-### Plot samples
+### Plot one sample
 
@@
 plot_samples_fjords_hf_e=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_e[data_fjords_hf_e$ID==8,],mapping=aes(y1,y2),size=0.1)+
-   geom_path(data=data_fjords_hf_e[data_fjords_hf_e$ID==8,],mapping=aes(y1,y2),linewidth=0.1)+
-  geom_point(data = data_fjords_hf_e[data_fjords_hf_e$ID==8,] %>% filter(!duplicated(ID)),
+  geom_point(data=data_fjords_hf_e[data_fjords_hf_e$ID==2,],mapping=aes(y1,y2),size=0.1)+
+   geom_path(data=data_fjords_hf_e[data_fjords_hf_e$ID==2,],mapping=aes(y1,y2),linewidth=0.1)+
+  geom_point(data = data_fjords_hf_e[data_fjords_hf_e$ID==2,] %>% filter(!duplicated(ID)),
              aes(x = y1, y = y2 ), shape = 3, size = 4,col = "red")+
@@
 
-### Low frequency and measurement error
-
-```{r subsamble hf e samples to low frequency (lf e)}
-
-data_fjords_lf_e=data_fjords_hf_e[seq(1,nrow(data_fjords_hf_e),by=5),]
-
-```
-
-### Plot samples
-
-```{r plot lf e samples}
-
-plot_samples_fjords_lf_e=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_lf_e[data_fjords_lf_e$ID==8,],mapping=aes(1000*y1,1000*y2),size=0.1)+
-   geom_path(data=data_fjords_lf_e[data_fjords_lf_e$ID==8,],mapping=aes(1000*y1,1000*y2),linewidth=0.1)+
-  geom_point(data = data_fjords_lf_e[data_fjords_lf_e$ID==8,] %>% filter(!duplicated(ID)),
-             aes(x = y1* 1000, y = y2 * 1000), shape = 3, size = 4,col = "red")+
-  xlab("x") + ylab("y")
-
-ggsave(filename="samples_fjords_lf_e_omega.png",plot=plot_samples_fjords_lf_e)
-ggplotly(plot_samples_fjords_lf_e)
-```
+# ESTIMATION
 
@@
 data_fjords_hf_ne=add_covs(data_fjords_hf_ne)
-data_fjords_lf_ne=add_covs(data_fjords_lf_ne)
 
 data_fjords_hf_e=add_covs(data_fjords_hf_e)
-data_fjords_lf_e=add_covs(data_fjords_lf_e)
 ```
 
-## Estimations
+## Estimate from high frequency data without error : ExpShore and tensor splines ti
 
-### High frequency data without error : ExpShore and tensor splines ti
+### Use ExpShore covariate and tensor splines te
+
+#### Fit sde
 
@@
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
+                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_fjords_hf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[data_fjords_hf_ne$ID==6,],type = "RACVM1",
+crcvm_fjords_hf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 ```
+
+#### Plot estimated smooth parameter omega in 3D
 
@@
 link=list("ExpShore"=(\(x) 1/x))
-res=crcvm_fjords_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-res$plot_omega_theta_ExpShore
+true_smooths=list("omega"=fomega)
+res=crcvm_fjords_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne1",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_ExpShore
 ```
 
-### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.
+#### Violin plot of the estimated spline coefficients
 
-```{r compare hf e estimates with Expshore to true smooth parameters for fixed values of ExpShore}
+```{r}
 
-#plot spline estimates of the parameter omega close and far to the shore
+true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
+rownames(true_coeff)=NULL
 
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/1,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))
+p=crcvm_fjords_hf_ne1$plot_violin_coeff(model_name="crcvm_fjords_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-sde_par_close <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_far)
+ggplotly(p)
+```
 
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
+#### Simulate from fitted model
 
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
+```{r,cache=TRUE}
 
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(1,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))
+fExpShore=function(z,v,p) {
 
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 1 km")
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
 
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 3 km")
+  #distance to shore
+  Dshore=sqrt(normal[1]^2+normal[2]^2)
+  if (Dshore>0.5){
+    return (0)
+  }
+  else if (Dshore<0.01) {
+    return (100)
+  }
+  else {
+    return (1/Dshore)
+  }
+}
 
-plot_omega_close
-plot_omega_far
+fangle=function(z,v,p) {
 
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_close.png",plot=plot_omega_close)
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_far.png",plot=plot_omega_far)
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #return angle
+  return (signed_angle(normal,v))
+}
+
+atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
+sim_data=crcvm_fjords_hf_ne1$simulate(z0=x0[1,],data=crcvm_fjords_hf_ne1$data(),atw=atw,land=land)
 
@@
 
-### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.
+#### Plot simulation from fitted model
 
-```{r compare hf e estimates with Expshore to true smooth parameters for fixed values of Theta}
+```{r}
 
-#plot spline estimates of the parameter omega close and far to the shore
+plot_sim=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_point(data = sim_data%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2 ), shape = 3, size = 4, col = "red")+
+    xlab("x") + ylab("y")
 
-predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
-predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))
-
-sde_par_away <- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_away)
-sde_CI_away <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_away)
-sde_par_toward<- crcvm_fjords_hf_ne1$par(t = "all",new_data=predict_toward)
-sde_CI_toward <- crcvm_fjords_hf_ne1$CI_simultaneous(t = "all",new_data=predict_toward)
-
-# Data frame for point estimates
-sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
-sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
-                          lowomega = sde_CI_away["omega", "low",],
-                          upomega = sde_CI_away["omega", "upp",])
-sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
-                          lowomega = sde_CI_toward["omega", "low",],
-                          upomega = sde_CI_toward["omega", "upp",])
-
-omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
-                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))
-
-plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
-  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
-  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))
-
-plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
-  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
-  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))
-
-plot_omega_away
-plot_omega_toward
-
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_away.png",plot=plot_omega_away)
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_toward.png",plot=plot_omega_toward)
-
+ggsave("sim_crcvm_fjords_hf_ne1.png",plot=plot_sim,width=10,height=5,path="crcvm_fjords_hf_ne1")
+ggplotly(plot_sim)
 ```
 
-### Low frequency data without error
+###  UseDistanceShore covariate and tensor splines ti
 
-```{r estimate from lf ne samples with cov ExpShore}
+#### Fit sde
 
+```{r,cache=TRUE}
 #estimation with tensor splines
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
+                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_fjords_lf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_lf_ne[data_fjords_lf_ne$ID==6,],type = "RACVM1",
+crcvm_fjords_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("log_sigmaobs0"=log(0.003)))
-crcvm_fjords_lf_ne1$fit(method="BFGS")
+                      other_data=list("H"=H0_hf))
+crcvm_fjords_hf_ne2$fit(method="BFGS")
 ```
 
-```{r get plots from lf ne estimates with Expshore}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.3)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-res=crcvm_fjords_lf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_lf_ne1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-res$plot_omega_theta_DistanceShore
+#### Plots estimated smooth in 3D
+
+```{r}
+xmin=list("theta"=-pi,"DistanceShore"=0.3)
+xmax=list("theta"=pi,"DistanceShore"=2)
+res=crcvm_fjords_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne2",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_DistanceShore
 ```
 
-### High frequency data with error : ExpShore and tensor splines te
+#### Violin plot of the estimated spline coefficients
 
-```{r estimate from hf e samples with cov ExpShore}
+```{r}
 
+true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
+rownames(true_coeff)=NULL
+
+p=crcvm_fjords_hf_ne2$plot_violin_coeff(model_name="crcvm_fjords_hf_ne2",substring="omega",true_coeff=true_coeff,save=TRUE)
+
+ggplotly(p)
+```
+
+## Estimate from high frequency data with error :
+
+### Use ExpShore and tensor splines te
+
+#### Fit sde
+
+```{r, cache=TRUE}
 #estimation with tensor splines
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
+                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_fjords_hf_e1 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==8,],type = "RACVM1",
+crcvm_fjords_hf_e1 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==1,],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
@@
 
-```{r get plots from hf e estimates with Expshore}
+#### Plot estimated smooth in 3D
+
+```{r}
 xmin=list("theta"=-pi,"ExpShore"=1/2)
@@
 link=list("ExpShore"=(\(x) 1/x))
-res=crcvm_fjords_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-res$plot_omega_theta_ExpShore
+true_smooths=list("omega"=fomega)
+res=crcvm_fjords_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e1",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_ExpShore
 ```
 
-### Plot true and estimated omega as a function of theta for fixed values of DistanceShore, with confidence intervals.
+#### Violin plot of the estimated spline coefficients
 
@@
 
-#plot spline estimates of the parameter omega close and far to the shore
+true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
+rownames(true_coeff)=NULL
 
-predict_close=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/0.5,100))
-predict_far=data.frame(theta=seq(-pi,pi,length.out=100),ExpShore=rep(1/3,100))
+p=crcvm_fjords_hf_e1$plot_violin_coeff(model_name="crcvm_fjords_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-sde_par_close <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_close)
-sde_CI_close <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_close)
-sde_par_far <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_far)
-sde_CI_far <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_far)
-
-# Data frame for point estimates
-sde_par_df_close <- data.frame(theta=predict_close$theta,omega_estimates = sde_par_close[, "omega"])
-sde_par_df_far<- data.frame(theta=predict_far$theta,omega_estimates = sde_par_far[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_close<- data.frame(theta=predict_close$theta,
-                          lowomega = sde_CI_close["omega", "low",],
-                          upomega = sde_CI_close["omega", "upp",])
-sde_ci_df_far<- data.frame(theta=predict_far$theta,
-                          lowomega = sde_CI_far["omega", "low",],
-                          upomega = sde_CI_far["omega", "upp",])
-
-omega_df=data.frame(theta=predict_close$theta,omega_close=fomega(rep(0.5,100),predict_close$theta),omega_far=fomega(rep(3,100),predict_close$theta))
-
-plot_omega_close=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_close) + 
-  geom_ribbon(data=sde_ci_df_close,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_close),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 0.5 km")
-
-plot_omega_far=ggplot() +geom_line(aes(theta, omega_estimates),data=sde_par_df_far) + 
-  geom_ribbon(data=sde_ci_df_far,aes(x=theta,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(theta,omega_far),col="red")+xlab("Theta")+
-  ylab("omega")+ggtitle("Distance to shore = 3 km")
-
-plot_omega_close
-plot_omega_far
-
-ggsave(filename="crcvm_fjords_hf_e1/crcvm_fjords_hf_e1_omega_close.png",plot=plot_omega_close)
-ggsave(filename="crcvm_fjords_hf_e1/crcvm_fjords_hf_e1_omega_far.png",plot=plot_omega_far)
-
+ggplotly(p)
 ```
 
-### Plot true and estimated omega as a function of DistanceShore for fixed values of theta, with confidence intervals.
+### Use DistanceShore and tensor splines te
+
+#### Fit sde
+
+```{r,cache=TRUE}
+#estimation with tensor splines
+formulas <- list(mu1=~1,mu2=~1,tau=~1,
+                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
+par0 <- c(0,0,1,4,0)
+
+crcvm_fjords_hf_e2 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==1,],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
+                      other_data=list("H"=H1_hf))
+crcvm_fjords_hf_e2$fit(method="BFGS")
+```
+
+#### Plot estimated smooth in 3D
+
+```{r}
+xmin=list("theta"=-pi,"DistanceShore"=0.3)
+xmax=list("theta"=pi,"DistanceShore"=2)
+res=crcvm_fjords_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e2",xmin=xmin,xmax=xmax,
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+res$fe_omega_theta_DistanceShore
+```
+
+#### Violin plot of the estimated spline coefficients
 
@@
 
-#plot spline estimates of the parameter omega close and far to the shore
+true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
+rownames(true_coeff)=NULL
 
-predict_away=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(pi/3,100))
-predict_toward=data.frame(ExpShore=seq(1/2,1/0.3,length.out=100),theta=rep(2*pi/3,100))
+p=crcvm_fjords_hf_e2$plot_violin_coeff(model_name="crcvm_fjords_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
 
-sde_par_away <- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_away)
-sde_CI_away <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_away)
-sde_par_toward<- crcvm_fjords_hf_e1$par(t = "all",new_data=predict_toward)
-sde_CI_toward <- crcvm_fjords_hf_e1$CI_simultaneous(t = "all",new_data=predict_toward)
-
-# Data frame for point estimates
-sde_par_df_away<- data.frame(ExpShore=predict_away$ExpShore,omega_estimates = sde_par_away[, "omega"])
-sde_par_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,omega_estimates = sde_par_toward[, "omega"])
-
-# Data frame for CIs
-sde_ci_df_away<- data.frame(ExpShore=predict_away$ExpShore,
-                          lowomega = sde_CI_away["omega", "low",],
-                          upomega = sde_CI_away["omega", "upp",])
-sde_ci_df_toward<- data.frame(ExpShore=predict_toward$ExpShore,
-                          lowomega = sde_CI_toward["omega", "low",],
-                          upomega = sde_CI_toward["omega", "upp",])
-
-omega_df=data.frame(DistanceShore=1/predict_away$ExpShore,omega_away=fomega(1/predict_away$ExpShore,rep(pi/3,100)),
-                    omega_toward=fomega(1/predict_toward$ExpShore,rep(2*pi/3,100)))
-
-plot_omega_away=ggplot()+geom_line(aes(1/ExpShore,omega_estimates),data=sde_par_df_away) + 
-  geom_ribbon(data=sde_ci_df_away,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_away),col="red")+xlab("Distance to shore (km)")+
-  ylab(TeX(r'(\omega)'))+ggtitle(TeX(r'(\Theta = \pi/3)'))
-
-plot_omega_toward=ggplot()+geom_line(aes(1/ExpShore, omega_estimates),data=sde_par_df_toward) + 
-  geom_ribbon(data=sde_ci_df_toward,aes(x=1/ExpShore,ymin=lowomega,ymax=upomega),alpha=0.2)+
-  geom_line(data=omega_df,aes(DistanceShore,omega_toward),col="red")+xlab("Distance to shore (km)")+
-  ylab("Omega")+ggtitle(TeX(r'(\Theta = 2\pi/3)'))
-
-plot_omega_away
-plot_omega_toward
-
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_away.png",plot=plot_omega_away)
-ggsave(filename="crcvm_fjords_hf_ne1/crcvm_fjords_hf_ne1_omega_toward.png",plot=plot_omega_toward)
-
+ggplotly(p)
 ```
-
-### Low frequency data with error : ExpShore and tensor splines ti
-
-```{r}
-
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~ti(theta,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_fjords_lf_e1 <- SDE$new(formulas = formulas,data = data_fjords_lf_e[data_fjords_lf_e$ID==1,],type = "RACVM1",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_lf))
-crcvm_fjords_lf_e1$fit(method="BFGS")
-```
-
-```{r}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.2)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-crcvm_fjords_lf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_lf_e1",npost=1000,level=0.95,xmin=xmin,xmax=xmax,xlabel=xlabel,link=link)
-
-```

### 2024-04-15 R/simulation_study/simulation_study_rect.Rmd
Commit Latex and simulation study changes

@@
   pdf_document: default
+  date: "2024-03-13"
 ---
@@
 
-#used in function split_term to get covariates names from the formulas
-library(stringr)
-
 #to fit smooth function with tensor splines
@@
 source("CVM_functions.R")
+```
+
+# global variables
+
+```{r}
+#number of tracks
+N_BATCH=1
+N_ID=3
+
+#degree of freedom in tensor splines
+SP_DF=5
 ```
@@
 
-#number of tracks
-n_samples=1
-
 #initial velocity and location
@@
 #generate random initial points
-x0=matrix(rep(NA,n_samples*2),ncol=2)
+x0=matrix(rep(NA,N_ID*2),ncol=2)
 colnames(x0)=c("x1","x2")
 
-for (i in 1:n_samples) {
+for (i in 1:N_ID) {
   #choose location uniformly in the domain
@@
 }
+
 ```
@@
 #observation times
-Tmax=3*24 #max time in hour
+Tmax=1*24 #max time in hour
 
@@
 sigma_obs_0=0.005
-H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
+H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
 #GPS measurement error 
 sigma_obs_1=0.03
-H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf),dim=c(2,2,n_hf))
+H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
 
@@
 #constant tau
-ftau_constant=function(cov_data,tau=1) {
-  return (tau)
+ftau_constant=function(cov_data,tau=1,sigma_re=0.1) {
+  b=rnorm(1,mean=0,sd=sigma_re)
+  return (tau+b)
 }
@@
 
-nu=4
+#constant nu
+fnu_constant=function(cov_data,nu=1,sigma_re=0.3) {
+    b=rnorm(1,mean=0,sd=sigma_re)
+  return (nu+b)
+}
+
 ```
@@
 ```{r}
-sp_df=5
+SP_DF=5
 #number of points
@@
 #fit with bivariate splines te
-m1 <- gam(y~te(theta,DistanceShore,k=sp_df))
+m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))
 
@@
 #fit with bivariate splines te
-m2 <- gam(y~te(theta,ExpShore,k=sp_df))
+m2 <- gam(y~te(theta,ExpShore,k=SP_DF))
 
@@
 ```
+
 ### Get estimated spline coefficients
@@
 ### Generate sample
+
 ```{r, cache=TRUE,results='hide'}
@@
 #initialize data frame
-data_rect_hf_ne=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+data_rect_hf_ne=list()
 
-for (i in 1:n_samples) {
+for (k in 1:N_BATCH) {
 
-  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
+  data=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+
+  for (i in 1:N_ID) {
+
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
 
@@
     data_shore=res$shore[,c("p1","p2")]
-  data_rect_hf_ne=rbind(data_rect_hf_ne,cbind(data_sim,data_shore))
+    data=rbind(data,cbind(data_sim,data_shore))
   }
@@
   #for the first value we cannot compute the angle theta
-data_rect_hf_ne=data_rect_hf_ne[-1,]
+  data=data[-1,]
+  data_rect_hf_ne[[paste("BATCH",k,sep="")]]=data
+}
 
@@
 count=0
-for (id in unique(data_rect_hf_ne$ID)){
-  sub_data=data_rect_hf_ne[data_rect_hf_ne$ID==id,]
+for (k in 1:length(names(data_rect_hf_ne))) {
+
+  data=data_rect_hf_ne[[paste("BATCH",k,sep="")]]
+  for (id in unique(data$ID)) {
+    sub_data=data[data$ID==id,]
     if (nrow(sub_data) < n_hf) {
-    print(id)
       count=count+1
+      cat("BATCH",k,id,"\n",sep="")
     }
   }
-cat(count/n_samples*100,"percent of the samples reached land")
+}
+cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
 ```
 
-### plot one sample
+### plot one batch
 
@@
 plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(y1,y2),size=0.1)+
-  geom_path(data=data_rect_hf_ne[data_rect_hf_ne$ID==1,],mapping=aes(y1,y2),size=0.1)+
-  geom_point(data = data_rect_hf_ne[data_rect_hf_ne$ID==1,]%>% filter(!duplicated(ID)),
+  geom_point(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_rect_hf_ne$BATCH1%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
@@
 #initialize data frame
-data_rect_hf_e=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+data_rect_hf_e=list()
 
-for (i in 1:n_samples) {
+for (k in 1:N_BATCH) {
 
-  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
+  data=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+
+  for (i in 1:N_ID) {
+
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
 
@@
     data_shore=res$shore[,c("p1","p2")]
-  data_rect_hf_e=rbind(data_rect_hf_e,cbind(data_sim,data_shore))
+    data=rbind(data,cbind(data_sim,data_shore))
   }
@@
   #for the first value we cannot compute the angle theta
-data_rect_hf_e=data_rect_hf_e[-1,]
-
+  data=data[-1,]
+  data_rect_hf_e[[paste("BATCH",k,sep="")]]=data
+}
 ```
@@
 count=0
-for (id in unique(data_rect_hf_ne$ID)){
-  sub_data=data_rect_hf_e[data_rect_hf_e$ID==id,]
+for (k in 1:length(names(data_rect_hf_e))) {
+
+  data=data_rect_hf_e[[paste("BATCH",k,sep="")]]
+  for (id in unique(data$ID)) {
+    sub_data=data[data$ID==id,]
     if (nrow(sub_data) < n_hf) {
-    print(id)
       count=count+1
+      cat("BATCH",k,id,"\n",sep="")
     }
   }
-cat(count/n_samples*100,"percent of the samples reached land")
+}
+cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
 ```
@@
 plot_samples_rect_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(y1,y2),size=0.1)+
-  geom_path(data=data_rect_hf_e[data_rect_hf_e$ID==1,],mapping=aes(y1,y2),size=0.1)+
-  geom_point(data = data_rect_hf_e[data_rect_hf_e$ID==1,]%>% filter(!duplicated(ID)),
+  geom_point(data=data_rect_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_rect_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_rect_hf_e$BATCH1%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
@@
   #data has at least columns "time","y1" "y2", "p1","p2"
-  #this function modifies data to add columns "theta", "DistanceShore" and "ExpShore"
+  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"
 
@@
 
-data_rect_hf_ne=add_covs(data_rect_hf_ne)
+for (k in 1:N_BATCH) {
 
-data_rect_hf_e=add_covs(data_rect_hf_e)
+  #dataframes for this batch
+  df_ne=data_rect_hf_ne[[paste("BATCH",k,sep="")]]
+  df_e=data_rect_hf_e[[paste("BATCH",k,sep="")]]
+
+  #add covariates columns
+  df_ne=add_covs(df_ne)
+  df_e=add_covs(df_e)
+
+  #change in the list of batch
+  data_rect_hf_ne[[paste("BATCH",k,sep="")]]=df_ne
+  data_rect_hf_e[[paste("BATCH",k,sep="")]]=df_e
+}
+
 ```
 
-## Estimate from high frequency data without low error
+## Estimate from high frequency data with low error
 
@@
 #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=sp_df,bs="cs"))
+formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
 par0 <- c(0,0,1,4,0)
 
-crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
+crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne$BATCH1,type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
                       other_data=list("H"=H0_hf))
@@
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
+                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
 par0 <- c(0,0,1,4,0)

### 2024-04-16 R/simulation_study/simulation_study_rect.Rmd
parallelization in simulation_study_rect + other stuff

@@
 
-# Import libraries and data
+# SET UP
 
-```{r}
+## Import libraries
+
+```{r libraries}
 #seed for reproducibility 
 set.seed(1)
+
+#parallel computing
+library(foreach)
+library(doParallel)
 
@@
 
-# global variables
+## global variables
 
-```{r}
-#number of tracks
-N_BATCH=1
+```{r global variables}
+#number of batch
+N_BATCH=10
+
+#number of individual tracks per batch
 N_ID=3
+
+#duration of each track in hour
+TMAX=1*24
 
@@
 We create a square with sides of length 8km.
-```{r}
+
+```{r create rectangle}
 
@@
 
-```{r}
+```{r plot rectangle}
 
@@
 
-```{r}
+```{r initial positions}
 
@@
 
-```{r}
-#observation times
-Tmax=1*24 #max time in hour
+```{r time steps}
+#time step
+hf=1/60
 
-hf=1/60 #1 minute time step
-
-#high frequency time steps
-times_hf=seq(0,Tmax,by=hf)
+#high frequency time points
+times_hf=seq(0,TMAX,by=hf)
 n_hf=length(times_hf)-1
@@
 
-```{r}
+```{r measurement error}
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
@@
 
-```{r}
+```{r smooth parameters}
 
@@
 #constant nu
-fnu_constant=function(cov_data,nu=1,sigma_re=0.3) {
+fnu_constant=function(cov_data,nu=4,sigma_re=0.3) {
     b=rnorm(1,mean=0,sd=sigma_re)
@@
 
-```{r}
+```{r plot smooth parameters}
 
@@
 
-```{r}
-SP_DF=5
+```{r approximate smooth omega with DistanceShore splines}
+
 #number of points
-n <- 500
+n <- 1000
 
@@
 
-```{r}
+```{r approximate smooth omega with ExpShore splines}
 ExpShore=1/DistanceShore
@@
 
-```{r}
+```{r estimated spline coeffs}
 sp_coeff_Dshore=m1$coefficients
@@
 
-## Simulation with measurement errors and high frequency
+### Define spline smooth functions for omega based on the approximations
+
+```{r spline smooth omega definition}
+
+fomega_splines=function(cov_data) {
+
+  if (is.null(cov_data$ExpShore)){
+    omega=predict(m1,newdata=cov_data)
+  }
+  else if (is.null(cov_data$DistanceShore)){
+    omega=predict(m2,newdata=cov_data)
+  }
+
+  return(as.numeric(omega))
+}
+
+```
+
+### Plot
+
+```{r plot spline smooth omega}
+
+Dshore=seq(from=0.3,to=2,length.out=30)
+theta=seq(from=-pi,to=pi,length.out=30)
+grid<- as.data.frame(expand.grid(DistanceShore,theta))
+colnames(grid)=c("DistanceShore","theta")
+true_par=grid
+true_par$omega=fomega_splines(grid)
+
+plot_true_omega <- plot_ly(true_par,type = "mesh3d",
+                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
+                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_true_omega
+```
+
+## Simulation with low measurement errors and high frequency
 
@@
 
-```{r, cache=TRUE,results='hide'}
+```{r parallel generate hf low errors samples}
 
-#initialize data frame
-data_rect_hf_ne=list()
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
 
-for (k in 1:N_BATCH) {
-
-  data=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
-
-  for (i in 1:N_ID) {
+data_rect_hf_ne <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+  data=foreach (i=1:N_ID,.combine='rbind') %do% {
 
@@
     data_shore=res$shore[,c("p1","p2")]
-    data=rbind(data,cbind(data_sim,data_shore))
+    cbind(data_sim,data_shore)
+  }
+  data 
 }
 
-  #for the first value we cannot compute the angle theta
-  data=data[-1,]
-  data_rect_hf_ne[[paste("BATCH",k,sep="")]]=data
-}
+#stop cluster
+stopCluster(cl)
 
+names(data_rect_hf_ne)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 
-```{r}
+```{r percentage of low error samples that reached land}
 count=0
@@
       count=count+1
-      cat("BATCH",k,id,"\n",sep="")
+      cat(paste("BATCH",k,sep=""),":",id,"\n",sep=" ")
     }
@@
 
-```{r}
+```{r plot one batch low error samples}
 
@@
 
-```{r, cache=TRUE,results='hide'}
+```{r high error samples, cache=TRUE,results='hide'}
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
 
-#initialize data frame
-data_rect_hf_e=list()
-
-for (k in 1:N_BATCH) {
-
-  data=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
-
-  for (i in 1:N_ID) {
+data_rect_hf_e <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+  data=foreach (i=1:N_ID,.combine='rbind') %do% {
 
@@
     data_shore=res$shore[,c("p1","p2")]
-    data=rbind(data,cbind(data_sim,data_shore))
+    cbind(data_sim,data_shore)
+  }
+  data 
 }
 
-  #for the first value we cannot compute the angle theta
-  data=data[-1,]
-  data_rect_hf_e[[paste("BATCH",k,sep="")]]=data
-}
+#stop cluster
+stopCluster(cl)
+
+#change names of the list elements
+names(data_rect_hf_e)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 
-```{r}
+```{r percentage of high error samples that reached land}
 count=0
@@
 
-```{r}
+```{r plot one batch of high error samples}
 
@@
 
-```{r}
+```{r compute covs}
 
@@
 
-```{r,cache=TRUE}
+```{r estimate te ExpShore for low error samples,cache=TRUE}
 
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_rect_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
@@
                  nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_rect_hf_ne1 <- SDE$new(formulas = formulas,data = data_rect_hf_ne$BATCH1,type = "RACVM",
+  par0 <- c(0,0,1,1,0)
+  m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
                       other_data=list("H"=H0_hf))
-crcvm_rect_hf_ne1$fit(method="BFGS")
+  m$fit(method="BFGS")
+  m 
+}
+
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_rect_hf_ne1)=paste("BATCH",1:N_BATCH,sep="")
+
 ```
 
-#### Plot estimated smooth parameter omega in 3D
+#### Plot estimated smooth parameter omega in 3D for one batch
 
-```{r}
+```{r plot estimates te ExpShore for one batch of low error samples}
 xmin=list("theta"=-pi,"ExpShore"=1/2)
@@
 link=list("ExpShore"=(\(x) 1/x))
-true_smooths=list("omega"=fomega)
-res=crcvm_rect_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",xmin=xmin,xmax=xmax,
+true_smooths=list("omega"=fomega_splines)
+res=crcvm_rect_hf_ne1$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-#### Violin plot of the estimated spline coefficients
+#### Violin plot of the estimated spline coefficients for one batch
 
-```{r}
+```{r violin plots spline coeff te ExpShore for one batch of low error samples}
 
-true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
+true_coeff=data.frame(value=sp_coeff_ExpShore,name=c(paste("omega.",names(sp_coeff_ExpShore),sep="")))
 rownames(true_coeff)=NULL
 
-p=crcvm_rect_hf_ne1$plot_violin_coeff(model_name="crcvm_rect_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_rect_hf_ne1$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
+
+ggplotly(p)
+```
+
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te ExpShore all batches with low measurement error}
+
+n_coeffs=length(sp_coeff_ExpShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_rect_hf_ne1[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+ggsave("all_batches_coeffs_crcvm_rect_hf_ne1.png",plot=p,path="crcvm_rect_hf_ne1")
 
@@
 
-```{r,cache=TRUE}
+```{r simulate fitted te ExpShore low error samples,cache=TRUE}
 
@@
   Dshore=sqrt(normal[1]^2+normal[2]^2)
-  if (Dshore>0.5){
-    return (0)
-  }
-  else if (Dshore<0.01) {
+  if (Dshore<0.01) {
     return (100)
@@
 atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
-sim_data=crcvm_rect_hf_ne1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne1$data(),atw=atw,land=domain)
+sim_data=crcvm_rect_hf_ne1$BATCH1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne1$BATCH1$data(),atw=atw,land=domain)
 
@@
 
-```{r}
+```{r plot sim te ExpSHore low error samples}
 
@@
 
-###  UseDistanceShore covariate and tensor splines ti
+###  Use DistanceShore covariate and tensor splines te
 
@@
 
-```{r,cache=TRUE}
+```{r estimate te DistanceShore for low error samples,cache=TRUE}
 #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-par0 <- c(0,0,1,4,0)
+formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
 
-crcvm_rect_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
+crcvm_rect_hf_ne2=list()
+for (k in 1:N_BATCH) {
+  m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH"),k,sep=""]],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
                       other_data=list("H"=H0_hf))
-crcvm_rect_hf_ne2$fit(method="BFGS")
+  m$fit(method="BFGS")
+  crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]<-m
+}
 ```
@@
 
-```{r}
+```{r plot estimates te DistanceShore for one batch of low error samples}
 xmin=list("theta"=-pi,"DistanceShore"=0.3)
 xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_rect_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",xmin=xmin,xmax=xmax,
+res=crcvm_rect_hf_ne2$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-```{r}
+```{r violin plots spline coeffs te DistanceShore for one batch of low error sample}
 
-true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
+true_coeff=data.frame(value=sp_coeff_Dshore,name=c(paste("omega.",names(sp_coeff_Dshore),sep="")))
 rownames(true_coeff)=NULL
 
-p=crcvm_rect_hf_ne2$plot_violin_coeff(model_name="crcvm_rect_hf_ne2",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_rect_hf_ne2$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_ne2",substring="omega",
+                                             true_coeff=true_coeff,save=TRUE)
 
@@
 
-## Estimate from high frequency data with error :
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
+
+n_coeffs=length(sp_coeff_DShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+ggsave("all_batches_coeffs_crcvm_rect_hf_ne2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_ne2")
+
+```
+
+## Estimate from high frequency data with error
 
@@
 
-```{r, cache=TRUE}
+```{r estimate te ExpShore for high error samples, cache=TRUE}
 #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
+formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
 
-crcvm_rect_hf_e1 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_hf))
-crcvm_rect_hf_e1$fit(method="BFGS")
+crcvm_rect_hf_e1=list()
+for (k in 1:N_BATCH) {
+  m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
+                      other_data=list("H"=H0_hf))
+  m$fit(method="BFGS")
+  crcvm_rect_hf_e1[[paste("BATCH",k,sep="")]]<-m
+}
 ```
 
-#### Plot estimated smooth in 3D
+#### Plot estimated smooth in 3D for one batch
 
-```{r}
+```{r plots estimates te ExpShore for one batch of high error samples}
 xmin=list("theta"=-pi,"ExpShore"=1/2)
@@
 true_smooths=list("omega"=fomega)
-res=crcvm_rect_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",xmin=xmin,xmax=xmax,
+res=crcvm_rect_hf_e1$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-#### Violin plot of the estimated spline coefficients
+#### Violin plot of the estimated spline coefficients for one batch
 
-```{r}
+```{r violin plots spline coeffs te ExpShore for one batch of high error sample}
 
@@
 
-p=crcvm_rect_hf_e1$plot_violin_coeff(model_name="crcvm_rect_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_rect_hf_e1$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
 ggplotly(p)
+```
+
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te ExpShore all batches with high measurement error}
+
+n_coeffs=length(sp_coeff_ExpShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_rect_hf_e1[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+ggsave("all_batches_coeffs_crcvm_rect_hf_e1.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e1")
+
+ggplotly(p)
+
 ```
@@
 
-```{r,cache=TRUE}
+```{r estimate te DistanceShore for high error samples,cache=TRUE}
 #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-par0 <- c(0,0,1,4,0)
+formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
 
-crcvm_rect_hf_e2 <- SDE$new(formulas = formulas,data = data_rect_hf_e[data_rect_hf_e$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_hf))
-crcvm_rect_hf_e2$fit(method="BFGS")
+crcvm_rect_hf_e2=list()
+
+for (k in 1:N_BATCH) {
+  m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
+                      other_data=list("H"=H0_hf))
+  m$fit(method="BFGS")
+  crcvm_rect_hf_e2[[paste("BATCH",k,sep="")]]<-m
+}
 ```
 
-#### Plot estimated smooth in 3D
+#### Plot estimated smooth in 3D for one batch
 
-```{r}
+```{r plots estimates te DistanceShore for one batch of high error samples}
 xmin=list("theta"=-pi,"DistanceShore"=0.3)
 xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_rect_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",xmin=xmin,xmax=xmax,
+res=crcvm_rect_hf_e2$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-#### Violin plot of the estimated spline coefficients
+#### Violin plot of the estimated spline coefficients for one batch
 
-```{r}
+```{r violin plots spline coeffs te DistanceShore for one batch of high error sample}
 
@@
 
-p=crcvm_rect_hf_e2$plot_violin_coeff(model_name="crcvm_rect_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_rect_hf_e2$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
 
@@
 ```
+
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te DistanceShore all batches with high measurement error}
+
+n_coeffs=length(sp_coeff_DShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_rect_hf_e2[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+ggsave("all_batches_coeffs_crcvm_rect_hf_e2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e2")
+```

### 2024-04-16 R/simulation_study/simulation_study_rect.Rmd
changes in simulation_study_rect

@@
 #number of batch
-N_BATCH=10
+N_BATCH=2
 
@@
 #duration of each track in hour
-TMAX=1*24
+TMAX=0.5*24
 
@@
 
-```{r high error samples, cache=TRUE,results='hide'}
+```{r parallel generate hf high error samples, cache=TRUE,results='hide'}
 #setup parallel backend to use many processors
@@
 ```{r estimate te DistanceShore for low error samples,cache=TRUE}
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_rect_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
@@
                  nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-
-crcvm_rect_hf_ne2=list()
-for (k in 1:N_BATCH) {
-  m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH"),k,sep=""]],type = "RACVM",
+  par0 <- c(0,0,1,1,0)
+  m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
@@
   m$fit(method="BFGS")
-  crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]<-m
+  m 
 }
+
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_rect_hf_ne2)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 ```{r estimate te ExpShore for high error samples, cache=TRUE}
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_rect_hf_e1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
@@
                  nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
-
-crcvm_rect_hf_e1=list()
-for (k in 1:N_BATCH) {
+  par0 <- c(0,0,1,1,0)
   m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
@@
   m$fit(method="BFGS")
-  crcvm_rect_hf_e1[[paste("BATCH",k,sep="")]]<-m
+  m 
 }
+
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_rect_hf_e1)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 ```{r estimate te DistanceShore for high error samples,cache=TRUE}
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_rect_hf_e2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
@@
                  nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-
-crcvm_rect_hf_e2=list()
-
-for (k in 1:N_BATCH) {
+  par0 <- c(0,0,1,1,0)
   m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
@@
   m$fit(method="BFGS")
-  crcvm_rect_hf_e2[[paste("BATCH",k,sep="")]]<-m
+  m 
 }
+
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_rect_hf_e2)=paste("BATCH",1:N_BATCH,sep="")
 ```

### 2024-04-17 R/simulation_study/simulation_study_rect.Rmd
changes in simulation_study_rect
fix error in simulation_study_rect + other trials
fix error in simulation_study_rect

@@
 #number of batch
-N_BATCH=2
+N_BATCH=30
 
 #number of individual tracks per batch
-N_ID=3
+N_ID=6
 
@@
 
-n_coeffs=length(sp_coeff_DShore)
+n_coeffs=length(sp_coeff_Dshore)
 coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)

### 2024-04-19 Latex/main.tex
add new files due to changes in smoothSDE package

@@
 
-By integrating the velocity equation over time one gets the transition density of the location process.
+By integrating the velocity equation over time one gets the following distribution for the location process :
 \begin{equation}
@@
 \]
-The position process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
+The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
 Both components of the location and velocity processes are independent.
@@
  \[
- \omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\frac{\pi}{2}))+\tanh(\lambda(\theta-\frac{\pi}{2})))\right)\times c(D_{shore})
+ \omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
  \]
- for some $\omega_0 \in \R$, $\lambda>0$
+ for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
 
@@
  	\label{fig: trueomega}
+ \end{figure}
+
+ \begin{figure}[H]
+ 	\centering
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_ne_omega.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_e_omega.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\caption{Samples of the CRCVM in a rectangular domain.
+x and y axis are in metres and initial positions are indicated by a red cross (a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+ 	\label{fig: samples_rect_CRCVM}
+ \end{figure}
+
+ \begin{figure}[H]
+ 	\centering
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_ne_omega.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_e_omega.png}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\caption{Samples of the CRCVM in Scoresby Sound coastline geometry.
+x and y axis are in metres and initial positions are indicated by a red cross.
+(a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+ 	\label{fig: samples_fjords_CRCVM}
  \end{figure}
@@
 Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_ne_omega.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_e_omega.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Samples of the CRCVM.
-x and y axis are in metres and initial positions are indicated by a red cross (a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: samples_rect_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_ne_omega.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_e_omega.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Samples of the CRCVM.
-x and y axis are in metres and initial positions are indicated by a red cross.
-(a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: samples_fjords_CRCVM}
-\end{figure}
-
-\begin{comment}
-
-To assess whether the CRCVM can describe a tendecy to move along the boundary, $\tau$ was then also expressed as a smooth function of $D_{shore}$ and $\Theta$.
-\begin{align*}	
-	\tau(\theta,D_{shore})&=(1-c(D_{shore}))\times \tau_0 +c(D_{shore}) \times   \\
-	&+\left(\Delta_0+(\tau_0-\Delta_0)\exp\left(\frac{-\theta^2}{\pi^2}\right)+
-	(\tau_1-\Delta_0)\exp\left(-\lambda_{\tau}\left(\theta-\frac{\pi}{2}\right)^2\right)+(\tau_2-\Delta_0)\exp\left(-\lambda_{\tau}\left(\theta+\frac{\pi}{2}\right)^2\right)\right)
-\end{align*}
-for some $0<\Delta_0<\tau_0<\tau_1,\tau_2$ and $\lambda_{\tau}>0$.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.7\linewidth]{"images/simulation study/true_tau"}
-	\caption{$\tau$ (unit: $h$) as a smoth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$}
-	\label{fig:truetau}
-\end{figure}
-
-\end{comment}
 

### 2024-04-19 R/simulation_study/simulation_study_rect.Rmd
add new files due to changes in smoothSDE package

@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_ne1.png",plot=p,path="crcvm_rect_hf_ne1")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_ne1/all_batches_coeffs_crcvm_rect_hf_ne1.png")
 
-ggplotly(p)
 ```
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_ne2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_ne2")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.png")
 
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_e1.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e1")
-
-ggplotly(p)
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_e1/all_batches_coeffs_crcvm_rect_hf_e1.png")
 
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_e2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e2")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_e2/all_batches_coeffs_crcvm_rect_hf_e2.png")
 ```

### 2024-04-19 R/simulation_study/simulation_study_fjords.Rmd
parallelization in simulation_study_rect + other stuff
changes in simulation study

@@
 
-# Import libraries and data
+# SET UP
+## Import libraries and data
 
@@
 
-# Get land polygons
+## Global variables
+
+```{r}
+#number of batch
+N_BATCH=2
+
+#number of individual tracks per batch
+N_ID=3
+
+#duration of each track in hour
+TMAX=0.5*24
+
+#degree of freedom in tensor splines
+SP_DF=5
+```
+
+# SIMULATION
+
+## Get land polygons
 
@@
 ```
-
-# SIMULATION
 
@@
 ```{r fix time steps}
-#observation times
-Tmax=3*24 #max time in hour
 
@@
 #constant tau
-ftau_constant=function(Dshore,theta,tau=1) {
-  return (tau)
+ftau_constant=function(cov_data,tau=1,sigma_re=0.1) {
+  b=rnorm(1,mean=0,sd=sigma_re)
+  return (tau+b)
 }
@@
 
-nu=4
+#constant nu
+fnu_constant=function(cov_data,nu=4,sigma_re=0.3) {
+    b=rnorm(1,mean=0,sd=sigma_re)
+  return (nu+b)
+}
 
@@
 #number of points
-n <- 500
+n <- 1000
 
@@
 
-## Simulation with measurement errors and high frequency
+### Define spline smooth functions for omega based on the approximation
 
-### Generate samples
+```{r spline smooth omega definition}
 
-```{r high frequency samples no error (hf ne)}
+fomega_splines=function(cov_data) {
 
-#initialize data frame
-data_fjords_hf_ne=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+  if (is.null(cov_data$ExpShore)){
+    omega=predict(m1,newdata=cov_data)
+  }
+  else if (is.null(cov_data$DistanceShore)){
+    omega=predict(m2,newdata=cov_data)
+  }
 
-for (i in 1:n_samples) {
+  return(as.numeric(omega))
+}
 
-  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=land,verbose=FALSE)
+```
+
+## Simulation with low measurement errors and high frequency
+
+### Generate sample
+
+```{r parallel generate hf low errors samples}
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+data_fjords_hf_ne <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+  data=foreach (i=1:N_ID,.combine='rbind') %do% {
+
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=land,verbose=FALSE)
 
@@
     data_shore=res$shore[,c("p1","p2")]
-  data_fjords_hf_ne=rbind(data_fjords_hf_ne,cbind(data_sim,data_shore))
+    cbind(data_sim,data_shore)
+  }
+  data 
 }
 
-data_fjords_hf_ne=data_fjords_hf_ne[-1,]
+#stop cluster
+stopCluster(cl)
 
+names(data_fjords_hf_ne)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 
-```{r percentage of hf ne samples that reached land}
+```{r percentage of low error samples that reached land}
 count=0
-for (id in unique(data_fjords_hf_ne$ID)){
-  sub_data=data_fjords_hf_ne[data_fjords_hf_ne$ID==id,]
+for (k in 1:length(names(data_rect_hf_ne))) {
+
+  data=data_fjords_hf_ne[[paste("BATCH",k,sep="")]]
+  for (id in unique(data$ID)) {
+    sub_data=data[data$ID==id,]
     if (nrow(sub_data) < n_hf) {
-    print(id)
       count=count+1
+      cat(paste("BATCH",k,sep=""),":",id,"\n",sep=" ")
     }
   }
-cat(count/n_samples*100,"percent of the samples reached land")
+}
+cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
 ```
 
-### Plot samples
+### plot one batch
 
-```{r plot hf ne samples}
+```{r plot one batch low error samples}
 
-plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(y1,y2),size=0.1)+
-  geom_point(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(p1,p2),size=0.01,col="red",shape=5)+
-  geom_path(data=data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],mapping=aes(y1,y2),linewidth=0.1)+
-  geom_point(data = data_fjords_hf_ne[data_fjords_hf_ne$ID==2,] %>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4,col = "blue")+
+plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_fjords_hf_ne$BATCH1%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
   xlab("x") + ylab("y")
@@
 
-## sImulate with high frequency and measurement error
+## Simulation with high measurement error and high frequency
 
@@
 
-```{r generate high frequency samples with error (hf e)}
+```{r parallel generate hf high error samples, cache=TRUE,results='hide'}
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
 
-#initialize data frame
-data_fjords_hf_e=data.frame("y1"=NA,"y2"=NA,"time"=NA,"p1"=NA,"p2"=NA,"ID"=NA)
+data_fjords_hf_e <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+  data=foreach (i=1:N_ID,.combine='rbind') %do% {
 
-for (i in 1:n_samples) {
-
-  res=sim_theta_CRCVM(ftau_constant,fomega,nu,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=land,verbose=FALSE)
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=land,verbose=FALSE)
 
@@
     data_shore=res$shore[,c("p1","p2")]
-  data_fjords_hf_e=rbind(data_fjords_hf_e,cbind(data_sim,data_shore))
+    cbind(data_sim,data_shore)
+  }
+  data 
 }
 
-data_fjords_hf_e=data_fjords_hf_e[-1,]
+#stop cluster
+stopCluster(cl)
 
+#change names of the list elements
+names(data_fjords_hf_e)=paste("BATCH",1:N_BATCH,sep="")
 ```
 
-### Percentage of samples that reached land
-```{r percentage of hf e samples that reached land}
+### percentage of samples that reached land
+
+```{r percentage of high error samples that reached land}
 count=0
-for (id in unique(data_fjords_hf_e$ID)){
-  sub_data=data_fjords_hf_e[data_fjords_hf_e$ID==id,]
+for (k in 1:length(names(data_fjords_hf_e))) {
+
+  data=data_fjords_hf_e[[paste("BATCH",k,sep="")]]
+  for (id in unique(data$ID)) {
+    sub_data=data[data$ID==id,]
     if (nrow(sub_data) < n_hf) {
-    print(id)
       count=count+1
+      cat("BATCH",k,id,"\n",sep="")
     }
   }
-cat(count/n_samples*100,"percent of the samples reached land")
+}
+cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
 ```
 
-### Plot one sample
+### plot one sample
 
-```{r plot hf e samples}
+```{r plot one batch of high error samples}
 
-plot_samples_fjords_hf_e=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_e[data_fjords_hf_e$ID==2,],mapping=aes(y1,y2),size=0.1)+
-   geom_path(data=data_fjords_hf_e[data_fjords_hf_e$ID==2,],mapping=aes(y1,y2),linewidth=0.1)+
-  geom_point(data = data_fjords_hf_e[data_fjords_hf_e$ID==2,] %>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2 ), shape = 3, size = 4,col = "red")+
+plot_samples_fjords_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_fjords_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_fjords_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_fjords_hf_e$BATCH1%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
   xlab("x") + ylab("y")
@@
 
-```{r compute covariates}
+```{r compute covs}
 
@@
   #data has at least columns "time","y1" "y2", "p1","p2"
-  #this function returns a dataframe with three mùore columns : "theta", "DistanceShore" and "ExpShore"
+  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"
 
@@
 
-data_fjords_hf_ne=add_covs(data_fjords_hf_ne)
+for (k in 1:N_BATCH) {
 
-data_fjords_hf_e=add_covs(data_fjords_hf_e)
+  #dataframes for this batch
+  df_ne=data_fjords_hf_ne[[paste("BATCH",k,sep="")]]
+  df_e=data_fjords_hf_e[[paste("BATCH",k,sep="")]]
+
+  #add covariates columns
+  df_ne=add_covs(df_ne)
+  df_e=add_covs(df_e)
+
+  #change in the list of batch
+  data_fjords_hf_ne[[paste("BATCH",k,sep="")]]=df_ne
+  data_fjords_hf_e[[paste("BATCH",k,sep="")]]=df_e
+}
+
 ```
+
+# ESTIMATION
 
@@
 
-```{r estimate from hf e samples with cov ExpShore}
+```{r estimate te ExpShore for low error samples,cache=TRUE}
 
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_fjords_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
+  formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
+  par0 <- c(0,0,1,1,0)
+  m <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
+                      other_data=list("H"=H0_hf))
+  m$fit(method="BFGS")
+  m 
+}
 
-crcvm_fjords_hf_ne1 <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[data_fjords_hf_ne$ID==2,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H0_hf))
-crcvm_fjords_hf_ne1$fit(method="BFGS")
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_fjords_hf_ne1)=paste("BATCH",1:N_BATCH,sep="")
+
 ```
 
-#### Plot estimated smooth parameter omega in 3D
+#### Plot estimated smooth parameter omega in 3D for one batch
 
@@
 
-#### Violin plot of the estimated spline coefficients
+#### Violin plot of the estimated spline coefficients for one batch
 
@@
 
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te ExpShore all batches with low measurement error}
+
+n_coeffs=length(sp_coeff_ExpShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_fjords_hf_ne1[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_fjords_hf_ne1/all_batches_coeffs_crcvm_fjords_hf_ne1.png")
+
+```
+
 #### Simulate from fitted model
 
-```{r,cache=TRUE}
+```{r simulate fitted te ExpShore low error samples,cache=TRUE}
 
@@
 atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
-sim_data=crcvm_fjords_hf_ne1$simulate(z0=x0[1,],data=crcvm_fjords_hf_ne1$data(),atw=atw,land=land)
+sim_data=crcvm_rect_hf_ne1$BATCH1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne1$BATCH1$data(),atw=atw,land=domain)
 

### 2024-04-19 R/simulation_study/simulation_study_rect.Rmd
changes in simulation study

--- 2024-04-16 R/simulation_study/simulation_study_rect.Rmd
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_ne1.png",plot=p,path="crcvm_rect_hf_ne1")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_ne1/all_batches_coeffs_crcvm_rect_hf_ne1.html")
 
-ggplotly(p)
 ```
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_ne2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_ne2")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.html")
 
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_e1.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e1")
-
-ggplotly(p)
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_e1/all_batches_coeffs_crcvm_rect_hf_e1.html")
 
@@
 
-ggsave("all_batches_coeffs_crcvm_rect_hf_e2.png",plot=p,width=10,height=5,path="crcvm_rect_hf_e2")
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_e2/all_batches_coeffs_crcvm_rect_hf_e2.html")
 ```

### 2024-04-20 R/simulation_study/simulation_study_fjords.Rmd
changes in simulation study

@@
 #number of batch
-N_BATCH=2
+N_BATCH=1
 
 #number of individual tracks per batch
-N_ID=3
+N_ID=6
 
@@
 
-#number of tracks
-n_samples=5
-
 #initial velocity and location
@@
 #generate random initial points
-x0=matrix(rep(NA,n_samples*2),ncol=2)
+x0=matrix(rep(NA,N_ID*2),ncol=2)
 colnames(x0)=c("x1","x2")
@@
 i=1
-while (i<=n_samples) {
+while (i<=N_ID) {
   #choose location uniformly in the map
@@
 #high frequency time steps
-times_hf=seq(0,Tmax,by=hf)
+times_hf=seq(0,TMAX,by=hf)
 n_hf=length(times_hf)-1
@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs_0=0.003
-H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf),dim=c(2,2,n_hf))
-#GPS measurement error 
-sigma_obs_1=0.03
-H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf),dim=c(2,2,n_hf))
+sigma_obs=0.005
+H_hf=array(rep(sigma_obs^2*diag(2),n_hf),dim=c(2,2,n_hf))
 
@@
 #mixture of independent gaussian
-ftau_gauss=function(cov_data,D0=0.3,tau0=1,tau1=3) {
+ftau_gauss=function(cov_data,D0=0.5,tau0=1,tau1=3) {
   Dshore=cov_data$DistanceShore
   theta=cov_data$theta
-  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/10))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/10))^2-(Dshore/D0)^2)
+  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/5))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/5))^2-(Dshore/D0)^2)
   return (tau)
@@
   return (tau+b)
+}
+
+ftau_step=function(cov_data,epsilon=pi/10,alpha=pi/5) {
+    Dshore=cov_data$DistanceShore
+    theta=cov_data$theta
+    tau=1.5+0.5*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
+      0.5*(tanh((theta+(-pi/2+epsilon))/alpha)-tanh((theta+(-pi/2-epsilon))/alpha))
+    return (tau)
 }
@@
 true_par=grid
-true_par$tau=ftau_constant(grid)
+true_par$tau=ftau_step(grid)
 true_par$omega=fomega(grid)
@@
 ```{r}
-sp_df=5
 #number of points
-n <- 1000
+n <- 500
 
@@
 
+### Plot
+
+```{r plot spline smooth omega}
+
+Dshore=seq(from=0.3,to=2,length.out=30)
+theta=seq(from=-pi,to=pi,length.out=30)
+grid<- as.data.frame(expand.grid(DistanceShore,theta))
+colnames(grid)=c("DistanceShore","theta")
+true_par=grid
+true_par$omega=fomega_splines(grid)
+
+plot_true_omega <- plot_ly(true_par,type = "mesh3d",
+                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
+                         colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_true_omega
+```
 ## Simulation with low measurement errors and high frequency
@@
 
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=land,verbose=FALSE)
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs),v0,x0[i,],times_hf,land=land,verbose=FALSE)
 
@@
 
-plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
   geom_point(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
@@
 ggplotly(plot_samples_fjords_hf_ne)
-```
-
-## Simulation with high measurement error and high frequency
-
-### Generate samples
-
-```{r parallel generate hf high error samples, cache=TRUE,results='hide'}
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-data_fjords_hf_e <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
-  data=foreach (i=1:N_ID,.combine='rbind') %do% {
-
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=land,verbose=FALSE)
-
-    data_sim=res$sim
-    data_sim$ID=factor(rep(i,length(data_sim$y1)))
-    data_shore=res$shore[,c("p1","p2")]
-    cbind(data_sim,data_shore)
-  }
-  data 
-}
-
-#stop cluster
-stopCluster(cl)
-
-#change names of the list elements
-names(data_fjords_hf_e)=paste("BATCH",1:N_BATCH,sep="")
-```
-
-### percentage of samples that reached land
-
-```{r percentage of high error samples that reached land}
-count=0
-for (k in 1:length(names(data_fjords_hf_e))) {
-
-  data=data_fjords_hf_e[[paste("BATCH",k,sep="")]]
-  for (id in unique(data$ID)) {
-    sub_data=data[data$ID==id,]
-    if (nrow(sub_data) < n_hf) {
-      count=count+1
-      cat("BATCH",k,id,"\n",sep="")
-    }
-  }
-}
-cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
-```
-
-### plot one sample
-
-```{r plot one batch of high error samples}
-
-plot_samples_fjords_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_path(data=data_fjords_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_point(data = data_fjords_hf_e$BATCH1%>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
-  xlab("x") + ylab("y")
-
-ggsave(filename="samples_fjords_hf_e_omega.png",plot=plot_samples_fjords_hf_e)
-ggplotly(plot_samples_fjords_hf_e)
 ```
@@
   df_ne=data_fjords_hf_ne[[paste("BATCH",k,sep="")]]
-  df_e=data_fjords_hf_e[[paste("BATCH",k,sep="")]]
 
@@
   df_ne=add_covs(df_ne)
-  df_e=add_covs(df_e)
 
@@
   data_fjords_hf_ne[[paste("BATCH",k,sep="")]]=df_ne
-  data_fjords_hf_e[[paste("BATCH",k,sep="")]]=df_e
 }
@@
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
-                      other_data=list("H"=H0_hf))
+                      other_data=list("H"=H_hf))
   m$fit(method="BFGS")
@@
 true_smooths=list("omega"=fomega)
-res=crcvm_fjords_hf_ne1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne1",xmin=xmin,xmax=xmax,
+res=crcvm_fjords_hf_ne1$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne1",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-p=crcvm_fjords_hf_ne1$plot_violin_coeff(model_name="crcvm_fjords_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_fjords_hf_ne1$BATCH1$plot_violin_coeff(model_name="crcvm_fjords_hf_ne1",substring="omega",true_coeff=true_coeff,save=TRUE)
 
@@
 
-```{r,cache=TRUE}
+```{r estimate te DistanceShore for low error samples,cache=TRUE}
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+crcvm_fjords_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
-par0 <- c(0,0,1,4,0)
+  formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
+  par0 <- c(0,0,1,1,0)
+  m <- SDE$new(formulas = formulas,data = data_fjords_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
+                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
+                      other_data=list("H"=H_hf))
+  m$fit(method="BFGS")
+  m 
+}
 
-crcvm_fjords_hf_ne2 <- SDE$new(formulas = formulas,data = data_rect_hf_ne[data_rect_hf_ne$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H0_hf))
-crcvm_fjords_hf_ne2$fit(method="BFGS")
+#stop cluster
+stopCluster(cl)
+
+names(crcvm_fjords_hf_ne2)=paste("BATCH",1:N_BATCH,sep="")
 ```
 
-#### Plots estimated smooth in 3D
+#### Plots estimated smooth in 3D for one batch
 
@@
 xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_fjords_hf_ne2$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne2",xmin=xmin,xmax=xmax,
+res=crcvm_fjords_hf_ne2$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_ne2",xmin=xmin,xmax=xmax,
                                     xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
@@
 
-p=crcvm_fjords_hf_ne2$plot_violin_coeff(model_name="crcvm_fjords_hf_ne2",substring="omega",true_coeff=true_coeff,save=TRUE)
+p=crcvm_fjords_hf_ne2$BATCH1$plot_violin_coeff(model_name="crcvm_fjords_hf_ne2",substring="omega",true_coeff=true_coeff,save=TRUE)
 
@@
 
-## Estimate from high frequency data with error :
+#### Simulate from fitted model
 
-### Use ExpShore and tensor splines te
+```{r simulate fitted te DistanceShore low error samples,cache=TRUE}
 
-#### Fit sde
+fExpShore=function(z,v,p) {
 
-```{r, cache=TRUE}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,ExpShore,k=5,bs="cs"))
-par0 <- c(0,0,1,4,0)
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
 
-crcvm_fjords_hf_e1 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_hf))
-crcvm_fjords_hf_e1$fit(method="BFGS")
+  #distance to shore
+  Dshore=sqrt(normal[1]^2+normal[2]^2)
+  if (Dshore<0.01) {
+    return (100)
+  }
+  else {
+    return (1/Dshore)
+  }
+}
+
+fangle=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #return angle
+  return (signed_angle(normal,v))
+}
+
+atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
+sim_data=crcvm_fjords_hf_ne2$BATCH1$simulate(z0=x0[1,],data=crcvm_fjords_hf_ne2$BATCH1$data(),atw=atw,land=land)
+
 ```
 
-#### Plot estimated smooth in 3D
+#### Plot simulation from fitted model
 
-```{r}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.3)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-true_smooths=list("omega"=fomega)
-res=crcvm_fjords_hf_e1$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e1",xmin=xmin,xmax=xmax,
-                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
-res$fe_omega_theta_ExpShore
+```{r plot sim te DistanceShore low error samples}
+
+plot_sim=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_point(data = sim_data%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2 ), shape = 3, size = 4, col = "red")+
+    xlab("x") + ylab("y")
+
+ggsave("sim_crcvm_fjords_hf_ne2.png",plot=plot_sim,width=10,height=5,path="crcvm_fjords_hf_ne2")
+ggplotly(plot_sim)
 ```
 
-#### Violin plot of the estimated spline coefficients
+#### Violin plot of the estimated spline coefficients for all batches
 
-```{r}
+```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
 
-true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
-rownames(true_coeff)=NULL
+n_coeffs=length(sp_coeff_Dshore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
 
-p=crcvm_fjords_hf_e1$plot_violin_coeff(model_name="crcvm_fjords_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
+for (k in 1:N_BATCH) {
+  m<-crcvm_fjords_hf_ne2[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
 
-ggplotly(p)
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_fjords_hf_ne2/all_batches_coeffs_crcvm_fjords_hf_ne2.html")
+
 ```
-
-### Use DistanceShore and tensor splines te
-
-#### Fit sde
-
-```{r,cache=TRUE}
-#estimation with tensor splines
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=sp_df,bs="cs"))
-par0 <- c(0,0,1,4,0)
-
-crcvm_fjords_hf_e2 <- SDE$new(formulas = formulas,data = data_fjords_hf_e[data_fjords_hf_e$ID==1,],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2","nu","tau"),
-                      other_data=list("H"=H1_hf))
-crcvm_fjords_hf_e2$fit(method="BFGS")
-```
-
-#### Plot estimated smooth in 3D
-
-```{r}
-xmin=list("theta"=-pi,"DistanceShore"=0.3)
-xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_fjords_hf_e2$get_all_plots(baseline=NULL,model_name="crcvm_fjords_hf_e2",xmin=xmin,xmax=xmax,
-                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
-res$fe_omega_theta_DistanceShore
-```
-
-#### Violin plot of the estimated spline coefficients
-
-```{r}
-
-true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
-rownames(true_coeff)=NULL
-
-p=crcvm_fjords_hf_e2$plot_violin_coeff(model_name="crcvm_fjords_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
-
-ggplotly(p)
-```

### 2024-04-20 R/simulation_study/simulation_study_rect.Rmd
changes in simulation study

@@
 #number of batch
-N_BATCH=2
+N_BATCH=50
 
 #number of individual tracks per batch
-N_ID=3
+N_ID=6
 
@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs_0=0.005
-H0_hf=array(rep(sigma_obs_0^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
-#GPS measurement error 
-sigma_obs_1=0.03
-H1_hf=array(rep(sigma_obs_1^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
+sigma_obs=0.005
+H_hf=array(rep(sigma_obs^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
 
@@
 
-```{r parallel generate hf low errors samples}
+```{r parallel generate hf low errors samples,cache=TRUE}
 
@@
 
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_0),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
+    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
 
@@
 ggplotly(plot_samples_rect_hf_ne)
-```
-
-## Simulation with high measurement error and high frequency
-
-### Generate samples
-
-```{r parallel generate hf high error samples, cache=TRUE,results='hide'}
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-data_rect_hf_e <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
-  data=foreach (i=1:N_ID,.combine='rbind') %do% {
-
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs_1),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
-
-    data_sim=res$sim
-    data_sim$ID=factor(rep(i,length(data_sim$y1)))
-    data_shore=res$shore[,c("p1","p2")]
-    cbind(data_sim,data_shore)
-  }
-  data 
-}
-
-#stop cluster
-stopCluster(cl)
-
-#change names of the list elements
-names(data_rect_hf_e)=paste("BATCH",1:N_BATCH,sep="")
-```
-
-### percentage of samples that reached land
-
-```{r percentage of high error samples that reached land}
-count=0
-for (k in 1:length(names(data_rect_hf_e))) {
-
-  data=data_rect_hf_e[[paste("BATCH",k,sep="")]]
-  for (id in unique(data$ID)) {
-    sub_data=data[data$ID==id,]
-    if (nrow(sub_data) < n_hf) {
-      count=count+1
-      cat("BATCH",k,id,"\n",sep="")
-    }
-  }
-}
-cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
-```
-
-### plot one sample
-
-```{r plot one batch of high error samples}
-
-plot_samples_rect_hf_e=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_path(data=data_rect_hf_e$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_point(data = data_rect_hf_e$BATCH1%>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
-  xlab("x") + ylab("y")
-
-ggsave(filename="samples_rect_hf_e_omega.png",plot=plot_samples_rect_hf_e)
-ggplotly(plot_samples_rect_hf_e)
 ```
@@
   df_ne=data_rect_hf_ne[[paste("BATCH",k,sep="")]]
-  df_e=data_rect_hf_e[[paste("BATCH",k,sep="")]]
 
@@
   data_rect_hf_ne[[paste("BATCH",k,sep="")]]=df_ne
-  data_rect_hf_e[[paste("BATCH",k,sep="")]]=df_e
 }
@@
 
+#### Simulate from fitted model
+
+```{r simulate fitted te DistanceShore low error samples,cache=TRUE}
+
+fExpShore=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #distance to shore
+  Dshore=sqrt(normal[1]^2+normal[2]^2)
+  if (Dshore<0.01) {
+    return (100)
+  }
+  else {
+    return (1/Dshore)
+  }
+}
+
+fangle=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #return angle
+  return (signed_angle(normal,v))
+}
+
+atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
+sim_data=crcvm_rect_hf_ne2$BATCH1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne2$BATCH1$data(),atw=atw,land=domain)
+
+```
+
+#### Plot simulation from fitted model
+
+```{r plot sim te DistanceShore low error samples}
+
+plot_sim=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=sim_data,mapping=aes(y1,y2),size=0.1)+
+  geom_point(data = sim_data%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2 ), shape = 3, size = 4, col = "red")+
+    xlab("x") + ylab("y")
+
+ggsave("sim_crcvm_rect_hf_ne2.png",plot=plot_sim,width=10,height=5,path="crcvm_rect_hf_ne2")
+ggplotly(plot_sim)
+```
+
 #### Violin plot of the estimated spline coefficients for all batches
@@
 ```
-
-## Estimate from high frequency data with error
-
-### Use ExpShore and tensor splines te
-
-#### Fit sde
-
-```{r estimate te ExpShore for high error samples, cache=TRUE}
-
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_rect_hf_e1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
-  #estimation with tensor splines
-  formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
-                 nu=~s(ID,bs="re"),omega=~te(theta,ExpShore,k=SP_DF,bs="cs"))
-  par0 <- c(0,0,1,1,0)
-  m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
-                      other_data=list("H"=H0_hf))
-  m$fit(method="BFGS")
-  m 
-}
-
-#stop cluster
-stopCluster(cl)
-
-names(crcvm_rect_hf_e1)=paste("BATCH",1:N_BATCH,sep="")
-```
-
-#### Plot estimated smooth in 3D for one batch
-
-```{r plots estimates te ExpShore for one batch of high error samples}
-xmin=list("theta"=-pi,"ExpShore"=1/2)
-xmax=list("theta"=pi,"ExpShore"=1/0.3)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
-true_smooths=list("omega"=fomega)
-res=crcvm_rect_hf_e1$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e1",xmin=xmin,xmax=xmax,
-                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
-res$fe_omega_theta_ExpShore
-```
-
-#### Violin plot of the estimated spline coefficients for one batch
-
-```{r violin plots spline coeffs te ExpShore for one batch of high error sample}
-
-true_coeff=data.frame(value=sp_coeff_ExpShore,name=paste("omega.",names(sp_coeff_ExpShore),sep=""))
-rownames(true_coeff)=NULL
-
-p=crcvm_rect_hf_e1$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_e1",substring="omega",true_coeff=true_coeff,save=TRUE)
-
-ggplotly(p)
-```
-
-#### Violin plot of the estimated spline coefficients for all batches
-
-```{r violin plots spline coeff te ExpShore all batches with high measurement error}
-
-n_coeffs=length(sp_coeff_ExpShore)
-coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
-
-for (k in 1:N_BATCH) {
-  m<-crcvm_rect_hf_e1[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
-  coeff_names=rownames(coeffs)
-  coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
-  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
-}
-
-#remove initialization row
-coeffs_df=coeffs_df[-1,]
-
-p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
-                xlab(" ")+labs(fill = "Estimated coefficients")+
-                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
-                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
-
-p=ggplotly(p)
-saveWidget(p, file ="crcvm_rect_hf_e1/all_batches_coeffs_crcvm_rect_hf_e1.html")
-
-```
-
-### Use DistanceShore and tensor splines te
-
-#### Fit sde
-
-```{r estimate te DistanceShore for high error samples,cache=TRUE}
-
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_rect_hf_e2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
-  #estimation with tensor splines
-  formulas <- list(mu1=~1,mu2=~1,tau=~s(ID,bs="re"),
-                 nu=~s(ID,bs="re"),omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-  par0 <- c(0,0,1,1,0)
-  m <- SDE$new(formulas = formulas,data = data_rect_hf_e[[paste("BATCH",k,sep="")]],type = "RACVM",
-                      response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
-                      other_data=list("H"=H0_hf))
-  m$fit(method="BFGS")
-  m 
-}
-
-#stop cluster
-stopCluster(cl)
-
-names(crcvm_rect_hf_e2)=paste("BATCH",1:N_BATCH,sep="")
-```
-
-#### Plot estimated smooth in 3D for one batch
-
-```{r plots estimates te DistanceShore for one batch of high error samples}
-xmin=list("theta"=-pi,"DistanceShore"=0.3)
-xmax=list("theta"=pi,"DistanceShore"=2)
-res=crcvm_rect_hf_e2$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_e2",xmin=xmin,xmax=xmax,
-                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
-res$fe_omega_theta_DistanceShore
-```
-
-#### Violin plot of the estimated spline coefficients for one batch
-
-```{r violin plots spline coeffs te DistanceShore for one batch of high error sample}
-
-true_coeff=data.frame(value=sp_coeff_Dshore,name=paste("omega.",names(sp_coeff_Dshore),sep=""))
-rownames(true_coeff)=NULL
-
-p=crcvm_rect_hf_e2$BATCH1$plot_violin_coeff(model_name="crcvm_rect_hf_e2",substring="omega",true_coeff=true_coeff,save=TRUE)
-
-ggplotly(p)
-```
-
-#### Violin plot of the estimated spline coefficients for all batches
-
-```{r violin plots spline coeff te DistanceShore all batches with high measurement error}
-
-n_coeffs=length(sp_coeff_Dshore)
-coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
-
-for (k in 1:N_BATCH) {
-  m<-crcvm_rect_hf_e2[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
-  coeff_names=rownames(coeffs)
-  coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
-  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
-}
-
-#remove initialization row
-coeffs_df=coeffs_df[-1,]
-
-p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
-                xlab(" ")+labs(fill = "Estimated coefficients")+
-                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
-                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
-
-p=ggplotly(p)
-saveWidget(p, file ="crcvm_rect_hf_e2/all_batches_coeffs_crcvm_rect_hf_e2.html")
-```

### 2024-04-23 R/simulation_study/simulation_study_rect.Rmd
push results of the simulation study launched on the cluster dahu
changes in simulation study global variables
fix bug in simulation_study_rect.Rmd

@@
 #number of batch
-N_BATCH=50
+N_BATCH=10
 
@@
 
+p
+
 ```
@@
 
+p
 ```

### 2024-04-24 Latex/main.tex
changes in tex files

@@
 \title{
- \textbf{Analysis of Greenland narwhals's behavioural responses to ship noise and airguns exposure using varying coefficients correlated velocity models}}
+ \textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}}
 
@@
 
-% Underwater noise in Arctic : lower than in other environments, anthropogenic nois : vessel traffic, ice breakers, drilling, sonars, seismic airguns for oil and gas exploration. (Halliday 2020)
-% vessel traffic likely to increase due to the melting of the ice cap
-% Largest impact on underwater noise due to seismic airguns (Halliday 2020)
-%Marine mammals sound exposure effects : behavioral and physical impacts (stranding events), more difficult to assess behavioral effect (southall 2007, 2021)
-%Sound threshold, regulation, research recommendations : behavioral response
-%Behavioral response study, control exposure experiments (examples: pile driving, seismic airguns, sonar)
-
-% Studies about a lot of species : narwhals, toothed whale,  mid frequency mammals
-% Greenland narwhals studies
-% effects on dive, vocalization and horizontal movement
-% data, diffusion models 
-% spatial constraints
-%context and literature review, objective of the paper, new contributions
- %% Southeast Greenland narwhals population in decline
-
 Anthropogenic underwater noise in the Arctic environment causes disturbances in the endemic wildlife.
@@
 Among those sources, an increase in vessel traffic is anticipated due to the melting of the ice cap, and seismic airguns are reported to have the largest impact on underwater noise.
-Two types of consequences can be considered : injury and behavior disturbance.
+Two types of consequences can be considered: injury and behavior disturbance.
 While physical harm is straightforward to assess and mitigate by determining species-specific received sound level thresholds, no canonical method is expected to be found to examine behavior disturbances, due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions.\\
 Research recommendations include a particular focus on measurement of behavioral reactions to various type of sound exposure \cite{southall_marine_2008,southall_marine_2019}.
-Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured: GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
-Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises \cite{tyack_beaked_2011,cioffi_trade-offs_2022}, belugas \cite{martin_exposure_2023}, sperm whales \cite{madsen_quantitative_2006}, blue whales \cite{friedlaender_preymediated_2016}, narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021} and many other marine mammals.
+Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured such as GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
+Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises \cite{tyack_beaked_2011,cioffi_trade-offs_2022}, belugas \cite{martin_exposure_2023}, sperm whales \cite{madsen_quantitative_2006}, blue whales \cite{friedlaender_preymediated_2016}, and narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}.
 \\
@@
 These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}.
-Resulting GPS positions of the narwhals have been analyzed with discrete time methods, typically relying on Markov chains.
+Resulting GPS positions of the narwhals have been analyzed with discrete time methods.
 Since GPS positions are retrieved irregularly in time, each narwhal track was linearly interpolated to obtain a position every second, and the positions were then classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second.
+The dynamic of the states was modeled with a Markov chain, assuming that the current state only depends on the last one in time.
 An exposure covariate defined as the inverse of the distance to the sound source was used to model the transition probabilities of the Markov Chain, and it was found that the probabilities of being in states "Close to the shore" and "Moving toward the shore" increased with exposure to the sound, while probability of being in state "Far from the shore" decreased.
@@
 
-A stochastic differential equation with time-varying coefficient is employed to describe the GPS tracks of the individuals \cite{michelot_varying-coefficient_2021}.
+We focus on continuous time models defined as a solution of a stochastic differential equation with time-varying coefficients \cite{michelot_varying-coefficient_2021}.
 One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}.
-It has been used extensively to analyze animal movement in two or even three dimensions \cite{johnson_continuoustime_2008,gurarie_correlated_2017,alt_correlation_1990,gurarie_estimating_2011,albertsen_generalizing_2018}, and can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
-The following formulation of the equations is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
+It has been used extensively to analyze animal movement in two or even three dimensions \cite{gurarie_correlated_2017,alt_correlation_1990,gurarie_estimating_2011,albertsen_generalizing_2018}.
+This model can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
+The following formulation of the equations represents an extension of the CVM \cite{johnson_continuoustime_2008}, which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
 \begin{equation} \left\{
@@
 \end{equation}
-$V(t)$ is the horizontal 2D velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a $2$ dimensional brownian motion. 
-The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives how much random variability there is in the velocity vector, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
+where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a two-dimensional brownian motion. 
+The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
 Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
-Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.\\
+Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.
+We follow \cite{gurarie_correlated_2017} and refer to this model as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" in the case $\mu \neq (0,0)$.\\
 
 In general, the motion of marine mammals is subject to spatial constraints induced by the coastline.
-This is particularly the case for the narwhal data we want to analyse : the animals move in a complex system of fjords in Scoresby Sound (East Greenland), imposing strong constraints on the location process.
-However, this phenomenon is not taken into account in model~\ref{eq: RACVM equation}, nor in most of the SDE used in the context of animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
-Recent research has been conducted to include constraints in these type of models.
-A reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
+This is particularly the case for the narwhal data we want to analyse: the animals move in a complex system of fjords in Scoresby Sound (South-East Greenland), imposing strong constraints on the location process.
+Thus, constraints for the shore are needed.
+However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
+Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
 Simulation of constrained paths was achieved by projecting on the coastline the  points that eventually reached land.
 Estimation of the SDE parameters relied on computationally intensive MCMC algorithm (Metroplis-Hastings), and required high frequency data.
-This requirement was met by linearly interpolating between the discrete time observations of the process.
-Another idea is to add a repulsive potential function in the drift.
-This has been done to constrain ants movement within a box \cite{russell_spatially_2018}, but requires very simple boundaries to get an explicit expression of the potential function.\\
+Another idea is to add a repulsive potential function in the drift, which has been done to constrain ants movement within a box \cite{russell_spatially_2018}.
+In our case study, in which the narwhals move in a complex fjords system, this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function.
+Thus, there is a need to develop new models adapted to such spatial constraints.\\
 Moreover, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}.
 For narwhals, it was observed from the data that they have a proclivity to move along the shoreline.
-Here we propose a new model based on the following simple idea : as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.
-The novelty of this paper consists in estimating the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector, therefore making the drift in the velocity equation dependent on the location process $x$ and the boundary of the domain.
+Here we propose a new model based on the following idea: as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline.
+The novelty of this paper consists in modeling the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector.
+This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.
 Note that in this framework, the special case $\omega=0$ would be the unconstrained correlated velocity model of \cite{johnson_continuoustime_2008}.
 \\ %More biblio about estimation methods
-The angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE}.
+We show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
 To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
@@
 
-Sound exposure will be considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}.
-This effect will be estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
-This choice of covariate is motivated by previous work in \cite{heide-jorgensen_behavioral_2021}.
+Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}.
+This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
+This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}.
 A response model will be compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
@@
 \begin{itemize}
-	\item Derivation of a state-space model for the so called "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}) \cite{gurarie_correlated_2017}, and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
-	\item Definition of a Constrained Rotational Correlated Velocity Model (CRCVM) for simulation study.
-Deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
-	\item Application to real narwhal data to asess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
+	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
+	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
+	\item Application to real narwhal data to assess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
 \end{itemize}
@@
 Section 2 gives an overview of the narwhal data available for the analysis.
-Then in section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models.
+Then in Section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models.
 Section 4 is dedicated to a comprehensive simulation study, whereas in the last section, the results of the application to the narwhal data are examined.
@@
 %trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
-As already mentioned, the dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}.
+
+\subsection{Description of the movement data}
+
+The dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
 Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
@@
 It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots.
-The guns were fired synchronously every 80 s,
-and the GPS navigation system recorded the
-location of every shot. 
+The guns were fired synchronously every 80 s, and the GPS navigation system recorded thelocation of every shot.
 Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres.
@@
 Figure \ref{fig:alltimestepshisto} shows the distribution of time steps in the data.
+While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.
 
@@
 
-The land geometry for this specific region in South-East Greenland was gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
-Supposing there are $n$ observations, for $j \in \{1,\cdots,n\}$, the GPS position obtained at time $t_j$ is denoted $y(t_j)$, and the nearest point on the shoreline is denoted $p(t_j)$.
-The distance to the shore, denoted $D_{shore}(t_j)$ in the sequel, is determined as the distance to this point.
+\subsection{Definition of the covariates}
+
+For each narwhal $i \in \{1,\cdots,N\}$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
+We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and denote $y_{ij}$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing, for $j \in \{1,\cdots,n_i\}$.
+
+The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
+The land polygons define the boundaries of the domain in which the narwhals can move.
+The closest point on the shoreline to the observed position $y_{ij}$  is denoted $p_{ij}$, and the distance to this point, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$.
 The resulting values range from $0$ to $7.5$ km.
 About $9\%$ of the GPS measurements in the dataset turned out to be on land.
-This was primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-For constrained movement models, these points were ignored.
-Then, the exposure to shore covariate was defined with the following formula
+This is primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
+For constrained movement models, these points are ignored.
+Then, exposure to the shore is defined with the following formula
 \begin{equation}
@@
 \begin{array}{ll}
-	 E_{shore}(t_j)=0 & \mbox{if } D_{shore}(t_j)>D_0 \\
-	E_{shore}(t_j)=\frac{1}{D_{shore}(t_j)} & \mbox{if } D_{shore}(t_j) \leq D_0
+	 E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{ref} \\
+	E^{shore}_{ij}=\frac{1}{D_{shore}(t_j)} & \mbox{if } D^{shore}_{ij} \leq D_{ref}
 \end{array}
@@
 \end{equation}
-where the threshold $D_0$ was chosen equal to $0.5$ km.
-The angle between the narwhal's heading and the shoreline is estimated as the angle $\Theta(t_j)$ between the observed empirical velocity $\hat{v}(t_j)=\frac{y(t_j+\Delta_j)-y(t_j)}{\Delta_j}$, where $\Delta_j=t_{j+1}-t_j$, and the vector $n(t_j)=y(t_j)-p(t_j)$.
+where the reference threshold $D_{ref}$ is chosen equal to $0.5$ km.
+We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
+The angle between the narwhal's heading and the shoreline, denoted $\Theta_{ij}$, is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-p_{ij}$.
 This is illustrated in figure~\ref{fig: illustration nearest shore points}.
-A value $\Theta(t_j)=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta(t_j) \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.
-The distance from the ship, denoted $D_{ship}(t_j)$, was computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
+A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.
+The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
 Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}.
-The covariates $E_{shore}$ and $\Theta$ will only be used for the smooth angular velocity parameter $\omega$ in the constrained SDE models.
 
@@
 		\node[left] at (0,0) {$\partial \mathcal{D}$};
-		\node[left] at (0,2) {$p(t_j)$};
-		\node[left] at (1.2,2.8) {$\hat{v}(t_j)$};
-		\node[below] at (1.8,2) {$y(t_j)$};
-		\node[below,red] at (3,2) {$\vec{n}(t_j)$};
-
-		\node[above,orange] at (2.3,2.3) {$\Theta(t_j)$};
+		\node[left] at (0,2) {$p_{ij}$};
+		\node[left] at (1.2,2.8) {$\hat{v}_{ij}$};
+		\node[below] at (1.8,2) {$y_{ij}$};
+		\node[below,red] at (3,2) {$\vec{n}_{ij}$};
+
+		\node[above,orange] at (2.3,2.3) {$\hat{\Theta}_{ij}$};
 	\end{tikzpicture}
@@
 
-The exposure to ship covariate, denoted $E_{ship}$, was defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
-$D_{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
+Exposure to ship covariate at time $t_{ij}$ for narwhal $i$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
+$D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
 This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source.

### 2024-04-24 R/preprocessing/data_preprocessing.Rmd
changes in tex file and mixed effects analysis
preprocessing updates

@@
                       proj4string=CRS("+proj=longlat +datum=WGS84"))
-utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84"))
+utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84 +units=km"))
 
@@
   #step lengths
-  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])/1000
-  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])/1000
+  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])
+  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])
 
@@
 
-# More data exploration
-
-## Time steps distribution
+# Time steps distribution
 
@@
 
-cat("Percentage of time steps > 2h :",length(delta[delta>2])/length(delta)*100)
+cat("Percentage of time steps > 2h :",length(dtimes$delta[dtimes$delta>2])/length(dtimes$delta)*100)
 ```
@@
 
-## Plot all GPS tracks (without interpolation) on a map
+# Time steps distribution during exposure vs no exposure
+
+# Plot all GPS tracks (without interpolation) on a map
 
@@
 ```{r histograms distances to shore}
-histo1=ggplot(dataprep, aes(x = DistanceShore*1000)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (m)")
-
-df=data.frame(DistanceShore=res_coastline$min_distances)
+histo1=ggplot(dataprep[dataprep$DistanceShore>0,], aes(x = DistanceShore)) +
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")
+
+df=data.frame(DistanceShore=res_coastline$min_distances[res_coastline$min_distances>0])
 histo2=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (m)")
-
-df=data.frame(DistanceShore=res_land$min_distances)
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (km)")
+
+df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
 histo3=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (m)")
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")
 ```
@@
 cat("Percentage of points on land with distance from land : \n",
-    length(which(res_land$min_distances/1000==0))/length(dataprep$time)*100,"\n")
+    length(which(res_land$min_distances==0))/length(dataprep$time)*100,"\n")
 
 cat("Percentage of points on land with distance from coastline : \n",
-    length(which(res_coastline$min_distances/1000==0))/length(dataprep$time)*100)
+    length(which(res_coastline$min_distances==0))/length(dataprep$time)*100)
 ```
@@
     coord_sf(datum=st_crs(32626))+
-  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
+  geom_point(data=df,aes(x,y,col=(DistanceShore<0.1)),size=0.2)+
     scale_color_viridis_d()
@@
 
-df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<100 & res_land$min_distances>0  ])
+df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<2 & res_land$min_distances>0  ])
 histo=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (m)")
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")
 
@@
 
-This seems to be pretty uniformly distributed, so let's choose arbitrarily a threshold of 50 metres.
+This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
+the shore in reality.
+Let's choose arbitrarily a threshold of 50 metres.
 
@@
 #replace distance to shore with distances calculated from land polygons (in km)
-dataprep$DistanceShore=min_distances/1000
+dataprep$DistanceShore=min_distances
 
@@
 dataprep$ExpShore <- ExposureFunction(X = dataprep$DistanceShore)
-
-```
-
-# Convert all UTM coordinates in km
-
-```{r convert to km}
-
-dataprep$px=dataprep$px/1000
-dataprep$py=dataprep$py/1000
-dataprep$nx=dataprep$nx/1000
-dataprep$ny=dataprep$ny/1000
-
-dataprep$x=dataprep$x/1000
-dataprep$y=dataprep$y/1000
 
@@
     coord_sf(datum=st_crs(32626))+
-  geom_point(data=sub_data,aes(1000*px,1000*py),shape=5,size=0.1,col="red")+
-  geom_point(data=sub_data,aes(1000*x,1000*y,col=time),size=0.1)+
-    geom_path(data=sub_data,mapping=aes(1000*x,1000*y),linewidth=0.1)+
+  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
+  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
+    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
     geom_point(data = sub_data %>% filter(!duplicated(ID)),
-             aes(1000*x,1000*y,col=ID), shape = 3, size = 4,col = "red")+
+             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
     scale_color_viridis_c()
@@
 
-ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExp)
+ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExpShore)
 ```
@@
 ```{r plots before and after exposure}
-pBE1=milne_land_map + geom_point(dataBE1, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
+pBE1=milne_land_map + 
+  geom_point(dataBE1[,c("ID","Long","Lat","time")],
+             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
+   geom_path(dataBE1[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
+  geom_point(data = dataBE1[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
+             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
   scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
 
-pBE2=milne_land_map + geom_point(dataBE2, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
+pBE2=milne_land_map + 
+  geom_point(dataBE2[,c("ID","Long","Lat","time")],
+             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
+   geom_path(dataBE2[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
+  geom_point(data = dataBE2[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
+             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
   scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
 
-pAE=milne_land_map + geom_point(dataAE, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
+pAE=milne_land_map + 
+  geom_point(dataAE[,c("ID","Long","Lat","time")],
+             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
+   geom_path(dataAE[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
+  geom_point(data = dataAE[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
+             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
   scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

### 2024-04-26 Latex/main.tex
updates in paper manuscript

@@
 More precisely, marine mammals, whose vital functions highly depend on sound perception for communication and orientation, are likely to react to such disturbances.
-The Arctic Council report \cite{halliday_underwater_2020} emphasized the need for more knowledge about the impact of underwater noise on Arctic marine mammals.
+The Arctic Council report \cite{halliday_underwater_2020} emphasizes the need for more knowledge about the impact of underwater noise on Arctic marine mammals.
 The main sources of noise in the region are vessel traffic, ice breakers, drilling, sonars and seismic airguns used for oil and gas exploration.
@@
 \\
-A severity scale was defined in \cite{southall_marine_2008} to help assessing which behavioral responses are more likely to alter a population's capacity to survive, reproduce or forage.
-Changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations  were ranked between 4 and 7 on a scale out of 9 (9 being stranding events that can directly lead to the death of the animal) depending on the magnitude of the changes and their duration.\\
-
-So far, the studies about narwhals behavioral responses to sound exposure proved that they exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}.
+A severity scale is defined in \cite{southall_marine_2008} to help assessing which behavioral responses are more likely to alter a population's capacity to survive, reproduce or forage.
+Changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations  are ranked between 4 and 7 on a scale out of 9 (9 being stranding events that can directly lead to the death of the animal) depending on the magnitude of the changes and their duration.\\
+
+So far, the studies about narwhals behavioral responses to sound exposure prove that they can exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}.
 Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source \cite{tervo_narwhals_2021}.
 These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}.
-Resulting GPS positions of the narwhals have been analyzed with discrete time methods.
-Since GPS positions are retrieved irregularly in time, each narwhal track was linearly interpolated to obtain a position every second, and the positions were then classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second.
-The dynamic of the states was modeled with a Markov chain, assuming that the current state only depends on the last one in time.
-An exposure covariate defined as the inverse of the distance to the sound source was used to model the transition probabilities of the Markov Chain, and it was found that the probabilities of being in states "Close to the shore" and "Moving toward the shore" increased with exposure to the sound, while probability of being in state "Far from the shore" decreased.
-In this paper, we adopt a different approach to describe horizontal movement and make use of a continuous time model to analyse the GPS positions.
-The objective is to be able to quantify  the effect of sound exposure on different characteristics of the movement.\\
-
-We focus on continuous time models defined as a solution of a stochastic differential equation with time-varying coefficients \cite{michelot_varying-coefficient_2021}.
+Though GPS positions are retrieved only when the narwhals emerge to the surface, that is irregularly in time, they have so far been analyzed with discrete time methods relying on linear interpolation to obtain data every second.
+In \cite{heide-jorgensen_behavioral_2021}, the positions are classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second, and the dynamic of the states is modeled with a Markov chain, assuming that the current state only depends on the last one in time.
+An exposure covariate defined as the inverse of the distance to the sound source is then used to model the transition probabilities of the Markov Chain, and the probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases.
+Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model.
+Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one single model.
+\\
+
+We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients.
 One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}.
-It has been used extensively to analyze animal movement in two or even three dimensions \cite{gurarie_correlated_2017,alt_correlation_1990,gurarie_estimating_2011,albertsen_generalizing_2018}.
-This model can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
+It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
 The following formulation of the equations represents an extension of the CVM \cite{johnson_continuoustime_2008}, which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
@@
 \end{equation}
-where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. $W(t)$ is a two-dimensional brownian motion. 
+where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates.
+Here, $W(t)$ is a two-dimensional brownian motion.
 The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
+The case $\omega=0$ corresponds to a standard CVM.
+In most applications, this model is satisfactory, but in case of birds movement, or merely tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
+
 Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
 Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.
-We follow \cite{gurarie_correlated_2017} and refer to this model as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" in the case $\mu \neq (0,0)$.\\
+We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\
 
 In general, the motion of marine mammals is subject to spatial constraints induced by the coastline.
-This is particularly the case for the narwhal data we want to analyse: the animals move in a complex system of fjords in Scoresby Sound (South-East Greenland), imposing strong constraints on the location process.
-Thus, constraints for the shore are needed.
+This is particularly the case for the narwhal data we want to analyse: the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland).
+Thus, constraints for the shore are needed in the location process $X$.
 However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
 Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
-Simulation of constrained paths was achieved by projecting on the coastline the  points that eventually reached land.
-Estimation of the SDE parameters relied on computationally intensive MCMC algorithm (Metroplis-Hastings), and required high frequency data.
-Another idea is to add a repulsive potential function in the drift, which has been done to constrain ants movement within a box \cite{russell_spatially_2018}.
-In our case study, in which the narwhals move in a complex fjords system, this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function.
-Thus, there is a need to develop new models adapted to such spatial constraints.\\
-Moreover, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}.
+In this study, simulation of constrained paths is achieved by projecting on the coastline the  points that eventually reach land.
+Estimation of the SDE parameters then relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data.
+Another idea is to add a repulsive potential function in the drift, which is done to constrain ants movement within a box in \cite{russell_spatially_2018}.
+In our case study, the narwhals move in a complex fjords system, which means that this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function.
+As a consequence, there is a need to develop new models adapted to such spatial constraints.\\
+Additionally, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}.
 For narwhals, it was observed from the data that they have a proclivity to move along the shoreline.
@@
 This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.
-Note that in this framework, the special case $\omega=0$ would be the unconstrained correlated velocity model of \cite{johnson_continuoustime_2008}.
 \\ %More biblio about estimation methods
 We show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
-To make this possible, we formulated a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
+To make this possible, we formulate a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
 Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\
@@
 This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}.
-A response model will be compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
-The effect of sound exposure is assessed both in the constrained and unconstrained ($\omega=0$) frameworks.
+A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
+The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model.
 
@@
 It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots.
-The guns were fired synchronously every 80 s, and the GPS navigation system recorded thelocation of every shot.
+The guns were fired synchronously every 80 s, and the GPS navigation system recorded the location of every shot.
 Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres.
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.5]{images/data_exploration/all_time_steps_histo}
-	\caption{Histogram of time step between consecutive GPS positions}
-	\label{fig:alltimestepshisto}
-\end{figure}
-
 Experiments were divided into unexposed periods - for which the narwhals are not in line of sight with the ship - trial periods - when the narwhals are exposed to the ship and airguns are shot - and intertrial periods - when the narwhals are exposed to the ship but airguns are not shot.
@@
 Figure~\ref{fig: trials and intertrials distributions} shows how these periods are distributed among the $6$ narwhals that were tracked.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[scale=0.4]{images/data_exploration/trials.png}
+In this analysis, we do not distinguish between Trial and Intertrial periods, and treat them both as exposure periods, where the degree of exposure depends on the narwhals positions relative to the ship.
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}{0.49\textwidth}
+    	\centering
+    \includegraphics[width=0.9\textwidth]{images/data_exploration/trials.png}
     \caption{Trial and Intertrial periods for each narwhal}
     \label{fig: trials and intertrials distributions}
-\end{figure}
-
-For each narwhal, the entire track was separated into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period during exposure, after the narwhal has been in line of sight with the ship for the first time.
+    \end{subfigure}
+	\begin{subfigure}{0.49\textwidth}
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/data_exploration/all_time_steps_histo}
+	\caption{Histogram of time steps}
+	\label{fig:alltimestepshisto}
+	\end{subfigure}
+\end{figure}
+
+For each narwhal, the entire track was split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time.
 We kept only GPS measures 24
@@
 The splitting between data before and after exposure resulted in $935$ measures before exposure and $3257$ measures after exposure.
-Table~\ref{table: data distribution} shows how the data is distributed among the different narwhals.
-Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and during exposure.
-
-\begin{table}[H]
-	\centering
-	\begin{tabular}{|c|c|c|}
-		\hline
-		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
-		\hline
-		A1 & 258 & 576\\
-		\hline
-		A2  & 40 & 515 \\
-		\hline
-		A3 & 222 & 680 \\
-		\hline
-		A4 & 291 & 642  \\
-		\hline
-		A5 & 48 & 419\\
-		\hline
-		A6 & 76 & 425 \\
-		\hline
-		\textbf{Total} & \textbf{935} & \textbf{3257} \\
-		\hline
-	\end{tabular}
-	\caption{Distribution of the data among the 6 individuals}
-	\label{table: data distribution}
-\end{table}
-
-\begin{figure}[H]
-    \centering
-\begin{subfigure}{0.49\textwidth}
-    \centering
-    \includegraphics[scale=0.38]{images/data_exploration/all_tracksBE.png}
-    \caption{Tracks before exposure experiments}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-    \centering
-    \includegraphics[scale=0.38]{images/data_exploration/all_tracksAE.png}
-    \caption{Tracks during exposure experiments}
-\end{subfigure}
-\caption{Movement data of East Greenland narwhals}
-\label{fig: tracks before and after exposure}
-\end{figure}
+Table~\ref{table: data distribution} in Appendix shows how the data is distributed among the different narwhals.
+Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.
 
@@
 
-For each narwhal $i \in \{1,\cdots,N\}$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and denote $y_{ij}$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing, for $j \in \{1,\cdots,n_i\}$.
+For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
+We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing..
 
@@
 This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source.
-This covariate is meant to be a proxy for sound exposure levels received by the narwhals.
+This provides a conservative estimate of the effect of the noise exposure.
+The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals.
 Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure.
@@
 \hline
-$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1\}$\\
+$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1,2\}$\\
 \hline
@@
  \hline
- $D_{shore}(t)$ & km & distance between the narwhal and the nearest point in the shore at time $t$ & $\R^+$ \\
+ $D_{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
  \hline
@@
 
+The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
+
 \subsection{Standard Rotational correlated velocity model (RCVM)}
 \label{section: RCVM}
+
+%Discussion, different options
 
@@
 \appendix
+
+\section{More details about the data}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|}
+		\hline
+		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
+		\hline
+		A1 & 258 & 576\\
+		\hline
+		A2  & 40 & 515 \\
+		\hline
+		A3 & 222 & 680 \\
+		\hline
+		A4 & 291 & 642  \\
+		\hline
+		A5 & 48 & 419\\
+		\hline
+		A6 & 76 & 425 \\
+		\hline
+		\textbf{Total} & \textbf{935} & \textbf{3257} \\
+		\hline
+	\end{tabular}
+	\caption{Distribution of the data among the 6 individuals}
+	\label{table: data distribution}
+\end{table}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.49\textwidth}
+		\centering
+		\includegraphics[scale=0.38]{images/data_exploration/all_tracksBE.png}
+		\caption{Tracks before exposure experiments}
+	\end{subfigure}
+	\begin{subfigure}{0.49\textwidth}
+		\centering
+		\includegraphics[scale=0.38]{images/data_exploration/all_tracksAE.png}
+		\caption{Tracks during exposure experiments}
+	\end{subfigure}
+	\caption{Movement data of East Greenland narwhals}
+	\label{fig: tracks before and after exposure}
+\end{figure}
 

### 2024-04-26 R/simulation_study/simulation_study_fjords.Rmd
various updates, including removing parallelization in batches fit cause it takes too much RAM

@@
 #number of batch
-N_BATCH=1
+N_BATCH=10
 
@@
 #duration of each track in hour
-TMAX=0.5*24
+TMAX=1*24
 
@@
 
-#get the coastline geometry from the geojson file
+#get the land and coastline geometries from the geojson file
 land<-st_read(file.path(greenland_data_path,"land_utm.shp"))
+coastline<-st_read(file.path(greenland_data_path,"coastline_utm.geojson"))
 ```
@@
 
-hf=1/60 #1 minute time step
+hf=1/120 #1 minute time step
 
@@
 #mixture of independent gaussian
-ftau_gauss=function(cov_data,D0=0.5,tau0=1,tau1=3) {
+ftau_gauss=function(cov_data,D0=1,tau0=1,tau1=3,alpha=pi/10) {
   Dshore=cov_data$DistanceShore
   theta=cov_data$theta
-  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/(pi/5))^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/(pi/5))^2-(Dshore/D0)^2)
+  tau=tau0+(tau1-tau0)*exp(-((theta-pi/2)/alpha)^2-(Dshore/D0)^2)+(tau1-tau0)*exp(-((theta+pi/2)/alpha)^2-(Dshore/D0)^2)
   return (tau)
@@
 
-ftau_step=function(cov_data,epsilon=pi/10,alpha=pi/5) {
+ftau_bump=function(cov_data,epsilon=pi/10,alpha=pi/5,D0=1) {
     Dshore=cov_data$DistanceShore
     theta=cov_data$theta
-    tau=1.5+0.5*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
-      0.5*(tanh((theta+(-pi/2+epsilon))/alpha)-tanh((theta+(-pi/2-epsilon))/alpha))
+    tau=(1.5+0.5*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
+      0.5*(tanh((theta+(-pi/2+epsilon))/alpha)-tanh((theta+(-pi/2-epsilon))/alpha)))
     return (tau)
@@
 
-fomega=function(cov_data,D0=0.3,omega0=60*pi/2,lambda=2,kappa=0.2) {
+fomega=function(cov_data,D0=0.1,omega0=60*pi,lambda=2,kappa=0.2) {
     Dshore=cov_data$DistanceShore
@@
 colnames(grid)=c("DistanceShore","theta")
-true_par=grid
-true_par$tau=ftau_step(grid)
-true_par$omega=fomega(grid)
+true_tau=matrix(ftau_constant(grid),30,30)
+true_omega=matrix(fomega(grid),30,30)
 
-plot_true_tau <- plot_ly(true_par,type = "mesh3d",
-                         x = ~theta,y = ~DistanceShore,z=~tau,intensity=~tau,
+plot_true_tau <- plot_ly(type = "surface",
+                         x = ~theta,y = ~Dshore,z=~true_tau,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-                layout(title=paste("True tau"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                layout(title=paste("True tau"),
+                       scene=list(xaxis = list(title = "theta",showgrid = F),
                       yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "tau")))
 
-plot_true_omega <- plot_ly(true_par,type = "mesh3d",
-                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
+plot_true_omega <- plot_ly(type = "surface",
+                           contours = list(z = list(show = TRUE, start = min(true_omega), end = max(true_omega), size = 5,color="black")),
+                         x = ~theta,y = ~Dshore,z=~true_omega,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                layout(title=paste("True omega"),
+                       scene=list(xaxis = list(title = "Theta",showgrid = F),
                       yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
@@
 plot_true_omega
+
 ```
@@
 
+#k-th basis spline function
+fomega_spline_basis=function(cov_data,k=1) {
+
+  if (is.null(cov_data$ExpShore)){
+    Xp=predict(m1, newdata = cov_data, type = "lpmatrix")
+
+  }
+  else if (is.null(cov_data$DistanceShore)){
+    Xp=predict(m2, newdata = cov_data, type = "lpmatrix")
+  }
+  coeffs=rep(0,ncol(Xp))
+  coeffs[k]=1
+  return(as.numeric(Xp%*%coeffs))
+}
+
 ```
@@
 theta=seq(from=-pi,to=pi,length.out=30)
-grid<- as.data.frame(expand.grid(DistanceShore,theta))
+grid<- as.data.frame(expand.grid(Dshore,theta))
 colnames(grid)=c("DistanceShore","theta")
-true_par=grid
-true_par$omega=fomega_splines(grid)
+spline_omega=matrix(fomega_splines(grid),30,30)
 
-plot_true_omega <- plot_ly(true_par,type = "mesh3d",
-                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
+# Extract the basis function values for the k-th basis function on the grid
+z2=matrix(fomega_spline_basis(grid,k=2),30,30)
+z3=matrix(fomega_spline_basis(grid,k=3),30,30)
+z4=matrix(fomega_spline_basis(grid,k=4),30,30)
+
+plot_spline_omega <- plot_ly(type = "surface",
+                             contours = list(z = list(show = TRUE, start = min(spline_omega), end = max(spline_omega), size = 5,color="black")),
+                         x = ~Dshore,y = ~theta,z=~spline_omega,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                layout(title=paste("Spline smooth omega"),scene=list(xaxis = list(title = "Theta",showgrid = T),
                                                                  yaxis = list(title = "Distance to shore",showgrid = F),
@@
 
-plot_true_omega
+plot_basis2 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z2), end = max(z2), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z2, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 2"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_basis3 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z3), end = max(z3), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z3, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 3"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_basis4 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z4), end = max(z4), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z4, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 4"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega"))) 
 ```
+
+```{r}
+plot_spline_omega
+```
+
+```{r}
+plot_basis2
+```
+
+```{r}
+plot_basis3
+```
+
+```{r}
+plot_basis4
+```
+
 ## Simulation with low measurement errors and high frequency
@@
   geom_path(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_point(data = data_fjords_hf_ne$BATCH1%>% filter(!duplicated(ID)),
+  geom_point(data = data_fjords_hf_ne$BATCH2%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+

### 2024-04-26 R/simulation_study/simulation_study_rect.Rmd
change number of batches
various updates, including removing parallelization in batches fit cause it takes too much RAM

@@
 #number of batch
-N_BATCH=10
+N_BATCH=1
 
 #number of individual tracks per batch
-N_ID=6
+N_ID=3
 
@@
 SP_DF=5
+
 ```
@@
 colnames(grid)=c("DistanceShore","theta")
-true_par=grid
-true_par$tau=ftau_constant(grid)
-true_par$omega=fomega(grid)
+true_tau=matrix(ftau_constant(grid),30,30)
+true_omega=matrix(fomega(grid),30,30)
 
-plot_true_tau <- plot_ly(true_par,type = "mesh3d",
-                         x = ~theta,y = ~DistanceShore,z=~tau,intensity=~tau,
+plot_true_tau <- plot_ly(type = "surface",
+                         x = ~Dshore,y = ~theta,z=~true_tau,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-                layout(title=paste("True tau"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                layout(title=paste("True tau"),
+                       scene=list(xaxis = list(title = "Theta",showgrid = F),
                       yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "tau")))
 
-plot_true_omega <- plot_ly(true_par,type = "mesh3d",
-                         x = ~theta,y = ~DistanceShore,z=~omega,intensity=~omega,
+plot_true_omega <- plot_ly(type = "surface",
+                           contours = list(z = list(show = TRUE, start = min(true_omega), end = max(true_omega), size = 5,color="black")),
+                         x = ~Dshore,y = ~theta,z=~true_omega,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-                layout(title=paste("True omega"),scene=list(xaxis = list(title = "Theta",showgrid = F),
+                layout(title=paste("True omega"),
+                       scene=list(xaxis = list(title = "Theta",showgrid = F),
                       yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
@@
 
+#k-th basis spline function
+fomega_spline_basis=function(cov_data,k=1) {
+
+  if (is.null(cov_data$ExpShore)){
+    Xp=predict(m1, newdata = cov_data, type = "lpmatrix")
+
+  }
+  else if (is.null(cov_data$DistanceShore)){
+    Xp=predict(m2, newdata = cov_data, type = "lpmatrix")
+  }
+  coeffs=rep(0,ncol(Xp))
+  coeffs[k]=1
+  return(as.numeric(Xp%*%coeffs))
+}
+
 ```
 
-### Plot
+### Plot spline omega and spline basis functions
 
@@
 
-plot_true_omega
+plot_basis2 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z2), end = max(z2), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z2, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 2"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_basis3 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z3), end = max(z3), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z3, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 3"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega")))
+
+plot_basis4 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z4), end = max(z4), size = 0.05,color="black")),
+                         x = ~Dshore,y = ~theta,z=~z4, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                layout(title=paste("Spline basis 4"),scene=list(xaxis = list(title = "Theta",showgrid = T),
+                                                                 yaxis = list(title = "Distance to shore",showgrid = F),
+                                                            zaxis = list(title = "omega"))) 
+```
+
+```{r}
+plot_spline_omega
+```
+
+```{r}
+plot_basis2
+```
+
+```{r}
+plot_basis3
+```
+
+```{r}
+plot_basis4
 ```
@@
 
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_rect_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
+crcvm_rect_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines
@@
 }
-
-#stop cluster
-stopCluster(cl)
 
@@
 
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_rect_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
+crcvm_rect_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines
@@
 ggplotly(p)
+```
+
+#### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
+
+n_coeffs=length(sp_coeff_Dshore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+  m<-crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]
+  coeffs=m$coeff_re()
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+  sub_index=grepl("omega",coeff_names)
+  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.html")
+
+p
 ```
@@
 ```
-
-#### Violin plot of the estimated spline coefficients for all batches
-
-```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
-
-n_coeffs=length(sp_coeff_Dshore)
-coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
-
-for (k in 1:N_BATCH) {
-  m<-crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
-  coeff_names=rownames(coeffs)
-  coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
-  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
-}
-
-#remove initialization row
-coeffs_df=coeffs_df[-1,]
-
-p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
-                xlab(" ")+labs(fill = "Estimated coefficients")+
-                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
-                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
-
-p=ggplotly(p)
-saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.html")
-
-p
-```

### 2024-04-26 R/preprocessing/data_preprocessing.Rmd
some changes in exploratory graphs

@@
 \item Keep GPS positions indicated by value 1 in \texttt{pos\_org} column
-\item  Project to UTM zone 26. The narwhal positions are in latitude and longitude.To analyze the track and visualize it better we project it on a plane in UTM coordinates.
+\item  Project to UTM zone 26. The narwhal positions are in latitude and longitude.To analyze the track and visualize it better we project it on a plane in UTM coordinastes.
 The positions of the narwhals are in the UTM zone 26 north (in Greenland).
@@
 
+#remove rows with unknown positions
 dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]
+
+#keep rows indicated as GPS measurements
 dataprep=dataprep[dataprep$pos_org==1,]
+
+#number of rows
 nrows=length(dataprep[,1])
@@
 ```{r display land , fig.cap="Land polygons"}
-plot_land
+ggplotly(plot_land)
 ```
@@
 ```{r compare distances to shore}
-plot(res_land$min_distances/1000-dataprep$DistanceShore)
-plot(res_coastline$min_distances/1000-dataprep$DistanceShore)
+plot(res_land$min_distances-dataprep$DistanceShore)
+plot(res_coastline$min_distances-dataprep$DistanceShore)
 ```
@@
 summary(res_coastline$min_distances)
-summary(dataprep$DistanceShore*1000)
+summary(dataprep$DistanceShore)
 ```
@@
     coord_sf(datum=st_crs(32626))+
-  geom_point(data=df,aes(x,y,col=(DistanceShore<0.1)),size=0.2)+
+  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
     scale_color_viridis_d()

### 2024-04-26 Latex/main.tex
changes in part 2

@@
 
-For each narwhal, the entire track was split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time.
-We kept only GPS measures 24
+For each narwhal, the entire track is split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time.
+We keep only GPS measures 24
 hours after tagging, to avoid any tagging effects on the narwhals's movement, as recommended in \cite{heide-jorgensen_behavioral_2021}.
-Measures that resulted in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
-Overall, $4192$ GPS positions were kept for the analysis and projected to UTM coordinates zone 26.
-The splitting between data before and after exposure resulted in $935$ measures before exposure and $3257$ measures after exposure.
+Measures that result in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
+Overall, $4192$ GPS positions are kept for the analysis and projected to UTM coordinates zone 26.
+The splitting between data before and after exposure results in $935$ measures before exposure and $3257$ measures after exposure.
 Table~\ref{table: data distribution} in Appendix shows how the data is distributed among the different narwhals.
@@
 For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing..
+We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.
 
@@
 	\end{tikzpicture}
-	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain.}
+	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain, $y_{ij}$ is the observed GPS position of narwhal $i$ at time $t_{ij}$.}
 	\label{fig: illustration nearest shore points}
@@
 
-Exposure to ship covariate at time $t_{ij}$ for narwhal $i$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
+Exposure to ship covariate for narwhal $i$ at time $t_{ij}$ , denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
 $D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
@@
 \end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
-This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2019}.
 There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
@@
 
-By integrating the velocity equation over time one gets the following distribution for the location process :
+By integrating the velocity equation over time one gets the following conditional distribution for the location process :
 \begin{equation}

### 2024-04-26 R/simulation_study/simulation_study_rect.Rmd
updates

@@
 #number of batch
-N_BATCH=1
+N_BATCH=10
 
 #number of individual tracks per batch
-N_ID=3
+N_ID=6
 
@@
 xmax=list("theta"=pi,"ExpShore"=1/0.3)
-xlabel=list("ExpShore"="Distance to shore")
-link=list("ExpShore"=(\(x) 1/x))
+xlabel=list("ExpShore"="Distance to shore","theta"="Theta")
+link=list("ExpShore"=(\(x) 1/x),"theta"=identity)
 true_smooths=list("omega"=fomega_splines)
 res=crcvm_rect_hf_ne1$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne1",xmin=xmin,xmax=xmax,
-                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=TRUE)
+                                    xlabel=xlabel,link=link,show_CI="pointwise",true_smooths=true_smooths,save=FALSE)
 res$fe_omega_theta_ExpShore

### 2024-04-27 Latex/main.tex
paper updates

@@
 The case $\omega=0$ corresponds to a standard CVM.
-In most applications, this model is satisfactory, but in case of birds movement, or merely tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
+In most applications, this model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
 
@@
 \\ %More biblio about estimation methods
-We show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
+
+Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
 To make this possible, we formulate a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
@@
 
-Section 2 gives an overview of the narwhal data available for the analysis.
-Then in Section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models.
-Section 4 is dedicated to a comprehensive simulation study, whereas in the last section, the results of the application to the narwhal data are examined.
-
-\section{Movement data of Greenland endemic narwhals}
-%describe how the data was obtained, the geography, type of whales, year, exposure to boats etc..
-%trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime
+The diffusion models are discussed in Section 2
+Then,  Section 3 is dedicated to a comprehensive simulation study, whereas in the last section,we give an overview of the narwhal data available for the analysis and the results of the application of our models to the narwhal data.
+
+\section{Diffusion models}
+
+The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
+First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement.
+
+\subsection{Standard Rotational Advective correlated velocity model (RACVM)}
+\label{section: RACVM}
+
+\subsubsection{Solution of the SDE}
+
+We write $A=\begin{pmatrix} 
+	\frac{1}{\tau} & -\omega \\
+	\omega & \frac{1}{\tau}
+\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}
+
+\begin{equation*} \left\{
+	\begin{array}{l}
+		dX(t)=V(t)dt \\
+		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
+	\end{array}
+	\right.
+	\label{eq: RACVM equation bis}
+\end{equation*}
+where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
+A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
+In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
+Then the Euler-scheme is used to get the transitions of the location process.
+Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula.
+Though this model is discussed in \cite{gurarie_correlated_2017}, we found no mention of these explicit formulas.
+\\
+
+The transition density for the velocity process is such that 
+\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
+\end{equation}
+
+where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
+Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.
+
+%stationary distribution, distribution of the norm, distribution of the phase
+This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}.
+After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$.
+In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.
+
+By integrating the velocity equation over time one gets the following conditional distribution for the location process :
+\begin{equation}
+	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
+\end{equation}
+Here, the covariance matrix $Q(\Delta)$ has an explicit form.
+
+\[
+Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
+\]
+The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
+Both components of the location and velocity processes are independent.
+and one can check that for $\omega=0$, the formulas match the ones of the  CTCRW model \cite{johnson_continuoustime_2008}.
+Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase.
+Derivation of the formulas given in this paragraph is detailed in appendix~\ref{section: detailed RACVM computations}.\\
+
+%samples with non zero and zero mu
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
+		\caption{}
+	\end{subfigure}
+	\caption{(a) Samples of a RACVM with parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h.
+The black cross indicates the starting point of the process.
+The samples were obtained from the exact transition densities.
+(b) Histogram of the empirical velocities obtained from the samples, along with the theoretical probability density function.}
+	\label{fig: sample RACVM}
+\end{figure}
+
+\subsubsection{State-space model}
+In most applications, it is needed to consider measurement errors through a state-space model.
+We suppose that positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$.
+$Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$.
+
+\begin{equation}  \left\{
+	\begin{array}{l}
+		Y_j=X_j+\varepsilon_j \quad j \in \{1,\cdots,n\}\\
+		\varepsilon_j \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
+		dX(t)=V(t)dt  \\
+		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
+	\end{array}
+	\right.
+\end{equation}
+
+Here, we consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
+We make the approximation that these parameters are constant on each time step $[t_j,t_{j+1}]$ .
+Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $V_j=V(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
+	\frac{1}{\tau_j} & -\omega_j \\
+	\omega_j & \frac{1}{\tau_j}
+\end{pmatrix}$.
+
+\begin{equation}
+	\forall j \in \{1, \cdots, n\} \quad 
+	\left\{ \begin{array}{l}
+		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
+		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
+	\end{array}\right. \\
+	\label{eq: RCVM autoregressive formula}
+\end{equation}
+
+where $\zeta_j$ and $\xi_j$ are centered normal random vectors.
+Denote $\alpha_j=\begin{pmatrix} X_{j,1}  & X_{j,2} & V_{j,1} & V_{j,2}\end{pmatrix}^\top$. The state space equations are
+
+\[
+\left\{
+\begin{array}{l}
+	Y_j=Z\alpha_j+\varepsilon_j \\
+	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
+\end{array}
+\right.\]
+where the link matrices are 
+\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
+	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
+	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
+\]
+
+$\varepsilon_j \sim \mathcal{N}(0,\sigma_{obs}^2 I_2)$ and the vector $\eta_j=\begin{pmatrix} \xi_{j,1} & \xi_{j,2} & \zeta_{j,1} & \zeta_{j,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero.
+Its covariance matrix $Q_j$ can be computed explicitly.
+\begin{equation}
+	Q_j=\begin{pmatrix}
+		M_j & L_j \\
+		L_j & N_j
+	\end{pmatrix}
+	\label{eq: state space covariance}
+\end{equation}
+where $M_j$, $N_j$and $L_j$ are block matrices of size $2 \times 2$ such that 
+\[M_j=\left( \Delta-2 \frac{\omega_j \sin(\omega_j \Delta_j)-\frac{1}{\tau_j} \cos(\omega_j \Delta_j)}{\frac{1}{\tau_j^2}+\omega_j^2 } \exp\left( -\frac{\Delta_j}{\tau_j} \right) +\frac{\tau_j}{2} \left( \frac{\omega_j^2-\frac{3}{\tau_j^2}}{\frac{1}{\tau_j^2}+\omega_j^2}-\exp\left( -\frac{2\Delta_j}{\tau_j}\right)\right) \right)I_2\]
+
+\[N_j=\frac{2\nu_j^2}{\pi}\left(1-e^{-\frac{2 \Delta_j}{\tau_j}}\right)I_2\]
+
+\[L_j=\begin{pmatrix} l_1 & l_2 \\
+	l_2 & l_1\end{pmatrix}\]
+where 
+\[l_1=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times \left( 1+\exp\left( -\frac{2\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right) \cos(\omega_j \Delta_j)\right)\]
+\[l_2=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times\left( \exp\left( -\frac{\Delta_j}{\tau_j}\right) \sin(\omega_j \Delta_j)-\frac{\omega_j \tau_j}{2} \left(1-\exp\left( -2 \frac{\Delta_j}{\tau_j}\right) \right)\right)\]
+
+One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model \cite{johnson_continuoustime_2008}.
+This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter.
+This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} the special case $\omega=0$.
+Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.
+
+\subsection{Constrained rotational correlated velocity model}
+\label{section: CRCVM}
+
+Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
+To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
+This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
+Denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
+\begin{equation} \left\{
+	\begin{array}{l}
+		dX(t)=V(t) dt \\
+		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)
+
+	\end{array}
+	\right.
+	\label{eq: CRCVM equation}
+\end{equation}
+
+Here, $A(X(t),V(t))=\begin{pmatrix} 
+	\frac{1}{\tau} & -\omega(t) \\
+	\omega(t) & \frac{1}{\tau}
+\end{pmatrix}$
+where $\omega(t)=f_{\omega}(D_{shore}(t),\Theta(t))$, $D_{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
+$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
+
+Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
+For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
+\[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
+In this case, the formulas of section~\ref{section: RACVM} are still valid.
+The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
+We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
+An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation.
+For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$.
+We hint that the condition should also include the regularity of the domain, but this will be a subject for further work.
+Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D_{shore}$ approached $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
+The closer we are to the shore, the faster the velocity should rotate. 
+whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
+Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
+A weight based on the distance to the shore can be defined as follows :
+\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
+
+Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :
+
+\begin{equation}
+f_{\omega}(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
+\label{eq: smooth omega}
+\end{equation}
+for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
+
+This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
+Any smooth function that meets minimum requirements could be used instead.
+The general model is very flexible, and different species-specific behaviours could be described.
+An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
+\\
+Figure~\ref{fig: trueomega} shows a smooth function obtained from~\ref{eq: smooth omega} for the parameter $\omega$.
+Figure~\ref{fig: splineomega} shows three spline basis functions and a smooth function $f_{\omega}$ obtained as a combination of these basis functions.
+
+\begin{figure}[H]
+	\centering
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
+		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
+			(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
+		\label{fig: trueomega}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
+		\caption{}
+	\end{subfigure}
+	\caption{}
+	\label{fig: splineomega}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_rect.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_fjords.png}
+		\caption{}
+	\end{subfigure}
+	\caption{Samples of the CRCVM in different domains.
+x and y axis are in metres and initial positions are indicated by a red cross and red point represent the nearest points on the boundary of the domain (a) One sample three day sample with time step $h=1$ min in a rectangular domain (b) One sample three day sample with time step $h=1$ min in Scoresby Sound fjords system in South-East Greenland}
+	\label{fig: samples_rect_CRCVM}
+\end{figure}
+
+\section{Simulation study}
+
+\label{section: simulation study}
+%samples and estimation
+%discussion about tensor splines vs additive splines
+
+\subsection{Simulation of constrained motion}
+In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.\\
+
+$3$ days long trajectories were sampled in a rectangular domain with sides of length $8$ km from the CRCVM with the smooth function for $\omega$ displayed in figure~\ref{fig: trueomega}(a) and other parameters kept constant ($\nu = 4$ km/h, $\tau=1$ h).
+For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
+Time steps were chosen equal to $1$ min for high frequency and subsampled to $5$ minutes for low frequency.
+Gaussian measurement errors with low standard deviation $\sigma_{obs}=3$ m and large standard deviations $\sigma_{obs}=30$ m were added to provide the actual observations.
+Of these $100$ trajectories, none reach the boundary for $\sigma_{obs}=3$ m and only one for $\sigma_{obs}=30$m.
+However, too small values of $D_0$ and $\omega_0$, as well as too large measurement errors may produce samples that reach the boundary of the domain.
+Figure~\ref{fig: samples_rect_CRCVM} shows some of the 100 samples obtained in the rectangular domain.
+\\
+
+10 trajectories were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints.
+The parameter $\omega$ was defined as the smooth function displayed in~\ref{fig: trueomega}(b).
+Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
+Time steps were also kept constant, equal to $1$ min for high frequency and subsampled to $5$ min for low frequency.
+None of the 10 trajectories reached land.
+Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
+For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
+Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}.
+Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
+
+\subsection{Estimation with tensor splines}
+
+The samples from previous section were used to estimate the function $\omega$, both in the rectangular domain and in Scoresby Sound fjords system.
+Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
+The marginal spline degree of freedom was set to $5$ for each covariate, which corresponds to $3$ knots in the splines.
+Hence, $25$ coefficients (including the intercept) were to be estimated.
+The other parameters of the diffusion were supposed to be known, as well as the measurement error.
+Underestimation of the smooth parameter $\omega$ is observed close to the shore as $\Theta$ approaches $\pm \pi$ (movement toward the shore).
+This is easily seen in figures \ref{fig: marginal_estimates_rect_CRCVM}(a)-(b) and and \ref{fig: marginal_estimates_fjords_CRCVM}(a)-(b).
+
+As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging.
+THta is why slight differences in figures ~\ref{fig: marginal_estimates_rect_CRCVM}(a)-(b).
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in the rectangular domain.
+(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+	\label{fig: estimate_rect_CRCVM}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_close.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_close.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_far.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_far.png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in the rectangular domain.
+The red curve represents the true value of $\omega$.
+(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
+(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+\label{fig: marginal_estimates_rect_CRCVM}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in Scoresby Sound fjords system.
+(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+	\label{fig: estimate_fjords_CRCVM}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
+		\caption{}
+	\end{subfigure}
+
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in Scoresby Sound fjords system.
+The red curve represents the true value of $\omega$.
+(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
+(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
+	\label{fig: marginal_estimates_fjords_CRCVM}
+
+\end{figure}
+
+\section{Application to narwhals behavioral response}
 
@@
 
-\subsection{Definition of the covariates}
+\subsection{Notations and definition of the covariates}
 
@@
 The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
-The land polygons define the boundaries of the domain in which the narwhals can move.
+The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move.
 The closest point on the shoreline to the observed position $y_{ij}$  is denoted $p_{ij}$, and the distance to this point, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$.
@@
 
-\section{Diffusion models}
-
-The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
-
-\subsection{Standard Rotational correlated velocity model (RCVM)}
-\label{section: RCVM}
-
-%Discussion, different options
-
-A general formulation for a rotational correlated velocity model is \cite{albertsen_generalizing_2018}:
-
-\begin{equation} \left\{
-	\begin{array}{l}
-		dX(t)=V(t) dt \\
-		dV(t)=-A(V(t)-\mu)dt+\Sigma dW(t)
-
-	\end{array}
-	\right.
-	\label{eq: general RACVM equation}
-\end{equation}
-where $A=\begin{pmatrix} 
-	\frac{1}{\tau_1} & -\omega \\
-	\omega & \frac{1}{\tau_2}
-\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
-There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
-
-It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$ \cite{johnson_continuoustime_2008}.
-The diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$, where $\nu$ is homogeneous to a velocity.
-This allows easier biological interpretation  of the parameters in the equation \cite{gurarie_correlated_2017}.
-These simplifications produce the equation~\ref{eq: RACVM equation}.
-
-When parameters $\tau$, $\mu$ and $\nu$ are constant over time, both the velocity and the position processes can be solved explicitly.
-The transition density for the velocity process is such that 
-\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
-\end{equation}
-
-Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a rotation matrix of angle $-\omega \Delta$.
-Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.
-
-%stationary distribution, distribution of the norm, distribution of the phase
-The stationary distribution of the velocity process is $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$.
-This result can be used to get the long term distribution of the velocity norm and the velocity phase angle.
-After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$.
-In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$.
-Hence, the order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm.
-The distribution of the phase angle $\Phi$ of the velocity has been derived in \cite{pawula_distribution_1982}.
-When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$.
-Hence, taking a look at the histograms of the phase angle of the data can help to assess whether a non zero mean velocity is necessary in the model or not.
-
-By integrating the velocity equation over time one gets the following conditional distribution for the location process :
-\begin{equation}
-	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
-\end{equation}
-Here, the covariance matrix $Q(\Delta)$ has an explicit form.
-
-\[
-	Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
-\]
-The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
-Both components of the location and velocity processes are independent.
-and one can check that for $\omega=0$, the formulas match the ones of the  CTCRW model \cite{johnson_continuoustime_2008}.
-Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase.
-Derivation of the formulas given in this paragraph is detailed in appendix~\ref{section: detailed RACVM computations}.\\
-
-%samples with non zero and zero mu
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
-		\caption{}
-	\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/rcvm/tau5_nu1_omega0mu10_sigmaobs0.03 .png}
-	\caption{}
-\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
-		\caption{}
-	\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
-	\caption{}
-\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
-	\caption{}
-\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
-	\caption{}
-\end{subfigure}
-	\caption{(a)-(b) Two samples of a RCVM with respective parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h  and $\tau=5 h$, $\nu=1$ km/h,$\omega=0$ rad/s, $\mu=(0,0)$ km/h.
-The black cross indicates the starting point of the process.
-(c) - (d) Histograms of the empirical velocities obtained from the samples, along with the theoretical probability density function.
-(e) - (f) Histograms of the empirical velocity phase angles obtained from the samples, along with the theoretical probability density function.
-}
-	\label{fig: sample RACVM}
-\end{figure}
-
-\subsection{Constrained rotational correlated velocity model}
-\label{section: CRCVM}
-
-To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
-This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
-Denote $\mathcal{D} \subset \R^2$ the domain of the process $X$.
-\begin{equation} \left\{
-	\begin{array}{l}
-		dX(t)=V(t) dt \\
-		dV(t)=-A(X(t),V(t))v(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)
-
-	\end{array}
-	\right.
-	\label{eq: CRCVM equation}
-\end{equation}
-
-Here, $A(X(t),V(t))=\begin{pmatrix} 
-	\frac{1}{\tau} & -\omega(D_{shore}(t),\Theta(t)) \\
-	\omega(D_{shore}(t),\Theta(t)) & \frac{1}{\tau}
-\end{pmatrix}$
-with $D_{shore}(t)=d(X(t),\mathcal{D})$,
- $\Theta(t)=\widehat{(X(t)-p(t),V(t))}$ and $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$.
-
- Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
- For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
- \[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
- In this case, the formulas of section~\ref{section: RCVM} are still valid.
- The question still hanging is : what are the conditions on the function $\omega(D_{shore},\Theta)$ to ensure that the process is constrained ?
- We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $x$.
- An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation.
-For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$.
-Moreover, a rotation needs to be considered only for some specific values of the angle $\Theta$, more precisely $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$, that is when the narwhal is heading in the direction of the shore.
- The closer we are to the shore, the faster the velocity should rotate. 
- whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
-Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
- A weight based on the distance to the shore can be defined as follows :
- \[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
-
- Based on this weight, a relationship between $\omega$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :
-
- \[
- \omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
- \]
- for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
-
- This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
-Any smooth function that meets the minimum requirements could be used instead.
-The general model is very flexible, and different species-specific behaviours could be described.
-For the narwhals, a tendency to move along the shore could be considered by specifying a function $\omega$ that is non zero even when $\Theta \in \left[\frac{\pi}{2}-\varepsilon,\frac{\pi}{2}\right] \cup \left[-\frac{\pi}{2},-\frac{\pi}{2}+\varepsilon\right]$, so that when the animal would be on the verge of leaving the shore, its velocity would be tweaked to align with the shoreline.
-It is also possible  to specify $\tau$ similarly as a function of $D_{shore}$ and $\Theta$ to model specific behaviour near the shore (e.g higher persistence along the shoreline).\\
- Figure~\ref{fig: trueomega} shows the smooth function obtained for the parameter $\omega$ and different values of $D_0$, $\lambda$ and $\kappa$.
-
- \begin{figure}[H]
- 	\centering
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.3]{images/simulation study/true_omega_rect.png}
- 		\caption{}
- 	\end{subfigure}
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.3]{images/simulation study/true_omega_fjords.png}
- 		\caption{}
- 	\end{subfigure}
- 	\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
- 		(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min. 
- 		(b)  $D_0=0.3$ km, $\lambda=2$, $\kappa=0.2$ and $\omega_0=\frac{\pi}{2}$ rad/min.}
- 	\label{fig: trueomega}
- \end{figure}
-
- \begin{figure}[H]
- 	\centering
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_ne_omega.png}
- 		\caption{}
- 	\end{subfigure}
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.4]{images/simulation study/samples_rect_hf_e_omega.png}
- 		\caption{}
- 	\end{subfigure}
- 	\caption{Samples of the CRCVM in a rectangular domain.
-x and y axis are in metres and initial positions are indicated by a red cross (a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
- 	\label{fig: samples_rect_CRCVM}
- \end{figure}
-
- \begin{figure}[H]
- 	\centering
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_ne_omega.png}
- 		\caption{}
- 	\end{subfigure}
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.5]{images/simulation study/samples_fjords_hf_e_omega.png}
- 		\caption{}
- 	\end{subfigure}
- 	\caption{Samples of the CRCVM in Scoresby Sound coastline geometry.
-x and y axis are in metres and initial positions are indicated by a red cross.
-(a)-(b) Two samples of time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
- 	\label{fig: samples_fjords_CRCVM}
- \end{figure}
-
-%conditions on the shore (regularity) and time steps
-%discuss conditions on omega and tau, existence of solution etc
-
 \subsection{Mixed effet model for ship exposure and shore effects}
@@
 Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations are discussed in section~\ref{section: inference}.\\
-
-\subsection{Inference}
-\label{section: inference}
-
-\subsubsection{State space model for the rotational CVM}
-
-As discussed in \cite{ditlevsen_hypoelliptic_2019}, estimation for such hypoelliptic models is challenging. The main difficulties come from
-\begin{itemize}
-	\item degenerate noise structure that prevents the use of standard approximation schemes as Euler-Maruyama.
-	 \item irregular times of observations : we only get position when the marine mammal reaches the sea surface.
-	 \item GPS measurement errors, with gaussian or student distribution.
-	 \item partial observations : two processes are involved in equation~\ref{eq: RACVM equation}, however only one of them can be observed, that is the position.
-\end{itemize}
-
-Several method already exist to estimate constant parameters $\tau$, $\nu$ and $\mu$.
-If the parameters are constant, it is possible to use a method of moments based on the distributional results about the velocity process \cite{gurarie_correlated_2017}, but it is only feasible with high frequency data and negligible measurement errors.
-Maximum likelihood with Kalman filter allows to estimate more efficiently from observations with errors \cite{johnson_continuoustime_2008}.
-If the parameters are allowed to depend on time through external covariates, then \texttt{smoothSDE} package, which is based on \cite{johnson_continuoustime_2008} but exploit the capabilities of both the \texttt{mgcv} package \cite{wood_generalized_2017} to specify the time dependency through covariates, and the \texttt{TMB} \cite{kristensen_tmb_2016} package to integrate over the random effect and optimize the likelihodd, is a convenient tool for estimation.
-%Discuss the improvement/differences compared to Gurarie and Albertsen, Michelot etc
-
-Consider equation~\ref{eq: RACVM equation} with parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
-If we suppose these parameters to be constant on each time step $[t_j,t_{j+1}]$ it can be solved explicitly.
-Denote $\Delta_j=t_{j+1}-t_j$, $v_j=v(t_j)$, $x_j=x(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
-	\frac{1}{\tau_j} & -\omega_j \\
-	\omega_j & \frac{1}{\tau_j}
-\end{pmatrix}$. Then applying Itô's formula gives
-
-\begin{equation}
-	\forall j \in \{1, \cdots, n\} \quad 
-	\left\{ \begin{array}{l}
-		v_{j+1}=\mu_j +\exp(-A_j\Delta_j)(v_j-\mu_j)+\zeta_j \\
-		x_{j+1}=x_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(v_j-\mu_j)+\xi_j
-	\end{array}\right. \\
-	\label{eq: RCVM autoregressive formula}
-\end{equation}
-
-where $\zeta_j$ and $\xi_j$ are centered normal random variables.\\
-Denote $\alpha_j=\begin{pmatrix} x_{j,1}  & x_{j,2} & v_{j,1} & v_{j,2}\end{pmatrix}^\top$ the hidden state and $y_j$ the observed GPS position at time $t_j$.
-Measurement errors are supposed to be gaussian in order to be able to use the basic Kalman filter equations.
-The state space equations are
-
-\[
-\left\{
-\begin{array}{l}
-	y_j=Z\alpha_j+\varepsilon_j \\
-	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
-\end{array}
-\right.\]
-where the link matrices are 
-\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
-	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
-	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
-\]
-
-$\varepsilon_j \sim \mathcal{N}(0,\sigma_{obs}^2 I_2)$ and the vector $\eta_j=\begin{pmatrix} \xi_{j,1} & \xi_{j,2} & \zeta_{j,1} & \zeta_{j,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero.
-Its covariance matrix $Q_j$ can be computed explicitly.
-\begin{equation}
-	Q_j=\begin{pmatrix}
-		M_j & L_j \\
-		L_j & N_j
-	\end{pmatrix}
-	\label{eq: state space covariance}
-\end{equation}
-where $M_j$, $N_j$and $L_j$ are block matrices of size $2 \times 2$ such that 
-\[M_j=\left( \Delta-2 \frac{\omega_j \sin(\omega_j \Delta_j)-\frac{1}{\tau_j} \cos(\omega_j \Delta_j)}{\frac{1}{\tau_j^2}+\omega_j^2 } \exp\left( -\frac{\Delta_j}{\tau_j} \right) +\frac{\tau_j}{2} \left( \frac{\omega_j^2-\frac{3}{\tau_j^2}}{\frac{1}{\tau_j^2}+\omega_j^2}-\exp\left( -\frac{2\Delta_j}{\tau_j}\right)\right) \right)I_2\]
-
-\[N_j=\frac{2\nu_j^2}{\pi}\left(1-e^{-\frac{2 \Delta_j}{\tau_j}}\right)I_2\]
-
-\[L_j=\begin{pmatrix} l_1 & l_2 \\
-	l_2 & l_1\end{pmatrix}\]
-where 
-\[l_1=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times \left( 1+\exp\left( -\frac{2\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right) \cos(\omega_j \Delta_j)\right)\]
-\[l_2=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times\left( \exp\left( -\frac{\Delta_j}{\tau_j}\right) \sin(\omega_j \Delta_j)-\frac{\omega_j \tau_j}{2} \left(1-\exp\left( -2 \frac{\Delta_j}{\tau_j}\right) \right)\right)\]
-
-One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model.
-We coded these formulas in the \texttt{smoothSDE} package to allow fast maximum likelihood inference of smooth parameters $\tau$,$\nu$ and $\omega$ by making the most of the power of the R package \texttt{TMB}.
 
@@
 
-\section{Simulation study}
-\label{section: simulation study}
-
-%samples and estimation
-%discussion about tensor splines vs additive splines
-
-\subsection{Simulation of constrained motion}
-In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.\\
-
-$3$ days long trajectories were sampled in a rectangular domain with sides of length $8$ km from the CRCVM with the smooth function for $\omega$ displayed in figure~\ref{fig: trueomega}(a) and other parameters kept constant ($\nu = 4$ km/h, $\tau=1$ h).
-For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
-Time steps were chosen equal to $1$ min for high frequency and subsampled to $5$ minutes for low frequency.
-Gaussian measurement errors with low standard deviation $\sigma_{obs}=3$ m and large standard deviations $\sigma_{obs}=30$ m were added to provide the actual observations.
-Of these $100$ trajectories, none reach the boundary for $\sigma_{obs}=3$ m and only one for $\sigma_{obs}=30$m.
-However, too small values of $D_0$ and $\omega_0$, as well as too large measurement errors may produce samples that reach the boundary of the domain.
-Figure~\ref{fig: samples_rect_CRCVM} shows some of the 100 samples obtained in the rectangular domain.
-\\
-
-10 trajectories were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints.
-The parameter $\omega$ was defined as the smooth function displayed in~\ref{fig: trueomega}(b).
-Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
-Time steps were also kept constant, equal to $1$ min for high frequency and subsampled to $5$ min for low frequency.
-None of the 10 trajectories reached land.
-Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
-For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
-Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}.
-Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-
-\subsection{Estimation with tensor splines}
-
-The samples from previous section were used to estimate the function $\omega$, both in the rectangular domain and in Scoresby Sound fjords system.
-Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom was set to $5$ for each covariate, which corresponds to $3$ knots in the splines.
-Hence, $25$ coefficients (including the intercept) were to be estimated.
-The other parameters of the diffusion were supposed to be known, as well as the measurement error.
-Underestimation of the smooth parameter $\omega$ is observed close to the shore as $\Theta$ approaches $\pm \pi$ (movement toward the shore).
-This is easily seen in figures \ref{fig: marginal_estimates_rect_CRCVM}(a)-(b) and and \ref{fig: marginal_estimates_fjords_CRCVM}(a)-(b).
-
-As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging.
-THta is why slight differences in figures ~\ref{fig: marginal_estimates_rect_CRCVM}(a)-(b).
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in the rectangular domain.
-(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: estimate_rect_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in the rectangular domain.
-The red curve represents the true value of $\omega$.
-(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
-(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-\label{fig: marginal_estimates_rect_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in Scoresby Sound fjords system.
-(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: estimate_fjords_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in Scoresby Sound fjords system.
-The red curve represents the true value of $\omega$.
-(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
-(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: marginal_estimates_fjords_CRCVM}
-
-\end{figure}
+%conditions on the shore (regularity) and time steps
+%discuss conditions on omega and tau, existence of solution e
 
@@
 
-\section{Discussion and further work}
+\section{Discussion}
+
+As discussed in \cite{ditlevsen_hypoelliptic_2019}, estimation for such hypoelliptic models is challenging. The main difficulties come from
+\begin{itemize}
+	\item degenerate noise structure that prevents the use of standard approximation schemes as Euler-Maruyama.
+	\item irregular times of observations : we only get position when the marine mammal reaches the sea surface.
+	\item GPS measurement errors, with gaussian or student distribution.
+	\item partial observations : two processes are involved in equation~\ref{eq: RACVM equation}, however only one of them is observed, that is the position.
+\end{itemize}
+Several method already exist to estimate the parameters $\tau$, $\nu$, $\mu$ and $\omega$.
+If the parameters are constant, it is possible to use a method of moments based on the distributional results about the velocity process \cite{gurarie_correlated_2017}, but it is only feasible with high frequency data and negligible measurement errors.
+Maximum likelihood with Kalman filter allows to estimate more efficiently from observations with errors \cite{johnson_continuoustime_2008}.
+If the parameters are allowed to depend on time through external covariates, then \texttt{smoothSDE} package, which is based on \cite{johnson_continuoustime_2008} but exploit the capabilities of both the \texttt{mgcv} package \cite{wood_generalized_2017} to specify the time dependency through covariates, and the \texttt{TMB} \cite{kristensen_tmb_2016} package to integrate over the random effect and optimize the likelihood, is a convenient tool for estimation.
+\\
+
+For the narwhals, a tendency to move along the shore could be considered by specifying a function $\omega$ that is non zero even when $\Theta \in \left[\frac{\pi}{2}-\varepsilon,\frac{\pi}{2}\right] \cup \left[-\frac{\pi}{2},-\frac{\pi}{2}+\varepsilon\right]$, so that when the animal would be on the verge of leaving the shore, its velocity would be tweaked to align with the shoreline.
+It is also possible  to specify $\tau$ similarly as a function of $D_{shore}$ and $\Theta$ to model specific behaviour near the shore (e.g higher persistence along the shoreline).
 
@@
 \subsubsection{Distribution of the velocity norm and phase}
+ The order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm.
+The distribution of the phase angle $\Phi$ of the velocity is derived in \cite{pawula_distribution_1982}.
+When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$.
+Hence, taking a look at the histograms of the phase angle of the data can help to assess whether a non zero mean velocity is necessary in the model or not.
 

### 2024-04-27 R/simulation_study/simulation_study_fjords.Rmd
changes in simulation study code

@@
 #seed for reproducibility 
-set.seed(1)
+set.seed(42)
 
@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs=0.005
+sigma_obs=0.01
 H_hf=array(rep(sigma_obs^2*diag(2),n_hf),dim=c(2,2,n_hf))
@@
 
-ftau_bump=function(cov_data,epsilon=pi/10,alpha=pi/5,D0=1) {
+ftau_bump=function(cov_data,epsilon=pi/10,alpha=pi/5,D0=1,tau0=0.8,tau1=2) {
     Dshore=cov_data$DistanceShore
     theta=cov_data$theta
-    tau=(1.5+0.5*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
-      0.5*(tanh((theta+(-pi/2+epsilon))/alpha)-tanh((theta+(-pi/2-epsilon))/alpha)))
+    tau=(tau0+(tau1-tau0)*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
+      (tau1-tau0)*(tanh((theta+(-pi/2+epsilon))/alpha)-tanh((theta+(-pi/2-epsilon))/alpha)))
     return (tau)
@@
 
-fomega=function(cov_data,D0=0.1,omega0=60*pi,lambda=2,kappa=0.2) {
+fomega=function(cov_data,D0=0.3,omega0=60*pi/2,lambda=2,kappa=0.2) {
     Dshore=cov_data$DistanceShore
@@
 colnames(grid)=c("DistanceShore","theta")
-true_tau=matrix(ftau_constant(grid),30,30)
+true_tau=matrix(ftau_bump(grid),30,30)
 true_omega=matrix(fomega(grid),30,30)
@@
 
+saveWidget(plot_true_omega, file ="true_smooth_omega_fjords.html")
+
 plot_true_tau
@@
 theta <- runif(n,-pi,pi);
-DistanceShore <- runif(n,0.2,2);
+DistanceShore <- runif(n,0.2,3);
 samples=data.frame(theta=theta,DistanceShore=DistanceShore)
@@
 theta_v <- seq(-pi,pi,length=30)
-Dshore_v<- seq(0.2,2,length=30)
+Dshore_v<- seq(0.2,3,length=30)
 pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
@@
                              contours = list(z = list(show = TRUE, start = min(spline_omega), end = max(spline_omega), size = 5,color="black")),
-                         x = ~Dshore,y = ~theta,z=~spline_omega,
+                         x = ~theta,y = ~Dshore,z=~spline_omega,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
@@
 plot_basis2 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z2), end = max(z2), size = 0.05,color="black")),
-                         x = ~Dshore,y = ~theta,z=~z2, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                         x = ~theta,y = ~Dshore,z=~z2, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                 layout(title=paste("Spline basis 2"),scene=list(xaxis = list(title = "Theta",showgrid = T),
@@
 plot_basis3 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z3), end = max(z3), size = 0.05,color="black")),
-                         x = ~Dshore,y = ~theta,z=~z3, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                         x = ~theta,y = ~Dshore,z=~z3, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                 layout(title=paste("Spline basis 3"),scene=list(xaxis = list(title = "Theta",showgrid = T),
@@
 plot_basis4 <- plot_ly(type = "surface",contours = list(z = list(show = TRUE, start = min(z4), end = max(z4), size = 0.05,color="black")),
-                         x = ~Dshore,y = ~theta,z=~z4, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
+                         x = ~theta,y = ~Dshore,z=~z4, colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
                 layout(title=paste("Spline basis 4"),scene=list(xaxis = list(title = "Theta",showgrid = T),
@@
                                                             zaxis = list(title = "omega")))
+
+saveWidget(plot_spline_omega, file ="spline_smooth_omega_fjords.html")
 ```
@@
 ### Generate sample
+
+```{r}
+res=sim_theta_CRCVM(ftau=ftau_bump,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[2,],times=seq(0,5*24,by=1/60),land=land,verbose=FALSE)
+
+data_sim=res$sim
+data_sim$ID=factor(rep(1,length(data_sim$y1)))
+data_shore=res$shore[,c("p1","p2")]
+data_sim=cbind(data_sim,data_shore)
+
+data_sim=data_sim[seq(1,length(data_sim$time),by=5),]
+
+plot_illust_fjords=ggplot()+geom_sf(data=land$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
+   geom_point(data=data_sim,aes(p1,p2),shape=5,size=0.1,col="red")+
+  geom_point(data=data_sim,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=data_sim,mapping=aes(y1,y2),size=0.1)+
+  geom_point(data = data_sim%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
+  xlab("x") + ylab("y")
+
+ggplotly(plot_illust_fjords)
+
+ggsave(filename="illustrative_sample_fjords.png",plot=plot_illust_fjords,width=10,height=5)
+```
 
@@
 
-ggsave(filename="samples_fjords_hf_ne_omega.png",plot=plot_samples_fjords_hf_ne)
+ggsave(filename="samples_fjords_hf_ne_omega.png",plot=plot_samples_fjords_hf_ne,width=10,height=5)
 ggplotly(plot_samples_fjords_hf_ne)

### 2024-04-27 R/simulation_study/simulation_study_rect.Rmd
changes in simulation study code

@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs=0.005
+sigma_obs=0.01
 H_hf=array(rep(sigma_obs^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
@@
 plot_true_tau <- plot_ly(type = "surface",
-                         x = ~Dshore,y = ~theta,z=~true_tau,
+                         x = ~theta,y = ~Dshore,z=~true_tau,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
@@
                            contours = list(z = list(show = TRUE, start = min(true_omega), end = max(true_omega), size = 5,color="black")),
-                         x = ~Dshore,y = ~theta,z=~true_omega,
+                         x = ~theta,y = ~Dshore,z=~true_omega,
                          colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
@@
 plot_true_omega
+
+saveWidget(plot_true_omega, file ="true_smooth_omega_rect.html")
+
 ```
@@
                                                             zaxis = list(title = "omega")))
+
+saveWidget(plot_spline_omega, file ="spline_smooth_omega_rect.html")
 ```
@@
 plot_basis2
+
+saveWidget(plot_basis2, file ="spline_basis2_omega_rect.html")
 ```
@@
 plot_basis3
+saveWidget(plot_basis3, file ="spline_basis3_omega_rect.html")
 ```
@@
 plot_basis4
+saveWidget(plot_basis4, file ="spline_basis4_omega_rect.html")
 ```
@@
 
-### Generate sample
+### Generate samples
 
-```{r parallel generate hf low errors samples,cache=TRUE}
+```{r illustrative sample for the paper}
+
+res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[1,],times=seq(0,2*24,by=1/60),land=domain,verbose=FALSE)
+
+data_sim=res$sim
+data_sim$ID=factor(rep(1,length(data_sim$y1)))
+data_shore=res$shore[,c("p1","p2")]
+data_sim=cbind(data_sim,data_shore)
+
+data_sim=data_sim[seq(1,length(data_sim$time),by=5),]
+
+plot_illust_rect=ggplot()+geom_sf(data=domain$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_sim,aes(p1,p2),shape=5,size=0.1,col="red")+
+  geom_point(data=data_sim,mapping=aes(y1,y2),size=0.1)+
+  geom_path(data=data_sim,mapping=aes(y1,y2),size=0.1)+
+  geom_point(data = data_sim%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
+  xlab("x") + ylab("y")
+
+plot_illust_rect
+
+ggsave(filename="illustrative_sample_rect.png",plot=plot_illust_rect,width=10,height=5)
+```
+
+```{r parallel generate batches of hf low errors samples,cache=TRUE}
 
@@
 
-### plot one batch
+### plot one sample and one batch
 
@@
 
-plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_path(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
+plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
+  geom_path(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
   geom_point(data = data_rect_hf_ne$BATCH1%>% filter(!duplicated(ID)),
@@
 
-ggsave(filename="samples_rect_hf_ne_omega.png",plot=plot_samples_rect_hf_ne)
+ggsave(filename="samples_rect_hf_ne_omega.png",plot=plot_samples_rect_hf_ne,width=10,height=5)
 ggplotly(plot_samples_rect_hf_ne)

### 2024-04-28 Latex/main.tex
reorganization of the paper

@@
 		\centering
-		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_rect.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_fjords.png}
+		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_rect.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_fjords.png}
 		\caption{}
@@
 
-\subsection{Estimation with tensor splines}
+\subsection{Estimation of angular velocity $\omega$ with tensor splines}
 
@@
 		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_fjords_hf_ne1.png}
+		\includegraphics[scale=0.3]{}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{}
 		\caption{}
@@
 
+\begin{equation}  \left\{
+	\begin{array}{l}
+		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
+		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
+		dX_i(t)=V_i(t)dt  \\
+		dV_i(t)=-\begin{pmatrix} 
+			\frac{1}{\tau_i} & -\omega_i \\
+			\omega_i & \frac{1}{\tau_i}
+		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
+		\log(\tau_i)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
+	\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
+	\omega_i=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{\omega,pre}^{(i)}
+	\end{array}
+	\right.
+\end{equation}
+
 \begin{equation} 
@@
 
-		\log(\tau^{(i)}_j)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
-		\log(\nu^{(i)}_j)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
-		\omega^{(i)}_j=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{\omega,pre}^{(i)}
 	\end{array}

### 2024-04-28 R/simulation_study/simulation_study_rect.Rmd
update simulation study

@@
 #number of batch
-N_BATCH=10
+N_BATCH=100
 
@@
 #degree of freedom in tensor splines
-SP_DF=5
+SP_DF=3
 
@@
 
-```{r}
+```{r plot spline omega}
 plot_spline_omega
@@
 
-```{r}
+```{r plot basis2}
 plot_basis2
@@
 
-```{r}
+```{r plot basis3}
 plot_basis3
@@
 
-```{r}
+```{r plot basis4}
 plot_basis4
@@
 
-crcvm_rect_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %do% {
+crcvm_rect_hf_ne1<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines
@@
 
-crcvm_rect_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %do% {
+crcvm_rect_hf_ne2<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines

### 2024-04-30 Latex/main.tex
reorganization of the paper and new results in simulation study

@@
 A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
-The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model.
+The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.
 
@@
 
-The diffusion models are discussed in Section 2
-Then,  Section 3 is dedicated to a comprehensive simulation study, whereas in the last section,we give an overview of the narwhal data available for the analysis and the results of the application of our models to the narwhal data.
+The diffusion models are discussed in Section 2 and simulation study results are shown.
+Then,  in Section 3 we give an overview of the narwhal data available for the analysis and the results of the application of our models to this data.
+The last section is dedicated to a discussion about these results, and potential extension of the models.
 
@@
 The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
-First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement.
+First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement and constrain the position within a polygon.
 
@@
 where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
+In this section, we obtain explicit distributional results about the process $X$.
+Though this model is considered in several papers, we found no mention of the closed form formulas derived here.
+In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
 A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
 In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
-Then the Euler-scheme is used to get the transitions of the location process.
+But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
 Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula.
-Though this model is discussed in \cite{gurarie_correlated_2017}, we found no mention of these explicit formulas.
 \\
@@
 \end{equation}
-Here, the covariance matrix $Q(\Delta)$ has an explicit form.
+Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :
 
@@
 Both components of the location and velocity processes are independent.
-and one can check that for $\omega=0$, the formulas match the ones of the  CTCRW model \cite{johnson_continuoustime_2008}.
-Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase.
-Derivation of the formulas given in this paragraph is detailed in appendix~\ref{section: detailed RACVM computations}.\\
+and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
+Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm.
+Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\
 
@@
 \subsubsection{State-space model}
-In most applications, it is needed to consider measurement errors through a state-space model.
-We suppose that positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$.
+\label{section: state space model}
+In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
+We suppose here that $n$ positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$.
 $Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$.
+The statistical model is the following:
 
@@
 
-Here, we consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
-We make the approximation that these parameters are constant on each time step $[t_j,t_{j+1}]$ .
-Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $V_j=V(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
+We consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time and make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that these parameters are constant on each time step $[t_j,t_{j+1}]$, equal to their value at time $t_j$.
+Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $X_j=X(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
 	\frac{1}{\tau_j} & -\omega_j \\
@@
 \begin{equation}
-	\forall j \in \{1, \cdots, n\} \quad 
+	\forall j \in \{1, \cdots, n\}, \quad 
 	\left\{ \begin{array}{l}
@@
 
-One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model \cite{johnson_continuoustime_2008}.
+Again, one can check that for $\omega_j=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
 This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter.
-This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} the special case $\omega=0$.
+This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$.
 Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.
@@
 \end{pmatrix}$
-where $\omega(t)=f_{\omega}(D_{shore}(t),\Theta(t))$, $D_{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
+where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
 $\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
@@
 For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
-\[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
+\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
 In this case, the formulas of section~\ref{section: RACVM} are still valid.
@@
 We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
-An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation.
-For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$.
-We hint that the condition should also include the regularity of the domain, but this will be a subject for further work.
-Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D_{shore}$ approached $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
-The closer we are to the shore, the faster the velocity should rotate. 
+An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation.
+It should always remain significantly lower than $D^{shore}$ for the process to be constrained.
+Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
+The closer we are to the shore, the faster the velocity should rotate,
 whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
@@
 A weight based on the distance to the shore can be defined as follows :
-\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
-
-Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :
+\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
+
+Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :
 
 \begin{equation}
-f_{\omega}(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
+f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D^{shore})
 \label{eq: smooth omega}
@@
 \\
-Figure~\ref{fig: trueomega} shows a smooth function obtained from~\ref{eq: smooth omega} for the parameter $\omega$.
-Figure~\ref{fig: splineomega} shows three spline basis functions and a smooth function $f_{\omega}$ obtained as a combination of these basis functions.
+Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
+In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
+A sample of a CRCVM with smooth function $f_{\omega}$ defined as y equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
+The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
 
@@
 		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-			(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
+		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
 		\label{fig: trueomega}
@@
 \begin{figure}[H]
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
-		\caption{}
-	\end{subfigure}
-	\caption{}
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
+		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
+		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
 	\label{fig: splineomega}
@@
 		\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_rect.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_fjords.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Samples of the CRCVM in different domains.
-x and y axis are in metres and initial positions are indicated by a red cross and red point represent the nearest points on the boundary of the domain (a) One sample three day sample with time step $h=1$ min in a rectangular domain (b) One sample three day sample with time step $h=1$ min in Scoresby Sound fjords system in South-East Greenland}
-	\label{fig: samples_rect_CRCVM}
-\end{figure}
-
-\section{Simulation study}
+		\includegraphics[scale=0.7]{images/simulation study/illustrative_sample_fjords.png}
+	\caption{Sample of a CRCVM.
+x and y axis are in metres and initial positions are indicated by a red cross and red points represent the nearest points on the boundary of the domain.
+One  five day sample with time step $h=5$ min within the spatial constraints of Scoresby Sound fjords system in South-East Greenland.
+Parameters are $\nu=4$ km/h and $\tau=1$ h.
+}
+	\label{fig: samples_fjords_CRCVM}
+\end{figure}
+
+\subsection{Simulation study}
 
@@
 
-\subsection{Simulation of constrained motion}
-In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.\\
-
-$3$ days long trajectories were sampled in a rectangular domain with sides of length $8$ km from the CRCVM with the smooth function for $\omega$ displayed in figure~\ref{fig: trueomega}(a) and other parameters kept constant ($\nu = 4$ km/h, $\tau=1$ h).
+\subsubsection{Simulation of constrained motion}
+\label{section: simulation constrained motion}
+In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\
+
+100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM.
+Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case.
+Denoting $n$ the number of observations for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:
+
+\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
+	\begin{array}{l}
+		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
+		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
+		dX_i(t)=V_i(t) dt \quad \\
+		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
+		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
+		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
+		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
+		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
+		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+	\end{array}
+	\right.
+	\label{eq: rect simulation parameters}
+\end{equation}
+Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
+1.8$ construction of tensor splines basis function.
+We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h.
 For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
-Time steps were chosen equal to $1$ min for high frequency and subsampled to $5$ minutes for low frequency.
-Gaussian measurement errors with low standard deviation $\sigma_{obs}=3$ m and large standard deviations $\sigma_{obs}=30$ m were added to provide the actual observations.
-Of these $100$ trajectories, none reach the boundary for $\sigma_{obs}=3$ m and only one for $\sigma_{obs}=30$m.
-However, too small values of $D_0$ and $\omega_0$, as well as too large measurement errors may produce samples that reach the boundary of the domain.
-Figure~\ref{fig: samples_rect_CRCVM} shows some of the 100 samples obtained in the rectangular domain.
-\\
-
-10 trajectories were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints.
-The parameter $\omega$ was defined as the smooth function displayed in~\ref{fig: trueomega}(b).
+Of the $600$ simulated trajectories, none reach the boundary.\\
+
+100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
+The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain.
 Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
-Time steps were also kept constant, equal to $1$ min for high frequency and subsampled to $5$ min for low frequency.
-None of the 10 trajectories reached land.
+Time steps were also kept constant, equal to $1$ min.
+None of the $600$ trajectories reached land.
 Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
@@
 
-\subsection{Estimation of angular velocity $\omega$ with tensor splines}
-
-The samples from previous section were used to estimate the function $\omega$, both in the rectangular domain and in Scoresby Sound fjords system.
-Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom was set to $5$ for each covariate, which corresponds to $3$ knots in the splines.
-Hence, $25$ coefficients (including the intercept) were to be estimated.
-The other parameters of the diffusion were supposed to be known, as well as the measurement error.
-Underestimation of the smooth parameter $\omega$ is observed close to the shore as $\Theta$ approaches $\pm \pi$ (movement toward the shore).
-This is easily seen in figures \ref{fig: marginal_estimates_rect_CRCVM}(a)-(b) and and \ref{fig: marginal_estimates_fjords_CRCVM}(a)-(b).
+\subsubsection{Parameters estimation with \texttt{smoothSDE}}
+
+Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
+The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
+Laplace's approximation of the integral of this complete likelihood over the random effects is computed using $\texttt{TMB}$ \cite{kristensen_tmb_2016} and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method.
+We refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the regularization terms based on a wiggliness measure in the mixed effect models involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
+\\
+
+$100$ batches of noisy samples obtained as described in section~\ref{section: simulation constrained motion} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
+We thus obtain $100$ parameters estimate in the two cases.
+The covariates $D^{shore}$ and $\Theta$ are computed from the observed positions and the smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
+The marginal spline degree of freedom was set to $3$ for each covariate, which corresponds to one single knot in each dimension.
+Hence, $9$ coefficients (including the intercept) are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
+Measurement error $\sigma_{obs}$ is supposed to be known.
+We fixed initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h,$\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories for different values of the distance to the shore in the rectangular and the fjords domain.
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q4_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q2_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+\begin{subfigure}{0.48\textwidth}
+	\centering
+	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
+	\caption{}
+\end{subfigure}
+\begin{subfigure}{0.48\textwidth}
+	\centering
+	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
+	\caption{}
+\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only,.
+The red curve represents the true value of $\omega$.
+(a)-(b) Estimation from samples in the rectangular domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+(c)-(d) Estimation from samples in the fjords domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+}
+\label{fig: marginal_estimates_CRCVM}
+\end{figure}
+
+For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$ and compute the root mean squared error and the absolute biais of the estimator.
+The results a summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
+
+\begin{table}[H]
+		\centering
+	\begin{tabular}{|c|c|c|c|}
+		\hline
+		Coefficient  & True value & Biais & Root mean squared error  \\
+		\hline
+		$\log(\tau_0)$  & 0& 0.24 &  \\
+		\hline
+		$\log(\nu_0)$  & 1.38 & 0.03 &  \\
+		\hline
+		$\omega_0$  &  -0.19  & 0.80 &  \\
+		\hline
+		$\omega_1$  & 16.68 & 11.54 &  \\
+		\hline
+		$\omega_2$  & -3.58 & 2.58 &  \\
+		\hline
+		$\omega_3$  & 13.07 &  8.74 &  \\
+		\hline
+		$\omega_4$  & 42.11 & 26.36 &  \\
+		\hline
+		$\omega_5$  & 13.89 & 8.44 &  \\
+		\hline
+		$\omega_6$  &  61.53 & 51.45 &  \\
+		\hline
+		$\omega_7$  & 10.15 & 2.54 &  \\
+		\hline
+		$\omega_8$  &  12.31 & 4.22 &  \\
+		\hline
+		$\sigma_{\tau}$ & 0.1 & & \\
+		\hline
+		$\sigma_{\nu}$ & 0.3 & & \\
+		\hline
+	\end{tabular}
+	\caption{Measures of estimation performance in the rectangular domain}
+	\label{table: estimation performance rect}
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|}
+		\hline
+		Coefficient  & True value & Biais & Root mean squared error  \\
+		\hline
+		$\log(\tau_0)$  &  0 &  &  \\
+		\hline
+		$\log(\nu_0)$  &  1.38 &  &  \\
+		\hline
+		$\omega_0$  & -0.34 &  &  \\
+		\hline
+		$\omega_1$  & 14.17  &  &  \\
+		\hline
+		$\omega_2$  & -2.57 &  &  \\
+		\hline
+		$\omega_3$  & 11.14 &  &  \\
+		\hline
+		$\omega_4$  & 36.48  &  &  \\
+		\hline
+		$\omega_5$  & 11.73 &  &  \\
+		\hline
+		$\omega_6$  & 53.69 &  &  \\
+		\hline
+		$\omega_7$  & 9.23 &  &  \\
+		\hline
+		$\omega_8$  & 10.19 &  &  \\
+		\hline
+		$\sigma_{\tau}$ & 0.1 & & \\
+		\hline
+		$\sigma_{\nu}$ & 0.3 & & \\
+		\hline
+	\end{tabular}
+	\caption{Measures of estimation performance in the Scoresby Sound domain}
+	\label{table: estimation performance fjords}
+\end{table}
 
 As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging.
-THta is why slight differences in figures ~\ref{fig: marginal_estimates_rect_CRCVM}(a)-(b).
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in the rectangular domain.
-(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: estimate_rect_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in the rectangular domain.
-The red curve represents the true value of $\omega$.
-(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
-(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-\label{fig: marginal_estimates_rect_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in Scoresby Sound fjords system.
-(a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: estimate_fjords_CRCVM}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_close.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/crcvm_fjords_hf_ne1_omega_far.png}
-		\caption{}
-	\end{subfigure}
-
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in Scoresby Sound fjords system.
-The red curve represents the true value of $\omega$.
-(a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.
-(c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
-	\label{fig: marginal_estimates_fjords_CRCVM}
-
-\end{figure}
 
@@
 
+Here, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
+
 \subsection{Description of the movement data}
 
-The dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
+The dataset analysed in this section has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
 Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
@@
 
+\begin{comment}
 \begin{figure}[H]
@@
 \end{figure}
+\end{comment}
 
@@
 \end{figure}
+
+\begin{comment}
 \begin{figure}[H]
@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{images/data_exploration/realExp_through_time}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.42]{images/data_exploration/realExpShip_through_time.png}
 		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.42]{images/data_exploration/realExpShore_through_time.png}
+		\caption{}
+	\end{subfigure}
+	\caption{(a) Exposure to the ship (in $km^{-1}$) over time for each narwhal. (b) Exposure to the shore (in $km^{-1}$) over time for each narwhal.}  
 	\label{fig: realexpthroughtime}
@@
 This is taken into account by using a mixed effect model for the RCVM parameters.
-For each narwhal $i \in \{1, \cdots, N\}$, write $t_{exp}^{(i)}$ the first time when it gets in line of sight with the ship (start of exposure).
-Then for $t \leq t_{exp}^{(i)}$, that is before exposure to the ship, the narwhals motion processes $V^{(i)}(t)=\begin{pmatrix} V^{(i)}_1(t) & V^{(i)}_2(t) \end{pmatrix}^\top$ and $X^{(i)}(t)=\begin{pmatrix} X^{(i)}_1(t) & X^{(i)}_2(t) \end{pmatrix}^\top$ are supposed to solve equation~\ref{eq: CRCVM equation}.
+For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations $Y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$ and the true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
 Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters.
@@
 
-Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau^{(i)}$, $\nu^{(i)}$  and $\omega^{(i)}$ the parameters of the diffusion for this narwhal $i$ and  $\tau^{(i)}_j$, $\nu_j^{(i)}$, $\omega_j^{(i)}$ their values at time $t_j^{(i)}$ for $j \in \{1, \cdots,n_{i,pre}\}$,  the mixed effect model is written
-
-\begin{equation}  \left\{
+Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau_i$, $\nu_i$  and $\omega_i$ the parameters of the diffusion for the narwhal $i$ and  $\tau_{ij}$, $\nu_{ij}$, $\omega_{ij}$ their values at time $t_{ij}$ for $j \in \{1, \cdots,n_{i,pre}\}$,  the mixed effect model is written
+
+\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
 	\begin{array}{l}
@@
 		dV_i(t)=-\begin{pmatrix} 
-			\frac{1}{\tau_i} & -\omega_i \\
-			\omega_i & \frac{1}{\tau_i}
+			\frac{1}{\tau_i} & -\omega_i(t) \\
+			\omega_i(t) & \frac{1}{\tau_i}
 		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
@@
 	\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
-	\omega_i=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E_{shore}^{(i)}(t_j^{(i)}),\Theta^{(i)}(t_j^{(i)})) + b_{\omega,pre}^{(i)}
+	\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t))\\
+	f_{\omega}(D^{shore},\Theta)=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E^{shore},\Theta) + b_{\omega,pre}^{(i)}, \quad (D^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
 	\end{array}
 	\right.
-\end{equation}
-
-\begin{equation} 
-	\forall j \in \{1,\cdots, n_{i,pre}\},
-	\left\{
-	\begin{array}{l}
-
-	\end{array}
-	\right.
-	\label{eq: approximate baseline model}
+	\label{eq: exact statistical model}
 \end{equation}
@@
 Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations are discussed in section~\ref{section: inference}.\\
-
-\subsubsection{Maximum likelihood estimation}
-The formulation of a state space model now allows for maximum likelihood estimation using Kalman filter.
-We detail the maximum likelihood estimation for the response model in section~\ref{subsection: ship exposure effect}. \\
-
-The spline coefficients in equation~\ref{eq: response model} will be treated as random effects to include a penalization in the log-likelihood. 
-Introduce smoothing parameters $\lambda_{\tau}$,$\lambda_{\nu}$, $\lambda_{\omega,E}$ and $\lambda_{\omega,\Theta}$.
-The vectors of spline regression coefficients $\log(\tau_{post})$, $\log(\nu_{post})$, $\omega_{post}$,  conditioned on the smoothing parameter are supposed to follow independent multivariate normal distributions with means zero and respective precision matrices $\lambda_{\tau}S_{\tau}$, $\lambda_{\nu}S_{\nu}$ and $\lambda_{\omega,E}S_{\omega,E}+\lambda_{\omega,\Theta}S_{\omega,\Theta}$.
-$S_{\tau}$, $S_{\nu}$, $S_{\omega,E}$ and $S_{\omega,\Theta}$ are penalization matrices of known coefficients such that
-$\log(\tau_{post})^\top S_{\tau}\log(\tau_{post})$, $\log(\nu_{post})^\top S_{\nu}\log(\nu_{post})$, and $\omega_{post}^\top (S_{\omega,E}+S_{\omega,\Theta})\omega_{post}$ measure the wiggliness of the fitted splines \cite{michelot_varying-coefficient_2021,wood_generalized_2017}.
-Thus, the chosen prior favours smooth functions over wiggly ones.
-Now, the mixed effect model equation can be rewritten as $\rho_{post}=X_{post}\beta_{post}+Z_{post} b_{post}$ where the fixed effect coefficients are changed to $\beta_{post}=\begin{pmatrix} \log(\tau_{0,post}) & \log(\nu_{0,post}) & \omega_{0,post} \end{pmatrix} $ and the random effects to
-\[b_{post}=\begin{pmatrix} \log(\tau_{post}) & b_{\tau,post} & \log(\nu_{post}) & b_{\nu,post} & \omega_{post} & b_{\omega,post}\end{pmatrix}\]
-The actual random effects $b_{\tau,post}$, $b_{\nu,post}$,$b_{\omega,post}$ also enter the same framework with a "penalization" matrix which is identity and smoothing parameters that are the inverse of the variances to be estimated.
-All these smoothing parameters are gathered in a vector $\log(\lambda_{post})$ that is estimated from the data.
-\\
-
-Denote $y^{(i)}$,  $i \in \{1, \cdots, 6\}$, the observed track for the $i$-th narwhal after exposure, which consists in $n_i$ GPS positions.
-
-The complete likelihood of the data and the random effects  is
-
-\begin{align*}
-	L_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs}))&= 
-	\prod_{i=1}^{6} p\left(\pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post} \vert \log(\lambda_{post})\right) \\
-	&\times p\left(y^{(i)} \vert \pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs})\right)
-	\label{eq: random effects likelihood}
-\end{align*}
-
-The terms $p\left(b^{(i)}_{post},  
-\log(\tau_{post}),\log(\nu_{post}),\omega_{post}\vert \log(\lambda_{post})\right)$ can be computed since $b^{(i)}_{post}$, $\log(\tau_{post})$ and  $\log(\nu_{post})$ are independent and their distributions are known multivariate gaussians.\\
-
-For each individual $i \in \{1,\cdots,6\}$, the distribution of the hidden state $\alpha^{(i)}_j=\begin{pmatrix} x^{(i)}_{j,1} & x^{(i)}_{j,2} & v^{(i)}_{j,1} & v^{(i)}_{j,2}\end{pmatrix}^\top$ conditioned on previous observations and random effects, i.e $\alpha_j^{(i)} \vert (y^{(i)}_{1:j-1},b_{post}^{(i)},\log(\tau_{post}),\log(\nu_{post}),\omega_{post})$, is $\mathcal{N}(\alpha_j^{(i)-}, R_j^{(i)-})$ where the matrices $\alpha_j^{(i)-}$ and $R_j^{(i)-}$ are computed by the Kalman filter with parameter values according to the conditioned values of the random effects and the equation~\ref{eq: response model}.
-Hence
-\begin{equation} y_j^{(i)} \vert (y_{1:j-1}^{(i)}, b_{pre}^{(i)}) \sim \mathcal{N} \left( Z \alpha_j^{(i)-},Z R_j^{(i)-} Z^\top+\sigma_{obs}^2 I_2 \right)
-	\label{eq: ditribution of observations conditioned on random effects}
-\end{equation}
-and the density $p\left(y^{(i)} \vert \pmb{b^{(i)}_{post}},\log(\tau_{post}),\log(\nu_{post}),\omega_{post},\beta_{post},\log(\lambda),\log(\sigma_{obs})\right)$ is deduced from this result.
-
-Let $\mathcal{L}_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)= - \log( L_c(\pmb{b_{post}},\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs})))$ be the joint negative log-likelihood of the data and the random effects.
-Then the likelihood of the data is the integral 
-\[
-L(\beta_{post},\log(\lambda_{post}),\log(\sigma_{obs});y)=\int_{\mathbb{R}^{d}} \exp(-\mathcal{L}_c(\pmb{b_{post}},\theta_{pre},\log(\lambda_{pre}),\log(\sigma_{obs});y) )d\pmb{b_{post}}
-\]
-where in this specific case, $d=3N+q_{\tau}+q_{\nu}+L$.
-
-The package \texttt{smooothSDE} relies on the package \texttt{TMB} to optimize this likelihood and get estimations of the coefficients.
-Laplace's approximation consists in finding a minimum of the function $\mathcal{L}_c$ with respect to $b_{post}$ and then approximating the integral over this parameter by the integral of a gaussian curve.
-This approximation is then optimized over $\beta_{post}$, $\log(\lambda_{post})$ and $\log(\sigma_{obs})$ using gradient methods and auto-differentiation \cite{kristensen_tmb_2016}.
-More precisely, optimization is  performed by using the usual \texttt{optim} R function, and plugging in the gradient function calculated by \texttt{TMB} via automatic differentiation.
-The BFGS method (Broyden–Fletcher–Goldfarb–Shanno algorithm) is generally used for non linear unconstrained problems.
-If we set constraints on the SDE parameter (maximum values of $\tau$ and $\nu$) based on biological expertise for instance, then Limited Memory BFGS with box constraints method should be preferred.
-In this study, BFGS method was used.\\
-Results of the optimization are estimated parameters intercepts $\hat{\beta}_{post}$, estimated smooothing parameters $\log(\hat{\lambda}_{post})$, estimated modes $\hat{b}_{post}^{(i)}$ of the random effects, and possibly estimated log of the standard deviation $\log(\hat{\sigma}_{obs})$ for the measurement error.
-Uncertainties on the estimates are derived from the inverse of the hessian matrix of the negative log-likelihood at estimated values and the delta method \cite{kristensen_tmb_2016}.
-
-%Talk about the consistency and asymptotic properties of such estimators
-
-%conditions on the shore (regularity) and time steps
-%discuss conditions on omega and tau, existence of solution e
 

### 2024-04-30 R/simulation_study/simulation_study_fjords.Rmd
add calculation of biais and rmse in simulation_study_rect.Rmd

@@
 #number of batch
-N_BATCH=10
+N_BATCH=1
 
@@
 #degree of freedom in tensor splines
-SP_DF=5
+SP_DF=3
 ```
@@
 
+```{r}
+sp_coeff_Dshore
+```
+
+```{r}
+sp_coeff_ExpShore
+```
+
 ### Define spline smooth functions for omega based on the approximation
@@
 
-```{r}
-res=sim_theta_CRCVM(ftau=ftau_bump,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[2,],times=seq(0,5*24,by=1/60),land=land,verbose=FALSE)
+```{r illustrative sample}
+res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[2,],times=seq(0,5*24,by=1/60),land=land,verbose=FALSE)
 
@@
 
-data_fjords_hf_ne <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+data_fjords_hf_ne <- foreach(k=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
   data=foreach (i=1:N_ID,.combine='rbind') %do% {
@@
 count=0
-for (k in 1:length(names(data_rect_hf_ne))) {
+for (k in 1:length(names(data_fjords_hf_ne))) {
 
@@
   geom_path(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_point(data = data_fjords_hf_ne$BATCH2%>% filter(!duplicated(ID)),
+  geom_point(data = data_fjords_hf_ne$BATCH1%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
@@
 
-crcvm_fjords_hf_ne1<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
+crcvm_fjords_hf_ne1<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
   #estimation with tensor splines

### 2024-04-30 R/simulation_study/simulation_study_rect.Rmd
add calculation of biais and rmse in simulation_study_rect.Rmd

@@
 #number of batch
-N_BATCH=100
+N_BATCH=10
 
@@
 #no measurement error (we cannot put 0 because only the log of this quantity is used)
-sigma_obs=0.01
+sigma_obs=0.005
 H_hf=array(rep(sigma_obs^2*diag(2),n_hf*N_ID),dim=c(2,2,n_hf*N_ID))
@@
 sp_coeff_ExpShore=m2$coefficients
+```
+
+```{r}
+sp_coeff_Dshore
+```
+
+```{r}
+sp_coeff_ExpShore
 ```
@@
 plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_rect_hf_ne$BATCH1,aes(p1,p2),shape=5,size=0.1,col="red")+
   geom_point(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
@@
 
-ggsave(filename="samples_rect_hf_ne_omega.png",plot=plot_samples_rect_hf_ne,width=10,height=5)
+ggsave(filename="samples_rect_hf_ne.png",plot=plot_samples_rect_hf_ne,width=10,height=5)
 ggplotly(plot_samples_rect_hf_ne)
@@
 
-#### Violin plot of the estimated spline coefficients for one batch
+#### Violin plot of the estimated coefficients for one batch
 
@@
   m<-crcvm_rect_hf_ne1[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
+  coeffs=rbind(m$coeff_re(),m$coeff_fe())
   coeff_names=rownames(coeffs)
@@
 
+```
+
+### Biais and RMSE
+
+```{r compute biais and rmse}
+biais=c()
+rmse=c()
+for (string in names(sp_coeff_ExpShore)) {
+  name=paste("omega.",string,sep="")
+  true_value=sp_coeff_ExpShore[string]
+  estimates=coeffs_df[coeffs_df$coeff_name==name,"estimate"]
+  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value))/true_value)
+  rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_value)^2)))
+}
+
+names(biais)=paste("omega.",names(sp_coeff_ExpShore),sep="")
+names(rmse)=paste("omega.",names(sp_coeff_ExpShore),sep="")
+```
+
+```{r print biais}
+biais
+```
+
+```{r print rmse}
+rmse
 ```
@@
 xmax=list("theta"=pi,"DistanceShore"=2)
+xlabel=list("DistanceShore"="Distance to shore","theta"="Theta")
+link=list("DistanceShore"=identity,"theta"=identity)
+true_smooths=list("omega"=fomega_splines)
 res=crcvm_rect_hf_ne2$BATCH1$get_all_plots(baseline=NULL,model_name="crcvm_rect_hf_ne2",xmin=xmin,xmax=xmax,
@@
 for (k in 1:N_BATCH) {
+
+  #fitted model
   m<-crcvm_rect_hf_ne2[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
+
+  #re and fe coeffs
+  coeffs=rbind(m$coeff_re(),m$coeff_fe())
+
+  #names and values of coeffs
   coeff_names=rownames(coeffs)
   coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+
+  #keep coeffs that do not contain ID
+  sub_index=grep("ID",coeff_names,invert=TRUE)
+
+  #standard deviation of random effects
+  sdev=m$sdev()
+  sdev_names=rownames(sdev)[1:2]
+  sdev_values=as.numeric(sdev)[1:2]
+
+  #bind dataframe
+  sub_coeffs_df=data.frame("estimate"=c(coeff_values[sub_index],sdev_values),"coeff_name"=factor(c(coeff_names[sub_index],sdev_names)))
   coeffs_df=rbind(coeffs_df,sub_coeffs_df)
@@
 p=ggplotly(p)
-saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.html")
+saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne1.html")
 
 p
+
+```
+
+### Biais and RMSE
+
+```{r compute biais and rmse DistanceShore splines}
+biais=c()
+rmse=c()
+
+#biais and rmse for omega coefficients
+for (string in names(sp_coeff_Dshore)) {
+  name=paste("omega.",string,sep="")
+  true_value=sp_coeff_Dshore[string]
+  estimates=coeffs_df[coeffs_df$coeff_name==name,"estimate"]
+  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value)))
+  rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_value)^2)))
+}
+
+#add biais and rmse for tau
+true_log_tau=0
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_tau)^2)))
+
+#add biais and rmse for nu
+true_log_nu=log(4)
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_nu)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_tau=0.1
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_tau)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_nu=0.3
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_nu)^2)))
+
+names(biais)=c(paste("omega.",names(sp_coeff_Dshore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+names(rmse)=c(paste("omega.",names(sp_coeff_Dshore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+
+```
+
+```{r print biais DistanceShore splines}
+biais
+```
+
+```{r print rmse DistanceShore splines}
+rmse
 ```
@@
 
-fExpShore=function(z,v,p) {
+fDshore=function(z,v,p) {
 
@@
   Dshore=sqrt(normal[1]^2+normal[2]^2)
-  if (Dshore<0.01) {
-    return (100)
-  }
-  else {
-    return (1/Dshore)
-  }
+  return (Dshore)
 }
@@
 
-atw=list("ExpShore"=fExpShore,"AngleNormal"=fangle)
+atw=list("DistanceShore"=fDshore,"AngleNormal"=fangle)
 sim_data=crcvm_rect_hf_ne2$BATCH1$simulate(z0=x0[1,],data=crcvm_rect_hf_ne2$BATCH1$data(),atw=atw,land=domain)

### 2024-05-02 R/simulation_study/simulation_study_fjords.Rmd
improve simulation study

@@
 # SET UP
-## Import libraries and data
+
+## Import libraries
 
@@
 set.seed(42)
+
+#parallel computing
+library(foreach)
+library(doParallel)
 
@@
 #number of batch
-N_BATCH=1
+N_BATCH=5
 
@@
 
-# ESTIMATION
+## Estimate with ExpShore covariate and tensor splines te
 
-## Estimate from high frequency data without error : ExpShore and tensor splines ti
-
-### Use ExpShore covariate and tensor splines te
-
-#### Fit sde
+### Fit sde
 
@@
 
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_fjords_hf_ne1<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
+crcvm_fjords_hf_ne1<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines
@@
 
-#stop cluster
-stopCluster(cl)
-
 names(crcvm_fjords_hf_ne1)=paste("BATCH",1:N_BATCH,sep="")
@@
 
-#### Plot estimated smooth parameter omega in 3D for one batch
+### Plot estimated smooth parameter omega in 3D for one batch
 
@@
 
-#### Violin plot of the estimated spline coefficients for one batch
+### Violin plot of the estimated spline coefficients for one batch
 
-```{r}
+```{r violin plots spline coeff te ExpShore for one batch of low error samples}
 
@@
 
-#### Violin plot of the estimated spline coefficients for all batches
+### Violin plot of the estimated spline coefficients for all batches
 
@@
 for (k in 1:N_BATCH) {
+
+  #fitted model
   m<-crcvm_fjords_hf_ne1[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
+
+  #re and fe coeffs
+  coeffs=rbind(m$coeff_re(),m$coeff_fe())
+
+  #names and values of coeffs
   coeff_names=rownames(coeffs)
   coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+
+  #keep coeffs that do not contain ID
+  sub_index=grep("ID",coeff_names,invert=TRUE)
+
+  #standard deviation of random effects
+  sdev=m$sdev()
+  sdev_names=rownames(sdev)[1:2]
+  sdev_values=as.numeric(sdev)[1:2]
+
+  #bind dataframe
+  sub_coeffs_df=data.frame("estimate"=c(coeff_values[sub_index],sdev_values),"coeff_name"=factor(c(coeff_names[sub_index],sdev_names)))
   coeffs_df=rbind(coeffs_df,sub_coeffs_df)
@@
 p=ggplotly(p)
-saveWidget(p, file ="crcvm_fjords_hf_ne1/all_batches_coeffs_crcvm_fjords_hf_ne1.png")
+saveWidget(p, file ="crcvm_fjords_hf_ne1/all_batches_coeffs_crcvm_fjords_hf_ne1.html")
+
+p
+```
+
+### Biais and RMSE
+
+```{r compute biais and rmse ExpShore splines}
+biais=c()
+rmse=c()
+
+#biais and rmse for omega coefficients
+for (string in names(sp_coeff_ExpShore)) {
+  name=paste("omega.",string,sep="")
+  true_value=sp_coeff_ExpShore[string]
+  estimates=coeffs_df[coeffs_df$coeff_name==name,"estimate"]
+  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value)))
+  rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_value)^2)))
+}
+
+#add biais and rmse for tau
+true_log_tau=0
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_tau)^2)))
+
+#add biais and rmse for nu
+true_log_nu=log(4)
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_nu)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_tau=0.1
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_tau)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_nu=0.3
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_nu)^2)))
+
+names(biais)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+names(rmse)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
 
@@
 
-#### Simulate from fitted model
+```{r print biais ExpShore splines}
+biais
+```
+
+```{r print rmse ExpShore splines}
+rmse
+```
+
+### Simulate from fitted model
 
@@
 
-#### Plot simulation from fitted model
+### Plot simulation from fitted model
 
@@
 
-###  UseDistanceShore covariate and tensor splines ti
+##  Estimate with DistanceShore covariate and tensor splines te
 
-#### Fit sde
+### Fit sde
 
@@
 
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-crcvm_fjords_hf_ne2<- foreach(i=1:N_BATCH,.packages=c("smoothSDE")) %dopar% {
+crcvm_fjords_hf_ne2<- foreach(k=1:N_BATCH,.packages=c("smoothSDE")) %do% {
   #estimation with tensor splines
@@
 
-#stop cluster
-stopCluster(cl)
-
 names(crcvm_fjords_hf_ne2)=paste("BATCH",1:N_BATCH,sep="")
@@
 
-#### Plots estimated smooth in 3D for one batch
+### Plots estimated smooth in 3D for one batch
 
-```{r}
+```{r get plots from hf e estimates with DistanceShore}
 xmin=list("theta"=-pi,"DistanceShore"=0.3)
@@
 
-#### Violin plot of the estimated spline coefficients for one batch
+### Violin plot of the estimated spline coefficients for one batch
 
-```{r}
+```{r violin plots spline coeff te DistanceShore for one batch of low error samples}
 
@@
 
-#### Simulate from fitted model
+### Violin plot of the estimated spline coefficients for all batches
+
+```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
+
+n_coeffs=length(sp_coeff_DistanceShore)
+coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
+
+for (k in 1:N_BATCH) {
+
+  #fitted model
+  m<-crcvm_fjords_hf_ne2[[paste("BATCH",k,sep="")]]
+
+  #re and fe coeffs
+  coeffs=rbind(m$coeff_re(),m$coeff_fe())
+
+  #names and values of coeffs
+  coeff_names=rownames(coeffs)
+  coeff_values=as.numeric(coeffs)
+
+  #keep coeffs that do not contain ID
+  sub_index=grep("ID",coeff_names,invert=TRUE)
+
+  #standard deviation of random effects
+  sdev=m$sdev()
+  sdev_names=rownames(sdev)[1:2]
+  sdev_values=as.numeric(sdev)[1:2]
+
+  #bind dataframe
+  sub_coeffs_df=data.frame("estimate"=c(coeff_values[sub_index],sdev_values),"coeff_name"=factor(c(coeff_names[sub_index],sdev_names)))
+  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
+}
+
+#remove initialization row
+coeffs_df=coeffs_df[-1,]
+
+p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
+                xlab(" ")+labs(fill = "Estimated coefficients")+
+                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
+
+p=ggplotly(p)
+saveWidget(p, file ="crcvm_fjords_hf_ne2/all_batches_coeffs_crcvm_fjords_hf_ne2.html")
+
+p
+```
+
+### Biais and RMSE
+
+```{r compute biais and rmse DistanceShore splines}
+biais=c()
+rmse=c()
+
+#biais and rmse for omega coefficients
+for (string in names(sp_coeff_ExpShore)) {
+  name=paste("omega.",string,sep="")
+  true_value=sp_coeff_ExpShore[string]
+  estimates=coeffs_df[coeffs_df$coeff_name==name,"estimate"]
+  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value)))
+  rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_value)^2)))
+}
+
+#add biais and rmse for tau
+true_log_tau=0
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_tau)^2)))
+
+#add biais and rmse for nu
+true_log_nu=log(4)
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_nu)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_tau=0.1
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_tau)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_nu=0.3
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_nu)^2)))
+
+names(biais)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+names(rmse)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+
+```
+
+```{r print biais DistanceShore splines}
+biais
+```
+
+```{r print rmse DistanceShore splines}
+rmse
+```
+
+### Simulate from fitted model
 
@@
 ```
-
-#### Violin plot of the estimated spline coefficients for all batches
-
-```{r violin plots spline coeff te DistanceShore all batches with low measurement error}
-
-n_coeffs=length(sp_coeff_Dshore)
-coeffs_df=data.frame("estimate"=NA,"coeff_name"=NA)
-
-for (k in 1:N_BATCH) {
-  m<-crcvm_fjords_hf_ne2[[paste("BATCH",k,sep="")]]
-  coeffs=m$coeff_re()
-  coeff_names=rownames(coeffs)
-  coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
-  coeffs_df=rbind(coeffs_df,sub_coeffs_df)
-}
-
-#remove initialization row
-coeffs_df=coeffs_df[-1,]
-
-p=ggplot()+geom_violin(data=coeffs_df,aes(x=coeff_name,y=estimate,fill=coeff_name))+
-                xlab(" ")+labs(fill = "Estimated coefficients")+
-                theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
-                geom_point(data = true_coeff, aes(x = name, y = value), color = "red", size = 1,shape=4)
-
-p=ggplotly(p)
-saveWidget(p, file ="crcvm_fjords_hf_ne2/all_batches_coeffs_crcvm_fjords_hf_ne2.html")
-
-```

### 2024-05-02 R/simulation_study/simulation_study_rect.Rmd
improve simulation study

@@
 
-### plot one sample and one batch
+### plot one batch
 
@@
 
-## Estimate from high frequency data with low error
+## Estimate with ExpShore covariate and tensor splines te
 
-### Use ExpShore covariate and tensor splines te
-
-#### Fit sde
+### Fit sde
 
@@
 
-#### Violin plot of the estimated coefficients for one batch
+### Violin plot of the estimated spline coefficients for one batch
 
@@
 for (k in 1:N_BATCH) {
+
+  #fitted model
   m<-crcvm_rect_hf_ne1[[paste("BATCH",k,sep="")]]
+
+  #re and fe coeffs
   coeffs=rbind(m$coeff_re(),m$coeff_fe())
+
+  #names and values of coeffs
   coeff_names=rownames(coeffs)
   coeff_values=as.numeric(coeffs)
-  sub_index=grepl("omega",coeff_names)
-  sub_coeffs_df=data.frame("estimate"=coeff_values[sub_index],"coeff_name"=factor(coeff_names[sub_index]))
+
+  #keep coeffs that do not contain ID
+  sub_index=grep("ID",coeff_names,invert=TRUE)
+
+  #standard deviation of random effects
+  sdev=m$sdev()
+  sdev_names=rownames(sdev)[1:2]
+  sdev_values=as.numeric(sdev)[1:2]
+
+  #bind dataframe
+  sub_coeffs_df=data.frame("estimate"=c(coeff_values[sub_index],sdev_values),"coeff_name"=factor(c(coeff_names[sub_index],sdev_names)))
   coeffs_df=rbind(coeffs_df,sub_coeffs_df)
@@
 
-```{r compute biais and rmse}
+```{r compute biais and rmse ExpShore splines}
 biais=c()
 rmse=c()
+
+#biais and rmse for omega coefficients
 for (string in names(sp_coeff_ExpShore)) {
@@
   estimates=coeffs_df[coeffs_df$coeff_name==name,"estimate"]
-  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value))/true_value)
+  biais=c(biais,1/N_BATCH*sum(abs(estimates-true_value)))
   rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_value)^2)))
@@
 
-names(biais)=paste("omega.",names(sp_coeff_ExpShore),sep="")
-names(rmse)=paste("omega.",names(sp_coeff_ExpShore),sep="")
+#add biais and rmse for tau
+true_log_tau=0
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_tau)^2)))
+
+#add biais and rmse for nu
+true_log_nu=log(4)
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(Intercept)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_log_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_log_nu)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_tau=0.1
+estimates=coeffs_df[coeffs_df$coeff_name=="tau.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_tau)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_tau)^2)))
+
+#add biais and rmse for sdev tau
+true_sdev_nu=0.3
+estimates=coeffs_df[coeffs_df$coeff_name=="nu.(ID)","estimate"]
+biais=c(biais,1/N_BATCH*sum(abs(estimates-true_sdev_nu)))
+rmse=c(rmse,sqrt(1/N_BATCH*sum((estimates-true_sdev_nu)^2)))
+
+names(biais)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+names(rmse)=c(paste("omega.",names(sp_coeff_ExpShore),sep=""),"tau.(Intercept)","nu.(Intercept)","tau.(ID)","nu.(ID)")
+
 ```
 
-```{r print biais}
+```{r print biais ExpShore splines}
 biais
@@
 
-```{r print rmse}
+```{r print rmse ExpShore splines }
 rmse
@@
 
-#### Simulate from fitted model
+### Simulate from fitted model
 
@@
 
-###  Use DistanceShore covariate and tensor splines te
+## Estimate with ExpShore covariate and tensor splines te
 
-#### Fit sde
+### Fit sde
 
@@
 
-#### Plots estimated smooth in 3D
+### Plots estimated smooth in 3D
 
@@
 
-#### Violin plot of the estimated spline coefficients
+### Violin plot of the estimated spline coefficients
 
@@
 p=ggplotly(p)
-saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne1.html")
+saveWidget(p, file ="crcvm_rect_hf_ne2/all_batches_coeffs_crcvm_rect_hf_ne2.html")
 

### 2024-05-02 Latex/main.tex
new changes to Latex file

@@
 %%PACKAGES
+\usepackage{authblk}
 \usepackage{hyperref} 
@@
 \usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
-\usepackage[affil-it]{authblk}
 \usepackage{fancyhdr}
@@
 
-\author{Alexandre Delporte%
-	\thanks{\texttt{alexandre.delporte@univ-grenoble-alpes.fr}} }
-	\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}
-\author{Susanne Ditlevsen%
-\thanks{\texttt{susanne@math.ku.dk}}}
-\affil{Department of Statistics, University of Copenhagen, Denmark}
-
-\author{Adeline Samson%
-	\thanks{\texttt{adeline.leclercq-samson@univ-grenoble-alpes.fr}}}
-	\affil{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France}
+\author[1]{Alexandre Delporte}
+\author[1]{Adeline Samson}
+\author[2]{Susanne Ditlevsen}
+
+\affil[1]{Laboratoire Jean Kuntzmann, Université Grenoble-Alpes, France \authorcr
+	\{\tt alexandre.delporte, adeline.samson\}@univ-grenoble-alpes.fr}
+\affil[2]{Department of Statistics, University of Copenhagen, Denmark \authorcr
+	\tt  susanne@math.ku.dk}
 
@@
 \begin{itemize}
-	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of \texttt{smoothSDE} \texttt{R} package, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
+	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
 	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
@@
 
-The transition density for the velocity process is such that 
+The transition density of the velocity Ornstein-Uhlenbeck process is  
 \begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
@@
 \]
-The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
-Both components of the location and velocity processes are independent.
+The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
+The two components of the location and velocity processes are independent.
 and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
@@
 	\end{array}\right. \\
-	\label{eq: RCVM autoregressive formula}
+	\label{eq: RACVM autoregressive formula}
 \end{equation}
@@
 
-\[
+\begin{equation}
 \left\{
@@
 \end{array}
-\right.\]
+\right.
+\label{eq: RACVM state space}
+\end{equation}
 where the link matrices are 
@@
 
+\subsection{Formulation of the model}
+
 Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
-To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
-This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
-Denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
+To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore.
+To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector.
+The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
+We denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
 \begin{equation} \left\{
@@
 
-Here, $A(X(t),V(t))=\begin{pmatrix} 
+with $A(X(t),V(t))=\begin{pmatrix} 
 	\frac{1}{\tau} & -\omega(t) \\
@@
 
-Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
+Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
 For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
-\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
+\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
 In this case, the formulas of section~\ref{section: RACVM} are still valid.
-The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
+The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
 We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
@@
 whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
-Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
+Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
 A weight based on the distance to the shore can be defined as follows :
@@
 An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
-\\
 Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
 In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
-A sample of a CRCVM with smooth function $f_{\omega}$ defined as y equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
+A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
 The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
@@
 
-\subsection{Simulation study}
-
 \label{section: simulation study}
@@
 
-\subsubsection{Simulation of constrained motion}
+\subsubsection{Simulation study}
 \label{section: simulation constrained motion}
@@
 Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case.
-Denoting $n$ the number of observations for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:
+Denoting $n$ the number of observations, which is the same for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:
 
@@
 Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
-1.8$ construction of tensor splines basis function.
+1.8$ for the construction of tensor splines basis function.
 We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h.
-For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
-Of the $600$ simulated trajectories, none reach the boundary.\\
+The values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ are given in Table~\ref{table: estimation performance rect}.
+For each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
+To simulate the trajectories, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
+\begin{equation}
+	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
+\end{equation}
+  where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
+The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
+None of the $600$ simulations reached the boundary.
+\\
 
@@
 The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain.
-Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
-Time steps were also kept constant, equal to $1$ min.
+The exact values of the spline coefficients can be found in Table~\ref{table: estimation performance fjords}.
+Again, for each trajectory, initial velocity was set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
+Time steps are also kept constant, equal to $1$ min.
 None of the $600$ trajectories reached land.
@@
 For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
-Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}.
 Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-
-\subsubsection{Parameters estimation with \texttt{smoothSDE}}
 
@@
 The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
-Laplace's approximation of the integral of this complete likelihood over the random effects is computed using $\texttt{TMB}$ \cite{kristensen_tmb_2016} and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method.
-We refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the regularization terms based on a wiggliness measure in the mixed effect models involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
+Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation.
+We refer to \cite{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations.
+Similarly, we refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the mixed effect model for the parameter of the SDE and the regularization terms that are based on a wiggliness measure involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
 \\
 
-$100$ batches of noisy samples obtained as described in section~\ref{section: simulation constrained motion} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
+$100$ batches of noisy samples obtained as described in equation~\ref{eq: rect simulation parameters} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
 We thus obtain $100$ parameters estimate in the two cases.
-The covariates $D^{shore}$ and $\Theta$ are computed from the observed positions and the smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom was set to $3$ for each covariate, which corresponds to one single knot in each dimension.
-Hence, $9$ coefficients (including the intercept) are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
+The covariates $D^{shore}$ and $\Theta$ are computed from the observed noisy positions.
+It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging.
+The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
+The marginal spline degree of freedom is set to $3$ for each covariate to match the simulated data from equation~\ref{eq: rect simulation parameters}.
+This corresponds to one single knot in each dimension.
+Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
 Measurement error $\sigma_{obs}$ is supposed to be known.
-We fixed initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h,$\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories for different values of the distance to the shore in the rectangular and the fjords domain.
+We fix initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h, $\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectoriesin the rectangular and the fjords domain.
+The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
 
@@
 
-For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$ and compute the root mean squared error and the absolute biais of the estimator.
-The results a summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
+For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the absolute biais of the estimator.
+The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
 
@@
 		\hline
-		$\log(\tau_0)$  & 0& 0.24 &  \\
-		\hline
-		$\log(\nu_0)$  & 1.38 & 0.03 &  \\
-		\hline
-		$\omega_0$  &  -0.19  & 0.80 &  \\
-		\hline
-		$\omega_1$  & 16.68 & 11.54 &  \\
-		\hline
-		$\omega_2$  & -3.58 & 2.58 &  \\
-		\hline
-		$\omega_3$  & 13.07 &  8.74 &  \\
-		\hline
-		$\omega_4$  & 42.11 & 26.36 &  \\
-		\hline
-		$\omega_5$  & 13.89 & 8.44 &  \\
-		\hline
-		$\omega_6$  &  61.53 & 51.45 &  \\
-		\hline
-		$\omega_7$  & 10.15 & 2.54 &  \\
-		\hline
-		$\omega_8$  &  12.31 & 4.22 &  \\
+		$\log(\tau_0)$  & 0& 0.24 &  0.26\\
+		\hline
+		$\log(\nu_0)$  & 1.38 & 0.03 & 0.04 \\
+		\hline
+		$\omega_0$  &  -0.19  & 0.80 & 0.82 \\
+		\hline
+		$\omega_1$  & 16.68 & 11.54 & 12.32 \\
+		\hline
+		$\omega_2$  & -3.58 & 2.58 &  3.04\\
+		\hline
+		$\omega_3$  & 13.07 &  8.74 &  8.84\\
+		\hline
+		$\omega_4$  & 42.11 & 26.36 &  29.04\\
+		\hline
+		$\omega_5$  & 13.89 & 8.44 & 8.54 \\
+		\hline
+		$\omega_6$  &  61.53 & 51.45 &  51.60\\
+		\hline
+		$\omega_7$  & 10.15 & 2.54 & 3.37 \\
+		\hline
+		$\omega_8$  &  12.31 & 4.22 & 4.45 \\
 		\hline
@@
 
-As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging.
-
 \section{Application to narwhals behavioral response}
 
-Here, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
+In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
 
@@
 The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.
-Figure \ref{fig:alltimestepshisto} shows the distribution of time steps in the data.
 While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.
@@
 These periods are indicated by a categorical variable $T_{ship}$ in the dataset.
-Figure~\ref{fig: trials and intertrials distributions} shows how these periods are distributed among the $6$ narwhals that were tracked.
 In this analysis, we do not distinguish between Trial and Intertrial periods, and treat them both as exposure periods, where the degree of exposure depends on the narwhals positions relative to the ship.
-
-\begin{comment}
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.9\textwidth]{images/data_exploration/trials.png}
-		\caption{Trial and Intertrial periods for each narwhal}
-		\label{fig: trials and intertrials distributions}
-	\end{subfigure}
-	\begin{subfigure}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.9\textwidth]{images/data_exploration/all_time_steps_histo}
-		\caption{Histogram of time steps}
-		\label{fig:alltimestepshisto}
-	\end{subfigure}
-\end{figure}
-\end{comment}
 
@@
 The splitting between data before and after exposure results in $935$ measures before exposure and $3257$ measures after exposure.
-Table~\ref{table: data distribution} in Appendix shows how the data is distributed among the different narwhals.
-Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.
 
@@
 The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
-Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}.
 
@@
 Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.
-All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
 
@@
 \end{figure}
-
-\begin{table}[H]
-	\centering
-	\begin{tabular}{|c|c|p{8cm}|c|}
-		\hline
-		Covariate & Unit & Description & Domain \\
-		\hline
-		$D_{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
-		\hline
-		$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1,2\}$\\
-		\hline
-		$E_{ship}(t)=\frac{1}{D_{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
-		\hline
-		$E_{ship,I}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=1}$& $km^{-1}$ & intertrial exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
-		\hline
-		$E_{ship,T}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=2}$ & $km^{-1}$ & trial exposure level of the narwhal to the ship disturbance at time $t$. & $\R^+$\\
-		\hline
-		$D_{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
-		\hline
-		$E_{shore}(t)=\frac{1}{D_{shore}(t)} \1_{D_{shore}(t)<D_0}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
-		\hline
-		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical step vector at time $t$  & $[-\pi,\pi]$ \\
-		\hline 
-	\end{tabular}
-	\caption{Summary of the covariates}
-	\label{tab: covariates}
-\end{table}
 
@@
 Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters.
-The angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $D_{shore}$ through splines.
-Since both covariates are not on the same scale (one is in $rad$ while the other is in $km$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\
-
-Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau_i$, $\nu_i$  and $\omega_i$ the parameters of the diffusion for the narwhal $i$ and  $\tau_{ij}$, $\nu_{ij}$, $\omega_{ij}$ their values at time $t_{ij}$ for $j \in \{1, \cdots,n_{i,pre}\}$,  the mixed effect model is written
+The angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $E^{shore}$ through splines.
+Since both covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\
+
+Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau_i$, $\nu_i$  and $\omega_i$ the parameters of the diffusion for the narwhal $i$.
 
@@
 	\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
-	\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t))\\
-	f_{\omega}(D^{shore},\Theta)=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E^{shore},\Theta) + b_{\omega,pre}^{(i)}, \quad (D^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+	\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
+	f_{\omega}(E^{shore},\Theta)=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E^{shore},\Theta) + b_{\omega,pre}^{(i)}, \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
 	\end{array}
 	\right.
-	\label{eq: exact statistical model}
+	\label{eq: exact baseline model}
 \end{equation}
@@
 
-The $\psi_k$ in equation~\ref{eq: approximate baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
+The $\psi_k$ in equation~\ref{eq: exact baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
 %Add plot of the spline basis functions \\
@@
 It is needed to choose them before hand.
-One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.
-Equation \ref{eq: approximate baseline model} is based on the approximation that the covariates are constant on each time step $[t_j^{(i)},t_{j+1}^{(i)}]$, so that the parameters are piecewise constant, which allows to form a state-space model for estimation as described in section \ref{section: inference}.\\
-In matrix form, equation~\ref{eq: approximate baseline model} can be written as
-\begin{equation}
-	\rho_{pre}= X_{pre} \beta_{pre}
-	+Z_{pre}b_{pre}
-\end{equation}
-
-where $\rho_{pre}$ is the vector of the parameter values on each time step, that has size $3n=$ number of observations $\times$ number of parameters  \[\rho_{pre}=\begin{pmatrix}
-	\log(\tau^{(1)}) & \cdots & \log(\tau^{(N)}) & \log(\nu^{(1)}) &
-	\cdots & \log(\nu^{(N)}) & \omega^{(1)}  &\cdots & \omega^{(N)}
-\end{pmatrix}^\top\]
-with $	\log(\tau^{(i)})=\begin{pmatrix} \log(\tau^{(i)}_1) & \cdots \log(\tau^{(i)}_{n_i})\end{pmatrix}$ and similarly for the $\log(\nu^{(i)})$ and $\omega^{(i)}$.
-$X_{pre}$ and $Z_{pre}$ are respectively the fixed and random effects model matrices with sizes $3n \times (L+3)$ and $3n \times 3N$.
-Expressions of these matrices are given in Appendix \ref{section: precisions about statistical models}.
-
-$\beta_{pre}=\begin{pmatrix}
-	\log(\tau_{pre}) &
-	\log(\nu_{pre})  &
-	\omega_{0,pre} &
-	\omega_{1,pre} &
-	\cdots & 
-	\omega_{q_E q_{\Theta},pre}
-\end{pmatrix}^\top$ is the vector of coefficients to estimate for the pre-exposure model and $b_{pre}=\begin{pmatrix}  b_{\tau,pre}^{(1)} &\cdots& b_{\tau,pre}^{(N)} &b_{\nu,pre}^{(1)}& \cdots & b_{\nu,pre}^{(N)} & b_{\omega,pre}^{(1)} & \cdots & b_{\omega,pre}^{(N)} \end{pmatrix}$ is the vector of random effects.\\
+One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.\\
 
@@
 
-\begin{equation} 
-	\forall j \in \{1,\cdots, n_{i,post}\}, \quad 
-	\left\{\begin{array}{l}
-
+\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
+	\begin{array}{l}
+		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}\\
+		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
+		dX_i(t)=V_i(t)dt  \\
+		dV_i(t)=-\begin{pmatrix} 
+			\frac{1}{\tau_i} & -\omega_i(t) \\
+			\omega_i(t) & \frac{1}{\tau_i}
+		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
 		\log(\tau^{(i)}_j)=\log(\tau_{0,post}) +\sum_{k=1}^{q_{\tau}} \log(\tau_{k,post})\phi_{k}(E_{ship}(t))+b^{(i)}_{\tau,post} \\
 		\log(\nu^{(i)}_j)=\log(\nu_{0,post}) +  \sum_{k=1}^{q_{\nu}} \log(\nu_{k,post}) \xi_{k}(E_{ship}(t)) +b^{(i)}_{\nu,post}  \\
-		\omega^{(i)}_j=\omega_{0,post}+\sum_{k=1}^{q_E\times q_{\Theta}} \omega_{k,post} \psi_k(D_{shore}(t),\Theta(t)) + b_{\omega,post}^{(i)}
+		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
+		f_{\omega}(E^{shore},\Theta)=\omega_{0,post}+\sum_{k=1}^{L} \omega_{k,post} \psi_k(E^{shore},\Theta) + b_{\omega,post}^{(i)}, \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
 	\end{array}
 	\right.
-	\label{eq: response model}
-\end{equation}
-
-Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are unknown intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$.$q_{\tau}$ and $q_{\nu}$ are the spline degrees of freedom for the exposure to the ship covariate.
-We can write the mixed effect models in matrix form with the same notations as above.
-\begin{equation}
-	\rho_{post}
-	=X_{post} \beta_{post}
-	+Z_{post} b_{post}
-\end{equation}
-
-Estimation procedure for the vectors of coefficient $\beta_{pre}$, $\beta_{post}$ and the expected values of the random effects $b_{pre}$, $b_{post}$ given the observations are discussed in section~\ref{section: inference}.\\
-
-\section{Estimation of ship exposure effect on real narwhal data}
+	\label{eq: exact response model}
+\end{equation}
+
+Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$ and $q_{\tau}$ and $q_{\nu}$ are the spline degrees of freedom for the exposure to the ship covariate.
+
+\subsection{Estimation of ship exposure effect on real narwhal data}
 
@@
 
-\subsection{Estimation without constraints}
-
-First, a simple CVM (see equation \ref{eq: CVM equation}) was used to assess any behavioural response of the narwhals to the ship exposure.
-
-\subsubsection{Baseline estimations}
+\subsubsection{Estimation without constraints}
+
+First, a simple CVM was used to assess any behavioural response of the narwhals to the ship exposure.
+
+\paragraph{Baseline estimations}
 Due to the low variability of the estimates of $\tau$ for each individual, it was chosen to only estimate one intercept value for this parameter.
@@
 The estimates for the long term velocity $\mu$ show that all the narwhals have a tendency to move toward the south ($\mu_2<0$) with more individual variability for the east-west direction $\mu_1$.
-However, confidence intervals are large and  forcin $\mu$ to $(0,0)$ is also a reasonable choice.
+However, confidence intervals are large and  forcing $\mu$ to $(0,0)$ is also a reasonable choice.
 
@@
 
-\subsubsection{Response estimations}
+\paragraph{Response estimations}
 
@@
 
-\subsection{Estimation with constraints}
-
-\subsubsection{Baseline estimations}
+\subsubsection{Estimation with constraints}
+
+\paragraph{Baseline estimations}
 
@@
 
+\textcolor{red}{Further work: 
+\begin{itemize}
+	\item Error in \texttt{smoothSDE} approximation based on the exact state space model, which is, in the case $\omega=0$ 
+	\[
+	\left\{
+	\begin{array}{l}
+		y_{c,j}=Z\alpha_{c,j}+\varepsilon_{c,j} \\
+		\alpha_{c,j+1}=T_j \alpha_{c,j}+\begin{pmatrix} \Delta_j \\ 0 \end{pmatrix} \mu_c+\eta_{c,j} 
+	\end{array}
+	\right.
+	\]
+	where $Z=\begin{pmatrix} 1 & 0 \end{pmatrix}$, $\varepsilon_{c,j} \sim \mathcal{N}(0,H_{c,j})$, 
+	$T_j=\begin{pmatrix} 
+		1 & \int_{t_j}^{t_j+\Delta_j} e^{-I(v,j)}dv \\
+		0 & e^{-I(t_j+\Delta_j,t_j)}
+	\end{pmatrix}$  and $\eta_{c,j}=\begin{pmatrix} \xi_{c,j} \\ \zeta_{c,j} \end{pmatrix} \sim \mathcal{N}(0,Q_j)$ with
+	\[Q_i=\begin{pmatrix}
+		\int_{t_j}^{t_j+\Delta_j} \sigma(s)^2 \left(\int_{s}^{t_j+\Delta_j} e^{-I(v,s)} dv\right)^2 ds
+		& \int_{t_j}^{t_j+\Delta_j} \sigma(s)^2e^{-I(t_j+\Delta_j,s)} \int_s^{t_j+\Delta_j} e^{-I(v,s)} dv ds \\
+		\int_{t_j}^{t_j+\Delta_j} \sigma(s)^2e^{-I(t_j+\Delta_j,s)} \int_s^{t_j+\Delta_j} e^{-I(v,s)} dv ds & \int_{t_j}^{t_j+\Delta_j} \sigma(s)^2 e^{-2I(t_j+\Delta_j,s)} ds
+	\end{pmatrix}
+	\]
+	and $I(t,s)=\int_{s}^{t} \beta(u) du$.
+	How to control the error between these integrals and the approximated covariance matrix in \texttt{smoothSDE} ?
+	\item Proof that the process $X$ in section~\ref{section: CRCVM} is constrained for simple domains (infinite vertical or horizontal barrier, rectangle for instance) which involves the diameter of rotation $\Leftrightarrow$ proof that the process $D^{shore}(t)$ is strictly positive (something like the non negativity condition for the CIR process ?)
+\end{itemize}}
+
+\begin{comment}
 As discussed in \cite{ditlevsen_hypoelliptic_2019}, estimation for such hypoelliptic models is challenging. The main difficulties come from
@@
 It is also possible  to specify $\tau$ similarly as a function of $D_{shore}$ and $\Theta$ to model specific behaviour near the shore (e.g higher persistence along the shoreline).
+\end{comment}
 
@@
 \section{More details about the data}
+
+Figure~\ref{fig: trials and intertrials distributions} shows how these periods are distributed among the $6$ narwhals that were tracked.
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.9\textwidth]{images/data_exploration/trials.png}
+			\caption{Trial and Intertrial periods for each narwhal}
+			\label{fig: trials and intertrials distributions}
+		\end{subfigure}
+		\begin{subfigure}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.9\textwidth]{images/data_exploration/all_time_steps_histo}
+			\caption{Histogram of time steps}
+			\label{fig:alltimestepshisto}
+		\end{subfigure}
+	\end{figure}
+ Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals.
+Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.
 
@@
 \end{figure}
+
+ All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|p{8cm}|c|}
+		\hline
+		Covariate & Unit & Description & Domain \\
+		\hline
+		$D_{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
+		\hline
+		$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1,2\}$\\
+		\hline
+		$E_{ship}(t)=\frac{1}{D_{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
+		\hline
+		$E_{ship,I}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=1}$& $km^{-1}$ & intertrial exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
+		\hline
+		$E_{ship,T}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=2}$ & $km^{-1}$ & trial exposure level of the narwhal to the ship disturbance at time $t$. & $\R^+$\\
+		\hline
+		$D_{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
+		\hline
+		$E_{shore}(t)=\frac{1}{D_{shore}(t)} \1_{D_{shore}(t)<D_0}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
+		\hline
+		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical step vector at time $t$  & $[-\pi,\pi]$ \\
+		\hline 
+	\end{tabular}
+	\caption{Summary of the covariates}ubsubsection
+	\label{tab: covariates}
+\end{table}
 

### 2024-05-05 R/simulation_study/fjords/simulation_study_fjords.Rmd
separate Rscript for simulation study

--- 2024-05-02 R/simulation_study/simulation_study_fjords.Rmd
@@
 ### Generate sample
-
-```{r illustrative sample}
-res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[2,],times=seq(0,5*24,by=1/60),land=land,verbose=FALSE)
-
-data_sim=res$sim
-data_sim$ID=factor(rep(1,length(data_sim$y1)))
-data_shore=res$shore[,c("p1","p2")]
-data_sim=cbind(data_sim,data_shore)
-
-data_sim=data_sim[seq(1,length(data_sim$time),by=5),]
-
-plot_illust_fjords=ggplot()+geom_sf(data=land$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
-   geom_point(data=data_sim,aes(p1,p2),shape=5,size=0.1,col="red")+
-  geom_point(data=data_sim,mapping=aes(y1,y2),size=0.1)+
-  geom_path(data=data_sim,mapping=aes(y1,y2),size=0.1)+
-  geom_point(data = data_sim%>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
-  xlab("x") + ylab("y")
-
-ggplotly(plot_illust_fjords)
-
-ggsave(filename="illustrative_sample_fjords.png",plot=plot_illust_fjords,width=10,height=5)
-```
 

### 2024-05-05 R/simulation_study/rect/simulation_study_rect.Rmd
separate Rscript for simulation study

--- 2024-05-02 R/simulation_study/simulation_study_rect.Rmd
@@
 ```{r global variables}
-#number of batch
-N_BATCH=10
 
@@
 theta <- runif(n,-pi,pi);
-DistanceShore <- runif(n,0.2,2);
+DistanceShore <- runif(n,0.2,5);
 samples=data.frame(theta=theta,DistanceShore=DistanceShore)
@@
 theta_v <- seq(-pi,pi,length=30)
-Dshore_v<- seq(0.2,2,length=30)
+Dshore_v<- seq(0.2,5,length=30)
 pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
@@
 #fit with bivariate splines te
-m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))
+m1 <- gam(y~te(theta,DistanceShore,k=SP_DF,bs="cs"),knots=list(theta=seq(-pi,pi,len=SP_DF),DistanceShore=seq(0.2,5,len=SP_DF)))
 
@@
 #fit with bivariate splines te
-m2 <- gam(y~te(theta,ExpShore,k=SP_DF))
+m2 <- gam(y~te(theta,ExpShore,k=SP_DF,bs="cs"),knots=list(theta=seq(-pi,pi,len=SP_DF),ExpShore=seq(1/5,1/0.2,len=SP_DF)))
 
@@
 
-data_rect_hf_ne <- foreach(i=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
+data_rect_hf_ne <- foreach(k=1:N_BATCH,.packages=c("progress","MASS","sf","foreach","mgcv")) %dopar% {
   data=foreach (i=1:N_ID,.combine='rbind') %do% {
 
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs),v0,x0[i,],times_hf,land=domain,verbose=FALSE)
+    res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
 
@@
   par0 <- c(0,0,1,1,0)
+  knots=list("omega"=list(theta=seq(-pi,pi,len=SP_DF),ExpShore=seq(1/5,1/0.2,len=SP_DF)))
+
   m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
-                      other_data=list("H"=H_hf))
+                      other_data=list("H"=H_hf),knots=knots)
   m$fit(method="BFGS")
@@
   Dshore=sqrt(normal[1]^2+normal[2]^2)
-  if (Dshore<0.01) {
-    return (100)
+  if (Dshore<0.005) {
+    return (2000)
   }
@@
 
-```{r plot sim te ExpSHore low error samples}
+```{r plot sim te ExpShore low error samples}
 
@@
   par0 <- c(0,0,1,1,0)
+  knots=list("omega"=list(theta=seq(-pi,pi,len=SP_DF),DistanceShore=seq(0.2,5,len=SP_DF)))
+
   m <- SDE$new(formulas = formulas,data = data_rect_hf_ne[[paste("BATCH",k,sep="")]],type = "RACVM",
                       response = c("y1","y2"),par0 = par0,fixpar=c("mu1","mu2"),
-                      other_data=list("H"=H_hf))
+                      other_data=list("H"=H_hf),knots=knots)
   m$fit(method="BFGS")

### 2024-05-05 R/simulation_study/rect/rect_DistanceShore45.R

@@ -1,0 +1,157 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 45
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
+  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
[…]

### 2024-05-05 R/simulation_study/rect/rect_DistanceShore91.R

@@ -1,0 +1,157 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 91
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
+  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
[…]

### 2024-05-05 R/simulation_study/rect/rect_ExpShore20.R

@@ -1,0 +1,158 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 20
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
[…]

### 2024-05-05 R/simulation_study/rect/rect_ExpShore24.R

@@ -1,0 +1,158 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 24
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
[…]

### 2024-05-05 R/simulation_study/rect/rect_ExpShore32.R

@@ -1,0 +1,158 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 32
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
[…]

### 2024-05-05 R/simulation_study/rect/rect_ExpShore43.R

@@ -1,0 +1,158 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 43
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
[…]

### 2024-05-05 R/simulation_study/rect/rect_ExpShore58.R

@@ -1,0 +1,158 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_rect.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 58
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data_rect=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=domain,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_rect$ID)) {
+  sub_data=data_rect[data_rect$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
[…]

### 2024-05-08 R/simulation_study/fjords/simulation_study_fjords.Rmd
update simulation study

@@
 #auxiliary function to simulate from RACVM with specified smooth function for omega
-source("CVM_functions.R")
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))
 ```
@@
 ```{r}
-#number of batch
-N_BATCH=5
 
@@
 #duration of each track in hour
-TMAX=0.5*24
+TMAX=5*24
 
@@
 # Set the path to the directory containing the data
-par_dir=dirname(dirname(getwd())) #parent directory 
+par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
 greenland_data_path <- file.path(par_dir,"Data", "Greenland")
@@
 
+fomega_fast=function(cov_data,D0=0.05,omega0=120*pi,lambda=2,kappa=0.2) {
+    Dshore=cov_data$DistanceShore
+    theta=cov_data$theta
+    if (is.null(Dshore)){
+      Dshore=1/cov_data$ExpShore
+    }
+    coeff=exp(-kappa*(Dshore/D0)^2)
+    omega=omega0/2*(tanh(lambda*(theta+pi/2-atanh(-0.9)/lambda))+tanh(lambda*(theta-(pi/2-atanh(-0.9)/lambda))))*coeff
+    return(omega)
+}
+
 #constant nu
@@
 
-Dshore=seq(from=0.1,to=2,length.out=30)
+Dshore=seq(from=0.01,to=2,length.out=30)
 theta=seq(from=-pi,to=pi,length.out=30)
@@
 true_tau=matrix(ftau_bump(grid),30,30)
-true_omega=matrix(fomega(grid),30,30)
+true_omega=matrix(fomega_fast(grid),30,30)
 
@@
 
-data_fjords_hf_ne <- foreach(k=1:N_BATCH,.packages=c("progress","MASS","sf","foreach")) %dopar% {
-  data=foreach (i=1:N_ID,.combine='rbind') %do% {
+data_fjords_hf_ne <- foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf")) %dopar% {
 
-    res=sim_theta_CRCVM(ftau_constant,fomega,fnu_constant,log_sigma_obs=log(sigma_obs),v0,x0[i,],times_hf,land=land,verbose=FALSE)
+    res=sim_theta_CRCVM(ftau_constant,fomega_fast,fnu_constant,log_sigma_obs=log(sigma_obs),v0,x0[i,],times_hf,land=land,verbose=FALSE)
 
@@
 }
-  data 
-}
 
@@
 
-names(data_fjords_hf_ne)=paste("BATCH",1:N_BATCH,sep="")
 ```
@@
 count=0
-for (k in 1:length(names(data_fjords_hf_ne))) {
-
-  data=data_fjords_hf_ne[[paste("BATCH",k,sep="")]]
-  for (id in unique(data$ID)) {
-    sub_data=data[data$ID==id,]
+for (id in unique(data_fjords_hf_ne$ID)) {
+  sub_data=data_fjords_hf_ne[data_fjords_hf_ne$ID==id,]
   if (nrow(sub_data) < n_hf) {
@@
 }
-}
-cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
+
+cat(count/(N_ID)*100,"percent of the samples reached land")
 ```
@@
 plot_samples_fjords_hf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_path(data=data_fjords_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID),size=0.1)+
-  geom_point(data = data_fjords_hf_ne$BATCH1%>% filter(!duplicated(ID)),
+  geom_point(data=data_fjords_hf_ne,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_fjords_hf_ne,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_fjords_hf_ne%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
@@
 ggplotly(plot_samples_fjords_hf_ne)
+```
+```{r plot subsampled data}
+
+data_fjords_lf_ne=data_fjords_hf_ne[seq(1,length(data_fjords_hf_ne$time),by=50),]
+plot_samples_fjords_lf_ne=ggplot()+geom_sf(data=land$geometry,fill=NA)+coord_sf(datum=st_crs(32626))+
+  geom_point(data=data_fjords_lf_ne,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_path(data=data_fjords_lf_ne,mapping=aes(y1,y2,color=ID),size=0.1)+
+  geom_point(data = data_fjords_lf_ne%>% filter(!duplicated(ID)),
+             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
+  xlab("x") + ylab("y")
+
+ggsave(filename="samples_fjords_lf_ne_omega.png",plot=plot_samples_fjords_lf_ne,width=10,height=5)
+ggplotly(plot_samples_fjords_lf_ne)
 ```

### 2024-05-08 R/simulation_study/fjords/fjords_DistanceShore64.R

@@ -1,0 +1,164 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_fjords.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 64
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+data_fjords=foreach (i=1:N_ID,.combine='rbind',.packages=c("mgcv","progress","MASS","sf")) %dopar% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=land,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+#stop cluster
+stopCluster(cl)
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_fjords$ID)) {
+  sub_data=data_fjords[data_fjords$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
[…]

### 2024-05-08 R/simulation_study/fjords/fjords_ExpShore77.R

@@ -1,0 +1,164 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_fjords.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 77
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+data_fjords=foreach (i=1:N_ID,.combine='rbind',.packages=c("mgcv","progress","MASS","sf")) %dopar% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=land,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+#stop cluster
+stopCluster(cl)
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_fjords$ID)) {
+  sub_data=data_fjords[data_fjords$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
[…]

### 2024-05-08 R/simulation_study/fjords/fjords_ExpShore95.R

@@ -1,0 +1,164 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+source("set_up_fjords.R")     #get set up for simulation study
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 95
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+#setup parallel backend to use many processors
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+data_fjords=foreach (i=1:N_ID,.combine='rbind',.packages=c("mgcv","progress","MASS","sf")) %dopar% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=land,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+#stop cluster
+stopCluster(cl)
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data_fjords$ID)) {
+  sub_data=data_fjords[data_fjords$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+
+# Add covariates to simulation -----------
+
+signed_angle <- function(u, v) {
[…]

### 2024-05-08 R/simulation_study/rect/rect_DistanceShore52.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+#OAR -n rect_DistanceShore52
+#OAR -l /core=16,walltime=01:30:00
+#OAR --stdout rect_DistanceShore52.out
+#OAR --stderr rect_DistanceShore52.err
+#OAR --project pr-formation-ced-calcul
+source /applis/site/guix-start.sh
+Rscript rect_DistanceShore52.R

### 2024-05-08 R/simulation_study/rect/rect_DistanceShore69.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+#OAR -n rect_DistanceShore69
+#OAR -l /core=16,walltime=01:30:00
+#OAR --stdout rect_DistanceShore69.out
+#OAR --stderr rect_DistanceShore69.err
+#OAR --project pr-formation-ced-calcul
+source /applis/site/guix-start.sh
+Rscript rect_DistanceShore69.R

### 2024-05-08 R/preprocessing/data_preprocessing.Rmd
preproseccing results with modified land polygons

@@
 
+```{r save points as geojson in data folder}
+points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
+st_crs(points_sf)="+init=EPSG:32626 +units=km"
+st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"))
+```
+
 # Coastline and land data
@@
 
-#get the coastline geometry from the geojson file
-coastline_utm<-st_read(file.path(greenland_data_path,"coastline_utm.geojson"))
-st_crs(coastline_utm)="+init=EPSG:32626"
-
 #get land polygons
@@
 
+#get land polygons
+modified_land_utm<-st_read(file.path(greenland_data_path,"modified_land_utm.shp"))
+modified_land_utm <- st_transform(modified_land_utm, crs = "+init=EPSG:32626 +units=km")
+
+#scoresby sound google
+land_google_utm<-st_read(file.path(greenland_data_path,"land_google_utm.shp"))
+st_crs(land_google_utm)="+init=EPSG:32626"
 ```
@@
 ```{r plot coastline and land}
-plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill=NA,col="brown")+
+plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill="grey",col="brown")+
   coord_sf(datum=st_crs(32626))
-plot_coastline=ggplot()+geom_sf(data=coastline_utm$geometry,fill=NA,col="blue")+
+plot_modified_land=ggplot()+geom_sf(data=modified_land_utm$geometry,fill="grey",col="blue")+
+  coord_sf(datum=st_crs(32626))
+plot_land_google=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey",col="blue")+
   coord_sf(datum=st_crs(32626))
@@
 ```{r display coastline , fig.cap="Coastline polygons and linestrings"}
-plot_coastline
+ggplotly(plot_modified_land)
+```
+
+```{r}
+ggplotly(plot_land_google)
 ```
@@
 res_land=nearest_points(positions=narwhal_positions,coastline=land_utm)
-res_coastline=nearest_points(positions=narwhal_positions,coastline=coastline_utm)
 })
@@
 plot(res_land$min_distances-dataprep$DistanceShore)
-plot(res_coastline$min_distances-dataprep$DistanceShore)
 ```
@@
 summary(res_land$min_distances)
-summary(res_coastline$min_distances)
 summary(dataprep$DistanceShore)
 ```
-Coastline distances are a bit closer to original distances in the dataset.
-Anyway, we shall keep distance computed from the land in the sequel, since the land polygons will be the reference domain
-for constrained movement of the narwhals.
 
@@
 
-df=data.frame(DistanceShore=res_coastline$min_distances[res_coastline$min_distances>0])
+df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
 histo2=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (km)")
-
-df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
-histo3=ggplot(df, aes(x =DistanceShore)) +
   geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")
@@
 histo2
-histo3
 ```
@@
 ```{=tex}
-There are pretty big differences between distances calculates with the coastline data and the land data.
-
-For the rest of the analysis, we'll keep the bounded domain defined by the polygons in \texttt{land\_utm} sf object as a reference.
+
+For the rest of the analysis, we'll keep the bounded domain defined by the polygons in \texttt{modified\_land\_utm} sf object as a reference.
 Doing that involves having some points on land, that is points with Distance to shore equal to 0.
@@
 
-cat("Percentage of points on land with distance from coastline : \n",
-    length(which(res_coastline$min_distances==0))/length(dataprep$time)*100)
+cat("Percentage of points on land with distance from modified : \n",
+    length(which(res_modified_land$min_distances==0))/length(dataprep$time)*100)
 ```
@@
 ```{r check original points on land}
-check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
+check_plot1=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
-  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
+  geom_point(data=dataprep,aes(x,y,col=(DistanceShore<0.05)),size=0.2)+
     scale_color_viridis_d()
@@
 
-check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
+check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
@@
 
-df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<2 & res_land$min_distances>0  ])
+df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<0.2 & res_land$min_distances>0  ])
 histo=ggplot(df, aes(x =DistanceShore)) +
@@
 
-```{=tex}
-$9 \%$ of the points are on land.
-If the points are too close to then land (let's say distance to shore < 50m), the covariates AngleShore and Angle Normal are likely to be irrelevant.
-```
-
 ```{r plot empirical velocities and deviation angles through time}

### 2024-05-08 Latex/main.tex
new paper version

@@
 
-The diffusion models are discussed in Section 2 and simulation study results are shown.
-Then,  in Section 3 we give an overview of the narwhal data available for the analysis and the results of the application of our models to this data.
-The last section is dedicated to a discussion about these results, and potential extension of the models.
-
-\section{Diffusion models}
-
-The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
-First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement and constrain the position within a polygon.
-
-\subsection{Standard Rotational Advective correlated velocity model (RACVM)}
-\label{section: RACVM}
-
-\subsubsection{Solution of the SDE}
-
-We write $A=\begin{pmatrix} 
-	\frac{1}{\tau} & -\omega \\
-	\omega & \frac{1}{\tau}
-\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}
-
-\begin{equation*} \left\{
-	\begin{array}{l}
-		dX(t)=V(t)dt \\
-		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
-	\end{array}
-	\right.
-	\label{eq: RACVM equation bis}
-\end{equation*}
-where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
-In this section, we obtain explicit distributional results about the process $X$.
-Though this model is considered in several papers, we found no mention of the closed form formulas derived here.
-In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
-A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
-In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
-But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
-Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula.
-\\
-
-The transition density of the velocity Ornstein-Uhlenbeck process is  
-\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
-\end{equation}
-
-where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
-Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.
-
-%stationary distribution, distribution of the norm, distribution of the phase
-This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}.
-After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$.
-In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.
-
-By integrating the velocity equation over time one gets the following conditional distribution for the location process :
-\begin{equation}
-	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
-\end{equation}
-Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :
-
-\[
-Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
-\]
-The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
-The two components of the location and velocity processes are independent.
-and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
-Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm.
-Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\
-
-%samples with non zero and zero mu
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
-		\caption{}
-	\end{subfigure}
-	\caption{(a) Samples of a RACVM with parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h.
-The black cross indicates the starting point of the process.
-The samples were obtained from the exact transition densities.
-(b) Histogram of the empirical velocities obtained from the samples, along with the theoretical probability density function.}
-	\label{fig: sample RACVM}
-\end{figure}
-
-\subsubsection{State-space model}
-\label{section: state space model}
-In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
-We suppose here that $n$ positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$.
-$Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$.
-The statistical model is the following:
-
-\begin{equation}  \left\{
-	\begin{array}{l}
-		Y_j=X_j+\varepsilon_j \quad j \in \{1,\cdots,n\}\\
-		\varepsilon_j \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
-		dX(t)=V(t)dt  \\
-		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
-	\end{array}
-	\right.
-\end{equation}
-
-We consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time and make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that these parameters are constant on each time step $[t_j,t_{j+1}]$, equal to their value at time $t_j$.
-Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $X_j=X(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix}
-	\frac{1}{\tau_j} & -\omega_j \\
-	\omega_j & \frac{1}{\tau_j}
-\end{pmatrix}$.
-
-\begin{equation}
-	\forall j \in \{1, \cdots, n\}, \quad 
-	\left\{ \begin{array}{l}
-		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
-		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
-	\end{array}\right. \\
-	\label{eq: RACVM autoregressive formula}
-\end{equation}
-
-where $\zeta_j$ and $\xi_j$ are centered normal random vectors.
-Denote $\alpha_j=\begin{pmatrix} X_{j,1}  & X_{j,2} & V_{j,1} & V_{j,2}\end{pmatrix}^\top$. The state space equations are
-
-\begin{equation}
-\left\{
-\begin{array}{l}
-	Y_j=Z\alpha_j+\varepsilon_j \\
-	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
-\end{array}
-\right.
-\label{eq: RACVM state space}
-\end{equation}
-where the link matrices are 
-\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
-	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
-	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
-\]
-
-$\varepsilon_j \sim \mathcal{N}(0,\sigma_{obs}^2 I_2)$ and the vector $\eta_j=\begin{pmatrix} \xi_{j,1} & \xi_{j,2} & \zeta_{j,1} & \zeta_{j,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero.
-Its covariance matrix $Q_j$ can be computed explicitly.
-\begin{equation}
-	Q_j=\begin{pmatrix}
-		M_j & L_j \\
-		L_j & N_j
-	\end{pmatrix}
-	\label{eq: state space covariance}
-\end{equation}
-where $M_j$, $N_j$and $L_j$ are block matrices of size $2 \times 2$ such that 
-\[M_j=\left( \Delta-2 \frac{\omega_j \sin(\omega_j \Delta_j)-\frac{1}{\tau_j} \cos(\omega_j \Delta_j)}{\frac{1}{\tau_j^2}+\omega_j^2 } \exp\left( -\frac{\Delta_j}{\tau_j} \right) +\frac{\tau_j}{2} \left( \frac{\omega_j^2-\frac{3}{\tau_j^2}}{\frac{1}{\tau_j^2}+\omega_j^2}-\exp\left( -\frac{2\Delta_j}{\tau_j}\right)\right) \right)I_2\]
-
-\[N_j=\frac{2\nu_j^2}{\pi}\left(1-e^{-\frac{2 \Delta_j}{\tau_j}}\right)I_2\]
-
-\[L_j=\begin{pmatrix} l_1 & l_2 \\
-	l_2 & l_1\end{pmatrix}\]
-where 
-\[l_1=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times \left( 1+\exp\left( -\frac{2\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right) \cos(\omega_j \Delta_j)\right)\]
-\[l_2=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times\left( \exp\left( -\frac{\Delta_j}{\tau_j}\right) \sin(\omega_j \Delta_j)-\frac{\omega_j \tau_j}{2} \left(1-\exp\left( -2 \frac{\Delta_j}{\tau_j}\right) \right)\right)\]
-
-Again, one can check that for $\omega_j=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
-This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter.
-This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$.
-Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.
-
-\subsection{Constrained rotational correlated velocity model}
-\label{section: CRCVM}
-
-\subsection{Formulation of the model}
-
-Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
-To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore.
-To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector.
-The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
-We denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
-\begin{equation} \left\{
-	\begin{array}{l}
-		dX(t)=V(t) dt \\
-		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)
-
-	\end{array}
-	\right.
-	\label{eq: CRCVM equation}
-\end{equation}
-
-with $A(X(t),V(t))=\begin{pmatrix} 
-	\frac{1}{\tau} & -\omega(t) \\
-	\omega(t) & \frac{1}{\tau}
-\end{pmatrix}$
-where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
-$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
-
-Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
-For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
-\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
-In this case, the formulas of section~\ref{section: RACVM} are still valid.
-The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
-We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
-An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation.
-It should always remain significantly lower than $D^{shore}$ for the process to be constrained.
-Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
-The closer we are to the shore, the faster the velocity should rotate,
-whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
-Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
-A weight based on the distance to the shore can be defined as follows :
-\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
-
-Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :
-
-\begin{equation}
-f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D^{shore})
-\label{eq: smooth omega}
-\end{equation}
-for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
-
-This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
-Any smooth function that meets minimum requirements could be used instead.
-The general model is very flexible, and different species-specific behaviours could be described.
-An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
-Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
-In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
-A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
-The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
-
-\begin{figure}[H]
-	\centering
-		\centering
-		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
-		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
-		\label{fig: trueomega}
-\end{figure}
-
-\begin{figure}[H]
-		\centering
-		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
-		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
-	\label{fig: splineomega}
-\end{figure}
-
-\begin{figure}[H]
-		\centering
-		\includegraphics[scale=0.7]{images/simulation study/illustrative_sample_fjords.png}
-	\caption{Sample of a CRCVM.
-x and y axis are in metres and initial positions are indicated by a red cross and red points represent the nearest points on the boundary of the domain.
-One  five day sample with time step $h=5$ min within the spatial constraints of Scoresby Sound fjords system in South-East Greenland.
-Parameters are $\nu=4$ km/h and $\tau=1$ h.
-}
-	\label{fig: samples_fjords_CRCVM}
-\end{figure}
-
-\label{section: simulation study}
-%samples and estimation
-%discussion about tensor splines vs additive splines
-
-\subsubsection{Simulation study}
-\label{section: simulation constrained motion}
-In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\
-
-100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM.
-Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case.
-Denoting $n$ the number of observations, which is the same for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:
-
-\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
-	\begin{array}{l}
-		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
-		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
-		dX_i(t)=V_i(t) dt \quad \\
-		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
-		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
-		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
-		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
-		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
-		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
-	\end{array}
-	\right.
-	\label{eq: rect simulation parameters}
-\end{equation}
-Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
-1.8$ for the construction of tensor splines basis function.
-We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h.
-The values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ are given in Table~\ref{table: estimation performance rect}.
-For each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
-To simulate the trajectories, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
-\begin{equation}
-	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
-\end{equation}
-  where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
-The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
-None of the $600$ simulations reached the boundary.
-\\
-
-100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
-The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain.
-The exact values of the spline coefficients can be found in Table~\ref{table: estimation performance fjords}.
-Again, for each trajectory, initial velocity was set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
-Time steps are also kept constant, equal to $1$ min.
-None of the $600$ trajectories reached land.
-Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
-For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
-Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-
-Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
-The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
-Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation.
-We refer to \cite{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations.
-Similarly, we refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the mixed effect model for the parameter of the SDE and the regularization terms that are based on a wiggliness measure involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
-\\
-
-$100$ batches of noisy samples obtained as described in equation~\ref{eq: rect simulation parameters} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
-We thus obtain $100$ parameters estimate in the two cases.
-The covariates $D^{shore}$ and $\Theta$ are computed from the observed noisy positions.
-It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging.
-The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom is set to $3$ for each covariate to match the simulated data from equation~\ref{eq: rect simulation parameters}.
-This corresponds to one single knot in each dimension.
-Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
-Measurement error $\sigma_{obs}$ is supposed to be known.
-We fix initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h, $\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectoriesin the rectangular and the fjords domain.
-The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q4_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q2_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
-	\caption{}
-\end{subfigure}
-\begin{subfigure}{0.48\textwidth}
-	\centering
-	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
-	\caption{}
-\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only,.
-The red curve represents the true value of $\omega$.
-(a)-(b) Estimation from samples in the rectangular domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-(c)-(d) Estimation from samples in the fjords domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-}
-\label{fig: marginal_estimates_CRCVM}
-\end{figure}
-
-For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the absolute biais of the estimator.
-The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
-
-\begin{table}[H]
-		\centering
-	\begin{tabular}{|c|c|c|c|}
-		\hline
-		Coefficient  & True value & Biais & Root mean squared error  \\
-		\hline
-		$\log(\tau_0)$  & 0& 0.24 &  0.26\\
-		\hline
-		$\log(\nu_0)$  & 1.38 & 0.03 & 0.04 \\
-		\hline
-		$\omega_0$  &  -0.19  & 0.80 & 0.82 \\
-		\hline
-		$\omega_1$  & 16.68 & 11.54 & 12.32 \\
-		\hline
-		$\omega_2$  & -3.58 & 2.58 &  3.04\\
-		\hline
-		$\omega_3$  & 13.07 &  8.74 &  8.84\\
-		\hline
-		$\omega_4$  & 42.11 & 26.36 &  29.04\\
-		\hline
-		$\omega_5$  & 13.89 & 8.44 & 8.54 \\
-		\hline
-		$\omega_6$  &  61.53 & 51.45 &  51.60\\
-		\hline
-		$\omega_7$  & 10.15 & 2.54 & 3.37 \\
-		\hline
-		$\omega_8$  &  12.31 & 4.22 & 4.45 \\
-		\hline
-		$\sigma_{\tau}$ & 0.1 & & \\
-		\hline
-		$\sigma_{\nu}$ & 0.3 & & \\
-		\hline
-	\end{tabular}
-	\caption{Measures of estimation performance in the rectangular domain}
-	\label{table: estimation performance rect}
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\begin{tabular}{|c|c|c|c|}
-		\hline
-		Coefficient  & True value & Biais & Root mean squared error  \\
-		\hline
-		$\log(\tau_0)$  &  0 &  &  \\
-		\hline
-		$\log(\nu_0)$  &  1.38 &  &  \\
-		\hline
-		$\omega_0$  & -0.34 &  &  \\
-		\hline
-		$\omega_1$  & 14.17  &  &  \\
-		\hline
-		$\omega_2$  & -2.57 &  &  \\
-		\hline
-		$\omega_3$  & 11.14 &  &  \\
-		\hline
-		$\omega_4$  & 36.48  &  &  \\
-		\hline
-		$\omega_5$  & 11.73 &  &  \\
-		\hline
-		$\omega_6$  & 53.69 &  &  \\
-		\hline
-		$\omega_7$  & 9.23 &  &  \\
-		\hline
-		$\omega_8$  & 10.19 &  &  \\
-		\hline
-		$\sigma_{\tau}$ & 0.1 & & \\
-		\hline
-		$\sigma_{\nu}$ & 0.3 & & \\
-		\hline
-	\end{tabular}
-	\caption{Measures of estimation performance in the Scoresby Sound domain}
-	\label{table: estimation performance fjords}
-\end{table}
-
-\section{Application to narwhals behavioral response}
-
-In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
-
-\subsection{Description of the movement data}
+In Section 2, we give an overview of the narwhal data available for the analysis.
+The diffusion models are discussed in Section 3. Inference and simulation study results are explained in section 4.
+Then, our models are applied to analyse a behavioral response of the narwhals in section 5. The last section is dedicated to a discussion about these results, and potential extension of the models.
+
+\section{Movement data of six narwals in South-East Greenland}
+
+\subsection{Description of the controlled exposure experiments}
 
@@
 For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.
+We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.
+These are noisy observations of the underlying unobserved true positions $x_{ij}$.\\
 
@@
 The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move.
-The closest point on the shoreline to the observed position $y_{ij}$  is denoted $p_{ij}$, and the distance to this point, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$.
+The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$.
+Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$.
 The resulting values range from $0$ to $7.5$ km.
@@
 A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.
-The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
 
@@
 		%nodes
+		\node[left] at (-1,3.5) {Land};
+		\node[right] at (1,3.5) {Water};
 		\node[left] at (0,0) {$\partial \mathcal{D}$};
@@
 
-Exposure to ship covariate for narwhal $i$ at time $t_{ij}$ , denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
+The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
+Exposure to ship covariate for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
 $D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
@@
 
+\section{Diffusion models}
+
+The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.
+First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement and constrain the position within a polygon.
+
+\subsection{Standard Rotational Advective correlated velocity model (RACVM)}
+\label{section: RACVM}
+
+We write $A=\begin{pmatrix} 
+	\frac{1}{\tau} & -\omega \\
+	\omega & \frac{1}{\tau}
+\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}
+
+\begin{equation*} \left\{
+	\begin{array}{l}
+		dX(t)=V(t)dt \\
+		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
+	\end{array}
+	\right.
+	\label{eq: RACVM equation bis}
+\end{equation*}
+where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
+In this section, we obtain explicit distributional results about the process $X$.
+Though this model is considered in several papers, we found no mention of the closed form formulas derived here.
+In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
+A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
+In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
+But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
+Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula.
+\\
+
+The transition density of the velocity Ornstein-Uhlenbeck process is  
+\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
+\end{equation}
+
+where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
+Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.
+
+%stationary distribution, distribution of the norm, distribution of the phase
+This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}.
+After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$.
+In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.
+
+By integrating the velocity equation over time one gets the following conditional distribution for the location process :
+\begin{equation}
+	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
+\end{equation}
+Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :
+
+\[
+Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
+\]
+The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
+The two components of the location and velocity processes are independent.
+and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
+Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm.
+Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\
+
+%samples with non zero and zero mu
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/illustrative_sample_standard.png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/illustrative_sample_fjords.png}
+		\caption{}
+	\end{subfigure}
+	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min.
+The red cross indicates the starting point of the process.
+The samples were obtained from the exact transition densities.
+(b) Samples of a CRCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
+Position are sampled from constant time step $h= 1$ min and then doownsampled to $5$ min.The samples were obtained from approximate transition densities  with constant covariates on each time step.
+}
+	\label{fig: sample RACVM}
+\end{figure}
+
+\subsection{Constrained rotational correlated velocity model}
+\label{section: CRCVM}
+
+Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
+To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore.
+To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector.
+The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
+We denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
+\begin{equation} \left\{
+	\begin{array}{l}
+		dX(t)=V(t) dt \\
+		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)
+
+	\end{array}
+	\right.
+	\label{eq: CRCVM equation}
+\end{equation}
+
+with $A(X(t),V(t))=\begin{pmatrix} 
+	\frac{1}{\tau} & -\omega(t) \\
+	\omega(t) & \frac{1}{\tau}
+\end{pmatrix}$
+where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
+$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
+
+Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
+For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
+\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
+In this case, the formulas of section~\ref{section: RACVM} are still valid.
+The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
+We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
+An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation.
+It should always remain significantly lower than $D^{shore}$ for the process to be constrained.
+Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
+The closer we are to the shore, the faster the velocity should rotate,
+whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
+Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
+A weight based on the distance to the shore can be defined as follows :
+\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
+
+Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :
+
+\begin{equation}
+f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D^{shore})
+\label{eq: smooth omega}
+\end{equation}
+for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
+
+This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
+Any smooth function that meets minimum requirements could be used instead.
+The general model is very flexible, and different species-specific behaviours could be described.
+An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
+Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
+In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
+A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
+The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
+
+\begin{figure}[H]
+	\centering
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
+		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
+		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
+		\label{fig: trueomega}
+\end{figure}
+
+\begin{figure}[H]
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
+		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
+		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
+	\label{fig: splineomega}
+\end{figure}
+
+\label{section: simulation study}
+%samples and estimation
+%discussion about tensor splines vs additive splines
+
+\section{Inference}
+
 \subsection{Mixed effet model for ship exposure and shore effects}
@@
 
-For the baseline model, only the effect of the shore is included through the shore exposure covariate $E_{shore}$ and the angle covariate $\Theta$.\\
+For the baseline model, only the effect of the shore is included through the shore exposure covariate $E^{shore}$ and the angle covariate $\Theta$.\\
 Although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance.
@@
 
-Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, $\tau_i$, $\nu_i$  and $\omega_i$ the parameters of the diffusion for the narwhal $i$.
+Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion .
 
@@
 		\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
-	\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
-	f_{\omega}(E^{shore},\Theta)=\omega_{0,pre}+\sum_{k=1}^{L} \omega_{k,pre} \psi_k(E^{shore},\Theta) + b_{\omega,pre}^{(i)}, \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
+		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\hat{\Theta}_i(t))\\
+		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
 	\end{array}
@@
 
-$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{pre}=\begin{pmatrix} b_{\tau,pre}^{(i)} & b_{\nu,pre}^{(i)} & b_{\omega,pre}^{(i)} \end{pmatrix}^\top \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{pre}}\right)\right)$
-are the individual parameters, where $\lambda_{pre}=\begin{pmatrix} \frac{1}{\sigma_{\tau,pre}^2} & \frac{1}{\sigma_{\nu,pre}^2} & \frac{1}{\sigma_{\omega,pre}^2} \end{pmatrix}^\top$ with variances $\sigma_{\tau,pre}^2$, $\sigma_{\nu,pre}^2$, $\sigma_{\omega,pre}^2$.
-
-The $\psi_k$ in equation~\ref{eq: exact baseline model} are  known bivariate cubic splines basis functions, constructed through tensor products from the univariate basis functions as described in \cite{wood_generalized_2017}, section 4.1.8.
+$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
+are the individual parameters.
+
+The functions $\psi_k$ in equation~\ref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions.
+We refer to \cite{wood_generalized_2017} section $4.
+1.8$ for the construction of tensor splines basis functions.
+
 %Add plot of the spline basis functions \\
-$L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$.
+The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$.
 It is needed to choose them before hand.
@@
 		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
-		\log(\tau^{(i)}_j)=\log(\tau_{0,post}) +\sum_{k=1}^{q_{\tau}} \log(\tau_{k,post})\phi_{k}(E_{ship}(t))+b^{(i)}_{\tau,post} \\
-		\log(\nu^{(i)}_j)=\log(\nu_{0,post}) +  \sum_{k=1}^{q_{\nu}} \log(\nu_{k,post}) \xi_{k}(E_{ship}(t)) +b^{(i)}_{\nu,post}  \\
+		\log(\tau_{ij})=\log(\tau_{0,post}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
+		\log(\nu_{ij})=\log(\nu_{0,post}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
 		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
-		f_{\omega}(E^{shore},\Theta)=\omega_{0,post}+\sum_{k=1}^{L} \omega_{k,post} \psi_k(E^{shore},\Theta) + b_{\omega,post}^{(i)}, \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
 	\end{array}
@@
 
-Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts and $b^{(i)}_{post} \sim \mathcal{N}\left( 0, \diag\left(\frac{1}{\lambda_{post}}\right)\right)$ with $\lambda_{post}=\begin{pmatrix} \frac{1}{\sigma_{\tau,post}^2} & \frac{1}{\sigma_{\nu,post}^2} & \frac{1}{\sigma_{\omega,post}^2} \end{pmatrix}^\top$ and $q_{\tau}$ and $q_{\nu}$ are the spline degrees of freedom for the exposure to the ship covariate.
+Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters.
+The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
+
+\subsection{State-space model}
+\label{section: state space model}
+In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
+This is the approach we use here.
+In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$.\\
+ Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
+	\frac{1}{\tau_{i}} & -\omega_{ij} \\
+	\omega_{ij} & \frac{1}{\tau_{i}}
+\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$.
+\begin{comment}
+\begin{equation}
+	\forall j \in \{1, \cdots, n_{i,pre}\}, \quad 
+	\left\{ \begin{array}{l}
+		V_{i,j+1}=\exp(-A_{ij}\Delta_{ij})V_{ij}+\zeta_{ij} \\
+		X_{i,j+1}=X_{ij}+A_{ij}^{-1}(I_2-\exp(-A_{ij}\Delta_{ij}))V_{ij}+\xi_{ij}
+	\end{array}\right. \\
+	\label{eq: RACVM autoregressive formula}
+\end{equation}
+
+where $\zeta_{ij}$ and $\xi_{ij}$ are centered normal random vectors.
+\end{comment}
+
+The state space equations are
+
+\begin{equation}
+	\left\{
+	\begin{array}{l}
+		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \\
+		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij}
+	\end{array}
+	\right.
+	\label{eq: RACVM state space}
+\end{equation}
+where the link matrices are 
+\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_{ij}=\begin{pmatrix} I_2 & A_{ij}^{-1}(I_2-\exp(-A_{ij} \Delta_j)) \\ 0_{2,2} & \exp(-A_{ij} \Delta_j) \end{pmatrix}\]
+
+The vector $\eta_{ij}=\begin{pmatrix} \xi_{ij,1} & \xi_{ij,2} & \zeta_{ij,1} & \zeta_{ij,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero.
+Its covariance matrix $Q_{ij}$ can be computed explicitly.
+\begin{equation}
+	Q_{ij}=\begin{pmatrix}
+		M_{ij} & L_{ij} \\
+		L_{ij} & N_{ij}
+	\end{pmatrix}
+	\label{eq: state space covariance}
+\end{equation}
+where $M_{ij}$, $N_{ij}$and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
+\[M_{ij}=\frac{\sigma_i^2}{C_{ij}}\left( \Delta-2 \frac{\omega_{ij} \sin(\omega_{ij} \Delta_{ij})-\frac{1}{\tau_{i}} \cos(\omega_{ij} \Delta_{ij})}{C_{ij} } \exp\left( -\frac{\Delta_{ij}}{\tau_{ij}} \right) +\frac{\tau_{i}}{2} \left( \frac{\omega_{ij}^2-\frac{3}{\tau_{ij}^2}}{C_{ij}}-\exp\left( -\frac{2\Delta_{ij}}{\tau_i}\right)\right) \right)I_2\]
+
+\[N_{ij}=\frac{\sigma_i^2}{2}\left(1-e^{-\frac{2 \Delta_{ij}}{\tau_{i}}}\right)I_2\]
+
+\[L_{ij}=\begin{pmatrix} a & b \\
+	b & a\end{pmatrix}\]
+with
+\[a=\frac{\sigma_i^2}{2 C_{ij}^2 } \times \left( 1+\exp\left( -\frac{2\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \cos(\omega_{ij}\Delta_{ij})\right)\]
+\[b=\frac{\sigma_{i}^2}{2 C_{ij}^2} \times\left( \exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \sin(\omega_{ij} \Delta_{ij})-\frac{\omega_{ij} \tau_{i}}{2} \left(1-\exp\left( -2 \frac{\Delta_{ij}}{\tau_{i}}\right) \right)\right)\]
+and $C_{ij}=\frac{1}{\tau_i^2}+\omega_{ij}^2$, $\sigma_{i}=\frac{2\nu_i}{\sqrt{\pi \tau_i}}$.
+Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
+This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter.
+This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$.
+Therefore, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models.
+
+\section{Simulation study for the baseline model}
+\label{section: simulation constrained motion}
+In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.
+We considered observation of a baseline model as described in equation~\ref{eq: exact baseline model}\\.
+
+We simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in a rectangular domain with sides of length $8$ km from a CRCVM.
+This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
+We fix $\sigma_{obs}=5$ m to represent low measurement errors.
+We choose $L=8$ for the degree of freedom of the bivariate splines of $D^{shore}$ and $\Theta$.
+The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{pre}$ and $\nu_{pre}$ are fixed respectively to $1$ h and $4$ km/h.
+The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance rect}.
+For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
+WE kept the same $6$ initial positions for all the batches of trajectories.
+Toobtain teh samples of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
+\begin{equation}
+	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
+\end{equation}
+where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
+The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
+None of the $600$ simulations reached the boundary.
+\\
+
+We also simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ with time step $h=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
+The parameters $\sigma_{obs}$,$L$,$\sigma_{\tau,pre}$,  $\sigma_{\nu,pre}$ , $\tau_{pre}$,$\nu_{pre}$ are kept the same as in the recangular case, and only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are changed to adapt to this specific fjords domain.
+The exact values of the spline coefficients are found in Table~\ref{table: estimation performance fjords}.
+Again, for each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
+None of the $M\times N=600$ trajectories reached land.
+
+\begin{comment}Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
+For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
+Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
+\end{comment}
+
+Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
+The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
+Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation.
+We refer to \cite{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations.
+Similarly, we refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the mixed effect model for the parameter of the SDE and the regularization terms that are based on a wiggliness measure involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
+\\
+
+$100$ batches of noisy samples obtained as described in equation~\ref{eq: rect simulation parameters} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
+We thus obtain $100$ parameters estimate in the two cases.
+The covariates $D^{shore}$ and $\Theta$ are computed from the observed noisy positions.
+It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging.
+The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
+The marginal spline degree of freedom is set to $3$ for each covariate to match the simulated data from equation~\ref{eq: rect simulation parameters}.
+This corresponds to one single knot in each dimension.
+Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
+Measurement error $\sigma_{obs}$ is supposed to be known.
+We fix initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h, $\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectoriesin the rectangular and the fjords domain.
+The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q4_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q2_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only,.
+The red curve represents the true value of $\omega$.
+(a)-(b) Estimation from samples in the rectangular domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+(c)-(d) Estimation from samples in the fjords domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+}
+	\label{fig: marginal_estimates_CRCVM}
+\end{figure}
+
+For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimator.
+The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+		\hline
+		Coefficient  & True value & Absolute biais  & Relative biais & RMSE\\
+		\hline
+		$\log(\tau_0)$  & 0& 0.08 &  & 0.16\\
+
+		$\log(\nu_0)$  & 1.38 & -0.07 & $-5.4 \%$ & 0.10 \\
+
+		$\omega_0$  &  0.20  & -0.11 & $-58\%$ & 0.30 \\
+
+		$\omega_1$  & 7.41 & -0.4 & $-5.4 \%$ & 1.11 \\
+
+		$\omega_2$  & -2.78 & -1.87 &  $67.3\%$ & 4.30\\
+
+		$\omega_3$  & 4.05 &  0.30 & $7.4\%$ & 0.96\\
+
+		$\omega_4$  & 14.34 & 0.53 & $3.71\%$ &  1.62\\
+
+		$\omega_5$  & 4.30 & -4.89 & $-113.7\%$ &  5.32\\
+
+		$\omega_6$  &  34.38 & -3.07 & $-8.9\%$ &   3.78\\
+
+		$\omega_7$  & -0.85 & 1.08 & $-126.4 \%$ & 1.41\\
+
+		$\omega_8$  &  8.81 & -3.91 & $-44.4 \%$ & 5.88\\
+
+		$\sigma_{\tau}$ & 0.1 & 0.015& $15\%$ & 0.10\\
+
+		$\sigma_{\nu}$ & 0.3 & 0.007 & $7 \%$ & 0.04\\
+		\hline
+	\end{tabular}
+	\caption{Measures of estimation performance in the rectangular domain}
+	\label{table: estimation performance rect}
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|}
+		\hline
+		Coefficient  & True value & Biais & RMSE  \\
+		\hline
+		$\log(\tau_0)$  &  0 &  &  \\
+
+		$\log(\nu_0)$  &  1.38 &  &  \\
+
+		$\omega_0$  & -0.34 &  &  \\
+
+		$\omega_1$  & 14.17  &  &  \\
+
+		$\omega_2$  & -2.57 &  &  \\
+
+		$\omega_3$  & 11.14 &  &  \\
+
+		$\omega_4$  & 36.48  &  &  \\
+
+		$\omega_5$  & 11.73 &  &  \\
+
+		$\omega_6$  & 53.69 &  &  \\
+
+		$\omega_7$  & 9.23 &  &  \\
+
+		$\omega_8$  & 10.19 &  &  \\
+
+		$\sigma_{\tau}$ & 0.1 & & \\
+
+		$\sigma_{\nu}$ & 0.3 & & \\
+		\hline
+	\end{tabular}
+	\caption{Measures of estimation performance in the Scoresby Sound domain}
+	\label{table: estimation performance fjords}
+\end{table}
+
+\section{Application to narwhals behavioral response}
+
+In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
 
@@
 
-\section{Precisions about the statistical models}
-\label{section: precisions about statistical models}
-In the baseline mixed effect model, the explicit model matrices are
-
-\[X_{pre}= \begin{bNiceArray}{ccc|ccc}[margin]
-	1 & 0 & 0 & \Block{3-3}<\Large>{0} & &  \\
-	\vdots & \vdots & \vdots & & &  \\
-	1 & 0 & 0 & & & \\
-	\hline
-	0 & 1 & 0 &  \Block{3-3}<\Large>{0} & & \\
-	\vdots & \vdots & \vdots & & &  \\
-	0 & 1 & 0 & & & \\
-	\hline
-	0 & 0 & 1 &  \Block{3-3}<\Large>{M_{pre}} & & \\
-	\vdots & \vdots & \vdots & & &  \\
-	0 & 0 & 1 & & & \\
-\end{bNiceArray} \quad 
-Z_{pre} = \begin{pmatrix}
-	J_{1,n_1} \\
-	\vdots \\
-	J_{6,n_6} \\
-	\vdots \\
-	J_{13,n_1} \\
-	\vdots \\
-	J_{18,n_6}
-\end{pmatrix} 
-\]
-where $M_{pre}$ is the matrix with elements  $\psi_k(E^{(i)}_{shore}(t^{(i)}_j))$ on column $k$ and row associated to observation $j$ for narwhal $i$, and $J_{l,m}$ is the matrix with $m$ rows and $18$ columns that has zero everywhere except on column $l$, where there are only ones.
-
 \end{document}

### 2024-05-08 Latex/main.tex
update latex

@@
 In \cite{heide-jorgensen_behavioral_2021}, the positions are classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second, and the dynamic of the states is modeled with a Markov chain, assuming that the current state only depends on the last one in time.
-An exposure covariate defined as the inverse of the distance to the sound source is then used to model the transition probabilities of the Markov Chain, and the probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases.
+An exposure covariate defined as the inverse of the distance to the sound source is then used to model the transition probabilities of the Markov Chain.
+The probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases.
 Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model.
@@
 It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
-The following formulation of the equations represents an extension of the CVM \cite{johnson_continuoustime_2008}, which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
+The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
 \begin{equation} \left\{
@@
 The case $\omega=0$ corresponds to a standard CVM.
-In most applications, this model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
-
-Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
+In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
+Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
 Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.
@@
 Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
-To make this possible, we formulate a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}.
+To make this possible, we formulate a state-space model by binding the velocity and location in one hidden state vector, following \cite{johnson_continuoustime_2008}.
 Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\
@@
 In Section 2, we give an overview of the narwhal data available for the analysis.
-The diffusion models are discussed in Section 3. Inference and simulation study results are explained in section 4.
-Then, our models are applied to analyse a behavioral response of the narwhals in section 5. The last section is dedicated to a discussion about these results, and potential extension of the models.
+Then, the diffusion models are discussed in Section 3.
+Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effect.
+Simulation study results are shown in section 5.
+Eventually, our models are applied to analyse a behavioral response of the narwhals in section 6.
 
@@
 \subsection{Notations and definition of the covariates}
+\label{subsection: covariates}
 
 For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.
-These are noisy observations of the underlying unobserved true positions $x_{ij}$.\\
+We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}.
+These points are noisy observations of the underlying unobserved true positions $x_{ij}$.\\
 
@@
 The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$.
-Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$.
-The resulting values range from $0$ to $7.5$ km.
+Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}=d(x_{ij},p_{ij})$.
 About $9\%$ of the GPS measurements in the dataset turned out to be on land.
-This is primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-For constrained movement models, these points are ignored.
-Then, exposure to the shore is defined with the following formula
+After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
+We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 downloaded from Google Earth Engine API.
+After modification,  $1.4 \%$ of the observations remained on land.
+Distance to shore values range from $0$ to $7.6$ km.
+We define an exposure to the shore covariate with the following formula
 \begin{equation}
@@
 	\begin{array}{ll}
-		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{ref} \\
-		E^{shore}_{ij}=\frac{1}{D_{shore}(t_j)} & \mbox{if } D^{shore}_{ij} \leq D_{ref}
+		E^{shore}_{ij}=0 & \mbox{if } D_{min} <D^{shore}_{ij}<D_{max} \\
+		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D^{shore}_{ij} \leq D_{max}
 	\end{array}
@@
 \end{equation}
-where the reference threshold $D_{ref}$ is chosen equal to $0.5$ km.
+where the reference thresholds are $D_{min}=58$ m, which is the $5\%$ quantile of observed distances, and $D_{max}=335$ m, which is the $25 \%$ quantile of observed distance.
+Thus we only consider the effect of the shore when the narwhals are close enough to it.
 We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
-The angle between the narwhal's heading and the shoreline, denoted $\Theta_{ij}$, is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-p_{ij}$.
+The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is defined as the angle between the vectors $x_{ij}-p_{ij}$ and the true velocity $v_{ij}$.
+It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-\hat{p}_{ij}$.
 This is illustrated in figure~\ref{fig: illustration nearest shore points}.
@@
 		\node[left] at (0,0) {$\partial \mathcal{D}$};
-		\node[left] at (0,2) {$p_{ij}$};
+		\node[left] at (0,2) {$\hat{p}_{ij}$};
 		\node[left] at (1.2,2.8) {$\hat{v}_{ij}$};
@@
 
-The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
+The distance from the ship denoted $D^{ship}_{ij}$, is estimated as the distance between the GPS ship and narwhals's positions.
+We denote $\hat{D}^{ship}_{ij}$ this estimation.
+The values are comprised between $2.68$ and $63.8$ km.
 Exposure to ship covariate for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
+Its estimate $\hat{E}^{ship}$ is the inverse of $\hat{D}^{ship}$.
 $D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
@@
 The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
-We denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
+We keep the notation $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
 \begin{equation} \left\{
@@
 
-with $A(X(t),V(t))=\begin{pmatrix} 
-	\frac{1}{\tau} & -\omega(t) \\
-	\omega(t) & \frac{1}{\tau}
-\end{pmatrix}$
-where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
-$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
+with 
+\begin{equation} 
+		A(X(t),V(t))=\begin{pmatrix} 
+			\frac{1}{\tau} & -\omega(X(t),V(t)) \\
+			\omega(X(t),V(t)) & \frac{1}{\tau}
+		\end{pmatrix}
+	\label{eq: CRCVM matrix A}
+\end{equation}
+where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(X(t)),\Theta(X(t),V(t)))$, $D^{shore}(X(t))$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
+and $\Theta(X(t),V(t))$ is the angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}.Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
 
@@
 This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
-Any smooth function that meets minimum requirements could be used instead.
-The general model is very flexible, and different species-specific behaviours could be described.
-An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
+Any smooth function that meets minimum requirements could be used instead.An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
 Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
 In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
-A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}.
+A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}.
 The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
@@
 
-\label{section: simulation study}
-%samples and estimation
+ The general model is very flexible, and different species-specific behaviours could be described.
+The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn.
+We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles.
+Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question.
+Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination.
+Again, this behavior can be statistically assessed from the observations.
+Examples of trajectories for different assumptions on the smooth parameters $\tau$ and $\nu$ are shown in Appendix.
+
 %discussion about tensor splines vs additive splines
 
-\section{Inference}
+\section{Statistical model}
 
@@
 \begin{equation}
-	\left\{
+	\forall i \in \{1,\cdots,N\}, \left\{
 	\begin{array}{l}
-		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \\
+		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
+		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
 		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij}
+		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
 	\end{array}
@@
 
-The vector $\eta_{ij}=\begin{pmatrix} \xi_{ij,1} & \xi_{ij,2} & \zeta_{ij,1} & \zeta_{ij,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero.
-Its covariance matrix $Q_{ij}$ can be computed explicitly.
+The vector $\eta_{ij}$ with covariance matrix 
 \begin{equation}
@@
 \end{equation}
-where $M_{ij}$, $N_{ij}$and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
+where $M_{ij}$, $N_{ij}$ and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
 \[M_{ij}=\frac{\sigma_i^2}{C_{ij}}\left( \Delta-2 \frac{\omega_{ij} \sin(\omega_{ij} \Delta_{ij})-\frac{1}{\tau_{i}} \cos(\omega_{ij} \Delta_{ij})}{C_{ij} } \exp\left( -\frac{\Delta_{ij}}{\tau_{ij}} \right) +\frac{\tau_{i}}{2} \left( \frac{\omega_{ij}^2-\frac{3}{\tau_{ij}^2}}{C_{ij}}-\exp\left( -\frac{2\Delta_{ij}}{\tau_i}\right)\right) \right)I_2\]
@@
 
-For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_0)}^{(k)}$, $\widehat{\log(\nu_0)}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimator.
+For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimator.
 The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
@@
 		\hline
-		$\log(\tau_0)$  & 0& 0.08 &  & 0.16\\
-
-		$\log(\nu_0)$  & 1.38 & -0.07 & $-5.4 \%$ & 0.10 \\
+		$\log(\tau_{pre})$  & 0& 0.08 &  & 0.16\\
+
+		$\log(\nu_{pre})$  & 1.38 & -0.07 & $-5.4 \%$ & 0.10 \\
 
@@
 
-		$\sigma_{\tau}$ & 0.1 & 0.015& $15\%$ & 0.10\\
-
-		$\sigma_{\nu}$ & 0.3 & 0.007 & $7 \%$ & 0.04\\
+		$\sigma_{\tau,pre}$ & 0.1 & 0.015& $15\%$ & 0.10\\
+
+		$\sigma_{\nu,pre}$ & 0.3 & 0.007 & $7 \%$ & 0.04\\
 		\hline
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|}
-		\hline
-		Coefficient  & True value & Biais & RMSE  \\
-		\hline
-		$\log(\tau_0)$  &  0 &  &  \\
-
-		$\log(\nu_0)$  &  1.38 &  &  \\
-
-		$\omega_0$  & -0.34 &  &  \\
-
-		$\omega_1$  & 14.17  &  &  \\
-
-		$\omega_2$  & -2.57 &  &  \\
-
-		$\omega_3$  & 11.14 &  &  \\
-
-		$\omega_4$  & 36.48  &  &  \\
-
-		$\omega_5$  & 11.73 &  &  \\
-
-		$\omega_6$  & 53.69 &  &  \\
-
-		$\omega_7$  & 9.23 &  &  \\
-
-		$\omega_8$  & 10.19 &  &  \\
-
-		$\sigma_{\tau}$ & 0.1 & & \\
-
-		$\sigma_{\nu}$ & 0.3 & & \\
+	\begin{tabular}{|c|c|c|c|c|}
+		\hline
+		Coefficient  & True value & Absolute biais & Relative biais & RMSE  \\
+		\hline
+		$\log(\tau_{pre})$  &  0 &  & & \\
+
+		$\log(\nu_{pre})$  &  1.38 &  & & \\
+
+		$\omega_0$  & -0.34 &  & & \\
+
+		$\omega_1$  & 14.17  &  & &  \\
+
+		$\omega_2$  & -2.57 &  & & \\
+
+		$\omega_3$  & 11.14 &  &  &\\
+
+		$\omega_4$  & 36.48  &  & & \\
+
+		$\omega_5$  & 11.73 &  & & \\
+
+		$\omega_6$  & 53.69 &  & & \\
+
+		$\omega_7$  & 9.23 &  & & \\
+
+		$\omega_8$  & 10.19 &  &  &\\
+
+		$\sigma_{\tau,pre}$ & 0.1 & & &\\
+
+		$\sigma_{\nu,pre}$ & 0.3 & & &\\
 		\hline

### 2024-05-08 R/preprocessing/data_preprocessing.Rmd
changes in preprocessing for land

@@
 
-#get land polygons
-modified_land_utm<-st_read(file.path(greenland_data_path,"modified_land_utm.shp"))
-modified_land_utm <- st_transform(modified_land_utm, crs = "+init=EPSG:32626 +units=km")
-
 #scoresby sound google
@@
   coord_sf(datum=st_crs(32626))
-plot_modified_land=ggplot()+geom_sf(data=modified_land_utm$geometry,fill="grey",col="blue")+
-  coord_sf(datum=st_crs(32626))
+
 plot_land_google=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey",col="blue")+
@@
 summary(dataprep$DistanceShore)
+quantile(res_land$min_distance,probs=c(0.05,0.95))
 ```
@@
 
-cat("Percentage of points on land with distance from modified : \n",
-    length(which(res_modified_land$min_distances==0))/length(dataprep$time)*100)
 ```

### 2024-05-12 R/simulation_study_CRCVM/simulation_study_fjords.Rmd
reset simulation study

--- 2024-05-08 R/simulation_study/fjords/simulation_study_fjords.Rmd
@@
 
-ftau_bump=function(cov_data,epsilon=pi/10,alpha=pi/5,D0=1,tau0=0.8,tau1=2) {
+ftau_bump=function(cov_data,epsilon=pi/8,alpha=pi/5,tau0=0.2,tau1=2) {
     Dshore=cov_data$DistanceShore

### 2024-05-12 R/simulation_study_CRCVM/simulation_study_rect.Rmd
reset simulation study

--- 2024-05-05 R/simulation_study/rect/simulation_study_rect.Rmd
@@
 #some auxiliary functions
-source("CVM_functions.R")
+par_dir=dirname(getwd())
+source(file.path(par_dir,"CVM_functions.R"))
 ```
@@
 #degree of freedom in tensor splines
-SP_DF=3
+SP_DF=4
 
@@
 #time step
-hf=1/60
+hf=1/120
 
@@
 
-Dshore=seq(from=0.3,to=2,length.out=30)
+Dshore=seq(from=0.3,to=4,length.out=30)
 theta=seq(from=-pi,to=pi,length.out=30)
@@
 
-### Generate samples
+```{r generate batch of hf low errors samples,cache=TRUE}
 
-```{r illustrative sample for the paper}
-
-res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,v0=v0,x0=x0[1,],times=seq(0,2*24,by=1/60),land=domain,verbose=FALSE)
-
-data_sim=res$sim
-data_sim$ID=factor(rep(1,length(data_sim$y1)))
-data_shore=res$shore[,c("p1","p2")]
-data_sim=cbind(data_sim,data_shore)
-
-data_sim=data_sim[seq(1,length(data_sim$time),by=5),]
-
-plot_illust_rect=ggplot()+geom_sf(data=domain$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_sim,aes(p1,p2),shape=5,size=0.1,col="red")+
-  geom_point(data=data_sim,mapping=aes(y1,y2),size=0.1)+
-  geom_path(data=data_sim,mapping=aes(y1,y2),size=0.1)+
-  geom_point(data = data_sim%>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
-  xlab("x") + ylab("y")
-
-plot_illust_rect
-
-ggsave(filename="illustrative_sample_rect.png",plot=plot_illust_rect,width=10,height=5)
-```
-
-```{r parallel generate batches of hf low errors samples,cache=TRUE}
-
-#setup parallel backend to use many processors
-cores=detectCores()
-cl <- makeCluster(cores[1]-1) #not to overload your computer
-registerDoParallel(cl)
-
-data_rect_hf_ne <- foreach(k=1:N_BATCH,.packages=c("progress","MASS","sf","foreach","mgcv")) %dopar% {
-  data=foreach (i=1:N_ID,.combine='rbind') %do% {
+data_rect_hf_ne =foreach (i=1:N_ID,.combine='rbind') %do% {
 
@@
 }
-  data 
-}
 
-#stop cluster
-stopCluster(cl)
-
-names(data_rect_hf_ne)=paste("BATCH",1:N_BATCH,sep="")
-```
-
-### percentage of samples that reached land
-
-```{r percentage of low error samples that reached land}
-count=0
-for (k in 1:length(names(data_rect_hf_ne))) {
-
-  data=data_rect_hf_ne[[paste("BATCH",k,sep="")]]
-  for (id in unique(data$ID)) {
-    sub_data=data[data$ID==id,]
-    if (nrow(sub_data) < n_hf) {
-      count=count+1
-      cat(paste("BATCH",k,sep=""),":",id,"\n",sep=" ")
-    }
-  }
-}
-cat(count/(N_BATCH*N_ID)*100,"percent of the samples reached land")
 ```
@@
 plot_samples_rect_hf_ne=ggplot()+geom_sf(data=domain$geometry,fill="grey")+coord_sf(datum=st_crs(32626))+
-  geom_point(data=data_rect_hf_ne$BATCH1,aes(p1,p2),shape=5,size=0.1,col="red")+
-  geom_point(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
-  geom_path(data=data_rect_hf_ne$BATCH1,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
-  geom_point(data = data_rect_hf_ne$BATCH1%>% filter(!duplicated(ID)),
+  geom_point(data=data_rect_hf_ne,aes(p1,p2),shape=5,size=0.1,col="red")+
+  geom_point(data=data_rect_hf_ne,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
+  geom_path(data=data_rect_hf_ne,mapping=aes(y1,y2,color=ID,shape=ID),size=0.1)+
+  geom_point(data = data_rect_hf_ne%>% filter(!duplicated(ID)),
              aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+

### 2024-05-12 R/simulation_study_CRCVM/CVM_functions.R

@@ -1,0 +1,1387 @@
+
+#SCRIPT CONTAINING FUNCTIONS TO SIMULATE AND COMPUTE LIKELIHOOD FOR DIFFERENT CVM MODELS 
+
+library(Matrix)
+library(MASS)
+library(progress)
+
+
+
+
+#' @description Compute likelihood for a 2D CVM with constant parameters and Kalman filter
+#' 
+#' @param obs data frame with at least columns matching with the names in the response argument 
+#' and one "time" column
+#' @param response vector of strings corresponding to the names of the response
+#' @param log_beta log of autocorrelation time scale parameter
+#' @param log_sigma log of of volatility parameter
+#' @param mu1 first component of mean velocity: scalar with size length(times)-1nent of the drift parameter,
+#'  scalar or vector of scalar with size length(times)-1
+#' @param mu2 second component of mean velocity: scalar with size length(times)-1nent of the drift parameter, 
+#' scalar or vector of scalar with size length(times)-1
+#' @param v0 initial velocity
+#' @param x0 initial location
+#' @param log_sigma_obs if not NULL, gaussian noise with standard deviation exp(log_sigma_obs) is added to the sampled locations 
+#' @return scalar corresponding to the likelihood of the data
+
+nllkCVM=function(obs,response,log_beta,log_sigma,mu1,mu2,v0,x0,log_sigma_obs) {
+  
+  
+  n=length(obs[,response[1]])
+  
+  beta=exp(log_beta)
+  sigma=exp(log_sigma)
+  mu=matrix(c(mu1,mu2),ncol=1,nrow=2,byrow=TRUE)
+  
+  times=obs$time
+  
+  
+  #initialization
+  alphax=c(x0[1],v0[1]-mu1)
+  alphay=c(x0[2],v0[2]-mu2)
+  alpha=c(alphax,alphay)
+  Rx=matrix(rep(0,4),ncol=2)
+  Ry=matrix(rep(0,4),ncol=2)
+  R=bdiag(Rx,Ry)
+  #log likelihood
+  llk=-n*log(2*pi)
+  
+  for (i in (2:n)) {
+    #time step
+    delta=times[i]-times[i-1]
+    
+    #link matrix for successive states
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore33.R

@@ -1,0 +1,205 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 33
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore40.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 40
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore47.R

@@ -1,0 +1,205 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 47
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore55.R

@@ -1,0 +1,205 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 55
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore65.R

@@ -1,0 +1,205 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 65
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore45.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 45
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore2.R

@@ -1,0 +1,205 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 2
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore20.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 20
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore51.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 51
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore68.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 68
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore71.R

@@ -1,0 +1,207 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(par_dir,"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 71
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data_rect[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-12 R/preprocessing/data_preprocessing.Rmd
update preprocessing

@@
 st_crs(points_sf)="+init=EPSG:32626 +units=km"
-st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"))
+st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
 ```
@@
 
-## Extract coastline and land polygons
+## Extract land polygons to define the border of the domain
 
@@
 
-#scoresby sound google
-land_google_utm<-st_read(file.path(greenland_data_path,"land_google_utm.shp"))
-st_crs(land_google_utm)="+init=EPSG:32626"
 ```
@@
   coord_sf(datum=st_crs(32626))
-
-plot_land_google=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey",col="blue")+
-  coord_sf(datum=st_crs(32626))
 ```
@@
 ggplotly(plot_land)
-```
-
-```{r display coastline , fig.cap="Coastline polygons and linestrings"}
-ggplotly(plot_modified_land)
-```
-
-```{r}
-ggplotly(plot_land_google)
 ```
@@
 quantile(res_land$min_distance,probs=c(0.05,0.95))
+D_low=quantile(res_land$min_distance,probs=0.05)
+D_up=median(res_land$min_distance)
 ```
@@
 ```{r check original points on land}
-check_plot1=ggplot()+geom_sf(data=land_google_utm$geometry,fill="grey")+
+check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
-  geom_point(data=dataprep,aes(x,y,col=(DistanceShore<0.05)),size=0.2)+
+  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
     scale_color_viridis_d()
@@
 
-df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<0.2 & res_land$min_distances>0  ])
+df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<1 & res_land$min_distances>0  ])
 histo=ggplot(df, aes(x =DistanceShore)) +
@@
 the shore in reality.
-Let's choose arbitrarily a threshold of 50 metres.
 
@@
 
-The UTM coordinates for the coastlines are in meters, so that we need to multiply by 1000 when plotting the nearest shore points.
-
 ```{r individual tracks}
@@
   sub_data=dataprep[dataprep$ID==id,]
-  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
+  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
@@
 ```{r}
-pExpShore=ggplot(dataprep[dataprep$DistanceShore>0.05,],aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+df=dataprep
+df[df$DistanceShore<D_low,"ExpShore"]=1/D_low
+df[df$DistanceShore>D_up,"ExpShore"]=0
+
+pExpShore=ggplot(df,aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
   theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")

### 2024-05-12 Latex/main.tex
latest version of the paper

@@
 After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 downloaded from Google Earth Engine API.
-After modification,  $1.4 \%$ of the observations remained on land.
+We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API.
+After adjustements,  approximately $1 \%$ of the observations remained on land.
 Distance to shore values range from $0$ to $7.6$ km.
@@
 	\begin{array}{ll}
-		E^{shore}_{ij}=0 & \mbox{if } D_{min} <D^{shore}_{ij}<D_{max} \\
-		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D^{shore}_{ij} \leq D_{max}
+		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
+		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
+		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
 	\end{array}
@@
 \end{equation}
-where the reference thresholds are $D_{min}=58$ m, which is the $5\%$ quantile of observed distances, and $D_{max}=335$ m, which is the $25 \%$ quantile of observed distance.
+where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances.
 Thus we only consider the effect of the shore when the narwhals are close enough to it.
@@
 The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is defined as the angle between the vectors $x_{ij}-p_{ij}$ and the true velocity $v_{ij}$.
-It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-\hat{p}_{ij}$.
+It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-\hat{p}_{ij}$.
 This is illustrated in figure~\ref{fig: illustration nearest shore points}.
@@
 But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
-Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula.
+Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
 \\
@@
 \begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
+	\label{eq: transition RACVM velocity}
 \end{equation}
@@
 	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
+	\label{eq: transition RACVM location}
 \end{equation}
@@
 \]
-The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
-The two components of the location and velocity processes are independent.
-and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
-Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm.
+The location process is not markovian, but the whole process $\begin{pmatrix}X(t) &V(t) \end{pmatrix}^\top$ is.
+The two components of the location and velocity processes are independent, and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
+Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}.
 Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\
@@
 		\centering
-		\includegraphics[scale=0.3]{images/illustrative_sample_standard.png}
+		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
 		\caption{}
@@
 		\centering
-		\includegraphics[scale=0.3]{images/illustrative_sample_fjords.png}
+		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
 		\caption{}
@@
 The red cross indicates the starting point of the process.
-The samples were obtained from the exact transition densities.
-(b) Samples of a CRCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
-Position are sampled from constant time step $h= 1$ min and then doownsampled to $5$ min.The samples were obtained from approximate transition densities  with constant covariates on each time step.
-}
+The samples were obtained from the exact transition densities in equations~\ref{eq: transition RACVM velocity} and~\ref{eq: transition RACVM location}.
+(b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
+Position are sampled from constant time step $h= 1$ min and then downsampled to $5$ min.
+The samples were obtained from approximate transition densities  with constant covariates on each time step.}
 	\label{fig: sample RACVM}
@@
 \end{equation}
-where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(X(t)),\Theta(X(t),V(t)))$, $D^{shore}(X(t))$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
-and $\Theta(X(t),V(t))$ is the angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}.Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
+where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
+and $\Theta(t)$ is the angle between the velocity $V(t)$ and the shoreline as defined in subsection ~\ref{subsection: covariates}.
+Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
 
@@
 This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal.
-Any smooth function that meets minimum requirements could be used instead.An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
-Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
-In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
-A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}.
+Any smooth function that meets minimum requirements could be used instead.
+An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
+Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
+In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
+A sample of a CRCVM with smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}.
 The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
@@
 	\centering
-		\centering
-		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
-		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
-		\label{fig: trueomega}
-\end{figure}
-
-\begin{figure}[H]
-		\centering
-		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
-		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
-		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
-	\label{fig: splineomega}
-\end{figure}
-
- The general model is very flexible, and different species-specific behaviours could be described.
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/crcvm/smooth_omega1.png}
+		\caption{}
+		\label{fig: smoothomega1}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/crcvm/smooth_omega2.png}
+		\caption{}
+	\label{fig: smoothomega2}
+	\end{subfigure}
+
+	\begin{subfigure}{0.48\textwidth}
+	\centering
+	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega1.png}
+	\caption{}
+	\label{fig: smoothomegasplines1}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+	\centering
+	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega2.png}
+	\caption{}
+	\label{fig: smoothomegasplines2}
+	\end{subfigure}
+	\caption{Examples of smooth functions $f_{\omega}$.
+Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.
+(a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega}.
+(c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivartiate cubic spline basis functions.}
+
+\end{figure}
+
+ This general model is very flexible, and different species-specific behaviours could be described.
 The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn.
-We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles.
+We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, observed distances and angles.
 Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question.
@@
 Again, this behavior can be statistically assessed from the observations.
-Examples of trajectories for different assumptions on the smooth parameters $\tau$ and $\nu$ are shown in Appendix.
+
+ More precisely, to model persistent movement along the shore line, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
+ \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] where $f_{\tau}$ is a smooth function from $[-\pi,\pi]$ to $\R$.
+ The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain.
+An example of such a function is given in figure~\ref{fig:smoothtaubump}.
+ \begin{figure}[H]
+ 	\centering
+ 	\includegraphics[scale=0.3]{images/crcvm/smooth_tau_bump}
+ 	\caption{Example of smooth function $f_{\tau}$ that can represent persistent movement along the boundary of the domain.}
+ 	\label{fig:smoothtaubump}
+ \end{figure}
+
+ We show side by side the samples obtained within different domain for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.
+
+ \begin{figure}[H]
+ 	\centering
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_standard}
+ 		\label{fig:illustrativesamplecircstandard}
+ 	\end{subfigure}
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_pers}
+ 		\label{fig:illustrativesamplecircpers}
+ 	\end{subfigure}
+
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_standard}
+ 		\label{fig:illustrativesamplerectstandard}
+ 	\end{subfigure}
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_pers}
+ 		\label{fig:illustrativesamplerectpers}
+ 	\end{subfigure}
+
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard}
+ 		\label{fig:illustrativesamplefjordsstandard}
+ 	\end{subfigure}	
+ 	\begin{subfigure}{0.48\textwidth}
+ 		\centering
+ 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_pers}
+ 		\label{fig:illustrativesamplefjordspers}
+ 	\end{subfigure}
+ 	\caption{Left: samples from a CRCVM with smooth $\omega$ and constant $\tau=1$ h in three different domains.
+Right: samples from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the shoreline.}
+ 	\label{fig: crcvm examples}
+ \end{figure}
 
@@
 
-Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion .
+Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion.
+The complete statistical model for the baseline estimations is the following:
 
@@
 		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
-		\log(\tau_i)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
-		\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
+		\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)} \\
+		\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}  \\
 		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
-		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\hat{\Theta}_i(t))\\
+		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
 		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
@@
 
-$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
+The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$, $\omega_{0}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
 are the individual parameters.
@@
 1.8$ for the construction of tensor splines basis functions.
-
-%Add plot of the spline basis functions \\
 The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$.
-It is needed to choose them before hand.
-One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.\\
+It is needed to choose them before hand.\\
 
@@
 		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
-		\log(\tau_{ij})=\log(\tau_{0,post}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
-		\log(\nu_{ij})=\log(\nu_{0,post}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
-		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
+		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
+		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
+		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
 		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
@@
 
-Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters.
+Again, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters.
 The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
@@
 This is the approach we use here.
-In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$.\\
+In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
+\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(\hat{E}^{shore}_{ij},\hat{\Theta}_{ij})\]
  Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
@@
 	\omega_{ij} & \frac{1}{\tau_{i}}
-\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$.
+\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.
 \begin{comment}
@@
 and $C_{ij}=\frac{1}{\tau_i^2}+\omega_{ij}^2$, $\sigma_{i}=\frac{2\nu_i}{\sqrt{\pi \tau_i}}$.
-Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
-This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter.
-This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$.
-Therefore, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models.
-
-\section{Simulation study for the baseline model}
-\label{section: simulation constrained motion}
-In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.
-We considered observation of a baseline model as described in equation~\ref{eq: exact baseline model}\\.
-
-We simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in a rectangular domain with sides of length $8$ km from a CRCVM.
-This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
-We fix $\sigma_{obs}=5$ m to represent low measurement errors.
-We choose $L=8$ for the degree of freedom of the bivariate splines of $D^{shore}$ and $\Theta$.
-The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{pre}$ and $\nu_{pre}$ are fixed respectively to $1$ h and $4$ km/h.
-The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance rect}.
-For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
-WE kept the same $6$ initial positions for all the batches of trajectories.
-Toobtain teh samples of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
-\begin{equation}
-	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
-\end{equation}
-where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
-The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
-None of the $600$ simulations reached the boundary.
-\\
-
-We also simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ with time step $h=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
-The parameters $\sigma_{obs}$,$L$,$\sigma_{\tau,pre}$,  $\sigma_{\nu,pre}$ , $\tau_{pre}$,$\nu_{pre}$ are kept the same as in the recangular case, and only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are changed to adapt to this specific fjords domain.
-The exact values of the spline coefficients are found in Table~\ref{table: estimation performance fjords}.
-Again, for each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.
-None of the $M\times N=600$ trajectories reached land.
-
-\begin{comment}Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
-For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
-Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-\end{comment}
-
-Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
+Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.\\
+
+Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
 The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
@@
 \\
-
-$100$ batches of noisy samples obtained as described in equation~\ref{eq: rect simulation parameters} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system.
+This is the technique is already available for the CVM, which corresponds to the special case $\omega=0$.
+Here, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models~\ref{eq: exact baseline model},~\ref{eq: exact response model}.
+
+\section{Simulation study for the baseline model}
+\label{section: simulation constrained motion}
+In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain.
+We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}\\.
+
+We simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in a rectangular domain with sides of length $8$ km with the statistical model~\ref{eq: exact baseline model}.
+This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
+We fix $\sigma_{obs}=5$ m to represent low measurement errors.
+We choose $L=8$ for the degree of freedom of the bivariate splines of $E^{shore}$ and $\Theta$.
+The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{0}$ and $\nu_{0}$ are fixed respectively to $1$ h and $4$ km/h.
+The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance rect}.
+For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
+We kept the same $6$ initial positions for all the batches of trajectories.
+To obtain the trajectories of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
+\begin{equation}
+	\omega_{ij}=f_{\omega}(E^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
+\end{equation}
+where $E^{shore}_{ij}=E^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
+The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
+None of the $600$ simulations reached the boundary.
+\\
+
+We also simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
+The parameters $\sigma_{obs}$, $L$, $\sigma_{\tau,pre}$, $\sigma_{\nu,pre}$, $\tau_{0}$,$\nu_{0}$ are kept the same as in the rectangular case, and only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are changed to adapt to this specific fjords domain.
+The exact values of the spline coefficients are found in Table~\ref{table: estimation performance fjords}.
+Again, for each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled within the water area, with the requirement that it should be at least $500$ metres away from the shore.
+None of the $M\times N=600$ trajectories reached land.\\
+
+\begin{comment}Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
+For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
+Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
+\end{comment}
+
+In the rectangular and fjords domains, these $100$ batches of noisy samples are used as data for estimation.
 We thus obtain $100$ parameters estimate in the two cases.
-The covariates $D^{shore}$ and $\Theta$ are computed from the observed noisy positions.
+The covariates $E^{shore}$ and $\Theta$ are computed from the observed noisy positions.
 It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging.
 The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom is set to $3$ for each covariate to match the simulated data from equation~\ref{eq: rect simulation parameters}.
+The marginal spline degree of freedom is set to $3$ for each covariate to match the fixed degree of freedom $L=8$.
 This corresponds to one single knot in each dimension.
-Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
+Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau,pre}$ and $\sigma_{\nu,pre}$.
 Measurement error $\sigma_{obs}$ is supposed to be known.
-We fix initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h, $\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectoriesin the rectangular and the fjords domain.
+We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau,pre}=1$ h, $\sigma_{\nu,pre}=1$ km/h and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories in the rectangular and the fjords domain, with confidence intervals obtained from the observed Fisher information matrix.
 The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
@@
 
-For each batch of $6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimator.
+For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimates.
+Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.
 The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
@@
 		\hline
-		Coefficient  & True value & Absolute biais  & Relative biais & RMSE\\
+		Coefficient  & True value & Absolute biais  & Relative biais ($\%$) & RMSE\\
 		\hline
@@
 
-		$\log(\nu_{pre})$  & 1.38 & -0.07 & $-5.4 \%$ & 0.10 \\
-
-		$\omega_0$  &  0.20  & -0.11 & $-58\%$ & 0.30 \\
-
-		$\omega_1$  & 7.41 & -0.4 & $-5.4 \%$ & 1.11 \\
-
-		$\omega_2$  & -2.78 & -1.87 &  $67.3\%$ & 4.30\\
-
-		$\omega_3$  & 4.05 &  0.30 & $7.4\%$ & 0.96\\
-
-		$\omega_4$  & 14.34 & 0.53 & $3.71\%$ &  1.62\\
-
-		$\omega_5$  & 4.30 & -4.89 & $-113.7\%$ &  5.32\\
-
-		$\omega_6$  &  34.38 & -3.07 & $-8.9\%$ &   3.78\\
-
-		$\omega_7$  & -0.85 & 1.08 & $-126.4 \%$ & 1.41\\
-
-		$\omega_8$  &  8.81 & -3.91 & $-44.4 \%$ & 5.88\\
-
-		$\sigma_{\tau,pre}$ & 0.1 & 0.015& $15\%$ & 0.10\\
-
-		$\sigma_{\nu,pre}$ & 0.3 & 0.007 & $7 \%$ & 0.04\\
+		$\log(\nu_{pre})$  & 1.38 & -0.07 & $-5.4 $ & 0.10 \\
+
+		$\omega_0$  &  0.20  & -0.11 & $-58$ & 0.30 \\
+
+		$\omega_1$  & 7.41 & -0.4 & $-5.4 $ & 1.11 \\
+
+		$\omega_2$  & -2.78 & -1.87 &  $67.3$ & 4.30\\
+
+		$\omega_3$  & 4.05 &  0.30 & $7.4$ & 0.96\\
+
+		$\omega_4$  & 14.34 & 0.53 & $3.71$ &  1.62\\
+
+		$\omega_5$  & 4.30 & -4.89 & $-113.7$ &  5.32\\
+
+		$\omega_6$  &  34.38 & -3.07 & $-8.9$ &   3.78\\
+
+		$\omega_7$  & -0.85 & 1.08 & $-126.4 $ & 1.41\\
+
+		$\omega_8$  &  8.81 & -3.91 & $-44.4 $ & 5.88\\
+
+		$\sigma_{\tau,pre}$ & 0.1 & 0.015& $15$ & 0.10\\
+
+		$\sigma_{\nu,pre}$ & 0.3 & 0.007 & $7 $ & 0.04\\
 		\hline
@@
 		\hline
-		Coefficient  & True value & Absolute biais & Relative biais & RMSE  \\
-		\hline
-		$\log(\tau_{pre})$  &  0 &  & & \\
-
-		$\log(\nu_{pre})$  &  1.38 &  & & \\
-
-		$\omega_0$  & -0.34 &  & & \\
-
-		$\omega_1$  & 14.17  &  & &  \\
-
-		$\omega_2$  & -2.57 &  & & \\
-
-		$\omega_3$  & 11.14 &  &  &\\
-
-		$\omega_4$  & 36.48  &  & & \\
-
-		$\omega_5$  & 11.73 &  & & \\
-
-		$\omega_6$  & 53.69 &  & & \\
-
-		$\omega_7$  & 9.23 &  & & \\
-
-		$\omega_8$  & 10.19 &  &  &\\
-
-		$\sigma_{\tau,pre}$ & 0.1 & & &\\
-
-		$\sigma_{\nu,pre}$ & 0.3 & & &\\
+		Coefficient  & True value  & Absolute biais & Relative biais ($\%$)& RMSE \\
+		\hline
+		$\log(\tau_{pre})$   & 0     & -0.16 &         & 0.2  \\
+		$\log(\nu_{pre}$)     & 1.39  & -0.02 & -1.77     & 0.07 \\
+		$\omega_0$   & 0.06  & -0.09 & -148.81   & 0.32 \\
+		$\omega_1$ & 11.04 & -1.87 & -16.94    & 2.66 \\
+		$\omega_2$ & -3.37 & 1.69  & -50.04    & 2.73 \\
+		$\omega_3$ & 7.64  & -2.01 & -26.31    & 2.41 \\
+		$\omega_4$ & 23.24 & -2.47 & -10.63    & 4.95 \\
+		$\omega_5$ & 7.37  & 1.66  & 22.5      & 5.84 \\
+		$\omega_6$ & 41.1  & -4.82 & -11.72    & 5.23 \\
+		$\omega_7$ & 2.02  & -0.03 & -1.35     & 1.71 \\
+		$\omega_8$ & 10.39 & -1.24 & -11.89    & 2.26 \\
+		$\sigma_{\tau,pre}$  & 0.1   & -0.03 & -34.31    & 0.07 \\
+		$\sigma_{\nu,pre}$   & 0.1   & -0.06 & -62.63    & 0.07 \\
 		\hline
@@
 In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
-
-\subsection{Estimation of ship exposure effect on real narwhal data}
-
 A baseline model is fitted on the narwhals's track before exposure.
@@
 Deviation from the baseline is then assessed by fitting a model with covariates $E$ on the tracks during exposure.
+
+\subsection{Baseline estimations}
+
+\begin{comment}
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\linewidth]{images/application/fe_baseline1_omega_AngleNormal}
+	\caption{}
+	\label{fig:febaseline1omegaanglenormal}
+\end{figure}
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\linewidth]{images/application/fe_baseline1_tau_AngleNormal}
+	\caption{}
+	\label{fig:febaseline1tauanglenormal}
+\end{figure}
 
@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{images/constrained_models/omega_estimations}
+	\includegraphics[width=0.7\linewidth]{images/application/omega_estimations}
 	\caption{Estimation of smooth parameter $\omega$ with tensor splines}
@@
 \paragraph{Response estimations}
+\end{comment}
 
@@
 
-Figure~\ref{fig: trials and intertrials distributions} shows how these periods are distributed among the $6$ narwhals that were tracked.
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.49\textwidth}
+The median time step between GPS measurements is $4.8$ minutes, and the mean time step is $9.3$ minutes. The histogram of the time steps is shown in figure~\ref{fig:alltimestepshisto}.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.5]{images/data_exploration/all_time_steps_histo}
+	\caption{Histogram of time steps}
+	\label{fig:alltimestepshisto}
+\end{figure}
+
+Figure~\ref{fig: trials and intertrials distributions} shows how the exposure periods are distributed among the $6$ narwhals that were tracked.
+
+	\begin{figure}[H]
+		\centering
 			\centering
@@
 			\label{fig: trials and intertrials distributions}
-		\end{subfigure}
-		\begin{subfigure}{0.49\textwidth}
-			\centering
-			\includegraphics[width=0.9\textwidth]{images/data_exploration/all_time_steps_histo}
-			\caption{Histogram of time steps}
-			\label{fig:alltimestepshisto}
-		\end{subfigure}
-	\end{figure}
+
+	\end{figure}
+
  Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals.
-Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.
+Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure on a map of Scoresby Sound fjords system.
 
@@
 		\centering
-		\includegraphics[scale=0.38]{images/data_exploration/all_tracksBE.png}
+		\includegraphics[scale=0.38]{images/data_exploration/tracksBE1.png}
 		\caption{Tracks before exposure experiments}
@@
 		\centering
-		\includegraphics[scale=0.38]{images/data_exploration/all_tracksAE.png}
+		\includegraphics[scale=0.38]{images/data_exploration/tracksAE.png}
 		\caption{Tracks during exposure experiments}
@@
 
- All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
+ All the relevant covariates that are used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.
 
@@
 	\end{tabular}
-	\caption{Summary of the covariates}ubsubsection
+	\caption{Summary of the covariates}
 	\label{tab: covariates}
@@
 
-\subsubsection{Solution to the SDE}
+\subsubsection{Solution of the SDE}
 
@@
 Its moments can be expressed with
-\[\E\left(\vert \vert v(t) \vert \vert\right)=\sqrt{\frac{2}{\pi}}\nu \times \sqrt{\frac{\pi}{2}} L_{\frac{1}{2}} \left( -\frac{\beta}{\sigma^2} \vert \vert \mu \vert \vert ^2 \right)\]
+\[\E\left(\vert \vert v(t) \vert \vert\right)=\sqrt{\frac{2}{\pi}}\nu \times \sqrt{\frac{\pi}{2}} L_{\frac{1}{2}} \left( -\frac{\pi}{4\nu^2} \vert \vert \mu \vert \vert ^2 \right)\]
 
@@
 Denote $\Phi=\arctan(v_2(t),v_1(t)) \in [-\pi, \pi]$ the phase angle of $v(t)$.
-According to \cite{letzepis_vonmises_2015,bennett_solving_1956}, the distribution of $\Phi$ is 
+According to \cite{letzepis_von_2015}, the distribution of $\Phi$ is 
 \[\forall \phi \in [-\pi,\pi], \quad f_{\Phi}(\phi)=\frac{e^{-\frac{\beta \vert \vert \mu \vert \vert^2}{2\sigma^2}}}{2\pi} +\frac{\sqrt{\beta} \vert \vert \mu \vert \vert}{2\sigma\sqrt{2\pi}} \cos(\phi-\theta)e^{-\frac{\beta\vert \vert \mu \vert \vert^2 \sin^2(\phi-\theta)}{2\sigma^2}} \times \left(1+\erf\left(\frac{\beta^{\frac{1}{4}}\vert \vert \mu \vert \vert \cos(\phi-\theta)}{\sqrt{2\sigma}}\right)\right)\]
-where $\erf$ is the Gauss error function defined by 
+where $\beta=\frac{1}{\tau}$, $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$ and $\erf$ is the Gauss error function defined by 
 \[\forall z \in \R, \quad \erf(z)=\frac{2}{\sqrt{\pi}} \int_0^z e^{-t^2} dt\]
@@
 
-\subsection{State space model, simulation and parameter estimation}
-WIth the same notation as before, for $j \in \{1,\cdots,,n\}$, denote $\alpha_j=\begin{pmatrix} x_{1j} & x_{2j} & v_{1j}-\mu_1 & v_{2j}-\mu_2 \end{pmatrix}^\top$.
+\subsection{State space model}
+With the same notation as before, for $j \in \{1,\cdots,,n\}$, denote $\alpha_j=\begin{pmatrix} x_{1j} & x_{2j} & v_{1j}-\mu_1 & v_{2j}-\mu_2 \end{pmatrix}^\top$.
 To formulate a state space model, we shall need the covariances between the velocity and the locations.
@@
 
-\subsection{Kalman filter equations and likelihood}
-
-Now, the conditional law of $\alpha_i$ knowing $y_{1:i}$ is $\mathcal{N}(\hat{\alpha}_i,R_i)$ and the conditional law of $\alpha_i$ knowing $y_{1:i-1}$ is $\mathcal{N}(\hat{\alpha}_i^-,R_i^-)$ where the expectation and covariance matrices are computed as follows.
-
-\paragraph{Kalman filter equations}
-\subparagraph{initialization:}
-\[\hat{\alpha}_0=\mathbb{E}(\alpha_0)=\begin{pmatrix} \mathbb{E}(x_{1,0}) \\ \mathbb{E}(x_{2,0})  \\ \mathbb{E}(v_{1,0})-\mu_1 \\ \mathbb{E}(v_{2,0})-\mu_2\end{pmatrix} \quad R_0=\begin{pmatrix} Var(x_{1,0}) & 0 & 0 & 0 \\ 0 & Var(x_{2,0})  & 0 \\ 0 & 0 & Var(v_{1,0}) & 0 \\ 0 & 0 & 0 & Var(v_{2,0}) \end{pmatrix}\]
-\subparagraph{prediction:}
-For $j \in \{2,\cdots,N\}$,
-\[\hat{\alpha}_j^-=\tilde{T}_{j-1} \hat{\alpha}_{j-1}+ \tilde{B}_{j-1} \mu\]
-\[ R_j^{-} = \tilde{T}_{j-1} R_{j-1} \tilde{T}_{j-1}^\top + \tilde{Q}_{j-1}\]
-\subparagraph{correction:}
-\[K_j=R_j^- \tilde{Z}^\top \left( \tilde{Z} R_j^- \tilde{Z}^\top +H_j\right)^{-1} \]
-\[\hat{\alpha}_j=\hat{\alpha_j}^{-}+K_j \left( y_j-\tilde{Z} \hat{\alpha}_j^-\right)\]
-\[R_j=\left(I_4-K_j \tilde{Z} \right) R_j^-\]
-where $I_4$ is the $4 \times 4$ identity matrix. \\
-
-From these equations we can deduce the likelihood of an observation $y^*=\begin{pmatrix} y_1^* &\cdots& y_n^* \end{pmatrix}$.\\
-Since $y_j=\tilde{Z}\alpha_j+\varepsilon_j$,  $\alpha_j \vert y_{1:j-1} \sim \mathcal{N}(\hat{\alpha}_j^-,R_j^-)$, and $\varepsilon_j \sim \mathcal{N}(0,H_j)$ is independent from the observations, we have
-\[y_j \vert y_{1:j-1} \sim \mathcal{N}\left(\tilde{Z} \hat{\alpha}_i^-,\tilde{Z} R_i^- \tilde{Z}^\top+H_i \right)\]
-
-Thus, if one writes $m_i=\tilde{Z} \hat{\alpha}_i^-$ and $\Gamma_i=\tilde{Z} R_i^- \tilde{Z}^\top +H_i$ then 
-\[\mathcal{L}(\sigma, \beta, \mu \vert y^*) =\mathbb{P}(y_0=y_0^*) \prod_{i=2}^n \mathbb{P}(y_i=y_i^* \vert y_{1:i-1}=y_{1:i-1}^*)\]
-If we suppose that the initial observation is fixed (not random), then
-\[\mathcal{L}(\sigma, \beta, \mu \vert y^*) = \prod_{i=2}^n \frac{1}{\sqrt{(2\pi)^2 \vert \det(\mu_i) \vert}} \exp \left( -\frac{1}{2} (y_i^*-m_i)^\top \Gamma_i^{-1} (y_i^*-m_i) \right) \]
-Hence, the log-likelihood is 
-\begin{align*}
-    \log \left(\mathcal{L}(\sigma,\beta,\mu \vert y^*)\right)&=\sum_{i=2}^n -\log(\sqrt{(2\pi)^2 \vert \det(\Gamma_i) \vert}) -\frac{1}{2} (y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i)\\
-    &=\sum_{i=2}^n -\frac{1}{2}\log\left((2\pi)^2 \vert \det(\Gamma_i) \vert\right) -\frac{1}{2} (y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i) \\
-    &=-(n-1)\log(2\pi)-\frac{1}{2} \sum_{i=2}^n \log(\vert \det(\Gamma_i) \vert) +(y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i)
-\end{align*}
-Consequently, the negative log-likelihood that we need to minimize to obtain parameter estimates is 
-\[
--\log \left(\mathcal{L}(\sigma,\beta,\mu \vert y^*)\right)=(n-1)\log(2\pi)+\frac{1}{2} \sum_{i=2}^n \log(\vert \det(\mu_i) \vert) +(y^*_i-m_i)^\top \mu_i^{-1} (y^*_i-m_i)
-\]
-
-The estimated parameters are 
-\[\hat{\theta}=(\hat{\sigma},\hat{\beta},\hat{\mu})=\underset{\sigma,\beta,\mu}{\mathrm{argmin}} \left( \sum_{i=2}^n \log(\vert \det(\Gamma_i) \vert) +(y^*_i-m_i)^\top \Gamma_i^{-1} (y^*_i-m_i) \right) \]
-
 \end{document}

### 2024-05-14 Latex/main.tex
add proposition for transition density of the whole process

@@
 \usepackage{amsfonts}
+\usepackage{amsthm}
 \usepackage{bbold}
@@
 
+%% MATHS
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[section]
+\newtheorem{lemma}{Lemma}[section]
+\newtheorem{proposition}{Proposition}[section]
+
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\theoremstyle{remark}
+\newtheorem*{example}{Example}
+
 %CITATIONS APPEARANCE
@@
 
-The transition density of the velocity Ornstein-Uhlenbeck process is  
-\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
-	\label{eq: transition RACVM velocity}
-\end{equation}
-
-where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
-Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.
-
-%stationary distribution, distribution of the norm, distribution of the phase
-This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}.
-After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$.
-In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.
-
-By integrating the velocity equation over time one gets the following conditional distribution for the location process :
+\begin{proposition}
+	Let $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}^\top$ for $t \geq 0$. Then the transition density of the Markov process $U$ is given by 
 	\begin{equation}
-	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
-	\label{eq: transition RACVM location}
-\end{equation}
-Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :
-
-\[
-Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
-\]
-The location process is not markovian, but the whole process $\begin{pmatrix}X(t) &V(t) \end{pmatrix}^\top$ is.
-The two components of the location and velocity processes are independent, and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}.
-Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}.
-Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\
+		U(t+\Delta) \vert U(t)=u \sim \mathcal{N}\left( T(\Delta) u +B(\Delta)\mu, Q(\Delta)\right) \mbox{ for all } \Delta>0 \mbox{ and } u \in \R^4
+		\label{eq: X V distribution}
+	\end{equation}
+where \begin{equation}
+	T(\Delta)=\begin{pmatrix} I_2 & A^{-1}(I_2-\exp(-A\Delta)) \\
+	0_2 & \exp(-A\Delta) \end{pmatrix} \mbox{ and } B(\Delta)=\begin{pmatrix}
+	\Delta I_2 -A^{-1}(I_2-\exp(-A\Delta))\\
+	I_2-\exp(-A\Delta)\end{pmatrix}
+	\label{eq: link matrices}
+	\end{equation}
+and the covariance block matrix is given by
+\begin{equation}
+	Q(\Delta)=\begin{pmatrix}
+		q_1(\Delta)I_2 & \Gamma(\Delta) \\
+		\Gamma(\Delta)^\top & q_2(\Delta)I_2
+	\end{pmatrix}
+	\label{eq: covariance matrix}
+\end{equation}
+with 
+\begin{equation}q_1(\Delta)=\frac{\sigma^2}{C}\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{C_{ij} } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{C}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
+\end{equation}
+
+\begin{equation}
+	q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-e^{-\frac{2 \Delta}{\tau}}\right)
+\end{equation}
+\begin{equation}\Gamma(\Delta)=\begin{pmatrix} \gamma_1 & \gamma_2 \\
+	-\gamma_2 & \gamma_1\end{pmatrix}
+\end{equation}
+where
+\begin{equation}\gamma_1=\frac{\sigma^2}{2 C } \times \left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega\Delta)\right)
+\end{equation}
+\begin{equation}\gamma_2=\frac{\sigma^2}{C} \times\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)
+\end{equation}
+and we denoted $C=\frac{1}{\tau^2}+\omega^2$ and $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$.
+\label{prop: transition density}
+\end{proposition}
+
+The formulas derived in \cite{johnson_continuoustime_2008} are a corollary of ~\ref{prop: transition density}, obtained with $\omega=0$. The proof of this proposition is detailed in Appendix.
+
+Equation~\ref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
+Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
+Intuitively, ~\ref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
+ Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}.\\
 
@@
 	\end{subfigure}
-	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min.
+	\caption{(a) Samples of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
 The red cross indicates the starting point of the process.
-The samples were obtained from the exact transition densities in equations~\ref{eq: transition RACVM velocity} and~\ref{eq: transition RACVM location}.
+The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.
 (b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
-Position are sampled from constant time step $h= 1$ min and then downsampled to $5$ min.
+Position are sampled from constant time step $\Delta= 1$ min and then downsampled to $5$ min.
 The samples were obtained from approximate transition densities  with constant covariates on each time step.}
@@
 \label{section: state space model}
-In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
+In most applications, the error between the observation $Y$ and the true position $X$ is not negligible, so that it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
 This is the approach we use here.
@@
 \end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.
-\begin{comment}
-\begin{equation}
-	\forall j \in \{1, \cdots, n_{i,pre}\}, \quad 
-	\left\{ \begin{array}{l}
-		V_{i,j+1}=\exp(-A_{ij}\Delta_{ij})V_{ij}+\zeta_{ij} \\
-		X_{i,j+1}=X_{ij}+A_{ij}^{-1}(I_2-\exp(-A_{ij}\Delta_{ij}))V_{ij}+\xi_{ij}
-	\end{array}\right. \\
-	\label{eq: RACVM autoregressive formula}
-\end{equation}
-
-where $\zeta_{ij}$ and $\xi_{ij}$ are centered normal random vectors.
-\end{comment}
 
@@
 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
-		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij}
+		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij} \\
 		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
@@
 \end{equation}
-where the link matrices are 
-\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_{ij}=\begin{pmatrix} I_2 & A_{ij}^{-1}(I_2-\exp(-A_{ij} \Delta_j)) \\ 0_{2,2} & \exp(-A_{ij} \Delta_j) \end{pmatrix}\]
-
-The vector $\eta_{ij}$ with covariance matrix 
-\begin{equation}
-	Q_{ij}=\begin{pmatrix}
-		M_{ij} & L_{ij} \\
-		L_{ij} & N_{ij}
-	\end{pmatrix}
-	\label{eq: state space covariance}
-\end{equation}
-where $M_{ij}$, $N_{ij}$ and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
-\[M_{ij}=\frac{\sigma_i^2}{C_{ij}}\left( \Delta-2 \frac{\omega_{ij} \sin(\omega_{ij} \Delta_{ij})-\frac{1}{\tau_{i}} \cos(\omega_{ij} \Delta_{ij})}{C_{ij} } \exp\left( -\frac{\Delta_{ij}}{\tau_{ij}} \right) +\frac{\tau_{i}}{2} \left( \frac{\omega_{ij}^2-\frac{3}{\tau_{ij}^2}}{C_{ij}}-\exp\left( -\frac{2\Delta_{ij}}{\tau_i}\right)\right) \right)I_2\]
-
-\[N_{ij}=\frac{\sigma_i^2}{2}\left(1-e^{-\frac{2 \Delta_{ij}}{\tau_{i}}}\right)I_2\]
-
-\[L_{ij}=\begin{pmatrix} a & b \\
-	b & a\end{pmatrix}\]
-with
-\[a=\frac{\sigma_i^2}{2 C_{ij}^2 } \times \left( 1+\exp\left( -\frac{2\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \cos(\omega_{ij}\Delta_{ij})\right)\]
-\[b=\frac{\sigma_{i}^2}{2 C_{ij}^2} \times\left( \exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \sin(\omega_{ij} \Delta_{ij})-\frac{\omega_{ij} \tau_{i}}{2} \left(1-\exp\left( -2 \frac{\Delta_{ij}}{\tau_{i}}\right) \right)\right)\]
-and $C_{ij}=\frac{1}{\tau_i^2}+\omega_{ij}^2$, $\sigma_{i}=\frac{2\nu_i}{\sqrt{\pi \tau_i}}$.
-Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.\\
+where $Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix}$, the link matrices $T_{ij}$ and $B_{ij}$ are computed according to~\ref{eq: link matrices} and the covariance matrix is computed according to~\ref{eq: covariance matrix}.
+\\
 
@@
 		\hline
-		Coefficient  & True value & Absolute biais  & Relative biais ($\%$) & RMSE\\
-		\hline
-		$\log(\tau_{pre})$  & 0& 0.08 &  & 0.16\\
-
-		$\log(\nu_{pre})$  & 1.38 & -0.07 & $-5.4 $ & 0.10 \\
-
-		$\omega_0$  &  0.20  & -0.11 & $-58$ & 0.30 \\
-
-		$\omega_1$  & 7.41 & -0.4 & $-5.4 $ & 1.11 \\
-
-		$\omega_2$  & -2.78 & -1.87 &  $67.3$ & 4.30\\
-
-		$\omega_3$  & 4.05 &  0.30 & $7.4$ & 0.96\\
-
-		$\omega_4$  & 14.34 & 0.53 & $3.71$ &  1.62\\
-
-		$\omega_5$  & 4.30 & -4.89 & $-113.7$ &  5.32\\
-
-		$\omega_6$  &  34.38 & -3.07 & $-8.9$ &   3.78\\
-
-		$\omega_7$  & -0.85 & 1.08 & $-126.4 $ & 1.41\\
-
-		$\omega_8$  &  8.81 & -3.91 & $-44.4 $ & 5.88\\
-
-		$\sigma_{\tau,pre}$ & 0.1 & 0.015& $15$ & 0.10\\
-
-		$\sigma_{\nu,pre}$ & 0.3 & 0.007 & $7 $ & 0.04\\
+		Coefficient  & True value  & Biais & Relative biais ($\%$) & RMSE \\
+		\hline
+		 $\log(\tau_0)$   & 0     & 0.03  &         & 0.11 \\
+		$\log(\nu_0)$     & 1.39  & -0.02 & -1.55     & 0.06 \\
+		$\omega_0$   & 0     & -0.02 &    & 0.22 \\
+		$\omega_1$ & 8.03  & -1.29 & -16.05    & 1.64 \\
+		$\omega_2$ & -3.14 & 1.55  & -49.39    & 4.13 \\
+		$\omega_3$ & 5.08  & 0.21  & 4.05      & 1.23 \\
+		$\omega_4$ & 16.8  & -1.86 & -11.08    & 2.32 \\
+		$\omega_5$ & 5.03  & -6.03 & -119.78   & 6.39 \\
+		$\omega_6$ & 22.52 & -2.83 & -12.56    & 3.27 \\
+		$\omega_7$ & 2.2   & -0.14 & -6.21     & 0.93 \\
+		$\omega_8$ & 6.17  & -4.57 & -74.09    & 6.14 \\
+		$\sigma_{\tau,pre}$  & 0.1   & 0.02  & 21.91     & 0.11 \\
+		$\sigma_{\nu,pre}$    & 0.1   & 0.02  & 18.53     & 0.05 \\
 		\hline
 	\end{tabular}
-	\caption{Measures of estimation performance in the rectangular domain}
-	\label{table: estimation performance rect}
+	\caption{Measures of estimation performance in the fjords for the first $12$h-long trajectories.}
+\label{table: estimation performance fjords 12h}
 \end{table}
@@
 		\hline
-		Coefficient  & True value  & Absolute biais & Relative biais ($\%$)& RMSE \\
-		\hline
-		$\log(\tau_{pre})$   & 0     & -0.16 &         & 0.2  \\
-		$\log(\nu_{pre}$)     & 1.39  & -0.02 & -1.77     & 0.07 \\
-		$\omega_0$   & 0.06  & -0.09 & -148.81   & 0.32 \\
-		$\omega_1$ & 11.04 & -1.87 & -16.94    & 2.66 \\
-		$\omega_2$ & -3.37 & 1.69  & -50.04    & 2.73 \\
-		$\omega_3$ & 7.64  & -2.01 & -26.31    & 2.41 \\
-		$\omega_4$ & 23.24 & -2.47 & -10.63    & 4.95 \\
-		$\omega_5$ & 7.37  & 1.66  & 22.5      & 5.84 \\
-		$\omega_6$ & 41.1  & -4.82 & -11.72    & 5.23 \\
-		$\omega_7$ & 2.02  & -0.03 & -1.35     & 1.71 \\
-		$\omega_8$ & 10.39 & -1.24 & -11.89    & 2.26 \\
-		$\sigma_{\tau,pre}$  & 0.1   & -0.03 & -34.31    & 0.07 \\
-		$\sigma_{\nu,pre}$   & 0.1   & -0.06 & -62.63    & 0.07 \\
+		Coefficient   & True value  & Biais & Relative biais ($\%$) & RMSE \\
+		\hline
+		$\log(\tau_0)$   & 0     & 0.03  &         & 0.09 \\
+		$\log(\nu_0)$  & 1.39  & -0.02 & -1.69     & 0.05 \\
+		$\omega_0$     & 0     & -0.02 &   & 0.13 \\
+		$\omega_1$ & 8.03  & -0.97 & -12.1     & 1.25 \\
+		$\omega_2$ & -3.14 & -0.71 & 22.71     & 2.64 \\
+		$\omega_3$ & 5.08  & 0.35  & 6.84      & 1.01 \\
+		$\omega_4$ & 16.8  & -1.16 & -6.93     & 1.66 \\
+		$\omega_5$ & 5.03  & -6.22 & -123.63   & 6.41 \\
+		$\omega_6$ & 22.52 & -2.04 & -9.06     & 2.35 \\
+		$\omega_7$ & 2.2   & 0.01  & 0.65      & 0.64 \\
+		$\omega_8$ & 6.17  & -2.61 & -42.31    & 3.54 \\
+		$\sigma_{\tau,pre}$   & 0.1   & 0.01  & 13.05     & 0.08 \\
+		$\sigma_{\nu,pre}$    & 0.1   & 0.03  & 29.72     & 0.04 \\
 		\hline
 	\end{tabular}
-	\caption{Measures of estimation performance in the Scoresby Sound domain}
-		\label{table: estimation performance fjords}
+	\caption{Measures of estimation performance in the fjords for the $24$h-long trajectories.}
+\label{table: estimation performance fjords 24h}
 \end{table}
@@
 
-\section{Detailed computations of the integrated rotational Ornstein-Uhlenbeck model}
-\label{section: detailed RACVM computations}
-
-\subsection{Velocity process}
-
-\subsubsection{Solution of the SDE}
-
+\section{Proof of proposition~\ref{prop: transition density}}
+\label{section: transition density proof}
+
+\begin{proof}
+The velocity process is an Ornstein-Uhlenbeck process. For $t \geq 0$ and $\Delta >0$,
 \begin{equation}
-	v(t)=\exp(-At) v(0)+ (I_2-\exp(-At))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)
+	V(t+\Delta)=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)
 	\label{eq: RACVM solution}
 \end{equation}
+
+It has gaussian transition density with mean 
+\begin{equation}
+	\E(V(t+\Delta \vert V(t)))=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu 
+\end{equation}
+and covariance matrix 
+\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top}\\
+	&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 
+\end{align*}
+Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$, the matrix product  $\exp(-Au) \exp(-Au)^{\top}$ is simply $\exp(-\frac{2u}{\tau}) I_2$.
+We deduce that the two components of the velocity are independent and have the same variance, written $q_2(\Delta)$ to follow the notations in~\ref{prop: transition density}.
+The variance is
+\begin{equation}
+	q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-\exp\left(-\frac{2\Delta}{\tau}\right)\right)
+\end{equation}
+These results are found in \cite{gurarie_correlated_2017}.
 In the sequel, we shall use the notation 
-\[\zeta(0,t) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)\]
-
-Replacing $\exp(At)$ by its expression gives more explicitly
+\[\zeta(t,s) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{s} \exp(A(u-s)) dW(u)\]
+
+\begin{align*}
+	X(t+\Delta)&=X(t)+\int_t^{t+\Delta} V(s) ds \mbox{ where } V(s)=\mu+\exp(-A(s-t))(V(t)-\mu)+\zeta(t,s) \\
+	&= X(t)+\mu \Delta+\int_t^{t+\Delta} \exp(-A(s-t))(X(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
+	&= X(t)+\mu \Delta +\int_0^{\Delta} \exp(-As)(X(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
+	&=X(t)+\mu \Delta + (A^{-1}(V(t)-\mu)-A^{-1}\exp(-A\Delta)(X(t)-\mu))+\int_t^{t+\Delta} \zeta(t,s) ds \\
+\end{align*}
+Thus, 
 \begin{equation}
-	v_1(t)=e^{-\frac{t}{\tau}} \left(\cos(\omega t)v_1(0)+\sin(\omega t)v_2(0)\right)+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_1-e^{-\frac{t}{\tau}} \sin(\omega t) \mu_2 +\zeta_1(0,t)
-	\label{eq: first random component RACVM}
-\end{equation}
-
+	X(t+\Delta)=X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu) +\xi(t,t+\Delta) 
+\end{equation}
+where $ \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds$.
+The location process is also gaussian with mean 
 \begin{equation}
-	v_2(t)=e^{-\frac{t}{\tau}} \left(-\sin(\omega t)v_1(0)+\cos(\omega t)v_2(0)\right)+e^{-\frac{t}{\tau}} \sin(\omega t) \mu_1+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_2 +\zeta_2(0,t)
-	\label{eq: second random component RACVM}
-\end{equation}
-where \begin{align*}
-	\zeta_1(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s) \\
-	\zeta_2(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_1(s) +\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_2(s)
+	\E(X(t+\Delta) \vert V(t),X(t)) =X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu)
+\end{equation}
+To get an expression of the covariance matrix, first rewrite
+
+\begin{align*}
+	\xi(t,t+\Delta)&=\int_t^{t+÷\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
+	&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
+	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
+	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
 \end{align*}
-\subsubsection{Autoregressive equation}
-
-For $t \geq 0$ and $\Delta >0$,
-
-\[v(t+\Delta) \vert v(t) \sim \mathcal{N}\left( \mu + \exp(-A\Delta)(v(t)-\mu), \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) \]
-
-Intuitively, the next velocity has a mean that is the sum of the long term mean velocity $\mu$ and the deviation of the previous velocity from this long term mean rotated by an angle $\omega \Delta$.
-\subsubsection{Stationary distribution}
+
+Then use Ito's isometry
+\begin{align*}
+	Var(X(t+\Delta)\vert X(t),V(t))&=\frac{4 \nu^2}{\pi \tau} \int_{t}^{t+\Delta} A^{-1} (I_2-\exp(A(u-t-\Delta)))(I_2-\exp(A(u-t-\Delta)))^\top A^{-\top} du \\
+	&=\frac{4 \nu^2}{\pi \tau} \int_{0}^{\Delta} A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top} dr
+\end{align*}
+Now this integral can be computed explicitly since 
+\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} \begin{pmatrix}  f(r) & 0 \\ 0 & f(r)\end{pmatrix}  \]
+where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
+We obtain that $X_1(t+\Delta)$ and $X_2(t+\Delta)$ are independent and have the same variance, denoted $q_1(\Delta)$.
+Writing $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$, the variance is
+\begin{align*}q_1(\Delta)=\frac{\sigma^2}{C} \left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
+\end{align*}
+
+Now there is still to compute the covariance between the processes $X$ and $V$ to get the full covariance matrix of $U$. With the same notation as in~\ref{prop: transition density}
+\begin{align*}
+	\Gamma(\Delta)&=\E\left(\left(\frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)\right) \left(\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)\right)^\top\right) \\
+	&= \frac{4\nu^2}{\pi \tau}\int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-(t+\Delta)))) \exp(A(u-(t+\Delta)))^\top du \\
+	&=\frac{4\nu^2}{\pi \tau}\int_0^{\Delta} A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top dr 
+\end{align*}
+Then, 
+\[A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top=\frac{1}{C}\exp\left(-\frac{r}{\tau} \right) \begin{pmatrix} g(r) & h(r) \\ -h(r) & g(r) \end{pmatrix}\]
+where $g(r)=\frac{1}{\tau}\left(\cos(\omega r)+\exp\left( -\frac{r}{\tau} \right)\right) -\omega \sin(\omega r)$ and $h(r)=-\frac{1}{\tau}\sin(\omega r)+\omega\left(\cos(\omega r)-\exp\left( -\frac{r}{ \tau}\right)\right)$.
+Eventually we get
+\[
+\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right)
+\]
+\[\gamma_2(\Delta)=\frac{\sigma^2}{C}\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)\]
+
+In the specific case $\omega=0$, we obtain $C=\frac{1}{\tau^2}$ and the variance of the location process $X$ becomes 
+\[q_1(\Delta)=\sigma^2 \tau^2\left( \Delta +\tau\exp\left( -\frac{\Delta}{\tau}\right)+\frac{\tau}{2}\left( -3 -\exp\left( -\frac{2\Delta}{\tau}\right) \right)\right)
+\]
+Writing $\beta=\frac{1}{\tau}$ and reorganizing the terms, we obtain
+\begin{equation}q_1(\Delta)=\frac{\sigma^2}{\beta^2}\left(\Delta -2 \frac{1-\exp(-\beta \Delta)}{\beta}+\frac{1-e^{-2\beta \Delta}}{2\beta}\right)
+\end{equation}
+This result match equation $(6)$ in \cite{johnson_continuoustime_2008}.
+Similarly, in the case $\omega=0$, we get $\gamma_2=0$ and the expression for $\gamma_1$ match equation $(7)$ in \cite{johnson_continuoustime_2008}.
+
+\end{proof}
+
+\section{More results about the standard RACVM}
+\subsection{Stationary distribution}
 
@@
 
-\subsubsection{Distribution of the velocity norm and phase}
+\subsection{Distribution of the velocity norm and phase}
  The order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm.
@@
 \[\E(\vert \vert v(t) \vert \vert)=\nu L_{\frac{1}{2}}(0)=\nu I_{0}(0)=\nu\]
+This result can be used to choose initial value of the parameter $\nu$ based on the mean observed empirical velocity.
 
@@
 
-\subsubsection{Velocity autocovariance function}
+\subsection{Velocity autocovariance function}
 
@@
 \end{equation}
-
-\subsection{Location process} \label{RACVM location process}
-For $t \geq 0$ and $\Delta >0$,
-
-\begin{align*}
-	x(t+\Delta)&=x(t)+\int_t^{t+\Delta} v(s) ds \mbox{ where } v(s)=\mu+\exp(-A(s-t))(v(t)-\mu)+\zeta(t,s) \\
-	&= x(t)+\mu \Delta+\int_t^{t+\Delta} \exp(-A(s-t))(v(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
-	&= x(t)+\mu \Delta +\int_0^{\Delta} \exp(-As)(v(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
-	&=x(t)+\mu \Delta + (A^{-1}(v(t)-\mu)-A^{-1}\exp(-A\Delta)(v(t)-\mu))+\int_t^{t+\Delta} \zeta(t,s) ds \\
-\end{align*}
-Thus, 
-\begin{equation}
-	x(t+\Delta)=x(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v(t)-\mu) +\xi(t,t+\Delta) \\
-	\mbox{ where } \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds
-\end{equation}
-Note that when $\Delta \rightarrow 0$, the deterministic part tends to $x(t)+\Delta v(t)$ as one would expect frome the Euler Maruyama scheme.
-It is known that $\xi$ is normally distributed with mean zero. We still have to compute its covariance matrix.
-
-\begin{align*}
-	\int_t^{t+\Delta} \zeta(t,s) ds &=\int_t^{t+÷\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
-	&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
-	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
-	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
-\end{align*}
-From this formula we deduce an expression for both components $\xi_1(t,t+\Delta)$ and $\xi_2(t,t+\Delta)$ of the gaussian vector.
-\begin{align*}
-	\begin{pmatrix}
-		\xi_1(t,t+\Delta) \\ \xi_2(t,t+\Delta) 
-	\end{pmatrix}
-	&=\frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} \frac{1}{\frac{1}{\tau^2}+\omega^2} 
-	\begin{pmatrix}
-		\frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau}
-	\end{pmatrix} \times \\
-	& \left[ \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}-\exp\left( -\frac{t+\Delta -u}{\tau}\right) 
-	\begin{pmatrix} \cos(\omega(t+\Delta-u)) & \sin(\omega(t+\Delta-u)) \\
-		-\sin(\omega(t+\Delta-u)) & \cos(\omega(t+\delta-u)) 
-	\end{pmatrix} \right]
-	\begin{pmatrix}
-		dW_1(u) \\ dW_2(u)
-	\end{pmatrix}\\
-	&= \frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \int_t ^{t+\Delta} \begin{pmatrix} \frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau} \end{pmatrix} \times \\
-	&\begin{pmatrix}  1-\exp\left(-\frac{(
-			t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) & -\exp\left( -\frac{t+\Delta-u}{\tau}]\right) \sin(\omega(t+\Delta-u)) \\
-		\exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u)) & 1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\end{pmatrix}
-	\times \begin{pmatrix} 
-		dW_1(u) \\ dW_2(u)
-	\end{pmatrix} \\
-	&= \frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \int_t ^{t+\Delta} \begin{pmatrix} \frac{1}{\tau} & \omega \\ -\omega & \frac{1}{\tau} \end{pmatrix} \times \\
-	& \begin{pmatrix} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right) dW_1(u)-\exp\left(  -\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u))dW_2(u) \\
-		\exp\left(  -\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u))dW_1(u)+\left(1- \exp\left( -\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_2(u)
-	\end{pmatrix}
-\end{align*}
-
-Eventually, 
-\begin{align*}
-	\xi_1(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \frac{1}{\tau} \times \Bigg( \textcolor{red}{\int_t^{t+\Delta} \left( 1-\exp\left(-\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_1(u)}\\
-	&-\textcolor{blue}{\int_t^{t+\Delta} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))dW_2(u)}\Bigg) \\
-	&+\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times \omega \Bigg( \textcolor{red}{\int_t^{t+\Delta} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))dW_1(u)}+ \\
-	& \textcolor{blue}{\int_t^{t+\Delta} \left( 1-\exp\left(-\frac{t+\Delta-u}{\tau}\right)\cos(\omega(t+\Delta-u))\right)dW_2(u)}\Bigg)
-\end{align*}
-
-\begin{align*}
-	\xi_1(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times  \Bigg( \textcolor{red}{\int_t^{t+\Delta} \bigg( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right)}  \\
-	& \textcolor{red}{+\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\bigg)dW_1(u)} \\
-	&\textcolor{blue}{+\int_t^{t+\Delta} \bigg( \omega \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta_u))\right)} \\
-	&\textcolor{blue}{-\frac{1}{\tau} \exp\left(-\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u)) \bigg) dW_2(u)} \Bigg)
-\end{align*}
-
-Similarly, 
-\begin{align*}
-	\xi_2(t,t+\Delta)&=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times \Bigg( \textcolor{red}{\int_t^{t+\Delta} \bigg( \frac{1}{\tau} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))} \\
-	&\textcolor{red}{-\omega \left( 1-\exp \left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) \bigg)dW_1(u)} \\
-	&  \textcolor{blue}{+\int_t^{t+\Delta} \bigg( \omega \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta_u))} \\
-	&\textcolor{blue}{+\frac{1}{\tau} \left(1-\exp\left(-\frac{t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) \right) \bigg) dW_2(u)} \Bigg)
-\end{align*}
-
-Denote 
-\begin{align*}
-	& \textcolor{red}{X_1=\int_t^{t+\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) +\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\right)dW_1(u)} \\
-	&\textcolor{red}{Y_1=\int_t^{t+\Delta} \left( \frac{1}{\tau} \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u)) -\omega \left( 1-\exp \left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) \right)dW_1(u)} \\
-	&\textcolor{blue}{X_2=\int_t^{t+\Delta} \left( \omega \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right)-\frac{1}{\tau} \exp\left(-\frac{t+\Delta-u}{\tau} \right) \sin(\omega(t+\Delta-u)) \right) dW_2(u)} \\
-	&\textcolor{blue}{Y_2=\int_t^{t+\Delta} \left( \omega \exp\left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))+\frac{1}{\tau} \left(1-\exp\left(-\frac{t+\Delta-u}{\tau} \right) \cos(\omega(t+\Delta-u)) \right) \right) dW_2(u)}
-\end{align*}
-so that both $X_1$,$Y_1$ and $X_2$
-,$Y_2$ are independent and 
-\[\xi_1(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times (X_1+Y_1) \quad \xi_2(t,t+\Delta)=\frac{2\nu}{\sqrt{\pi \tau} \times \left( \frac{1}{\tau^2}+\omega^2\right)}  \times (X_2+Y_2)\]
-\[Var(\xi_1(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} (Var(X_1)+Var(Y_1)) \quad Var(\xi_1(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} (Var(X_2)+Var(Y_2)) \]
-
-Itô's isometry yields
-\begin{align*}
-	Var(X_1)&=\int_t^{t+\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{t+\Delta-u}{\tau}\right) \cos(\omega(t+\Delta-u))\right) +\omega \exp \left( -\frac{t+\Delta-u}{\tau}\right) \sin(\omega(t+\Delta-u))\right)^2du \\
-	&= \int_0^{\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{u}{\tau}\right) \cos(\omega u)\right) +\omega \exp \left( -\frac{u}{\tau}\right) \sin(\omega u)\right)^2du 
-\end{align*}
-Similarly, 
-\[
-Var(Y_1)=\int_0^{\Delta} \left( \frac{1}{\tau} \exp\left( -\frac{u}{\tau}\right) \sin(\omega u) -\omega \left( 1-\exp \left( -\frac{u}{\tau}\right) \cos(\omega u)\right) \right)^2 du
-\]
-Hence,
-\begin{align*}
-	Var(\xi_1(t,t+\Delta))&=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)^2} \Bigg( \int_0^{\Delta} \left( \frac{1}{\tau} \left(1-\exp\left( -\frac{u}{\tau}\right) \cos(\omega u)\right) +\omega \exp \left( -\frac{u}{\tau}\right) \sin(\omega u)\right)^2+ \\
-	& \left( \frac{1}{\tau} \exp\left( -\frac{u}{\tau}\right) \sin(\omega u) -\omega \left( 1-\exp \left( -\frac{u}{\tau}\right) \cos(\omega u)\right) \right)^2 du\Bigg)
-\end{align*}
-
-It is possible to see that the formula is exactly the same for $\xi_2(t,t+\Delta)$.
-
-The time integral can be calculated explicitly to obtain
-\begin{align*}
-	Var(\xi_1(t,t+\Delta))&=Var(\xi_2(t,t+\Delta))=\frac{4\nu^2}{\pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
-	& \left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
-\end{align*}
-It is easy to check that when $\omega=0$, the formula match the one for the basic correlated velocity model the hidden state. \\
-With similar calculations, it can be shown that
-\[Cov(\xi_1(t,t+\Delta), \xi_2(t,t+\Delta)=0\]
 

### 2024-05-15 Latex/main.tex
latest paper version

@@
 Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
-To make this possible, we formulate a state-space model by binding the velocity and location in one hidden state vector, following \cite{johnson_continuoustime_2008}.
-Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\
+To make this possible, we derive explicit formulas for the transitioon density of the process $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}$ and then formulate a state-space model from the discretisation of these formulas.
+Maximum likelihood estimation based on Kalman filter is performed as for any other model in \texttt{smoothSDE}.\\
 
@@
 \begin{itemize}
-	\item Derivation of a state-space model for the"Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
+	\item Derivation of explicit formulas for the transition density of the "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
 	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore109.R
many updates

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 109
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore112.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 112
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore152.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 152
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore157.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 157
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore175.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 175
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore187.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 187
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore108.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 108
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore132.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 132
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore148.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 148
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore179.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 179
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore162.R

@@ -1,0 +1,209 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 162
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories ------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore139.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 139
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore142.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 142
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore167.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 167
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore184.R

@@ -1,0 +1,211 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 184
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+
+data=foreach (i=1:N_ID,.combine='rbind') %do% {
+    
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                        log_sigma_obs=log(sigma_obs),v0=v0,x0=x0[i,],times=times_hf,land=border,verbose=FALSE)
+    
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_hf) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+# Save plot of the trajectories -------------------
+
+plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data,mapping=aes(y1,y2,col=ID),size=0.1)+
+  geom_point(data = data%>% filter(!duplicated(ID)),
[…]

### 2024-05-21 Latex/main.tex
last paper version

@@
 
-To summarize, the main contributions of this paper are
+To summarize, the main contributions of this paper are:
 \begin{itemize}
@@
 For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}.
-These points are noisy observations of the underlying unobserved true positions $x_{ij}$.\\
+We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, projected in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}.
+These points are noisy observations of the underlying unobserved true positions.\\
 
@@
 The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move.
-The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$.
-Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}=d(x_{ij},p_{ij})$.
+The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$.
+Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary.
 About $9\%$ of the GPS measurements in the dataset turned out to be on land.
 After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API.
+We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy the of land polygons.
+We modified  the land polygons using \texttt{qgis}.
 After adjustements,  approximately $1 \%$ of the observations remained on land.
@@
 		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
-		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
+		E^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
 		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
@@
 Thus we only consider the effect of the shore when the narwhals are close enough to it.
-We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
-The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is defined as the angle between the vectors $x_{ij}-p_{ij}$ and the true velocity $v_{ij}$.
-It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-\hat{p}_{ij}$.
+The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$.
 This is illustrated in figure~\ref{fig: illustration nearest shore points}.
@@
 		\node[left] at (0,0) {$\partial \mathcal{D}$};
-		\node[left] at (0,2) {$\hat{p}_{ij}$};
+		\node[left] at (0,2) {$p_{ij}$};
 		\node[left] at (1.2,2.8) {$\hat{v}_{ij}$};
@@
 		\node[below,red] at (3,2) {$\vec{n}_{ij}$};
-
-		\node[above,orange] at (2.3,2.3) {$\hat{\Theta}_{ij}$};
+		\node[above,orange] at (2.3,2.3) {$\Theta_{ij}$};
 	\end{tikzpicture}
@@
 
-The distance from the ship denoted $D^{ship}_{ij}$, is estimated as the distance between the GPS ship and narwhals's positions.
-We denote $\hat{D}^{ship}_{ij}$ this estimation.
-The values are comprised between $2.68$ and $63.8$ km.
+The distance from the ship denoted $D^{ship}_{ij}$, is estimated as the distance between the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
 Exposure to ship covariate for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
-Its estimate $\hat{E}^{ship}$ is the inverse of $\hat{D}^{ship}$.
 $D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case.
@@
 with 
-\begin{equation}q_1(\Delta)=\frac{\sigma^2}{C}\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{C_{ij} } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{C}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
+\begin{equation}q_1(\Delta)=\frac{\sigma^2}{C}\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{C} \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{C}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
 \end{equation}
@@
 	\end{subfigure}
-	\caption{(a) Samples of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
+	\caption{(a) Trajectory of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
 The red cross indicates the starting point of the process.
 The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.
-(b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
+(b) Trajectory of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
 Position are sampled from constant time step $\Delta= 1$ min and then downsampled to $5$ min.
-The samples were obtained from approximate transition densities  with constant covariates on each time step.}
+The samples are obtained from approximate transition densities  with constant covariates on each time step.}
 	\label{fig: sample RACVM}
@@
 \end{equation}
-where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
-and $\Theta(t)$ is the angle between the velocity $V(t)$ and the shoreline as defined in subsection ~\ref{subsection: covariates}.
+where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the observed distance to the boundary of the domain $\partial \mathcal{D}$
+and $\Theta(t)$ is the observed angle between the velocity $V(t)$ and the shoreline as defined in subsection ~\ref{subsection: covariates}.
 Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
@@
 It should always remain significantly lower than $D^{shore}$ for the process to be constrained.
-Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
+Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
 The closer we are to the shore, the faster the velocity should rotate,
-whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$.
+whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be close to $0$.
 Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
@@
 A sample of a CRCVM with smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}.
-The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.
+The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system in South-East Greenland.
 
@@
 
- We show side by side the samples obtained within different domain for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.
+ We show side by side the samples obtained with a circular border for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.
 
@@
  		\centering
- 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_standard}
+ 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_standard}
  		\label{fig:illustrativesamplecircstandard}
@@
  		\centering
- 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_pers}
+ 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_pers}
  		\label{fig:illustrativesamplecircpers}
@@
 
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_standard}
- 		\label{fig:illustrativesamplerectstandard}
- 	\end{subfigure}
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_pers}
- 		\label{fig:illustrativesamplerectpers}
- 	\end{subfigure}
-
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard}
- 		\label{fig:illustrativesamplefjordsstandard}
- 	\end{subfigure}	
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_pers}
- 		\label{fig:illustrativesamplefjordspers}
- 	\end{subfigure}
- 	\caption{Left: samples from a CRCVM with smooth $\omega$ and constant $\tau=1$ h in three different domains.
-Right: samples from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the shoreline.}
+ 	\caption{Left: sample from a CRCVM with smooth $\omega$ and constant $\tau=1$ h.
+Right: sample from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the boundary.}
  	\label{fig: crcvm examples}
@@
 This is taken into account by using a mixed effect model for the RCVM parameters.
-For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations $Y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$ and the true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
+For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations $y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$ and the true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
 Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters.
@@
 	\begin{array}{l}
-		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
+		y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
@@
 		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
-		\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)} \\
-		\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}  \\
-		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
-		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
+		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}\\
+		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}  \\
+		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
+		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t)) \quad t \geq 0\\
 		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
@@
 
-The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$, $\omega_{0}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
+The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$, $\omega_{0}$ are population intercepts and $b_{\tau,i}$, $b_{\nu,i}$
 are the individual parameters.
@@
 
-\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
+\begin{equation}  \forall j \in \{1,\cdots,n_{i,post}\}, \quad \left\{
 	\begin{array}{l}
-		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}\\
-		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
-		dX_i(t)=V_i(t)dt  \\
-		dV_i(t)=-\begin{pmatrix} 
-			\frac{1}{\tau_i} & -\omega_i(t) \\
-			\omega_i(t) & \frac{1}{\tau_i}
-		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
-		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
-		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
-		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
-		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k})\phi_{k}(E^{ship}_i(t))+b_{\tau,i} \\
+		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k}) \phi_{k}(E^{ship}_i(t)) +b_{\nu,i}  \\
 	\end{array}
@@
 
-Again, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters.
-The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
+Again, the functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
 
@@
 In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
-\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(\hat{E}^{shore}_{ij},\hat{\Theta}_{ij})\]
+\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(E^{shore}_{ij},\Theta_{ij})\]
  Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
@@
 
-The state space equations are
+Discretisation of~\ref{eq: X V distribution} gives the following state-space equations:
 
@@
 	\begin{array}{l}
-		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
+		y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
@@
 \label{section: simulation constrained motion}
-In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain.
+In order to experiment the constrained models we simulate a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain.
 We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}\\.
 
-We simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in a rectangular domain with sides of length $8$ km with the statistical model~\ref{eq: exact baseline model}.
-This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
+We simulate $M=100$ batches of $N=6$ individual trajectories over $T=24$ h with time step $\Delta=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application..
+This amounts to $n_{i,pre}=1440$ observations for each individual $i \in \{1,\cdots,N\}$.
 We fix $\sigma_{obs}=5$ m to represent low measurement errors.
@@
 The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{0}$ and $\nu_{0}$ are fixed respectively to $1$ h and $4$ km/h.
-The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance rect}.
-For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.
+The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance fjords 12h}.
+For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $1$ km away from the boundary.
 We kept the same $6$ initial positions for all the batches of trajectories.
-To obtain the trajectories of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to
-\begin{equation}
-	\omega_{ij}=f_{\omega}(E^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
-\end{equation}
-where $E^{shore}_{ij}=E^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$.
+To obtain the trajectories of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step.
 The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
@@
 \\
-
-We also simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application.
-The parameters $\sigma_{obs}$, $L$, $\sigma_{\tau,pre}$, $\sigma_{\nu,pre}$, $\tau_{0}$,$\nu_{0}$ are kept the same as in the rectangular case, and only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are changed to adapt to this specific fjords domain.
-The exact values of the spline coefficients are found in Table~\ref{table: estimation performance fjords}.
-Again, for each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled within the water area, with the requirement that it should be at least $500$ metres away from the shore.
-None of the $M\times N=600$ trajectories reached land.\\
 
@@
 
-In the rectangular and fjords domains, these $100$ batches of noisy samples are used as data for estimation.
-We thus obtain $100$ parameters estimate in the two cases.
+These $100$ batches of noisy samples are used as data for estimation.
+First, we estimate using only the first $12$h of the trajectories, that is $720$ observations for each narwhals.
+Then, we use the whole $24$h-long trajectory.
 The covariates $E^{shore}$ and $\Theta$ are computed from the observed noisy positions.
@@
 We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau,pre}=1$ h, $\sigma_{\nu,pre}=1$ km/h and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories in the rectangular and the fjords domain, with confidence intervals obtained from the observed Fisher information matrix.
+For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the root mean squared error and the biais of the estimates.
+Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.
+The results are summarized in Tables~\ref{table: estimation performance fjords 12h} and~\ref{table: estimation performance fjords 24h}.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with confidence intervals obtained from the observed Fisher information matrix.
 The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
@@
 	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q4_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q2_ExpShore .png}
-		\caption{}
-	\end{subfigure}
 	\begin{subfigure}{0.48\textwidth}
@@
 	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only,.
-The red curve represents the true value of $\omega$.
-(a)-(b) Estimation from samples in the rectangular domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-(c)-(d) Estimation from samples in the fjords domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+The red curve represents the true value of $\omega$.}
 	\label{fig: marginal_estimates_CRCVM}
 \end{figure}
-
-For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,100\}$ and compute the root mean squared error and the biais of the estimates.
-Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.
-The results are summarized in Tables~\ref{table: estimation performance rect} and~\ref{table: estimation performance fjords}.
 
@@
 
-In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic ariguns exposure.
+In this section, we apply our diffusion models to analyse a behavioral response of the narwhals to ship and seismic airguns exposure.
 A baseline model is fitted on the narwhals's track before exposure.
 It is supposed to capture the characteristics of the movement under normal conditions.
-Deviation from the baseline is then assessed by fitting a model with covariates $E$ on the tracks during exposure.
+Deviation from the baseline is then assessed by fitting a model with covariate $E^{ship}$ on the tracks during exposure.
 
 \subsection{Baseline estimations}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|}
+		\hline
+		Coefficient   & Estimate  & $95\%$ CI \\
+		\hline
+		$\log(\tau_0)$   & 0     & 0.03 \\
+		$\log(\nu_0)$  & 1.39  & -0.02 \\
+		$\omega_0$     & 0     & -0.02 \\
+		$\sigma_{\tau,pre}$   & 0.1   & 0.01  \\
+		$\sigma_{\nu,pre}$    & 0.1   & 0.03  \\
+		\hline
+	\end{tabular}
+	\caption{Intercepts baseline estimations}
+	\label{table: baseline estilations}
+\end{table}
+
+\subsection{Response estimations}
 
@@
 Now this integral can be computed explicitly since 
-\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} \begin{pmatrix}  f(r) & 0 \\ 0 & f(r)\end{pmatrix}  \]
+\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} f(r) I_2 \]
 where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
@@
 In the specific case $\omega=0$, we obtain $C=\frac{1}{\tau^2}$ and the variance of the location process $X$ becomes 
-\[q_1(\Delta)=\sigma^2 \tau^2\left( \Delta +\tau\exp\left( -\frac{\Delta}{\tau}\right)+\frac{\tau}{2}\left( -3 -\exp\left( -\frac{2\Delta}{\tau}\right) \right)\right)
+\[q_1(\Delta)=\sigma^2 \tau^2\left( \Delta +2\tau\exp\left( -\frac{\Delta}{\tau}\right)+\frac{\tau}{2}\left( -3 -\exp\left( -\frac{2\Delta}{\tau}\right) \right)\right)
 \]
@@
 
-\subsection{State space model}
-With the same notation as before, for $j \in \{1,\cdots,,n\}$, denote $\alpha_j=\begin{pmatrix} x_{1j} & x_{2j} & v_{1j}-\mu_1 & v_{2j}-\mu_2 \end{pmatrix}^\top$.
-To formulate a state space model, we shall need the covariances between the velocity and the locations.
-Calculations similar to  section \ref{RACVM location process} give for all $t \geq 0$, $\Delta >0$
-\begin{align*}
-	Cov(\xi_1(t,t+\Delta),\zeta_1(t,t+\Delta))&=Cov(\xi_2(t,t+\Delta)\zeta_2(t,t+\Delta))=\frac{4\nu^2}{2 \pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
-	& \left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right)
-\end{align*}
-
-\begin{align*}
-	Cov(\xi_1(t,t+\Delta),\zeta_2(t,t+\Delta))&=Cov(\xi_2(t,t+\Delta),\zeta_1(t,t+\Delta))=\frac{4\nu^2}{2 \pi \tau \times \left( \frac{1}{\tau^2}+\omega^2\right)} \times \\
-	&\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)
-\end{align*}
-
-Therefore, the recursive equations for the state space are
-\begin{align*}
-	\forall j \in \{1, \cdots, n\}&, \quad 
-	\left\{ \begin{array}{l}
-		y_j=x_j+\varepsilon_j \\
-		v_{j+1}=\mu +\exp(-A\Delta_j)(v_j-\mu)+\zeta_j \\
-		x_{j+1}=x_j+\mu \Delta_j+A^{-1}(I_2-\exp(-A\Delta_j))(v_j-\mu)+\xi_j
-	\end{array}\right. \\
-	&\Leftrightarrow \left\{ \begin{array}{l}
-		y_j=Z\alpha_j+\varepsilon_j \\
-		\alpha_{j+1}=T_j \alpha_j+B_j \mu + \eta_j
-	\end{array}\right.
-\end{align*}
-where $Z=\begin{pmatrix} I_2 & 0_2 \end{pmatrix}$, $T_j=\begin{pmatrix} I_2 & A^{-1}(I_2-\exp(-A\Delta_j)) \\
-	0_2 & \exp(-A\Delta_j) \end{pmatrix}$, $B_j=\begin{pmatrix}
-	\Delta_j I_2 \\
-	0_2 \end{pmatrix}$ and $\eta_j=\begin{pmatrix} \xi_j \\ \zeta_j\end{pmatrix}=\begin{pmatrix} \xi(t_j,t_j+\Delta_j) \\ \zeta(t_j,t_{j+\Delta_j})\end{pmatrix} \sim \mathcal{N}(0,Q_j)$ with 
-\[Q_j=\begin{pmatrix} 
-	Var(\xi_{1,j} & 0 & Cov(\xi_{1,j},\zeta_{1,j}) & Cov(\xi_{1,j},\zeta_{2,j})) \\
-	\cdots & Var(\xi_{2,j}) & Cov(\xi_{2,j}, \zeta_{1,j}) & Cov(\xi_{2,j},\zeta_{2,j}) \\
-	\cdots & \cdots & Var(\zeta_{1,j}) & 0 \\
-	\cdots & \cdots & \cdots & Var(\zeta_{2,j}) 
-\end{pmatrix}\]
-
 \end{document}

### 2024-05-21 Latex/main.tex
fix typos in paper

@@
 		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}  \\
-		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
+		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
 		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t)) \quad t \geq 0\\
@@
 	\omega_{ij} & \frac{1}{\tau_{i}}
-\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.
+\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu}^{(i)}$.
 
@@
 In order to experiment the constrained models we simulate a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain.
-We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}\\.
+We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}.\\
 
@@
 We choose $L=8$ for the degree of freedom of the bivariate splines of $E^{shore}$ and $\Theta$.
-The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{0}$ and $\nu_{0}$ are fixed respectively to $1$ h and $4$ km/h.
+The random effects standard deviations are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu,pre}=0.1$ to represent the individual variability, and $\tau_{0}$ and $\nu_{0}$ are fixed respectively to $1$ h and $4$ km/h.
 The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance fjords 12h}.
@@
 This corresponds to one single knot in each dimension.
-Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau,pre}$ and $\sigma_{\nu,pre}$.
+Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
 Measurement error $\sigma_{obs}$ is supposed to be known.
-We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau,pre}=1$ h, $\sigma_{\nu,pre}=1$ km/h and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
-For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{pre})}^{(k)}$, $\widehat{\log(\nu_{pre})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the root mean squared error and the biais of the estimates.
+We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
+For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the root mean squared error and the biais of the estimates.
 Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.

### 2024-05-27 Latex/main.tex
updates

@@
 \usepackage{subcaption}
+\usepackage{float}
+\usepackage{multirow}
+
 \usepackage{amssymb}
@@
 \usepackage{bbold}
-\usepackage{float}
+
 \usepackage[margin=0.8in]{geometry}
@@
 Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model.
-Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one single model.
+Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one holistic model.
 \\
@@
 Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
-Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions.
+Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source.
 We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\
@@
 Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}.
-To make this possible, we derive explicit formulas for the transitioon density of the process $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}$ and then formulate a state-space model from the discretisation of these formulas.
-Maximum likelihood estimation based on Kalman filter is performed as for any other model in \texttt{smoothSDE}.\\
+To make this possible, we derive explicit formulas for the transition density of the process $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}$ and then formulate a state-space model from the discretisation of these formulas.
+Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\
 
@@
 Then, the diffusion models are discussed in Section 3.
-Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effect.
+Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effects.
 Simulation study results are shown in section 5.
@@
 
-The dataset analysed in this section has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
+The dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
 Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
@@
 We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy the of land polygons.
-We modified  the land polygons using \texttt{qgis}.
-After adjustements,  approximately $1 \%$ of the observations remained on land.
+After adjustement the land polygons using \texttt{qgis},  approximately $1 \%$ of the observations remained on land.
 Distance to shore values range from $0$ to $7.6$ km.
@@
 	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
-		\caption{}
-	\end{subfigure}
-	\caption{(a) Trajectory of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
+	\includegraphics[scale=0.5]{images/rcvm/illustrative_sample_standard.png}
+	\caption{Trajectory of a RCVM with parameters $\tau=1$h, $\nu=4$ km/h, $\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
 The red cross indicates the starting point of the process.
-The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.
-(b) Trajectory of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}.
-Position are sampled from constant time step $\Delta= 1$ min and then downsampled to $5$ min.
-The samples are obtained from approximate transition densities  with constant covariates on each time step.}
+The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.}
 	\label{fig: sample RACVM}
@@
 
+Unconstrained movement of marine mammals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, while other animals or organisms whose motion appear tortuous require the more general model \ref{eq: RACVM equation}.
+In the case of motion under spatial constraints, an increase in tortuosity enables the animals to avoid the boundary and adapt their heading to the constraint. 
 \subsection{Constrained rotational correlated velocity model}
@@
 
-Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
-To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore.
+We thus propose to make use of a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to include how the animals turn in reaction to the shore. 
 To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
@@
 \[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
-In this case, the formulas of section~\ref{section: RACVM} are still valid.
+In this case, we can use the formulas of section~\ref{section: RACVM} to get an approximation of the exact transition density.
 The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
@@
 
-Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :
+Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated  :
 
@@
 In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
-A sample of a CRCVM with smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}.
-The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system in South-East Greenland.
 
@@
 
- We show side by side the samples obtained with a circular border for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.
-
- \begin{figure}[H]
- 	\centering
+ We illustrate this flexibility in Figure~\ref{fig: crcvm examples}.Trajectories obtained within the fjords domain for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
+
+ \begin{figure}[H]
+
  	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_standard}
- 		\label{fig:illustrativesamplecircstandard}
- 	\end{subfigure}
- 	\begin{subfigure}{0.48\textwidth}
- 		\centering
- 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_pers}
- 		\label{fig:illustrativesamplecircpers}
- 	\end{subfigure}
-
- 	\caption{Left: sample from a CRCVM with smooth $\omega$ and constant $\tau=1$ h.
-Right: sample from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the boundary.}
+		\includegraphics[scale=0.65]{images/crcvm/illustrative_sample_fjords}
+		\label{fig:illustrativesamplefjords}
+	\end{subfigure}
+
+ 	\caption{Five day trajectories with time step $\Delta=5$ min in Scoresby Sound fjords domain}
  	\label{fig: crcvm examples}
@@
 
-\section{Statistical model}
-
-\subsection{Mixed effet model for ship exposure and shore effects}
+\section{Inference of the sound exposure effect}
+
+\subsection{Mixed effet model for sound exposure and shore effects}
 \label{subsection: ship exposure effect}
@@
 		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
-		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t)) \quad t \geq 0\\
-		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
+		\omega_i(t)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E_i^{shore},\Theta_i)
 	\end{array}
@@
 
-\begin{equation}  \forall j \in \{1,\cdots,n_{i,post}\}, \quad \left\{
+\begin{equation}   \left\{
 	\begin{array}{l}
-		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k})\phi_{k}(E^{ship}_i(t))+b_{\tau,i} \\
-		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k}) \phi_{k}(E^{ship}_i(t)) +b_{\nu,i}  \\
+		y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots,n_{i,post}\} \\
+		\log(\tau_{i}(t))=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k})\phi_{k}(E^{ship}_i(t))+b_{\tau,i} \\
+		\log(\nu_{i}(t))=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k}) \phi_{k}(E^{ship}_i(t)) +b_{\nu,i}  \\
 	\end{array}
@@
 
-Again, the functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
+The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are considered are offset.
+The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
+Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$ are estimated from the trajectories during exposure to the sound
 
@@
 	\omega_{ij} & \frac{1}{\tau_{i}}
-\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu}^{(i)}$.
+\end{pmatrix}$ and $U_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu}^{(i)}$.
 
@@
 	\begin{array}{l}
-		y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
+		y_{ij}=ZU_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
-		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij} \\
+		U_{i,j+1}=T_{ij} U_{ij} + \eta_{ij} \\
 		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
@@
 Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
-The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
+The complete likelihood is computed from the observations $y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
 Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation.
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-		\hline
-		Coefficient  & True value  & Biais & Relative biais ($\%$) & RMSE \\
-		\hline
-		 $\log(\tau_0)$   & 0     & 0.03  &         & 0.11 \\
-		$\log(\nu_0)$     & 1.39  & -0.02 & -1.55     & 0.06 \\
-		$\omega_0$   & 0     & -0.02 &    & 0.22 \\
-		$\omega_1$ & 8.03  & -1.29 & -16.05    & 1.64 \\
-		$\omega_2$ & -3.14 & 1.55  & -49.39    & 4.13 \\
-		$\omega_3$ & 5.08  & 0.21  & 4.05      & 1.23 \\
-		$\omega_4$ & 16.8  & -1.86 & -11.08    & 2.32 \\
-		$\omega_5$ & 5.03  & -6.03 & -119.78   & 6.39 \\
-		$\omega_6$ & 22.52 & -2.83 & -12.56    & 3.27 \\
-		$\omega_7$ & 2.2   & -0.14 & -6.21     & 0.93 \\
-		$\omega_8$ & 6.17  & -4.57 & -74.09    & 6.14 \\
-		$\sigma_{\tau,pre}$  & 0.1   & 0.02  & 21.91     & 0.11 \\
-		$\sigma_{\nu,pre}$    & 0.1   & 0.02  & 18.53     & 0.05 \\
+		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
+		\hline
+		&\multicolumn{3}{|c|}{Intercepts} & \multicolumn{8}{|c|}{Spline coefficients} & \multicolumn{2}{|c|}{SD} \\
+		\hline
+		&$\log(\tau_0)$ & $\log(\nu_0)$ & $\omega_0$ & $\omega_1$ & $\omega_2$ & $\omega_3$ & $\omega_4$  & $\omega_5$ & $\omega_6$ & $\omega_7$ & $\omega_8$ & $\sigma_{\tau}$ & $\sigma_{\nu}$ \\
+		\hline 
+		 True values & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
+		\hline
+			\multicolumn{14}{|c|}{Estimates for $d(X_0,\mathcal{D}) > 4$ km} \\
+			\hline
+	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
+	& & & & & & & & & & & & & \\
+	\hline
+	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
+	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
+	0.09\\
+	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
+	0.08 & 0.03  \\
 	\hline	
 \end{tabular}
-	\caption{Measures of estimation performance in the fjords for the first $12$h-long trajectories.}
-\label{table: estimation performance fjords 12h}
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-		\hline
-		Coefficient   & True value  & Biais & Relative biais ($\%$) & RMSE \\
-		\hline
-		$\log(\tau_0)$   & 0     & 0.03  &         & 0.09 \\
-		$\log(\nu_0)$  & 1.39  & -0.02 & -1.69     & 0.05 \\
-		$\omega_0$     & 0     & -0.02 &   & 0.13 \\
-		$\omega_1$ & 8.03  & -0.97 & -12.1     & 1.25 \\
-		$\omega_2$ & -3.14 & -0.71 & 22.71     & 2.64 \\
-		$\omega_3$ & 5.08  & 0.35  & 6.84      & 1.01 \\
-		$\omega_4$ & 16.8  & -1.16 & -6.93     & 1.66 \\
-		$\omega_5$ & 5.03  & -6.22 & -123.63   & 6.41 \\
-		$\omega_6$ & 22.52 & -2.04 & -9.06     & 2.35 \\
-		$\omega_7$ & 2.2   & 0.01  & 0.65      & 0.64 \\
-		$\omega_8$ & 6.17  & -2.61 & -42.31    & 3.54 \\
-		$\sigma_{\tau,pre}$   & 0.1   & 0.01  & 13.05     & 0.08 \\
-		$\sigma_{\nu,pre}$    & 0.1   & 0.03  & 29.72     & 0.04 \\
-		\hline
-	\end{tabular}
-	\caption{Measures of estimation performance in the fjords for the $24$h-long trajectories.}
-\label{table: estimation performance fjords 24h}
 \end{table}

### 2024-05-28 Latex/main.tex
update Latex

@@
 
+The persistence is estimated to $\hat{\tau}_{0}=1.3 $ h.
+In comparison, harbour seal in Alaska were proved to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \cite{johnson_continuoustime_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \cite{gurarie_correlated_2017}.
+The intercept value for $\nu$ is estimated to be $\hat{\nu}_{0}=4.19$ km/h with $95\%$ CI $3.79-4.63$.
 \begin{table}[H]
@@
 		\hline
-		$\log(\tau_0)$   & 0     & 0.03 \\
-		$\log(\nu_0)$  & 1.39  & -0.02 \\
-		$\omega_0$     & 0     & -0.02 \\
-		$\sigma_{\tau,pre}$   & 0.1   & 0.01  \\
-		$\sigma_{\nu,pre}$    & 0.1   & 0.03  \\
+		$\log(\tau_0)$   & 0.3    &  \\
+		$\log(\nu_0)$  & 1.48  &  \\
+		$\omega_0$     & -0.13     &  \\
+		$\sigma_{\tau,pre}$   & 0.1   &   \\
+		$\sigma_{\nu,pre}$    & 0.1   &  \\
 		\hline
@@
 \end{table}
+
+Figure \ref{fig:febaseline3omegaanglenormalq0} shows the estimated smooth $\omega$ as a function of $\Theta$ for different values of the distance to shore.
+As distance increases, the magnitude of the angular velocity decreases.
+% PUT ON THE SAME PLOT WITH DIFFERENT COLOR FOR EACH CURVE
+\begin{figure}[H]
+	\begin{subfigure}{0.33\textwidth}
+		\centering
+		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.95_ExpShore .png"}
+	\end{subfigure}
+	\begin{subfigure}{0.33\textwidth}
+		\centering
+		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.75_ExpShore .png"}
+	\end{subfigure}
+	\begin{subfigure}{0.33\textwidth}
+		\centering
+		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.25_ExpShore .png"}
+	\end{subfigure}
+	\caption{}
+
+	\label{fig:febaseline3omegaanglenormalq0}
+\end{figure}
 

### 2024-05-29 Latex/main.tex
new version

@@
 The probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases.
-Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model.
+Additionally, an overall speed increase due to sound exposure is revealed in the 2018 a data by an analysis of the consecutive GPS positions through a mixed effect model.
 Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one holistic model.
@@
 \end{equation}
-where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates.
-Here, $W(t)$ is a two-dimensional brownian motion.
-The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
+where $V(t)$ is the horizontal two-dimensional velocity at time $t$, $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates and $W(t)$ is a two-dimensional brownian motion. 
+The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, and $\omega$ is an angular velocity that controls how fast the velocity vector rotates.
 The case $\omega=0$ corresponds to a standard CVM.
@@
 For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure.
-We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, projected in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}.
+The positions are observed at discrete times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, projected in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}.
 These points are noisy observations of the underlying unobserved true positions.\\
@@
 After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy the of land polygons.
-After adjustement the land polygons using \texttt{qgis},  approximately $1 \%$ of the observations remained on land.
+We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons.
+After adjustement of the polygon data using \texttt{qgis},  approximately $1 \%$ of the observations remained on land.
 Distance to shore values range from $0$ to $7.6$ km.
@@
 
-The distance from the ship denoted $D^{ship}_{ij}$, is estimated as the distance between the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
+The distance from the ship denoted $D^{ship}_{ij}$, is estimated as the distance between observed GPS locations of the ship and the narwhals. The values are comprised between $2.68$ and $63.8$ km.
 Exposure to ship covariate for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
@@
 Intuitively, ~\ref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
- Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}.\\
+ Figure~\ref{fig: sample RACVM} shows one sample of a RCVM.\\
 
@@
 
-Unconstrained movement of marine mammals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, while other animals or organisms whose motion appear tortuous require the more general model \ref{eq: RACVM equation}.
-In the case of motion under spatial constraints, an increase in tortuosity enables the animals to avoid the boundary and adapt their heading to the constraint. 
+Unconstrained movement of marine mammals  is often modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, or a RACVM with low angular velocity $\omega$.
+In the case of motion under spatial constraints, we claim that an increase in tortuosity is the symptom of movement to avoid the boundary and adapt the heading to the spatial constraint.
+
 \subsection{Constrained rotational correlated velocity model}
@@
 where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the observed distance to the boundary of the domain $\partial \mathcal{D}$
-and $\Theta(t)$ is the observed angle between the velocity $V(t)$ and the shoreline as defined in subsection ~\ref{subsection: covariates}.
+and $\Theta(t)$ is the observed angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}.
 Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
@@
 \[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
-In this case, we can use the formulas of section~\ref{section: RACVM} to get an approximation of the exact transition density.
+In this case, we can use the formulas of section~\ref{section: RACVM} to get approximate transition densities.
 The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
@@
 It should always remain significantly lower than $D^{shore}$ for the process to be constrained.
-Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
+Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the animal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
 The closer we are to the shore, the faster the velocity should rotate,
@@
 	\caption{Examples of smooth functions $f_{\omega}$.
-Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.
-(a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega}.
-(c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivartiate cubic spline basis functions.}
-
-\end{figure}
-
- This general model is very flexible, and different species-specific behaviours could be described.
+Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
+(c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions.}
+
+\end{figure}
+
+ This general model is very flexible, and different species-specific behaviours can be described.
 The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn.
@@
 
- More precisely, to model persistent movement along the shore line, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
+ More precisely, to model persistent movement along the shoreline, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
  \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] where $f_{\tau}$ is a smooth function from $[-\pi,\pi]$ to $\R$.
@@
 
- We illustrate this flexibility in Figure~\ref{fig: crcvm examples}.Trajectories obtained within the fjords domain for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
-
- \begin{figure}[H]
-
+ We illustrate the flexibility of the CRCVMs in Figure~\ref{fig: crcvm examples}.
+Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
+
+ \begin{figure}[H]
+
+   	\hspace{2cm} 
  	 \begin{subfigure}{0.48\textwidth}
-		\includegraphics[scale=0.65]{images/crcvm/illustrative_sample_fjords}
+ 	 	\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_circ}
+ 	 	\label{fig:illustrativesamplecirc}
+ 	 \end{subfigure}
+
+   	\hspace{2cm} 
+  	 \begin{subfigure}{0.48\textwidth}
+  		\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_rect}
+  		\label{fig:illustrativesamplerect}
+ 	 \end{subfigure}
+
+ 	\hspace{2cm} 
+ 	\begin{subfigure}{0.48\textwidth}
+		\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_fjords}
 		\label{fig:illustrativesamplefjords}
@@
 
- 	\caption{Five day trajectories with time step $\Delta=5$ min in Scoresby Sound fjords domain}
+ 	\caption{Five day trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains.
+Plot labeled persistent and standard have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h.
+Trajectories labeled "tortuous" are obtained from a contant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
  	\label{fig: crcvm examples}
@@
 \section{Inference of the sound exposure effect}
+
+To analyse potential perturbations in the narwhals motion due to sound exposure, we formulate a mixed effect model for the parameters of the CRCVM, and discretise~\ref{eq: RACVM solution} to get an approximate state space model for inference with the Kalman filter.
 
@@
 
-%WHY TENSOR SPLINES : because covariates THeta and Eshore are not on the same scale, and identity link function
-% FOR TAU: this makes sense to use additive splines since the link function is log
-To assess if the exposure to the ship has a significative effect on the narwhals movement, a baseline model is first fitted only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the disturbance.
-Then, a response model is fitted on the data during exposure and check whether the estimated smooth parameters deviate significatively from the baseline values.
+First a baseline model is first fitted only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the airguns sound.
+Then, a response model is fitted on the data during exposure and we check whether the estimated smooth parameters deviate significatively from the baseline values.
 We present here the baseline and response models.\\
@@
 For the baseline model, only the effect of the shore is included through the shore exposure covariate $E^{shore}$ and the angle covariate $\Theta$.\\
-Although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance.
-This is taken into account by using a mixed effect model for the RCVM parameters.
-For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations $y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$ and the true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
-Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters.
-The angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $E^{shore}$ through splines.
-Since both covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\
-
-Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion.
-The complete statistical model for the baseline estimations is the following:
+We add random effects to account for individual variability among the narwhals.
+For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations before exposure $y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$.
+The true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
+Since $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters and the angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $E^{shore}$ through splines.
+Since the two covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\
+
+We denote $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion. The complete statistical model for the baseline estimations is the following:
 
@@
 The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
-Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$ are estimated from the trajectories during exposure to the sound
+Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$ are estimated from the trajectories during exposure to the sound.
 
@@
 \label{section: state space model}
-In most applications, the error between the observation $Y$ and the true position $X$ is not negligible, so that it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}.
+Since the GPS measurement error between the observation $y$ and the true position $X$ is not negligible, it is needed to include the observations in a state-space model as done by \cite{johnson_continuoustime_2008}.
 This is the approach we use here.
@@
 
-Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
+Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation.
 The complete likelihood is computed from the observations $y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
@@
 \\
-This is the technique is already available for the CVM, which corresponds to the special case $\omega=0$.
+This is technique is already available for the CVM, which corresponds to the special case $\omega=0$ in \texttt{smoothSDE}.
 Here, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models~\ref{eq: exact baseline model},~\ref{eq: exact response model}.
@@
 \label{section: simulation constrained motion}
-In order to experiment the constrained models we simulate a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain.
-We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}.\\
-
-We simulate $M=100$ batches of $N=6$ individual trajectories over $T=24$ h with time step $\Delta=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application..
-This amounts to $n_{i,pre}=1440$ observations for each individual $i \in \{1,\cdots,N\}$.
-We fix $\sigma_{obs}=5$ m to represent low measurement errors.
-We choose $L=8$ for the degree of freedom of the bivariate splines of $E^{shore}$ and $\Theta$.
-The random effects standard deviations are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu,pre}=0.1$ to represent the individual variability, and $\tau_{0}$ and $\nu_{0}$ are fixed respectively to $1$ h and $4$ km/h.
-The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance fjords 12h}.
-For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $1$ km away from the boundary.
-We kept the same $6$ initial positions for all the batches of trajectories.
-To obtain the trajectories of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step.
-The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}.
-None of the $600$ simulations reached the boundary.
+In order to experiment the constrained models we simulate a CRCVM in Scoresby Sound fjords system in which the narwhals move in our application.
+We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}.
+We set $L=9$ degrees of freedom for the bivariates splines in the parameter $\omega$ and  we place the knots regularly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $0.5$ and $5$ km for $D^{shore}$.
+The values of the spline coefficients are chosen before hand to get a process that meet the spatial constraints.
+We fix $\tau_0=1$ h and $\nu_0=4$ km/h.
+Random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$.
+Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released for the controlled exposure experiment.
 \\
+
+Two setups are considered: high frequency data with short trajectories and low frequency data with long trajectories.
+For each setup, we simulate $M=100$ batches of $N$ individual trajectories for $N \in \{6,12\}$.
+In the high frequency framework, the time step between consecutive observations is $\Delta=1$ min, the overall observation time is $12$h and measurement error is gaussian with standard deviation $\sigma_{obs}=5$ m.
+This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
+In the low frequency framework, the time step between consecutive observations is $\Delta=5$ min, the overall observation time is $2.5$ days and measurement error is gaussian with standard deviation $\sigma_{obs}=25$ m.
+This also amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
+
+None of the $600$ simulations reached the boundary. \\
 
@@
 		\hline
-			\multicolumn{14}{|c|}{Estimates for $d(X_0,\mathcal{D}) > 4$ km} \\
+			\multicolumn{14}{|c|}{High frequency} \\
+			\hline
+	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
+	& & & & & & & & & & & & & \\
+	\hline
+	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
+	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
+	0.09\\
+	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
+	0.08 & 0.03  \\
+	\hline	
+	\multicolumn{14}{|c|}{Low frequency} \\
 	\hline

### 2024-05-29 Latex/main.tex
updates

@@
 \usepackage{multirow}
+\usepackage{lscape}
 
@@
  	\caption{Five day trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains.
-Plot labeled persistent and standard have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h.
-Trajectories labeled "tortuous" are obtained from a contant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
+Plots labeled "persistent" and "standard" have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h.
+Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
  	\label{fig: crcvm examples}
@@
 \label{section: simulation constrained motion}
-In order to experiment the constrained models we simulate a CRCVM in Scoresby Sound fjords system in which the narwhals move in our application.
+In order to experiment estimation of the constrained models we simulate CRCVM trajectories in Scoresby Sound fjords system where the narwhals move in our application.
 We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}.
@@
 Random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$.
-Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released for the controlled exposure experiment.
+Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released during the controlled exposure experiment.
 \\
@@
 
-None of the $600$ simulations reached the boundary. \\
+A small percentage (around $2\%$) of all the simulated trajectories reached the shoreline, due to regions of acute corner in the polygon defining land.
+For the estimation, we only kept the trajectories that did not reached land and whose observed distance to the shore in km were included in the interval $[0.5,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$.
+\\
 
@@
 
-These $100$ batches of noisy samples are used as data for estimation.
-First, we estimate using only the first $12$h of the trajectories, that is $720$ observations for each narwhals.
-Then, we use the whole $24$h-long trajectory.
-The covariates $E^{shore}$ and $\Theta$ are computed from the observed noisy positions.
-It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging.
-The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}.
-The marginal spline degree of freedom is set to $3$ for each covariate to match the fixed degree of freedom $L=8$.
-This corresponds to one single knot in each dimension.
-Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
+These batches of noisy samples are used as data for estimation.
+It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging in the low frequency framework.
+For the smooth $\omega$, $9$ coefficients, including the intercept, are to be estimated, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
 Measurement error $\sigma_{obs}$ is supposed to be known.
 We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
-For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the root mean squared error and the biais of the estimates.
+For each batch of $N$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard deviation of the estimates.
 Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.
-The results are summarized in Tables~\ref{table: estimation performance fjords 12h} and~\ref{table: estimation performance fjords 24h}.
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with confidence intervals obtained from the observed Fisher information matrix.
-The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-The red curve represents the true value of $\omega$.}
-	\label{fig: marginal_estimates_CRCVM}
-\end{figure}
-
+The results are summarized in Table~\ref{table: simulation study results}.
+
+\begin{landscape}
 \begin{table}[H]
@@
 		\hline 
-		 True values & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
+		 True & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
 		\hline
 			\multicolumn{14}{|c|}{High frequency} \\
+			\hline
+	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
+	& & & & & & & & & & & & & \\
+	\hline
+	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
+	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
+	0.09\\
+	& $\pm$ 0.09 & $\pm$ 0.04 & $\pm$ 0.17 & $\pm$ 0.64 & $\pm$ 0.83 & $\pm$ 0.81 & $\pm$ 1.08 & $\pm$ 1.83 & $\pm$ 1.02 & $\pm$ 0.65 & $\pm$ 0.77 &
+	$\pm$ 0.08 & $\pm$ 0.03  \\
+	\hline	
+	\multicolumn{14}{|c|}{Low frequency} \\
 	\hline
@@
 	\hline	
-	\multicolumn{14}{|c|}{Low frequency} \\
-	\hline
-	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
-	& & & & & & & & & & & & & \\
-	\hline
-	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
-	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
-	0.09\\
-	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
-	0.08 & 0.03  \\
-	\hline	
 \end{tabular}
+\label{table: simulation study results}
+\caption{Simulation study results}
 \end{table}
+\end{landscape}
+
+In practical case, confidence intervals on the estimates are derived from the observed Fisher information matrix.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with these confidence intervals.
+The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+The red curve represents the true value of $\omega$.}
+	\label{fig: marginal_estimates_CRCVM}
+\end{figure}
 

### 2024-06-01 Latex/main.tex
Last paper version

@@
 
-\newpage
+\begin{abstract}
+Narwhals in Arctic seas are increasingly exposed to human activities that can temporarily or permanently threaten their survival by modifying their behavior.
+We collected GPS data from a population of narwhals in the Scoresby Sound fjord system in Southeast Greenland, which were exposed to ships and seismic airguns during a controlled experiment in 2018.
+We propose a new continuous-time correlated velocity model with varying coefficients that includes spatial constraints on movement and sound exposure effects through covariates.
+Our model, applied to the narwhal data, demonstrates that sound exposure can disturb narwhals motion up to tens of kilometers.
+Specifically, we found a $40\%$ and $25\%$ increase in velocity and a $50\%$ and $30\%$ drop in persistence at distances of 4 km and 10 km from the sound source respectively.
+\end{abstract}
 
@@
 Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source.
-We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\
+We follow \cite{gurarie_correlated_2017} and refer to~\eqref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\
 
@@
 Thus, constraints for the shore are needed in the location process $X$.
-However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
+However, this is not the case in model~\eqref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
 Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
@@
 
-Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}.
+Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \eqref{eq: RACVM equation}.
 This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
@@
 \begin{itemize}
-	\item Derivation of explicit formulas for the transition density of the "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
+	\item Derivation of explicit formulas for the transition density of the "Rotational Advective Correlated Velocity Model" (RACVM, \eqref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
 	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
@@
 We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons.
-After adjustement of the polygon data using \texttt{qgis},  approximately $1 \%$ of the observations remained on land.
+After adjustement of the polygon data using \texttt{qgis},  approximately $1 \%$ of the observations remain on land.
 Distance to shore values range from $0$ to $7.6$ km.
@@
 \end{equation}
-where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances.
-Thus we only consider the effect of the shore when the narwhals are close enough to it.
+where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=3$ km.
+Thus we only consider the effect of the shore when the narwhals are within a $3$ km range from the shore.
 The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$.
@@
 
-We write $A=\begin{pmatrix} 
+We denote $A=\begin{pmatrix} 
 	\frac{1}{\tau} & -\omega \\
 	\omega & \frac{1}{\tau}
-\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}
+\end{pmatrix}$ and rewrite~\eqref{eq: RACVM equation}
 
@@
 \end{equation*}
-where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
-In this section, we obtain explicit distributional results about the process $X$.
+In this section, we obtain explicit transition density for the process $U=\begin{pmatrix} X & V \end{pmatrix}^\top$.
 Though this model is considered in several papers, we found no mention of the closed form formulas derived here.
@@
 But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
-Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
+Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\eqref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
 \\
@@
 
-Equation~\ref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
+Equation~\eqref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
 Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
-Intuitively, ~\ref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
+Intuitively, ~\eqref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
  Figure~\ref{fig: sample RACVM} shows one sample of a RCVM.\\
@@
 	\centering
-	\includegraphics[scale=0.5]{images/rcvm/illustrative_sample_standard.png}
+	\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
 	\caption{Trajectory of a RCVM with parameters $\tau=1$h, $\nu=4$ km/h, $\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min.
+Values of $x$ and $y$ are in km.
 The red cross indicates the starting point of the process.
@@
 
-Unconstrained movement of marine mammals  is often modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, or a RACVM with low angular velocity $\omega$.
+Unconstrained movement of marine mammals  is often modeled through a CVM, that is~\eqref{eq: RACVM equation} with $\omega=0$, or a RACVM with low angular velocity $\omega$.
 In the case of motion under spatial constraints, we claim that an increase in tortuosity is the symptom of movement to avoid the boundary and adapt the heading to the spatial constraint.
@@
 
-We thus propose to make use of a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to include how the animals turn in reaction to the shore. 
-To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
+We thus propose a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to describe how the animals turn in reaction to the shore. 
+To do so, we consider $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
 The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
@@
 \end{equation}
-where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the observed distance to the boundary of the domain $\partial \mathcal{D}$
-and $\Theta(t)$ is the observed angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}.
-Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.
+where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$ with  $f_{\omega}$ a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$ meeting minimum requirements.
+In the sequel, such models will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).\\
 
@@
 \[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
-In this case, we can use the formulas of section~\ref{section: RACVM} to get approximate transition densities.
+In this case, we use the formulas of section~\ref{section: RACVM} to get approximate transition densities.
 The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
@@
 An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
-Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions.
-In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).
+Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show function~\ref{eq: smooth omega}, while Figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represent a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions approximating~\ref{eq: smooth omega}.
+Sampled trajectories of these models are shown in Figure~\ref{fig: crcvm examples}.
 
@@
 	\caption{Examples of smooth functions $f_{\omega}$.
-Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
+Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
 (c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions.}
@@
 The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn.
-We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, observed distances and angles.
+We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles.
 Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question.
 Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination.
-Again, this behavior can be statistically assessed from the observations.
-
- More precisely, to model persistent movement along the shoreline, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
- \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] where $f_{\tau}$ is a smooth function from $[-\pi,\pi]$ to $\R$.
+More precisely, to model persistent movement along the shoreline, we modify~\eqref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
+ \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] with $f_{\tau}:[-\pi,\pi] \longrightarrow \R$.
  The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain.
 An example of such a function is given in figure~\ref{fig:smoothtaubump}.
- \begin{figure}[H]
+ \begin{figure}[h]
  	\centering
@@
 
- We illustrate the flexibility of the CRCVMs in Figure~\ref{fig: crcvm examples}.
+ The flexibility of CRCVMs is illustrated in Figure~\ref{fig: crcvm examples}.
 Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
@@
 
-   	\hspace{2cm} 
+   	\hspace{3cm} 
  	 \begin{subfigure}{0.48\textwidth}
@@
 
-   	\hspace{2cm} 
+   	\hspace{3cm} 
   	 \begin{subfigure}{0.48\textwidth}
@@
 
- 	\hspace{2cm} 
+ 	\hspace{3cm} 
  	\begin{subfigure}{0.48\textwidth}
@@
 
- 	\caption{Five day trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains.
-Plots labeled "persistent" and "standard" have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h.
-Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
+ 	\caption{Five days trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains.
+Plots labeled "persistent" and "standard" have the same smooth parameter $\omega$ defined by~\eqref{eq: smooth omega}, but the persistent trajectory has a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$ h.
+Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\omega$ defined with spline basis function as in Figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
  	\label{fig: crcvm examples}
@@
 
-To analyse potential perturbations in the narwhals motion due to sound exposure, we formulate a mixed effect model for the parameters of the CRCVM, and discretise~\ref{eq: RACVM solution} to get an approximate state space model for inference with the Kalman filter.
+To analyse potential perturbations in the narwhals motion due to sound exposure, we formulate a mixed effect model for the parameters of the CRCVM, and discretise~\eqref{eq: RACVM solution} to get an approximate state space model for inference with the Kalman filter.
 
@@
 
-First a baseline model is first fitted only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the airguns sound.
-Then, a response model is fitted on the data during exposure and we check whether the estimated smooth parameters deviate significatively from the baseline values.
+First a baseline model is fitted only on the data before exposure, to get an estimation on how the narwhals move when they are not exposed to the airguns sound.
+Then, a response model is fitted on the data during exposure and we test whether the estimated smooth parameters deviate significantly from the baseline values.
 We present here the baseline and response models.\\
@@
 For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations before exposure $y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$.
-The true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}.
+The true motion processes $V_i$ and $X_i$ are supposed to solve equation~\eqref{eq: CRCVM equation}.
 Since $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters and the angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $E^{shore}$ through splines.
@@
 
-The functions $\psi_k$ in equation~\ref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions.
+The functions $\psi_k$ in equation~\eqref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions.
 We refer to \cite{wood_generalized_2017} section $4.
@@
 
-Then, the response model is designed to assess a deviation from the baseline model due to exposure to the ship.
+The response model is then designed to assess a deviation from the baseline model due to exposure to the ship.
 We illustrate this by addding a dependency on the covariate $E_{ship}$ in the parameters $\tau$ and $\nu$.
@@
 
-The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are considered are offset.
-The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
-Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$ are estimated from the trajectories during exposure to the sound.
+The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are considered as offsets.
+The functions $\phi_k$ are univariate cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.
+Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$, $k \in \{1,\cdots,q\}$ are estimated from the trajectories during exposure to the sound.
 
@@
 \label{section: state space model}
-Since the GPS measurement error between the observation $y$ and the true position $X$ is not negligible, it is needed to include the observations in a state-space model as done by \cite{johnson_continuoustime_2008}.
+To consider the GPS measurement error between the observation $y$ and the true position $X$, it is needed to include the observations in a state-space model as done by \cite{johnson_continuoustime_2008}.
 This is the approach we use here.
-In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
+We consider i.i.d gaussian measurement errors with variance $\sigma_{obs}^2$.
+In the statistical model~\eqref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
 \[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(E^{shore}_{ij},\Theta_{ij})\]
@@
 
-Discretisation of~\ref{eq: X V distribution} gives the following state-space equations:
+Discretisation of~\eqref{eq: X V distribution} gives the following state-space equations:
 
@@
 \end{equation}
-where $Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix}$, the link matrices $T_{ij}$ and $B_{ij}$ are computed according to~\ref{eq: link matrices} and the covariance matrix is computed according to~\ref{eq: covariance matrix}.
+where $Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix}$, the link matrices $T_{ij}$ and $B_{ij}$ are computed according to~\eqref{eq: link matrices} and the covariance matrix $Q_{ij}$ is computed according to~\eqref{eq: covariance matrix}.
 \\
@@
 \\
-This is technique is already available for the CVM, which corresponds to the special case $\omega=0$ in \texttt{smoothSDE}.
-Here, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models~\ref{eq: exact baseline model},~\ref{eq: exact response model}.
+This method is already implemented for the CVM, which corresponds to the special case $\omega=0$, in \texttt{smoothSDE}.
+Here, we extend this inference method to the more general model~\eqref{eq: CRCVM equation} and use it for estimation of the baseline and the response models~\eqref{eq: exact baseline model},~\eqref{eq: exact response model}.
 
@@
 In order to experiment estimation of the constrained models we simulate CRCVM trajectories in Scoresby Sound fjords system where the narwhals move in our application.
-We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}.
-We set $L=9$ degrees of freedom for the bivariates splines in the parameter $\omega$ and  we place the knots regularly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $0.5$ and $5$ km for $D^{shore}$.
+We consider observations of a baseline model as described in~\eqref{eq: exact baseline model}.
+We set $L=9$ degrees of freedom for the bivariate splines in the parameter $\omega$ and  we place the knots evenly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $0.5$ and $5$ km for $D^{shore}$.
 The values of the spline coefficients are chosen before hand to get a process that meet the spatial constraints.
-We fix $\tau_0=1$ h and $\nu_0=4$ km/h.
+We fix $\tau_0=1.5$ h and $\nu_0=4$ km/h.
 Random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$.
-Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released during the controlled exposure experiment.
+Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $4$ km away from the shore.
 \\
 
-Two setups are considered: high frequency data with short trajectories and low frequency data with long trajectories.
-For each setup, we simulate $M=100$ batches of $N$ individual trajectories for $N \in \{6,12\}$.
+Two setups are considered: high frequency data with short trajectories and low frequency data with longer trajectories.
+For each setup, we simulate $M=100$ batches of $N$ individual trajectories with $N \in \{6,12\}$.
 In the high frequency framework, the time step between consecutive observations is $\Delta=1$ min, the overall observation time is $12$h and measurement error is gaussian with standard deviation $\sigma_{obs}=5$ m.
@@
 A small percentage (around $2\%$) of all the simulated trajectories reached the shoreline, due to regions of acute corner in the polygon defining land.
-For the estimation, we only kept the trajectories that did not reached land and whose observed distance to the shore in km were included in the interval $[0.5,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$.
-\\
-
-\begin{comment}Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often.
-For the fjords system, this already happens if we try to sample directly with $5$ min time steps.
-Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.
-\end{comment}
-
+For the estimation, we only kept the trajectories that did not reach land and whose observed distances to the shore in km were included in the interval $[0.5,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$.
 These batches of noisy samples are used as data for estimation.
@@
 For each batch of $N$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard deviation of the estimates.
-Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h.
-The results are summarized in Table~\ref{table: simulation study results}.
+Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ CPU cores within less than $2$ h.
+The results are summarized in Table~\ref{table: simulation study results}.\\
+
+Estimation of the spline coefficients has overall less than $20 \%$ biais, and becomes more precise as $N$ increases.
+The variances of the random effects are the most challenging parameters to estimate in the model.
+At least one of the standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$ is often estimated close to $0$.
+This tendency is clear for $N=6$ and $N=12$ individuals, though it seems to attenuate as $N$ increases.
+
+In practical cases, confidence intervals on the estimates are derived from the observed Fisher information matrix.
+Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with these confidence intervals.
+The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[scale=0.25]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
+		\caption{}
+	\end{subfigure}
+	\caption{Estimates of the smooth parameter $\omega$ for one batch of high frequency trajectories as a function of $\Theta$ only, with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
+The red curve represents the true value of $\omega$.}
+	\label{fig: marginal_estimates_CRCVM}
+\end{figure}
 
 \begin{landscape}
-\begin{table}[H]
+
+\begin{table}
 	\centering
@@
 		\hline 
-		 True & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
+		 True & $0.4$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
 		\hline
@@
 			\hline
-	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
-	& & & & & & & & & & & & & \\
-	\hline
-	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
+	\multirow{2}{4em}{$N=6$} & 0.38 & 1.44 & 0.00 & 6.01 &
+	-1.77 & 3.66 & 14.77 & 3.80 & 19.41 & 3.21
+	 & 3.99 & 0.11 &
+	 0.06 \\
+	&	$\pm0.17$ & $\pm 0.09$ & $\pm 0.20$ & $\pm 1.20$ &$\pm1.255$ & $\pm0.72$ & $\pm1.82$ & $\pm2.68$ & $\pm 1.38$ & $\pm1.19$ & $\pm1.26$ & $\pm 0.08$ & $\pm0.04$
+	 \\
+	\hline
+	\multirow{2}{4em}{$N=12$} & 0.42 & 1.38 & -0.02
 	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
@@
 	\hline
-	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
-	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
-	0.09\\
-	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
-	0.08 & 0.03  \\
+	\multirow{2}{4em}{$N=12$} &  &  & 
+	& &  & & & & & & & & \\
+	&  & &  & & & &  & & & & &
+	 &  \\
 	\hline	
 \end{tabular}
+\caption{Simulation study results}
 \label{table: simulation study results}
-\caption{Simulation study results}
 \end{table}
+
 \end{landscape}
 
-In practical case, confidence intervals on the estimates are derived from the observed Fisher information matrix.
-Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with these confidence intervals.
-The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.
-
-\begin{figure}[H]
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
-		\caption{}
-	\end{subfigure}
-	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances.
-The red curve represents the true value of $\omega$.}
-	\label{fig: marginal_estimates_CRCVM}
-\end{figure}
-
 \section{Application to narwhals behavioral response}
+\label{section: application}
 
@@
 It is supposed to capture the characteristics of the movement under normal conditions.
-Deviation from the baseline is then assessed by fitting a model with covariate $E^{ship}$ on the tracks during exposure.
+Deviation from the baseline is then assessed by fitting a model with covariate $E^{ship}$ on the tracks after exposure.
 
@@
 
-The persistence is estimated to $\hat{\tau}_{0}=1.3 $ h.
+Estimation is performed in \texttt{smoothSDE} by defining a template model for the complete negative  log-likelihood with \texttt{TMB}.
+The choice of initial parameters is crucial in the optimization.
+Based on the results on measurement errors of Fastloc-GPS found in \cite{wensveen_path_2015}, we initialize the measurement error standard deviation to $\sigma_{obs}=50$ m.
+Then we estimate model ~\eqref{eq: exact baseline model} on all the available data with degree of freedom of the marginal splines set to $5$ and we use the estimated spline coefficients as initial values to fit the baseline model on the data before exposure.
+We initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=0.2$ and fix the smoothing penalties of the tensor splines to $1$.
+
+Estimation and standard error on the parameter scale are shown in table~\ref{table: baseline estimations}. 
+The persistence is estimated to $\hat{\tau}_{0}=1.08 $ h.
 In comparison, harbour seal in Alaska were proved to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \cite{johnson_continuoustime_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \cite{gurarie_correlated_2017}.
-The intercept value for $\nu$ is estimated to be $\hat{\nu}_{0}=4.19$ km/h with $95\%$ CI $3.79-4.63$.
+The intercept value for $\nu$ is estimated to be $\hat{\nu}_{0}=4.61$ km/h which is of the order of magnitude of the mean of the observed empirical velocities.
 \begin{table}[H]
@@
 		\hline
-		Coefficient   & Estimate  & $95\%$ CI \\
-		\hline
-		$\log(\tau_0)$   & 0.3    &  \\
-		$\log(\nu_0)$  & 1.48  &  \\
-		$\omega_0$     & -0.13     &  \\
-		$\sigma_{\tau,pre}$   & 0.1   &   \\
-		$\sigma_{\nu,pre}$    & 0.1   &  \\
+		Coefficient   & Estimate  & Standard error \\
+		\hline
+		$\tau_0$ (h)   & $1.08$    &  $0.16$ \\
+		$\nu_0$ (km/h)  & $4.64$   &  $0.25$\\
+		$\omega_0$ (rad/h)    & $0.01$    &  $0.08$ \\
+		$\sigma_{\tau,pre}$    & 0.29   &  $0.12$ \\
+		$\sigma_{\nu,pre}$    & 0.10   &  $0.05$ \\
+		$\sigma_{obs}$ (m) & $35$ & $1$ \\
 		\hline
 	\end{tabular}
-	\caption{Intercepts baseline estimations}
-	\label{table: baseline estilations}
+	\caption{Baseline estimations}
+	\label{table: baseline estimations}
 \end{table}
@@
 Figure \ref{fig:febaseline3omegaanglenormalq0} shows the estimated smooth $\omega$ as a function of $\Theta$ for different values of the distance to shore.
-As distance increases, the magnitude of the angular velocity decreases.
-% PUT ON THE SAME PLOT WITH DIFFERENT COLOR FOR EACH CURVE
-\begin{figure}[H]
-	\begin{subfigure}{0.33\textwidth}
-		\centering
-		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.95_ExpShore .png"}
-	\end{subfigure}
-	\begin{subfigure}{0.33\textwidth}
-		\centering
-		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.75_ExpShore .png"}
-	\end{subfigure}
-	\begin{subfigure}{0.33\textwidth}
-		\centering
-		\includegraphics[width=0.95\linewidth]{"images/application/baseline/fe_baseline3_omega_AngleNormal_q0.25_ExpShore .png"}
-	\end{subfigure}
+Movement toward the shoreline matches values of $\Theta$ such that $\vert \Theta \vert > \frac{\pi}{2}$, for which the angular velocity $\omega$ is estimated to be close to $2$ rad/h for low distances to shore.
+As distance increases, the magnitude of the angular velocity decreases, which means that the movement becomes less tortuous.
+This indicates that the spatial constraints influence the tortuosity of the movement by forcing the narwhals velocity to rotate when they are close to the shoreline.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\linewidth]{"images/application/baseline/omega_DistanceShore_levels.png"}
+	\caption{Estimated smooth $\omega$ as a function of the angle $\Theta$ for fixed exposures to the shore : $250$ metres for "close", $1.41$ km for medium and zero exposure, that is distance higher than $3$ km, for "far".}
+
+	\label{fig:febaseline3omegaanglenormalq0}
+\end{figure}
+
+\subsection{Response estimations}
+
+The baseline estimates of the spline coefficients $\omega_k$, the intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the standard deviations $\sigma_{\tau}$, $\sigma_{\nu}$ are introduced as offsets in the response model.
+Only the effect of the exposure to the sound on $\tau$ and $\nu$ is estimated from the data after exposure.\\
+
+Figure~\ref{fig: me response} shows that exposure to sound has an effect on both the persistence and the velocity of the movement.
+The value of $\nu$ increases with increased exposure to the sound, suggesting motion at higher speed in presence of the ship.
+This should not come as a surprise since, for instance, the mean experimental velocity norm when the narwhals are less than $5$ km away from the ship is $5.7$ km/h, which is more than $1$ km/h higher than the mean experimental velocity norm before exposure.
+We draw specific attention to the parameter $\tau$, that decreases with increased exposure to the ship, implying lower persistence and lower autocorrelation in the velocity of the narwhals.
+This is a new finding that had been hinted in \cite{heide-jorgensen_behavioral_2021} since it was acknowledged that the narwhals had more tendency to change direction and move toward the shore in the presence of the ship.
+This can now be interpreted as a drop in persistence due to the appearance of the ship.
+Our analysis suggests that the magnitude of this drop might be alarming : for most of the narwhals, at distance less than $5$ km away from the ship, the persistence is half the baseline value.
+This shows evidence of a strong shift in the behavior of the whales, which is likely to impede their capacity to mate and forage at short term.
+The baseline value is attained for a distance to the sound source higher than $20$ km for $\tau$, and  $40$ km for $\nu$.
+This indicates that sound exposure can be perceived and can disturb the narwhals motion up to $40$ km.
+
+ \begin{figure}[H]
+ 	\begin{subfigure}{0.98\textwidth}\centering
+ 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_tau_ExpShip.png"}
  		\caption{}
-
-	\label{fig:febaseline3omegaanglenormalq0}
-\end{figure}
-
-\subsection{Response estimations}
+ 	\end{subfigure}
+
+ 	\begin{subfigure}{0.98\textwidth}
+ 		\centering
+ 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_nu_ExpShip.png"}
+ 		\caption{}
+ 	\end{subfigure}
+ 	\caption{Estimated effect of ship exposure on the parameters.
+The horizontal dashed lines represent the baseline values for each individual.
+The values on the $x$-axis are in kilometres.
+(a)  Smooth parameter $\tau$.
+Values on the $y$-axis are in hour.
+(b) Smooth parameter $\nu$.
+Values on the $y$-axis are in km/h.}
+ 	\label{fig: me response}
+ \end{figure}
+
+We categorize narwhals position relative to the ship as follows :
+\begin{itemize}
+	\item very close when the distance to the ship is lower than the $5\%$. quantile, which is $4.04$ km.
+	\item close when the distance to the ship is between $5\%$. quantile and the $25\%$ quantile, which is $9.35$ km.
+	\item medium when the distance to the ship is between the $25\%$ and $75\%$ quantile, that is between $9.35$ and $25.80$ km.
+	\item far otherwise.
+\end{itemize}
+Table~\ref{table: baseline vs response parameters comparison} shows the changes in persistence and mean velocity parameters $\tau$ and $\nu$ for the four categories.
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+		\hline
+		Position relative to the ship & mean population $\hat{\tau}$  & mean population $\hat{\nu}$ & $\frac{\hat{\tau}-\hat{\tau}_{0}}{\tau_{0}}$ & $\frac{\hat{\nu}-\hat{\nu}_{0}}{\hat{\nu}_{0}}$   \\
+		\hline
+		Very close & 0.54 & 6.4 & $-50\%$ & $+39\%$ \\
+		\hline
+		Close & 0.74 & 5.8 & $-31\%$ & $+26\%$ \\
+		\hline
+		Medium&  0.96 &  5.0 & $-11\%$ & $+9\%$\\
+		\hline
+		Far &  1.07 & 4.7 &  $-1\%$ & $+2\%$\\
+		\hline
+	\end{tabular}
+	\caption{Comparison of response and baseline values of the parameters for different levels of exposure to the ship}
+	\label{table: baseline vs response parameters comparison}
+\end{table}
 
@@
 
+\begin{comment}
 \textcolor{red}{Further work: 
@@
 
-\begin{comment}
 As discussed in \cite{ditlevsen_hypoelliptic_2019}, estimation for such hypoelliptic models is challenging. The main difficulties come from
@@
 	\centering
-	\includegraphics[scale=0.5]{images/data_exploration/all_time_steps_histo}
+	\includegraphics[scale=0.4]{images/data_exploration/all_time_steps_histo}
 	\caption{Histogram of time steps}
@@
 Figure~\ref{fig: trials and intertrials distributions} shows how the exposure periods are distributed among the $6$ narwhals that were tracked.
-
-	\begin{figure}[H]
-		\centering
-			\centering
-			\includegraphics[width=0.9\textwidth]{images/data_exploration/trials.png}
+Our analysis in section \ref{section: application} does not distinguish between intertrial and trial periods.
+They are both treated as exposure periods, though the nature and intensity of the behavioral response might differ for the two periods.
+The lack of intertrial data as well as a potential persistence in time of the behavior shift due to airguns exposure during trial periods justify the use of this simple model.
+
+	\begin{figure}[H]
+		\centering
+			\centering
+			\includegraphics[scale=0.5]{images/data_exploration/trials.png}
 			\caption{Trial and Intertrial periods for each narwhal}
@@
  Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals.
-Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure on a map of Scoresby Sound fjords system.
+Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure with a base map of Scoresby Sound fjords system.
 
@@
 	\end{subfigure}
-	\caption{Movement data of East Greenland narwhals}
+	\caption{Movement data of East Greenland narwhals. The red crosses indicate initial positions.}
 	\label{fig: tracks before and after exposure}
@@
 		\hline
-		$D_{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
-		\hline
-		$T_{ship}(t)$  & categorical & $0$ when there is no trial, $1$ if it is an intertrial period, and $2$ when it is a trial period & $\{0,1,2\}$\\
-		\hline
-		$E_{ship}(t)=\frac{1}{D_{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
-		\hline
-		$E_{ship,I}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=1}$& $km^{-1}$ & intertrial exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
-		\hline
-		$E_{ship,T}(t)=\frac{1}{D_{ship}(t)}\mathbb{1}_{T_{ship}(t)=2}$ & $km^{-1}$ & trial exposure level of the narwhal to the ship disturbance at time $t$. & $\R^+$\\
-		\hline
-		$D_{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
-		\hline
-		$E_{shore}(t)=\frac{1}{D_{shore}(t)} \1_{D_{shore}(t)<D_0}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
-		\hline
-		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical step vector at time $t$  & $[-\pi,\pi]$ \\
+		$D^{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
+		\hline
+		\hline
+		$E^{ship}(t)=\frac{1}{D^{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
+		\hline
+		$D^{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
+		\hline
+		$E^{shore}(t)=\frac{1}{D^{shore}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
+		\hline
+		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical velocity vector at time $t$  & $[-\pi,\pi]$ \\
 		\hline 
@@
 Writing $\beta=\frac{1}{\tau}$ and reorganizing the terms, we obtain
-\begin{equation}q_1(\Delta)=\frac{\sigma^2}{\beta^2}\left(\Delta -2 \frac{1-\exp(-\beta \Delta)}{\beta}+\frac{1-e^{-2\beta \Delta}}{2\beta}\right)
+\begin{equation}q_1(\Delta)=\frac{\sigma^2}{\beta^2}\left(\Delta -2 \frac{1-\exp(-\beta \Delta)}{\beta}+\frac{1-\exp(-2\beta \Delta)}{2\beta}\right)
 \end{equation}
@@
 
-\section{More results about the standard RACVM}
-\subsection{Stationary distribution}
-
-If $v(0) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ then for all $t \geq 0$, $v(t) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$
-
-\subsection{Distribution of the velocity norm and phase}
- The order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm.
-The distribution of the phase angle $\Phi$ of the velocity is derived in \cite{pawula_distribution_1982}.
-When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$.
-Hence, taking a look at the histograms of the phase angle of the data can help to assess whether a non zero mean velocity is necessary in the model or not.
-
-Suppose that $t\geq \frac{3}{\beta}$ so that the velocity has reached its stationary distribution
+\section{Some distributional results about the standard RACVM}
+
+If $v(0) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ then for all $t \geq 0$, $v(t) \sim \mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$.
+This is the stationary distribution of the velocity process.
+
+Suppose that $t\geq 3\tau$ so that the velocity has reached its stationary distribution
 \begin{align*}
@@
 		% Add node beside the circle with text $\frac{\sigma^2}{2\beta}$
-		\node[blue] at (4.5,3) [anchor=west] {$\frac{\sigma^2}{2\beta}$};
+		\node[blue] at (4.5,3) [anchor=west] {$\frac{2\nu^2}{\pi}$};
 	\end{tikzpicture}
@@
 \[\E(\vert \vert v(t) \vert \vert)=\nu L_{\frac{1}{2}}(0)=\nu I_{0}(0)=\nu\]
-This result can be used to choose initial value of the parameter $\nu$ based on the mean observed empirical velocity.
-
-Then, in the plane, $v(t)$ can be thought of as a complex random variable $v(t)=v_1(t)+iv_2(t)$ that has complex normal distribution with mean $\mu_1+i\mu_2$, variance $\E\left((v(t)-\E(v(t)))\overline{(v(t)-\E(v(t)))}\right)=\frac{4\nu^
+This result can be used to choose initial value of the parameter $\nu$ based on the mean observed empirical velocity. The paremeter $\nu$ is interpreted as the mean velocity norm of the movement.
+
+In the plane, $v(t)$ can be thought of as a complex random variable $v(t)=v_1(t)+iv_2(t)$ that has complex normal distribution with mean $\mu_1+i\mu_2$, variance $\E\left((v(t)-\E(v(t)))\overline{(v(t)-\E(v(t)))}\right)=\frac{4\nu^
 	2}{\pi}$ and pseudo-variance $\E\left((v(t)-\E(v(t)))^2\right)=0$.
@@
 Denote $\Phi=\arctan(v_2(t),v_1(t)) \in [-\pi, \pi]$ the phase angle of $v(t)$.
-According to \cite{letzepis_von_2015}, the distribution of $\Phi$ is 
+According to \cite{letzepis_von_2015} and \cite{pawula_distribution_1982}, the distribution of $\Phi$ is 
 \[\forall \phi \in [-\pi,\pi], \quad f_{\Phi}(\phi)=\frac{e^{-\frac{\beta \vert \vert \mu \vert \vert^2}{2\sigma^2}}}{2\pi} +\frac{\sqrt{\beta} \vert \vert \mu \vert \vert}{2\sigma\sqrt{2\pi}} \cos(\phi-\theta)e^{-\frac{\beta\vert \vert \mu \vert \vert^2 \sin^2(\phi-\theta)}{2\sigma^2}} \times \left(1+\erf\left(\frac{\beta^{\frac{1}{4}}\vert \vert \mu \vert \vert \cos(\phi-\theta)}{\sqrt{2\sigma}}\right)\right)\]
@@
 
-\subsection{Velocity autocovariance function}
-
-Suppose that $v(0)$ has the stationary distribution.The autocovariance function is the expected dot product of the velocity vectors over different lags.
-
-\[\mathbb{E}(\langle v(t), v(t+\Delta) \rangle =Cov(v_1(t)v_1(t+\Delta))+Cov(v_2(t),v_2(t+\Delta))+\mathbb{E}(v_1(t)) \mathbb{E}(v_1(t+\Delta))+\mathbb{E}(v_2(t)) \mathbb{E}(v_2(t+\Delta))\]
-
-Therefore, the velocity autocovariance function is
-
-\begin{equation}
-	\mathbb{E}(\langle v(t), v(t+\Delta) \rangle )=\frac{4\nu^2}{\pi} \cos(\omega \Delta) e^{-\frac{\Delta}{\tau}}+\vert \vert \mu \vert \vert^2
-	\label{eq: RACVM autocovariance function}
-\end{equation}
-
 \end{document}

### 2024-06-02 R/preprocessing/data_preprocessing.Rmd
code updates

@@
 for (id in unique(data$ID)) {
-  narwhalDataBE=dataBE1[dataBE1$ID==id,]
+  narwhalDataBE=dataBE2[dataBE2$ID==id,]
   narwhalDataAE=dataAE[dataAE$ID==id,]

### 2024-06-02 Latex/main.tex
paper updates

@@
 The probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases.
-Additionally, an overall speed increase due to sound exposure is revealed in the 2018 a data by an analysis of the consecutive GPS positions through a mixed effect model.
+Additionally, an overall speed increase due to sound exposure is revealed for the 2018 data by an analysis of the consecutive GPS positions through a mixed effect model.
 Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one holistic model.
@@
 
-Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \eqref{eq: RACVM equation}.
+Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in \eqref{eq: RACVM equation}.
 This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
@@
 The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.
-While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.
-
-Experiments were divided into unexposed periods - for which the narwhals are not in line of sight with the ship - trial periods - when the narwhals are exposed to the ship and airguns are shot - and intertrial periods - when the narwhals are exposed to the ship but airguns are not shot.
-These periods are indicated by a categorical variable $T_{ship}$ in the dataset.
-In this analysis, we do not distinguish between Trial and Intertrial periods, and treat them both as exposure periods, where the degree of exposure depends on the narwhals positions relative to the ship.
+While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.\\
 
 For each narwhal, the entire track is split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time.
-We keep only GPS measures 24
-hours after tagging, to avoid any tagging effects on the narwhals's movement, as recommended in \cite{heide-jorgensen_behavioral_2021}.
+We keep only GPS measures 12
+hours after tagging, to avoid any tagging effects on the narwhals's movement.
 Measures that result in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
-Overall, $4192$ GPS positions are kept for the analysis and projected to UTM coordinates zone 26.
-The splitting between data before and after exposure results in $935$ measures before exposure and $3257$ measures after exposure.
+Overall, $4815$ GPS positions are kept for the analysis and projected to UTM coordinates zone 26.
+The splitting between data before and after exposure results in $1558$ measures before exposure and $3257$ measures after exposure.
 
@@
 In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
-A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
+A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
 In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.
@@
 In the case of motion under spatial constraints, we claim that an increase in tortuosity is the symptom of movement to avoid the boundary and adapt the heading to the spatial constraint.
+We use this remark to define a constrained version of the standard RCVM.
 
@@
 
-We thus propose a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to describe how the animals turn in reaction to the shore. 
+We propose a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to describe how the animals turn in reaction to the shore. 
 To do so, we consider $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
@@
 An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions.
-Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show function~\ref{eq: smooth omega}, while Figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represent a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions approximating~\ref{eq: smooth omega}.
+Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show function~\eqref{eq: smooth omega}, while Figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represent a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions approximating~\eqref{eq: smooth omega}.
 Sampled trajectories of these models are shown in Figure~\ref{fig: crcvm examples}.
@@
 	\caption{Examples of smooth functions $f_{\omega}$.
-Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
-(c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions.}
+Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.
+(a)-(b) Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
+(c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions approximating~\eqref{eq: smooth omega} .}
 
@@
 The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn.
-We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles.
+We will show in section~\ref{section: simulation constrained motion} that it can be estimated from the observed positions, distances and angles.
 Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question.
@@
 
- The flexibility of CRCVMs is illustrated in Figure~\ref{fig: crcvm examples}.
-Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
-
  \begin{figure}[H]
@@
 Plots labeled "persistent" and "standard" have the same smooth parameter $\omega$ defined by~\eqref{eq: smooth omega}, but the persistent trajectory has a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$ h.
-Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\omega$ defined with spline basis function as in Figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
+Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\omega$ defined with spline basis function as in Figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.
+All trajectories have constant velocity parameter $\nu=4$ km/h.}
  	\label{fig: crcvm examples}
  \end{figure}
+
+The flexibility of CRCVMs is illustrated in Figure~\ref{fig: crcvm examples}.
+Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.
 
@@
 
-We denote $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion. The complete statistical model for the baseline estimations is the following:
+We denote $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion for the narwhal $i$. The complete statistical model for the baseline estimations is the following:
 
@@
 The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$, $\omega_{0}$ are population intercepts and $b_{\tau,i}$, $b_{\nu,i}$
-are the individual parameters.
+are the individual parameters. THe standard deviation $\sigma_{obs}$ represent GPS measurement errors.
 
@@
 This is the approach we use here.
-We consider i.i.d gaussian measurement errors with variance $\sigma_{obs}^2$.
 In the statistical model~\eqref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
@@
 Figure \ref{fig:febaseline3omegaanglenormalq0} shows the estimated smooth $\omega$ as a function of $\Theta$ for different values of the distance to shore.
-Movement toward the shoreline matches values of $\Theta$ such that $\vert \Theta \vert > \frac{\pi}{2}$, for which the angular velocity $\omega$ is estimated to be close to $2$ rad/h for low distances to shore.
+Movement toward the shoreline matches values of $\Theta$ such that $\vert \Theta \vert > \frac{\pi}{2}$, for which the angular velocity $\omega$ is estimated to be close to $\pm 2$ rad/h for low distances to shore.
 As distance increases, the magnitude of the angular velocity decreases, which means that the movement becomes less tortuous.
@@
 The value of $\nu$ increases with increased exposure to the sound, suggesting motion at higher speed in presence of the ship.
-This should not come as a surprise since, for instance, the mean experimental velocity norm when the narwhals are less than $5$ km away from the ship is $5.7$ km/h, which is more than $1$ km/h higher than the mean experimental velocity norm before exposure.
+This should not come as a surprise since, for instance, the mean empirical velocity norm when the narwhals are less than $5$ km away from the ship is $5.7$ km/h, which is more than $1$ km/h higher than the mean experimental velocity norm before exposure.
 We draw specific attention to the parameter $\tau$, that decreases with increased exposure to the ship, implying lower persistence and lower autocorrelation in the velocity of the narwhals.
@@
 	\item very close when the distance to the ship is lower than the $5\%$. quantile, which is $4.04$ km.
-	\item close when the distance to the ship is between $5\%$. quantile and the $25\%$ quantile, which is $9.35$ km.
+	\item close when the distance to the ship is between $5\%$ quantile and the $25\%$ quantile, which is $9.35$ km.
 	\item medium when the distance to the ship is between the $25\%$ and $75\%$ quantile, that is between $9.35$ and $25.80$ km.
@@
 
+Experiments were divided into unexposed periods - for which the narwhals are not in line of sight with the ship - trial periods - when the narwhals are exposed to the ship and airguns are shot - and intertrial periods - when the narwhals are exposed to the ship but airguns are not shot.
+These periods are indicated by a categorical variable $T_{ship}$ in the dataset.
+In this analysis, we do not distinguish between Trial and Intertrial periods, and treat them both as exposure periods, where the degree of exposure depends on the narwhals positions relative to the ship.
 Figure~\ref{fig: trials and intertrials distributions} shows how the exposure periods are distributed among the $6$ narwhals that were tracked.
@@
 		\hline
-		A1 & 258 & 576\\
-		\hline
-		A2  & 40 & 515 \\
-		\hline
-		A3 & 222 & 680 \\
-		\hline
-		A4 & 291 & 642  \\
-		\hline
-		A5 & 48 & 419\\
-		\hline
-		A6 & 76 & 425 \\
+		A1 & 354 & 576\\
+		\hline
+		A2  & 151 & 515 \\
+		\hline
+		A3 & 397 & 680 \\
+		\hline
+		A4 & 127 & 642  \\
+		\hline
+		A5 & 207 & 419\\
+		\hline
+		A6 & 322 & 425 \\
 		\hline

### 2024-06-17 R/application/CRCVM/fit_crcvm.Rmd
update code

---
title: "fit_crcvm"
author: "Alexandre Delporte"
date: "2024-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

```{r packages}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
```

```{r get data}

# Set the path to the directory containing the data
par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#data before exposure with first 24h removed
dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

#data after exposure
dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"), header = TRUE,dec = ".")

```

We choose thresholds on the distance to the shore to define the ExpShore covariate.

```{r preprocess data}

D_low=0.07
D_up=3
dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low

dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low

dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low

N=length(unique(dataBE2$ID))

n_pre=length(dataBE2$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

## Baseline models

We can estimate the measurement error standard deviation directly from the data. Based on accuracy studies about fastloc GPS, it should be a few tens of metres. The estimate we get from the data before exposure is 35m (std error 1m), which seems consistent with previous studies. However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same. If we fix the measurement error to 35m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters. This cannot be solved by changing the other initial parameters in the optimisation. Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get. Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.

### Baselines with constant parameters

```{r fit baseline0}

#initial parameters
par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1,omega=~1)

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_100m$fit()

```

```{r results baseline0}
estimates_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Est")
std_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Std")

estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")

estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")

estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")

```

### Baseline with AngleNormal covariate in omega

Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau. Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.

```{r fit baseline1}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,0.01)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_35m$update_lambda(init_lambda)

baseline1_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_50m$update_lambda(init_lambda)

baseline1_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_75m$update_lambda(init_lambda)

baseline1_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_100m$update_lambda(init_lambda)

baseline1_100m$fit()

```

```{r results baseline1}
estimates_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Est")
std_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Std")

estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")

estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")

estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")

plots_bas1_35m=baseline1_35m$get_all_plots(baseline=NULL,model_name="baseline1_35m",show_CI="pointwise",save=TRUE)
plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)

```

### Baseline with AngleNormal and DistanceShore in omega

```{r fit baseline2}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_35m$update_lambda(init_lambda)
baseline2_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_50m$update_lambda(init_lambda)
baseline2_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_75m$update_lambda(init_lambda)
baseline2_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_100m$update_lambda(init_lambda)
baseline2_100m$fit()

```

```{r get results baseline2}

estimates_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Est")
std_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Std")

estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")

estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")

estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("DistanceShore"=0.07)
xmax=list("DistanceShore"=5)

plots_bas2_35m=baseline2_35m$get_all_plots(baseline=NULL,model_name="baseline2_35m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
```

```{r estimates on parameter scale baseline2}

post_coeff_bas2_35m=baseline2_35m$post_coeff(n_post=1000)
post_par_bas2_35m=list("tau0"=exp(post_coeff_bas2_35m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_35m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_35m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,2])))
mean_par_bas2_35m=lapply(post_par_bas2_35m,mean)
sd_par_bas2_35m=lapply(post_par_bas2_35m,sd)

post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)

post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)

#get 95% confidence intervals
post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
```

```{r smooth omega baseline2}

#get angular velocity for different distances to shore

plot1=plots_bas2_50m$fe_omega_AngleNormal_q1_DistanceShore
plot2=plots_bas2_50m$fe_omega_AngleNormal_q2_DistanceShore
plot4=plots_bas2_50m$fe_omega_AngleNormal_q4_DistanceShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Close"
omega2$Category <- "Medium"
omega4$Category <- "Far"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Close"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Far"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline2_50m",width = 10, height = 6, units = "in")

```

### Baseline with AngleNormal and ExpShore in omega

```{r fit baseline3}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1,1,0.01,1,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_35m$update_lambda(init_lambda)
baseline3_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_50m$update_lambda(init_lambda)
baseline3_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_75m$update_lambda(init_lambda)
baseline3_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_100m$update_lambda(init_lambda)
baseline3_100m$fit()
```
```{r get results baseline3}

estimates_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Est")
std_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Std")

estimates_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Est")
std_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Std")

estimates_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Est")
std_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Std")

estimates_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Est")
std_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("ExpShore"=1/D_up)
xmax=list("ExpShore"=1/D_low)
link=list("ExpShore"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore")

plots_bas3_35m=baseline3_35m$get_all_plots(baseline=NULL,model_name="baseline3_35m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale baseline3}

post_coeff_bas3_35m=baseline3_35m$post_coeff(n_post=1000)
post_par_bas3_35m=list("tau0"=exp(post_coeff_bas3_35m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_35m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_35m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,2])))
mean_par_bas3_35m=lapply(post_par_bas3_35m,mean)
sd_par_bas3_35m=lapply(post_par_bas3_35m,sd)

post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,2])))
mean_par_bas3_50m=lapply(post_par_bas3_50m,mean)
sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)

post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,2])))
mean_par_bas3_75m=lapply(post_par_bas3_75m,mean)
sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)

post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,2])))
mean_par_bas3_100m=lapply(post_par_bas3_100m,mean)
sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)

```

```{r smooth omega baseline3}

plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
plot2=plots_bas3_50m$fe_omega_AngleNormal_q2_ExpShore
plot4=plots_bas3_50m$fe_omega_AngleNormal_q4_ExpShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Far"
omega2$Category <- "Medium"
omega4$Category <- "Close"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Far"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Close"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")

```

## Response models with splines of ExpShip

### Response models based on baseline1

```{r fit response1}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.04
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_35m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_35m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_35m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_35m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_35m$lambda()["tau.s(ID)",1],1,baseline1_35m$lambda()["nu.s(ID)",1],
             baseline1_35m$lambda()["omega.s(AngleNormal)",1])
response1_sp_35m$update_coeff_re(new_coeff_re)
response1_sp_35m$update_coeff_fe(baseline1_35m$coeff_fe()[,1])

response1_sp_35m$update_lambda(new_lambda)

response1_sp_35m$fit()

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_50m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_50m$lambda()["tau.s(ID)",1],1,baseline1_50m$lambda()["nu.s(ID)",1],
             baseline1_50m$lambda()["omega.s(AngleNormal)",1])
response1_sp_50m$update_coeff_re(new_coeff_re)
response1_sp_50m$update_coeff_fe(baseline1_50m$coeff_fe()[,1])

response1_sp_50m$update_lambda(new_lambda)

response1_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_75m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_75m$lambda()["tau.s(ID)",1],1,baseline1_75m$lambda()["nu.s(ID)",1],
             baseline1_75m$lambda()["omega.s(AngleNormal)",1])
response1_sp_75m$update_coeff_re(new_coeff_re)
response1_sp_75m$update_coeff_fe(baseline1_75m$coeff_fe()[,1])

response1_sp_75m$update_lambda(new_lambda)

response1_sp_75m$fit()

# Fit response with 50m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_100m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_100m$lambda()["tau.s(ID)",1],1,baseline1_100m$lambda()["nu.s(ID)",1],
             baseline1_100m$lambda()["omega.s(AngleNormal)",1])
response1_sp_100m$update_coeff_re(new_coeff_re)
response1_sp_100m$update_coeff_fe(baseline1_100m$coeff_fe()[,1])

response1_sp_100m$update_lambda(new_lambda)

response1_sp_100m$fit()
```

```{r get results response1}

estimates_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Est")
std_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Std")

estimates_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Est")
std_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Std")

estimates_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Est")
std_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Std")

estimates_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Est")
std_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plots_res1_sp_35m=response1_sp_35m$get_all_plots(baseline=baseline1_35m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_35m",show_CI="pointwise",save=TRUE)
plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_50m",show_CI="pointwise",save=TRUE)
plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_75m",show_CI="pointwise",save=TRUE)
plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_100m",show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale response1}

post_coeff_res1_sp_35m=response1_sp_35m$post_coeff(n_post=1000)
post_par_res1_sp_35m=list("tau0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_35m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,2])))
mean_par_res1_sp_35m=lapply(post_par_res1_sp_35m,mean)
sd_par_res1_sp_35m=lapply(post_par_res1_sp_35m,sd)

post_coeff_res1_sp_50m=response1_sp_50m$post_coeff(n_post=1000)
post_par_res1_sp_50m=list("tau0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,2])))
mean_par_res1_sp_50m=lapply(post_par_res1_sp_50m,mean)
sd_par_res1_sp_50m=lapply(post_par_res1_sp_50m,sd)

post_coeff_res1_sp_75m=response1_sp_75m$post_coeff(n_post=1000)
post_par_res1_sp_75m=list("tau0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,2])))
mean_par_res1_sp_75m=lapply(post_par_res1_sp_75m,mean)
sd_par_res1_sp_75m=lapply(post_par_res1_sp_75m,sd)

post_coeff_res1_sp_100m=response1_sp_100m$post_coeff(n_post=1000)
post_par_res1_sp_100m=list("tau0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,2])))
mean_par_res1_sp_100m=lapply(post_par_res1_sp_100m,mean)
sd_par_res1_sp_100m=lapply(post_par_res1_sp_100m,sd)

```

### Response model with splines of ExpShip based on baseline2

```{r fit response2}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
response2_sp_50m$update_coeff_re(new_coeff_re)
response2_sp_50m$update_coeff_fe(baseline2$coeff_fe()[,1])

response2_sp_50m$update_lambda(new_lambda)

response2_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
response2_sp_75m$update_coeff_re(new_coeff_re)
response2_sp_75m$update_coeff_fe(baseline2$coeff_fe()[,1])

response2_sp_75m$update_lambda(new_lambda)

response2_sp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
response2_sp_100m$update_coeff_re(new_coeff_re)
response2_sp_100m$update_coeff_fe(baseline2$coeff_fe()[,1])

response2_sp_100m$update_lambda(new_lambda)

response2_sp_100m$fit()
```

```{r get results response2}
estimates_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
std_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")

estimates_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
std_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")

estimates_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
std_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_sp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_sp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_sp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
```

```{r}

post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)

post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)

post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)

```

### Response model with splines of ExpShip based on baseline3

```{r fit response3}

#initialize parameters
par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

#Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_50m$lambda()["tau.s(ID)",1],1,baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
response3_sp_50m$update_coeff_re(new_coeff_re)
response3_sp_50m$update_coeff_fe(baseline3_50m$coeff_fe()[,1])

response3_sp_50m$update_lambda(new_lambda)
response3_sp_50m$fit()

#Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_75m$lambda()["tau.s(ID)",1],1,baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
response3_sp_75m$update_coeff_re(new_coeff_re)
response3_sp_75m$update_coeff_fe(baseline3_75m$coeff_fe()[,1])

response3_sp_75m$update_lambda(new_lambda)
response3_sp_75m$fit()

#Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_100m$lambda()["tau.s(ID)",1],1,baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
response3_sp_100m$update_coeff_re(new_coeff_re)
response3_sp_100m$update_coeff_fe(baseline3_100m$coeff_fe()[,1])

response3_sp_100m$update_lambda(new_lambda)
response3_sp_100m$fit()
```

```{r get results response3}
estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")

estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")

estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_sp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_sp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_sp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale response3}

post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)

post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)

post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)

```

### Response model with exponential of ExpShip based on baseline2

```{r}

par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
response2_exp_50m$update_coeff_re(new_coeff_re)
response2_exp_50m$update_coeff_fe(new_coeff_fe)

response2_exp_50m$update_lambda(new_lambda)

response2_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
response2_exp_75m$update_coeff_re(new_coeff_re)
response2_exp_75m$update_coeff_fe(new_coeff_fe)

response2_exp_75m$update_lambda(new_lambda)

response2_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
response2_exp_100m$update_coeff_re(new_coeff_re)
response2_exp_100m$update_coeff_fe(new_coeff_fe)

response2_exp_100m$update_lambda(new_lambda)

response2_exp_100m$fit()
```

```{r}
estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")

estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")

estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res2_exp=response2_exp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_exp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_exp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_exp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

### Response model with exponential of ExpShip based on baseline3

```{r}

par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_50m$coeff_fe()[1:3,1],0,baseline3_50m$coeff_fe()[4,1],0,baseline3_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_50m$lambda()["tau.s(ID)",1],baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
response3_exp_50m$update_coeff_re(new_coeff_re)
response3_exp_50m$update_coeff_fe(new_coeff_fe)

response3_exp_50m$update_lambda(new_lambda)
response3_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_75m$coeff_fe()[1:3,1],0,baseline3_75m$coeff_fe()[4,1],0,baseline3_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_75m$lambda()["tau.s(ID)",1],baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
response3_exp_75m$update_coeff_re(new_coeff_re)
response3_exp_75m$update_coeff_fe(new_coeff_fe)

response3_exp_75m$update_lambda(new_lambda)
response3_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_100m$coeff_fe()[1:3,1],0,baseline3_100m$coeff_fe()[4,1],0,baseline3_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_100m$lambda()["tau.s(ID)",1],baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
response3_exp_100m$update_coeff_re(new_coeff_re)
response3_exp_100m$update_coeff_fe(new_coeff_fe)

response3_exp_100m$update_lambda(new_lambda)
response3_exp_100m$fit()
```

```{r}
estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")

estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")

estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_exp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_exp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_exp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore209.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 209
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore286.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 286
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore230.R

@@ -1,0 +1,201 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="circ"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 230
+set.seed(seed)
+
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore212.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 212
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore216.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 216
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore234.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 234
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore230.R

@@ -1,0 +1,201 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 230
+set.seed(seed)
+
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore257.R

@@ -1,0 +1,201 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 257
+set.seed(seed)
+
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore285.R

@@ -1,0 +1,201 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 285
+set.seed(seed)
+
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore287.R

@@ -1,0 +1,201 @@
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="fjords"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 287
+set.seed(seed)
+
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore236.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 236
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore265.R

@@ -1,0 +1,200 @@
+#
+# SETUP ------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+domain_name="rect"
+par_dir=dirname(getwd())
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
+library(smoothSDE)          #sde models
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+seed= 265
+set.seed(seed)
+
+
+
+# Generate samples ---------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+                      log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore[,c("p1","p2")]
+  cbind(data_sim,data_shore)
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+
+
+# Points that reached land ---------------
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID*100,"percent of the samples reached land")
+
+#subsample
+data=data[seq(1,n_obs*N_ID,by=BY),]
+
+
[…]

### 2024-06-17 Latex/main.tex
update Latex

@@
 \usepackage{lscape}
+\usepackage{calc}
+\usepackage{array}
 
@@
 
+%TABLES APPEARANCE
+\newlength\lenA \newlength\lenB
+\settowidth\lenA{Reach distance (km)}
+\setlength\lenB{(\lenA-2\tabcolsep-\arrayrulewidth)/2}
+
 %TITLE
 \title{
- \textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}}
+ \textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhals behavioral response to ship exposure and airgun exposure}}
 
@@
 But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
-Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\eqref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
+Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in~\eqref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
 \\
@@
 
+\begin{comment}
 \begin{equation}   \left\{
@@
 Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$, $k \in \{1,\cdots,q\}$ are estimated from the trajectories during exposure to the sound.
+\end{comment}
+
+\begin{equation}   \left\{
+	\begin{array}{l}
+		y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots,n_{i,post}\} \\
+		\log(\tau_{i}(t))=\log(\tau_{0})+\alpha_{\tau} E^{ship}_i(t)+b_{\tau,i} \\
+		\log(\nu_{i}(t))=\log(\nu_{0}) +  \alpha_{\nu} E^{ship}_i(t) +b_{\nu,i}  \\
+	\end{array}
+	\right.
+	\label{eq: exact response model}
+\end{equation}
+
+The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are considered as offsets.
+The coefficients $\alpha_{\tau}$ and $\alpha_{\nu}$ in the log-linear models control the distance at which the parameters $\tau_i$ and $\nu_i$ reach the baseline values $\tau_0$ and $\nu_0$.
 
@@
 Estimation is performed in \texttt{smoothSDE} by defining a template model for the complete negative  log-likelihood with \texttt{TMB}.
-The choice of initial parameters is crucial in the optimization.
-Based on the results on measurement errors of Fastloc-GPS found in \cite{wensveen_path_2015}, we initialize the measurement error standard deviation to $\sigma_{obs}=50$ m.
-Then we estimate model ~\eqref{eq: exact baseline model} on all the available data with degree of freedom of the marginal splines set to $5$ and we use the estimated spline coefficients as initial values to fit the baseline model on the data before exposure.
-We initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=0.2$ and fix the smoothing penalties of the tensor splines to $1$.
+We fixed the measurement error standard deviation to $\sigma_{obs}=50$ m based on the results on measurement errors of Fastloc-GPS found in \cite{wensveen_path_2015}.
+Then we estimate model ~\eqref{eq: exact baseline model} on all the available data with degree of freedom of the marginal splines set to $5$.
+We initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=1$, spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ to 0 and fix the smoothing penalties of the tensor splines to $1$.
 
 Estimation and standard error on the parameter scale are shown in table~\ref{table: baseline estimations}. 
-The persistence is estimated to $\hat{\tau}_{0}=1.08 $ h.
+The persistence is estimated to $\hat{\tau}_{0}=1.35 $ h.
 In comparison, harbour seal in Alaska were proved to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \cite{johnson_continuoustime_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \cite{gurarie_correlated_2017}.
-The intercept value for $\nu$ is estimated to be $\hat{\nu}_{0}=4.61$ km/h which is of the order of magnitude of the mean of the observed empirical velocities.
+The intercept value for $\nu$ is estimated to be $\hat{\nu}_{0}=4.64$ km/h which is of the order of magnitude of the mean of the observed empirical velocities.
 \begin{table}[H]
@@
 		\hline
-		$\tau_0$ (h)   & $1.08$    &  $0.16$ \\
-		$\nu_0$ (km/h)  & $4.64$   &  $0.25$\\
-		$\omega_0$ (rad/h)    & $0.01$    &  $0.08$ \\
-		$\sigma_{\tau,pre}$    & 0.29   &  $0.12$ \\
-		$\sigma_{\nu,pre}$    & 0.10   &  $0.05$ \\
-		$\sigma_{obs}$ (m) & $35$ & $1$ \\
+		$\tau_0$ (h)   & $1.35$    &  $0.16$ \\
+		$\nu_0$ (km/h)  & $4.60$   &  $0.25$\\
+		$\omega_0$ (rad/h)    & $-0.01$    &  $0.07$ \\
+		$\sigma_{\tau,pre}$    & 0.20   &  $0.13$ \\
+		$\sigma_{\nu,pre}$    & 0.08   &  $0.05$ \\
 		\hline
@@
 This indicates that the spatial constraints influence the tortuosity of the movement by forcing the narwhals velocity to rotate when they are close to the shoreline.
+However, the magnitude of the estimated $\omega$ is not sufficient to effectively obtain constrained trajectories when simulating from the fitted model.
 \begin{figure}[H]
@@
 
+We observed that fixing a high standard deviation for the measurement error tends to increase the estimate of the parameter $\tau$.
+Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.
+We tried fixing the measurement errors $\sigma_{obs}$ to different values and observed that the lowest negative log-likelihood value was obtained for a standard deviation $\sigma_{obs}$ close to $50$m.
+More details can be found in Appendix.
+
 \subsection{Response estimations}
@@
 The baseline estimates of the spline coefficients $\omega_k$, the intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the standard deviations $\sigma_{\tau}$, $\sigma_{\nu}$ are introduced as offsets in the response model.
-Only the effect of the exposure to the sound on $\tau$ and $\nu$ is estimated from the data after exposure.\\
-
-Figure~\ref{fig: me response} shows that exposure to sound has an effect on both the persistence and the velocity of the movement.
+Only the effect of the exposure to the sound on $\tau$ and $\nu$ is estimated from the data after exposure.
+The estimates for the coefficients are $\alpha_{\tau}=-3.43$ with standard error $0.70$ and $\alpha_{\nu}=0.74$ with standard error $0.27$.\\
+
+The smooth parameters $\tau$ and $\nu$ as a function of the distance to the sound source are shown in Figure~\ref{fig: me response}.
+They indicate that exposure to sound has an effect on both the persistence and the velocity of the movement.
 The value of $\nu$ increases with increased exposure to the sound, suggesting motion at higher speed in presence of the ship.
@@
 This shows evidence of a strong shift in the behavior of the whales, which is likely to impede their capacity to mate and forage at short term.
-The baseline value is attained for a distance to the sound source higher than $20$ km for $\tau$, and  $40$ km for $\nu$.
-This indicates that sound exposure can be perceived and can disturb the narwhals motion up to $40$ km.
+For the parameter $\tau$, $95\%$ of the baseline value is recovered at a distance of $66$ km.
+For the parameter $\nu$, $95\%$ of the baseline value is recovered at a distance of $14$ km.
+This indicates that sound exposure can be perceived and can disturb the narwhals motion up to tens of kilometres.
 
@@
  	\begin{subfigure}{0.98\textwidth}\centering
- 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_tau_ExpShip.png"}
+ 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_exp_tau_ExpShip.png"}
  		\caption{}
@@
  		\centering
- 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_nu_ExpShip.png"}
+ 		\includegraphics[width=0.7\linewidth]{"images/application/response/me_response3_exp_nu_ExpShip.png"}
  		\caption{}
@@
 
-We categorize narwhals position relative to the ship as follows :
-\begin{itemize}
-	\item very close when the distance to the ship is lower than the $5\%$. quantile, which is $4.04$ km.
-	\item close when the distance to the ship is between $5\%$ quantile and the $25\%$ quantile, which is $9.35$ km.
-	\item medium when the distance to the ship is between the $25\%$ and $75\%$ quantile, that is between $9.35$ and $25.80$ km.
-	\item far otherwise.
-\end{itemize}
-Table~\ref{table: baseline vs response parameters comparison} shows the changes in persistence and mean velocity parameters $\tau$ and $\nu$ for the four categories.
+The distances to the ship $D^{ship}_{\tau}(p)$ and $D^{ship}_{\nu}(p)$  at which a percentage $p$ of the population baseline values $\tau_0$ and $\nu_0$ are reached are given by 
+\begin{equation}
+	D^{ship}_{\tau}(p)=\frac{\alpha_{\tau}}{\log(p)} \quad  \mbox{and} \quad  D^{ship}_{\nu}(p)=-\frac{\alpha_{\nu}}{\log(p)}
+\end{equation}
+Table~\ref{table: baseline vs response parameters comparison} shows these values for different percentages $p$.
 \begin{table}[H]
 	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-		\hline
-		Position relative to the ship & mean population $\hat{\tau}$  & mean population $\hat{\nu}$ & $\frac{\hat{\tau}-\hat{\tau}_{0}}{\tau_{0}}$ & $\frac{\hat{\nu}-\hat{\nu}_{0}}{\hat{\nu}_{0}}$   \\
-		\hline
-		Very close & 0.54 & 6.4 & $-50\%$ & $+39\%$ \\
-		\hline
-		Close & 0.74 & 5.8 & $-31\%$ & $+26\%$ \\
-		\hline
-		Medium&  0.96 &  5.0 & $-11\%$ & $+9\%$\\
-		\hline
-		Far &  1.07 & 4.7 &  $-1\%$ & $+2\%$\\
-		\hline
+	\begin{tabular}{| c | *{2}{w{c}{\lenB}|}}
+		\hline
+		\multirow{2}{*}{Percentage of the baseline value} & \multicolumn{2}{c|}{Reach distance (km) }  \\
+		\cline{2-3} 
+		& $D^{ship}_{\tau}$ & $D^{ship}_{\nu}$ \\
+		\hline
+		50 & 4.9 &  1.1 \\
+		75 & 11.9 & 2.6 \\
+		95 & 66.9 & 14.4 \\
+		\hline
+
 	\end{tabular}
-	\caption{Comparison of response and baseline values of the parameters for different levels of exposure to the ship}
+	\caption{Percentage of the baseline value reached at different distances to the ship}
 	\label{table: baseline vs response parameters comparison}
@@
 \end{comment}
-
+\begin{comment}
 \section{Discussion}
 
+We were able to estimate an increase tortuosity due to the proximity with the shoreline in the baseline model.
+However, the estimated angular velocity is too low to obtain constrained samples of the fitted model as the ones shown in~\ref{fig: crcvm examples}.
+
+\end{comment}
 \begin{comment}
@@
 		\hline
-		\textbf{Total} & \textbf{935} & \textbf{3257} \\
+		\textbf{Total} & \textbf{1558} & \textbf{3257} \\
 		\hline
@@
 		\centering
-		\includegraphics[scale=0.38]{images/data_exploration/tracksBE1.png}
+		\includegraphics[scale=0.38]{images/data_exploration/tracksBE2.png}
 		\caption{Tracks before exposure experiments}
@@
 and covariance matrix 
-\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top}\\
+\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top} ds\\
 	&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 

### 2024-06-17 Latex/main.tex
update Latex

@@
 Research recommendations include a particular focus on measurement of behavioral reactions to various type of sound exposure \cite{southall_marine_2008,southall_marine_2019}.
-Rich behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured such as GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
+Behavioral studies typically involve controlled exposure experiments during which animals are exposed to disturbances in a precise and monitored set up where different  statistics can be measured such as GPS positions, depth, sound exposure level, temperature, pressure, heart beat, stroke rate, vocalizations...
 Numerous surveys have been conducted about beaked whales, which have been regularly implicated in stranding events following sonar exercises \cite{tyack_beaked_2011,cioffi_trade-offs_2022}, belugas \cite{martin_exposure_2023}, sperm whales \cite{madsen_quantitative_2006}, blue whales \cite{friedlaender_preymediated_2016}, and narwhals \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}.
@@
 Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one holistic model.
+The benefit of such a modelisation are multiple : it does not rely on any interpolation of the positions each second since irregular time step are easily handled in the continuous-time framework, it does not require labelling states by hand and therefore results are more reliable, it allows to quantify the magnitude of the changes in the movement depending on the distance to the sound source.
 \\
@@
 It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}.
-The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
+An extension of the CVM which involves a rotational component and is based on biologically interpretable parameters is formulated in  \cite{gurarie_correlated_2017}.
+The equations are as follows:
 \begin{equation} \left\{
@@
 Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in \eqref{eq: RACVM equation}.
-This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship.
+These effects are estimated as smooth functions of a sound exposure covariate defined as the inverse of the distance to the ship.
 This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}.
@@
 Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effects.
-Simulation study results are shown in section 5.
-Eventually, our models are applied to analyse a behavioral response of the narwhals in section 6.
+Simulation study results are shown in Section 5.
+Eventually, our models are applied to analyse a behavioral response of the narwhals in Section 6.
 
@@
 
-The dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
+The dataset analysed in this paper has been subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
 Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters.
@@
 Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure.
-Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.
+Exposure levels for each narwhal are displayed in figure~\ref{fig: realexpthroughtime}.
 
@@
 	\omega & \frac{1}{\tau}
-\end{pmatrix}$ and rewrite~\eqref{eq: RACVM equation}
+\end{pmatrix}$ and rewrite~\eqref{eq: RACVM equation} as
 
@@
 But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$.
+This may not be a big deal when sucessive positions of the animal are known with high frequency.
+However, when it comes to marine mammals that dive at great depth such as the narwhal, the time steps between GPS measurements are generally irregular and can reach from several minutes to several hours, therefore invalidating the use of the Euler scheme.
+Moreover, to be able to include covariates in the SDE parameters we often need to approximate the transition density by supposing that the covariates are constant on each time step, which would then create two levels of approximations, likely making estimation unreliable.
  Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in~\eqref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula.
-\\
+This is stated in the following proposition.\\
 
@@
 A weight based on the distance to the shore can be defined as follows :
-\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]
-
-Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated  :
+\[c(D^{shore})=\mbox{ for some } \]
+
+A relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :
 
 \begin{equation}
-f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D^{shore})
+f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times \exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)
 \label{eq: smooth omega}
 \end{equation}
-for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
+for some $D_0>0$, $\kappa>0$, $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.
 

### 2024-11-08 Latex/appendix/supp-aoas-template.tex
reorganize folders

%% Template for the submission to:
%%   The Annals of Applied Statistics [AOAS]
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% In this template, the places where you   %%
%% need to fill in your information are     %%
%% indicated by '???'.                      %%
%%                                          %%
%% Please do not use \input{...} to include %%
%% other tex files. Submit your LaTeX       %%
%% manuscript as one .tex document.         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[aoas]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage[authoryear]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures

%%PACKAGES
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{lscape}
\usepackage{calc}
\usepackage{array}
\usepackage{bbold}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\usepackage{fancyhdr}

% set up for custom R code
\usepackage{listings}
\definecolor{darkgreen}{rgb}{0.0, 0.5, 0.0} % Define a dark green color for the comments

\lstdefinestyle{customR}{
	language=R,
	backgroundcolor=\color{lightgray},
	basicstyle=\ttfamily\footnotesize,
	keywordstyle=\bfseries\color{blue},
	commentstyle=\itshape\color{darkgreen}, % Use the custom dark green color
	stringstyle=\color{red},
	numbers=left,
	numberstyle=\tiny\color{gray},
	stepnumber=1,
	numbersep=10pt,
	showspaces=false,
	showstringspaces=false,
	showtabs=false,
	frame=single,
	rulecolor=\color{black},
	tabsize=2,
	captionpos=b,
	breaklines=true,
	breakatwhitespace=false,
	title=\lstname,
	escapeinside={\%*}{*)},
	morekeywords={*,...} % if you want to add more keywords to highlight
}

\lstset{style=customR}

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{plain}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{remark}                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{remark}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{proposition}{Proposition}[section]

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{remark}
\newtheorem*{example}{Example}

%% SHORTCUTS
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\erf}{erf}
\newcommand {\R}{\mathbb{R}}
\newcommand {\E}{\mathbb{E}}
\newcommand {\prob}{\mathbb{P}}
\newcommand {\N}{\mathbb{N}}
\newcommand {\Z}{\mathbb{Z}}
\newcommand {\1}{\mathbb{1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endlocaldefs

\begin{document}

\begin{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Supplementary file to "Varying coefficients correlated velocity models in complex landscapes with boundaries applied to narwhal responses to noise exposure"}
%\title{A sample article title with some additional note\thanksref{T1}}
\runtitle{Movement models in complex landscapes}
%\thankstext{T1}{A sample of additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Alexandre}~\snm{Delporte}\ead[label=e1]{alexandre.delporte@univ-grenoble-alpes.fr}},
\author[B]{\fnms{Susanne}~\snm{Ditlevsen}\ead[label=e2]{susanne@math.ku.dk}}
\and
\author[A]{\fnms{Adeline}~\snm{Samson}\ead[label=e3]{adeline.leclercq-samson@univ-grenoble-alpes.fr}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Laboratoire Jean Kuntzmann,
	Université Grenoble-Alpes\printead[presep={,\ }]{e1,e3}}

\address[B]{Department of Mathematical Sciences,
	University of Copenhagen \printead[presep={,\ }]{e2}}
\end{aug}

\begin{keyword}
\kwd{behavioral response study}
\kwd{stochastic differential equations}
\kwd{mixed effect model}
\end{keyword}

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix---Please move all appendices to %%
%% a Supplementary file.                    %%

\appendix

\section{More details about the data}
This section is a complement to Section 2 of the paper. We provide more information about the narwhal movement data used for the analysis of behavioral disturbance.\\

The time step between consecutive GPS observations is not constant. Its median is $4.8$ minutes and its mean is $9.3$ minutes. We show the histogram of the time steps in Figure~\ref{fig:alltimestepshisto}.

\begin{figure}[ht!]
	\centering
	\includegraphics[scale=0.4]{images/data_exploration/all_time_steps_histo}
	\caption{Histogram of time steps}
	\label{fig:alltimestepshisto}
\end{figure}

The observations are divided into unexposed periods, for which the narwhals are not in line of sight with the ship; trial periods, when the narwhals are exposed to the ship and airguns are shot; and intertrial periods, when the narwhals are exposed to the ship but airguns are not shot. These periods are indicated by a categorical variable $T_{ship}$ in the dataset. Figure~\ref{fig: trials and intertrials distributions} shows how the exposure periods are distributed among the $6$ narwhals that were tracked. Our analysis in section 6 does not distinguish between intertrial and trial periods. They are both treated as exposure periods, though the nature and intensity of the behavioral response might differ for the two periods. We adopted this approach due to the lack of intertrial data as well as a potential persistence in time of the behavior shift due to airgun exposure during trial periods.

\begin{figure}[ht!]
	\centering
	\centering
	\includegraphics[scale=0.5]{images/data_exploration/trials.png}
	\caption{Trial and Intertrial periods for each narwhal}
	\label{fig: trials and intertrials distributions}

\end{figure}

Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals. Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure with a base map of Scoresby Sound fjords system.

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
		\hline
		A1 & 354 & 576\\
		\hline
		A2  & 151 & 515 \\
		\hline
		A3 & 397 & 680 \\
		\hline
		A4 & 127 & 642  \\
		\hline
		A5 & 207 & 419\\
		\hline
		A6 & 322 & 425 \\
		\hline
		\textbf{Total} & \textbf{1558} & \textbf{3257} \\
		\hline
	\end{tabular}
	\caption{Distribution of the data among the 6 individuals}
	\label{table: data distribution}
\end{table}

\begin{figure}[ht!]
	\centering
	\begin{subfigure}{0.8\linewidth}
		\centering
		\includegraphics[scale=0.5]{images/data_exploration/tracksBE2.png}
		\caption{Tracks before exposure experiments}
	\end{subfigure}

	\begin{subfigure}{0.8\linewidth}
		\centering
		\includegraphics[scale=0.5]{images/data_exploration/tracksAE.png}
		\caption{Tracks during exposure experiments}
	\end{subfigure}
	\caption{Movement data of East Greenland narwhals. The red crosses indicate initial positions.}
	\label{fig: tracks before and after exposure}
\end{figure}

All the relevant covariates   used for the analysis of narwhals movement are summarized in Table \ref{tab: summary covariates}.

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|p{8cm}|c|}
		\hline
		Covariate & Unit & Description & Domain \\
		\hline
		$D^{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
		\hline
		$E^{ship}(t)=\frac{1}{D^{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
		\hline
		$D^{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
		\hline
		$I^{shore}(t)=\frac{1}{D^{shore}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
		\hline
		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical velocity vector at time $t$  & $[-\pi,\pi]$ \\
		\hline 
	\end{tabular}
	\caption{Summary of the covariates}
	\label{tab: summary covariates}
\end{table}

\section{Proof of Proposition 3.1}
\label{section: transition density proof}

Here, we prove Proposition 3.1 The proof is inspired by the results in \citep{gurarie_correlated_2017,johnson_continuous_2008} and the proof of the transition density of the velocity process in \citep{albertsen_generalizing_2018}.
\begin{proof}
	The velocity process is an Ornstein-Uhlenbeck process. For $t \geq 0$ and $\Delta >0$,
	\begin{equation}
		V(t+\Delta)=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)
		\label{eq: RACVM solution}
	\end{equation}

	It has Gaussian transition density with mean 
	\begin{equation}
		\E(V(t+\Delta \vert V(t)))=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu 
	\end{equation}
	and covariance matrix 
	\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top} ds\\
		&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 
	\end{align*}
	Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$ where $R_{-\omega u}$ is the rotation matrix with angle $-\omega u$, the matrix product  $\exp(-Au) \exp(-Au)^{\top}$ is simply $\exp(-\frac{2u}{\tau}) I_2$. We deduce that the two components of the velocity are independent and have the same variance, denoted  $q_2(\Delta)$. The variance is 
	\begin{equation}
		q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-\exp\left(-\frac{2\Delta}{\tau}\right)\right).
	\end{equation}
	These results are found in \cite{gurarie_correlated_2017}.
	In the sequel, we  use the notation 
	\[\zeta(t,s) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{s} \exp(A(u-s)) dW(u).\] 
	Using that $V(s)=\mu+\exp(-A(s-t))(V(t)-\mu)+\zeta(t,s)$, we have
\begin{align*}
    X(t+\Delta) &= X(t) + \int_t^{t+\Delta} V(s) \, ds \\
    &= X(t) + \mu \Delta + \int_t^{t+\Delta} \exp(-A(s-t))(X(t) - \mu) \, ds \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    % &= X(t) + \mu \Delta + \int_0^{\Delta} \exp(-As)(X(t) - \mu) \, ds + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    &= X(t) + \mu \Delta + \left(A^{-1}(V(t) - \mu) - A^{-1}\exp(-A\Delta)(X(t) - \mu)\right) \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds
\end{align*}
	Thus, 
	\begin{equation}
		X(t+\Delta)=X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu)+\xi(t,t+\Delta) 
	\end{equation}
	where $ \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds$.
	The location process is also Gaussian with mean 
	\begin{equation}
		\E(X(t+\Delta) \vert V(t),X(t)) =X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu).
	\end{equation}
	To get an expression of the covariance matrix, first rewrite
	\begin{align*}
		\xi(t,t+\Delta)&=\int_t^{t+\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
		%&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
	\end{align*}
	Then use Ito's isometry
 \begin{align*}
    \mbox{Var}(X(t+\Delta)\vert X(t),V(t))&=\frac{4 \nu^2}{\pi \tau} 
    \int_{t}^{t+\Delta} A^{-1} (I_2-\exp(A(u-t-\Delta)))\\
    &\quad \times (I_2-\exp(A(u-t-\Delta)))^\top A^{-\top} \, du \\
    &=\frac{4 \nu^2}{\pi \tau} \int_{0}^{\Delta} A^{-1} (I_2-\exp(-Ar))\\
    &\quad \times (I_2-\exp(-Ar))^\top A^{-\top} \, dr.
\end{align*}
This integral can be computed explicitly since 
\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} f(r) I_2 \]
where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
We obtain that $X_1(t+\Delta)$ and $X_2(t+\Delta)$ are independent and have the same variance, denoted $q_1(\Delta)$.
Writing $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$, the variance is
\begin{align*}
    q_1(\Delta) &= \frac{\sigma^2}{C} \left( \Delta 
    - 2 \frac{\omega \sin(\omega \Delta) - \frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2} + \omega^2 } 
    \exp\left( -\frac{\Delta}{\tau} \right) \right. \\
    &\quad \left. + \frac{\tau}{2} \left( \frac{\omega^2 - \frac{3}{\tau^2}}{\frac{1}{\tau^2} + \omega^2} 
    - \exp\left( -\frac{2\Delta}{\tau}\right) \right) 
    \right)
\end{align*}

Now we compute the covariance between   $X$ and $V$ to get the full covariance matrix of $U$:
\begin{align*}
    \Gamma(\Delta) &= \frac{4\nu^2}{\pi \tau} \E\left(\left( \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) \, dW(u)\right) \right. \\
    &\quad \left. \times \left(\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \, dW(s)\right)^\top\right) \\
    &= \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-(t+\Delta)))) \exp(A(u-(t+\Delta)))^\top \, du \\
    &= \frac{4\nu^2}{\pi \tau} \int_0^{\Delta} A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top \, dr.
\end{align*}
	Then, 
	\[A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top=\frac{1}{C}\exp\left(-\frac{r}{\tau} \right) \begin{pmatrix} g(r) & h(r) \\ -h(r) & g(r) \end{pmatrix}\]
	where 
\begin{align*}
    g(r) &= \frac{1}{\tau} \left(\cos(\omega r) + \exp\left( -\frac{r}{\tau} \right)\right) 
    - \omega \sin(\omega r), \\
    h(r) &= -\frac{1}{\tau} \sin(\omega r) 
    + \omega \left(\cos(\omega r) - \exp\left( -\frac{r}{\tau}\right)\right).
\end{align*}
	Finally we get
	\[
	\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right),
	\]
	\[\gamma_2(\Delta)=\frac{\sigma^2}{C}\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right).\]

	In the specific case $\omega=0$, we obtain $C=\frac{1}{\tau^2}$ and the variance of  $X$ becomes 
	\[q_1(\Delta)=\sigma^2 \tau^2\left( \Delta +2\tau\exp\left( -\frac{\Delta}{\tau}\right)+\frac{\tau}{2}\left( -3 -\exp\left( -\frac{2\Delta}{\tau}\right) \right)\right).
	\]
	Writing $\beta=\frac{1}{\tau}$ and reorganizing the terms, we obtain
	\begin{equation}q_1(\Delta)=\frac{\sigma^2}{\beta^2}\left(\Delta -2 \frac{1-\exp(-\beta \Delta)}{\beta}+\frac{1-\exp(-2\beta \Delta)}{2\beta}\right).
	\end{equation}
	This result match equation $(6)$ in \citep{johnson_continuous_2008}. Similarly, in the case $\omega=0$, we get $\gamma_2=0$ and the expression for $\gamma_1$ match equation $(7)$ in \citep{johnson_continuous_2008}.

\end{proof}

\section{Measurement error}

In the application in Section 6, different values of the measurement error were estimated for the data before and after exposure: $35$ m before and $48$ m after. Both these values are consistent with the results in \citep{wensveen_path_2015}. However, the post exposure estimation gave non-positive definite Hessian matrix for the negative log-likelihood, which prevents from using the information matrix equality to get confidence intervals of the estimates. Fixing a $35$ m measurement error value when fitting the response model led to the same issue. We therefore tried different values of the measurement error, and kept the one that gave a positive definite hessian matrix and had the highest log-likelihood value. It turned out to be $50$ m, very close to the initially estimated $48$ m. Table~\ref{tab: estimates for fixed measurement errors} shows these results. In comparison, the final values of the log-likelihood when $\sigma_{obs}$ is estimated from the data are respectively $4273$ and $8043$ before and after exposure, while the estimate of $\tau_0$ is $1.10$, and the estimates of $\alpha_\tau$ and $\alpha_\nu$ are respectively $-4.19$ and $0.66$, which is in the confidence interval of the final estimations we kept (those obtained for $\sigma_{obs}=50$ m).
\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline
		$\sigma_{obs}$ (m) & $\hat{\tau_0}$ & Baseline llk & Response llk & P.d hessian & $\alpha_{\tau}$ & $\alpha_\nu$ \\
		\hline
		$30$  & $0.96 \pm 0.15$ & $4261$  & 8038 & No & $0.29$ & $2.17$ \\
		\hline
		$40$  & $1.18 \pm 0.16$   & $4266$ & 8014  & No  & $-2.06$ & $0.60$\\
		\hline
		$45$  & $1.29 \pm 0.17$   & $4243$ & 7965  & No & $-3.64$ & $0.49$\\
		\hline
		$\mathbf{50}$  & $\mathbf{1.35 \pm 0.16}$ & $\mathbf{4208}$ & $\mathbf{7861}$ & \textbf{Yes} & $\mathbf{-3.43 \pm 0.70}$ & $\mathbf{0.74 \pm 0.27}$\\
		\hline
		$75$ &  $1.63 \pm 0.19$ & $3944$ & $7225$ & Yes & $-3.88 \pm 0.73$ & $0.76 \pm 0.30$\\
		\hline 
		$100$ & $1.85 \pm 0.22$ & $3640$ & $6590$ & Yes & $-4.27 \pm 0.74$ & $0.63 \pm 0.29$ \\
		\hline 
	\end{tabular}
	\caption{Estimate for the baseline and response models for several fixed measurement errors.}
	\label{tab: estimates for fixed measurement errors}
\end{table}

\section{Code example}

We illustrate briefly how to fit our baseline SDE model and obtain the results with \texttt{smoothSDE} R package \cite{michelot_varying-coefficient_2021}. The version of the package including our new model is available \href{https://github.com/alexandre-delporte/smoothSDE}{here} . We suppose the package has been loaded. We consider a dataframe \texttt{dataBE} containing the preprocessed observations before exposure to the ship in the columns \texttt{x} and \texttt{y}, an animal identifier in a column \texttt{ID} and columns \texttt{Ishore} and \texttt{AngleNormal} for the covariates $I^{shore}$ and $\Theta$. The first step consists in choosing initial SDE parameters and model formulas. For the model we consider, there are five parameters $\mu_1$, $\mu_2$, $\tau$, $\nu$, and $\omega$, and each of them needs a formula. Specification of the formulas is identical to the \texttt{R} package \texttt{mgcv}. Among the parameters, $\mu_1$, $\mu_2$ will be set to 0, while $\tau$, $\nu$ and $\omega$ are expressed as in section 4.1.

\begin{lstlisting}[language=R]
#number of observation
n_pre<-nrow(dataBE)

#initial parameters
par0 <- c(0,0,1,4,0)

#model formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))
\end{lstlisting}

We then specify the measurement error for each observation in an array of covariance matrices. We suppose they are all diagonal with the same standard deviation \texttt{sigma\_obs}. We will fix this measurement error. \\

\begin{lstlisting}[language=R]
# 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
\end{lstlisting}

We can then create the SDE object as in \cite{michelot_varying-coefficient_2021}. We choose the type of SDE in the argument \texttt{type}. Here, it is \texttt{RACVM} (see Section 3.1) since we want to include a non zero rotation parameter $\omega$. The name of the columns where the observations are found is specified in the \texttt{response} argument.
We specify the measurement error matrix \texttt{H} in the argument \texttt{other\_data}. Fixed parameters are indicated in the argument \texttt{fixpar}.

\begin{lstlisting}[language=R]
#create SDE object
baseline_50m<- SDE$new(formulas = formulas,data = dataBE,type = "RACVM",
response = c("x","y"),par0 = par0,other_data=list("H"=H),
fixpar=c("mu1","mu2"))
\end{lstlisting}

To fix specific parameters in the statistical model, we need to use the \texttt{map} attribute. Here we use it to specify that the smoothing parameters should be fixed. Then we update the smoothing parameters to $1$, and fit the SDE model.

\begin{lstlisting}[language=R]
#update map to fix smoothig parameters
baseline_50m$update_map(list("log_lambda"=factor(c(1,2,rep(NA,4))))

#update smoothing parameters values
init_lambda=rep(1,6)
baseline_50m$update_lambda(init_lambda)

#fit the model
baseline_50m$fit()
\end{lstlisting}

The results of the optimization are stored in the attribute \texttt{tmb\_rep}.
We can extract the estimated parameters along with the standard errors.

\begin{lstlisting}[language=R]
#estimates
estimates_bas_50m=as.list(baseline_50m$tmb_rep(),what="Est")
#standard error
std_bas_50m=as.list(baseline_50m$tmb_rep(),what="Std")
\end{lstlisting}

Finally, we would like to plot all the smooth parameters as a function of the covariates. 
We can do it with the \texttt{get\_all\_plots} method.
We only need to specify the range of each covariate value we want to plot, a link function if we don't want to have directly the covariate on the x-axis but rather a function of the covariate, and the x-axis label of the plots. We put the option \texttt{show\_CI="pointwise"} to show the pointwise confidence intervals on the plots.

\begin{lstlisting}[language=R]
#range of the covariates
D_low=0.073
D_up=3
xmin=list("Ishore"=1/D_up)
xmax=list("Ishore"=1/D_low)
#link function
link=list("Ishore"=(\(x) 1/x))
#label
xlabel=list("Ishore"="Distance to shore")

#draw plots
plots_bas_50m=baseline_50m$get_all_plots(model_name="baseline_50m",
xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  imsart-nameyear.bst  will be used to                   %%
%%  create a .BBL file for submission.                     %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%  MR numbers will be added by VTeX.                      %%
%%                                                         %%
%%  Use \citep{...} to cite references in text.             %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{imsart-nameyear} % Style BST file
\bibliography{biblio-supp}       % Bibliography file (usually '*.bib')
\end{document}

### 2024-11-08 Latex/main/aoas-template.tex
reorganize folders

%% Template for the submission to:
%%   The Annals of Applied Statistics [AOAS]
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% In this template, the places where you   %%
%% need to fill in your information are     %%
%% indicated by '???'.                      %%
%%                                          %%
%% Please do not use \input{...} to include %%
%% other tex files. Submit your LaTeX       %%
%% manuscript as one .tex document.         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[aoas]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage[authoryear]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures

%%PACKAGES
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{lscape}
\usepackage{calc}
\usepackage{array}
\usepackage{bbold}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\usepackage{fancyhdr}

% set up for custom R code
\usepackage{listings}
\definecolor{darkgreen}{rgb}{0.0, 0.5, 0.0} % Define a dark green color for the comments

\lstdefinestyle{customR}{
	language=R,
	backgroundcolor=\color{lightgray},
	basicstyle=\ttfamily\footnotesize,
	keywordstyle=\bfseries\color{blue},
	commentstyle=\itshape\color{darkgreen}, % Use the custom dark green color
	stringstyle=\color{red},
	numbers=left,
	numberstyle=\tiny\color{gray},
	stepnumber=1,
	numbersep=10pt,
	showspaces=false,
	showstringspaces=false,
	showtabs=false,
	frame=single,
	rulecolor=\color{black},
	tabsize=2,
	captionpos=b,
	breaklines=true,
	breakatwhitespace=false,
	title=\lstname,
	escapeinside={\%*}{*)},
	morekeywords={*,...} % if you want to add more keywords to highlight
}

\lstset{style=customR}

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{plain}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{remark}                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{remark}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{proposition}{Proposition}[section]

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{remark}
\newtheorem*{example}{Example}

%% SHORTCUTS
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\erf}{erf}
\newcommand {\R}{\mathbb{R}}
\newcommand {\E}{\mathbb{E}}
\newcommand {\prob}{\mathbb{P}}
\newcommand {\N}{\mathbb{N}}
\newcommand {\Z}{\mathbb{Z}}
\newcommand {\1}{\mathbb{1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endlocaldefs

\begin{document}

\begin{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Varying coefficients correlated velocity models in complex landscapes with boundaries applied to narwhal responses to noise exposure}
%\title{A sample article title with some additional note\thanksref{T1}}
\runtitle{Movement models in complex landscapes}
%\thankstext{T1}{A sample of additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Alexandre}~\snm{Delporte}\ead[label=e1]{alexandre.delporte@univ-grenoble-alpes.fr}},
\author[B]{\fnms{Susanne}~\snm{Ditlevsen}\ead[label=e2]{susanne@math.ku.dk}}
\and
\author[A]{\fnms{Adeline}~\snm{Samson}\ead[label=e3]{adeline.leclercq-samson@univ-grenoble-alpes.fr}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Laboratoire Jean Kuntzmann,
	Université Grenoble-Alpes\printead[presep={,\ }]{e1,e3}}

\address[B]{Department of Mathematical Sciences,
	University of Copenhagen \printead[presep={,\ }]{e2}}
\end{aug}

\begin{abstract}
Narwhals in the Arctic are increasingly exposed to human activities that can temporarily or permanently threaten their survival by modifying their behavior. We examine GPS data from a population of narwhals exposed to ship and seismic airgun noise during a controlled experiment in 2018 in the Scoresby Sound fjord system in Southeast Greenland. The fjord system has a complex shore line, restricting the behavioral response options for the narwhals to escape the threats. We propose a new continuous-time correlated velocity model with varying coefficients that includes spatial constraints on movement. To assess the sound exposure effect we compare a baseline model for the movement before exposure to a response model for the movement during exposure. Our model, applied to the narwhal data, suggests increased tortuosity of the trajectories as a consequence of the spatial constraints, and further indicates that sound exposure can disturb narwhal motion up to a couple of tens of kilometers. Specifically, we found an increase in velocity and a decrease in the movement persistence.
\end{abstract}

\begin{keyword}
\kwd{behavioral response study}
\kwd{stochastic differential equations}
\kwd{mixed effect model}
\end{keyword}

\end{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please use \tableofcontents for articles %%
%% with 50 pages and more                   %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Main text entry area:

\section{Introduction}

%general context
There is a large variety of sources of anthropogenic underwater noise in the Arctic region, including sonars, ice breakers, vessel traffic, drilling  and seismic airguns used for oil and gas exploration \citep{halliday_underwater_2020}.
A better understanding of the impact of these noises on marine life is critical for conservation policies. Marine mammals, whose vital functions highly depend on sound perception for communication and orientation, are particularly vulnerable. Besides causing physical harm, anthropogenic noise can also disturb their behavior and interrupt their natural foraging habits. However, there is still no clear criteria for measuring behavioral disturbances due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions \citep{southall_marine_2008,southall_marine_2019}. Comprehensive behavioral studies are thus essential.\\

%behavioral responses
Several surveys of behavioral responses have been conducted on beaked whales, which have been regularly involved in stranding events following sonar exercises \citep{tyack_beaked_2011,cioffi_trade-offs_2022}. Studies have also been conducted on belugas \citep{martin_exposure_2023}, sperm whales \citep{madsen_quantitative_2006}, blue whales \citep{friedlaender_preymediated_2016}, and narwhals \citep{heide-jorgensen_behavioral_2021, tervo_narwhals_2021, tervo_stuck_2023}.
The most critical behavioral responses - those potentially altering a population's capacity to survive, reproduce or forage - include changes in movement speed and direction, avoidance reactions as well as modified dive profiles or vocalizations \citep{southall_marine_2008}. 
Studies have shown that narwhals can exhibit some of these critical behavioral responses to sound exposure, such as avoidance reactions or changes of direction to move towards the shore \citep{heide-jorgensen_behavioral_2021}. A significant decrease in their buzzing rate has also been assessed as far as 40 km away from the sound source \citep{tervo_narwhals_2021}. These results are based on controlled exposure experiments conducted in 2017 and 2018 in a pristine area in South-East Greenland, still largely unaffected by noise pollution, and which is home to a declining population of narwhals \citep{garde_biological_2022}. Our study is based on these data collected in 2018. \\ %%% link sentence missing

%general purpose of the paper
This paper concentrates on the horizontal motion of narwhals. Previous analysis showed that the probabilities of the narwhals being close to the shore or moving towards the shore increase with exposure to the sound. However, these analyses were based on discrete-time discrete-space Markov chains, despite the continuous nature of narwhals movement in both time and space. To address this, we employ a continuous-time continuous-state model to analyze GPS positions, aiming to quantify the effects of sound exposure on the  speed and heading of the narwhals. This modeling approach offers multiple advantages: it avoids the need for positional interpolation at regular intervals, eliminates manual state labeling, incorporates GPS measurement errors, and allows to precisely quantify the magnitude of the changes in the movement depending on the distance to the sound source.\\

%the models we consider
We specifically focus on continuous-time models defined as a solution of a stochastic differential equation (SDE) with time-varying coefficients. One widely known model in this context is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \citep{johnson_continuous_2008}. It can be viewed as a special case of velocity potential models, where the potential is a second degree polynomial \citep{preisler_analyzing_2013}.  Varieties of CVMs have been applied to birds \citep{janaswamy_state_2018}, marine algae \citep{gurarie_estimating_2011}, ants \citep{russell_spatially_2018}, seals \citep{johnson_continuous_2008,albertsen_generalizing_2018}, bowhead whales \citep{gurarie_estimating_2011} and sea lion motion \citep{hanks_reflected_2017}.\\

% specificity of our data
In our case study, the narwhals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland), the largest fjord system in the world, known to be the summer residence for an isolated population of narwhals. Thus, the modeling process must include shoreline boundaries through movement constraints. Most existing CVMs for animal movement do not account for such constraints \citep{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
Example models including landscape boundaries are constrained SDEs  \citep{hanks_reflected_2017} in which a reflected version of the CVM is considered to study sea lion telemetry data, and SDEs with drift described by a potential function \citep{russell_spatially_2018} where a repulsive potential function  constrains ants movement within a box. \citep{brillinger_simulating_2003} also illustrates different methods of simulation for constrained animal movement, and emphasizes that the way to model motion near the boundary of the domain should be species-specific and account for the particular behaviors and spatial use of the studied animal. Narwhals have a proclivity to rotate to avoid or align with the shoreline without reaching the boundary. A reflected SDE as in \citep{hanks_reflected_2017} does not allow this type of behavior. In addition, a simple exponential potential function is suitable only with basic boundaries, such as the rectangular box in \citep{russell_spatially_2018}, but would break down for a complex boundary such as Scoresby Sound fjords. As a consequence, there is a need to develop new models adapted to such spatial constraints and behaviors. Reflected SDEs have certain limitations. On the one hand, it is not a modelling of shoreline avoidance behaviour. The model only uses a projection operator to ensure that the process remains within the domain. On the other hand, the transition density of these models is not known, which complicates the statistical inference of the parameters. We therefore propose to work with the second class of model, SDEs whose drift constraints movement within the system of fjords. \\

%new contributions
The novelty of this paper consists in extending the rotational CVM defined in \citep{gurarie_correlated_2017} with a rotation parameter expressed as a smooth function of the distance to the boundary and the angle between the animal's heading and the boundary normal vector. This makes the drift in the velocity equation dependent on the location process and the boundary of the domain, and allows the velocity to rotate as the narwhal approaches the shore. We show that the rotational parameter can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \citep{michelot_varying-coefficient_2021}. To make this possible, we derive explicit formulas for the transition density of the location and velocity process and formulate a linear Gaussian state-space model from the discretisation of these formulas. Maximum likelihood estimation based on the Kalman filter is then performed as for any other model in \texttt{smoothSDE}. The estimation procedure exploits the capabilities of the \texttt{R} package \texttt{TMB} to get approximate likelihood using Laplace's approximation \citep{kristensen_tmb_2016,albertsen_fast_2015}. It is considerably less computationally intensive than an estimation procedure based on Markov Chain Monte Carlo posterior samples as done in \citep{hanks_reflected_2017,russell_spatially_2018}.\\

In this general framework, sound exposure is introduced in the model through smooth functions of a covariate defined as the inverse of the distance to the sound source. This choice is motivated by \citep{heide-jorgensen_behavioral_2021, tervo_narwhals_2021,tervo_stuck_2023}.  We estimate the effect of the exposure variables on both the velocity and the persistence parameters of the diffusion process.  A response model is compared to a baseline model for the narwhal movement before exposure to the disturbance, that is, under normal conditions \citep{michelot_continuous-time_2022}. The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters. 
\\

%summary
To summarize, the main contributions of this paper are:
\begin{itemize}
	\item Definition of a constrained version of the CVM, where deviation angles from shoreline and distance to shore are used to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Derivation of explicit formulas for the transition density of the rotational CVM defined in \citep{gurarie_correlated_2017}, and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Analysis of narwhal data to assess a behavioural response using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

In Section 2, we give an overview of the narwhal data available for the analysis.
The diffusion models are then discussed in Section 3. Section 4 is dedicated to the statistical model used to infer the shore constraint and the sound exposure effects. Results of a simulation study are shown in Section 5. Finally, our models are applied to analyse the behavioral response of the narwhals in Section 6.

\section{Movement data of six narwals in South-East Greenland}
\label{section: movement data}

\subsection{Description of the controlled exposure experiments}
\label{subsection: data description}
The dataset analysed in this paper has been subject of several studies, focusing mainly on vocalizations and avoidance reactions \citep{heide-jorgensen_behavioral_2021,tervo_narwhals_2021,tervo_stuck_2023}. 
Here we briefly describe the data, for more details see \citep{heide-jorgensen_behavioral_2021}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound in South-East Greenland by biologists from the Greenland Institute of Natural Ressources, with the help of local Inuit hunters. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two airguns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every $80$ s, while the GPS navigation system recorded the location of every shot. 
The data include latitude and longitude of the narwhal positions, distance relative to the ship in metres, GPS time, and distance to the shore in metres. GPS positions are known only at times when the narwhals are at the surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours. While the statistical analysis in \citep{heide-jorgensen_behavioral_2021} relied on augmented data with positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.\\

For each narwhal, the entire track is split into two periods: a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and the period of exposure starting after exposure onset. We discard the GPS measurements from the first 12 hours after tagging, to avoid any tagging effects on the behavior. Measurements that result in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
Overall, $4815$ GPS positions are kept for the analysis. The splitting between data before and after exposure results in $1558$ measurements before exposure and $3257$ measurements after exposure.

\subsection{Notation and definition of covariates}
\label{subsection: covariates}

For each narwhal $i \in \{1,\ldots,N\}$ with $N=6$, we denote $n_i$ the total number of observations,   decomposed as $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure onset. The positions are observed at discrete times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, projected in UTM zone 26 North coordinates with the \texttt{R} package \texttt{rgdal}. These points are noisy observations of the underlying unobserved true positions.\\

The land geometry for the specific region in South-East Greenland is gathered from the \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. For each narwhal $i\in\{1, \ldots N\}$ and each time $j\in\{1, \ldots, n_i\}$, the closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$. Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This was due  to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons. After adjustement of the polygon data using \texttt{qgis} software,  approximately $1 \%$ of the observations remain on land. We kept these positions for the analysis. Distance to shore values range from $0$ to $7.6$ km. The shapefile defining the adjusted shoreline is made available in Supplementary materials. We define a new covariate $I^{shore}$ based on the inverse of the distance to the shore in kilometres with the following formula
\begin{equation}
	\left\{
	\begin{array}{lll}
		I^{shore}_{ij}=0 && \mbox{if } D^{shore}_{ij}>D_{up} \\
		I^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} && \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		I^{shore}_{ij}=\frac{1}{D_{low}} && \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of the observed distances, and $D_{up}=3$ km, assuming narwhals are influenced by the shore within a $3$ km range. 
The angle $\Theta_{ij}$ between the narwhal's heading and the shoreline is defined as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{ij+1}-y_{ij}}{\Delta_{ij}}$, where $\Delta_{ij}=t_{ij+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$, as illustrated in Figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates a movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement towards the shore.

\begin{figure}[ht!]
	\centering
	\begin{tikzpicture}
		%boundary
		\draw[gray, thick] (0,0) -- (0,4);
		%position
		\draw[red,fill=gray] (2,2) circle (.3ex);
		%nearest shore point
		\draw[red,fill=red] (0,2) circle (.3ex);
		%velocity
		\draw[->,gray] (4,0) -- (2,2);
		\draw[->,gray,dashed,thick] (2,2) coordinate (O) -- (0,4) coordinate (A);
		%normal vector
		\draw[->,red,dashed] (O) -- (4,2) coordinate (B);
		%trigonometric circle
		\draw[gray,dashed] (2,2) circle (2cm);
		%angle 
		\pic[draw=orange,->,angle eccentricity=1.2,angle radius=0.5cm] {angle=B--O--A};

		%nodes
		\node[left] at (-1,3.5) {Land};
		\node[right] at (1,3.5) {Water};
		\node[left] at (0,0) {$\partial \mathcal{D}$};
		\node[left] at (0,2) {$p_{ij}$};
		\node[left] at (1.2,2.8) {$\hat{v}_{ij}$};
		\node[below] at (1.8,2) {$y_{ij}$};
		\node[below,red] at (3,2) {$\vec{n}_{ij}$};
		\node[above,orange] at (2.3,2.3) {$\Theta_{ij}$};
	\end{tikzpicture}
	\caption{Example of nearest point on the shore and angle $\Theta$. $\partial D$ represents the boundary of the domain, $y_{ij}$ is the observed GPS position of narwhal $i$ at time $t_{ij}$.}
	\label{fig: illustration nearest shore points}
\end{figure}

The distance to the ship $D^{ship}$ (km) is defined as the distance between observed GPS locations of the ship and the narwhal. The values are comprised between $2.68$ and $63.8$ km.
Exposure to the ship for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \citep{heide-jorgensen_behavioral_2021}. The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals. Reasonably, the closer the narwhal is to the ship, the louder is the sound, and thus the exposure is larger. 
$D^{ship}$ is only defined when the narwhal is in line of sight with the ship and exposures are set to $0$ when this is not this case. This implies that in the statistical model, the ship noise is only allowed to affect the narwhal when the ship is in line of sight, though it is likely that narwhals can perceive the disturbance even when they are not in line of sight with the sound source. This provides a conservative estimate of the effect of the noise exposure. Levels of the covariates $I^{shore}$ and $E^{ship}$ for each narwhal are displayed in Figure~\ref{fig: realexpthroughtime}.

\begin{figure}[ht!]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.40]{images/data_exploration/realExpShip_through_time.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.40]{images/data_exploration/realExpShore_through_time.png}
		\caption{}
	\end{subfigure}
	\caption{Covariates of exposure and distance to shore over time for each narwhal. Time 0 corresponds to 12 hours after tagging for each individual whale. (a) Exposure to the ship (in $km^{-1}$). (b) Inverse of the distance to the shore (in $km^{-1}$).}  
	\label{fig: realexpthroughtime}
\end{figure}

\section{Movement models within a constrained landscape}

The dynamics of the true positions of the narwhals is modelled by an SDE. First, we recall the standard rotational CVM, and then we extend it to include the effect of the distance to shore on the movement and constrain the position within a polygon.

\subsection{Standard Rotational Advective correlated velocity model (RACVM)}
\label{section: RACVM}
\mbox{}\\

Let $A=\begin{pmatrix} 
	\frac{1}{\tau} &&-\omega \\
	\omega && \frac{1}{\tau}
\end{pmatrix}$ and define

\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$, $X(t)$ is the two-dimensional horizontal position, typically in longitude and latitude or in UTM coordinates, $W(t)$ is a two-dimensional brownian motion. Parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu=\begin{pmatrix} \mu_1 && \mu_2 \end{pmatrix}^\top$ is the long term velocity, and $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \citep{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}. Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it makes sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source. 
We refer to~\eqref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) when $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) when $\mu \neq (0,0)$.\\

In this section, we derive the explicit transition density for the process $U=\begin{pmatrix} X && V \end{pmatrix}^\top$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular with positive diagonal elements and two distinct autocorrelation parameters are considered in each direction allowing for anisotropic movement. In this case, \cite{albertsen_generalizing_2018} derives the Gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum. However, the formulas for the process $X$
are not provided, and the Euler scheme is employed to approximate the transition densities of $X$. This approach is generally sufficient when the animal’s positions are recorded at high frequency. Yet, for marine mammals such as narwhals that dive to great depths, GPS measurements are usually taken at irregular intervals, typically of several minutes, which can render the Euler scheme unsuitable. Additionally, incorporating covariates into the parameters of the SDE often requires approximating the transition density by assuming the covariates remain constant during each time step \citep{michelot_varying-coefficient_2021}. This results in two layers of approximation, possibly leading to unreliable estimation.\\

Here, we show that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in~\eqref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula. This is stated in the following proposition.\\

\begin{proposition}
	Let $U(t)=\begin{pmatrix} X(t) && V(t) \end{pmatrix}^\top$ for $t \geq 0$ be the solution to \eqref{eq: RACVM equation}. Then the exact transition density of the Markov process $U$ is given by 
	\begin{equation}
		U(t+\Delta) \vert U(t)=u \sim \mathcal{N}\left( T(\Delta) u +B(\Delta)\mu, Q(\Delta)\right) \mbox{ for all } \Delta>0 \mbox{ and } u \in \R^4
		\label{eq: X V distribution}
	\end{equation}
	where \begin{equation}
		T(\Delta)=\begin{pmatrix} I_2 && A^{-1}(I_2-\exp(-A\Delta)) \\
			0_2 && \exp(-A\Delta) \end{pmatrix} \mbox{, } B(\Delta)=\begin{pmatrix}
			\Delta I_2 -A^{-1}(I_2-\exp(-A\Delta))\\
			I_2-\exp(-A\Delta)\end{pmatrix}
		\label{eq: link matrices}
	\end{equation}
	and the covariance block matrix is given by
	\begin{equation}
		Q(\Delta)=\begin{pmatrix}
			q_1(\Delta)I_2 && \Gamma(\Delta) \\
			\Gamma(\Delta)^\top && q_2(\Delta)I_2
		\end{pmatrix}
		\label{eq: covariance matrix}
	\end{equation}
	with 
	\begin{equation*}q_1(\Delta)=\frac{\sigma^2}{C}\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{C} \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{C}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
	\end{equation*}

	\begin{equation*}
		q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-e^{-\frac{2 \Delta}{\tau}}\right)
	\end{equation*}
	\begin{equation*}\Gamma(\Delta)=\begin{pmatrix} \gamma_1 && \gamma_2 \\
			-\gamma_2 && \gamma_1\end{pmatrix}
	\end{equation*}
	where
	\begin{equation*}\gamma_1=\frac{\sigma^2}{2 C } \times \left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega\Delta)\right)
	\end{equation*}
	\begin{equation*}\gamma_2=\frac{\sigma^2}{C} \times\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)
	\end{equation*}
	and we denoted $C=\frac{1}{\tau^2}+\omega^2$ and $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$.
	\label{prop: transition density}
\end{proposition}

The formulas derived in \citep{johnson_continuous_2008} are a corollary of Proposition~\ref{prop: transition density}, obtained with $\omega=0$. The proof of this proposition is detailed in Appendix.

Equation~\eqref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) && \sin(\omega \Delta) \\ -\sin(\omega \Delta) && \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, ~\eqref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.\\

For locomotion under spatial constraints, we hypothesize that an increase in tortuosity is a sign of avoiding the boundary and adapting the heading to the spatial constraint. We use this to define a constrained version of the standard RCVM.

\subsection{Constrained rotational correlated velocity model}
\label{section: CRCVM}

We propose a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to describe how the animals turn in reaction to the shore. 
To do so, we consider $\omega$ as a smooth function both of the distance to the shore $D^{shore}$ and the angle between the velocity vector and the shore normal vector $\Theta$. 
We write $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

	\end{array}
	\right.
	\label{eq: CRCVM equation}
\end{equation}
with 
\begin{equation} 
	A(X(t),V(t))=\begin{pmatrix} 
		\frac{1}{\tau} && -\omega(X(t),V(t)) \\
		\omega(X(t),V(t)) && \frac{1}{\tau}
	\end{pmatrix}
	\label{eq: CRCVM matrix A}
\end{equation}
where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$ with  $f_{\omega}$ a smooth function from $\R^{+} \times [-\pi,\pi]$ to $\R$. In the sequel, such models will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).\\

To induce a rotation when the animal is close to the boundary and heading in the direction of the shore,
the following assumptions must guide the shape of $f_{\omega}$:
\begin{itemize}
	\item[1)] for a distance to the shore lower than some threshold, $f_{\omega}(D^{shore},\cdot)$ should be positive and increasing on $]\frac{\pi}{2},\pi[$, 
	\item[2)] for a distance to shore  lower than some threshold, $f_{\omega}(D^{shore},\cdot)$ should be negative and decreasing on $]-\pi,-\frac{\pi}{2}[$,
	\item[3)] as distance to shore decreases, the magnitude of the functions  $f_{\omega}(D^{shore},\cdot)$ on $]\frac{\pi}{2},\pi[$  and $]-\pi,-\frac{\pi}{2}[$ should increase.
\end{itemize}

We start by proposing a parametric model for the function $f_{\omega}$ and then discuss a non-parametric representation. For $\Theta \in [-\pi,\pi]$ and $D^{shore} \geq0$, we define $f_{\omega}$ as follows

\begin{equation}
	\label{eq: smooth omega}
	f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\Theta+\theta^{*}))+\tanh(\lambda(\Theta-\theta^{*})))\right)\times \exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)
\end{equation}
for some $D_0>0$, $\kappa>0$, $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,\pi]$.
The parameter $\omega_0$ controls the magnitude of the angular velocity, $D_0$ controls the distance at which the velocity starts to rotate to avoid the boundary, $\lambda$ and $\kappa$ control the steepness of the curves in each dimension, and $\theta^*$ is chosen so that the rotation starts when the angle $\Theta$ gets close to $\frac{\pi}{2}$ or $-\frac{\pi}{2}$. \\

This parametrization of $\omega$ is motivated by the observations above, and do not pretend to be optimal. 
Any smooth function, including~\eqref{eq: smooth omega}, could be approximated by a finite number of weighted spline basis functions of the form

\[f_\omega(\Theta,D^{shore})=\sum_{k=1}^L \omega_k \psi_k(\Theta,D^{shore})\]
The $\psi_k, k=1, \ldots, L$ can be constructed in several ways based on thin plate regression splines, or on the tensor product of any univariate spline basis. The coefficients $\omega_k$ control the contribution of each basis function to the final shape of the smooth function \citep{wood_generalized_2017}. Figure~\ref{fig: combinedperspplots}  shows function~\eqref{eq: smooth omega} in the top, along with a smooth function $f_{\omega}$ obtained as a combination of $L=9$ cubic spline basis functions approximating~\eqref{eq: smooth omega} in the bottom plots.

\begin{figure}[ht!]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.25]{images/crcvm/combined_persp_plots.png}
		\caption{}
		\label{fig: combinedperspplots}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\raisebox{0.5cm}{
			\includegraphics[scale=0.25]{images/crcvm/smooth_tau_bump.pdf}
		}
		\caption{}
		\label{fig: smoothtau}
	\end{subfigure}
\caption{(a) Examples of smooth functions $f_{\omega}$. Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.  Top: Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min. Bottom: Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions approximating~\eqref{eq: smooth omega}. (b) Example of a smooth parameter $\tau$ to get persistent trajectories along the shore. The function used is $\tau_0+\frac{\tau_1-\tau_0}{2}\times \left(\tanh(\frac{\Theta+\pi/2+\varepsilon}{\alpha})-\tanh(\frac{\Theta+\pi/2-\varepsilon}{\alpha})+\tanh(\frac{\Theta-\pi/2+\varepsilon}{\alpha})-\tanh(\frac{\Theta-\pi/2-\varepsilon}{\alpha})\right)$ for some parameters $\tau_0$, $\tau_1$, $\varepsilon$, $\alpha$.}

\end{figure}

Model~\eqref{eq: CRCVM equation} is flexible, and different species-specific behaviours can be described. Function $f_{\omega}$ controls at which distance the animal will turn away from the shore, and how fast it will turn.  Specific constraints on the smooth function $f_{\omega}$ can be set from biological knowledge about the species in question.

The tendency of an animal to move along the boundary of its spatial domain can be captured by the parameter $\tau$, as more persistent movement when the angle $\theta$ is close to $\pm\frac{\pi}{2}$ would induce such behavior. More precisely, to model persistent movement along the shoreline, we modify~\eqref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function:
\[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] with $f_{\tau}:[-\pi,\pi] \longrightarrow \R$.
The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain. An example of such a function is given in Figure~\ref{fig: smoothtau}. However, for simplicity we will assume $\tau$ constant in our analysis of the data.\\

Solving explicitly~\eqref{eq: CRCVM equation} is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$. But it is possible to get an approximate solution. We choose a time step $\Delta$ and approximate $D^{shore}$ and $\Theta$ by piecewise constant functions on each time step. We then use the formulas of Section~\ref{section: RACVM} to simulate the process and get the next position from the approximate transition densities. Sampled trajectories obtained within circular, rectangular and fjord domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown in Figure~\ref{fig: crcvm examples}. Depending on the smooth functions $f_{\omega}$ and $f_{\tau}$, the trajectories near the boundary have different characteristics. The trajectories labelled "persistent" tend to keep the same direction along the boundary, while the trajectories labelled standard turn whenever they are about to hit the boundary, and behave like an unconstrained CVM with $\omega=0$ otherwise.  The trajectories labelled "tortuous" also turn when going toward the shore, but keep turning when leaving the shore.\\

\begin{figure}[ht!]
	\includegraphics[scale=0.4]{images/crcvm/final_combined_plot.png}

	\caption{Five days trajectories with time step $\Delta=5$ min within circular (bottom), rectangular (middle) and fjords (top) domains. Plots labelled "Persistent CRCVM" and "Standard CRCVM" have the same smooth parameter $\omega$ defined by~\eqref{eq: smooth omega}, whereas the persistent trajectory has a smooth varying parameter $\tau$ as in Figure~\ref{fig: smoothtau}. The standard trajectory has a constant $\tau =1$h. Trajectories labelled "tortuous" are obtained from a constant $\tau=1$h and a smooth parameter $\omega$ defined with spline basis functions as in Figures~\ref{fig: combinedperspplots}. All trajectories have constant velocity parameter $\nu=4$km/h.}
 \label{fig: crcvm examples}
\end{figure}

While we have not established formal theoretical conditions for the function $f_{\omega}$ to guarantee constrained motion,   illustrative examples demonstrate how smooth, interpretable functions can produce constrained trajectories for the process $X$. A critical parameter in this model is the spatial scale of the rotational movement, defined as $\rho=\frac{\nu}{\vert \omega \vert}$, representing the diameter of rotation. To prevent the animal from hitting the boundary, $\rho$ should remain smaller than $D^{shore}$.\\

We will use the CRCVM \eqref{eq: CRCVM equation} allowing $\omega$ to vary to analyze narwhals trajectories while considering spatial constraints, whereas we keep $\tau$ fixed. In section~\ref{section: simulation constrained motion}, we show that a smooth function $f_{\omega}$ depending on $D^{shore}$ and $\Theta$ can be estimated from discrete time obsevations of the process $X$.

\section{Inference of the sound exposure effect}

To analyse potential perturbations in the locomotion of the narwhals due to sound exposure, we formulate a statistical mixed effect model for the parameters of the CRCVM, and discretise~\eqref{eq: X V distribution} to get an approximate state space model for inference with the Kalman filter.

\subsection{Mixed effect CRCVM model for  shore influence and sound exposure}
\label{subsection: ship exposure effect}

First a baseline model is fitted only on the data before exposure, to get an estimate of natural behavior, when the narwhals are not exposed to the ship and airgun sounds. Then, a response model is fitted on the data during exposure to test whether the estimated smooth parameters deviate significantly from the baseline values. We present here the baseline and response models.\\

For the baseline model, only the effect of the shore is included through the covariates $I^{shore}$ and $\Theta$. For each narwhal $i \in \{1, \cdots, N\}$ and each time $j\in\{1, \cdots, n_{i,pre}\}$, $y_{ij}$ is an observation with measurement error of position $X_{ij}$:

\begin{equation}
	y_{ij}=X_{ij}+\varepsilon_{ij} \quad 
	\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  
	\label{eq: baseline observations}
\end{equation}
%The process  $V_{ij}$ is  unobserved. 
The standard deviation $\sigma_{obs}$ represents GPS measurement errors variability. The assumption of i.i.d. Gaussian measurement errors is a simplified model of actual GPS measurement errors. In practice, the accuracy of GPS positions is influenced by the number of satellites processing the GPS signal—more satellites generally result in more accurate positions. Furthermore, the error can differ between the $x$ and $y$ directions. For a comprehensive study on Fastloc-GPS measurement errors, we refer to \citep{wensveen_path_2015}. It is often more appropriate to use a distribution with heavier tails, such as the Student's $t$-distribution, to model measurement error. However, this introduces additional complexity to the inference process, as a standard linear Kalman filter cannot be directly applied in such cases. Since the Student's $t$-distribution approaches a Gaussian distribution as its degrees of freedom increase, the Gaussian distribution is considered a reasonable trade-off between simplicity and suitability.\\

%We denote $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the CRCVM for narwhal $i$. 
The latent processes $X_i$ and $V_i$ solve

\begin{equation}  %\forall i \in \{1,\cdots,N\}, \quad 
	\left\{
	\begin{array}{l}
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} && -\omega_i(t) \\
			\omega_i(t) && \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
	\end{array}
	\right.
	\label{eq:baseline latent processes}
\end{equation}
%
where $\tau_i$, $\nu_i$ are individual parameters   to account for variability between narwhals:
%
\begin{equation}   
	%\begin{array}{l}
	\begin{pmatrix}	\log(\tau_i)\\ \log(\nu_i) \end{pmatrix}
	= \begin{pmatrix}\log(\tau_0)\\ \log(\nu_0) \end{pmatrix}
	+
	\begin{pmatrix} b_{\tau,i}\\  b_{\nu,i}  \end{pmatrix}
	\quad \mbox{ with } \quad
	\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 && 0 \\ 0 && \sigma_{\nu}^2 \end{pmatrix}\right) 
	%\end{array}
	\label{eq:baseline random effects}
\end{equation}

Since $\tau$ and $\nu$ are positive, a log link function is used for these parameters. The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$ are population intercepts and $b_{\tau,i}$, $b_{\nu,i}$
are the individual random effects. \\

Finally, as discussed in Section \ref{section: CRCVM}, the angular velocity function of individual $i$, $\omega_i$, is expressed as a smooth function of $\Theta$ and $I^{shore}$ through splines: 
%
\begin{equation} 
	\omega_i(t)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(I_i^{shore}(t),\Theta_i(t))
	\label{eq: baseline splines}
\end{equation}

Since the two covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines are more appropriate than thin plate regression splines.
The functions $\psi_k$ in equation~\eqref{eq: baseline splines} are the bivariate basis functions constructed from univariate cubic spline basis functions. We refer to \citep{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis functions. The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the number of knots in the splines are $q_E-1$ and $q_{\Theta}-1$ for the covariates $I_{shore}$ and $\Theta$, respectively.\\

The response model is then designed to assess a deviation from the baseline model due to exposure to the ship. The observations are supposed to have the same measurement errors as the baseline. For $i \in \{1, \ldots, N\}$ and $j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}$, we have

\begin{equation}  		y_{ij}=X_{ij}+\varepsilon_{ij}, \quad 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) 
	\label{eq: response observations}
\end{equation}

Then we add a dependency on the covariate $E_{ship}$ in the parameters $\tau_i$ and $\nu_i$: 
\begin{equation}    
	\begin{array}{l}
		\log(\tau_{i}(t))=\log(\tau_{0})+\alpha_{\tau} E^{ship}_i(t)+b_{\tau,i} \\
		\log(\nu_{i}(t))=\log(\nu_{0}) +  \alpha_{\nu} E^{ship}_i(t) +b_{\nu,i}  \\
	\end{array}
	\label{eq:response loglinear model}
\end{equation}

The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the smooth $\omega$ from the baseline model enter as offsets in the response model.
Thus, only the coefficients $\alpha_{\tau}$ and $\alpha_{\nu}$ are estimated from the data during exposure. These parameters control the deviation of $\tau_i$ and $\nu_i$ from their baseline values $\tau_0$ and $\nu_0$ as a function of the distance to the ship. Coefficients $\alpha_{\tau}$ and $\alpha_{\nu}$ significantly deviating from $0$ indicate an effect of the sound exposure on the narwhals horizontal motion.

\subsection{Linear Gaussian state-space model}
\label{section: state space model}
In the baseline statistical model \eqref{eq:baseline latent processes}, we make the approximation suggested in \citep{michelot_varying-coefficient_2021} that the function $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is 
\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(I_{ij}^{shore},\Theta_{ij}).\]
We write %$\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, 
$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
	\frac{1}{\tau_{i}} && -\omega_{ij} \\
	\omega_{ij} && \frac{1}{\tau_{i}}
\end{pmatrix}$ and $U_{ij}=\begin{pmatrix} X_{ij,1}  && X_{ij,2} && V_{ij,1} && V_{ij,2}\end{pmatrix}^\top$. % where $\log(\tau_i)=\log(\tau_0)+b_{\tau}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu}^{(i)}$.
%
We use the exact formulas from Proposition \ref{prop: transition density} to obtain the state-space matrix equations for $i\in\{1,\ldots, N\}$, and $j\in\{1, \ldots, n_{i,pre}\}$:%.
%Discretisation of~\eqref{eq: X V distribution} gives the following system:

\begin{equation}
	\begin{array}{l}
		y_{ij}=ZU_{ij}+\varepsilon_{ij},  \quad 
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
		U_{i,j+1}=T_{ij}\,U_{ij} + \eta_{ij}, \quad 
		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
	\end{array}
	\label{eq: RACVM state space}
\end{equation}
where $Z=\begin{pmatrix} I_2 && 0_{2,2}\end{pmatrix}$, the link matrix $T_{ij}$ is computed according to~\eqref{eq: link matrices} and the covariance matrix $Q_{ij}$ is computed according to~\eqref{eq: covariance matrix}.
\\
A linear Gaussian state-space model is obtained similarly for the response model. Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation.
The complete likelihood is computed from the observations $y_{ij}$ as a by-product of the Kalman filter algorithm \citep{michelot_varying-coefficient_2021}.
Laplace's approximation of the integral of this complete likelihood over the random effects is computed using the \texttt{R} package \texttt{TMB}  and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method, with the gradient being calculated by automatic differenciation. We refer to \citep{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations. We refer to \cite[section 6.6]{wood_generalized_2017} for the specification of the mixed effect model and the regularization terms that are based on a wiggliness measure involving Gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$. \\
This method is already implemented for the CVM ($\omega=0$) in \texttt{smoothSDE}. Here, we extend this inference method to the more general model~\eqref{eq: CRCVM equation} and use it for estimation of the baseline and the response models.

\section{Simulation study for the baseline model}
\label{section: simulation constrained motion}
To test the estimator of the constrained model for our application, we simulate CRCVM trajectories in the Scoresby Sound fjord system. We consider observations of a baseline model as described in Section~\ref{subsection: ship exposure effect}. We set $L=9$ degrees of freedom for the bivariate splines in the parameter $\omega$ and  we place the knots evenly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $1$ and $5$ km for $D^{shore}$. The values of the spline coefficients are chosen to get a process that turns fast enough to meet the spatial constraints. We fix $\tau_0=1.5$ h and $\nu_0=4$ km/h. The random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$. Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in the fjord system. \\

Two setups are considered: high frequency data with short trajectories and low frequency data with longer trajectories. For each setup, we simulate $M=100$ batches of $N$ individual trajectories with $N \in \{6,12\}$. In the high frequency framework, the time step between consecutive observations is $\Delta=1$ min, the overall observation time is $12$h and measurement error is Gaussian with standard deviation $\sigma_{obs}=5$ m. This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$. In the low frequency framework, the trajectories are simulated with one minute time steps and then subsampled to a time step $\Delta=5$ min. The overall observation time is $2.5$ days and measurement error is Gaussian with standard deviation $\sigma_{obs}=25$ m. This also amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.\\

For the estimation, we only keep the trajectories whose observed distances to the shore in km include the interval $[1,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$. The reason is that if a trajectory happens to be confined at a few hundreds metres from the shore in a corner of the fjords there is no sense in trying to estimate its movement $5$ km from the shore, and conversely, if the trajectories occur in open sea far from the shore, there is no sense in trying to estimate the movement characteristics $1$ km from the shore.\\

For the smooth function $\omega$, $9$ coefficients, including the intercept, are to be estimated, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the random effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. The measurement error $\sigma_{obs}$ is supposed to be known, usually given by the known GPS precision.  
We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$ in the optimization. 
For each batch of $N$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard error of the estimates. Simulation and optimization of the log-likelihood for one single batch of trajectories are performed on $16$ CPU cores within less than $2$ h. Parallelization within \texttt{R} is used to simulate the trajectories since it can be time consuming due to the calculation of the nearest shore point for each new sampled position. \\

The results are summarized in Table~\ref{table: simulation study results}. We show the mean and the standard error of the estimates for each coefficient in the high frequency and low frequency cases.

Estimation of the spline coefficients has overall less than $20 \%$ bias in the high frequency framework, and becomes more precise as $N$ increases. The variances of the random effects are the most challenging parameters to estimate. At least one of the standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$ is often estimated close to $0$. This tendency is clear for $N=6$ and $N=12$ individuals, though it seems to attenuate as $N$ increases. This is expected, since $N$ is the number of independent observations used for estimating these standard deviations. 
Regarding the intercepts, we obtain better estimates for the velocity parameter $\log(\nu_0)$ than for the persistence parameter $\log(\tau_0)$. The biases are about $4 \%$ and $15\%$, respectively,  and the standard error is lower for $\log(\nu_0)$.

Estimation is more difficult in the low frequency framework with higher measurement errors. Indeed, as the measurement error grows, the covariates are becoming less accurate, which clearly makes estimation less precise. In particular, we are unable to estimate the intercept $\log(\tau_0)$, as it remains close to the initial value. However, in practice this can be solved step by step by trying several initial values to find the coefficients giving the highest log-likelihood value. Some of the spline coefficients cannot be estimated as well. Perhaps surprisingly, the estimates of the random effects standard deviations seem slightly better in the low-frequency framework.\\

In practice, confidence intervals are often derived using the observed Fisher information matrix, which is done in Section~\ref{section: application}. 
It requires the Hessian matrix of the log-likelihood at the estimated coefficients to be positive definite. If this is not the case, this may indicate numerical errors or improper convergence.

\begin{landscape}

	\begin{table}
		\centering
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			&\multicolumn{3}{|c|}{Intercepts} & \multicolumn{8}{|c|}{Spline coefficients} & \multicolumn{2}{|c|}{SD} \\
			\hline
			&$\log(\tau_0)$ & $\log(\nu_0)$ & $\omega_0$ & $\omega_1$ & $\omega_2$ & $\omega_3$ & $\omega_4$  & $\omega_5$ & $\omega_6$ & $\omega_7$ & $\omega_8$ & $\sigma_{\tau}$ & $\sigma_{\nu}$ \\
			\hline 
			True & $0.41$ & $1.39$ & $0.02$ & $8.87$& $-2.54$& $6.72$&
			$22.50$& $6.73$& $30.88$& $4.78$& $6.62$& $0.2$ & $0.1$ \\
			\hline
			\multicolumn{14}{|c|}{High frequency} \\
			\hline
			\multirow{2}{4em}{$N=6$} & $0.35$ &
			$1.34$ &
			$-0.04$ &
			$7.68$&
			$-2.23$&
			$6.89$&
			$20.38$&
			$4.57$&
			$28.11$&
			$4.24$&
			$5.04$&

			$0.20$&
			$0.13$

			\\
			&	$\pm 0.16$ &
			$\pm 0.09$&
			$\pm 0.23$ & $\pm 1.37$ &
			$\pm 1.59$ &
			$\pm 1.74$ &
			$\pm 2.56$ &
			$\pm 3.67$ &
			$\pm 1.51$ &
			$\pm 1.27$ &
			$\pm 1.59$ 
			& $\pm 0.16$ &
			$\pm 0.08$
			\\
			\hline
			\multirow{2}{4em}{$N=12$} & $0.35$ &
			$1.42$ & $0.00$ & $7.54$&
			$-2.04$ &
			$7.02$ &
			$20.91$ &
			$5.33$ &
			$28.59$ &
			$4.74$ &
			$5.48$

			& $0.14$ &
			$0.09$

			\\
			& $ \pm 0.10$ &
			$ \pm0.05$ &
			$\pm 0.22$ &
			$\pm 0.94$ &
			$\pm 1.25$ &
			$\pm 1.10$ &
			$\pm 1.59$ &
			$\pm 2.70$ &
			$\pm 1.34$ &
			$\pm 0.86$ &
			$\pm 1.09$ &

			$\pm 0.09$ &
			$\pm 0.05$ 
			\\
			\hline	
			\multicolumn{14}{|c|}{Low frequency} \\
			\hline
			\multirow{2}{4em}{$N=6$}  & $0.06$ &
			$1.36$ &
			$0.00$ & $5.67$ &
			$0.78$&
			$6.20$ &
			$16.31$ &
			$6.26$ &
			$25.19$ &
			$4.52$ &
			$3.92$ &
			$0.22$ &
			$0.14$
			\\
			& $\pm 0.11$ &
			$\pm 0.03$ &
			$\pm 0.09$ & 
			$\pm 2.11$ &
			$\pm 2.45$ &
			$\pm 1.36$ &
			$\pm 6.33$ &
			$\pm 3.90$ &
			$\pm 2.96$ &
			$\pm 1.61$ &
			$\pm 2.03$ &
			$\pm 0.13$ &
			$\pm 0.05$

			\\
			\hline
			\multirow{2}{4em}{$N=12$} & $0.04$ &
			$1.39$ &
			$-0.02$ & $6.05$ &
			$-0.11$ &
			$6.19$ &
			$17.39$ &
			$4.03$ &
			$24.98$ &
			$5.01$ &
			$3.40$ & $0.25$ &
			$0.08 $
			\\
			& $\pm 0.06$  &
			$ \pm 0.03$ &
			$ \pm 0.11 $ & $ \pm 2.04$ &
			$ \pm 2.24$&
			$\pm 0.78$&
			$\pm 6.12$&
			$\pm 4.11$&
			$\pm 1.40$&
			$\pm 1.74$&
			$\pm 1.01$& $\pm 0.08$ &
			$\pm 0.03$

			\\
			\hline	
		\end{tabular}
		\caption{Results from the simulation study. Numbers are average $\pm$ standard error of the estimates obtained from $M=100$ simulated data sets in four scenarios: high and low frequency, and $N=6$ or $12$ individuals. High frequency are sampled at $\Delta = 1$ min, low frequency is sampled at $\Delta = 5$ min. There are 720 observations per individual in each simulated data set. The true values used in the simulations are given in the upper row.}
		\label{table: simulation study results}
	\end{table}

\end{landscape}

\section{Application to narwhal data}
\label{section: application}

In this section, we apply our CRCVM diffusion model to analyse the  behavioral response of the narwhals to ship and seismic airgun exposure.
The baseline model is fitted on the tracks before exposure to capture normal behavior. Deviations from the baseline is then assessed by fitting a model with covariate $E^{ship}$ on the tracks after exposure with the offsets estimated at baseline.

\subsection{Baseline estimation}
We estimate the baseline model as described in Section~\ref{subsection: ship exposure effect} on the data before exposure. We set the degree of freedom of the marginal splines to $5$ leading to a total number of $25$ degrees of freedom, fix the smoothing penalties of the tensor splines to $1$, and initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=1$, and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ to 0.
The measurement error standard deviation is fixed to $\sigma_{obs}=50$ m based on the measurement errors of Fastloc-GPS found in \citep{wensveen_path_2015}.\\

Estimation and confidence intervals on the parameter scale are shown in Table~\ref{table: baseline estimations}. 
The persistence is estimated to $\hat{\tau}_{0}=1.35 $ h.
In comparison, harbour seal in Alaska were shown to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \citep{johnson_continuous_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \citep{gurarie_correlated_2017}.
The intercept value for $\nu$ is estimated to $\hat{\nu}_{0}=4.60$ km/h which is of the order of magnitude of the average of the observed velocities. 
\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Coefficient   & Estimate  & $95\%$ CI \\
		\hline
		$\tau_0$ (h)   & $1.36$    & $[1.06; 1.71]$  \\
		$\nu_0$ (km/h)  & $4.60$   &  $[4.12; 5.12]$\\
		$\omega_0$ (rad/h)    & $0.00$    &  $[-0.14; 0.13]$ \\
		$\sigma_{\tau,pre}$    & 0.21   &  $[0.05; 0.56]$ \\
		$\sigma_{\nu,pre}$    & 0.09   &  $[0.02; 0.24]$ \\
		\hline
	\end{tabular}
	\caption{Baseline estimations}
	\label{table: baseline estimations}
\end{table}

Figure \ref{fig:febaseline3omegaanglenormalq0} shows the estimated smooth $\omega$ as a function of $\Theta$ for different values of the distance to shore. Movement towards the shoreline matches values of $\Theta$ such that $\vert \Theta \vert > \frac{\pi}{2}$. At $250$ m away from the shore, there is a clear rotation of the velocity estimated to be $\pm 2.5$ rad/h as $\Theta$ approaches $\pm\pi$. As distance increases, the magnitude of the angular velocity decreases, which means that the movement becomes less tortuous. This indicates that the spatial constraints influence the tortuosity of the movement by forcing the narwhal's velocity to rotate when they are close to the shoreline.   %However, the magnitude of the estimated $\omega$ is not sufficient to effectively obtain constrained trajectories similar to Figure~\ref{fig: crcvm examples} when simulating from the fitted model.
\begin{figure}[ht!]
	\centering
	\includegraphics[width=0.7\linewidth]{"images/application/baseline/omega_DistanceShore_levels.pdf"}
	\caption{Estimated smooth $\omega$ as a function of the angle $\Theta$ for fixed distances to the shore: $250$ metres for close, $1.41$ km for medium, and $3$ km or more for far (no effect of the shore).}

	\label{fig:febaseline3omegaanglenormalq0}
\end{figure}

Fixing a high standard deviation for the measurement error tends to increase the estimate of the parameter $\tau$. Intuitively, for a low value of the measurement error, a tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, and therefore a higher estimate of the persistence parameter. We fixed the measurement errors $\sigma_{obs}$ at different values and found that the highest log-likelihood value was obtained for a standard deviation $\sigma_{obs}$ close to $50$m. More details can be found in Appendix.

\subsection{Response estimation}

The baseline estimates of the spline coefficients $\omega_k$, the intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the standard deviations $\sigma_{\tau}$, $\sigma_{\nu}$ are used as offsets in the response model. Only the sound exposure effects on $\tau$ and $\nu$ are estimated from the data after exposure.
Estimates and $95\%$ confidence intervals are shown in Table~\ref{table: response estimations}. As the coefficients $\alpha_\tau$ and $\alpha_\nu$ significatively deviate from zero, this suggests that sound exposure influences both the persistence and the velocity of the movement.\\
%However, we miss the uncertainty coming from the baseline estimates here. To better quantify this uncertainty, we draw $100$ "posterior" samples of the coefficients of the baseline using the Hessian matrix of the log-likelihood to approximate the covariance, and we use the drawn coefficients as offsets for the response model. The mean of the estimates and the standard error over the $100$ estimations are $mean(\hat{\alpha}_\tau)=-3.26$, $mean(\hat{\alpha}_\nu)=0.86$, $mean(\hat{\sigma}_{\alpha_\tau})=0.57$ and $mean(\hat{\sigma}_{\alpha_\nu})=0.27$. \\

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Coefficient   & Estimate  & $95 \%$ CI \\
		\hline
		$\alpha_\tau$    & $-3.43$    &  $[-4.77; -2.04]$ \\
		$\alpha_\nu$  & $0.74$   &  $[0.23; 1.25]$\\
		\hline
	\end{tabular}
	\caption{Estimate of the parameters in the log-linear response model}
	\label{table: response estimations}
\end{table}

The estimated $\tau$ and $\nu$ are shown in Figure~\ref{fig: me response} as a function of the distance to the sound source. The value of $\nu$ increases with increased exposure to the sound. This should not come as a surprise since, for instance, the average empirical velocity norm when the narwhals are less than $5$ km away from the ship is $5.7$ km/h, which is more than $1$ km/h higher than the average empirical velocity norm before exposure.

We draw special attention to the parameter $\tau$, which decreases with increased exposure to the ship, implying lower persistence and lower autocorrelation in the velocity of the narwhals. This had been hinted at in \citep{heide-jorgensen_behavioral_2021}, where it was shown that the narwhals have more tendency to change direction and move towards the shore in presence of the ship. This can be interpreted as a drop in persistence due to the appearance of the ship. Our analysis suggests that the magnitude of this drop might be alarming: for most of the narwhals, at distance less than $5$ km from the ship, the persistence is half the baseline value. This is evidence of a strong shift in the behavior of the whales, which might have consequences on their capacity to rest and forage at short term.

\begin{figure}[ht!]
	\begin{subfigure}{0.98\textwidth}\centering
		\includegraphics[width=0.7\linewidth]{"images/application/response/plot_me_tau_updated.pdf"}
		\caption{}
	\end{subfigure}

	\begin{subfigure}{0.98\textwidth}
		\centering
		\includegraphics[width=0.7\linewidth]{"images/application/response/plot_me_nu_updated.pdf"}
		\caption{}
	\end{subfigure}
	\caption{Estimated effect of ship exposure on the parameters. The horizontal dashed lines represent the baseline values for each individual. The values on the $x$-axis are in kilometres. (a)  Persistence parameter $\tau$. Values on the $y$-axis are in hour. (b) Velocity parameter $\nu$. Values on the $y$-axis are in km/h.}
	\label{fig: me response}
\end{figure}

The distances to the ship $D^{ship}_{\tau}(p)$ and $D^{ship}_{\nu}(p)$  for which a percentage $p$ of deviation from the population baseline values $\tau_0$ and $\nu_0$ are reached are given by 
\begin{equation}
	D^{ship}_{\tau}(p)=\frac{\alpha_{\tau}}{\log(1-p)} \quad  \mbox{and} \quad  D^{ship}_{\nu}(p)=\frac{\alpha_{\nu}}{\log(1+p)}
\end{equation}
Table~\ref{table: baseline vs response parameters comparison} shows these values for different proportions $p$.
For $\tau$, $90\%$ of the baseline value is recovered at a distance about $32$ km. For $\nu$, $110\%$ of the baseline value is reached at a distance about $7$ km. This indicates that sound exposure can be perceived and can disturb the narwhals motion up to tens of kilometres.

We want to point out that the standard errors for these recovery distances are likely to be underestimated, due to the uncertainty in the baseline estimates, and the dependence between the measurement error and the estimates of the persistence parameters. However, the general take-away remains valid: there is evidence of a decrease in persistence and an increase in velocity up to a couple of tens of kilometers from the sound source. Along with other studies, we believe it can serve as a guideline for mitigation measures towards the effects of anthropogenic noise on the narwhals behavior.

\renewcommand{\arraystretch}{1.5} % Increase row height by 1.5 times

\begin{table}[ht!]
	\centering
	\begin{tabular}{| c | *{2}{w{c}{4em}|}}
		\hline
		\multirow{2}{*}{Percentage of deviation from baseline} & \multicolumn{2}{c|}{Recovery distance (km) }  \\
		\cline{2-3} 
		& $D^{ship}_{\tau}$ & $D^{ship}_{\nu}$ \\
		\hline
		$50 $ & $4.9 \pm 1.0$ &  $1.8 \pm 0.4$ \\
		$10$ & $32.5 \pm 6.7 $ & $7.8 \pm 2.9$ \\
		\hline

	\end{tabular}

	\caption{Estimated recovery distances of the baseline values within a $50 \%$ and $10\%$ range with standard errors.}
	\label{table: baseline vs response parameters comparison}
\end{table}

\section{Conclusion and perspectives}

%summary of the results

We introduced a new method to constrain a stochastic differential equation for animal movement in a bounded region of $\R^2$. Our approach relies on modeling the angular velocity as a function of the distance to the boundary and the angle between the velocity and the boundary normal vector. The additional term that constrains the motion is included in the drift, and, for this reason, acts as a confining potential.
We demonstrated how to simulate such diffusions and how to model different behaviors close to the boundary.\\

This new SDEs with smooth parameters depending on covariates was used to estimate the movement of the narwhals in the fjords. We managed to show an increased tortuosity of the trajectories when approaching the shore. More importantly, we showed that noise exposure has a significant effect on some parameters that drive the motion by comparing a baseline and a response model. We believe this method can be used as a basis for the assessment of behavioral responses in many contexts, and hope it will help understanding better the effects of anthropogenic noise on marine mammals movement.\\

We emphasize that, even if we were able to estimate quite reliably the smooth parameters of the SDE, a part of uncertainty still remains. First, the log-likelihood function that we rely on for the estimation is approximate, based on piecewise constant parameter values on each time step. Given that observations are relatively high frequency (5 min between consecutive observations in median), the error introduced is expected to be low, though we don't have theoretical bounds for it. Moreover, the constant covariate value that is used on each time step is computed from the GPS observations, even though these observations come with GPS measurement error. We don't integrate the measurement error in the covariates $\Theta$, $E^{ship}$ and $I^{shore}$ that we are considering. Since GPS measurement error are pretty low (a few tens of metres) compared to ARGOS for instance, this may not introduce significant error in the estimations. Eventually, Laplace approximation is used to approach the integral of likelihood over the random effects. We did not discuss the error that may be introduced by this approximation, but the simulation study show that despite all the approximations made along the way, we are indeed able to get reliable estimates even for pretty complex models.\\

In the future, studying theoretical properties of the SDE models we used here might be of significant interest. For instance, we did not prove any result or exhibit clear assumptions that would guarantee that the process is effectively constrained. We noticed that in the literature of animal movement modeling, whether that would be with SDEs, HMM or step-selection functions, spatial constraints are rarely considered. We hint that better considerations of the spatial constraints can give new insights into the interactions between marine mammals and their environment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Support information, if any,             %%
%% should be provided in the                %%
%% Acknowledgements section.                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Funding information, if any,             %%
%% should be provided in the                %%
%% funding section.                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{funding}
% The first author was supported by ...
The authors would like to thank the CNRS and IRN Madef for funding A. Delporte, and MIAI - ANR-19-P3IA-0003 for the funding of A Samson.
We also thank the project MATH-AmSud 23-MATH-12.
% The second author was supported in part by ...
\end{funding}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, including data   %%
%% sets and code, should be provided in     %%
%% {supplement} environment with title      %%
%% and short description. It cannot be      %%
%% available exclusively as external link.  %%
%% All Supplementary Material must be       %%
%% available to the reader on Project       %%
%% Euclid with the published article.       %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Shapefile for land geometry}
\sdescription{Shapefile based on OpenStreetMap data for the land polygons in Scoresby Sound fjords system. We used high resolution satellite images to improve it with \texttt{qgis} software}.
\end{supplement}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  imsart-nameyear.bst  will be used to                   %%
%%  create a .BBL file for submission.                     %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%  MR numbers will be added by VTeX.                      %%
%%                                                         %%
%%  Use \citep{...} to cite references in text.             %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{imsart-nameyear} % Style BST file
\bibliography{aoas-template}       % Bibliography file (usually '*.bib')
\end{document}

### 2024-11-08 R/application/CRCVM/fit_crcvm.Rmd
reorganize folders

@@
 
-D_low=0.07
+D_low=0.073
 D_up=3
@@
 #model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1,omega=~1)
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
+formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),omega=~1)
+
+# Fit baseline 
+baseline0<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+baseline0$fit()
+
+# Fit baseline with fixed 25m measurement error
+sigma_obs=0.025
 H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
 
-baseline0_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
+baseline0_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                     par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_35m$fit()
+baseline0_25m$fit()
 
@@
 ```{r results baseline0}
-estimates_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Est")
-std_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Std")
+
+estimates_bas0=as.list(baseline0$tmb_rep(),what="Est")
+std_bas0=as.list(baseline0$tmb_rep(),what="Std")
+
+estimates_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Est")
+std_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Std")
 
@@
 par0 <- c(0,0,1,4,0)
-init_lambda=c(1/0.2^2,1/0.2^2,0.01)
+init_lambda=c(1,1,1)
 
@@
                  nu=~s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=5,bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_35m$update_lambda(init_lambda)
-
-baseline1_35m$fit()
+                 omega=~s(AngleNormal,k=5,bs="cp"))
+
+# Fit baseline 
+baseline1<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+
+baseline1$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
+baseline1$fit()
 
@@
 baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
 baseline1_50m$update_lambda(init_lambda)
+
+baseline1_50m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
 
@@
 baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
 baseline1_75m$update_lambda(init_lambda)
+
+baseline1_75m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
 
@@
 baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 
+baseline1_100m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
+
 baseline1_100m$fit()
@@
 ```{r results baseline1}
-estimates_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Est")
-std_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Std")
+
+#get estimates
+estimates_bas1=as.list(baseline1$tmb_rep(),what="Est")
+std_bas1=as.list(baseline1$tmb_rep(),what="Std")
 
@@
 
-plots_bas1_35m=baseline1_35m$get_all_plots(baseline=NULL,model_name="baseline1_35m",show_CI="pointwise",save=TRUE)
-plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
-plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
-plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)
-
+#get plots
+
+plots_bas1=baseline1$get_all_plots(baseline=NULL,show_CI="simultaneous")
+plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,show_CI="pointwise")
+plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,show_CI="pointwise")
+plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,show_CI="pointwise")
+
+```
+
+```{r display plots baseline1}
+
+plots_bas1
+plots_bas1_50m
+plots_bas1_75m
+plots_bas1_100m
 ```
@@
 par0 <- c(0,0,1,4,0)
-init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)
+init_lambda=c(1,1,0.1,1,1,1)
 
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_35m$update_lambda(init_lambda)
-baseline2_35m$fit()
+                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cp")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
+
+# Fit baseline 
+baseline2<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.035)),fixpar=c("mu1","mu2"))
+
+baseline2$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",rep(NA,4)))))
+baseline2$fit()
 
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline2_50m$update_lambda(init_lambda)
+baseline2_50m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",rep(NA,4)))))
 baseline2_50m$fit()
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline2_75m$update_lambda(init_lambda)
+baseline2_75m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",rep(NA,4)))))
 baseline2_75m$fit()
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline2_100m$update_lambda(init_lambda)
+baseline2_100m$update_map(new_map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",rep(NA,4)))))
 baseline2_100m$fit()
@@
 
-estimates_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Est")
-std_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Std")
+estimates_bas2=as.list(baseline2$tmb_rep(),what="Est")
+std_bas2=as.list(baseline2$tmb_rep(),what="Std")
 
@@
 
-xmin=list("DistanceShore"=0.07)
+xmin=list("DistanceShore"=0.073)
 xmax=list("DistanceShore"=5)
 
-plots_bas2_35m=baseline2_35m$get_all_plots(baseline=NULL,model_name="baseline2_35m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
+plots_bas2=baseline2$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,show_CI="pointwise")
+
+plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,show_CI="pointwise")
+plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,show_CI="pointwise")
+plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,show_CI="pointwise")
+```
+
+```{r display plots baseline2}
+
+#print plots
+plots_bas2
+plots_bas2_50m
+plots_bas2_75m
+plots_bas2_100m
+
+```
+
+```{r save plots baseline2}
+# Create directory
+if (!dir.exists("baseline2_50m")) {
+    dir.create("baseline2_50m", recursive = TRUE)
+}
+
+# Use lapply to save each plot
+lapply(seq_along(plots_bas2), function(i) {
+
+    plot<-plots_bas2[[i]]
+    plot_name <- names(plots_bas2)[i]
+    file_path <- file.path("baseline2_50m", plot_name)
+
+    # Check if the plot is a ggplot or plotly object
+    if (inherits(plot, "ggplot")) {
+      ggsave(file_path, plot, device = "pdf")
+    } else if (inherits(plot, "plotly")) {
+      saveWidget(plot, file_path)
+  }
+    })
 ```
@@
 
-post_coeff_bas2_35m=baseline2_35m$post_coeff(n_post=1000)
-post_par_bas2_35m=list("tau0"=exp(post_coeff_bas2_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,2])))
-mean_par_bas2_35m=lapply(post_par_bas2_35m,mean)
-sd_par_bas2_35m=lapply(post_par_bas2_35m,sd)
+post_coeff_bas2=baseline2$post_coeff(n_post=1000)
+post_par_bas2=list("tau0"=exp(post_coeff_bas2$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_bas2$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_bas2$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_bas2$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_bas2$log_lambda[,2])))
+mean_par_bas2=lapply(post_par_bas2,mean)
+sd_par_bas2=lapply(post_par_bas2,sd)
 
@@
 par0 <- c(0,0,1,4,0)
-init_lambda=c(1,1,0.01,1,1,1)
-
+init_lambda=c(1,1,1,1,1,1)
+
+#mapping for fixed smooth penalties
+new_map=list("log_lambda"=factor(c("tau.s(ID)","nu.s(ID)",rep(NA,4))))
+
+#parameters formulas
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
+                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+
+                   ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
+
+# Fit baseline
+baseline3<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
+                    response = c("x","y"),par0 = par0,other_data=list("log_sigma_obs0"=log(0.035)),
+                    fixpar=c("mu1","mu2"))
+
+baseline3$update_lambda(init_lambda)
+baseline3$update_map(new_map)
+baseline3$fit()
+
+# Fit baseline with 30m measurement error
+sigma_obs=0.03
 H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
 
-baseline3_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
+baseline3_30m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_35m$update_lambda(init_lambda)
-baseline3_35m$fit()
+                    fixpar=c("mu1","mu2"))
+
+baseline3_30m$update_lambda(init_lambda)
+baseline3$update_map(new_map)
+baseline3_30m$fit()
+
+# Fit baseline with 40m measurement error
+sigma_obs=0.04
+H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
+
+baseline3_40m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+                    fixpar=c("mu1","mu2"))
+
+baseline3_40m$update_lambda(init_lambda)
+baseline3_40m$update_map(new_map)
+baseline3_40m$fit()
+
+# Fit baseline with 45m measurement error
+sigma_obs=0.045
+H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
+
+baseline3_45m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+                    fixpar=c("mu1","mu2"))
+
+baseline3_45m$update_lambda(init_lambda)
+baseline3_45m$update_map(new_map)
+baseline3_45m$fit()
 
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline3_50m$update_lambda(init_lambda)
+baseline3_50m$update_map(new_map)
 baseline3_50m$fit()
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline3_75m$update_lambda(init_lambda)
+baseline3_75m$update_map(new_map)
 baseline3_75m$fit()
@@
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
+                    fixpar=c("mu1","mu2"))
 
 baseline3_100m$update_lambda(init_lambda)
+baseline3_100m$update_map(new_map)
 baseline3_100m$fit()
 ```
+
 ```{r get results baseline3}
 
-estimates_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Est")
-std_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Std")
+estimates_bas3=as.list(baseline3$tmb_rep(),what="Est")
+std_bas3=as.list(baseline3$tmb_rep(),what="Std")
+
+estimates_bas3_30m=as.list(baseline3_30m$tmb_rep(),what="Est")
+std_bas3_30m=as.list(baseline3_30m$tmb_rep(),what="Std")
+
+estimates_bas3_40m=as.list(baseline3_40m$tmb_rep(),what="Est")
+std_bas3_40m=as.list(baseline3_40m$tmb_rep(),what="Std")
+
+estimates_bas3_45m=as.list(baseline3_45m$tmb_rep(),what="Est")
+std_bas3_45m=as.list(baseline3_45m$tmb_rep(),what="Std")
 
@@
 
-plots_bas3_35m=baseline3_35m$get_all_plots(baseline=NULL,model_name="baseline3_35m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
+plots_bas3=baseline3$get_all_plots(baseline=NULL,xmin=xmin,
+                                   xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_30m=baseline3_30m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                           link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_40m=baseline3_40m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                           link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_45m=baseline3_45m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                           link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                           link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                           link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                                             link=link,xlabel=xlabel,show_CI="pointwise")
+
+```
+
+```{r display plots baseline3}
+
+#print plots
+plots_bas3
+plots_bas3_50m
+plots_bas3_75m
+plots_bas3_100m
+
+```
+
+```{r}
+# Create directory
+if (!dir.exists("baseline3_50m")) {
+    dir.create("baseline3_50m", recursive = TRUE)
+}
+
+# Use lapply to save each plot
+lapply(seq_along(plots_bas2), function(i) {
+
+    plot<-plots_bas2[[i]]
+    plot_name <- names(plots_bas2)[i]
+    file_path <- file.path("baseline3_50m", plot_name)
+
+    # Check if the plot is a ggplot or plotly object
+    if (inherits(plot, "ggplot")) {
+      ggsave(file_path, plot, device = "pdf")
+    } else if (inherits(plot, "plotly")) {
+      saveWidget(plot, file_path)
+  }
+    })
 ```
@@
 
-post_coeff_bas3_35m=baseline3_35m$post_coeff(n_post=1000)
-post_par_bas3_35m=list("tau0"=exp(post_coeff_bas3_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,2])))
-mean_par_bas3_35m=lapply(post_par_bas3_35m,mean)
-sd_par_bas3_35m=lapply(post_par_bas3_35m,sd)
-
-post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
+post_coeff_bas3=baseline3$post_coeff(n_post=10000)
+post_par_bas3=list("sigma_obs"=exp(post_coeff_bas3$log_sigma_obs),
+              "tau0"=exp(post_coeff_bas3$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_bas3$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_bas3$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_bas3$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_bas3$log_lambda[,2])))
+mean_par_bas3=lapply(post_par_bas3,mean)
+sd_par_bas3=lapply(post_par_bas3,sd)
+quant_par_bas3 <- lapply(post_par_bas3, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_30m=baseline3_30m$post_coeff(n_post=10000)
+post_par_bas3_30m=list("tau0"=exp(post_coeff_bas3_30m$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_bas3_30m$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_bas3_30m$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_30m$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_30m$log_lambda[,2])))
+mean_par_bas3_30m=lapply(post_par_bas3_30m,mean)
+sd_par_bas3_30m=lapply(post_par_bas3_30m,sd)
+quant_par_bas3_30m <- lapply(post_par_bas3_30m, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_40m=baseline3_40m$post_coeff(n_post=10000)
+post_par_bas3_40m=list("tau0"=exp(post_coeff_bas3_40m$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_bas3_40m$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_bas3_40m$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_40m$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_40m$log_lambda[,2])))
+mean_par_bas3_40m=lapply(post_par_bas3_40m,mean)
+sd_par_bas3_40m=lapply(post_par_bas3_40m,sd)
+quant_par_bas3_40m <- lapply(post_par_bas3_40m, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_45m=baseline3_45m$post_coeff(n_post=10000)
+post_par_bas3_45m=list("tau0"=exp(post_coeff_bas3_45m$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_bas3_45m$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_bas3_45m$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_45m$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_45m$log_lambda[,2])))
+mean_par_bas3_45m=lapply(post_par_bas3_45m,mean)
+sd_par_bas3_45m=lapply(post_par_bas3_45m,sd)
+quant_par_bas3_45m <- lapply(post_par_bas3_45m, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=10000)
 post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
@@
 sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)
-
-post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
+quant_par_bas3_50m <- lapply(post_par_bas3_50m, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=10000)
 post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
@@
 sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)
-
-post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
+quant_par_bas3_75m <- lapply(post_par_bas3_75m, quantile, probs = c(0.025, 0.975))
+
+post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=10000)
 post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
@@
 sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)
+quant_par_bas3_100m <- lapply(post_par_bas3_100m, quantile, probs = c(0.025, 0.975))
 
@@
 ```{r smooth omega baseline3}
-
 plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
@@
 combined_plot <- ggplot() +
-  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
-  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
-  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
-  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
-  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
-  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
-  theme_minimal()
-
-ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")
+  geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
+  geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
+  labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
+  geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
+  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
+  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
+  theme_minimal() +
+  theme(
+    axis.title.x = element_text(size=14),
+    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
+    axis.text.y = element_text(size=12),
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    legend.title = element_text(size=14),
+    legend.text = element_text(size=12))
+
+ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline3", width=10, height=6, units="in")
 
@@
 
-# Fit response with 50m measurement error
-sigma_obs=0.04
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_35m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
+#mapping for fiwed coefficients
+new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                     "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_35m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_35m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_35m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_35m$lambda()["tau.s(ID)",1],1,baseline1_35m$lambda()["nu.s(ID)",1],
-             baseline1_35m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_35m$update_coeff_re(new_coeff_re)
-response1_sp_35m$update_coeff_fe(baseline1_35m$coeff_fe()[,1])
-
-response1_sp_35m$update_lambda(new_lambda)
-
-response1_sp_35m$fit()
+                    log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,NA)))
+#Fit baseline
+response1_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+
+new_coeff_re=c(rep(0,2),baseline1$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
+               baseline1$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
+               baseline1$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
+new_lambda=c(1,baseline1$lambda()["tau.s(ID)",1],1,baseline1$lambda()["nu.s(ID)",1],
+             baseline1$lambda()["omega.s(AngleNormal)",1])
+
+response1_sp$update_map(new_map)
+response1_sp$update_coeff_re(new_coeff_re)
+response1_sp$update_coeff_fe(baseline1$coeff_fe()[,1])
+
+response1_sp$update_lambda(new_lambda)
+
+response1_sp$fit()
 
@@
 response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
              baseline1_50m$lambda()["omega.s(AngleNormal)",1])
+
+response1_sp_50m$update_map(new_map)
 response1_sp_50m$update_coeff_re(new_coeff_re)
@@
 response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
              baseline1_75m$lambda()["omega.s(AngleNormal)",1])
+
+response1_sp_75m$update_map(new_map)
 response1_sp_75m$update_coeff_re(new_coeff_re)
@@
 response1_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
              baseline1_100m$lambda()["omega.s(AngleNormal)",1])
+
+response1_sp_100m$update_map(new_map)
 response1_sp_100m$update_coeff_re(new_coeff_re)
@@
 
-estimates_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Est")
-std_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Std")
+estimates_res1_sp=as.list(response1_sp$tmb_rep(),what="Est")
+std_res1_sp=as.list(response1_sp$tmb_rep(),what="Std")
 
@@
 
-plots_res1_sp_35m=response1_sp_35m$get_all_plots(baseline=baseline1_35m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_35m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_50m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_75m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_100m",show_CI="pointwise",save=TRUE)
+plots_res1_sp=response1_sp$get_all_plots(baseline=baseline1,xmin=xmin,
+                                         xmax=xmax,link=link,xlabel=xlabel,show_CI="simultaneous")
+plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,
+                                                 xmax=xmax,link=link,xlabel=xlabel,show_CI="simultaneous")
+plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,
+                                                 link=link,xlabel=xlabel,show_CI="simultaneous")
+plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,
+                                                   xmax=xmax,link=link,xlabel=xlabel,show_CI="simultaneous")
+
+```
+
+```{r display plots response1_sp}
+
+#print plots
+plots_res1_sp
+plots_res1_sp_50m
+plots_res1_sp_75m
+plots_res1_sp_100m
 
@@
 ```{r estimates on parameter scale response1}
-
-post_coeff_res1_sp_35m=response1_sp_35m$post_coeff(n_post=1000)
-post_par_res1_sp_35m=list("tau0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,2])))
-mean_par_res1_sp_35m=lapply(post_par_res1_sp_35m,mean)
-sd_par_res1_sp_35m=lapply(post_par_res1_sp_35m,sd)
+post_coeff_res1_sp=response1_sp$post_coeff(n_post=1000)
+post_par_res1_sp=list("tau0"=exp(post_coeff_res1_sp$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_res1_sp$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_res1_sp$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,2])))
+mean_par_res1_sp=lapply(post_par_res1_sp,mean)
+sd_par_res1_sp=lapply(post_par_res1_sp,sd)
 
@@
 
+#mapping for fixed coefficients
+new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",
+        rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),
+        coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))
+
+#Fit baseline
+response2_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+
+new_coeff_re=c(rep(0,2),baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
+               rep(0,2),baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
+new_lambda=c(1,baseline2$lambda()["tau.s(ID)",1],1,baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])
+
+response2_sp$update_map(new_map)
+response2_sp$update_coeff_re(new_coeff_re)
+response2_sp$update_coeff_fe(baseline2$coeff_fe()[,1])
+response2_sp$update_lambda(new_lambda)
+
+response2_sp$fit()
+
 # Fit response with 50m measurement error
@@
 response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
+
+response2_sp_50m$update_map(new_map)
 response2_sp_50m$update_coeff_re(new_coeff_re)
-response2_sp_50m$update_coeff_fe(baseline2$coeff_fe()[,1])
+response2_sp_50m$update_coeff_fe(baseline2_50m$coeff_fe()[,1])
 
@@
 response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
+
+response2_sp_75m$update_map(new_map)
 response2_sp_75m$update_coeff_re(new_coeff_re)
-response2_sp_75m$update_coeff_fe(baseline2$coeff_fe()[,1])
+response2_sp_75m$update_coeff_fe(baseline2_75m$coeff_fe()[,1])
 
@@
 response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
+
+response2_sp_100m$update_map(new_map)
 response2_sp_100m$update_coeff_re(new_coeff_re)
-response2_sp_100m$update_coeff_fe(baseline2$coeff_fe()[,1])
+response2_sp_100m$update_coeff_fe(baseline2_100m$coeff_fe()[,1])
 
@@
 ```{r get results response2}
-estimates_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
-std_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")
-
-estimates_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
-std_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")
-
-estimates_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
-std_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")
+
+estimates_res2_sp=as.list(response2_sp$tmb_rep(),what="Est")
+std_res2_sp=as.list(response2_sp$tmb_rep(),what="Std")
+
+estimates_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
+std_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")
+
+estimates_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
+std_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")
+
+estimates_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
+std_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")
 
@@
 
-plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
+plot_res2_sp=response2_sp$get_all_plots(baseline=baseline2,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+```
+
+```{r display plots response2_sp}
+
+#print plots
+plot_res2_sp
+plot_res2_sp_50m
+plot_res2_sp_75m
+plot_res2_sp_100m
+
 ```
@@
 ```{r}
+
+post_coeff_res2_sp=response2_sp$post_coeff(n_post=1000)
+post_par_res2_sp=list("tau0"=exp(post_coeff_res2_sp$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_res2_sp$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_res2_sp$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,2])))
+mean_par_res2_sp=lapply(post_par_res2_sp,mean)
+sd_par_res2_sp=lapply(post_par_res2_sp,sd)
 
@@
 #initialize parameters
-par0=c(0,0,1,4,0)
+par0=c(0,0,1.5,4.5,0)
 
@@
 
+#mapping for fixed coeffs
+new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
+                    "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
+                    log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))
+
+#Fit response
+response3_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("log_sigma_obs0"=log(0.07)))
+
+new_coeff_re=c(rep(0,2),baseline3$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
+               rep(0,2),baseline3$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
+new_lambda=c(1,baseline3$lambda()["tau.s(ID)",1],1,baseline3$lambda()["nu.s(ID)",1],baseline3$lambda()[3:6,1])
+
+response3_sp$update_map(new_map)
+response3_sp$update_coeff_re(new_coeff_re)
+response3_sp$update_coeff_fe(baseline3$coeff_fe()[,1])
+
+response3_sp$update_lambda(new_lambda)
+response3_sp$fit()
+
 #Fit response with 50m measurement error
@@
 response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(1,baseline3_50m$lambda()["tau.s(ID)",1],1,baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
+
+response3_sp_50m$update_map(new_map)
 response3_sp_50m$update_coeff_re(new_coeff_re)
@@
 response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(1,baseline3_75m$lambda()["tau.s(ID)",1],1,baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
+
+response3_sp_75m$update_map(new_map)
 response3_sp_75m$update_coeff_re(new_coeff_re)
@@
 response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(1,baseline3_100m$lambda()["tau.s(ID)",1],1,baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
+
+response3_sp_100m$update_map(new_map)
 response3_sp_100m$update_coeff_re(new_coeff_re)
@@
 ```{r get results response3}
+
+estimates_res3_sp=as.list(response3_sp$tmb_rep(),what="Est")
+std_res3_sp=as.list(response3_sp$tmb_rep(),what="Std")
+
 estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
@@
 
-plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
+plot_res3_sp=response3_sp$get_all_plots(baseline=baseline3,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+```
+
+```{r display plots response3_sp}
+
+#print plots
+plot_res3_sp
+plot_res3_sp_50m
+plot_res3_sp_75m
+plot_res3_sp_100m
 
@@
 ```{r estimates on parameter scale response3}
+
+post_coeff_res3_sp=response3_sp$post_coeff(n_post=1000)
+post_par_res3_sp=list("tau0"=exp(post_coeff_res3_sp$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_res3_sp$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_res3_sp$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,2])))
+mean_par_res3_sp=lapply(post_par_res3_sp,mean)
+sd_par_res3_sp=lapply(post_par_res3_sp,sd)
 
@@
                  nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
+                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")
+                 +ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
+
+new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
+                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))
+
+# Fit response
+response2_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),
+                    other_data=list("log_sigma_obs0"=log(0.05)))
+
+new_coeff_fe=c(baseline2$coeff_fe()[1:3,1],0,baseline2$coeff_fe()[4,1],0,baseline2$coeff_fe()[5,1])
+new_coeff_re=c(baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
+new_lambda=c(baseline2$lambda()["tau.s(ID)",1],baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])
+
+response2_exp$update_map(new_map)
+response2_exp$update_coeff_re(new_coeff_re)
+response2_exp$update_coeff_fe(new_coeff_fe)
+
+response2_exp$update_lambda(new_lambda)
+
+response2_exp$fit()
 
@@
 response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
+
+response2_exp_50m$update_map(new_map)
 response2_exp_50m$update_coeff_re(new_coeff_re)
@@
 response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
+
+response2_exp_75m$update_map(new_map)
 response2_exp_75m$update_coeff_re(new_coeff_re)
 response2_exp_75m$update_coeff_fe(new_coeff_fe)
-
 response2_exp_75m$update_lambda(new_lambda)
@@
 response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
+
+response2_exp_100m$update_map(new_map)
 response2_exp_100m$update_coeff_re(new_coeff_re)
@@
 ```{r}
+
+estimates_res2_exp=as.list(response2_exp$tmb_rep(),what="Est")
+std_res2_exp=as.list(response2_exp$tmb_rep(),what="Std")
+
 estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
@@
 
-plots_res2_exp=response2_exp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
+plots_res2_exp=response2_exp$get_all_plots(baseline=baseline2,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res2_exp_50m=response2_exp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+```
+
+```{r display plots response3_sp}
+
+#print plots
+plots_res2_exp
+plots_res2_exp_50m
+plots_res2_exp_75m
+plots_res2_exp_100m
 
@@
 
-par0=c(0,0,1,4,0)
+par0=c(0,0,1,4.5,0)
 
@@
 
+#mapping for fixed coeffs
+new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
+                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))
+
+#Fit response
+sigma_obs=exp(estimates_bas3$log_sigma_obs)
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+response3_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
+
+new_coeff_fe=c(baseline3$coeff_fe()[1:3,1],0,baseline3$coeff_fe()[4,1],0,baseline3$coeff_fe()[5,1])
+new_coeff_re=c(baseline3$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
+               baseline3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
+new_lambda=c(baseline3$lambda()["tau.s(ID)",1],baseline3$lambda()["nu.s(ID)",1],baseline3$lambda()[3:6,1])
+
+response3_exp$update_map(new_map)
+response3_exp$update_coeff_re(new_coeff_re)
+response3_exp$update_coeff_fe(new_coeff_fe)
+
+response3_exp$update_lambda(new_lambda)
+response3_exp$fit()
+
+# Fit response with 30m measurement error
+sigma_obs=0.03
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+response3_exp_30m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
+
+new_coeff_fe=c(baseline3_30m$coeff_fe()[1:3,1],0,baseline3_30m$coeff_fe()[4,1],0,baseline3_30m$coeff_fe()[5,1])
+new_coeff_re=c(baseline3_30m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_30m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
+               baseline3_30m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline3_30m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
+               baseline3_30m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
+new_lambda=c(baseline3_30m$lambda()["tau.s(ID)",1],baseline3_30m$lambda()["nu.s(ID)",1],baseline3_30m$lambda()[3:6,1])
+
+response3_exp_30m$update_map(new_map)
+response3_exp_30m$update_coeff_re(new_coeff_re)
+response3_exp_30m$update_coeff_fe(new_coeff_fe)
+
+response3_exp_30m$update_lambda(new_lambda)
+response3_exp_30m$fit()
+
+# Fit response with 40m measurement error
+sigma_obs=0.04
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+response3_exp_40m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
+
+new_coeff_fe=c(baseline3_40m$coeff_fe()[1:3,1],0,baseline3_40m$coeff_fe()[4,1],0,baseline3_40m$coeff_fe()[5,1])
+new_coeff_re=c(baseline3_40m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_40m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
+               baseline3_40m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline3_40m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
+               baseline3_40m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
+new_lambda=c(baseline3_40m$lambda()["tau.s(ID)",1],baseline3_40m$lambda()["nu.s(ID)",1],baseline3_40m$lambda()[3:6,1])
+
+response3_exp_40m$update_map(new_map)
+response3_exp_40m$update_coeff_re(new_coeff_re)
+response3_exp_40m$update_coeff_fe(new_coeff_fe)
+
+response3_exp_40m$update_lambda(new_lambda)
+response3_exp_40m$fit()
+
+# Fit response with 45m measurement error
+sigma_obs=0.045
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+response3_exp_45m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
+
+new_coeff_fe=c(baseline3_45m$coeff_fe()[1:3,1],0,baseline3_45m$coeff_fe()[4,1],0,baseline3_45m$coeff_fe()[5,1])
+new_coeff_re=c(baseline3_45m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_45m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
+               baseline3_45m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
+               baseline3_45m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
+               baseline3_45m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
+new_lambda=c(baseline3_45m$lambda()["tau.s(ID)",1],baseline3_45m$lambda()["nu.s(ID)",1],baseline3_45m$lambda()[3:6,1])
+
+response3_exp_45m$update_map(new_map)
+response3_exp_45m$update_coeff_re(new_coeff_re)
+response3_exp_45m$update_coeff_fe(new_coeff_fe)
+response3_exp_45m$update_lambda(new_lambda)
+
+response3_exp_45m$fit()
+
 # Fit response with 50m measurement error
@@
 response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline3_50m$lambda()["tau.s(ID)",1],baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
+
+response3_exp_50m$update_map(new_map)
 response3_exp_50m$update_coeff_re(new_coeff_re)
@@
 response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline3_75m$lambda()["tau.s(ID)",1],baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
-response3_exp_75m$update_coeff_re(new_coeff_re)
+
+response3_exp_75m$update_map(new_map)
 response3_exp_75m$update_coeff_fe(new_coeff_fe)
-
 response3_exp_75m$update_lambda(new_lambda)
+
 response3_exp_75m$fit()
@@
 response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
 
@@
 new_lambda=c(baseline3_100m$lambda()["tau.s(ID)",1],baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
+
+response3_exp_100m$update_map(new_map)
 response3_exp_100m$update_coeff_re(new_coeff_re)
 response3_exp_100m$update_coeff_fe(new_coeff_fe)
-
 response3_exp_100m$update_lambda(new_lambda)
+
 response3_exp_100m$fit()
@@
 ```{r}
+
+estimates_res3_exp=as.list(response3_exp$tmb_rep(),what="Est")
+std_res3_exp=as.list(response3_exp$tmb_rep(),what="Std")
+
+estimates_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Est")
+std_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Std")
+
+estimates_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Est")
+std_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Std")
+
+estimates_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Est")
+std_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Std")
+
 estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
@@
 
-plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
+plots_res3_exp=response3_exp$get_all_plots(baseline=baseline3,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res3_exp_30m=response3_exp_30m$get_all_plots(baseline=baseline3_30m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res3_exp_40m=response3_exp_40m$get_all_plots(baseline=baseline3_40m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,xmin=xmin,
+                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+```
+
+```{r display plots response3_sp}
+
+#print plots
+plots_res3_exp
+plots_res3_exp_50m
+plots_res3_exp_75m
+plots_res3_exp_100m
+
+```
+
+We select the response 3 model with 50m measurement error
+
+```{r}
+# Create directory
+if (!dir.exists("response3_exp_50m")) {
+    dir.create("response3_exp_50m", recursive = TRUE)
+}
+
+# Use lapply to save each plot
+lapply(seq_along(plots_bas2), function(i) {
+
+    plot<-plots_bas2[[i]]
+    plot_name <- names(plots_bas2)[i]
+    file_path <- file.path("response3_exp_50m", plot_name)
+
+    # Check if the plot is a ggplot or plotly object
+    if (inherits(plot, "ggplot")) {
+      ggsave(file_path, plot, device = "pdf")
+    } else if (inherits(plot, "plotly")) {
+      saveWidget(plot, file_path)
+  }
+    })
+```
+### Plots improvements
+
+```{r}
+
+#get angular velocity for different distances to shore
+plot_me_tau <- plots_res3_exp_50m$me_tau_ExpShip
+plot_me_nu <- plots_res3_exp_50m$me_nu_ExpShip
+
+# Update plot_me_tau
+plot_me_tau <- plot_me_tau +
+  labs(y = expression(tau)) +
+  theme_minimal() +
+  theme(
+    axis.title.x = element_text(size=18),
+    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=14),
+    axis.text.y = element_text(size=14),
+    legend.title = element_text(size=14),
+    legend.text = element_text(size=14)
+  )
+
+# Update plot_me_nu
+plot_me_nu <- plot_me_nu +
+  labs(y = expression(nu)) +
+  theme_minimal() +
+  theme(
+    axis.title.x = element_text(size=18),
+    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=14),
+    axis.text.y = element_text(size=14),
+    legend.title = element_text(size=14),
+    legend.text = element_text(size=14)
+  )
+
+# Save the updated plots
+ggsave("plot_me_tau_updated.pdf", plot=plot_me_tau, width=10, height=6, units="in")
+ggsave("plot_me_nu_updated.pdf", plot=plot_me_nu, width=10, height=6, units="in")
+
+```
+
+## Posterior samples
+```{r}
+post_coeff_res3_exp_50m=response3_exp_50m$post_coeff(n_post=10000)
+post_par_res3_exp_50m=list(
+              "tau0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"tau.(Intercept)"]),
+              "nu0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"nu.(Intercept)"]),
+              "omega0"=post_coeff_res3_exp_50m$coeff_fe[,"omega.(Intercept)"],
+              "sigma_tau"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,1])),
+              "sigma_nu"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,2])),
+              "tau.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"tau.ExpShip"],
+              "nu.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"nu.ExpShip"])
+mean_par_res3=lapply(post_par_res3_exp_50m,mean)
+sd_par_res3=lapply(post_par_res3_exp_50m,sd)
+quant_par_res3 <- lapply(post_par_res3_exp_50m, quantile, probs = c(0.025, 0.975))
+
+```
+
+#### Thorough uncertainty quantification for response3_exp_50m
+
+```{r}
+
+#posterior samples of the coefficient of the baseline model
+post_coeff_re=post_coeff_bas3_50m$coeff_re[1:100,]
+post_coeff_fe=post_coeff_bas3_50m$coeff_fe[1:100,]
+post_coeff_lambda=exp(post_coeff_bas3_50m$log_lambda[1:100,])
+
+# Fit response with 50m measurement error
+
+par0=c(0,0,1,4.5,0)
+formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
+                 nu=~ExpShip+s(ID,bs="re"),
+                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
+
+sigma_obs=0.05
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+
+estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+
+for (i in 1:nrow(post_coeff_fe)) {
+
+    response3_exp_50m_temp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
+                      response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
+                      map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
+                               log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+
+    new_coeff_fe=c(post_coeff_fe[i,1:3],0,post_coeff_fe[i,4],0,post_coeff_fe[i,5])
+    new_coeff_re=post_coeff_re[i,]
+    new_lambda=post_coeff_lambda[i,]
+    response3_exp_50m_temp$update_coeff_re(new_coeff_re)
+    response3_exp_50m_temp$update_coeff_fe(new_coeff_fe)
+
+    response3_exp_50m_temp$update_lambda(new_lambda)
+    response3_exp_50m_temp$fit()
+
+    est_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+    std_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+
+    estimates[i,]=est_alpha
+    std[i,]=std_alpha
+}
+
+colnames(estimates)=c("tau.ExpShip","nu.ExpShip")
+colnames(std)=c("tau.ExpShip","nu.ExpShip")
+
+```
+
+```{r}
+#The 9th value in estimates is inconsistent, we drop it
+estimates=estimates[-9,]
+std=std[-9,]
+
+mean_est=apply(X=estimates,MARGIN=2,FUN=mean)
+mean_std=apply(X=na.omit(std),MARGIN=2,FUN=mean)
+```

### 2024-11-08 R/application/CVM/fit_cvm.Rmd
reorganize folders

---
title: "fit_crcvm"
author: "Alexandre Delporte"
date: "2024-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

```{r packages}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
```

```{r get data}

# Set the path to the directory containing the data
par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#data before exposure with first 24h removed
dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

#data after exposure
dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"), header = TRUE,dec = ".")

```

We choose thresholds on the distance to the shore to define the ExpShore covariate.

```{r preprocess data}

D_low=0.07
D_up=3
dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low

dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low

dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low

N=length(unique(dataBE2$ID))

n_pre=length(dataBE2$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

## Baseline models

We can estimate the measurement error standard deviation directly from the data. Based on accuracy studies about fastloc GPS, it should be a few tens of metres. The estimate we get from the data before exposure is 25m (std error 1m), which seems consistent with previous studies. However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same. If we fix the measurement error to 25m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters. This cannot be solved by changing the other initial parameters in the optimisation. Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get. Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.

### Baselines with constant parameters

```{r fit baseline0}

#initial parameters
par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),omega=~1)

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_100m$fit()

```

```{r results baseline0}
estimates_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Est")
std_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Std")

estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")

estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")

estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")

```

### Baseline with AngleNormal covariate in omega

Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau. Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.

```{r fit baseline1}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1,1,1)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_25m$update_lambda(init_lambda)

baseline1_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_50m$update_lambda(init_lambda)

baseline1_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_75m$update_lambda(init_lambda)

baseline1_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_100m$update_lambda(init_lambda)

baseline1_100m$fit()

```

```{r results baseline1}
estimates_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Est")
std_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Std")

estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")

estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")

estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")

plots_bas1_25m=baseline1_25m$get_all_plots(baseline=NULL,model_name="baseline1_25m",show_CI="pointwise",save=TRUE)
plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)

```

### Baseline with AngleNormal and DistanceShore in omega

```{r fit baseline2}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_25m$update_lambda(init_lambda)
baseline2_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_50m$update_lambda(init_lambda)
baseline2_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_75m$update_lambda(init_lambda)
baseline2_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_100m$update_lambda(init_lambda)
baseline2_100m$fit()

```

```{r get results baseline2}

estimates_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Est")
std_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Std")

estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")

estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")

estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("DistanceShore"=0.07)
xmax=list("DistanceShore"=5)

plots_bas2_25m=baseline2_25m$get_all_plots(baseline=NULL,model_name="baseline2_25m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
```

```{r estimates on parameter scale baseline2}

post_coeff_bas2_25m=baseline2_25m$post_coeff(n_post=1000)
post_par_bas2_25m=list("tau0"=exp(post_coeff_bas2_25m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_25m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_25m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,2])))
mean_par_bas2_25m=lapply(post_par_bas2_25m,mean)
sd_par_bas2_25m=lapply(post_par_bas2_25m,sd)

post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)

post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)

#get 95% confidence intervals
post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
```

```{r smooth omega baseline2}

#get angular velocity for different distances to shore

plot1=plots_bas2_50m$fe_omega_AngleNormal_q1_DistanceShore
plot2=plots_bas2_50m$fe_omega_AngleNormal_q2_DistanceShore
plot4=plots_bas2_50m$fe_omega_AngleNormal_q4_DistanceShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Close"
omega2$Category <- "Medium"
omega4$Category <- "Far"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Close"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Far"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline2_50m",width = 10, height = 6, units = "in")

```

### Baseline with AngleNormal and ExpShore in omega

```{r fit baseline3}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1,1,0.01,1,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_25m$update_lambda(init_lambda)
baseline3_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_50m$update_lambda(init_lambda)
baseline3_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_75m$update_lambda(init_lambda)
baseline3_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_100m$update_lambda(init_lambda)
baseline3_100m$fit()
```
```{r get results baseline3}

estimates_bas3_25m=as.list(baseline3_25m$tmb_rep(),what="Est")
std_bas3_25m=as.list(baseline3_25m$tmb_rep(),what="Std")

estimates_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Est")
std_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Std")

estimates_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Est")
std_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Std")

estimates_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Est")
std_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("ExpShore"=1/D_up)
xmax=list("ExpShore"=1/D_low)
link=list("ExpShore"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore")

plots_bas3_25m=baseline3_25m$get_all_plots(baseline=NULL,model_name="baseline3_25m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale baseline3}

post_coeff_bas3_25m=baseline3_25m$post_coeff(n_post=1000)
post_par_bas3_25m=list("tau0"=exp(post_coeff_bas3_25m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_25m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_25m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_25m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_25m$log_lambda[,2])))
mean_par_bas3_25m=lapply(post_par_bas3_25m,mean)
sd_par_bas3_25m=lapply(post_par_bas3_25m,sd)

post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,2])))
mean_par_bas3_50m=lapply(post_par_bas3_50m,mean)
sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)

post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,2])))
mean_par_bas3_75m=lapply(post_par_bas3_75m,mean)
sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)

post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,2])))
mean_par_bas3_100m=lapply(post_par_bas3_100m,mean)
sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)

```

```{r smooth omega baseline3}

plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
plot2=plots_bas3_50m$fe_omega_AngleNormal_q2_ExpShore
plot4=plots_bas3_50m$fe_omega_AngleNormal_q4_ExpShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Far"
omega2$Category <- "Medium"
omega4$Category <- "Close"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Far"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Close"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")

```

## Response models with splines of ExpShip

### Response models based on baseline1

```{r fit response1}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_50m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_50m$lambda()["tau.s(ID)",1],1,baseline1_50m$lambda()["nu.s(ID)",1],
             baseline1_50m$lambda()["omega.s(AngleNormal)",1])
response1_sp_50m$update_coeff_re(new_coeff_re)
response1_sp_50m$update_coeff_fe(baseline1_50m$coeff_fe()[,1])

response1_sp_50m$update_lambda(new_lambda)

response1_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_75m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_75m$lambda()["tau.s(ID)",1],1,baseline1_75m$lambda()["nu.s(ID)",1],
             baseline1_75m$lambda()["omega.s(AngleNormal)",1])
response1_sp_75m$update_coeff_re(new_coeff_re)
response1_sp_75m$update_coeff_fe(baseline1_75m$coeff_fe()[,1])

response1_sp_75m$update_lambda(new_lambda)

response1_sp_75m$fit()

# Fit response with 50m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_100m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_100m$lambda()["tau.s(ID)",1],1,baseline1_100m$lambda()["nu.s(ID)",1],
             baseline1_100m$lambda()["omega.s(AngleNormal)",1])
response1_sp_100m$update_coeff_re(new_coeff_re)
response1_sp_100m$update_coeff_fe(baseline1_100m$coeff_fe()[,1])

response1_sp_100m$update_lambda(new_lambda)

response1_sp_100m$fit()
```

```{r get results response1}

estimates_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Est")
std_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Std")

estimates_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Est")
std_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Std")

estimates_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Est")
std_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_50m",show_CI="pointwise",save=TRUE)
plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_75m",show_CI="pointwise",save=TRUE)
plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
                            model_name="response1_sp_100m",show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale response1}

post_coeff_res1_sp_50m=response1_sp_50m$post_coeff(n_post=1000)
post_par_res1_sp_50m=list("tau0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,2])))
mean_par_res1_sp_50m=lapply(post_par_res1_sp_50m,mean)
sd_par_res1_sp_50m=lapply(post_par_res1_sp_50m,sd)

post_coeff_res1_sp_75m=response1_sp_75m$post_coeff(n_post=1000)
post_par_res1_sp_75m=list("tau0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,2])))
mean_par_res1_sp_75m=lapply(post_par_res1_sp_75m,mean)
sd_par_res1_sp_75m=lapply(post_par_res1_sp_75m,sd)

post_coeff_res1_sp_100m=response1_sp_100m$post_coeff(n_post=1000)
post_par_res1_sp_100m=list("tau0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,2])))
mean_par_res1_sp_100m=lapply(post_par_res1_sp_100m,mean)
sd_par_res1_sp_100m=lapply(post_par_res1_sp_100m,sd)

```

### Response model with splines of ExpShip based on baseline2

```{r fit response2}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
response2_sp_50m$update_coeff_re(new_coeff_re)
response2_sp_50m$update_coeff_fe(baseline2_50m$coeff_fe()[,1])

response2_sp_50m$update_lambda(new_lambda)

response2_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
response2_sp_75m$update_coeff_re(new_coeff_re)
response2_sp_75m$update_coeff_fe(baseline2_75m$coeff_fe()[,1])

response2_sp_75m$update_lambda(new_lambda)

response2_sp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
response2_sp_100m$update_coeff_re(new_coeff_re)
response2_sp_100m$update_coeff_fe(baseline2_100m$coeff_fe()[,1])

response2_sp_100m$update_lambda(new_lambda)

response2_sp_100m$fit()
```

```{r get results response2}
estimates_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
std_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")

estimates_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
std_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")

estimates_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
std_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_sp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_sp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_sp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
```

```{r}

post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)

post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)

post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)

```

### Response model with splines of ExpShip based on baseline3

```{r fit response3}

#initialize parameters
par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

#Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_50m$lambda()["tau.s(ID)",1],1,baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
response3_sp_50m$update_coeff_re(new_coeff_re)
response3_sp_50m$update_coeff_fe(baseline3_50m$coeff_fe()[,1])

response3_sp_50m$update_lambda(new_lambda)
response3_sp_50m$fit()

#Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_75m$lambda()["tau.s(ID)",1],1,baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
response3_sp_75m$update_coeff_re(new_coeff_re)
response3_sp_75m$update_coeff_fe(baseline3_75m$coeff_fe()[,1])

response3_sp_75m$update_lambda(new_lambda)
response3_sp_75m$fit()

#Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))

new_coeff_re=c(rep(0,2),baseline3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline3_100m$lambda()["tau.s(ID)",1],1,baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
response3_sp_100m$update_coeff_re(new_coeff_re)
response3_sp_100m$update_coeff_fe(baseline3_100m$coeff_fe()[,1])

response3_sp_100m$update_lambda(new_lambda)
response3_sp_100m$fit()
```

```{r get results response3}
estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")

estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")

estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_sp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_sp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_sp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

```{r estimates on parameter scale response3}

post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)

post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)

post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)

```

### Response model with exponential of ExpShip based on baseline2

```{r}

par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
response2_exp_50m$update_coeff_re(new_coeff_re)
response2_exp_50m$update_coeff_fe(new_coeff_fe)

response2_exp_50m$update_lambda(new_lambda)

response2_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
response2_exp_75m$update_coeff_re(new_coeff_re)
response2_exp_75m$update_coeff_fe(new_coeff_fe)

response2_exp_75m$update_lambda(new_lambda)

response2_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
response2_exp_100m$update_coeff_re(new_coeff_re)
response2_exp_100m$update_coeff_fe(new_coeff_fe)

response2_exp_100m$update_lambda(new_lambda)

response2_exp_100m$fit()
```

```{r}
estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")

estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")

estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res2_exp=response2_exp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_exp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_exp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_exp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

### Response model with exponential of ExpShip based on baseline3

```{r}

par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_50m$coeff_fe()[1:3,1],0,baseline3_50m$coeff_fe()[4,1],0,baseline3_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_50m$lambda()["tau.s(ID)",1],baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
response3_exp_50m$update_coeff_re(new_coeff_re)
response3_exp_50m$update_coeff_fe(new_coeff_fe)

response3_exp_50m$update_lambda(new_lambda)
response3_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_75m$coeff_fe()[1:3,1],0,baseline3_75m$coeff_fe()[4,1],0,baseline3_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_75m$lambda()["tau.s(ID)",1],baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
response3_exp_75m$update_coeff_re(new_coeff_re)
response3_exp_75m$update_coeff_fe(new_coeff_fe)

response3_exp_75m$update_lambda(new_lambda)
response3_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

new_coeff_fe=c(baseline3_100m$coeff_fe()[1:3,1],0,baseline3_100m$coeff_fe()[4,1],0,baseline3_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline3_100m$lambda()["tau.s(ID)",1],baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
response3_exp_100m$update_coeff_re(new_coeff_re)
response3_exp_100m$update_coeff_fe(new_coeff_fe)

response3_exp_100m$update_lambda(new_lambda)
response3_exp_100m$fit()
```

```{r}
estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")

estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")

estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_exp_50m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_exp_75m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_exp_100m",xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```

### 2024-11-08 R/preprocessing/narwhals/data_preprocessing.Rmd
reorganize folders

--- 2024-06-02 R/preprocessing/data_preprocessing.Rmd
@@
 
+#to get root of git repo
+library(here)
+
 #for tex code
@@
 
-plots_folder <-"plots"
+plots_folder <-"exploratory_plots"
 
@@
 ```{r read raw data,cache=TRUE}
-# Set the path to the directory containing the data
-par_dir=dirname(dirname(getwd())) #parent directory 
-narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")
+# Set the path to the directory containing the raw data for narwhals
+raw_narwhal_data_path <- here(file.path("Data","raw_data","narwhals"))
 
@@
 data_file="db_narwhal_2017_2018_with_gps.txt"
-rawData <- read.table(file.path(narwhal_data_path,data_file),
+rawData <- read.table(file.path(raw_narwhal_data_path,data_file),
                       header = TRUE, sep = "\t",dec = ".")
@@
 The main differences are that only small guns were used in 2017 while only big guns were used in 2018, and the tagged whales used a much larger part of Scoresby Sound in 2018.
+More interesting results were found in 2018 in the paper of Mads-Peter, Adeline and Susanne.
+We primarily focus on this data rather than the 2017 data.
 
@@
 \begin{itemize}
-\item keeping only 2018 data - adding a time since tagging column 
-\item converting distance to ship and distance to shore to km - adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.
-\item Renaming the "Ind" column to "ID" and converting it to factors.
-\item Converting column "Type" in factor with levels trial and intertrial 
+\item keeping only 2018 data and adding a time since tagging column 
+\item converting distance to ship and distance to shore to km and adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.\texttt{ExpShip} is a proxy for sound exposure level (SEL).
+\item Renaming the "Ind" column to "ID" for clarity and converting it to factors.
+\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to shio during trial and intertrial separately
 \item Converting column "Posixct" to POSIXct format for date and time
@@
 The first measurement time is set to $1$ second and then measures are interpolated every second (see Jorgensen et al.).
-Thus, the time $T_k$ of the $k$-th measurement is $$T_{k}=\frac{k}{3600}\mbox{ h }$$ As proposed in Jorgensen et al., we also add a column with a covariate $E$ defined as $$E(t)=\frac{1}{\mbox{distance to ship}(t)}$$ where distance to ship has been converted in km.
-
-\texttt{ExpShip} is a proxy for sound exposure level (SEL).
+Thus, the time $T_k$ of the $k$-th measurement is $$T_{k}=\frac{k}{3600}\mbox{ h }$.
+
 ```
@@
 data=rawData[rawData$Year==2018,]
-
-#convert Distance Shore and Dist_ship to km
-data$Dist_Ship=data$Dist_Ship/1000
-data$DistanceShore=data$DistanceShore/1000
 
@@
 
+#convert Distance Shore and Dist_ship to km
+data$Dist_Ship=data$Dist_Ship/1000
+data$DistanceShore=data$DistanceShore/1000
+
 ## Defining exposures
@@
 
+data$ExpShipI=data$ExpShip
+data$ExpShipT=data$ExpShip
+
+data$ExpShipI[data$Type=="trial"]=0
+data$ExpShipT[data$Type=="intertrial"]=0
+
 #convert Posixct to POSIXct class for time and dates
@@
 
-# Data exploration after first preprocessing
+# First data exploration
 
@@
 #plot with all the narwhals rescaled so that the experiments start at the same time 
-trials1=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
+trials_rescaled=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
   xlab("Number of days after tagging")+ylab("Exposure")+
@@
 #plot with all the narwhals at the real experiments time
-trials2=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
+trials=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
   xlab("Day")+ylab("Exposure")+
@@
 
-ggsave(path=plots_folder,filename="trials_rescaled.png",plot=trials1,width=10,height=5)
-ggsave(path=plots_folder,filename="trials.png",plot=trials2,width=10,height=5)
-```
-
-## Ship exposure through time and depending on trial type
-
-```{r define ExpShipI and ExpShipT}
-
-data$ExpShipI=data$ExpShip
-data$ExpShipT=data$ExpShip
-
-data$ExpShipI[data$Type=="trial"]=0
-data$ExpShipT[data$Type=="intertrial"]=0
-
-```
-
-It appears that the exposure is not so smooth with respect to time, meaning that a narwhal can be highly exposed at some time and be completely out of sight from the airguns a few seconds later.
+ggsave(path=plots_folder,filename="trials_rescaled.png",plot=trials_rescaled,width=10,height=5)
+ggsave(path=plots_folder,filename="trials.png",plot=trials,width=10,height=5)
+```
+
+The exposure is not smooth with respect to time, because a narwhal can be highly exposed at some time and be completely out of sight from the airguns a few seconds later, and also because the airguns shooting stops abruptly.
 This phenomenon appears when the values in DistShip suddenly change to NA.
@@
 Moreover, exposure times are relatively short compare to the overall duration of the measurement period, so that most of the time, ExpShip is zero but short peaks appear when trials or intertrials are happening.
-We would need to zoom on these peaks to assess their smoothness.
+We can zoom on these peaks for a better visualization.
 
@@
 
-#plot exposure durinh intertrial
+#plot exposure during intertrial
 pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
@@
 
+# Second preprocesssing step
+
+```{=tex}
+\textbf{\texttt{Lat} and \texttt{Long} columns have been interpolated every second, and distance to ship has been computed from these interpolations.The true measured distance to ship are in the rows for which the column \texttt{pos\_org}  has value 1.}
+We will drop the interpolated positions for the analysis.
+```
+
+```{=tex}
+The next step consists in :
+\begin{itemize}
+\item  Dropping missing latitude and longitude.
+\item Keeping only GPS positions indicated by value 1 in \texttt{pos\_org} column.
+\item  The narwhal positions are in latitude and longitude.
+To analyze the track and visualize it better we project it on a plane in UTM coordinastes.
+Hence, we project to UTM zone north 26 (https://epsg.io/32626).
+\item We add a column "speed" with empirical horizontal speed obtained from the GPS positions (not interpolated)
+\end{itemize}
+
+```
+
+```{r preprocess data 2}
+
+#remove rows with missing positions
+dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]
+
+#keep rows indicated as GPS measurements
+dataprep=dataprep[dataprep$pos_org==1,]
+
+#number of rows
+nrows=length(dataprep[,1])
+
+#project to UTM (Universal Transverse Mercator coordinate system)
+
+llcoord=SpatialPoints(dataprep[,c("Long","Lat")],
+                      proj4string=CRS("+proj=longlat +datum=WGS84"))
+utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84 +units=km"))
+
+#add UTM locations in km to data frame
+dataprep$x=attr(utmcoord,"coords")[,1]
+dataprep$y=attr(utmcoord,"coords")[,2]
+
+#compute empirical horizontal speed
+dataprep$horizontal_speed=rep(NA,nrows)
+
+rows_index=c()
+for (id in unique(dataprep$ID)){
+
+  #filter data with specific id
+  sub_ind=(dataprep$ID==id)
+  sub_data=dataprep[sub_ind,]
+  n_sub=length(sub_data$time)
+
+  #time steps
+  dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]
+
+  #step lengths
+  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])
+  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])
+
+  speed=c(NA,sqrt((dx/dtimes)^2+(dy/dtimes)^2))
+
+  dataprep[sub_ind,"horizontal_speed"]=speed
+}
+
+```
+
+# Data exploration after second preprocessing
+
 ## True exposure per individual
@@
 ```{=tex}
-\textbf{\texttt{Lat} and \texttt{Long} columns have been interpolated every second, and distance to ship has been computed from these interpolations.
-The true measured distance to ship are in the rows for which the column \texttt{pos\_org}  has value 1.}
+We plot exposure from interpolated positions and highlight points that are directly observed exposures, that is exposure levels computed from the actual GPS position of the narhwal compared to the GPS position of the ship.
+
 ```
@@
 
-# Second preprocesssing step
-
-```{=tex}
-\begin{itemize}
-\item  Drop missing latitude and longitude
-\item Keep GPS positions indicated by value 1 in \texttt{pos\_org} column
-\item  Project to UTM zone 26. The narwhal positions are in latitude and longitude.To analyze the track and visualize it better we project it on a plane in UTM coordinastes.
-The positions of the narwhals are in the UTM zone 26 north (in Greenland).
-\item Add a column "speed" with empirical horizontal speed obtained from the GPS positions (not interpolated)
-\end{itemize}
-
-```
-
-```{r preprocess data 2}
-
-#remove rows with unknown positions
-dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]
-
-#keep rows indicated as GPS measurements
-dataprep=dataprep[dataprep$pos_org==1,]
-
-#number of rows
-nrows=length(dataprep[,1])
-
-#project to UTM (Universal Transverse Mercator coordinate system)
-
-llcoord=SpatialPoints(dataprep[,c("Long","Lat")],
-                      proj4string=CRS("+proj=longlat +datum=WGS84"))
-utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84 +units=km"))
-
-#add UTM locations in km to data frame
-dataprep$x=attr(utmcoord,"coords")[,1]
-dataprep$y=attr(utmcoord,"coords")[,2]
-
-#compute empirical horizontal speed
-n=length(dataprep$time)
-dataprep$horizontal_speed=rep(NA,n)
-
-rows_index=c()
-for (id in unique(dataprep$ID)){
-
-  #filter data with specific id
-  sub_ind=(dataprep$ID==id)
-  sub_data=dataprep[sub_ind,]
-  n_sub=length(sub_data$time)
-
-  #time steps
-  dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]
-
-  #step lengths
-  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])
-  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])
-
-  speed=c(NA,sqrt((dx/dtimes)^2+(dy/dtimes)^2))
-
-  dataprep[sub_ind,"horizontal_speed"]=speed
-}
-
-```
-
-# Data exploration after second preprocessing
-
-## measurement periods for each narwhal
+## Measurement periods for each narwhal
+
 ```{r measurement periods}
@@
 the two points.
-But this is obviously not true, especially for long time steps, because the narwhal make U-turns for instance, and more importantly, it dives to find food.
+But this is obviously not true, especially for long time steps, because the narwhal make U-turns for instance.
 Therefore the empirical horizontal speed always underestimates the true horizontal speed.
-Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram
+Anormal values of the speed may indicate outliers to remove from the data.
+While there is no reason to drop low speed values, since the narwhals can well rest for some time, high speed values that are clearly outliers and could not have been reached by a narwhal might be removed from the data.
+Here we clearly see some outliers in the histogram.
 ```
@@
 
+```{=tex}
+Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram.
+We thus decide to filter by setting a threshold of $20$ km/h for horizontal speed.
+```
+
 ```{r speed higher than 20km/h }
@@
 
-# Time steps distribution
+```{r histogram filtrered horizontal speed,warnings=FALSE}
+speed_histo_filtered=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
+  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+  scale_color_viridis_d(guide="none")+
+  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")
+
+speed_histo_filtered
+ggsave(path=plots_folder,"speed_histo_filtered.png", plot = speed_histo_filtered)
+```
+## Time steps distribution
+
+```{=tex}
+Now that we removed the interpolated positions, we can check the time step distribution to get the frequence of acquisition of the narwhals Fastloc GPS device.
+```
 
@@
 
-# Time steps distribution during exposure vs no exposure
-
-# Plot all GPS tracks (without interpolation) on a map
+## Time steps distribution during exposure vs no exposure
+
+## Plot all GPS tracks (without interpolation) on a map
 
@@
 
-```{r save points as geojson in data folder}
+```{r save points as shapefile in data folder}
+# Set the path to the directory containing the preprocessed data
+preprocessed_narwhal_data_path <- here(file.path("Data","preprocessed_data","narwhals"))
+
 points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
 st_crs(points_sf)="+init=EPSG:32626 +units=km"
-st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
-```
-
-# Coastline and land data
+st_write(points_sf,file.path(preprocessed_narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
+```
+
+# Land polygons data for the domain boundary
 
@@
 
-```{r extract coastline and land}
+We read the preprocessed land polygons extracted from OpenStreetMap database.
+
+```{r extract land polygons}
 
 # Set the path to the directory containing coastline data
-par_dir=dirname(dirname(getwd())) #parent directory 
-greenland_data_path <- file.path(par_dir,"Data", "Greenland")
+preprocessed_greenland_data_path <- here(file.path("Data","preprocessed_data", "greenland"))
 
 #get land polygons
-land_utm<-st_read(file.path(greenland_data_path,"land_utm.shp"))
-
-```
-
-## Plot coastline and land
+scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"scoresby_sound_utm.shp"))
+
+```
+
+## Plot scoresby sound land
 
 ```{r plot coastline and land}
-plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill="grey",col="brown")+
+plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")+
   coord_sf(datum=st_crs(32626))
@@
 
-## Compute nearest point on land and coastline for each GPS position
+## Compute nearest point on land for each GPS position
 
@@
 
-nearest_points=function(positions,coastline) {
+nearest_points=function(positions,border) {
   #positions: dataframe of positions in UTM coordinates in metres
-  #coastline : geometry of the coastline projected in the same UTM zone as positions in metres
+  #border : geometry of the coastline projected in the same UTM zone as positions in metres
 
@@
 
-    # Iterate over each polygon in coastline
-    for (j in 1:length(coastline$geometry)) {
+    # Iterate over each polygon in border
+    for (j in 1:length(border$geometry)) {
 
       # Get the current polygon
-      polygon <- coastline$geometry[[j]]
+      polygon <- border$geometry[[j]]
 
@@
 
-#nearest points in coastline and land
+#nearest points in border and land
 
@@
 
-res_land=nearest_points(positions=narwhal_positions,coastline=land_utm)
+res_land=nearest_points(positions=narwhal_positions,border=scoresby_sound_utm)
 })
@@
 
-## Compare computed distances to shore
+## Compare distances to shore from original data and computed distances to shore
 
@@
 
-```{=tex}
-
-For the rest of the analysis, we'll keep the bounded domain defined by the polygons in \texttt{modified\_land\_utm} sf object as a reference.
-Doing that involves having some points on land, that is points with Distance to shore equal to 0.
-This should be essentially due to the error in the land polygons than to the error in GPS measurements.
-```
-
 ## Points on land
-
-For coastline, distance is 0 only if the position of the narwhal is exactly on the coastline or on a small island.
-A proper "land domain" is not defined in this case, but only the boundary of the domain (the coastline).
 
@@
 
-cat("Percentage of points on land with distance from land : \n",
+cat("Percentage of points on land with distance from osm land polygons : \n",
     length(which(res_land$min_distances==0))/length(dataprep$time)*100,"\n")
@@
 ```{r check original points on land}
-check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
+plot_original_inland_pts=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
@@
 
-ggplotly(check_plot1)
-```
-
-```{r inspect points on land}
+ggplotly(plot_original_inland_pts)
+```
+
+```{r inspect depth of points on land}
 dataprep[dataprep$DistanceShore==0,c("Depth","time")]
@@
 
-```{r check new points on land}
+```{r check osm points on land}
 df=dataprep
@@
 
-check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
+plot_osm_inland_pts=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
@@
 
-ggplotly(check_plot2)
-```
-
-```{r percentage points on land per ID}
+ggplotly(plot_osm_inland_pts)
+```
+
+```{r percentage points on land per ID with osm map}
 
@@
 
+## Use land polygons updated with QGIS
+
+```{=tex}
+We need to know the nearest point on land to compute how the narwhal moves with respect to the land.
+For this, we need use the land polygons.
+In the original data, we have values for the Distance to shore, but we don't know exactly how it was computed.
+Approximately $2 \%$ of the narwhals positions are inland.
+This may be due to GPS measurement error, or errors in distance to shore computations.
+We notice that some points that have non zero distance to shore in the original data appear inland on the osm map.
+This suggests the osm map is not accurate.
+Indeed, as we have just seen, if we compute the distance to shore based on the osm polygon data, $9\%$ of the narwhal positions are inland.
+Thus, we decided to manually inspect the polygon data from OpenStreetMap with qgis software.
+We overlayed Sentinel-2 satellite images with the osm polygons data and the narwhals positions, and modified the shoreline to better match the original distance to shore and the satellite image.
+The resulting shapefile is updated_scoresby_sound_utm.shp.
+```
+
+```{r extract updated land polygons}
+
+#get updated land polygons
+updated_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"updated_scoresby_sound_utm.shp"))
+
+updated_scoresby_sound_utm <- st_transform(updated_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
+
+```
+
+```{r compute updated distance to shore}
+res_updated_land=nearest_points(positions=narwhal_positions,border=updated_scoresby_sound_utm)
+
+cat("Percentage of points on land with distance from updated osm land polygons : \n",
+    length(which(res_updated_land$min_distances==0))/length(dataprep$time)*100,"\n")
+
+```
+
+```{r check updated osm points on land}
+df=dataprep
+df$DistanceShore=res_updated_land$min_distances
+
+plot_updated_osm_inland_pts=ggplot()+geom_sf(data=updated_scoresby_sound_utm$geometry,fill="grey")+
+    coord_sf(datum=st_crs(32626))+
+  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
+    scale_color_viridis_d()
+
+ggplotly(plot_updated_osm_inland_pts)
+```
+
 ## Points too close to land
@@
 
-df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<1 & res_land$min_distances>0  ])
+df=data.frame(DistanceShore=res_updated_land$min_distances[res_updated_land$min_distances<1 & res_updated_land$min_distances>0  ])
 histo=ggplot(df, aes(x =DistanceShore)) +
@@
 
-This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
-the shore in reality.
+```{=tex}
+
+For the rest of the analysis, we'll keep the bounded domain defined by the updated polygons in "updated_scoresby_sound_utm" sf object as a reference.
+Doing that involves having some points on land, that is points with Distance to shore equal to 0.
+```
 
@@
 #keep points calculated from land polygons
-nearest_points=res_land$nearest_points
-min_distances=res_land$min_distances
+nearest_points=res_updated_land$nearest_points
+min_distances=res_updated_land$min_distances
 
@@
 ```{r}
-#add ExpShore covariate based on 1/distance in km
-dataprep$ExpShore <- ExposureFunction(X = dataprep$DistanceShore)
+#add Ishore covariate based on 1/distance in km
+dataprep$Ishore <- ExposureFunction(X = dataprep$DistanceShore)
 
@@
 
-```{=tex}
-It turns out that the effect was due to points on land. 
-We can now observe a slight decrease in the angle of deviation from the coastline, which may indicate that narwhals tend to move parallel to the coast when close enough.
-However, no effects of Distance to shore is observed on the empirical deviation between consecutive velocities.
-This may be due to the averaging that is made to get those plots. The narwhal only turns for a few positions and then keep its heading paralell to the shore.
-```
-
 # Visualize individual tracks
@@
   sub_data=dataprep[dataprep$ID==id,]
-  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
+  id_track=ggplot()+geom_sf(data=updated_scoresby_sound_utm$geometry,fill="grey")+
     coord_sf(datum=st_crs(32626))+
@@
 ```
-# Further details
-
-## Plot ship exposure without interpolated points
-
-```{r plot real exposure}
-pExpShip=ggplot(dataprep,aes(time,ExpShip,col=ID))+geom_line(color="black")+
-  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")
-
-ggsave(path=plots_folder,"realExpShip_through_time.png", plot=pExpShip)
-```
-
-```{r display pExp, fig.cap="Measured exposure to ship through time"}
-pExpShip
-```
-## Plot shore exposure
-
-```{r}
-df=dataprep
-df[df$DistanceShore<D_low,"ExpShore"]=1/D_low
-df[df$DistanceShore>D_up,"ExpShore"]=0
-
-pExpShore=ggplot(df,aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")
-
-ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExpShore)
-```
-
-```{r}
-pExpShore
-```
 
@@
   maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
-  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > maxtime) & (time>tagtime)))
+  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
 }
@@
   maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
-  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > maxtime) & (time>tagtime/2)))
+  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
 }
@@
   maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
-  dataAE =rbind(dataAE,subset(narwhalData, (time > maxtime) & (time>tagtime)))
+  dataAE =rbind(dataAE,subset(narwhalData, (time >= maxtime) & (time>tagtime)))
 }
@@
 ```{r save preprocessed data}
-
-write.csv(dataprep,file.path(narwhal_data_path,"allData.csv"))
-write.csv(dataBE1,file.path(narwhal_data_path,"DataBE1.csv"))
-write.csv(dataBE2,file.path(narwhal_data_path,"DataBE2.csv"))
-write.csv(dataAE,file.path(narwhal_data_path,"DataAE.csv"))
-```
+# Set the path to the directory containing the raw data for narwhals
+preprocessed_narwhal_data_path <- here(file.path("Data","preprocessed_data","narwhals"))
+
+write.csv(dataprep,file.path(preprocessed_narwhal_data_path,"allData.csv"))
+write.csv(dataBE1,file.path(preprocessed_narwhal_data_path,"DataBE1.csv"))
+write.csv(dataBE2,file.path(preprocessed_narwhal_data_path,"DataBE2.csv"))
+write.csv(dataAE,file.path(preprocessed_narwhal_data_path,"DataAE.csv"))
+```
+
+## Plot ship exposure 12h after tagging
+
+```{r plot real exposure}
+
+cropped_data=dataprep[dataprep$time>12,]
+pExpShip=ggplot(cropped_data,aes(time,ExpShip,col=ID))+geom_line(color="black")+
+  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")
+
+ggsave(path=plots_folder,"realExpShip_through_time.png", plot=pExpShip)
+```
+
+```{r display pExp, fig.cap="Measured exposure to ship through time"}
+pExpShip
+```
+## Plot shore exposure
+
+```{r}
+D_low=0.07
+D_up=3
+df=cropped_data
+df[df$DistanceShore<D_low,"Ishore"]=1/D_low
+df[df$DistanceShore>D_up,"Ishore"]=0
+
+pIshore=ggplot(df,aes(time,Ishore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+  theme(legend.position="none")+xlab("Time (h)")+ylab("Inverse distance to shore")
+
+ggsave(path=plots_folder,"realIshore_through_time.png", plot=pIshore)
+```
+
+```{r}
+pIshore
+```

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore118.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 118
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_DistanceShore63.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 63
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore114.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 114
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore152.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 152
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore179.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 179
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore82.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 82
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/circ/circ_ExpShore93.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="circ"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 93
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore108.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 108
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore114.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 114
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore166.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 166
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_DistanceShore87.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 87
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore162.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 162
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore181.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 181
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore188.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 188
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore22.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 22
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore25.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 25
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/fjords_ExpShore50.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="fjords"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="fjords"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 50
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/fjords/set_up_fjords.R

@@ -1,0 +1,190 @@
+# HEADER ----------------------------------
+#
+# Author:     Alexandre Delporte
+# Copyright     Copyright 2024 - Alexandre Delporte
+# Email:      alexandre.delporte@univ-grenoble-alpes.fr
+#
+# Date:     2024-05-03
+#
+# Script Name:    set_up_fjords.R
+#
+# Script Description: Set up of the simulation study for the CRCVM within 
+# a Scoresby Sound fjords system
+#
+#
+# SETUP ------------------------------------
+
+set.seed(42)                #seed for reproductibility
+
+
+library(ggplot2)            #plots
+library(plotly)             #html plots
+library(sf)                 #geometry for constraints
+library(mgcv)               #tensor splines
+library(smoothSDE)          #to compute nearest shore point
+
+
+# READ HYPERPARAMETERS FILE
+
+# Read hyperparameters from a user specified file
+hyperparams_file <- "hyperparams_file.txt"  # Update this with a bash variable
+path=here("R","simulation_study_CRCVM","spline_estimation")
+
+# Read the file and evaluate each line
+lines <- readLines(file.path(path,hyperparams_file))
+for (line in lines) {
+  eval(parse(text = line))
+}
+
+# Land polygons --------------------
+
+
+# Set the path to the directory containing the data
+greenland_data_path <- file.path("/home","delporta","Documents","Research","Projects","narwhals_smoothSDE","Data","Greenland")
+
+#get the land and coastline geometries from the geojson file
+border<-st_read(file.path(greenland_data_path,"land_utm.shp"))
+st_crs(border)="+init=EPSG:32626"
+border<-st_make_valid(border)
+
+v0=c(0,0)
+
+#generate random initial points
+x0=matrix(rep(NA,N_ID*2),ncol=2)
+colnames(x0)=c("x1","x2")
+
+i=1
+while (i<=N_ID) {
+  #choose location uniformly in the map
+  x=c(runif(1,min=430,max=500),runif(1,min=7760,max=7900))
+  p=nearest_shore_point(st_point(x),border)
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore160.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 160
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore53.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 53
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore62.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 62
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_DistanceShore72.R

@@ -3,14 +3,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 72
 set.seed(seed)
@@ -36,7 +40,7 @@ data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv
   ftau_constant=function(cov_data) {
     return (exp(true_log_tau[i]))
   }
-  res=sim_theta_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_splines,fnu=fnu_constant,
                       log_sigma_obs=log(SIGMA_OBS),v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
   
   data_sim=res$sim
@@ -151,7 +155,7 @@ add_covs=function(data) {
 
 data=add_covs(data)
 
-#only estimate if observed Distances to shore include the interval [0.5,5]
+#only estimate if observed Distances to shore include the interval [D_low,D_up]
 
 if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
[…]

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore136.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 136
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore140.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 140
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore146.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 146
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore15.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 15
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore169.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 169
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore177.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 177
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore25.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 25
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore31.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 31
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2024-11-08 R/simulation_study_CRCVM/spline_estimation/rect/rect_ExpShore90.R

@@ -2,14 +2,18 @@
 cat("\014")                 # Clears the console
 rm(list = ls())             # Remove all variables of the work space
 
-domain_name="rect"
-par_dir=dirname(getwd())
-set_up_file=paste("set_up_",domain_name,".R",sep="")
-source(file.path(par_dir,set_up_file))     #get set up for simulation study
-source(file.path(dirname(par_dir),"CVM_functions.R"))  #get functions to simulate trajectories
-library(smoothSDE)          #sde models
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
 library(foreach)            #foreach loop
 library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
 
 seed= 90
 set.seed(seed)
@@ -153,7 +157,7 @@ data=add_covs(data)
 
 #only estimate if observed Distances to shore include the interval [D_low,D_up]
 
-if (min(data$DistanceShore)>D_low | max(data$DistanceShore)<D_up) {
+if (min(data$DistanceShore)>D_low+0.2 | max(data$DistanceShore)<D_up-0.2) {
   stop("Observed distance to shore are irrelevant with the fixed knots")
 }
 
@@ -196,6 +200,5 @@ true_df=data.frame("true"=c(tau_re,nu_re,sp_coeff_ExpShore[2:(SP_DF[1]*SP_DF[2])
 
 coeffs_df=cbind(coeffs_df,true_df)
 
-write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_ExpShore",seed,".csv",sep=""), row.names=FALSE)
-
-
+write.csv(coeffs_df,paste("result_",domain_name,"_",TMAX,"h_",N_ID,"ID_",DMIN,"km_",
+                          SP_DF[1],"-",SP_DF[2],"spdf_ExpShore",seed,".csv",sep=""), row.names=FALSE)

### 2025-01-07 R/application/CRCVM/fit_crcvm.Rmd
R code updates

--- 2024-06-17 R/application/CRCVM/fit_crcvm.Rmd
@@
 author: "Alexandre Delporte"
-date: "2024-06-07"
+date: "2024-12-12"
 output: html_document
 ---
+To circumvent this issue, we can fix a smooth parametric function for $\omega$, based on the observed distances to shore and the observed angular velocities.
+
+# Set up
 
@@
 
-## Set up
-
-```{r packages}
+```{r packages,message=FALSE}
 set.seed(42)
@@
 library(mgcv)
+library(here)
+library(xtable)
+library(sf)
+library(gganimate) # animated trajectories
 ```
 
-```{r get data}
+```{r get data,message=FALSE}
 
 # Set the path to the directory containing the data
-par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
-narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")
+par_dir=here() #parent directory 
+narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")
 
 #data before exposure with first 24h removed
-dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")
+dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")
 
 #data before exposure with first 12h removed
-dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")
+dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")
 
@@
 
-We choose thresholds on the distance to the shore to define the ExpShore covariate.
+```{r}
 
-```{r preprocess data}
+N=length(unique(dataBE12$ID))
 
-D_low=0.07
-D_up=3
-dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
-dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
-dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
-dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-N=length(unique(dataBE2$ID))
-
-n_pre=length(dataBE2$time) #number of observations before exposure
+n_pre12=length(dataBE12$time) #number of observations before exposure
+n_pre24=length(dataBE24$time) #number of observations before exposure
 n_post=length(dataAE$time) #number of observations after exposure
-```
-
-## Baseline models
-
-We can estimate the measurement error standard deviation directly from the data.
-Based on accuracy studies about fastloc GPS, it should be a few tens of metres.
-The estimate we get from the data before exposure is 35m (std error 1m), which seems consistent with previous studies.
-However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same.
-If we fix the measurement error to 35m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters.
-This cannot be solved by changing the other initial parameters in the optimisation.
-Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get.
-Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.
-
-### Baselines with constant parameters
-
-```{r fit baseline0}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1,omega=~1)
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_35m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_100m$fit()
-
-```
-
-```{r results baseline0}
-estimates_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Est")
-std_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Std")
-
-estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
-std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")
-
-estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
-std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")
-
-estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
-std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")
-
-```
-
-### Baseline with AngleNormal covariate in omega
-
-Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau.
-Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.
-
-```{r fit baseline1}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1/0.2^2,1/0.2^2,0.01)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,
-                 tau =~s(ID,bs="re"),
-                 nu=~s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=5,bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_35m$update_lambda(init_lambda)
-
-baseline1_35m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_50m$update_lambda(init_lambda)
-
-baseline1_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_75m$update_lambda(init_lambda)
-
-baseline1_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_100m$update_lambda(init_lambda)
-
-baseline1_100m$fit()
-
-```
-
-```{r results baseline1}
-estimates_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Est")
-std_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Std")
-
-estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
-std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")
-
-estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
-std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")
-
-estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
-std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")
-
-plots_bas1_35m=baseline1_35m$get_all_plots(baseline=NULL,model_name="baseline1_35m",show_CI="pointwise",save=TRUE)
-plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
-plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
-plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)
-
-```
-
-### Baseline with AngleNormal and DistanceShore in omega
-
-```{r fit baseline2}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)
-
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_35m$update_lambda(init_lambda)
-baseline2_35m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_50m$update_lambda(init_lambda)
-baseline2_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_75m$update_lambda(init_lambda)
-baseline2_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_100m$update_lambda(init_lambda)
-baseline2_100m$fit()
-
-```
-
-```{r get results baseline2}
-
-estimates_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Est")
-std_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Std")
-
-estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
-std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")
-
-estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
-std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")
-
-estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
-std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")
-
-#plot parameters
-
-xmin=list("DistanceShore"=0.07)
-xmax=list("DistanceShore"=5)
-
-plots_bas2_35m=baseline2_35m$get_all_plots(baseline=NULL,model_name="baseline2_35m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-```
-
-```{r estimates on parameter scale baseline2}
-
-post_coeff_bas2_35m=baseline2_35m$post_coeff(n_post=1000)
-post_par_bas2_35m=list("tau0"=exp(post_coeff_bas2_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,2])))
-mean_par_bas2_35m=lapply(post_par_bas2_35m,mean)
-sd_par_bas2_35m=lapply(post_par_bas2_35m,sd)
-
-post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
-post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
-mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
-sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)
-
-post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
-post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
-mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
-sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)
-
-#get 95% confidence intervals
-post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
-post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
-mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
-sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
-```
-
-```{r smooth omega baseline2}
-
-#get angular velocity for different distances to shore
-
-plot1=plots_bas2_50m$fe_omega_AngleNormal_q1_DistanceShore
-plot2=plots_bas2_50m$fe_omega_AngleNormal_q2_DistanceShore
-plot4=plots_bas2_50m$fe_omega_AngleNormal_q4_DistanceShore
-
-omega1 <- ggplot_build(plot1)$data[[1]]
-omega2 <- ggplot_build(plot2)$data[[1]]
-omega4 <- ggplot_build(plot4)$data[[1]]
-
-omega1$Category <- "Close"
-omega2$Category <- "Medium"
-omega4$Category <- "Far"
-
-# Combine the data frames
-combined_omega <- rbind(omega1, omega2, omega4)
-
-ci_omega1 <- ggplot_build(plot1)$data[[2]]
-ci_omega2 <- ggplot_build(plot2)$data[[2]]
-ci_omega4 <- ggplot_build(plot4)$data[[2]]
-
-ci_omega1$Category <- "Close"
-ci_omega2$Category <- "Medium"
-ci_omega4$Category <- "Far"
-
-# Combine the data frames
-combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)
-
-combined_plot <- ggplot() +
-  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
-  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
-  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
-  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
-  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
-  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
-  theme_minimal()
-
-ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline2_50m",width = 10, height = 6, units = "in")
-
-```
-
-### Baseline with AngleNormal and ExpShore in omega
-
-```{r fit baseline3}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1,1,0.01,1,1,1)
-
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 35m measurement error
-sigma_obs=0.035
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_35m$update_lambda(init_lambda)
-baseline3_35m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_50m$update_lambda(init_lambda)
-baseline3_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_75m$update_lambda(init_lambda)
-baseline3_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_100m$update_lambda(init_lambda)
-baseline3_100m$fit()
-```
-```{r get results baseline3}
-
-estimates_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Est")
-std_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Std")
-
-estimates_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Est")
-std_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Std")
-
-estimates_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Est")
-std_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Std")
-
-estimates_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Est")
-std_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Std")
-
-#plot parameters
-
-xmin=list("ExpShore"=1/D_up)
-xmax=list("ExpShore"=1/D_low)
-link=list("ExpShore"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore")
-
-plots_bas3_35m=baseline3_35m$get_all_plots(baseline=NULL,model_name="baseline3_35m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
-
-```{r estimates on parameter scale baseline3}
-
-post_coeff_bas3_35m=baseline3_35m$post_coeff(n_post=1000)
-post_par_bas3_35m=list("tau0"=exp(post_coeff_bas3_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,2])))
-mean_par_bas3_35m=lapply(post_par_bas3_35m,mean)
-sd_par_bas3_35m=lapply(post_par_bas3_35m,sd)
-
-post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
-post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,2])))
-mean_par_bas3_50m=lapply(post_par_bas3_50m,mean)
-sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)
-
-post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
-post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,2])))
-mean_par_bas3_75m=lapply(post_par_bas3_75m,mean)
-sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)
-
-post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
-post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,2])))
-mean_par_bas3_100m=lapply(post_par_bas3_100m,mean)
-sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)
-
-```
-
-```{r smooth omega baseline3}
-
-plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
-plot2=plots_bas3_50m$fe_omega_AngleNormal_q2_ExpShore
-plot4=plots_bas3_50m$fe_omega_AngleNormal_q4_ExpShore
-
-omega1 <- ggplot_build(plot1)$data[[1]]
-omega2 <- ggplot_build(plot2)$data[[1]]
-omega4 <- ggplot_build(plot4)$data[[1]]
-
-omega1$Category <- "Far"
-omega2$Category <- "Medium"
-omega4$Category <- "Close"
-
-# Combine the data frames
-combined_omega <- rbind(omega1, omega2, omega4)
-
-ci_omega1 <- ggplot_build(plot1)$data[[2]]
-ci_omega2 <- ggplot_build(plot2)$data[[2]]
-ci_omega4 <- ggplot_build(plot4)$data[[2]]
-
-ci_omega1$Category <- "Far"
-ci_omega2$Category <- "Medium"
-ci_omega4$Category <- "Close"
-
-# Combine the data frames
-combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)
-
-combined_plot <- ggplot() +
-  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
-  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
-  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
-  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
-  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
-  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
-  theme_minimal()
-
-ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")
-
-```
-
-## Response models with splines of ExpShip
-
-### Response models based on baseline1
-
-```{r fit response1}
-
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,
-                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=5,bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.04
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_35m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_35m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_35m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_35m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_35m$lambda()["tau.s(ID)",1],1,baseline1_35m$lambda()["nu.s(ID)",1],
-             baseline1_35m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_35m$update_coeff_re(new_coeff_re)
-response1_sp_35m$update_coeff_fe(baseline1_35m$coeff_fe()[,1])
-
-response1_sp_35m$update_lambda(new_lambda)
-
-response1_sp_35m$fit()
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_50m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_50m$lambda()["tau.s(ID)",1],1,baseline1_50m$lambda()["nu.s(ID)",1],
-             baseline1_50m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_50m$update_coeff_re(new_coeff_re)
-response1_sp_50m$update_coeff_fe(baseline1_50m$coeff_fe()[,1])
-
-response1_sp_50m$update_lambda(new_lambda)
-
-response1_sp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_75m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_75m$lambda()["tau.s(ID)",1],1,baseline1_75m$lambda()["nu.s(ID)",1],
-             baseline1_75m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_75m$update_coeff_re(new_coeff_re)
-response1_sp_75m$update_coeff_fe(baseline1_75m$coeff_fe()[,1])
-
-response1_sp_75m$update_lambda(new_lambda)
-
-response1_sp_75m$fit()
-
-# Fit response with 50m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_100m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_100m$lambda()["tau.s(ID)",1],1,baseline1_100m$lambda()["nu.s(ID)",1],
-             baseline1_100m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_100m$update_coeff_re(new_coeff_re)
-response1_sp_100m$update_coeff_fe(baseline1_100m$coeff_fe()[,1])
-
-response1_sp_100m$update_lambda(new_lambda)
-
-response1_sp_100m$fit()
-```
-
-```{r get results response1}
-
-estimates_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Est")
-std_res1_sp_35m=as.list(response1_sp_35m$tmb_rep(),what="Std")
-
-estimates_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Est")
-std_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Std")
-
-estimates_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Est")
-std_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Std")
-
-estimates_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Est")
-std_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShip"="Distance to ship")
-
-plots_res1_sp_35m=response1_sp_35m$get_all_plots(baseline=baseline1_35m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_35m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_50m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_75m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_100m",show_CI="pointwise",save=TRUE)
-
-```
-
-```{r estimates on parameter scale response1}
-
-post_coeff_res1_sp_35m=response1_sp_35m$post_coeff(n_post=1000)
-post_par_res1_sp_35m=list("tau0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_35m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_35m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_35m$log_lambda[,2])))
-mean_par_res1_sp_35m=lapply(post_par_res1_sp_35m,mean)
-sd_par_res1_sp_35m=lapply(post_par_res1_sp_35m,sd)
-
-post_coeff_res1_sp_50m=response1_sp_50m$post_coeff(n_post=1000)
-post_par_res1_sp_50m=list("tau0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,2])))
-mean_par_res1_sp_50m=lapply(post_par_res1_sp_50m,mean)
-sd_par_res1_sp_50m=lapply(post_par_res1_sp_50m,sd)
-
-post_coeff_res1_sp_75m=response1_sp_75m$post_coeff(n_post=1000)
-post_par_res1_sp_75m=list("tau0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,2])))
-mean_par_res1_sp_75m=lapply(post_par_res1_sp_75m,mean)
-sd_par_res1_sp_75m=lapply(post_par_res1_sp_75m,sd)
-
-post_coeff_res1_sp_100m=response1_sp_100m$post_coeff(n_post=1000)
-post_par_res1_sp_100m=list("tau0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,2])))
-mean_par_res1_sp_100m=lapply(post_par_res1_sp_100m,mean)
-sd_par_res1_sp_100m=lapply(post_par_res1_sp_100m,sd)
-
-```
-
-### Response model with splines of ExpShip based on baseline2
-
-```{r fit response2}
-
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-response2_sp_50m$update_coeff_re(new_coeff_re)
-response2_sp_50m$update_coeff_fe(baseline2$coeff_fe()[,1])
-
-response2_sp_50m$update_lambda(new_lambda)
-
-response2_sp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-response2_sp_75m$update_coeff_re(new_coeff_re)
-response2_sp_75m$update_coeff_fe(baseline2$coeff_fe()[,1])
-
-response2_sp_75m$update_lambda(new_lambda)
-
-response2_sp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-response2_sp_100m$update_coeff_re(new_coeff_re)
-response2_sp_100m$update_coeff_fe(baseline2$coeff_fe()[,1])
-
-response2_sp_100m$update_lambda(new_lambda)
-
-response2_sp_100m$fit()
-```
-
-```{r get results response2}
-estimates_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
-std_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")
-
-estimates_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
-std_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")
-
-estimates_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
-std_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShip"="Distance to ship")
-
-plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
 ```
@@
 
-post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
-post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
-mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
-sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)
+plot_Dshore=ggplot(data=dataBE12)+geom_point(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.5)+
+  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)+
+  theme_minimal()
 
-post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
-post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
-mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
-sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)
+ggsave("observed_Dshore.pdf",plot=plot_Dshore)
 
-post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
-post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
-mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
-sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)
+quantile(dataBE12$DistanceShore,probs=0.05)
 
@@
 
-### Response model with splines of ExpShip based on baseline3
-
-```{r fit response3}
-
-#initialize parameters
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-#Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_50m$lambda()["tau.s(ID)",1],1,baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
-response3_sp_50m$update_coeff_re(new_coeff_re)
-response3_sp_50m$update_coeff_fe(baseline3_50m$coeff_fe()[,1])
-
-response3_sp_50m$update_lambda(new_lambda)
-response3_sp_50m$fit()
-
-#Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_75m$lambda()["tau.s(ID)",1],1,baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
-response3_sp_75m$update_coeff_re(new_coeff_re)
-response3_sp_75m$update_coeff_fe(baseline3_75m$coeff_fe()[,1])
-
-response3_sp_75m$update_lambda(new_lambda)
-response3_sp_75m$fit()
-
-#Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_100m$lambda()["tau.s(ID)",1],1,baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
-response3_sp_100m$update_coeff_re(new_coeff_re)
-response3_sp_100m$update_coeff_fe(baseline3_100m$coeff_fe()[,1])
-
-response3_sp_100m$update_lambda(new_lambda)
-response3_sp_100m$fit()
-```
-
-```{r get results response3}
-estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
-std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")
-
-estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
-std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")
-
-estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
-std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
+```{r}
+dataBE12$angular_velocity=c(NA,diff(dataBE12$AngleNormal)/diff(dataBE12$time))
+plot_angular_velocity=ggplot(data=dataBE12)+geom_point(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.5)+
+  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)+
+  theme_minimal()
+ggsave("observed_angular_velocity.pdf",plot=plot_angular_velocity)
+max(dataBE12$angular_velocity,na.rm=TRUE)/(pi*(pi-pi/2)*(pi+pi/2)*exp(-1))
 
@@
 
-```{r estimates on parameter scale response3}
+```{r}
 
-post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
-post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
-mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
-sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)
+renamed_dataBE12=dataBE12
+colnames(renamed_dataBE12)[48]="theta"
+renamed_dataBE12[renamed_dataBE12$DistanceShore==0,]=0.02
 
-post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
-post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
-mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
-sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)
+#Fixed measurement error
+sigma_obs=0.05
+H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))
 
-post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
-post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
-mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
-sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)
+#parameters formulas
+formulas <- list(tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~s(theta,k=6,bs="cs"),D0=~1)
+
+# Fit baseline
+baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
+                    response = c("x","y"),par0 = c(2,4,0,0.05),other_data=list("H"=H),
+                    fixpar=c("D0"))
+
+baseline_crcvm_cubic$update_lambda(c(exp(3),exp(3),10^-12))
+baseline_crcvm_cubic$update_map(list("log_lambda"=factor(c(1,2,NA))))
+baseline_crcvm_cubic$fit()
 
@@
 
-### Response model with exponential of ExpShip based on baseline2
+```{r}
+plots_bas_crcvm=baseline_crcvm_cubic$get_all_plots(baseline=NULL,show_CI="simultaneous")
 
-```{r}
-
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
-                 nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-response2_exp_50m$update_coeff_re(new_coeff_re)
-response2_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_50m$update_lambda(new_lambda)
-
-response2_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-response2_exp_75m$update_coeff_re(new_coeff_re)
-response2_exp_75m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_75m$update_lambda(new_lambda)
-
-response2_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-response2_exp_100m$update_coeff_re(new_coeff_re)
-response2_exp_100m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_100m$update_lambda(new_lambda)
-
-response2_exp_100m$fit()
+plots_bas_crcvm
 ```
-
-```{r}
-estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
-std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")
-
-estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
-std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")
-
-estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
-std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res2_exp=response2_exp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
-
-### Response model with exponential of ExpShip based on baseline3
-
-```{r}
-
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
-                 nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_50m$coeff_fe()[1:3,1],0,baseline3_50m$coeff_fe()[4,1],0,baseline3_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_50m$lambda()["tau.s(ID)",1],baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
-response3_exp_50m$update_coeff_re(new_coeff_re)
-response3_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_50m$update_lambda(new_lambda)
-response3_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_75m$coeff_fe()[1:3,1],0,baseline3_75m$coeff_fe()[4,1],0,baseline3_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_75m$lambda()["tau.s(ID)",1],baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
-response3_exp_75m$update_coeff_re(new_coeff_re)
-response3_exp_75m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_75m$update_lambda(new_lambda)
-response3_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_100m$coeff_fe()[1:3,1],0,baseline3_100m$coeff_fe()[4,1],0,baseline3_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_100m$lambda()["tau.s(ID)",1],baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
-response3_exp_100m$update_coeff_re(new_coeff_re)
-response3_exp_100m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_100m$update_lambda(new_lambda)
-response3_exp_100m$fit()
-```
-
-```{r}
-estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
-std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")
-
-estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
-std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")
-
-estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
-std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```

### 2025-01-07 R/application/CVM/fit_cvm.Rmd
R code updates

@@
 ---
-title: "fit_crcvm"
+title: "Preliminary analysis of Greenland narwhals trajectories with smoothSDE"
 author: "Alexandre Delporte"
 date: "2024-06-07"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
+  word_document: default
+header-includes:
+- \usepackage{float}
+
 ---
+
+In this analysis, we fit SDE models to the narwhal data.
+In this preliminary work, the analysis is made without considering any constraints on the motion.
+Only the effect of the distance to the ship will be considered in the response model.
+
+# Set up
 
@@
 
-## Set up
-
-```{r packages}
+```{r packages,message=FALSE}
 set.seed(42)
 
-library(smoothSDE)
-library(ggplot2)
-library(dplyr)
-library(plotly)
-library(htmlwidgets)
-library(mgcv)
+library(smoothSDE) #sde fit
+library(ggplot2)  #plots
+library(dplyr)    #manipulate dataframes
+library(tidyr)    # same
+library(plotly)   # dynamic plots
+library(htmlwidgets)  #save dynamic plots
+library(mgcv)    # mixed models
+library(here)    # path management
+library(xtable)  # for latex tables
+library(sf)      # read shapefile for land polygons
+library(gganimate) # animated trajectories
+
 ```
@@
 # Set the path to the directory containing the data
-par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
-narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")
+narwhal_data_path <- here("Data", "preprocessed_data","narwhals")
 
 #data before exposure with first 24h removed
-dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")
+dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), 
+                  header = TRUE,dec = ".")
 
 #data before exposure with first 12h removed
-dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")
+dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"),
+                  header = TRUE,dec = ".")
 
 #data after exposure
-dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"), header = TRUE,dec = ".")
-
-```
-
-We choose thresholds on the distance to the shore to define the ExpShore covariate.
-
-```{r preprocess data}
-
-D_low=0.07
-D_up=3
-dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
-dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
-dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
-dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low
-
-N=length(unique(dataBE2$ID))
-
-n_pre=length(dataBE2$time) #number of observations before exposure
+dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"),
+                header = TRUE,dec = ".")
+
+N=length(unique(dataBE12$ID))
+
+n_pre12=length(dataBE12$time) #nb of obs before exposure with first 12h removed
+n_pre24=length(dataBE24$time) #nb of obs before exposure with first 24h removed
 n_post=length(dataAE$time) #number of observations after exposure
-```
-
-## Baseline models
-
-We can estimate the measurement error standard deviation directly from the data.
-Based on accuracy studies about fastloc GPS, it should be a few tens of metres.
-The estimate we get from the data before exposure is 25m (std error 1m), which seems consistent with previous studies.
-However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same.
-If we fix the measurement error to 25m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters.
-This cannot be solved by changing the other initial parameters in the optimisation.
-Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get.
-Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.
-
-### Baselines with constant parameters
-
-```{r fit baseline0}
+
+```
+
+# Baseline CTCRW models
+
+## CTCRW without random individual effects
+
+We fit a baseline CTCRW with constant parameters (no covariates) on the data before exposure \texttt{dateBE12}.
+The mean velocity $\mu$ is fixed to $(0,0)$ because we are not investigating a long term migration movement.
+
+```{r fit baseline ctcrw without random effects,message=FALSE}
+
+#initial parameters
+par0 <- c(0,0,1,4)
+
+#model formula
+formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1)
+
+#Fit baseline with measurement error estimated from the data
+baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
+                          response = c("x","y"),
+                          par0 = par0,
+                          other_data=list("log_sigma_obs0"=log(0.05)),
+                          fixpar=c("mu1","mu2"))
+baseline_ctcrw0$fit()
+
+```
+
+We now extract the estimated parameters and their standard deviations. These estimates are on the log-scale.
+
+```{r results baseline_ctcrw0}
+
+estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
+std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")
+
+```
+
+We process the results and show a table with the estimates and their standard deviations.
+
+```{r render coefficient estimates table baseline_ctcrw0 }
+
+# Function to extract estimate values into a data frame format
+combine_to_df <- function(estimates, std_dev, param_name) {
+  data.frame(
+    Parameter = param_name,
+    Estimate = unlist(estimates, use.names = FALSE),
+    Std_Error = unlist(std_dev, use.names = FALSE)
+  )
+}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
+                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe, 
+                          std_bas_ctcrw0$coeff_fe, 
+                          rownames(estimates_bas_ctcrw0$coeff_fe))
+
+# Combine all data frames into one
+table_data_baseline_ctcrw0 <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis',message=FALSE}
+latex_table <- xtable(table_data_baseline_ctcrw0, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
+```
+
+From this we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$.
+
+```{r CI baseline_ctcrw0 parameters}
+
+post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
+post_par_bas_ctcrw0=
+  list("sigma_obs"=1000*exp(post_coeff_bas_ctrcw0$log_sigma_obs),
+              "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
+              "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]))
+mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
+sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
+quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
+                               quantile, probs = c(0.025, 0.975))
+
+```
+
+We include the results in a latex table.
+
+```{r CI table baseline_ctcrw0 parameters,results='asis',message=FALSE}
+
+parameter_names <- names(mean_par_bas_ctcrw0)
+estimates <- mean_par_bas_ctcrw0
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
+          quant_par_bas_ctcrw0[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = c("$\\sigma_{obs}$","$\\tau$", "$\\nu$"),
+  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+
+```{r,results='asis',echo=FALSE}
+# Create an xtable object
+xtab <- xtable(
+  table_data,
+  caption = "Baseline estimations",
+  label = "table:baseline_estimations"
+)
+
+# Customize xtable options
+print(
+  xtab,
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
+  hline.after = c(-1, 0, nrow(table_data)),
+  table.placement="H"
+)
+```
+
+The estimate for $\tau$ is $\bar \tau =1.08$ h and the estimate for $\nu$ is $\bar \nu =4.69$ km/h.
+The measurement error standard deviation is estimated to $\bar \sigma_{obs}\simeq 38$ m.
+The measurement error in Fastloc GPS data mainly depends on the number of satellite that were used to compute the location.
+This number might change from a data point to another.
+However, we don't know the number of satellites that computed the position in the data.
+That is why we assume a constant standard deviation for the measurement error.
+We know from Weensveen paper, table 3, that it should be at most around 30m when only 4 satellites are involved, and then decrease as the number of satellites increases.
+This suggest that the measurement error might be slightly overestimated here.
+
+## Sensitivity to measurement error standard deviation
+
+To investigate the sensitivity of the parameters estimates to the measurement error, we fix $\sigma_{obs}$ to different values and estimate the parameters again.
+
+```{r fit baseline ctcrw fixed error without random effects,message=FALSE}
+
+#fixed values for std of measurement error
+sigma_obs=c(0.025,0.05,0.075,0.1)
+list_baseline_ctcrw0=list()
+
+for (i in 1:length(sigma_obs)){
+
+  H=array(rep(sigma_obs[i]^2*diag(2),n_pre12),dim=c(2,2,n_pre12))
+
+  baseline_ctcrw0_fixed_error<- SDE$new(formulas = formulas,data = dataBE12,
+                                        type = "CTCRW",
+                              response = c("x","y"),
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
+  baseline_ctcrw0_fixed_error$fit()
+  list_baseline_ctcrw0[[i]]= baseline_ctcrw0_fixed_error
+}
+
+```
+
+```{r results for baselines with fixed measurement error}
+list_baseline_ctcrw0_est=list()
+list_baseline_ctcrw0_std=list()
+
+for (i in seq_along(list_baseline_ctcrw0)) {
+  estimates=as.list(list_baseline_ctcrw0[[i]]$tmb_rep(),what="Est")
+  std=as.list(list_baseline_ctcrw0[[i]]$tmb_rep(),what="Std")
+  list_baseline_ctcrw0_est[[i]]=estimates
+  list_baseline_ctcrw0_std[[i]]=std
+}
+
+```
+
+```{r render estimates table baselines with fixed mesaurement error,results='asis', message=FALSE}
+
+for (i in seq_along(list_baseline_ctcrw0)) {
+  estimates=list_baseline_ctcrw0_est[[i]]
+  std=list_baseline_ctcrw0_std[[i]]
+
+  # Create data frames for each set
+  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
+                               std$log_sigma_obs, "log_sigma_obs")
+  coeff_fe <- combine_to_df(estimates$coeff_fe, 
+                          std$coeff_fe, 
+                          rownames(estimates$coeff_fe))
+
+  # Combine all data frames into one
+  table_data <- bind_rows(log_sigma_obs, coeff_fe)
+
+  latex_table <- xtable(
+    table_data, 
+    caption = paste(
+      "Parameter Estimates and Standard Errors for fixed \\(\\sigma_{obs}=\\) ",
+      sigma_obs[i]," m"),
+    xtable.table.placement="H")
+
+  # Print LaTeX table
+  print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
+
+}
+
+```
+
+```{r plot tau and nu estimates as a function of measurement error}
+tau_estimates=sapply(list_baseline_ctcrw0_est,
+                     function(estimates) 
+                       {exp(estimates$coeff_fe["tau.(Intercept)",1])})
+nu_estimates=sapply(list_baseline_ctcrw0_est,
+                    function(estimates) 
+                      {exp(estimates$coeff_fe["nu.(Intercept)",1])})
+tau_std=sapply(list_baseline_ctcrw0_std,
+               function(std) {exp(std$coeff_fe["tau.(Intercept)",1])})
+nu_std=sapply(list_baseline_ctcrw0_std,
+              function(std) {exp(std$coeff_fe["nu.(Intercept)",1])})
+df=as.data.frame(cbind(tau_estimates,nu_estimates,sigma_obs))
+colnames(df)=c("tau","nu","sigma_obs")
+df <- df %>%
+  pivot_longer(cols = c(tau, nu), names_to = "parameter", values_to = "value")
+
+ggplot(df, aes(x = 1000*sigma_obs, y = value)) +
+  geom_point() +
+  geom_line() +
+  theme_minimal() +
+  xlab("Sigma_obs") +
+  ylab("Value") +
+  ggtitle("Estimated Tau and Nu as functions of fixed Sigma_obs") +
+  facet_wrap(~parameter, scales = "free_y")  # Facet by 'parameter' (tau or nu)
+
+```
+
+We see a clear effect of the fixed measurement error on the estimates of $\tau$, and a slight influence of $\nu$.
+We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ increases.
+Overall, knowing or estimating precisely the measurement error will be crucial in the analysis.
+
+### CTCRW with random individual effects
+
+```{r fit baseline ctcrw with random effects,cache=TRUE,message=FALSE}
+
+#model formula
+formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))
+
+#Fit baseline with measurement error estimated from the data
+baseline_ctcrw1<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
+                          response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
+                    fixpar=c("mu1","mu2"))
+baseline_ctcrw1$fit()
+
+```
+
+```{r results baseline_ctcrw1}
+
+estimates_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Est")
+std_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table baseline_ctcrw1}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_bas_ctcrw1$log_sigma_obs,
+                               std_bas_ctcrw1$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_bas_ctcrw1$coeff_fe,
+                          std_bas_ctcrw1$coeff_fe,
+                          rownames(estimates_bas_ctcrw1$coeff_fe))
+log_lambda <- combine_to_df(estimates_bas_ctcrw1$log_lambda, 
+                            std_bas_ctcrw1$log_lambda, 
+                            rownames(estimates_bas_ctcrw1$log_lambda))
+coeff_re <- combine_to_df(estimates_bas_ctcrw1$coeff_re, 
+                          std_bas_ctcrw1$coeff_re, 
+                          rownames(estimates_bas_ctcrw1$coeff_re))
+
+# Combine all data frames into one
+table_data_baseline_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe, log_lambda, coeff_re)
+
+```
+
+```{r, results='asis', echo=FALSE}
+latex_table <- xtable(table_data_baseline_ctcrw1, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
+```
+
+As previously, we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$, and include the results in a latex table.
+
+```{r CI baseline_ctcrw1 parameters}
+
+post_coeff_bas_ctrcw1=baseline_ctcrw1$post_coeff(n_post=10000)
+post_par_bas_ctcrw1=list(
+  "sigma_obs"=1000*exp(post_coeff_bas_ctrcw1$log_sigma_obs),
+  "tau"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"nu.(Intercept)"]),
+  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"tau.s(ID)"])),
+  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"nu.s(ID)"])))
+
+mean_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,mean)
+sd_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,sd)
+quant_par_bas_ctcrw1 <- lapply(post_par_bas_ctcrw1, 
+                               quantile, probs = c(0.025, 0.975))
+
+```
+
+```{r CI table baseline_ctcrw1 parameters, message=FALSE}
+
+parameter_names <- names(mean_par_bas_ctcrw1)
+estimates <- mean_par_bas_ctcrw1
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw1[[param]][1], 
+          quant_par_bas_ctcrw1[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = 
+    c("$\\sigma_{obs}$","$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
+  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw1),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+```{r,results='asis',echo=FALSE}
+xtab <- xtable(
+  table_data,
+  caption = "Baseline estimations with random effects",
+  label = "table:baseline_estimations_with_random_effects"
+)
+
+print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
+  hline.after = c(-1, 0, nrow(table_data)),  
+  table.placement="H"
+)
+```
+
+Note that the random effect is multiplicative and not additive since it appears in the $\log$ of the parameters. We can plot the estimates on the parameter scale for each narwhal.
+
+```{r plot baseline_ctcrw1 estimates,results='hide'}
+
+baseline_ctcrw1$get_all_plots(baseline=NULL,show_CI="simultaneous")
+
+```
+
+### RACVM without random individual effects
+
+We can also test whether adding a constant rotational $\omega$ in the SDE model would make sense. For this, we fit a RACVM model.
+
+```{r fit baseline racvm,message=FALSE}
 
@@
 #model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),omega=~1)
-
-# Fit baseline with 25m measurement error
-sigma_obs=0.025
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_25m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-baseline0_100m$fit()
-
-```
-
-```{r results baseline0}
-estimates_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Est")
-std_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Std")
-
-estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
-std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")
-
-estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
-std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")
-
-estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
-std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")
-
-```
-
-### Baseline with AngleNormal covariate in omega
-
-Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau.
-Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.
-
-```{r fit baseline1}
+formulas <- list(mu1=~1,mu2=~1,
+                 tau =~1,
+                 nu=~1,
+                 omega=~1)
+
+#Fit baseline with measurement error estimated from the data
+baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
+                          response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
+                    fixpar=c("mu1","mu2"))
+baseline_racvm0$fit()
+
+```
+
+```{r results baseline_racvm0}
+
+estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
+std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table baseline_racvm0}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
+                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe,
+                          std_bas_racvm0$coeff_fe, 
+                          rownames(estimates_bas_racvm0$coeff_fe))
+
+# Combine all data frames into one
+table_data_baseline_racvm0 <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis', echo=FALSE}
+latex_table <- xtable(table_data_baseline_racvm0, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
+```
+
+The estimate of $\omega$ is $0$.
+This indicates that a \textbf{constant} parameter $\omega$ is irrelevant in the model.
+As a consequence, the final baseline model we keep is \texttt{baseline\_ctcrw1} which corresponds to a CTCRW with constant parameters $\tau$ and $\nu$ and mean velocity fixed to zero.
+
+## Sensitivity analysis of the baseline CTCRW to the tagging effect threshold
+
+In this first analysis, we only removed 12h of data after the tagging for each narwhal.
+However, the narwhals might still be affected by the tagging and not behave "normally".
+This would invalidate all our conclusions in the response model, if the baseline model happens to be biased.
+Therefore, we need to assess the sensitivity of the baseline model to the time threshold we choose for the tagging effect.
+Here, we only repeat the fit for the model \texttt{baseline\_ctcrw1} with the sub dataset \texttt{dataBE24} where all the data until 24h after tagging have been removed.
+
+```{r fit baseline sensitivity analysis tagging effect,message=FALSE}
 
 #initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1,1,1)
+par0 <- c(0,0,1,4)
+
+#model formula
+formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))
+
+#Fit baseline with measurement error estimated from the data
+baseline_ctcrw1_24h<- SDE$new(
+  formulas = formulas,data = dataBE24,type = "CTCRW",
+  response = c("x","y"),par0 = par0,
+  other_data=list("log_sigma_obs0"=log(0.05)),
+  fixpar=c("mu1","mu2"))
+
+baseline_ctcrw1_24h$fit()
+
+```
+
+```{r results baseline_ctcrw1_24h}
+
+estimates_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Est")
+std_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table baseline_ctcrw1_24h}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(
+  estimates_bas_ctcrw1_24h$log_sigma_obs, 
+  std_bas_ctcrw1_24h$log_sigma_obs, "log_sigma_obs")
+
+coeff_fe <- combine_to_df(estimates_bas_ctcrw1_24h$coeff_fe, 
+                          std_bas_ctcrw1_24h$coeff_fe, 
+                          rownames(estimates_bas_ctcrw1_24h$coeff_fe))
+
+# Combine all data frames into one
+table_data_baseline_ctcrw1_24h <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis',echo=FALSE}
+latex_table <- xtable(table_data_baseline_ctcrw1_24h, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,
+      table.placement="H")
+```
+
+The standard deviations are wider due to the smaller amount of data, but also due to the fact that the measurement error is estimated higher than before.
+$\tau$ seems to be estimated slightly below the value we obtained for the dataset \texttt{dataBE12}, but the difference seems reasonable as the CI of the two estimates overlap.
+
+```{r fit baseline sensitivity analysis tagging effect fixed error,message=FALSE}
+
+sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+
+#Fit baseline with measurement error 
+baseline_ctcrw1_24h_fixed_error<- SDE$new(formulas = formulas,
+                                          data = dataBE24,type = "CTCRW",
+                                          response = c("x","y"),
+                                          par0 = par0,
+                                          other_data=list("H"=H),
+                                          fixpar=c("mu1","mu2"))
+baseline_ctcrw1_24h_fixed_error$fit()
+
+```
+
+```{r results baseline_ctcrw1_24h_fixed_error}
+
+estimates_bas_ctcrw1_24h_fixed_error=as.list(
+  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Est")
+std_bas_ctcrw1_24h_fixed_error=as.list(
+  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table baseline_ctcrw1_24h_fixed_error }
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(
+  estimates_bas_ctcrw1_24h_fixed_error$log_sigma_obs,
+  std_bas_ctcrw1_24h_fixed_error$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(
+  estimates_bas_ctcrw1_24h_fixed_error$coeff_fe, 
+  std_bas_ctcrw1_24h_fixed_error$coeff_fe, 
+  rownames(estimates_bas_ctcrw1_24h_fixed_error$coeff_fe))
+
+# Combine all data frames into one
+table_data_baseline_ctcrw1_24h_fixed_error <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis',echo=FALSE}
+latex_table <- xtable(table_data_baseline_ctcrw1_24h_fixed_error, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,
+      table.placement="H")
+```
+
+When the measurement error is fixed to the value we initially estimated from the whole dataset \texttt{dateBE12}, we observe a big decrease in the estimate of $\tau$.
+This indeed suggest a shift in the behavior of the whales within the time interval [12h,24h] after tagging, which might invalidate our results.
+
+# Response CTCRW models
+
+## Response CTCRW with constant parameters
+
+First, we try to fit the exact same model as \texttt{baseline\_ctcrw1} on the data during exposure to assess a potential shift in the behavior due to the presence of the ship.
+We fix the measurement error to the value that has been estimated from the data before exposure.
+
+```{r fit response constant parameters,message=FALSE}
+
+#initial parameters
+par0 <- c(0,0,1,4)
+
+#model formula
+formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))
+
+sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+
+#Fit respopnse with measurement error estimated from the baseline
+response_ctcrw1<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
+                          response = c("x","y"),
+                          par0 = par0,
+                          other_data=list("H"=H),
+                          fixpar=c("mu1","mu2"))
+response_ctcrw1$fit()
+
+```
+
+```{r results response_ctcrw1}
+
+estimates_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Est")
+std_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table response_ctcrw1}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_res_ctcrw1$log_sigma_obs, 
+                               std_res_ctcrw1$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_res_ctcrw1$coeff_fe,
+                          std_res_ctcrw1$coeff_fe, 
+                          rownames(estimates_res_ctcrw1$coeff_fe))
+
+# Combine all data frames into one
+table_data_response_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis', echo=FALSE}
+latex_table <- xtable(table_data_response_ctcrw1, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,
+      table.placement="H")
+```
+
+The estimated standard deviation of the random effects can be computed with \texttt{sdev} method in \texttt{smoothSDE}.
+
+```{r}
+response_ctcrw1$sdev()
+```
+
+The estimates are lower than those obtained from the data before exposure.
+This may happen because we have short trajectories before exposure, that may not represent the whole range of possible movement for each narwhal, and therefore there may be more varieties of movements between the different individuals.
+On the contrary, the trajectories during exposure are longer so we can reasonably say that they represent better the range of possible movement.
+The lower standard deviation of the random effects might then indicate that on the long run, the narwhals in the study tend to have similar movement characteristics.
+
+The estimated value for $\tau$ is slightly lower than the value we obtained from the data before exposure.
+The value for $\nu$ is also slightly lower.
+In both cases, the CI of the estimate overlaps with the CI of the estimate from the data before exposure.
+To better understand if this difference in the estimated values of the parameters is due to the presence of the ship, we will introduce the exposure to ship \texttt{ExpShip} (or $E_{ship}$ in the paper) covariate in the model.
+
+## Response models with splines of ExpShip
+
+The intercept of the parameters $\tau$ and $\nu$ are considered as offset in the statistical model, and fixed using the estimates we get from the baseline model \texttt{baseline\_ctcrw1}.
+The measurement error is an offset 7that is fixed to the value that has been estimated from the data before exposure.
+
+In the response model, we suppose that the presence of the ship has a similar effect on all the narwhals.
+Therefore, we estimate a mean reaction (if any) to the presence of the ship.
+We are aware that a variety of reactions might be possible, depending on the narwhals habituation to noise or on the specific place where it has been approached by the ship for instance.
+This variety of reaction can be assessed visually by looking at a small subset of the data, but is difficult to infer statistically (see [Heide-Jorgensen2021] figure 7).
+
+```{r response with splines,cache=TRUE}
 
@@
 formulas <- list(mu1=~1,mu2=~1,
-                 tau =~s(ID,bs="re"),
-                 nu=~s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=5,bs="cs"))
-
-# Fit baseline with 25m measurement error
-sigma_obs=0.025
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_25m$update_lambda(init_lambda)
-
-baseline1_25m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_50m$update_lambda(init_lambda)
-
-baseline1_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_75m$update_lambda(init_lambda)
-
-baseline1_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))
-
-baseline1_100m$update_lambda(init_lambda)
-
-baseline1_100m$fit()
-
-```
-
-```{r results baseline1}
-estimates_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Est")
-std_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Std")
-
-estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
-std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")
-
-estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
-std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")
-
-estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
-std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")
-
-plots_bas1_25m=baseline1_25m$get_all_plots(baseline=NULL,model_name="baseline1_25m",show_CI="pointwise",save=TRUE)
-plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
-plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
-plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)
-
-```
-
-### Baseline with AngleNormal and DistanceShore in omega
-
-```{r fit baseline2}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)
-
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 25m measurement error
-sigma_obs=0.025
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_25m$update_lambda(init_lambda)
-baseline2_25m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_50m$update_lambda(init_lambda)
-baseline2_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_75m$update_lambda(init_lambda)
-baseline2_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline2_100m$update_lambda(init_lambda)
-baseline2_100m$fit()
-
-```
-
-```{r get results baseline2}
-
-estimates_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Est")
-std_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Std")
-
-estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
-std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")
-
-estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
-std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")
-
-estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
-std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")
+                 tau =~s(ExpShip,k=10,bs="cs")+s(ID,bs="re"),
+                 nu=~s(ExpShip,k=10,bs="cs")+s(ID,bs="re"))
+
+# Fit response with measurement error fixed to baseline value
+sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
+H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+response_ctcrw_sp<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
+                            response = c("x","y"),
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
+
+# fix random effects to baseline values
+response_ctcrw_sp$update_coeff_re(
+  c(rep(0,9),baseline_ctcrw1$coeff_re()[1:6,1],rep(0,9)
+    ,baseline_ctcrw1$coeff_re()[7:12,1]))
+response_ctcrw_sp$update_lambda(
+  c(1,baseline_ctcrw1$lambda()[1,1],1,baseline_ctcrw1$lambda()[2,1]))
+response_ctcrw_sp$update_coeff_fe(baseline_ctcrw1$coeff_fe())
+#use the indices of the variable in the map for TMB 
+#to keep the coefficients in the right order
+#otherwise we may have issues in function post_coeff
+response_ctcrw_sp$update_map(list(coeff_re=factor(
+  c(1:9,rep(NA,6),10:18,rep(NA,6))),
+  coeff_fe=factor(rep(NA,4)),
+  log_lambda=factor(c(1,NA,2,NA))))
+
+response_ctcrw_sp$fit()
+
+```
+
+```{r results response spline}
+
+estimates_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Est")
+std_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table response_ctcrw_sp}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_res_ctcrw_sp$log_sigma_obs, 
+                               std_res_ctcrw_sp$log_sigma_obs, "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_res_ctcrw_sp$coeff_fe,
+                          std_res_ctcrw_sp$coeff_fe,
+                          rownames(estimates_res_ctcrw_sp$coeff_fe))
+coeff_re <- combine_to_df(estimates_res_ctcrw_sp$coeff_re, 
+                          std_res_ctcrw_sp$coeff_re, 
+                          rownames(estimates_res_ctcrw_sp$coeff_re))
+log_lambda <- combine_to_df(estimates_res_ctcrw_sp$log_lambda, 
+                            std_res_ctcrw_sp$log_lambda, 
+                            rownames(estimates_res_ctcrw_sp$log_lambda))
+
+# Combine all data frames into one
+table_data_response_ctcrw_sp <- bind_rows(
+  log_sigma_obs, coeff_fe,coeff_re,log_lambda)
+
+```
+
+```{r,results='asis',echo=FALSE}
+latex_table <- xtable(table_data_response_ctcrw_sp, 
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,
+      table.placement="H")
+```
+
+```{r plot smooth parameters response_ctcrw_sp}
+plot_smooth_par=response_ctcrw_sp$plot_par(var="ExpShip",
+                                           par_names=c("tau","nu"),
+                                           n_post=100)
+
+#get data for the plot
+plot_data <- ggplot_build(plot_smooth_par)$data[[1]]
+#rename columns
+colnames(plot_data)[1:5]=c("mle","val","var","group","par")
+
+# ELow exposure threshold
+threshold <- 1 / 45 # 1/ max(distance to ship)
+plot_data <- plot_data %>% filter(var > threshold)
+plot_data$distance <- rep(1/plot_data[plot_data$group==1,"var"],
+                          length(unique(plot_data$group)))
+
+plot_data <- plot_data %>%
+  mutate(mle= recode(mle, "#B300001A" = "no", "#000000FF" = "yes"))
+plot_data <- plot_data %>%
+  mutate(par= recode(par, "1" = "tau", "2" = "nu"))
+
+pal <- c("no" = rgb(0.7, 0, 0, 0.1), "yes" = rgb(0, 0, 0, 1))
+
+plot_smooth_par_distance=ggplot(plot_data, aes(x = distance, group = group)) +
+  scale_colour_manual(values = pal, guide = "none") +
+  facet_wrap(c("par"), scales = "free_y",strip.position = "left") +
+  theme_minimal() + xlab("Distance to ship")  + 
+  ggtitle("Smooth parameters as a function of distance to ship") +
+  theme(strip.background = element_blank(),strip.placement = "outside",
+        strip.text = element_text(colour = "black"))+
+  geom_line(aes(y = val,color=mle))
+
+plot_smooth_par_distance
+
+```
+
+The estimated smooth functions depend on the number of knots we choose in the splines.
+For $\nu$, the estimated smooth does not change as we increase gradually the degree of freedom to $10$, meaning that only a few knots (1 or 2) are needed.
+The effect of the presence of the ship on the parameter $\nu$ is easily interpreted : the closer the narwhals are to the ship, the higher their velocity, likely to move away or avoid the ship.
+Now, the estimated smooth function for $\tau$ requires more knots and is harder to interpret.
+Though there seem to be a decrease in persistence close to the ship, at a medium distance around $20$ km the persistence of the movement seem to increase.
+This is likely connected to figure 7 in [Heide-Jorgensen], where depending on how the narwhals were first exposed to the ship, they either change direction (lower persistence) or persist in their heading to flee away (higher persistence).
+
+However, we only plot 100 posterior draws here.
+If we increase to $n_{post}=1000$ posterior draws, the confidence intervals appear much wider.
+We now construct confidence intervals from 1000 posterior draws of the estimated coeffs and compare the smooth parameter with the baseline parameter.
+
+```{r plot CI response_ctcrw_sp and baseline_ctcrw0,results='hide'}
 
 #plot parameters
-
-xmin=list("DistanceShore"=0.07)
-xmax=list("DistanceShore"=5)
-
-plots_bas2_25m=baseline2_25m$get_all_plots(baseline=NULL,model_name="baseline2_25m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
-```
-
-```{r estimates on parameter scale baseline2}
-
-post_coeff_bas2_25m=baseline2_25m$post_coeff(n_post=1000)
-post_par_bas2_25m=list("tau0"=exp(post_coeff_bas2_25m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_25m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_25m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,2])))
-mean_par_bas2_25m=lapply(post_par_bas2_25m,mean)
-sd_par_bas2_25m=lapply(post_par_bas2_25m,sd)
-
-post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
-post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
-mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
-sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)
-
-post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
-post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
-mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
-sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)
-
-#get 95% confidence intervals
-post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
-post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
-mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
-sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
-```
-
-```{r smooth omega baseline2}
-
-#get angular velocity for different distances to shore
-
-plot1=plots_bas2_50m$fe_omega_AngleNormal_q1_DistanceShore
-plot2=plots_bas2_50m$fe_omega_AngleNormal_q2_DistanceShore
-plot4=plots_bas2_50m$fe_omega_AngleNormal_q4_DistanceShore
-
-omega1 <- ggplot_build(plot1)$data[[1]]
-omega2 <- ggplot_build(plot2)$data[[1]]
-omega4 <- ggplot_build(plot4)$data[[1]]
-
-omega1$Category <- "Close"
-omega2$Category <- "Medium"
-omega4$Category <- "Far"
-
-# Combine the data frames
-combined_omega <- rbind(omega1, omega2, omega4)
-
-ci_omega1 <- ggplot_build(plot1)$data[[2]]
-ci_omega2 <- ggplot_build(plot2)$data[[2]]
-ci_omega4 <- ggplot_build(plot4)$data[[2]]
-
-ci_omega1$Category <- "Close"
-ci_omega2$Category <- "Medium"
-ci_omega4$Category <- "Far"
-
-# Combine the data frames
-combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)
-
-combined_plot <- ggplot() +
-  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
-  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
-  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
-  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
-  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
-  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
-  theme_minimal()
-
-ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline2_50m",width = 10, height = 6, units = "in")
-
-```
-
-### Baseline with AngleNormal and ExpShore in omega
-
-```{r fit baseline3}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-init_lambda=c(1,1,0.01,1,1,1)
-
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-# Fit baseline with 25m measurement error
-sigma_obs=0.025
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_25m$update_lambda(init_lambda)
-baseline3_25m$fit()
-
-# Fit baseline with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_50m$update_lambda(init_lambda)
-baseline3_50m$fit()
-
-# Fit baseline with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_75m$update_lambda(init_lambda)
-baseline3_75m$fit()
-
-# Fit baseline with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))
-
-baseline3_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))
-
-baseline3_100m$update_lambda(init_lambda)
-baseline3_100m$fit()
-```
-```{r get results baseline3}
-
-estimates_bas3_25m=as.list(baseline3_25m$tmb_rep(),what="Est")
-std_bas3_25m=as.list(baseline3_25m$tmb_rep(),what="Std")
-
-estimates_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Est")
-std_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Std")
-
-estimates_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Est")
-std_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Std")
-
-estimates_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Est")
-std_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Std")
-
-#plot parameters
-
-xmin=list("ExpShore"=1/D_up)
-xmax=list("ExpShore"=1/D_low)
-link=list("ExpShore"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore")
-
-plots_bas3_25m=baseline3_25m$get_all_plots(baseline=NULL,model_name="baseline3_25m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
-                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
-
-```{r estimates on parameter scale baseline3}
-
-post_coeff_bas3_25m=baseline3_25m$post_coeff(n_post=1000)
-post_par_bas3_25m=list("tau0"=exp(post_coeff_bas3_25m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_25m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_25m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_25m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_25m$log_lambda[,2])))
-mean_par_bas3_25m=lapply(post_par_bas3_25m,mean)
-sd_par_bas3_25m=lapply(post_par_bas3_25m,sd)
-
-post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
-post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,2])))
-mean_par_bas3_50m=lapply(post_par_bas3_50m,mean)
-sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)
-
-post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
-post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,2])))
-mean_par_bas3_75m=lapply(post_par_bas3_75m,mean)
-sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)
-
-post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
-post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_bas3_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_bas3_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,2])))
-mean_par_bas3_100m=lapply(post_par_bas3_100m,mean)
-sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)
-
-```
-
-```{r smooth omega baseline3}
-
-plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
-plot2=plots_bas3_50m$fe_omega_AngleNormal_q2_ExpShore
-plot4=plots_bas3_50m$fe_omega_AngleNormal_q4_ExpShore
-
-omega1 <- ggplot_build(plot1)$data[[1]]
-omega2 <- ggplot_build(plot2)$data[[1]]
-omega4 <- ggplot_build(plot4)$data[[1]]
-
-omega1$Category <- "Far"
-omega2$Category <- "Medium"
-omega4$Category <- "Close"
-
-# Combine the data frames
-combined_omega <- rbind(omega1, omega2, omega4)
-
-ci_omega1 <- ggplot_build(plot1)$data[[2]]
-ci_omega2 <- ggplot_build(plot2)$data[[2]]
-ci_omega4 <- ggplot_build(plot4)$data[[2]]
-
-ci_omega1$Category <- "Far"
-ci_omega2$Category <- "Medium"
-ci_omega4$Category <- "Close"
-
-# Combine the data frames
-combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)
-
-combined_plot <- ggplot() +
-  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
-  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
-  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
-  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
-  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
-  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
-  theme_minimal()
-
-ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")
-
-```
-
-## Response models with splines of ExpShip
-
-### Response models based on baseline1
-
-```{r fit response1}
-
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,
-                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=5,bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_50m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_50m$lambda()["tau.s(ID)",1],1,baseline1_50m$lambda()["nu.s(ID)",1],
-             baseline1_50m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_50m$update_coeff_re(new_coeff_re)
-response1_sp_50m$update_coeff_fe(baseline1_50m$coeff_fe()[,1])
-
-response1_sp_50m$update_lambda(new_lambda)
-
-response1_sp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_75m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_75m$lambda()["tau.s(ID)",1],1,baseline1_75m$lambda()["nu.s(ID)",1],
-             baseline1_75m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_75m$update_coeff_re(new_coeff_re)
-response1_sp_75m$update_coeff_fe(baseline1_75m$coeff_fe()[,1])
-
-response1_sp_75m$update_lambda(new_lambda)
-
-response1_sp_75m$fit()
-
-# Fit response with 50m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response1_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
-                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
-                                                                                                          "nu.s(ExpShip)",NA,NA))))
-
-new_coeff_re=c(rep(0,2),baseline1_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
-               baseline1_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline1_100m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
-new_lambda=c(1,baseline1_100m$lambda()["tau.s(ID)",1],1,baseline1_100m$lambda()["nu.s(ID)",1],
-             baseline1_100m$lambda()["omega.s(AngleNormal)",1])
-response1_sp_100m$update_coeff_re(new_coeff_re)
-response1_sp_100m$update_coeff_fe(baseline1_100m$coeff_fe()[,1])
-
-response1_sp_100m$update_lambda(new_lambda)
-
-response1_sp_100m$fit()
-```
-
-```{r get results response1}
-
-estimates_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Est")
-std_res1_sp_50m=as.list(response1_sp_50m$tmb_rep(),what="Std")
-
-estimates_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Est")
-std_res1_sp_75m=as.list(response1_sp_75m$tmb_rep(),what="Std")
-
-estimates_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Est")
-std_res1_sp_100m=as.list(response1_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
+xmin=list("ExpShip"=1/45)
+xmax=list("ExpShip"=1/5)
 link=list("ExpShip"=(\(x) 1/x))
@@
 
-plots_res1_sp_50m=response1_sp_50m$get_all_plots(baseline=baseline1_50m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_50m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_75m=response1_sp_75m$get_all_plots(baseline=baseline1_75m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_75m",show_CI="pointwise",save=TRUE)
-plots_res1_sp_100m=response1_sp_100m$get_all_plots(baseline=baseline1_100m,xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,
-                            model_name="response1_sp_100m",show_CI="pointwise",save=TRUE)
-
-```
-
-```{r estimates on parameter scale response1}
-
-post_coeff_res1_sp_50m=response1_sp_50m$post_coeff(n_post=1000)
-post_par_res1_sp_50m=list("tau0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_50m$log_lambda[,2])))
-mean_par_res1_sp_50m=lapply(post_par_res1_sp_50m,mean)
-sd_par_res1_sp_50m=lapply(post_par_res1_sp_50m,sd)
-
-post_coeff_res1_sp_75m=response1_sp_75m$post_coeff(n_post=1000)
-post_par_res1_sp_75m=list("tau0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_75m$log_lambda[,2])))
-mean_par_res1_sp_75m=lapply(post_par_res1_sp_75m,mean)
-sd_par_res1_sp_75m=lapply(post_par_res1_sp_75m,sd)
-
-post_coeff_res1_sp_100m=response1_sp_100m$post_coeff(n_post=1000)
-post_par_res1_sp_100m=list("tau0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp_100m$log_lambda[,2])))
-mean_par_res1_sp_100m=lapply(post_par_res1_sp_100m,mean)
-sd_par_res1_sp_100m=lapply(post_par_res1_sp_100m,sd)
-
-```
-
-### Response model with splines of ExpShip based on baseline2
-
-```{r fit response2}
-
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
+response_ctcrw_sp$get_all_plots(baseline=baseline_ctcrw1,xmin=xmin,xmax=xmax,
+                                link=link,xlabel=xlabel,show_CI="poinwise")
+
+```
+
+Even if there seems to be an influence of ExpShip over the parameters, the confidence intervals are too wide to really conclude.
+
+## Log linear CTCRW response model with exponential of ExpShip
+
+To simplify the analysis, instead of using spline, we now try a simpler log-linear model for the parameters $\tau$ and $\nu$, meaning that $\tau$ and $\nu$ are exponentials of a linear function of $E_{ship}$.
+
+```{r fit response log linear}
+
+#define model
+formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
+                 nu=~ExpShip+s(ID,bs="re"))
+
+# Fit response with measurement error fixed to baseline value
+sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
 H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-response2_sp_50m$update_coeff_re(new_coeff_re)
-response2_sp_50m$update_coeff_fe(baseline2_50m$coeff_fe()[,1])
-
-response2_sp_50m$update_lambda(new_lambda)
-
-response2_sp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-response2_sp_75m$update_coeff_re(new_coeff_re)
-response2_sp_75m$update_coeff_fe(baseline2_75m$coeff_fe()[,1])
-
-response2_sp_75m$update_lambda(new_lambda)
-
-response2_sp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
-                    map=list(coeff_re=factor(c(1:2,rep(NA,N),3:4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                             log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-response2_sp_100m$update_coeff_re(new_coeff_re)
-response2_sp_100m$update_coeff_fe(baseline2_100m$coeff_fe()[,1])
-
-response2_sp_100m$update_lambda(new_lambda)
-
-response2_sp_100m$fit()
-```
-
-```{r get results response2}
-estimates_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
-std_res2_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")
-
-estimates_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
-std_res2_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")
-
-estimates_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
-std_res2_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")
+
+response_ctcrw_exp<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
+                             response = c("x","y"),par0 = par0,
+                             other_data=list("H"=H),fixpar=c("mu1","mu2"))
+
+# fix random effects to baseline values
+response_ctcrw_exp$update_coeff_re(baseline_ctcrw1$coeff_re())
+response_ctcrw_exp$update_lambda(c(baseline_ctcrw1$lambda()))
+response_ctcrw_exp$update_coeff_fe(
+  c(baseline_ctcrw1$coeff_fe()[1:3,1],0,baseline_ctcrw1$coeff_fe()[4,1],0))
+#use the indices of the variable in the map for TMB 
+#to keep the coefficients in the right order
+#otherwise we may have issues in function post_coeff
+response_ctcrw_exp$update_map(list(coeff_re=factor(rep(NA,12)),
+                             coeff_fe=factor(c(rep(NA,3),1,NA,2)),
+                             log_lambda=factor(c(NA,NA))))
+
+response_ctcrw_exp$fit()
+
+```
+
+```{r results response_ctcrw_exp}
+
+estimates_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Est")
+std_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Std")
+
+```
+
+```{r render estimates table response_ctcrw_exp}
+
+# Create data frames for each set
+log_sigma_obs <- combine_to_df(estimates_res_ctcrw_exp$log_sigma_obs, 
+                               std_res_ctcrw_exp$log_sigma_obs, 
+                               "log_sigma_obs")
+coeff_fe <- combine_to_df(estimates_res_ctcrw_exp$coeff_fe, 
+                          std_res_ctcrw_exp$coeff_fe,
+                          rownames(estimates_res_ctcrw_exp$coeff_fe))
+
+# Combine all data frames into one
+table_data_response_ctcrw_exp <- bind_rows(log_sigma_obs, coeff_fe)
+
+```
+
+```{r, results='asis', echo=FALSE}
+latex_table <- xtable(table_data_response_ctcrw_exp,
+                      caption = "Parameter Estimates and Standard Errors")
+
+# Print LaTeX table
+print(latex_table, type = "latex", include.rownames = FALSE,
+      table.placement="H")
+```
+
+We can also compute $95 \%$ CI for the coefficients in the log-linear model.
+
+```{r CI response_ctcrw_exp parameters}
+
+post_coeff_res_exp=response_ctcrw_exp$post_coeff(n_post=10000)
+post_par_res_exp=list(
+  "alpha_tau"=post_coeff_res_exp$coeff_fe[,"tau.ExpShip"],
+  "alpha_nu"=post_coeff_res_exp$coeff_fe[,"nu.ExpShip"])
+
+mean_par_res_exp=lapply(post_par_res_exp,mean)
+sd_par_res_exp=lapply(post_par_res_exp,sd)
+quant_par_res_exp <- lapply(post_par_res_exp, 
+                               quantile, probs = c(0.025, 0.975))
+
+```
+
+```{r CI table response_ctcrw_exp parameters, message=FALSE}
+
+parameter_names <- names(mean_par_res_exp)
+estimates <- mean_par_res_exp
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_par_res_exp[[param]][1], 
+          quant_par_res_exp[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = 
+    c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
+  Estimate = sprintf("$%.2f$", mean_par_res_exp),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+
+```{r,results='asis',echo=FALSE}
+xtab <- xtable(
+  table_data,
+  caption = "Response log linear estimations",
+  label = "table:response_log_linear_estimations"
+)
+
+print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
+  hline.after = c(-1, 0, nrow(table_data)),
+  table.placement="H"
+)
+```
+
+We can now plot the estimated smooth parameters.
+
+```{r plot smooth parameters response_ctcrw_exp}
+
+plot_smooth_par=response_ctcrw_exp$plot_par(var="ExpShip",
+                                            par_names=c("tau","nu"),n_post=100)
+
+#get data for the plot
+plot_data <- ggplot_build(plot_smooth_par)$data[[1]]
+#rename columns
+colnames(plot_data)[1:5]=c("mle","val","var","group","par")
+
+# ELow exposure threshold
+threshold <- 1 / 45 # 1/ max(distance to ship)
+plot_data <- plot_data %>% filter(var > threshold)
+plot_data$distance <- rep(1/plot_data[plot_data$group==1,"var"],
+                          length(unique(plot_data$group)))
+
+plot_data <- plot_data %>%
+  mutate(mle= recode(mle, "#B300001A" = "no", "#000000FF" = "yes"))
+plot_data <- plot_data %>%
+  mutate(par= recode(par, "1" = "tau", "2" = "nu"))
+
+pal <- c("no" = rgb(0.7, 0, 0, 0.1), "yes" = rgb(0, 0, 0, 1))
+
+plot_smooth_par_distance=ggplot(plot_data, aes(x = distance, group = group)) +
+  scale_colour_manual(values = pal, guide = "none") +
+  facet_wrap(c("par"), scales = "free_y",strip.position = "left") +
+                theme_minimal() + xlab("Distance to ship") + ylab(NULL) + 
+  ggtitle("Smooth parameters as a function of distance to ship") +
+  theme(strip.background = element_blank(),strip.placement = "outside",
+        strip.text = element_text(colour = "black"))+
+  geom_line(aes(y = val,color=mle))
+
+plot_smooth_par_distance
+```
+
+The results are indeed similar to what we obtain using cubic splines with a low degree of freedom.
+
+```{r plot CI response_ctcrw_exp and baseline,results='hide'}
 
 #plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
+xmin=list("ExpShip"=1/50)
+xmax=list("ExpShip"=1/3)
 link=list("ExpShip"=(\(x) 1/x))
@@
 
-plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
+response_ctcrw_exp$get_all_plots(baseline=baseline_ctcrw1,xmin=xmin,xmax=xmax,
+                                 link=link,xlabel=xlabel,show_CI="pointwise")
+
+```
+
+## Baseline uncertainty propagation
+
+In the foregoing, offset of the parameters are fixed to the baseline estimate.
+We did not propagate the uncertainty from the baseline estimate to the response model.
+To do this, we can draw values from the estimated distribution of the baseline parameters, and estimate the response for each drawn value.
+
+```{r propagate baseline uncertainty, message=FALSE,results='hide',cache=TRUE}
+
+#posterior samples of the baseline model coefficients
+post_coeff=baseline_ctcrw1$post_coeff(n_post=200)
+
+post_coeff_re=post_coeff$coeff_re
+post_coeff_fe=post_coeff$coeff_fe
+post_coeff_lambda=exp(post_coeff$log_lambda)
+post_coeff_sigma_obs=exp(post_coeff$log_sigma_obs)
+
+# Fit response for each draw of baseline coefficients
+
+formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip,nu=~ExpShip)
+
+alpha_estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+alpha_std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+
+models_list=list()
+
+for (i in 1:nrow(post_coeff_fe)) {
+
+    formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
+                 nu=~ExpShip+s(ID,bs="re"))
+
+    sigma_obs=post_coeff_sigma_obs[i,1]
+    H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+
+    response_ctcrw_exp_i<- SDE$new(formulas = formulas,data = dataAE,
+                                   type = "CTCRW",
+                                   response = c("x","y"),
+                                   par0 = par0,
+                                   other_data=list("H"=H),fixpar=c("mu1","mu2"))
+
+    response_ctcrw_exp_i$update_map(list(coeff_re=factor(rep(NA,12)),
+                             coeff_fe=factor(c(rep(NA,3),1,NA,2)),
+                             log_lambda=factor(c(NA,NA))))
+
+    new_coeff_fe=c(post_coeff_fe[i,1:3],0,post_coeff_fe[i,4],0)
+    response_ctcrw_exp_i$update_coeff_fe(new_coeff_fe)
+
+    response_ctcrw_exp_i$update_lambda(post_coeff_lambda[i,])
+
+    response_ctcrw_exp_i$update_coeff_re(post_coeff_re[i,])
+
+    response_ctcrw_exp_i$fit()
+
+    est_alpha=as.list(response_ctcrw_exp_i$tmb_rep(),
+                      what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+    std_alpha=as.list(response_ctcrw_exp_i$tmb_rep(),
+                      what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+
+    alpha_estimates[i,]=est_alpha
+    alpha_std[i,]=std_alpha
+
+    models_list[[i]]=response_ctcrw_exp_i
+}
+
+colnames(alpha_estimates)=c("tau.ExpShip","nu.ExpShip")
+colnames(alpha_std)=c("tau.ExpShip","nu.ExpShip")
+
+```
+
+We can get new and more reliable confidence intervals for the coefficients
+in the log-linear model.
+
+```{r propagate uncertainty coefficient log -linear model}
+
+mean_alpha=list(alpha_tau=mean(alpha_estimates[,1]),
+                alpha_nu=mean(alpha_estimates[,2]))
+quant_alpha<- list(
+  alpha_tau=quantile(alpha_estimates[,1],probs = c(0.025, 0.975)),
+  alpha_nu=quantile(alpha_estimates[,2],probs = c(0.025, 0.975)))
+
+```
+```{r propagated CI alpha log linear coefficients, message=FALSE}
+
+parameter_names <- names(mean_alpha)
+estimates <- mean_alpha
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1], 
+          quant_alpha[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = 
+    c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
+  Estimate = sprintf("$%.2f$", mean_alpha),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+
+```{r,results='asis',echo=FALSE}
+xtab <- xtable(
+  table_data,
+  caption = "Response log linear estimations with propagated uncertainty",
+  label = "table:response_log_linear_estimations_with_propagated_uncertainty"
+)
+
+print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  
+  hline.after = c(-1, 0, nrow(table_data)),
+  table.placement="H"
+
+)
+```
+
+```{r all CI plots,cache=TRUE}
+
+CI_plots=lapply(models_list,function(model) { 
+  model$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                      link=link,xlabel=xlabel,show_CI="pointwise")})
+
+```
+
+```{r plot smooth parameters with propagated uncertainty,results='hide'}
+
+# Extract data for each estimated smooth parameter
+tau_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+})
+nu_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+})
+
+# Get confidence intervals as quantiles of estimated parameters
+
+# Combine the x columns from all data frames into a matrix
+x_matrix_tau <- do.call(cbind, lapply(tau_CI_data, function(df) df$y))
+x_matrix_nu <- do.call(cbind, lapply(nu_CI_data, function(df) df$y))
+# Compute the 5% and 95% quantiles row-wise
+quantiles_tau <- apply(x_matrix_tau, 1, 
+                       function(row) quantile(row, probs = c(0.025, 0.975)))
+quantiles_nu <- apply(x_matrix_nu, 1,
+                      function(row) quantile(row, probs = c(0.025, 0.975)))
+
+# Transpose and format the results into a data frame
+quantiles_tau <- as.data.frame(t(quantiles_tau))
+colnames(quantiles_tau) <- c("low", "up")
+quantiles_nu <- as.data.frame(t(quantiles_nu))
+colnames(quantiles_nu) <- c("low", "up")
+
+# Add the "true" estimated smooth
+plots=response_ctcrw_exp$get_all_plots(
+  baseline=baseline_ctcrw1,xmin=xmin,xmax=xmax,
+  link=link,xlabel=xlabel,show_CI="pointwise")
+est_tau=ggplot_build(plots$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+est_nu=ggplot_build(plots$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+est_tau=cbind(est_tau,quantiles_tau)
+est_nu=cbind(est_nu,quantiles_nu)
+
+est_tau$baseline=ggplot_build(plots$fe_tau_ExpShip)$data[[2]][,"y"]
+est_nu$baseline=ggplot_build(plots$fe_nu_ExpShip)$data[[2]][,"y"]
+
+plot_tau_response_exp <- ggplot(data = est_tau) +
+  geom_line(aes(x = x, y = y), color = "steelblue", size = 1) +
+  geom_line(aes(x=x,y=baseline),color="steelblue",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "lightblue", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = "Value",
+    title = "Smooth Tau Estimates with propagated uncertainty"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = "none",
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
+  )
+
+# Enhanced plot for nu
+plot_nu_response_exp <- ggplot(data = est_nu) +
+  geom_line(aes(x = x, y = y), color = "darkorange", size = 1) +
+   geom_line(aes(x,baseline),color="darkorange",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "navajowhite", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = "Value",
+    title = "Smooth Nu Estimates with propagated uncertainty"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = "none",
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
+  )
+
 ```
@@
 ```{r}
-
-post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
-post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
-mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
-sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)
-
-post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
-post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
-mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
-sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)
-
-post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
-post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
-mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
-sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)
-
-```
-
-### Response model with splines of ExpShip based on baseline3
-
-```{r fit response3}
-
-#initialize parameters
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-#Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_50m$lambda()["tau.s(ID)",1],1,baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
-response3_sp_50m$update_coeff_re(new_coeff_re)
-response3_sp_50m$update_coeff_fe(baseline3_50m$coeff_fe()[,1])
-
-response3_sp_50m$update_lambda(new_lambda)
-response3_sp_50m$fit()
-
-#Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_75m$lambda()["tau.s(ID)",1],1,baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
-response3_sp_75m$update_coeff_re(new_coeff_re)
-response3_sp_75m$update_coeff_fe(baseline3_75m$coeff_fe()[,1])
-
-response3_sp_75m$update_lambda(new_lambda)
-response3_sp_75m$fit()
-
-#Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4)))))
-
-new_coeff_re=c(rep(0,2),baseline3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline3_100m$lambda()["tau.s(ID)",1],1,baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
-response3_sp_100m$update_coeff_re(new_coeff_re)
-response3_sp_100m$update_coeff_fe(baseline3_100m$coeff_fe()[,1])
-
-response3_sp_100m$update_lambda(new_lambda)
-response3_sp_100m$fit()
-```
-
-```{r get results response3}
-estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
-std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")
-
-estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
-std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")
-
-estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
-std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_sp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_sp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_sp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
-
-```{r estimates on parameter scale response3}
-
-post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
-post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
-mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
-sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)
-
-post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
-post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
-mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
-sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)
-
-post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
-post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
-mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
-sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)
-
-```
-
-### Response model with exponential of ExpShip based on baseline2
-
-```{r}
-
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
-                 nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-response2_exp_50m$update_coeff_re(new_coeff_re)
-response2_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_50m$update_lambda(new_lambda)
-
-response2_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-response2_exp_75m$update_coeff_re(new_coeff_re)
-response2_exp_75m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_75m$update_lambda(new_lambda)
-
-response2_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-response2_exp_100m$update_coeff_re(new_coeff_re)
-response2_exp_100m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_100m$update_lambda(new_lambda)
-
-response2_exp_100m$fit()
-```
-
-```{r}
-estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
-std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")
-
-estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
-std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")
-
-estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
-std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res2_exp=response2_exp_50m$get_all_plots(baseline=baseline2_50m,model_name="response2_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,model_name="response2_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,model_name="response2_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
-
-### Response model with exponential of ExpShip based on baseline3
-
-```{r}
-
-par0=c(0,0,1,4,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
-                 nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_50m$coeff_fe()[1:3,1],0,baseline3_50m$coeff_fe()[4,1],0,baseline3_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_50m$lambda()["tau.s(ID)",1],baseline3_50m$lambda()["nu.s(ID)",1],baseline3_50m$lambda()[3:6,1])
-response3_exp_50m$update_coeff_re(new_coeff_re)
-response3_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_50m$update_lambda(new_lambda)
-response3_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_75m$coeff_fe()[1:3,1],0,baseline3_75m$coeff_fe()[4,1],0,baseline3_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_75m$lambda()["tau.s(ID)",1],baseline3_75m$lambda()["nu.s(ID)",1],baseline3_75m$lambda()[3:6,1])
-response3_exp_75m$update_coeff_re(new_coeff_re)
-response3_exp_75m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_75m$update_lambda(new_lambda)
-response3_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                    map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
-
-new_coeff_fe=c(baseline3_100m$coeff_fe()[1:3,1],0,baseline3_100m$coeff_fe()[4,1],0,baseline3_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline3_100m$lambda()["tau.s(ID)",1],baseline3_100m$lambda()["nu.s(ID)",1],baseline3_100m$lambda()[3:6,1])
-response3_exp_100m$update_coeff_re(new_coeff_re)
-response3_exp_100m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_100m$update_lambda(new_lambda)
-response3_exp_100m$fit()
-```
-
-```{r}
-estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
-std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")
-
-estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
-std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")
-
-estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
-std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline3_50m,model_name="response3_exp_50m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline3_75m,model_name="response3_exp_75m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline3_100m,model_name="response3_exp_100m",xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)
-
-```
+plot_tau_response_exp
+plot_nu_response_exp
+
+```
+
+# Simulation from the baseline and response models
+
+We know want to investigate how the fitted model match the actual observations.
+To fo this, we simulate trajectories from the fitted model before and during exposure.
+First, we ignore that the narwhals move in a restricted area and only look at the shape of the trajectories.
+
+## Without land constraints
+
+```{r simulate baseline_ctcrw1 trajectories}
+#initial narwhals positions (convert to matrix to avoid bugs)
+z0_BE <- as.matrix(dataBE12[!duplicated(dataBE12$ID), c("x", "y")])
+
+sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
+
+baseline_ctcrw_sim=baseline_ctcrw1$simulate(z0 = z0_BE,
+                                            data=baseline_ctcrw1$data(),
+                                            sd_noise=sigma_obs)
+# Add a label to each dataset for plotting
+baseline_ctcrw_sim <- baseline_ctcrw_sim %>% mutate(data_type = "Simulated")
+dataBE12 <- dataBE12 %>% mutate(data_type = "Real")
+
+# Combine both datasets
+baseline_trajectories <- bind_rows(baseline_ctcrw_sim, dataBE12)
+
+# Plot both in a single plot with facets
+ggplot(baseline_trajectories) +
+  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
+  facet_wrap(~ data_type) + 
+  ggtitle("Trajectories before ship exposure") +
+  theme_minimal() +
+  labs(color = "Narwhal ID")
+
+```
+
+```{r simulate response trajectories}
+
+#initial narwhals positions (convert to matrix to avoid bugs)
+z0_AE <- as.matrix(dataAE[!duplicated(dataAE$ID), c("x", "y")])
+
+response_ctcrw_sim=response_ctcrw_exp$simulate(z0 = z0_AE,
+                                               data=response_ctcrw_exp$data(),
+                                               sd_noise=sigma_obs)
+# Add a label to each dataset for plotting
+response_ctcrw_sim <- response_ctcrw_sim %>% mutate(data_type = "Simulated")
+dataAE <- dataAE %>% mutate(data_type = "Real")
+
+# Combine both datasets
+response_trajectories <- bind_rows(response_ctcrw_sim, dataAE)
+
+# Plot both in a single plot with facets
+ggplot(response_trajectories) +
+  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
+  facet_wrap(~ data_type) + 
+  ggtitle("Trajectories during ship exposure") +
+  theme_minimal() +
+  labs(color = "Narwhal ID")
+
+```
+
+The simulated trajectories appear less persistent than the observed ones. We probably miss something in the model.
+
+```{r animate response trajectories, results=FALSE,eval=interactive()}
+
+# Basic ggplot setup
+plot_trajectories <- ggplot(
+  response_trajectories, aes(x = x, y = y, color =ExpShip,group=ID)) +
+  geom_path(size = 1, lineend = "round",alpha=0.1) +     
+  geom_point(aes(shape=ID),size =4 ) +                       
+  scale_color_gradient(low = "blue", high = "red") +  # 
+  labs(title = "Narwhal Trajectories with Ship Exposure", 
+       subtitle = "Time : {frame_along} h", 
+       x = "X Position", 
+       y = "Y Position") +
+  theme_minimal() +
+  facet_wrap(~ data_type)
+
+# Animation setup with transition_reveal for one-point-at-a-time appearance
+anim <- plot_trajectories + 
+  transition_reveal(along = time) + 
+  labs(subtitle = "Time: {round(frame_along,1)} hours")
+
+# Save the animation
+animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
+        renderer = gifski_renderer("response_trajectories.gif"))
+
+```
+
+<!-- ![Narwhal Trajectories](response_trajectories.gif) -->
+
+The baseline model seems to match better the observations than the response model.
+In both cases, the simulated trajectories appear less persistent than the actual observations.
+Indeed, the persistence in the movement is linked to the highly constrained environment in which the narwhals move.
+We want to add this effect of the shoreline on the movement.
+
+### With land reflection
+
+To illustrate, we simulate trajectories of the fitted model and project the points that reach land on the shoreline.
+This is equivalent to simulating from the reflected version of the fitted SDE, that is the fitted SDE with an additional term with bounded variations that keeps the process within the water domain.
+
+```{r simulate baseline trajectories with land reflection,cache=TRUE,message=FALSE}
+
+# Set the path to the directory containing the data
+par_dir=here()
+greenland_data_path <- file.path(par_dir,"Data", 
+                                 "preprocessed_data","greenland")
+
+#get the coastline geometry from the geojson file
+land<-st_read(file.path(greenland_data_path,
+                        "updated_scoresby_sound_utm.shp"))
+land <- st_transform(land, crs = "+init=EPSG:32626 +units=km")
+
+baseline_ctcrw_sim_land=baseline_ctcrw1$simulate(z0 = z0_BE,
+                                            data=baseline_ctcrw1$data(),
+                                            sd_noise=sigma_obs,land=land)
+# Add a label to each dataset for plotting
+baseline_ctcrw_sim_land <- baseline_ctcrw_sim_land %>% 
+  mutate(data_type = "Simulated with reflection on the boudary")
+dataBE12 <- dataBE12 %>% mutate(data_type = "Real")
+
+# Combine both datasets
+baseline_trajectories_land <- bind_rows(baseline_ctcrw_sim_land, dataBE12)
+
+# Plot both in a single plot with facets
+ggplot(baseline_trajectories_land) +geom_sf(data=land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
+  facet_wrap(~ data_type) + 
+  ggtitle("Trajectories before ship exposure") +
+  theme_minimal() +
+  labs(color = "Narwhal ID")
+
+```
+
+```{r simulate response trajectories with land reflection,cache=TRUE,message=FALSE}
+
+response_ctcrw_sim_land=response_ctcrw_exp$simulate(z0 = z0_AE,
+                                            data=response_ctcrw_exp$data(),
+                                            sd_noise=sigma_obs,land=land)
+# Add a label to each dataset for plotting
+response_ctcrw_sim_land <- response_ctcrw_sim_land %>% 
+  mutate(data_type = "Simulated with reflection on the boudary")
+dataAE <- dataAE %>% mutate(data_type = "Real")
+
+# Combine both datasets
+response_trajectories_land <- bind_rows(response_ctcrw_sim_land, dataAE)
+
+# Plot both in a single plot with facets
+ggplot(response_trajectories_land) +geom_sf(data=land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
+  facet_wrap(~ data_type) + 
+  ggtitle("Trajectories during ship exposure") +
+  theme_minimal() +
+  labs(color = "Narwhal ID")
+
+```
+
+```{r animate response trajectories with land reflection, results=FALSE,eval=interactive()}
+
+# Basic ggplot setup
+plot_trajectories_land <- ggplot() +
+  geom_sf(data=land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=response_trajectories_land, 
+            aes(x = x, y = y, color =ExpShip,group=ID),size = 1, 
+            lineend = "round",alpha=0.15) +     
+  geom_point(data=response_trajectories_land,
+             aes(x = x, y = y, color =ExpShip,group=ID,shape=ID),size =4 ) +  
+  scale_color_gradient(low = "blue", high = "red") +
+  labs(title = "Narwhal Trajectories with Ship Exposure", 
+       subtitle = "Time : {frame_along} h", 
+       x = "X Position", 
+       y = "Y Position") +
+  theme_minimal() +
+  facet_wrap(~ data_type)
+
+anim <- plot_trajectories_land + 
+  transition_reveal(along = time) + 
+  labs(subtitle = "Time: {round(frame_along,1)} hours")
+
+# Save the animation
+animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
+        renderer = gifski_renderer("response_trajectories_land.gif"))
+
+```
+
+<!-- ![Narwhal Trajectories](response_trajectories_land.gif)  -->
+
+In the real data, the trajectories seem to follow more the shoreline and be more persistent than in the simulated data.
+To better integrate this phenomenon in the model, we will need to use new covariates.

### 2025-01-07 R/application/RCVM/fit_rcvm.Rmd
R code updates

---
title: "Final analysis of Greenland narwhals trajectories with smoothSDE"
author: "Alexandre Delporte"
date: "2024-06-07"
output: html_document
---

In this analysis, we introduce covariates \texttt{AngleNormal} and \texttt{Ishore} ($\Theta$ and $I_{shore}$) in the SDE models for the narwhal data. This allows to better understand how the fjords shoreline influences the narwhals movement in the baseline model.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

```{r packages,message=FALSE}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate) # animated trajectories
```

```{r get data,message=FALSE}

# Set the path to the directory containing the data
par_dir=here() #parent directory 
narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")

#data before exposure with first 24h removed
dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

#data after exposure
dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"), header = TRUE,dec = ".")

```

We choose thresholds on the distance to the shore to define the Ishore covariate.

```{r preprocess data}

D_low=0.073
D_up=3
dataBE24[dataBE24$DistanceShore> D_up,"Ishore"]=0
dataBE24[dataBE24$DistanceShore< D_low,"Ishore"]=1/D_low

dataBE12[dataBE12$DistanceShore> D_up,"Ishore"]=0
dataBE12[dataBE12$DistanceShore< D_low,"Ishore"]=1/D_low

dataAE[dataAE$DistanceShore> D_up,"Ishore"]=0
dataAE[dataAE$DistanceShore< D_low,"Ishore"]=1/D_low

N=length(unique(dataBE12$ID))

n_pre12=length(dataBE12$time) #number of observations before exposure
n_pre24=length(dataBE24$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

# Fit baseline models

We try sde models where the parameters are smooth function of the covariates AngleNormal and Ishore.
Here, for simplicity we will always fix the measurement error to $30$ m. This choice is based on the paper of Weensven.
We will assess the sensitivity of our results to the hyperparameter later.

## Baseline CTCRW with AngleNormal covariate in $\tau$

The first model is a CTCRW where the covariate AngleNormal influences the persistence of the movement.

```{r fit baseline_ctcrw0,message=FALSE}

#initial parameters
par0 <- c(0,0,1,4)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

```

```{r results baseline_ctcrw0}

estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")

```

```{r render estimates table baseline_ctcrw0}

# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe,
                          std_bas_ctcrw0$coeff_fe, 
                          rownames(estimates_bas_ctcrw0$coeff_fe))

coeff_re <- combine_to_df(estimates_bas_ctcrw0$coeff_re,
                          std_bas_ctcrw0$coeff_re, 
                          rownames(estimates_bas_ctcrw0$coeff_re))

log_lambda <- combine_to_df(estimates_bas_ctcrw0$log_lambda,
                          std_bas_ctcrw0$log_lambda, 
                          rownames(estimates_bas_ctcrw0$log_lambda))

# Combine all data frames into one
table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)

```

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```
We can get confidence intervals for the intercept of each parameter and the random effects standard deviations.

```{r CI baseline_ctcrw0 parameters}

post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
post_par_bas_ctcrw0=list(
  "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"nu.s(ID)"])))

mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
                               quantile, probs = c(0.025, 0.975))

```

```{r CI table baseline_ctcrw0 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw0)
estimates <- mean_par_bas_ctcrw0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
          quant_par_bas_ctcrw0[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "CTCRW baseline estimations with random effects",
  label = "table:ctcrw_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

We can now plot the smooth parameter $\tau$ as a function of the covariate \texttt{AngleNormal}.

```{r plot smooth parameters baseline_ctcrw0 }
baseline_ctcrw0$plot_par(var="AngleNormal",
                         par_names=c("tau","nu"),n_post=100)

```

```{r plot smooth parameters CI baseline_ctcrw0}
baseline_ctcrw0$get_all_plots(baseline=NULL,show_CI="simultaneous")
```

The movement is clearly more persistent along the shoreline, meaning that $\tau$ increases when \texttt{AngleNormal} is close to $\pm \frac{\pi}{2}$).

## Baseline racvm with AngleNormal covariate in tau and omega

We now decide to introduce a rotational parameter $\omega$ in the SDE to allow the velocity to rotate depending on the narwhals position with respect to the shoreline.
We start by introducing the covariate \texttt{AngleNormal} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.

```{r fit baseline_racvm0,cache=TRUE}

#initial parameters
par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=10,bs="cs"))

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

baseline_racvm0$fit()

```

```{r results baseline_racvm0}

estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")
```

```{r render estimates table baseline_racvm0 }

# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe, 
                          std_bas_racvm0$coeff_fe, 
                          rownames(estimates_bas_racvm0$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm0$coeff_re, 
                          std_bas_racvm0$coeff_re, 
                          rownames(estimates_bas_racvm0$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm0$log_lambda, 
                          std_bas_racvm0$log_lambda, 
                          rownames(estimates_bas_racvm0$log_lambda))

# Combine all data frames into one
table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)

```

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```

```{r CI baseline_racvm0 parameters}

post_coeff_bas_racvm0=baseline_racvm0$post_coeff(n_post=10000)
post_par_bas_racvm0=list(
  "tau"=exp(post_coeff_bas_racvm0$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm0$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm0$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm0=lapply(post_par_bas_racvm0,mean)
sd_par_bas_racvm0=lapply(post_par_bas_racvm0,sd)
quant_par_bas_racvm0 <- lapply(post_par_bas_racvm0, 
                               quantile, probs = c(0.025, 0.975))

```

```{r CI table baseline_racvm0 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm0)
estimates <- mean_par_bas_racvm0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm0[[param]][1], 
          quant_par_bas_racvm0[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm0 baseline estimations",
  label = "table:racvm0_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

```{r plot smooth parameters baseline_racvm0}
baseline_racvm0$plot_par(var="AngleNormal",
                         par_names=c("omega","tau"),n_post=100)

```

$\omega$ close to $\pm \pi$ indicates movement toward the shoreline. Here, it seems that the narwhals have a slight tendency to rotate, presumably to move along the shoreline or change direction, when they are heading toward the shore. However, it is not very clear since the confidence intervals are wide.

```{r plot smooth parameter CI baseline_racvm0}
baseline_racvm0$get_all_plots(baseline=NULL,show_CI="pointwise")
```

## Baseline with AngleNormal and Ishore in omega and tau

We add the effect of the distance to the shore through the covariate \texttt{Ishore}.
This model has a lot of parameters and takes around 15 minutes to fit. 
Indeed, we will need to fix the penalties in the tensor splines in advance to ensure convergence of the optimizer. The standard deviation of the random effects and the smoothing penalties for the splines in $\tau$ will still be estimated from the data and not fixed. To illustrate this we first fit a model where the smoothing penalties are all estimated from the data.

```{r fit baseline_racvm1_unstable,cache=TRUE,message=FALSE}

#initial parameters
par0 <- c(0,0,1,4,0)

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1_unstable<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

baseline_racvm1_unstable$fit()

```

```{r get results baseline_racvm1_unstable,message=FALSE}

estimates_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Est")
std_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Std")

estimates_bas_racvm1_unstable

```
```{r std random effects baseline_racvm1_unstable}

baseline_racvm1_unstable$sdev()

```

We see that the estimate for the standard deviation $\sigma_\tau$ of the random effect for $\tau$ is very high while the estimate for the standard deviation $\sigma_\nu$ of the random effect for $\nu$ is very close to $0$. The smoothing penalties estimates are extremely high (the smoothing penalties are the exponential of \texttt{estimates\_bas\_racvm1\_unstable$log\_lambda}). This suggests that the optimization did not go well and numerical errors occurred. For this reason, we decide to fix the smoothing penalties to "reasonable" values in the tensor splines.

```{r fit baseline_racvm1,cache=TRUE}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=exp(c(3,3,4,-2,-2,-2,-2))

#mapping for fixed smoothing penalties
new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,4))))

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

baseline_racvm1$update_lambda(init_lambda)
baseline_racvm1$update_map(new_map)
baseline_racvm1$fit()

```

```{r render estimates table baseline_racvm1 }

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm1$log_sigma_obs, 
                               std_bas_racvm1$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm1$coeff_fe, 
                          std_bas_racvm1$coeff_fe, 
                          rownames(estimates_bas_racvm1$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm1$coeff_re, 
                          std_bas_racvm1$coeff_re, 
                          rownames(estimates_bas_racvm1$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm1$log_lambda, 
                          std_bas_racvm1$log_lambda, 
                          rownames(estimates_bas_racvm1$log_lambda))

# Combine all data frames into one
table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)

```

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```

```{r CI baseline_racvm1 parameters}

post_coeff_bas_racvm1=baseline_racvm1$post_coeff(n_post=10000)
post_par_bas_racvm1=list(
  "tau"=exp(post_coeff_bas_racvm1$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm1$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm1$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm1=lapply(post_par_bas_racvm1,mean)
sd_par_bas_racvm1=lapply(post_par_bas_racvm1,sd)
quant_par_bas_racvm1 <- lapply(post_par_bas_racvm1, 
                               quantile, probs = c(0.025, 0.975))

```

```{r CI table baseline_racvm1 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm1)
estimates <- mean_par_bas_racvm1
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm1[[param]][1], 
          quant_par_bas_racvm1[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm1),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm1 baseline estimations",
  label = "table:racvm1_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```
```{r display plots baseline_racvm1}

Ishore_values=c(1/0.25,1/2,1/10)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm1$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

```

```{r plot smooth parameters CI baseline_racvm1}

#plot parameters

xmin=list("Ishore"=1/D_up)
xmax=list("Ishore"=1/D_low)
link=list("Ishore"=(\(x) 1/x))
xlabel=list("Ishore"="Distance to shore")

plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_bas_racvm1

```

Now, we would like to demonstrate how the Distance to shore can influence the smooth parameter $\omega$.
We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.

```{r smooth omega baseline_racvm1}
plot1=plots_bas_racvm1$fe_omega_AngleNormal_q1_Ishore
plot2=plots_bas_racvm1$fe_omega_AngleNormal_q2_Ishore
plot4=plots_bas_racvm1$fe_omega_AngleNormal_q4_Ishore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Far"
omega2$Category <- "Medium"
omega4$Category <- "Close"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Far"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Close"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
  geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
  labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
  geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size=14),
    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
    axis.text.y = element_text(size=12),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    legend.title = element_text(size=14),
    legend.text = element_text(size=12))

ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline_racvm1", width=10, height=6, units="in")

combined_plot

```

There seems to be a rotation close to the shore when the angle $\Theta \in [-\pi,-\frac{\pi}{2}] \cup [\frac{\pi}{2},\pi]$, which means movement toward the shore. However, again the confidence intervals are too wide to really conclude.
Moreover, we fixed the smoothing penalties in the tensor splines. Changing their values can influence the shape of the estimated functions : lower values will produce functions that are more wiggily in each direction.

This model in indeed very complex and involves a large number of parameters given the amount of data points we have for the trajectories before exposure.
We can try a more parsimonious model to make estimation easier.

## Baseline with AngleNormal and Ishore in omega

We decide to remove the effect of \texttt{AngleNormal} on $\tau$ and only put covariates on $\omega$.

```{r fit baseline_racvm2,cache=TRUE}

#initial parameters
par0 <- c(0,0,2,4,0)
init_lambda=c(exp(4),exp(4),10^-6,10^-1)

#mapping for fixed smoothing penalties
new_map=list("log_lambda"=factor(c(1,2,rep(NA,2))))

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(AngleNormal,DistanceShore,k=5,bs="cs"))

# Fit baseline
baseline_racvm2<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

baseline_racvm2$update_lambda(init_lambda)
baseline_racvm2$update_map(new_map)
baseline_racvm2$fit()

```

```{r get results baseline_racvm2}

estimates_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Est")
std_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Std")

```

```{r render estimates table baseline_racvm2}
# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm2$log_sigma_obs, 
                               std_bas_racvm2$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm2$coeff_fe, 
                          std_bas_racvm2$coeff_fe, 
                          rownames(estimates_bas_racvm2$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm2$coeff_re, 
                          std_bas_racvm2$coeff_re, 
                          rownames(estimates_bas_racvm2$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm2$log_lambda, 
                          std_bas_racvm2$log_lambda, 
                          rownames(estimates_bas_racvm2$log_lambda))

# Combine all data frames into one
table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)

```

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```
```{r CI baseline_racvm2 parameters}

post_coeff_bas_racvm2=baseline_racvm2$post_coeff(n_post=10000)
post_par_bas_racvm2=list(
  "tau"=exp(post_coeff_bas_racvm2$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm2$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm2$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm2=lapply(post_par_bas_racvm2,mean)
sd_par_bas_racvm2=lapply(post_par_bas_racvm2,sd)
quant_par_bas_racvm2 <- lapply(post_par_bas_racvm2, 
                               quantile, probs = c(0.025, 0.975))

```

```{r CI table baseline_racvm2 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm2)
estimates <- mean_par_bas_racvm2
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm2[[param]][1], 
          quant_par_bas_racvm2[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm2),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm2 baseline estimations",
  label = "table:racvm2_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```
```{r display plots baseline_racvm2}

Ishore_values=c(1/0.25,1/2,1/5)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm2$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

```

```{r plot smooth parameters CI baseline_racvm2}
#plot parameters

xmin=list("Ishore"=1/D_up)
xmax=list("Ishore"=1/D_low)
link=list("Ishore"=(\(x) 1/x))
xlabel=list("Ishore"="Distance to shore")

plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_bas_racvm2

```

```{r save plots baseline_racvm2}
# Create directory
if (!dir.exists("baseline_racvm2")) {
    dir.create("baseline_racvm2", recursive = TRUE)
}

# Use lapply to save each plot
lapply(seq_along(plots_bas2), function(i) {

    plot<-plots_bas2[[i]]
    plot_name <- names(plots_bas2)[i]
    file_path <- file.path("baseline_racvm2", plot_name)

    # Check if the plot is a ggplot or plotly object
    if (inherits(plot, "ggplot")) {
      ggsave(file_path, plot, device = "pdf")
    } else if (inherits(plot, "plotly")) {
      saveWidget(plot, file_path)
  }
    })
```

```{r smooth omega baseline_racvm2}
plot1=plots_bas_racvm2$fe_omega_AngleNormal_q1_Ishore
plot2=plots_bas_racvm2$fe_omega_AngleNormal_q2_Ishore
plot4=plots_bas_racvm2$fe_omega_AngleNormal_q4_Ishore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Far"
omega2$Category <- "Medium"
omega4$Category <- "Close"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Far"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Close"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
  geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
  labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
  geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size=14),
    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
    axis.text.y = element_text(size=12),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    legend.title = element_text(size=14),
    legend.text = element_text(size=12))

ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline_racvm2", width=10, height=6, units="in")

combined_plot

```

Again, even though there seem to be a rotation in the movement close to the shore when the angle $\Theta \in [-\pi,-\frac{\pi}{2}] \cup [\frac{\pi}{2},\pi]$, the confidence intervals are too wide to conclude.
Moreover, the magnitude of the rotation is too low to effectively constrained the movement : it is estimated to be at most around $\pi$ rad/h, while the value we used in the simulation study to obtain constrained trajectories is $\frac{\pi}{2}$ rad/min, that is approximately $30 \pi $ rad/h.

## Simulate trajectories before exposure

```{r simulate baseline trajectories}
#initial narwhals positions (convert to matrix to avoid bugs)
z0_BE <- as.matrix(dataBE12[!duplicated(dataBE12$ID), c("x", "y")])

sigma_obs=0.03

baseline_ctcrw0_sim=baseline_ctcrw0$simulate(z0 = z0_BE,
                                            data=baseline_ctcrw0$data(),
                                            sd_noise=sigma_obs)

baseline_racvm0_sim=baseline_racvm0$simulate(z0 = z0_BE,
                                            data=baseline_racvm0$data(),
                                            sd_noise=sigma_obs)

baseline_racvm1_sim=baseline_racvm1$simulate(z0 = z0_BE,
                                            data=baseline_racvm1$data(),
                                            sd_noise=sigma_obs)

baseline_racvm2_sim=baseline_racvm2$simulate(z0 = z0_BE,
                                            data=baseline_racvm2$data(),
                                            sd_noise=sigma_obs)
# Add a label to each dataset for plotting
baseline_ctcrw0_sim <- baseline_ctcrw0_sim %>% 
  mutate(data_type = "Simulated baseline_ctcrw0")
baseline_racvm0_sim <- baseline_racvm0_sim %>% 
  mutate(data_type = "Simulated baseline_racvm0")
baseline_racvm1_sim <- baseline_racvm1_sim %>% 
  mutate(data_type = "Simulated baseline_racvm1")
baseline_racvm2_sim <- baseline_racvm2_sim %>% 
  mutate(data_type = "Simulated baseline_racvm2")
dataBE12 <- dataBE12 %>% mutate(data_type = "Observed")

# Combine both datasets
baseline_trajectories <- bind_rows(
  baseline_ctcrw0_sim,baseline_racvm0_sim,
  baseline_racvm1_sim,baseline_racvm2_sim, dataBE12)

# Plot both in a single plot with facets
ggplot(baseline_trajectories) +
  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
  facet_wrap(~ data_type) + 
  ggtitle("Trajectories before ship exposure") +
  theme_minimal() +
  labs(color = "Narwhal ID")

```

As in the preliminary analysis, the simulated trajectories still appear more wiggily or tortuous than the observed trajectories.
In comparison, look at what type of simulated trajectories we obtain when fixing a higher value for the persistence.

```{r simulate persistent trajectory,cache=TRUE}
#initial parameters
par0 <- c(0,0,3,4.6)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1)

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw_fixed<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2","tau","nu"))

baseline_ctcrw_fixed_sim=baseline_ctcrw_fixed$simulate(z0 = z0_BE,
                                            data=baseline_ctcrw_fixed$data(),
                                            sd_noise=sigma_obs)

baseline_ctcrw_fixed_sim <- baseline_ctcrw_fixed_sim %>% 
  mutate(data_type = "Simulated baseline_ctcrw_fixed")

# Combine both datasets
baseline_trajectories <- bind_rows(baseline_trajectories,
 baseline_ctcrw_fixed_sim)

# Plot both in a single plot with facets
ggplot(baseline_trajectories) +
  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
  facet_wrap(~ data_type) + 
  ggtitle("Trajectories before ship exposure") +
  theme_minimal() +
  labs(color = "Narwhal ID")

```
The simple ctcrw model where we fixed all parameters to some value based on observations seems to produce trajectories that match better the observations than the complicated models that we fitted. However, here we simulated with the covariates values that are in the data. But the angle and distance to shore should be recomputed as the process evolves. We now try to simulate and compute the new value of the covariates \texttt{AngleNormal} and \texttt{Ishore} at each step. We use high frequency time step, and not the low frequency times steps in the observed data to have accurate trajectories.

```{r,cache=TRUE}

# Get land polygons
par_dir=here()
greenland_data_path <- file.path(par_dir,"Data", 
                                 "preprocessed_data","greenland")  
land<-st_read(file.path(greenland_data_path,
                        "updated_scoresby_sound_utm.shp"))
land <- st_transform(land, crs = "+init=EPSG:32626 +units=km")

# Define functions to compute covariates along the way
fangle=function(z,v,p) {

  #normal vector
  normal=c(z[1]-p[1],z[2]-p[2])

  #return angle
  return (signed_angle(normal,v))
}

fIshore=function(z,v,p) {

  #normal vector
  normal=c(z[1]-p[1],z[2]-p[2])

  #distance to shore
  Dshore=sqrt(normal[1]^2+normal[2]^2)
  if (Dshore>3){
    return (0)
  }
  else if (Dshore<0.02) {
    return (50)
  }
  else {
    return (1/Dshore)
  }
}

atw=list("AngleNormal"=fangle,"Ishore"=fIshore)

# Define time steps
delta=1/60
times=unlist(sapply(unique(dataBE12$ID),function(ID) {seq(min(dataBE12[dataBE12$ID==ID,"time"]),max(dataBE12[dataBE12$ID==ID,"time"]),by=delta)}))
n=length(times)
N=length(unique(dataBE12$ID))
data_reg=data.frame("AngleNormal"=rep(0,n*N),
                    "Ishore"=rep(1,n*N),time=c(),ID=rep(unique(dataBE12$ID),each=n))

# Simulate trajectories with small time step
baseline_racvm1_trajectory=baseline_racvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 1)
baseline_racvm1_trajectory_adjusted=baseline_racvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 40)

baseline_racvm2_trajectory=baseline_racvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 1)
baseline_racvm2_trajectory_adjusted=baseline_racvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 40)

baseline_racvm1_trajectory <- baseline_racvm1_trajectory %>% 
  mutate(data_type = "Simulated baseline_racvm1")
baseline_racvm1_trajectory_adjusted <- baseline_racvm1_trajectory_adjusted %>% 
  mutate(data_type = "Simulated baseline_racvm1 adjusted")
baseline_racvm2_trajectory <- baseline_racvm2_trajectory %>% 
  mutate(data_type = "Simulated baseline_racvm2")
baseline_racvm2_trajectory_adjusted <- baseline_racvm2_trajectory_adjusted %>% 
  mutate(data_type = "Simulated baseline_racvm2 adjusted")

baseline_trajectories <- bind_rows(baseline_racvm1_trajectory,
                                   baseline_racvm1_trajectory_adjusted,
                                   baseline_racvm2_trajectory,
                                   baseline_racvm2_trajectory_adjusted,dataBE12)

```

```{r}

# Basic ggplot setup
plot_baseline_trajectories <- ggplot() +
  geom_sf(data=land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=baseline_trajectories, 
            aes(x = x, y = y,group=ID),size = 1, 
            lineend = "round",alpha=0.15) +     
  geom_point(data=baseline_trajectories,
             aes(x = x, y = y,group=ID,shape=ID),size =4 ) + facet_wrap(~data_type) 
  labs(title = "Narwhal Trajectories before Exposure", 
       subtitle = "Time : {frame_along} h", 
       x = "X Position", 
       y = "Y Position") +
  theme_minimal()

anim <- plot_baseline_trajectories + 
  transition_reveal(along = time) + 
  labs(subtitle = "Time: {round(frame_along,1)} hours")

# Save the animation
animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
        renderer = gifski_renderer("baseline_trajectories.gif"))

```

The adjusted SDE models where the smooth parameters $\omega$ has been multiplied by a factor $40$ produces trajectories that seem to follow the shoreline and look more similar to the actual data.
However, the trajectories obtained from the non adjusted fitted model does not produce constrained trajectories.
This suggests that we are indeed underestimating the smooth parameter $\omega$ with the tensor splines. This may be due to several factors : the low frequency at which the data is observed, the measurement error, the small amount of data that we have.

We can try to fit an interpretable parametric surface on the estimated spline surface for the parameter $\omega$ in the models \texttt{baseline\_racvm1} or \texttt{baseline\_racvm2}. However the surfaces are so different that it does not work very well.

```{r}

Ishore=seq(from=1/6,to=1/0.05,length.out=30)
theta=seq(from=-pi,to=pi,length.out=30)
grid<- as.data.frame(expand.grid(Ishore,theta))
colnames(grid) <- c("Ishore","AngleNormal")
grid$ID="Asgeir"

#get model matrices
mats=baseline_racvm1$make_mat(new_data=grid)
X_fe=mats$X_fe
X_re=mats$X_re

par_mat=baseline_racvm1$par(new_data=grid,X_fe=X_fe,X_re=X_re)

#matrix of values for surface plot
z=matrix(par_mat[,"omega"],30,30)

# Define the parametric function for omega
fomega<- function(cov_data, a =5,theta0=pi/2,D0=1) {
  Dshore=cov_data$DistanceShore
  theta=cov_data$AngleNormal
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }
  a*theta*(theta-theta0)*(theta+theta0)*exp(-Dshore/D0)/Dshore
}

data=grid
data$z=as.vector(z)

#fix magnitude and slope for omega on the AngleNormal axis
a_fixed <- 1  
theta0_fixed <- pi/2

fit_optim <- optim(par = c(log_D0 = log(1)),  # Exclude a from parameters
                   fn = function(params) {
                     log_D0 <- params[1]

                     # Use fixed values
                     a <- a_fixed
                     theta0 <- theta0_fixed

                      # Compute predicted z values
                     z_pred <- fomega(data, a,theta0,exp(log_D0))

                     # Return sum of squared residuals
                     sum((data$z - z_pred)^2)
                   })

est_par=fit_optim$par

# Compute the predicted values using the starting parameters
z_pred_starting <- fomega(grid,D0=exp(est_par))

# Reshape z_pred_starting to a matrix for plotting
z_pred_matrix <- matrix(z_pred_starting, nrow = length(Ishore), ncol = length(theta))

# Create the 3D surface plot using plot_ly
plot_ly(x = ~theta, y = ~1/Ishore, z = ~z_pred_matrix) %>%
  add_surface() %>%
  layout(title = "Parametric surface for omega (starting parameters)",
         scene = list(
           yaxis = list(title = "1/Ishore"),
           xaxis = list(title = "AngleNormal"),
           zaxis = list(title = "Predicted z")
         ))

# Create the 3D surface plot using plot_ly
plot_ly(x = ~theta, y = ~1/Ishore, z = ~z) %>%
  add_surface() %>%
  layout(title = "Predicted spline surface for omega",
         scene = list(
           yaxis = list(title = "Ishore"),
           xaxis = list(title = "AngleNormal"),
         ))

```

We continue the analysis with the fitted baseline models \texttt {baseline\_racvm1} and \texttt{baseline\_racvm2}.

# Fit response models

## Try a model similar to baseline racvm2

```{r}

par0 <- c(0,0,1,4,0)

sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~te(AngleNormal,Ishore,k=5,bs="cs"))

response0<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

response0$update_map(list("log_lambda"=factor(c(1,2,NA,NA))))
response0$update_lambda(exp(c(4,4,-5,-5)))

response0$fit()
```
```{r}
response0$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```

## Response models with splines of ExpShip

### Response models based on baseline_racvm2

```{r fit response1}

par0 <- c(0,0,1,4,0)

sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))

#mapping for fixed coefficients
new_map=list(coeff_re=factor(c(1,2,rep(NA,N),
                    3,4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                    log_lambda=factor(c(5,NA,6,NA,rep(NA,4))))
#Fit baseline
response1_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

new_coeff_re=c(rep(0,2),baseline_racvm2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline_racvm2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline_racvm2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm2$coeff_re()[paste("omega.ti(Ishore).",1:4,sep=""),1],
               baseline_racvm2$coeff_re()[paste("omega.ti(AngleNormal,Ishore).",1:16,sep=""),1])
new_lambda=c(1,baseline_racvm2$lambda()["tau.s(ID)",1],1,baseline_racvm2$lambda()["nu.s(ID)",1],
             baseline_racvm2$lambda()[3:6,1])

response1_sp$update_map(new_map)
response1_sp$update_coeff_re(new_coeff_re)
response1_sp$update_coeff_fe(baseline_racvm2$coeff_fe()[,1])

response1_sp$update_lambda(new_lambda)

response1_sp$fit()

```

```{r get results response1}

estimates_res1_sp=as.list(response1_sp$tmb_rep(),what="Est")
std_res1_sp=as.list(response1_sp$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plots_res1_sp=response1_sp$get_all_plots(baseline=baseline_racvm2,xmin=xmin,
                                         xmax=xmax,link=link,xlabel=xlabel,show_CI="simultaneous")

```

```{r display plots response1_sp}

#print plots
plots_res1_sp

```

```{r estimates on parameter scale response1}
post_coeff_res1_sp=response1_sp$post_coeff(n_post=1000)
post_par_res1_sp=list("tau0"=exp(post_coeff_res1_sp$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res1_sp$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res1_sp$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,2])))
mean_par_res1_sp=lapply(post_par_res1_sp,mean)
sd_par_res1_sp=lapply(post_par_res1_sp,sd)

```

### Response model with splines of ExpShip based on baseline2

```{r fit response2}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

#mapping for fixed coefficients
new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",
        rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),
        coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))

#Fit baseline
response2_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

new_coeff_re=c(rep(0,2),baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2$lambda()["tau.s(ID)",1],1,baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])

response2_sp$update_map(new_map)
response2_sp$update_coeff_re(new_coeff_re)
response2_sp$update_coeff_fe(baseline2$coeff_fe()[,1])
response2_sp$update_lambda(new_lambda)

response2_sp$fit()

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])

response2_sp_50m$update_map(new_map)
response2_sp_50m$update_coeff_re(new_coeff_re)
response2_sp_50m$update_coeff_fe(baseline2_50m$coeff_fe()[,1])

response2_sp_50m$update_lambda(new_lambda)

response2_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])

response2_sp_75m$update_map(new_map)
response2_sp_75m$update_coeff_re(new_coeff_re)
response2_sp_75m$update_coeff_fe(baseline2_75m$coeff_fe()[,1])

response2_sp_75m$update_lambda(new_lambda)

response2_sp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])

response2_sp_100m$update_map(new_map)
response2_sp_100m$update_coeff_re(new_coeff_re)
response2_sp_100m$update_coeff_fe(baseline2_100m$coeff_fe()[,1])

response2_sp_100m$update_lambda(new_lambda)

response2_sp_100m$fit()
```

```{r get results response2}

estimates_res2_sp=as.list(response2_sp$tmb_rep(),what="Est")
std_res2_sp=as.list(response2_sp$tmb_rep(),what="Std")

estimates_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
std_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")

estimates_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
std_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")

estimates_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
std_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShip"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship")

plot_res2_sp=response2_sp$get_all_plots(baseline=baseline2,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```

```{r display plots response2_sp}

#print plots
plot_res2_sp
plot_res2_sp_50m
plot_res2_sp_75m
plot_res2_sp_100m

```

```{r}

post_coeff_res2_sp=response2_sp$post_coeff(n_post=1000)
post_par_res2_sp=list("tau0"=exp(post_coeff_res2_sp$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,2])))
mean_par_res2_sp=lapply(post_par_res2_sp,mean)
sd_par_res2_sp=lapply(post_par_res2_sp,sd)

post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)

post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)

post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)

```

### Response model with splines of ExpShip based on baseline_racvm1

```{r fit response3}

#initialize parameters
par0=c(0,0,1.5,4.5,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

#mapping for fixed coeffs
new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                    "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
                    log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))

#Fit response
response3_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("log_sigma_obs0"=log(0.07)))

new_coeff_re=c(rep(0,2),baseline_racvm3$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline_racvm3$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline_racvm3$lambda()["tau.s(ID)",1],1,baseline_racvm3$lambda()["nu.s(ID)",1],baseline_racvm3$lambda()[3:6,1])

response3_sp$update_map(new_map)
response3_sp$update_coeff_re(new_coeff_re)
response3_sp$update_coeff_fe(baseline_racvm3$coeff_fe()[,1])

response3_sp$update_lambda(new_lambda)
response3_sp$fit()

#Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_re=c(rep(0,2),baseline_racvm3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline_racvm3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline_racvm3_50m$lambda()["tau.s(ID)",1],1,baseline_racvm3_50m$lambda()["nu.s(ID)",1],baseline_racvm3_50m$lambda()[3:6,1])

response3_sp_50m$update_map(new_map)
response3_sp_50m$update_coeff_re(new_coeff_re)
response3_sp_50m$update_coeff_fe(baseline_racvm3_50m$coeff_fe()[,1])

response3_sp_50m$update_lambda(new_lambda)
response3_sp_50m$fit()

#Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_re=c(rep(0,2),baseline_racvm3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline_racvm3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline_racvm3_75m$lambda()["tau.s(ID)",1],1,baseline_racvm3_75m$lambda()["nu.s(ID)",1],baseline_racvm3_75m$lambda()[3:6,1])

response3_sp_75m$update_map(new_map)
response3_sp_75m$update_coeff_re(new_coeff_re)
response3_sp_75m$update_coeff_fe(baseline_racvm3_75m$coeff_fe()[,1])

response3_sp_75m$update_lambda(new_lambda)
response3_sp_75m$fit()

#Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_re=c(rep(0,2),baseline_racvm3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
               rep(0,2),baseline_racvm3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(1,baseline_racvm3_100m$lambda()["tau.s(ID)",1],1,baseline_racvm3_100m$lambda()["nu.s(ID)",1],baseline_racvm3_100m$lambda()[3:6,1])

response3_sp_100m$update_map(new_map)
response3_sp_100m$update_coeff_re(new_coeff_re)
response3_sp_100m$update_coeff_fe(baseline_racvm3_100m$coeff_fe()[,1])

response3_sp_100m$update_lambda(new_lambda)
response3_sp_100m$fit()
```

```{r get results response3}

estimates_res3_sp=as.list(response3_sp$tmb_rep(),what="Est")
std_res3_sp=as.list(response3_sp$tmb_rep(),what="Std")

estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")

estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")

estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plot_res3_sp=response3_sp$get_all_plots(baseline=baseline_racvm3,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline_racvm3_50m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline_racvm3_75m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline_racvm3_100m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

```

```{r display plots response3_sp}

#print plots
plot_res3_sp
plot_res3_sp_50m
plot_res3_sp_75m
plot_res3_sp_100m

```

```{r estimates on parameter scale response3}

post_coeff_res3_sp=response3_sp$post_coeff(n_post=1000)
post_par_res3_sp=list("tau0"=exp(post_coeff_res3_sp$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,2])))
mean_par_res3_sp=lapply(post_par_res3_sp,mean)
sd_par_res3_sp=lapply(post_par_res3_sp,sd)

post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)

post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)

post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)

```

### Response model with exponential of ExpShip based on baseline2

```{r}

par0=c(0,0,1,4,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")
                 +ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))

# Fit response
response2_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

new_coeff_fe=c(baseline2$coeff_fe()[1:3,1],0,baseline2$coeff_fe()[4,1],0,baseline2$coeff_fe()[5,1])
new_coeff_re=c(baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2$lambda()["tau.s(ID)",1],baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])

response2_exp$update_map(new_map)
response2_exp$update_coeff_re(new_coeff_re)
response2_exp$update_coeff_fe(new_coeff_fe)

response2_exp$update_lambda(new_lambda)

response2_exp$fit()

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])

response2_exp_50m$update_map(new_map)
response2_exp_50m$update_coeff_re(new_coeff_re)
response2_exp_50m$update_coeff_fe(new_coeff_fe)

response2_exp_50m$update_lambda(new_lambda)

response2_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])

response2_exp_75m$update_map(new_map)
response2_exp_75m$update_coeff_re(new_coeff_re)
response2_exp_75m$update_coeff_fe(new_coeff_fe)
response2_exp_75m$update_lambda(new_lambda)

response2_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])

response2_exp_100m$update_map(new_map)
response2_exp_100m$update_coeff_re(new_coeff_re)
response2_exp_100m$update_coeff_fe(new_coeff_fe)

response2_exp_100m$update_lambda(new_lambda)

response2_exp_100m$fit()
```

```{r}

estimates_res2_exp=as.list(response2_exp$tmb_rep(),what="Est")
std_res2_exp=as.list(response2_exp$tmb_rep(),what="Std")

estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")

estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")

estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res2_exp=response2_exp$get_all_plots(baseline=baseline2,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res2_exp_50m=response2_exp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

```

```{r display plots response3_sp}

#print plots
plots_res2_exp
plots_res2_exp_50m
plots_res2_exp_75m
plots_res2_exp_100m

```

### Response model with exponential of ExpShip based on baseline_racvm3

```{r}

par0=c(0,0,1,4.5,0)

#define model
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

#mapping for fixed coeffs
new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))

#Fit response
sigma_obs=exp(estimates_bas3$log_sigma_obs)
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3$coeff_fe()[1:3,1],0,baseline_racvm3$coeff_fe()[4,1],0,baseline_racvm3$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3$lambda()["tau.s(ID)",1],baseline_racvm3$lambda()["nu.s(ID)",1],baseline_racvm3$lambda()[3:6,1])

response3_exp$update_map(new_map)
response3_exp$update_coeff_re(new_coeff_re)
response3_exp$update_coeff_fe(new_coeff_fe)

response3_exp$update_lambda(new_lambda)
response3_exp$fit()

# Fit response with 30m measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_30m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_30m$coeff_fe()[1:3,1],0,baseline_racvm3_30m$coeff_fe()[4,1],0,baseline_racvm3_30m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_30m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_30m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_30m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_30m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_30m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_30m$lambda()["tau.s(ID)",1],baseline_racvm3_30m$lambda()["nu.s(ID)",1],baseline_racvm3_30m$lambda()[3:6,1])

response3_exp_30m$update_map(new_map)
response3_exp_30m$update_coeff_re(new_coeff_re)
response3_exp_30m$update_coeff_fe(new_coeff_fe)

response3_exp_30m$update_lambda(new_lambda)
response3_exp_30m$fit()

# Fit response with 40m measurement error
sigma_obs=0.04
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_40m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_40m$coeff_fe()[1:3,1],0,baseline_racvm3_40m$coeff_fe()[4,1],0,baseline_racvm3_40m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_40m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_40m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_40m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_40m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_40m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_40m$lambda()["tau.s(ID)",1],baseline_racvm3_40m$lambda()["nu.s(ID)",1],baseline_racvm3_40m$lambda()[3:6,1])

response3_exp_40m$update_map(new_map)
response3_exp_40m$update_coeff_re(new_coeff_re)
response3_exp_40m$update_coeff_fe(new_coeff_fe)

response3_exp_40m$update_lambda(new_lambda)
response3_exp_40m$fit()

# Fit response with 45m measurement error
sigma_obs=0.045
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_45m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_45m$coeff_fe()[1:3,1],0,baseline_racvm3_45m$coeff_fe()[4,1],0,baseline_racvm3_45m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_45m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_45m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_45m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_45m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_45m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_45m$lambda()["tau.s(ID)",1],baseline_racvm3_45m$lambda()["nu.s(ID)",1],baseline_racvm3_45m$lambda()[3:6,1])

response3_exp_45m$update_map(new_map)
response3_exp_45m$update_coeff_re(new_coeff_re)
response3_exp_45m$update_coeff_fe(new_coeff_fe)
response3_exp_45m$update_lambda(new_lambda)

response3_exp_45m$fit()

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_50m$coeff_fe()[1:3,1],0,baseline_racvm3_50m$coeff_fe()[4,1],0,baseline_racvm3_50m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_50m$lambda()["tau.s(ID)",1],baseline_racvm3_50m$lambda()["nu.s(ID)",1],baseline_racvm3_50m$lambda()[3:6,1])

response3_exp_50m$update_map(new_map)
response3_exp_50m$update_coeff_re(new_coeff_re)
response3_exp_50m$update_coeff_fe(new_coeff_fe)

response3_exp_50m$update_lambda(new_lambda)
response3_exp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_75m$coeff_fe()[1:3,1],0,baseline_racvm3_75m$coeff_fe()[4,1],0,baseline_racvm3_75m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_75m$lambda()["tau.s(ID)",1],baseline_racvm3_75m$lambda()["nu.s(ID)",1],baseline_racvm3_75m$lambda()[3:6,1])

response3_exp_75m$update_map(new_map)
response3_exp_75m$update_coeff_fe(new_coeff_fe)
response3_exp_75m$update_lambda(new_lambda)

response3_exp_75m$fit()

# Fit response with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))

new_coeff_fe=c(baseline_racvm3_100m$coeff_fe()[1:3,1],0,baseline_racvm3_100m$coeff_fe()[4,1],0,baseline_racvm3_100m$coeff_fe()[5,1])
new_coeff_re=c(baseline_racvm3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
new_lambda=c(baseline_racvm3_100m$lambda()["tau.s(ID)",1],baseline_racvm3_100m$lambda()["nu.s(ID)",1],baseline_racvm3_100m$lambda()[3:6,1])

response3_exp_100m$update_map(new_map)
response3_exp_100m$update_coeff_re(new_coeff_re)
response3_exp_100m$update_coeff_fe(new_coeff_fe)
response3_exp_100m$update_lambda(new_lambda)

response3_exp_100m$fit()
```

```{r}

estimates_res3_exp=as.list(response3_exp$tmb_rep(),what="Est")
std_res3_exp=as.list(response3_exp$tmb_rep(),what="Std")

estimates_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Est")
std_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Std")

estimates_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Est")
std_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Std")

estimates_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Est")
std_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Std")

estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")

estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")

estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")

#plot parameters
xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")

plots_res3_exp=response3_exp$get_all_plots(baseline=baseline_racvm3,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res3_exp_30m=response3_exp_30m$get_all_plots(baseline=baseline_racvm3_30m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res3_exp_40m=response3_exp_40m$get_all_plots(baseline=baseline_racvm3_40m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline_racvm3_50m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline_racvm3_75m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline_racvm3_100m,xmin=xmin,
                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

```

```{r display plots response3_sp}

#print plots
plots_res3_exp
plots_res3_exp_50m
plots_res3_exp_75m
plots_res3_exp_100m

```

We select the response 3 model with 50m measurement error

```{r}
# Create directory
if (!dir.exists("response3_exp_50m")) {
    dir.create("response3_exp_50m", recursive = TRUE)
}

# Use lapply to save each plot
lapply(seq_along(plots_bas2), function(i) {

    plot<-plots_bas2[[i]]
    plot_name <- names(plots_bas2)[i]
    file_path <- file.path("response3_exp_50m", plot_name)

    # Check if the plot is a ggplot or plotly object
    if (inherits(plot, "ggplot")) {
      ggsave(file_path, plot, device = "pdf")
    } else if (inherits(plot, "plotly")) {
      saveWidget(plot, file_path)
  }
    })
```

### Plots improvements

```{r}

#get angular velocity for different distances to shore
plot_me_tau <- plots_res3_exp_50m$me_tau_ExpShip
plot_me_nu <- plots_res3_exp_50m$me_nu_ExpShip

# Update plot_me_tau
plot_me_tau <- plot_me_tau +
  labs(y = expression(tau)) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size=18),
    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
    axis.text.x = element_text(size=14),
    axis.text.y = element_text(size=14),
    legend.title = element_text(size=14),
    legend.text = element_text(size=14)
  )

# Update plot_me_nu
plot_me_nu <- plot_me_nu +
  labs(y = expression(nu)) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size=18),
    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
    axis.text.x = element_text(size=14),
    axis.text.y = element_text(size=14),
    legend.title = element_text(size=14),
    legend.text = element_text(size=14)
  )

# Save the updated plots
ggsave("plot_me_tau_updated.pdf", plot=plot_me_tau, width=10, height=6, units="in")
ggsave("plot_me_nu_updated.pdf", plot=plot_me_nu, width=10, height=6, units="in")

```

## Posterior samples

```{r}
post_coeff_res3_exp_50m=response3_exp_50m$post_coeff(n_post=10000)
post_par_res3_exp_50m=list(
              "tau0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_res3_exp_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,2])),
              "tau.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"tau.ExpShip"],
              "nu.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"nu.ExpShip"])
mean_par_res3=lapply(post_par_res3_exp_50m,mean)
sd_par_res3=lapply(post_par_res3_exp_50m,sd)
quant_par_res3 <- lapply(post_par_res3_exp_50m, quantile, probs = c(0.025, 0.975))

```

#### Thorough uncertainty quantification for response3_exp_50m

```{r}

#posterior samples of the coefficient of the baseline model
post_coeff_re=post_coeff_bas3_50m$coeff_re[1:100,]
post_coeff_fe=post_coeff_bas3_50m$coeff_fe[1:100,]
post_coeff_lambda=exp(post_coeff_bas3_50m$log_lambda[1:100,])

# Fit response with 50m measurement error

par0=c(0,0,1,4.5,0)
formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                 nu=~ExpShip+s(ID,bs="re"),
                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)

for (i in 1:nrow(post_coeff_fe)) {

    response3_exp_50m_temp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
                      response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
                      map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
                               log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))

    new_coeff_fe=c(post_coeff_fe[i,1:3],0,post_coeff_fe[i,4],0,post_coeff_fe[i,5])
    new_coeff_re=post_coeff_re[i,]
    new_lambda=post_coeff_lambda[i,]
    response3_exp_50m_temp$update_coeff_re(new_coeff_re)
    response3_exp_50m_temp$update_coeff_fe(new_coeff_fe)

    response3_exp_50m_temp$update_lambda(new_lambda)
    response3_exp_50m_temp$fit()

    est_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
    std_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]

    estimates[i,]=est_alpha
    std[i,]=std_alpha
}

colnames(estimates)=c("tau.ExpShip","nu.ExpShip")
colnames(std)=c("tau.ExpShip","nu.ExpShip")

```

```{r}
#The 9th value in estimates is inconsistent, we drop it
estimates=estimates[-9,]
std=std[-9,]

mean_est=apply(X=estimates,MARGIN=2,FUN=mean)
mean_std=apply(X=na.omit(std),MARGIN=2,FUN=mean)
```

### 2025-01-07 R/preprocessing/narwhals/data_preprocessing.Rmd
R code updates

@@
 \item Renaming the "Ind" column to "ID" for clarity and converting it to factors.
-\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to shio during trial and intertrial separately
+\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to ship during trial and intertrial separately
 \item Converting column "Posixct" to POSIXct format for date and time
@@
 ```{=tex}
-Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram.
 We thus decide to filter by setting a threshold of $20$ km/h for horizontal speed.
@@
 
-ggsave(path=plots_folder,filename="all_narwhals_tracks.png", 
+ggsave(path=plots_folder,filename="all_narwhals_tracks.pdf", 
        plot = map,width=10,height=5)
@@
 
+scoresby_sound_utm=st_transform(scoresby_sound_utm,crs= "+init=EPSG:32626 +units=km")
 ```
@@
 ```{r plot coastline and land}
-plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")+
-  coord_sf(datum=st_crs(32626))
+plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")
 ```
@@
 Some points that appear on land on the map don't have distance to shore 0. This may be due to the map inaccuracy.
-If we compute the distance to shore based on the map, as done in res_land, we get more points on the land.
 
@@
   facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")
-
-ggsave(path=plots_folder,"realExpShip_through_time.png", plot=pExpShip)
+  xlab("Time (h)")+ylab("Exposure to ship")+
+   theme(
+    axis.text = element_text(size = 12),   # Increase tick size
+    axis.title = element_text(size = 16)  # Increase label size
+  )
+
+ggsave(path=plots_folder,"realExpShip_through_time.pdf", plot=pExpShip)
 ```

### 2025-01-07 R/simulation_study_CRCVM/CRCVM_examples/illustrative_trajectories.Rmd
R code updates

---
title: "Illustrative trajectories"
author: "Alexandre Delporte"
date: "2024-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document, we simulate from constrained correlated velocity models with smooth parameters $omega$ and $tau$. The smooth functions will be parametric and the parameters will have direct interpretation in terms of movement.

# Set up

```{r set up}
cat("\014")                 # Clears the console

#seed for reproducibility 
seed=42

#parallel computing
library(foreach)
library(doParallel)

#for pipeline operator
library(dplyr)

#plots
library(ggplot2)
library(plotly)
library(htmlwidgets)
library(ggpubr)

#data managing
library(tidyr)

#progress bar
library(progress)

#for land polygons
library(sf)

#for spline estimation
library(mgcv)

#to get root of git repo
library(here)

#animated trajectories
library(gganimate)

#get functions to simulate trajectories
source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))

```

# Define initial velocity and location for trajectories

```{r initial velocity and positions}
set.seed(seed)
# Initial velocity and location in fjords -------------

# Set the path to the directory containing the data
greenland_data_path <- here("Data","preprocessed_data","greenland")

#get the land and coastline geometries from the geojson file
land_border<-st_read(file.path(greenland_data_path,"updated_scoresby_sound_utm.shp"))

land_border <- st_transform(land_border, crs = "+init=EPSG:32626 +units=km")

v0=c(0,0)

#generate random initial points
x0=matrix(rep(NA,12),ncol=2)
colnames(x0)=c("x1","x2")

N=6 #number of trajectories
i=1
while (i<=N) {
  #choose location uniformly in the map
  x=c(runif(1,min=430,max=510),runif(1,min=7760,max=7900))
  p=nearest_shore_point(st_point(x),land_border)
  Dshore=(x[1]-p[1])^2+(x[2]-p[2])^2
  if (Dshore<1.5 && Dshore >0.3) {
    x0[i,]=x
    i=i+1
  }
}
```

# Define domains

```{r rectangular domain}

# Rectangular domain around the initial location ---------------------

x.max=max(x0[,1])+5
x.min=min(x0[,1])-5
y.max=max(x0[,2])+5
y.min=min(x0[,2])-5

large_rectangle_coords <- matrix(c(x.min, y.min, x.min, y.max, x.max, y.max, x.max, y.min, x.min,  y.min), ncol = 2, byrow = TRUE)
small_rectangle_coords <- matrix(c(x.min+1, y.min+1, x.min+1, y.max-1, x.max-1, y.max-1, x.max-1, y.min+1, x.min+1,  y.min+1), ncol = 2, byrow = TRUE)

# Create an sf object representing the rectangles
large_rectangle <- st_sfc(st_polygon(list(large_rectangle_coords)))
large_rectangle<-st_sf(geometry = large_rectangle)

small_rectangle <- st_sfc(st_polygon(list(small_rectangle_coords)))
small_rectangle<-st_sf(geometry = small_rectangle)

#change its crs
rect_border<- st_sym_difference(large_rectangle, small_rectangle)

```

```{r circular domain}

# Circular domain around the initial position -------
middle=c(mean(x0[,1]),mean(x0[,2]))

# Create circle geometries
large_circle <- st_buffer(st_sfc(st_point(middle)), dist = 55)
small_circle <- st_buffer(st_sfc(st_point(middle)), dist = 54)

# Convert circles to polygons
large_circle_polygon <- st_cast(large_circle, "POLYGON")
small_circle_polygon <- st_cast(small_circle, "POLYGON")

# Create sf objects
large_circle_sf <- st_sf(geometry = large_circle_polygon)
small_circle_sf <- st_sf(geometry = small_circle_polygon)

# Calculate the symmetric difference to get the border
circ_border <- st_sym_difference(large_circle_sf, small_circle_sf)

```

```{r circular domains with islands}

#### Circular domain with islands

# Create small circles borders within the big circle
circle1 <- st_buffer(st_sfc(st_point(middle-c(15,0))), dist = 10)

# Convert circles to polygons
circle_polygon1 <- st_cast(circle1, "POLYGON")

# Create sf objects
circle_sf1 <- st_sf(geometry = circle_polygon1)

# Create small circles borders within the big circle
circle2 <- st_buffer(st_sfc(st_point(middle-c(0,15))), dist = 10)

# Convert circles to polygons
circle_polygon2 <- st_cast(circle2, "POLYGON")

# Create sf objects
circle_sf2 <- st_sf(geometry = circle_polygon2)

circ_island_border <- st_union(st_union(circ_border, circle_sf1), circle_sf2)

```

```{r plot domains}

# Plot using ggplot2
ggplot() +
  # Rectangular border
  geom_sf(data = rect_border, fill = "lightblue", color = "black") +
  ggtitle("Rectangular Border") +
  theme_minimal()

ggplot() +
  # Circular border
  geom_sf(data = circ_border, fill = "lightgreen", color = "black") +
  ggtitle("Circular Border") +
  theme_minimal()

ggplot() +
  # Circular domain with islands
  geom_sf(data = circ_island_border, fill = "lightcoral", color = "black") +
  ggtitle("Circular Border with Islands") +
  theme_minimal()
```

# Define parametric functions for smooth parameters $\tau$

First define the function for a constant value of $\tau$. By default, we set this value to $2$ h.

```{r function for constant tau}

#constant tau
ftau_constant=function(cov_data,tau=2) {
  return (tau)
}

```

Then we define smooth functions of the angle $\theta$ to model higher persistence along the shoreline. The function need to have bumps around $\frac{\pi}{2}$ and $-\frac{\pi}{2}$. This type of function can be obtained in multiple ways, using hyperbolic tangent or gaussian bumps for instance. We give an example with hyperbolic tangent.

```{r tanh tau}

ftau_tanh=function(cov_data,alpha=1,tau0=0.5,tau1=4) {
  #cov_data : dataframe of covariate with column for angle theta
  #alpha : steepness of the bump
  #tau0 : base value of tau
  #tau1 : maximum value of tau
  epsilon=0.1
  theta=cov_data$theta
  tau=tau0+(tau1-tau0)/2/tanh(epsilon/alpha)*(tanh((theta+(pi/2+epsilon))/alpha)-tanh((theta+(pi/2-epsilon))/alpha))+
         (tau1-tau0)/2/tanh(epsilon/alpha)*(tanh((theta+-pi/2+epsilon)/alpha)-tanh((theta-pi/2-epsilon)/alpha))

  return (tau)
}

```

```{r plot tanh tau tau1}
# Plot curves for different values of tau1
tau1_values <- c(2,4)

cov_data=data.frame("theta"=seq(from=-pi+0.1,to=pi-0.1,length.out=100))

plot_data_tau1 <- data.frame(
  theta = rep(cov_data$theta, times = length(tau1_values)),
  tau1 = rep(tau1_values, each = length(cov_data$theta)))
plot_data_tau1$tau=unlist(lapply(tau1_values,function(tau1) {ftau_tanh(cov_data,tau1=tau1)}))
# Plot using ggplot
ggplot(plot_data_tau1, aes(x = theta, y = tau, color = as.factor(tau1))) +
  geom_line() +
 ylab(expression(tau)) + 
  xlab(expression(Theta)) + 
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "black") + 
  annotate("text", x = -pi/2-0.3, y = 0.3, label = expression(-pi/2), color = "black", size = 6) + 
  annotate("text", x = pi/2+0.2, y = 0.3, label = expression(pi/2), color = "black", size = 6) + 
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("tau1 = 2", "tau1 = 4"))+
  theme_minimal() +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

```

```{r plot tanh tau alpha}
# Plot curves for different values of alpha
alpha_values<-c(0.5,1)

plot_data_alpha <- data.frame(
  theta = rep(cov_data$theta, times = length(tau1_values)),
  alpha = rep(alpha_values, each = length(cov_data$theta)))
plot_data_alpha$tau=unlist(lapply(alpha_values,function(alpha) {ftau_tanh(cov_data,alpha=alpha)}))
# Plot using ggplot
ggplot(plot_data_alpha, aes(x = theta, y = tau, color = as.factor(alpha))) +
  geom_line() +
 ylab(expression(tau)) + 
  xlab(expression(Theta)) + 
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "black") + 
  annotate("text", x = -pi/2-0.3, y = 0.3, label = expression(-pi/2), color = "black", size = 6) + 
  annotate("text", x = pi/2+0.2, y = 0.3, label = expression(pi/2), color = "black", size = 6) + 
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("alpha = 0.5", "alpha = 1"))+
  theme_minimal() +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

```

However this parametrization is quite complex and we can do better and easier with gaussian bumps.

```{r define tau gaussian}

ftau_gaussian=function(cov_data,sigma_bump=0.3,sigma_dip=0.5,tau0=1,dtau_bump=1,dtau_dip=0.5) {
  #cov_data : dataframe of covariates with column for angle theta
  #sigma_bump : width of the bumps at -pi/2 and pi/2
  #sigma_dip : width of the dip at 0
  #tau0 : base value of tau
  #dtau_bump : height of the bump
  #dtau_dip : depth of the dip

  theta=cov_data$theta
  bump1 <- dtau_bump*exp(-((theta - pi/2)^2) / (2 * sigma_bump^2))
  bump2 <- dtau_bump*exp(-((theta + pi/2)^2) / (2 * sigma_bump^2))
  dip<- dtau_dip*exp(-((theta)^2) / (2 * sigma_dip^2))
  tau=tau0+bump1+bump2-dip

  return (tau)
}

```

```{r plot tau gaussian dtau_bump}
dtau_bump_values=c(1,3)

plot_data_dtau <- data.frame(
  theta = rep(cov_data$theta, times = length(dtau_bump_values)),
  dtau = rep(dtau_bump_values, each = length(cov_data$theta)))
plot_data_dtau$tau=unlist(lapply(dtau_bump_values,function(dtau) {ftau_gaussian(cov_data,dtau_bump=dtau)}))

#ggplot for tau
plot_tau_gaussian_dtau <- ggplot(data=plot_data_dtau) + 
  geom_line(aes(x = theta, y = tau,color = as.factor(dtau))) + 
  ylab(expression(tau)) + 
  xlab(expression(Theta)) + 
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "black") + 
  annotate("text", x = -pi/2-0.3, y = 0.3, label = expression(-pi/2), color = "black", size = 6) + 
  annotate("text", x = pi/2+0.2, y = 0.3, label = expression(pi/2), color = "black", size = 6) + 
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("dtau_bump = 1", "dtau_bump = 3"))+
  theme_minimal() +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

plot_tau_gaussian_dtau

```

```{r plot tau gaussian sigma_bump}
sigma_bump_values=c(0.2,0.4)

plot_data_sigma_bump <- data.frame(
  theta = rep(cov_data$theta, times = length(sigma_bump_values)),
  sigma = rep(sigma_bump_values, each = length(cov_data$theta)))
plot_data_sigma_bump$tau=unlist(lapply(sigma_bump_values,function(sigma) {ftau_gaussian(cov_data,sigma_bump=sigma)}))

#ggplot for tau
plot_tau_gaussian_sigma <- ggplot(data=plot_data_sigma_bump) + 
  geom_line(aes(x = theta, y = tau,color = as.factor(sigma))) + 
  ylab(expression(tau)) + 
  xlab(expression(Theta)) + 
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "black") + 
  annotate("text", x = -pi/2-0.3, y = 0.3, label = expression(-pi/2), color = "black", size = 6) + 
  annotate("text", x = pi/2+0.2, y = 0.3, label = expression(pi/2), color = "black", size = 6) + 
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("sigma_bump = 0.2", "sigma_bump = 0.4"))+
  theme_minimal() +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

plot_tau_gaussian_sigma
```

# Define parametric functions for smooth parameter $\omega$

The parameter $\omega$ will drive how the process rotates depending on the current distance to the boundary and the current angle. It is crucial to rotate fast enough and in the right direction to avoid hitting the boundary. Obviously, what we mean by "fast enough" and "right direction" depends on the shape of the boundary. What is fast enough for a smooth boundary might not be fast enough in a corner for instance. $\vert \omega \vert$ should increase fast as $\vert \theta \vert$ ranges through $\left[ \frac{\pi}{2},\pi \right]$. The sign of $\omega$ depends on the sign of $\theta$. For positive angles $\theta \in \left[ \frac{\pi}{2},\pi \right]$, we need the velocity to rotate in the inverse trigonometric sense. Since the rotation induced in the velocity during $\Delta$ hours is $-\omega \Delta$, this means that we need a positive $\omega$ to effectively rotate in the inverse trigonometric sense. Similarly, for negative $\theta \in \left[ -\pi,-\frac{\pi}{2} \right]$, we need a negative $\omega$.

This type of behavior can be obtained with hyperbolic tangent function.

```{r parametric tanh function for omega}

fomega_tanh=function(cov_data,D0=0.5,omega0=60*pi/2,lambda=2,kappa=0.2) {
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$ExpShore
  }
  coeff=exp(-kappa*(Dshore/D0)^2)
  omega=omega0/2*(tanh(lambda*(theta+pi/2-atanh(-0.9)/lambda))+tanh(lambda*(theta-(pi/2-atanh(-0.9)/lambda))))*coeff
  return(omega)
}

```

```{r plot omega tanh}

#grid 
Dshore=seq(from=0.05,to=4,length.out=30)
theta=seq(from=-pi+0.1,to=pi-0.1,length.out=30)
grid<- as.data.frame(expand.grid(Dshore,theta))
colnames(grid)=c("DistanceShore","theta")

#smooth omega
omega_tanh=matrix(fomega_tanh(grid),30,30)

# dynamic plots for omega

plot_omega_tanh <- plot_ly(type = "surface",
                           contours = list(z = list(show = TRUE, start = min(omega_tanh), end = max(omega_tanh), size = 5,color="black")),
                           x = ~theta,y = ~Dshore,z=~omega_tanh,
                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
  layout(title=paste("Omega"),
         scene=list(xaxis = list(title = "Theta",showgrid = F),
                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_omega_tanh
```

A similar smooth function may also be obtained with the inverse of a square root near $-\pi$ and $\pi$.

```{r define omega  sqrt}

fomega_sqrt <- function(cov_data, a=10*pi, p = 2,D0=0.5) {

  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  epsilon <- pmax(0, Dshore - D0)^2  # Damping term
  ifelse(
    abs(theta) <= pi / 2,
    0,  # Flat in the middle
    sign(theta) * a * ((abs(theta) - pi / 2)^p) /
      sqrt((pi - abs(theta)) + epsilon)
  )
}

```

```{r plot omega sqrt}
#smooth omega
omega_sqrt=matrix(fomega_sqrt(grid),30,30)

# dynamic plots for omega

plot_omega_sqrt <- plot_ly(type = "surface",
                           contours = list(z = list(show = TRUE, start = min(omega_sqrt), end = max(omega_sqrt), size = 5,color="black")),
                           x = ~theta,y = ~Dshore,z=~omega_sqrt,
                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
  layout(title=paste("Omega"),
         scene=list(xaxis = list(title = "Theta",showgrid = F),
                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_omega_sqrt
```

These smooth functions $\omega$ can produce a SDE model that is constrained within a restricted domain of $\R^2$.
The trajectories avoid the shoreline, but are not attracted to it. To model a tendecy to follow the shoreline, we need to modify a bit the smooth function to get a rotation of the velocity toward the shore when it is on the verge of leaving it. We can do this in several ways. 
First, we use a sin function.

```{r define and plot omega sin}
# Define the extended function f(theta)
fomega_sin<- function(cov_data, a =4*pi,b=80*pi,D0=0.8) {
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }
  ifelse(abs(theta) < pi/2, 
         -a * sin(2*theta)*exp(-(Dshore/D0)^2),
         ifelse(theta>0,
               sign(theta)*b*abs(theta-pi/2)^2*exp(-(Dshore/D0)^2),
               sign(theta)*b*abs(theta+pi/2)^2*exp(-(Dshore/D0)^2)))
}

#smooth omega
omega_sin=matrix(fomega_sin(grid),30,30)

# dynamic plots for omega

plot_omega_sin <- plot_ly(type = "surface",
                           contours = list(z = list(show = TRUE, start = min(omega_sin), end = max(omega_sin), size = 5,color="black")),
                           x = ~theta,y = ~Dshore,z=~omega_sin,
                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
  layout(title=paste("Omega"),
         scene=list(xaxis = list(title = "Theta",showgrid = F),
                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_omega_sin

```

Now we use a cubic polynomial with an additional gaussian kernel.

```{r}
fomega_cubic=function(cov_data,a=1,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.2,b=5){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
      b*(exp(-1/2*(theta+pi/2/sqrt(3))^2/sigma_theta^2)-
           exp(-1/2*(theta-pi/2/sqrt(3))^2/(sigma_theta)^2))*exp(-1/2*(Dshore-D1)^2/sigma_D^2)
}

#smooth omega
omega_cubic=matrix(fomega_cubic(grid),30,30)

# dynamic plots for omega

plot_omega_cubic <- plot_ly(type = "surface",
                           contours = list(z = list(show = TRUE, start = min(omega_cubic), end = max(omega_cubic), size = 5,color="black")),
                           x = ~theta,y = ~Dshore,z=~omega_cubic,
                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
  layout(title=paste("Omega"),
         scene=list(xaxis = list(title = "Theta",showgrid = F),
                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_omega_cubic
```

We can also geometrically compute what should be the angular veolcity to avoid land. And we get the expression of $\omega$ like this.

```{r}

fomega_minimal=function(cov_data,nu=4,D0=0.5,D1=1.5,a=1,b=10,sigma_D=0.5,sigma_theta=pi/8) {
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  ifelse(abs(theta)>pi/2,a*sign(theta)*(pi/2-abs(theta))*nu*cos(theta)/Dshore*exp(-Dshore/D0),-sign(theta)*b*
    exp(-1/2*(theta-sign(theta)*pi/4)^2/sigma_theta^2)*exp(-Dshore/D1))
}

#smooth omega
omega_minimal=matrix(fomega_minimal(grid),30,30)

# dynamic plots for omega

plot_omega_minimal <- plot_ly(type = "surface",
                           contours = list(z = list(show = TRUE, start = min(omega_minimal), end = max(omega_minimal), size = 5,color="black")),
                           x = ~theta,y = ~Dshore,z=~omega_minimal,
                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
  layout(title=paste("Omega"),
         scene=list(xaxis = list(title = "Theta",showgrid = F),
                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))

plot_omega_minimal
```

We can approximate these surfaces using tensor splines

```{r approximate tanh omega with splines}
# Splines smooth for omega -------------------- 
n <- 100000
SP_DF=5

#sample theta and DistanceShore points in the domain
theta <- runif(n,-pi,pi);
DistanceShore <- runif(n,0.2,8);
samples=data.frame(theta=theta,DistanceShore=DistanceShore)

#define grid of values
theta_v <- seq(-pi,pi,length=30)
Dshore_v<- seq(0.2,8,length=30)
pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))

#points on the surface perturbed by gaussian noise
f <- fomega_tanh(samples)
y <- f+0.1*rnorm(n)

#fit with bivariate splines te
m1 <- gam(y~te(theta,DistanceShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
                                                       DistanceShore=seq(0.2,8,len=SP_DF)))

ExpShore=1/DistanceShore

#fit with bivariate splines te
m2 <- gam(y~te(theta,ExpShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
                                                  ExpShore=seq(1/8,1/0.2,len=SP_DF)))

fomega_tanh_splines=function(cov_data) {

  if (is.null(cov_data$ExpShore)){
    omega=predict(m1,newdata=cov_data)
  }
  else if (is.null(cov_data$DistanceShore)){
    omega=predict(m2,newdata=cov_data)
  }

  return(as.numeric(omega))
}

```

```{r plot tanh parametric and spline omega}

# Perspective plots for omega

# Grid 
Dshore = seq(from = 0, to = 2, length.out = 30)
theta = seq(from = -pi + 0.1, to = pi - 0.1, length.out = 30)
grid <- as.data.frame(expand.grid(Dshore, theta))
colnames(grid) = c("DistanceShore", "theta")

# Smooth omega
z = matrix(fomega_cubic_advanced(grid), 30, 30)

# Create a function interpolating colors in the range of specified colors
jet.colors <- colorRampPalette( c("blue", "green") )
# Generate the desired number of colors from this palette
nbcol <- 100
color <- jet.colors(nbcol)
# Compute the z-value at the facet centres
ncz=ncol(z)
nrz=nrow(z)
zfacet <- z[-1, -1] + z[-1, -ncz] + z[-nrz, -1] + z[-nrz, -ncz]
# Recode facet z-values into color indices
facetcol <- cut(zfacet, nbcol)

# Open PNG device 
png(file.path("perspective_plots.png"), width = 1000, height = 500)

par(mfrow = c(1, 2), mar = c(3, 3, 3, 3))  #

# First plot
res=persp(Dshore, theta, z, 
      phi = 15,
      axes = FALSE,
      col = color[facetcol], 
      cex.lab = 1.5, 
      ticktype = "simple",
      xlim = c(min(Dshore), max(Dshore)),  # Explicit xlim
      ylim = c(min(theta), max(theta)))

#lines(trans3d(seq(0.05,2,length.out=5), y = rep(-pi+0.1,5), z = rep(min(z),5), pmat = res), col = "black")
lines(trans3d(rep(1,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)
lines(trans3d(rep(0.5,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)
lines(trans3d(rep(0.1,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)

text(trans3d(0.1, -pi + 0.05, min(z)+30, pmat = res), labels = "0.1", pos = 4, cex = 1.2, col = "black")  # Line 4
text(trans3d(0.5, -pi + 0.05, min(z)+30, pmat = res), labels = "0.5", pos = 4, cex = 1.2, col = "black")  # Line 2
text(trans3d(1, -pi + 0.05, min(z)+30, pmat = res), labels = "1", pos = 4, cex = 1.2, col = "black")  # Line 3

mtext(expression(D^{shore}), side = 1, line = -2, cex = 1.5)
mtext(expression(omega), side = 2, line = 1, cex = 1.5, las = 2)

# Second plot
persp(Dshore, theta, z, 
      phi = 15,theta=-90,
      axes = FALSE,
      col = color[facetcol], 
      cex.lab = 1.5,  #
      ticktype = "simple")
mtext(expression(Theta), side = 1, line = -2, cex = 1.5)
mtext(expression(omega), side = 2, line = 1, cex = 1.5, las = 2)

#lines(trans3d(seq(0.05,2,length.out=5), y = rep(-pi+0.1,5), z = rep(min(z),5), pmat = res), col = "black")
lines(trans3d(rep(0.5,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)
lines(trans3d(rep(1,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)
lines(trans3d(rep(1.5,5), y = rep(-pi+0.1,5), z = seq(min(z)-20,min(z)+1030,5), pmat = res), col = "grey",lty=2)

text(trans3d(0.5,-pi+0.05, min(z)+30, pmat = res), labels = expression(-pi/2), pos = 4, cex = 1.2, col = "black") 
text(trans3d(1, -pi + 0.05, min(z)+30, pmat = res), labels = "0", pos = 4, cex = 1.2, col = "black")  
text(trans3d(1.5, -pi + 0.05, min(z)+30, pmat = res), labels = expression(pi/2), pos = 4, cex = 1.2, col = "black")

dev.off()

```

We will keep the velocity parameter $\nu$ constant in our simulations.

```{r constant function for nu}

#constant nu
fnu_constant=function(cov_data,nu=4) {
  return (nu)
}
```

# Trajectories in a rectangular domain.

To start we inspect what the trajectories look like in the rectangular domain. 
We will try the different parametric function that we used before.

We simulate over 3 days with a constant time step equal to 10 seconds.

```{r set time steps for simulation}
TMAX=3*24
dt=1/360
times=seq(0,TMAX,by=dt)

```

```{r simulate with omega tanh in rectangle,cache=TRUE}
# CRCVM with tanh smooth omega

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_tanh=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_minimal,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=rect_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_rect_tanh_sub=data_sim_rect_tanh[seq(1,length(data_sim_rect_tanh$y1),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

The process is avoiding land by rotating when it approaches the boundary. 
Now, to have trajectories that tends to remain close to the boundary and move along it we can use the function \texttt{fomega_cubic} instead.

```{r simulate with omega cubic in rectangle,cache=TRUE}

fomega_cubic_rect_attractive=function(cov_data) {

  fomega_cubic(cov_data,a=3,D0=0.5,D1=1.5,sigma_theta=pi/8,sigma_D=0.5,b=5)

}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=rect_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_rect_cubic_attractive_sub=data_sim_rect_cubic_attractive[seq(1,length(data_sim_rect_cubic_attractive$y1),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

We remove the attractive term in the function by taking parameter $b=0$.

```{r simulate with omega cubic in rectangle,cache=TRUE}

fomega_cubic_rect_repulsive=function(cov_data) {

  fomega_cubic(cov_data,a=2,D0=0.5,D1=1.5,sigma_theta=pi/10,sigma_D=0.5,b=0)

}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_rect_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=rect_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_rect_cubic_repulsive_sub=data_sim_rect_cubic_repulsive[seq(1,length(data_sim_rect_cubic_repulsive$y1),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

```{r simulate with omega minimal in rectangle,cache=TRUE}

fomega_minimal_rect_attractive=function(cov_data) {

  fomega_minimal(cov_data,nu=4,D0=0.5,D1=1.5,a=3,b=10,sigma_D=0.5,sigma_theta=pi/8)
}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_minimal_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_minimal_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=rect_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_rect_minimal_attractive_sub=data_sim_rect_minimal_attractive[seq(1,length(data_sim_rect_minimal_attractive$y1),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_minimal_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_minimal_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

Until now, we kept the parameter $\tau$ constant. We can emphasize the tendency to move along the boundary by using a smooth $\tau$ given by the functions \texttt{ftau_tanh} or \texttt{ftau_gaussian}. Here, we plot trajectories for the SDE model where the smoth $\omega$ parameter is given by \texttt{fomega_cubic} and the smooth $\tau$ parameter is given by \texttt{ftau\_gaussian}.

```{r simulate with omega cubic tau gaussian in rectangle,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_cubic_gaussian_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_gaussian,fomega=fomega_cubic_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=rect_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}
#subsample to 10 minutes
data_sim_rect_cubic_gaussian_attractive_sub=data_sim_rect_cubic_gaussian_attractive[seq(1,length(data_sim_rect_cubic_gaussian_attractive$y1),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_cubic_gaussian_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_cubic_gaussian_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

Finally, we compare it to Hanks simulation technique for reflected SDE.

```{r simulate hanks reflected in rectangle,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_rect_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  data_sim=sim_HANKSreflected(mu1=0,mu2=0,beta=0.5,sigma=8/sqrt(pi*2),x0[i,],v0,dt,TMAX,land=rect_border)
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_sim
}

#stop cluster
stopCluster(cl)

```

```{r}
#subsample to 10 minutes
data_sim_rect_hanks_sub=data_sim_rect_hanks[seq(1,length(data_sim_rect_hanks$time),by=60),]

ggplot() +
    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_rect_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_rect_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

# Trajectories in a circular domain.

The values for the parameter of the smooth function $\omega$ and $\tau$ that produce constrained trajectories in a domain may not work for another domain.

```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}

#change parameter values 
fomega_cubic_circ_attractive=function(cov_data) {

   fomega_cubic(cov_data,a=30,D0=0.5,D1=3,sigma_theta=pi/10,sigma_D=1,b=5)
}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_circ_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_circ_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=circ_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_circ_cubic_attractive_sub=data_sim_circ_cubic_attractive[seq(1,length(data_sim_circ_cubic_attractive$y1),by=60),]

ggplot() +
    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_circ_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_circ_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}

#change parameter values 
fomega_cubic_circ_repulsive=function(cov_data) {

   fomega_cubic(cov_data,a=30,D0=0.5,D1=1.5,sigma_theta=pi/10,sigma_D=0.5,b=0)
}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_circ_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_circ_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=circ_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}

#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_circ_cubic_repulsive_sub=data_sim_circ_cubic_repulsive[seq(1,length(data_sim_circ_cubic_repulsive$y1),by=60),]

ggplot() +
    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_circ_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_circ_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

Finally, we compare it to Hanks simulation technique for reflected SDE.

```{r simulate hanks reflected in rectangle,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_circ_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  data_sim=sim_HANKSreflected(mu1=0,mu2=0,beta=0.5,sigma=8/sqrt(pi*2),x0[i,],v0,dt,TMAX,land=circ_border)
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_sim
}

#stop cluster
stopCluster(cl)

```

```{r}
#subsample to 10 minutes
data_sim_circ_hanks_sub=data_sim_circ_hanks[seq(1,length(data_sim_circ_hanks$time),by=60),]

ggplot() +
    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_circ_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_circ_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```
# Trajectories in a circular domain with islands

```{r simulate with omega cubic and tau gaussian in circle with islands,cache=TRUE}

set.seed(42)

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_circ_island_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_gaussian_circ_island,fomega=fomega_cubic_circ_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=circ_island_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}
#stop cluster
stopCluster(cl)

```

```{r}

data_sim_circ_island_cubic_attractive_sub=data_sim_circ_island_cubic_attractive[seq(1,length(data_sim_circ_island_cubic_attractive$y1),by=10),]

ggplot() +
    geom_sf(data = circ_island_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_circ_island_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_circ_island_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

```{r simulate with omega cubic and tau gaussian in circle with islands,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_circ_island_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_gaussian_circ_island,fomega=fomega_cubic_circ_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=circ_island_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}
#stop cluster
stopCluster(cl)

```

```{r}

data_sim_circ_island_cubic_repulsive_sub=data_sim_circ_island_cubic_repulsive[seq(1,length(data_sim_circ_island_cubic_repulsive$y1),by=60),]

ggplot() +
    geom_sf(data = circ_island_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_circ_island_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_circ_island_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

# Trajectories in Scoresby Sound fjords

```{r simulate with omega tanh in fjords,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_fjords_tanh=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_tanh,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=land_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}
#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_fjords_tanh_sub=data_sim_fjords_tanh[seq(1,length(data_sim_fjords_tanh$y1),by=60),]

ggplot() +
    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_fjords_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_fjords_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none")

```

```{r simulate with omega cubic and tau gaussian in fjords,cache=TRUE}

#set parameters values
fomega_cubic_fjords_attractive<- function(cov_data){
 fomega_cubic(cov_data,a=1,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.5,b=5)
}

ftau_gaussian_fjords=function(cov_data) {
  ftau_gaussian(cov_data,sigma_bump=0.4,sigma_dip=0.4,tau0=1,dtau_bump=1.5,dtau_dip=0.5
                )
  }

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) 
registerDoParallel(cl)

data_sim_fjords_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_fjords_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=land_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}
#stop cluster
stopCluster(cl)

```

```{r}
#subsample to 10 minutes
data_sim_fjords_cubic_attractive_sub=data_sim_fjords_cubic_attractive[seq(1,length(data_sim_fjords_cubic_attractive$y1),by=60),]

ggplot() +
    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_fjords_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_fjords_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

```{r animate trajectory sim_fjords_cubic_gaussian_sub,eval=FALSE}

# Basic ggplot setup
plot_data_sim_fjords_cubic_attractive_sub <- ggplot() +
  geom_sf(data=land_border$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=data_sim_fjords_cubic_attractive_sub, 
            aes(x = y1, y = y2,group=ID),size = 1, 
            lineend = "round",alpha=0.15) +     
  geom_point(data=data_sim_fjords_cubic_attractive_sub,
             aes(x = y1, y = y2,shape=ID),size =4 ) +  
  labs(title = "Simulated trajectory in Scoresby Sound ", 
       subtitle = "Time : {frame_along} h", 
       x = "X Position", 
       y = "Y Position") +
  theme_minimal()

anim <- plot_data_sim_fjords_cubic_attractive_sub + 
  transition_reveal(along = time) + 
  labs(subtitle = "Time: {round(frame_along,1)} hours")

# Save the animation
animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
        renderer = gifski_renderer("sim_fjords_cubic_attractive_sub.gif"))

```

![Simulated Trajectories](sim_fjords_cubic_attractive_sub.gif)
```{r}

# Add covariates to simulation -----------

signed_angle <- function(u, v) {
  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
  # as in 
  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
  u <- matrix(u, ncol = 2)
  v <- matrix(v, ncol = 2)
  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
                                  rows")
  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
  ind1 <- which(result > pi)
  ind2 <- which(result <= -pi)
  result[ind1] <- result[ind1] - 2*pi
  result[ind2] <- result[ind2] + 2*pi
  return(result) 
}

add_covs=function(data) {
  #data has at least columns "time","y1" "y2", "p1","p2"
  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"

  new_data=data
  #observed deviation angles
  n=length(new_data$time)

  new_data$theta=rep(NA,n)

  for (id in unique(new_data$ID)) {

    #filter data with specific id
    sub_ind=(new_data$ID==id)
    sub_data=new_data[sub_ind,]
    n_sub=length(sub_data$time)

    #time steps
    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

    #step lengths
    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]

    #matrix of empirical velocity
    vexp_df=cbind(dx/dtimes,dy/dtimes)

    #matrix of normal vectors
    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])

    #angle between velocity and normal vector
    theta_coast=signed_angle(normal,vexp_df)

    #adjust lengths 
    theta_coast=c(theta_coast,1)

    new_data[sub_ind,"theta"]=theta_coast
  }

  #distance to shore
  Dshore=sqrt((new_data$y1-new_data$p1)^2+(new_data$y2-new_data$p2)^2)
  new_data$DistanceShore=Dshore
  new_data$ExpShore=1/Dshore

  return(new_data)
}

```

```{r}
data_sim_fjords_cubic_attractive_sub=add_covs(data_sim_fjords_cubic_attractive_sub)
data_sim_fjords_cubic_attractive_sub$angular_velocity=
  c(NA,diff(data_sim_fjords_cubic_attractive_sub$theta)/
      diff(data_sim_fjords_cubic_attractive_sub$time))

ggplot(data=data_sim_fjords_cubic_attractive_sub)+geom_point(aes(x=time,y=DistanceShore,col=ID))+
  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)

ggplot(data=data_sim_fjords_cubic_attractive_sub)+geom_point(aes(x=time,y=angular_velocity,col=ID))+
  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)
```

We now simulate trajectories without the attractive gaussian term in $\omega$.

```{r simulate with omega cubic advanced and tau gaussian in fjords,cache=TRUE}

fomega_cubic_fjords_repulsive=function(cov_data) {
  fomega_cubic(cov_data,a=40,D0=0.3,D1=0.8,sigma_theta=pi/10,sigma_D=0.3,b=0)
}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_fjords_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_fjords_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
                    times=times,land=land_border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_shore=res$shore[,c("p1","p2")]
  cbind(data_sim,data_shore)
}
#stop cluster
stopCluster(cl)

```

```{r}

#subsample to 10 minutes
data_sim_fjords_cubic_repulsive_sub=data_sim_fjords_cubic_repulsive[
  seq(1,length(data_sim_fjords_cubic_repulsive$y1),by=60),]

ggplot() +
    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_fjords_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_fjords_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none")

```

```{r animate trajectory sim_fjords_cubic_advanced_gaussian_sub,eval=FALSE}

# Basic ggplot setup
plot_data_sim_fjords_cubic_repulsive_sub <- ggplot() +
  geom_sf(data=land_border$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=data_sim_fjords_cubic_repulsive_sub, 
            aes(x = y1, y = y2,group=ID),size = 1, 
            lineend = "round",alpha=0.15) +     
  geom_point(data=data_sim_fjords_cubic_repulsive,
             aes(x = y1, y = y2,shape=ID),size =4 ) +  
  labs(title = "Simulated trajectory in Scoresby Sound ", 
       subtitle = "Time : {frame_along} h", 
       x = "X Position", 
       y = "Y Position") +
  theme_minimal()

anim <- plot_data_sim_fjords_cubic_repulsive_sub + 
  transition_reveal(along = time) + 
  labs(subtitle = "Time: {round(frame_along,1)} hours")

# Save the animation
animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
        renderer = gifski_renderer("sim_fjords_cubic_repulsive_sub.gif"))

```

![Simulated Trajectories](sim_fjords_cubic_repulsive_sub.gif)
```{r}
data_sim_fjords_cubic_advanced_gaussian_sub=add_covs(data_sim_fjords_cubic_advanced_gaussian_sub)
data_sim_fjords_cubic_advanced_gaussian_sub$angular_velocity=
  c(NA,diff(data_sim_fjords_cubic_advanced_gaussian_sub$theta)/
      diff(data_sim_fjords_cubic_advanced_gaussian_sub$time))

ggplot(data=data_sim_fjords_cubic_advanced_gaussian_sub)+geom_point(aes(x=time,y=DistanceShore,col=ID))+
  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)

ggplot(data=data_sim_fjords_cubic_advanced_gaussian_sub)+geom_point(aes(x=time,y=angular_velocity,col=ID))+
  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)
```

Finally, we compare it to Hanks simulation technique for reflected SDE.

```{r simulate hanks reflected in rectangle,cache=TRUE}

# Generate samples ---------------
cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data_sim_fjords_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N+i)

  data_sim=sim_HANKSreflected(mu1=0,mu2=0,beta=0.5,sigma=8/sqrt(pi*2),x0[i,],v0,dt,TMAX,land=land_border)
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_sim
}

#stop cluster
stopCluster(cl)

```

```{r}
#subsample to 10 minutes
data_sim_fjords_hanks_sub=data_sim_fjords_hanks[seq(1,length(data_sim_fjords_hanks$time),by=60),]

ggplot() +
    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_sim_fjords_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_sim_fjords_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none") 
```

```{r zoom in fjords}

#Zoom on Fjords for the plot --------------------

x.min= min(c(data_sim_fjords_cubic_repulsive_sub$y1,data_sim_fjords_cubic_attractive_sub$y1,
             data_sim_fjords_hanks_sub$y1))-1
x.max = max(c(data_sim_fjords_cubic_repulsive_sub$y1,data_sim_fjords_cubic_attractive_sub$y1,
              data_sim_fjords_hanks_sub$y1))+1
y.min = min(c(data_sim_fjords_cubic_repulsive_sub$y2,data_sim_fjords_cubic_attractive_sub$y2,
              data_sim_fjords_hanks_sub$y2))-1
y.max = max(c(data_sim_fjords_cubic_repulsive_sub$y2,data_sim_fjords_cubic_attractive_sub$y2,
              data_sim_fjords_hanks_sub$y2))+1

bbox <- st_bbox(c(xmin=x.min, xmax = x.max, ymin =y.min, ymax = y.max), 
                crs = st_crs(land_border))

bbox_polygon <- st_as_sfc(st_bbox(bbox)
                          , crs = st_crs(land_border))

cropped_land<- land_border%>% st_intersection(bbox_polygon)

```

```{r get final combined plot }

# Create the plots ----------------------

# Ensure all geometries have the same CRS
st_crs(cropped_land$geometry) <- st_crs("+init=EPSG:32626 +units=km")
st_crs(rect_border$geometry) <- st_crs("+init=EPSG:32626 +units=km")
st_crs(circ_border$geometry) <- st_crs("+init=EPSG:32626 +units=km")

# Combine geometries into a single sf object
geometries <- bind_rows(
  st_as_sf(data.frame(geometry = cropped_land$geometry, domain = "Fjords")),
  st_as_sf(data.frame(geometry = rect_border$geometry, domain = "Rect")),
  st_as_sf(data.frame(geometry = circ_border$geometry, domain = "Circ"))
)

# Combine all data into one dataframe
data_combined <- bind_rows(
  data_sim_fjords_cubic_attractive_sub %>% mutate(domain = "Fjords", model = "shore-attractive CVM"),
  data_sim_fjords_hanks_sub %>% mutate(domain = "Fjords", model = "shore-reflected CVM"),
  data_sim_rect_cubic_attractive_sub %>% mutate(domain = "Rect", model = "shore-attractive CVM"),
  data_sim_rect_hanks_sub %>% mutate(domain = "Rect", model = "shore-reflected CVM"),
  data_sim_circ_cubic_attractive_sub %>% mutate(domain = "Circ", model = "shore-attractive CVM"),
  data_sim_circ_hanks_sub %>% mutate(domain = "Circ", model = "shore-reflected CVM")

)

# Function to create plots with individual scales
create_plot <- function(domain, model) {
  data_filtered <- data_combined %>% filter(domain == !!domain & model == !!model)

  ggplot() +
    geom_sf(data = geometries %>% filter(domain == !!domain), aes(geometry = geometry), fill = "grey", inherit.aes = FALSE) +
    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
    geom_path(data = data_filtered, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = data_filtered, aes(x = y1, y = y2, col = ID), size = 0.1) +
    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
    theme_minimal() +
    xlab("x") + ylab("y") +
    theme(legend.position = "none")  # Remove legend from individual plots
}

# Get unique combinations of domain and model
unique_combinations <- expand.grid(model = unique(data_combined$model),domain = unique(data_combined$domain))

# Create a list of plots
plot_list <- lapply(seq_len(nrow(unique_combinations)), function(i) {
  domain <- unique_combinations$domain[i]
  model <- unique_combinations$model[i]
  create_plot(domain, model)
})

final_plot=ggarrange(plotlist=plot_list,
                     ncol=2, nrow=3, common.legend = TRUE, legend="bottom",
                     labels= c("shore-attractive CVM","shore-reflected CVM"))

ggsave("illustrative_trajectories.pdf", plot = final_plot, width = 14, height = 14)

```

```{r print final plot}
final_plot

```

### 2025-01-07 R/simulation_study_CRCVM/spline_estimation_local/simulation_study_spline.Rmd
R code updates

---
title: "simulation_study_spline"
author: "Alexandre Delporte"
date: "2024-12-01"
output: html_document
---

This is a simulation study dedicated to the estimation of the parameters in our CRCVM SDE model.
We try different framework to decide which one we choose for the calculations on the university clusters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(sf)
library(smoothSDE)
library(doParallel)
library(ggplot2)
library(dplyr)
library(mgcv)

#get functions to simulate trajectories
source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))

```

# Simple framework : no random effects

## Define hyperparameters

```{r hyperparameters}
N_ID=6
TMAX=1*24
SP_DF=c(7,4)
TAU_0=1.5
NU_0=4
DMIN=0.5
DELTA=1/360
PAR0_RACVM=c(0,0,1,1,0)
SIGMA_OBS_low=0.01
SIGMA_OBS_high=0.04
BY=30

```

```{r initial positions}

seed=100
set.seed(seed)
# Set the path to the directory containing the data
greenland_data_path <- here("Data","preprocessed_data","greenland")

#get the land geometry from shapefile
border<-st_read(file.path(greenland_data_path,"updated_scoresby_sound_utm.shp"))
border <- st_transform(border, crs = "+init=EPSG:32626 +units=km")

v0=c(0,0)

#generate random initial points
x0=matrix(rep(NA,N_ID*2),ncol=2)
colnames(x0)=c("x1","x2")

i=1
while (i<=N_ID) {
  #choose location uniformly in the map
  x=c(runif(1,min=430,max=500),runif(1,min=7760,max=7900))
  if (!(is_in_land(st_point(x),border))) {
    p=nearest_boundary_point(st_point(x),border)
    Dshore=(x[1]-p[1])^2+(x[2]-p[2])^2
    #keep it as initial position if it is at least 50 metres away from the shore
    if (Dshore>DMIN && Dshore<1) {
      x0[i,]=x
      i=i+1
    }
  }
}

```

## Define SDE parameters

```{r model parameters}
# Time steps ----------------

#high frequency time points
times=seq(0,TMAX,by=DELTA)

n_obs=length(times)-1

# Defintion of smooth parameter omega ----------

fomega_cubic=function(cov_data,a=2,D0=0.5,D1=0.8,sigma_theta=pi/8,sigma_D=0.3,b=5){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
}

ftau_constant=function(cov_data) {
  TAU_0
}

fnu_constant=function(cov_data) {
  NU_0
}

```

## Simulate true trajectories

```{r,cache=TRUE}

# Generate samples ---------------

cores=detectCores()
cl <- makeCluster(6) #not to overload your computer
registerDoParallel(cl)

data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {

  set.seed((seed-1)*N_ID+i)

  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic,fnu=fnu_constant,
                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)

  data_sim=res$sim
  data_sim$ID=factor(rep(i,length(data_sim$y1)))
  data_sim
}

#stop cluster
stopCluster(cl)

```

## Remove trajectories that reached land

```{r}

# Points that reached land ---------------
count=0
remove_ID=c()
for (id in unique(data$ID)) {
  id_data=data[data$ID==id,]
  if (nrow(id_data) < n_obs) {
    count=count+1
    remove_ID=c(remove_ID,id)
    cat("ID",id,"reached land","\n",sep=" ")
  }
}

cat(count/N_ID*100,"percent of the trajectories reached land","\n")

data_sub=data %>%  filter(!(ID %in% remove_ID))

```

## Add noise to get observed trajectories

```{r}
# add noise
low_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_low^2,2)))
high_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_high^2,2)))
observed_data_le=data_sub
observed_data_le[,c("y1","y2")]=data_sub[,c("y1","y2")]+low_noise
observed_data_he=data_sub
observed_data_he[,c("y1","y2")]=data_sub[,c("y1","y2")]+high_noise

#subsample
data_lf_he=observed_data_he[seq(1,length(data_sub$time),by=BY),]
data_lf_le=observed_data_le[seq(1,length(data_sub$time),by=BY),]
data_hf_le=observed_data_le[seq(1,length(data_sub$time),by=6),]
data_hf_he=observed_data_he[seq(1,length(data_sub$time),by=6),]

# Save plot of the trajectories ------------

plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_point(data=data_lf_he,mapping=aes(y1,y2,col=ID),size=0.1)+
  geom_path(data=data_lf_he,mapping=aes(y1,y2,col=ID),size=0.1)+
  geom_point(data = data%>% filter(!duplicated(ID)),
             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
  xlab("x") + ylab("y")

ggsave(filename=paste("plot_","border","_",TMAX,"h_",N_ID,"ID_",DMIN,"km",seed,".png",sep=""),plot=plot,width=10,height=5)

```

## Compute covariates from observations

```{r}

# Add covariates to data -----------

signed_angle <- function(u, v) {
  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
  # as in 
  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
  u <- matrix(u, ncol = 2)
  v <- matrix(v, ncol = 2)
  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
                                  rows")
  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
  ind1 <- which(result > pi)
  ind2 <- which(result <= -pi)
  result[ind1] <- result[ind1] - 2*pi
  result[ind2] <- result[ind2] + 2*pi
  return(result) 
}

add_covs=function(data) {
  #data has at least columns "time","y1" "y2", "p1","p2"
  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"

  new_data=data
  #observed deviation angles
  n=length(new_data$time)

  new_data$theta=rep(NA,n)

  for (id in unique(new_data$ID)) {

    #filter data with specific id
    sub_ind=(new_data$ID==id)
    sub_data=new_data[sub_ind,]
    n_sub=length(sub_data$time)

    #time steps
    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]

    #step lengths
    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]

    #matrix of empirical velocity
    vexp_df=cbind(dx/dtimes,dy/dtimes)

    #nearest points on shore
    sub_data=cbind(sub_data,nearest_boundary_points(as.matrix(sub_data[,c("y1","y2")]),border))

    #matrix of normal vectors
    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])

    #angle between velocity and normal vector
    theta_coast=signed_angle(normal,vexp_df)

    #adjust lengths 
    theta_coast=c(theta_coast,1)

    new_data[sub_ind,"theta"]=theta_coast

    Dshore=rep(0,n_sub)
    for (i in 1:n_sub) {
      y1=sub_data[i,"y1"]
      y2=sub_data[i,"y2"]
      if (!(is_in_land(st_point(c(y1,y2)),border))) {
        p1=sub_data[i,"p1"]
        p2=sub_data[i,"p2"]
        Dshore[i]=sqrt((y1-p1)^2+(y2-p2)^2)
      }
    }
    new_data[sub_ind,"DistanceShore"]=Dshore

  }

  return(new_data)
}

```

```{r}
data_lf_he=add_covs(data_lf_he)
data_hf_le=add_covs(data_hf_le)
data_lf_le=add_covs(data_lf_le)
data_hf_he=add_covs(data_hf_he)
```

```{r}

plot_Dshore=ggplot(data=data_lf_he)+geom_point(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.5)+
  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)+facet_wrap(~ID)+
  theme_minimal()

plot_Dshore

```

```{r}

ggplot(data_lf_he)+geom_histogram(aes(x=DistanceShore),bins=30)+facet_wrap(~ID)+theme_minimal()
quantile(data_lf$DistanceShore)
```

```{r}

data_lf_he$angular_velocity=as.vector(sapply(unique(data_sub$ID),function(id){
  c(NA,diff(data_lf_he[data_lf_he$ID==id,"theta"])/diff(data_lf_he[data_lf_he$ID==id,"time"]))}))

plot_angular_velocity=ggplot(data=data_lf_he)+geom_point(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.5)+
  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)+facet_wrap(~ID)+
  theme_minimal()
plot_angular_velocity

```

```{r}

ggplot(data_lf_he)+geom_histogram(aes(x=angular_velocity),bins=30)+facet_wrap(~ID)+theme_minimal()
quantile(data_lf_he$angular_velocity,na.rm=TRUE)

```

```{r}
data_lf_he$speed=as.vector(sapply(unique(data_sub$ID),function(id) {c(NA,sqrt(diff(data_lf_he[data_lf_he$ID==id,"y1"])^2+diff(data_lf_he[data_lf_he$ID==id,"y2"])^2)
                                                                      /diff(data_lf_he[data_lf_he$ID==id,"time"]))}))

ggplot(data_lf_he)+geom_histogram(aes(x=speed),bins=30)+facet_wrap(~ID)+theme_minimal()
quantile(data_lf_he$speed,na.rm=TRUE)
```

## Estimate CTCRW

### low frequency high error

First we estimate $\tau$ and $\nu$ while ignoring the spatial constraints that are included in $\omega$.
We only keep 2 of the twelve trajectories.

```{r fit sde model with splines,cache=TRUE}

formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)

ctcrw_lf_he<- SDE$new(formulas = formulas,data =data_lf_he[data_lf_he$ID %in% c(1,2,3),],type = "CTCRW",
                    response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

#fit
ctcrw_lf_he$fit(method="BFGS")
```
```{r}
ctcrw_lf_he$get_all_plots()
exp(as.list(ctcrw_lf_he$tmb_rep(),what="Estimate")$log_sigma_obs)
```

It seems that the estimate for $\tau$ is biaised. While $\nu$ is well estimated, 
$\tau$ seems to be significantly underestimated. The reason is probably that some of the turns in the trajectories that are due to the parameter $\omega$ are attributed to a lower persistence since we do not model $\omega$ here. This is indeed in agreement with the results we get from fitting a CTCRW to the narwhal data. The trajectories from the fitted model seemed Here, if we play with the number of trajectories used for the estimation, we see that for some trajectories we are able to get both $\tau$ and $\nu$ right. The parameter $\sigma_{obs}$ for the measurement error is also well estimated.

### Low frequency low error

```{r fit sde model with splines,cache=TRUE}

formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)

ctcrw_lf_le<- SDE$new(formulas = formulas,data =data_lf_le[data_lf_le$ID %in% c(1,2,3),],type = "CTCRW",
                    response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.03)))

#fit
ctcrw_lf_le$fit(method="BFGS")
```
```{r}
ctcrw_lf_le$get_all_plots()
exp(as.list(ctcrw_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)

```

We are still estimating $\nu$ and $\sigma_{obs}$ right, but $\tau$ is also still underestimated. Thus, reducing the measurement error does not help get a better estimate of $\tau$ here.

### High frequency low error

```{r fit sde model with splines,cache=TRUE}

formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)

ctcrw_hf_le<- SDE$new(formulas = formulas,data =data_hf_le,type = "CTCRW",
                    response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.03)))

#fit
ctcrw_hf_le$fit(method="BFGS")
```
```{r}
ctcrw_hf_le$get_all_plots()

exp(as.list(ctcrw_hf_le$tmb_rep(),what="Estimate")$log_sigma_obs)

```

The estimate is slightly less biased here, but still increasing the resolution of the data does not really help to correct the biais on $\tau$.

### High frequency high error

If we take all the data points, we get numerical errors during the estimation with a non definite positive Hessian matrix.

```{r fit sde model with splines,cache=TRUE}

formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)

ctcrw_hf_he<- SDE$new(formulas = formulas,data =data_hf_he[data_hf_he$ID %in% c(1,2,3),],type = "CTCRW",
                    response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

#fit
ctcrw_hf_he$fit(method="BFGS")
```
```{r}
ctcrw_hf_he$get_all_plots()

exp(as.list(ctcrw_hf_he$tmb_rep(),what="Estimate")$log_sigma_obs)

```

Whether we have low error or high error, we still have a biased estimate of $\tau$.
We now introduce the parameter $\omega$ in the model.

## Estimate RACVM with splines

Now we can try to add parameter $\omega$ in the model.
We are limited by the RAM of the computer that is why we would need to do this on the university clusters with longer or higher frequency trajectories.
We need to fix the smoothing penalties otherwise they are either estimated to very high or very low values meaning that numerical errors occurred.

First, we get good values for the smoothing penalty by approximating the true surface with splines.

```{r}
# Splines smooth for omega -------------------- 
n <- 100000
SP_DF=5

#sample theta and DistanceShore points in the domain
theta <- runif(n,-pi,pi);
DistanceShore <- runif(n,0.1,2)
samples=data.frame(theta=theta,DistanceShore=DistanceShore)

#define grid of values
theta_v <- seq(-pi,pi,length=30)
Dshore_v<- seq(0.2,8,length=30)
pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))

#points on the surface perturbed by gaussian noise
f <- fomega_cubic(samples)
y <- f+0.1*rnorm(n)

#fit with bivariate splines te
m1 <- gam(y~te(theta,DistanceShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
                                                       DistanceShore=seq(0.02,7,len=SP_DF)))

ExpShore=1/DistanceShore

#fit with bivariate splines te
m2 <- gam(y~te(theta,ExpShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
                                                  ExpShore=seq(1/8,1/0.2,len=SP_DF)))

fomega_cubic_splines=function(cov_data) {

  if (is.null(cov_data$ExpShore)){
    omega=predict(m1,newdata=cov_data)
  }
  else if (is.null(cov_data$DistanceShore)){
    omega=predict(m2,newdata=cov_data)
  }

  return(as.numeric(omega))
}

m1$sp
```

 ### Low frequency high error

```{r fit sde model with splines,cache=TRUE}

cropped_data=data_lf_he[data_lf_he$ID %in% 1:3,]

H=array(rep(SIGMA_OBS_high^2*diag(2),length(cropped_data$time)),dim=c(2,2,length(cropped_data$time)))

formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))

crcvm_splines_lf_he<- SDE$new(formulas = formulas,data = cropped_data,type = "RACVM_SSM",
                    response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

crcvm_splines_lf_he$update_map(list("log_lambda"=factor(rep(NA,2))))
crcvm_splines_lf_he$update_lambda(c(10^-7,10^-13))
#fit
crcvm_splines_lf_he$fit(method="BFGS")
```

```{r plot estimates}

crcvm_splines_lf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))

exp(as.list(crcvm_splines_lf_he$tmb_rep(),what="Estimate")$log_sigma_obs)

```

We are clearly underestimating the magnitude of the rotation $\omega$ near $-\pi$ and $\pi$. This is indeed very similar to the result we got for the narwhal GPS data. We are well estimating $\nu$ and $\sigma_{obs}$, and we corrected the bias in $\tau$ compared to the estimate we got for the CTCRW on the same data.

### Low frequency low error

```{r fit sde model with splines,cache=TRUE}

formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=c(10,5),bs="cs"))

crcvm_splines_lf_le<- SDE$new(formulas = formulas,data = data_lf_le,type = "RACVM_SSM",
                    response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.03)))

crcvm_splines_lf_le$update_map(list("log_lambda"=factor(rep(NA,2))))
crcvm_splines_lf_le$update_lambda(c(10^-12,10^-12))
#fit
crcvm_splines_lf_le$fit(method="BFGS")
```

```{r plot estimates}

crcvm_splines_lf_le$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))

exp(as.list(ctcrw_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)

```

Again, having low measurement error is not really of any help in the estimation. It is probably even more difficult to estimate since errors can occur in the matrix inversions in the Kalman filter when the coefficients in the covariance matrix are too close to $0$.

### High frequency high error

```{r fit sde model high frequency,cache=TRUE}

cropped_data=data_hf_he[data_hf_he$ID %in% 2:2,]

formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))

crcvm_splines_hf_he<- SDE$new(formulas = formulas,data = cropped_data,type = "RACVM_SSM",
                    response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

#initialize smoothing penalties and re variances
new_lambda=c(10^-6,10^-4)
crcvm_splines_hf_he$update_lambda(new_lambda)

crcvm_splines_hf_he$update_map(list(log_lambda=factor(c(NA,NA))))

#fit
crcvm_splines_hf_he$fit(method="BFGS")
```

```{r}

plot_results_hf=crcvm_splines_hf$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))

plot_results_hf

```

For high frequency data, we are able to get $\tau$ and $\nu$ right, but we are still underestimating $\omega$.
Now we force the parametrization of the surface for $\omega$ with a new SDE model in smoothSDE (called \texttt{CRCVM_SSM}. $\omega$ is a cubic function of the angle $\theta$, with an exponential factor.

### High frequency low error

```{r fit sde model high frequency,cache=TRUE}

cropped_data=data_hf_he[data_hf_le$ID %in% 1,]

H=array(rep(SIGMA_OBS^2*diag(2),length(data_hf$time)),dim=c(2,2,length(data_hf$time)))

formulas <- list(mu1=~1,mu2=~1,tau=~1,
                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))

crcvm_splines_hf<- SDE$new(formulas = formulas,data = data_hf,type = "RACVM_SSM",
                    response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
                    other_data=list("H"=H))

#initialize smoothing penalties and re variances
new_lambda=c(exp(-1),exp(-1))
crcvm_splines_hf$update_lambda(new_lambda)

crcvm_splines_hf$update_map(list(log_lambda=factor(c(NA,NA))))

#fit
crcvm_splines_hf$fit(method="BFGS")
```

```{r}

plot_results_hf=crcvm_splines_hf$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))

plot_results_hf

```

## Estimation with fixed cubic function of angle

### Low frequency high error

We will fix the new parameters that are introduced in the model, and estimate the others, that is $\tau$ and $\nu$.
These new parameters might be chosen based on the observed distance to shore and observed angular velocity in the data.

```{r}
H=array(rep(0.03^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
# Estimate from simulated data ------------------

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_lf_he<- SDE$new(formulas = formulas,data = data_lf_he[data_lf_he$DistanceShore>0.01,],type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,1,5,0.5,0.8,0.3,pi/8),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("log_sigma_obs0"=log(0.02)))
#fit
crcvm_cubic_lf_he$fit(method="BFGS")
```

```{r}

crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")
exp(as.list(crcvm_cubic_lf_he$tmb_rep(),what="Estimate")$log_sigma_obs)

```

We are unable to estimate $\tau$ while fixing $\omega$. $\tau$ is totally biased and underestimated, even more than when we neglected the constraints and used a simple CTCRW..

### Low frequency low error

```{r}
H=array(rep(0.007^2*diag(2),length(data_lf_le$time)),dim=c(2,2,length(data_lf_le$time)))
# Estimate from simulated data ------------------

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_lf_le<- SDE$new(formulas = formulas,data = data_lf_le,type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1.5,4,1,5,1,0.8,0.2,pi/8),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("log_sigma_obs0"=log(0.03)))
#fit
crcvm_cubic_lf_le$fit(method="BFGS")
```
```{r}

crcvm_cubic_lf_le$get_all_plots(show_CI="simultaneous")
exp(as.list(crcvm_cubic_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)

```
Let's now try with high frequency data.

### High frequency high error

```{r}

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_he<- SDE$new(formulas = formulas,data = data_hf_he[data,type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 =c(1,1,1,5,0.5,0.8,0.3,pi/8),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("log_sigma_obs0"=log(0.05)))

#fit
crcvm_cubic_hf_he$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_he$get_all_plots()
exp(as.list(crcvm_cubic_hf_he$tmb_rep(),what="Estimate")$log_sigma_obs)
```
### High frequency low error

```{r}

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_le<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% 4:5,],type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,40,5,0.15,0.8,0.2,pi/8),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("log_sigma_obs0"=log(0.03)))

#fit
crcvm_cubic_hf_le$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_le$get_all_plots()
exp(as.list(crcvm_cubic_hf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
```

```{r}
# Definition of parameters tau and nu ----------------
tau_re=rnorm(N_ID,mean=0,sd=SIGMA_TAU)
nu_re=rnorm(N_ID,mean=0,sd=SIGMA_NU)
true_log_nu=nu_re+log(NU_0)

ftau_gaussian=function(cov_data,sigma_bump=0.5,sigma_dip=0.5,tau0=TAU_0,dtau_bump=1.5,dtau_dip=0) {
  #cov_data : dataframe of covariates with column for angle theta
  #sigma_bump : width of the bumps at -pi/2 and pi/2
  #sigma_dip : width of the dip at 0
  #tau0 : base value of tau
  #dtau_bump : height of the bump
  #dtau_dip : depth of the dip

  theta=cov_data$theta
  bump1 <- dtau_bump*exp(-((theta - pi/2)^2) / (2 * sigma_bump^2))
  bump2 <- dtau_bump*exp(-((theta + pi/2)^2) / (2 * sigma_bump^2))
  dip<- dtau_dip*exp(-((theta)^2) / (2 * sigma_dip^2))
  tau=tau0+bump1+bump2-dip

  return (tau)
}

```

### 2025-01-08 R/simulation_study_CRCVM/CRCVM_examples/illustrative_trajectories.Rmd
new set up for simulation study on clusters

@@
 
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_minimal,fnu=fnu_constant,v0=v0,x0=x0[i,],
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_tanh,fnu=fnu_constant,v0=v0,x0=x0[i,],
                     times=times,land=rect_border,verbose=FALSE)
@@
 
-```{r simulate with omega cubic in rectangle,cache=TRUE}
+```{r simulate with omega cubic attractive in rectangle,cache=TRUE}
 
@@
 
-```{r simulate with omega cubic in rectangle,cache=TRUE}
+```{r simulate with omega cubic repulsive in rectangle,cache=TRUE}
 

### 2025-01-08 R/simulation_study_CRCVM/spline_estimation_local/simulation_study_spline.Rmd
new set up for simulation study on clusters

@@
 ```
+```{r}
+ggplot(data_lf_he)+geom_histogram(aes(x=theta),bins=30)+facet_wrap(~ID)+theme_minimal()
+```
 
@@
 
-cropped_data=data_lf_he[data_lf_he$ID %in% 1:3,]
+H=array(rep(SIGMA_OBS_high^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
 
-H=array(rep(SIGMA_OBS_high^2*diag(2),length(cropped_data$time)),dim=c(2,2,length(cropped_data$time)))
-
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
+formulas <- list(mu1=~1,mu2=~1,tau=~s(theta,k=8,bs="cc"),
                  nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
@@
 
-crcvm_splines_lf_he$update_map(list("log_lambda"=factor(rep(NA,2))))
-crcvm_splines_lf_he$update_lambda(c(10^-7,10^-13))
+crcvm_splines_lf_he$update_map(list("log_lambda"=factor(c(1,rep(NA,2)))))
+crcvm_splines_lf_he$update_lambda(c(1,10^-7,10^-13))
 #fit

### 2025-01-08 R/simulation_study_CRCVM/spline_estimation_cluster/circ/circ_Ishore71.R

@@ -1,0 +1,571 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 71
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-01-08 R/simulation_study_CRCVM/spline_estimation_cluster/rect/rect_Ishore3.R

@@ -1,0 +1,571 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 3
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-01-12 R/simulation_study_CRCVM/CRCVM_examples/illustrative_trajectories.Rmd
fix measurement error in simulation study

@@
 
-ftau_gaussian_fjords=function(cov_data) {
-  ftau_gaussian(cov_data,sigma_bump=0.4,sigma_dip=0.4,tau0=1,dtau_bump=1.5,dtau_dip=0.5
-                )
-  }
-
 # Generate samples ---------------
@@
 ```{r simulate hanks reflected in rectangle,cache=TRUE}
-
 # Generate samples ---------------
@@
 stopCluster(cl)
-
+```
+
+```{r simulate hanks reflected in rectangle,cache=TRUE}
 ```
@@
 
-x.min= min(c(data_sim_fjords_cubic_repulsive_sub$y1,data_sim_fjords_cubic_attractive_sub$y1,
+x.min= min(c(data_sim_fjords_cubic_attractive_sub$y1,
              data_sim_fjords_hanks_sub$y1))-1
-x.max = max(c(data_sim_fjords_cubic_repulsive_sub$y1,data_sim_fjords_cubic_attractive_sub$y1,
+x.max = max(c(data_sim_fjords_cubic_attractive_sub$y1,
               data_sim_fjords_hanks_sub$y1))+1
-y.min = min(c(data_sim_fjords_cubic_repulsive_sub$y2,data_sim_fjords_cubic_attractive_sub$y2,
+y.min = min(c(data_sim_fjords_cubic_attractive_sub$y2,
               data_sim_fjords_hanks_sub$y2))-1
-y.max = max(c(data_sim_fjords_cubic_repulsive_sub$y2,data_sim_fjords_cubic_attractive_sub$y2,
+y.max = max(c(data_sim_fjords_cubic_attractive_sub$y2,
               data_sim_fjords_hanks_sub$y2))+1
@@
     xlab("x") + ylab("y") +
-    theme(legend.position = "none")  # Remove legend from individual plots
+    theme(legend.position = "none") +
+  theme(
+    axis.title = element_text(size = 16),
+    axis.text = element_text(size = 12),
+    legend.text = element_text(size = 16),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
 }
@@
 ```
+
+```{r}
+
+#plot only trajectory in fjords
+
+data_combined_fjords=data_combined[data_combined$domain=="Fjords",]
+
+# Create a list of plots
+plot_list <- lapply(seq_len(length(unique(data_combined$model))), function(i) {
+  model <- unique_combinations$model[i]
+  create_plot(domain="Fjords",model)
+})
+
+final_plot_fjords=ggarrange(plotlist=plot_list,
+                     ncol=1, nrow=2, common.legend = TRUE, legend="bottom",
+                     labels= c("shore-attractive CVM","shore-reflected CVM"))
+
+final_plot_fjords
+
+ggsave("illustrative_trajectories_fjords.pdf", plot = final_plot_fjords, width = 14, height = 14)
+
+```

### 2025-01-19 R/preprocessing/narwhals/data_preprocessing.Rmd
redo preprocessing with shrunk polygons land to include new covariates DistanceShore_translated and theta_translated

@@
 
-# Read the raw data
+# Explore raw data
 
 ## Read
+Takes a few minutes to read the whole dataset.
 
@@
 
-# First data exploration
-
-## Trials and intertrials periods
+# Trials and intertrials periods exploration
 
@@
                     name="Trial type",labels=c("Intertrial","Trial"))+
-  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))
+  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))+theme_minimal()
 
@@
                      name="Trial type",labels=c("Intertrial","Trial"))+
-  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))
+  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))+theme_minimal()
 
@@
 pExpAll=ggplot(data,aes(time,ExpShip,col=ID))+geom_line()+
-  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
+  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()
 
@@
 pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
-  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
+  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()
 
@@
 pExpTr=ggplot(data,aes(time,ExpShipT,col=ID))+geom_line()+
-  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
+  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()
 
@@
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")
+  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")+theme_minimal()
 
@@
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Time step (min) ")
+  theme(legend.position="top")+xlab("Time step (min) ")+theme_minimal()
 
@@
 ```{r display hist , fig.cap="Histogram of time steps",warnings=FALSE}
-all_times_histo=ggplot(data=dtimes,aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")
+all_times_histo=ggplot(data=dtimes,aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")+
+  theme_minimal()
 all_times_histo
@@
 
-## Time steps distribution during exposure vs no exposure
+```{r}
+all_times_histo_cropped=ggplot(data=dtimes[dtimes$delta<1,],aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")+
+  theme_minimal()
+all_times_histo_cropped
+ggsave(path=plots_folder,"all_time_steps_histo_cropped.png", plot = all_times_histo_cropped)
+```
 
@@
              aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
-  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
+  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()
 
@@
 
-We read the preprocessed land polygons extracted from OpenStreetMap database.
+We read the preprocessed land polygons extracted from OpenStreetMap database without modification except for cropping the polygons in the right bounding box and projecting it in EPSG 32626 to get UTM coordinates in km.
 
@@
 summary(dataprep$DistanceShore)
-quantile(res_land$min_distance,probs=c(0.05,0.95))
-D_low=quantile(res_land$min_distance,probs=0.05)
-D_up=median(res_land$min_distance)
 ```
@@
 histo1=ggplot(dataprep[dataprep$DistanceShore>0,], aes(x = DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")+
+  theme_minimal()
 
@@
 histo2=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")+
+  theme_minimal()
 ```
@@
 ```{=tex}
-
-For the rest of the analysis, we'll keep the bounded domain defined by the updated polygons in "updated_scoresby_sound_utm" sf object as a reference.
-Doing that involves having some points on land, that is points with Distance to shore equal to 0.
+A lot of points are indeed still very close to land.
+We think that this means the shoreline is still not accurate.
+To solve the problem, we translate the polygon vertices 200m inward.
+We compute new distances to shore based on the translated polygon.
+
+```
+
+```{r get shrunk polygons}
+
+#get updated land polygons
+shrunk_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk_updated_scoresby_sound_utm.shp"))
+
+shrunk_scoresby_sound_utm <- st_transform(shrunk_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
+
+```
+
+```{r compute updated translated distance to shore}
+res_shrunk_land=nearest_points(positions=narwhal_positions,border=shrunk_scoresby_sound_utm)
+
+cat("Percentage of points on land with distance from shrunk updated osm land polygons : \n",
+    length(which(res_shrunk_land$min_distances==0))/length(dataprep$time)*100,"\n")
+
+```
+
+```{r distribution of points close to the shore}
+
+df=data.frame(DistanceShore=res_shrunk_land$min_distances[res_shrunk_land$min_distances<1 & res_shrunk_land$min_distances>0  ])
+histo=ggplot(df, aes(x =DistanceShore)) +
+  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")
+
+histo
 ```
@@
 
+#same for shrunk polygons
+nearest_points_shrunk=res_shrunk_land$nearest_points
+min_distances_shrunk=res_shrunk_land$min_distances
+
+#add columns px and py with coordinates of the nearest point on the shore line
+dataprep$px_translated=nearest_points_shrunk[,1]
+dataprep$py_translated=nearest_points_shrunk[,2]
+
+# add columns with difference between narwhal position and nearest point on the shoreline
+dataprep$nx_translated=dataprep$x-dataprep$px_translated
+dataprep$ny_translated=dataprep$y-dataprep$py_translated
+dataprep$DistanceShore_translated=min_distances_shrunk
 ```
@@
 dataprep$Ishore <- ExposureFunction(X = dataprep$DistanceShore)
+
+#add Ishore covariate based on 1/distance in km
+dataprep$Ishore_translated <- ExposureFunction(X = dataprep$DistanceShore_translated)
 
@@
 dataprep$AngleVelocities=rep(NA,n)
-dataprep$AngleNormal=rep(NA,n)
+dataprep$theta=rep(NA,n)
+dataprep$theta_translated=rep(NA,n)
 dataprep$VelocityPhase=rep(NA,n)
@@
   normal=as.matrix(sub_data[2:n_sub,c("nx","ny")])
+  normal_translated=as.matrix(sub_data[2:n_sub,c("nx_translated","ny_translated")])
 
   #angle between velocity and normal vector
-  theta_coast=signed_angle(vexp_df,normal)
+  theta=signed_angle(vexp_df,normal)
+  theta_translated=signed_angle(vexp_df,normal_translated)
 
   #deviation angle between consecutive velocities
-  theta_vel=signed_angle(vexp_df[1:(n_sub-2),],vexp_df[2:(n_sub-1),])
+  angle_vel=signed_angle(vexp_df[1:(n_sub-2),],vexp_df[2:(n_sub-1),])
 
@@
   #deviation from trajectory parallel to coastline
-  phi=ifelse(theta_coast > 0, theta_coast - pi/2, theta_coast + pi/2)
+  phi=ifelse(theta > 0, theta - pi/2, theta + pi/2)
 
   #adjust lengths by adding NA when necessary
-  theta_vel=c(NA,theta_vel,NA)
+  angle_vel=c(NA,angle_vel,NA)
   phi=c(NA,phi)
-  theta=c(NA,theta_coast)
+  theta=c(NA,theta)
+  theta_translated=c(NA,theta_translated)
   vel_phase=c(NA,vel_phase,NA)
@@
   dataprep[sub_ind,"AngleShore"]=phi
-  dataprep[sub_ind,"AngleVelocities"]=theta_vel
-  dataprep[sub_ind,"AngleNormal"]=-theta
+  dataprep[sub_ind,"AngleVelocities"]=angle_vel
+  dataprep[sub_ind,"theta"]=-theta
+  dataprep[sub_ind,"theta_translated"]=-theta_translated
   dataprep[sub_ind,"VelocityPhase"]=vel_phase
@@
   facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical velocity")
+  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical velocity")+
+   theme_minimal()
 
@@
   facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")
+  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")+
+  theme_minimal()
 
@@
   facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation between consecutive velocities")
-
-plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleNormal,col=ID))+
+  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation between consecutive velocities")+ theme_minimal()
+
+plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,theta,col=ID))+
   geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
   facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Angle between empirical velocity and normal vector")
+  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Angle between empirical velocity and normal vector")+ theme_minimal()
 
@@
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")
+  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")+ theme_minimal()
 
@@
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Empirical deviation between consecutive velocities (rad) ")
-
-AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleNormal,col=ID)) +
+  theme(legend.position="top")+xlab("Empirical deviation between consecutive velocities (rad) ")+ theme_minimal()
+
+AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = theta,col=ID)) +
   geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Angle between empirical velocity and normal vector (rad) ")
+  theme(legend.position="top")+xlab("Angle between empirical velocity and normal vector (rad) ")+ theme_minimal()
 
@@
   scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Velocity phase angele (rad) ")
+  theme(legend.position="top")+xlab("Velocity phase angele (rad) ")+ theme_minimal()
 
@@
 
-We also plot the density of the Dist_Ship variable to understand better its distribution.
+We also plot the histogram of the Dist_Ship variable to understand better its distribution.
 
@@
 
-dataBE1=data.frame()
+dataBE24=data.frame()
 for (id in unique(data$ID)) {
@@
   maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
-  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
-}
-
-dataBE2=data.frame()
+  dataBE24 =rbind(dataBE24,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
+}
+
+dataBE12=data.frame()
 for (id in unique(data$ID)) {
@@
   maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
-  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
+  dataBE12 =rbind(dataBE12,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
 }
@@
 for (id in unique(data$ID)) {
-  narwhalDataBE=dataBE2[dataBE2$ID==id,]
+  narwhalDataBE=dataBE12[dataBE12$ID==id,]
   narwhalDataAE=dataAE[dataAE$ID==id,]
@@
 ```{r plots before and after exposure}
-pBE1=milne_land_map + 
-  geom_point(dataBE1[,c("ID","Long","Lat","time")],
+pBE24=milne_land_map + 
+  geom_point(dataBE24[,c("ID","Long","Lat","time")],
              mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
-   geom_path(dataBE1[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
-  geom_point(data = dataBE1[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
+   geom_path(dataBE24[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
+  geom_point(data = dataBE24[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
              aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
@@
 
-pBE2=milne_land_map + 
-  geom_point(dataBE2[,c("ID","Long","Lat","time")],
+pBE12=milne_land_map + 
+  geom_point(dataBE12[,c("ID","Long","Lat","time")],
              mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
-   geom_path(dataBE2[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
-  geom_point(data = dataBE2[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
+   geom_path(dataBE12[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
+  geom_point(data = dataBE12[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
              aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
@@
 
-ggsave(path=plots_folder,"tracksBE1.png",pBE1,width=10,height=5)
-ggsave(path=plots_folder,"tracksBE2.png",pBE2,width=10,height=5)
+ggsave(path=plots_folder,"tracksBE12.png",pBE12,width=10,height=5)
+ggsave(path=plots_folder,"tracksBE24.png",pBE24,width=10,height=5)
 ggsave(path=plots_folder,"tracksAE.png",pAE,width=10,height=5)
@@
 ```{r display pBE1, fig.cap="Tracks before exposure, 24h after tagging"}
-pBE1
+pBE12
 ```
@@
 ```{r display pBE2, fig.cap="Tracks before exposure, 12h after tagging"}
-pBE2
+pBE24
 ```
@@
 write.csv(dataprep,file.path(preprocessed_narwhal_data_path,"allData.csv"))
-write.csv(dataBE1,file.path(preprocessed_narwhal_data_path,"DataBE1.csv"))
-write.csv(dataBE2,file.path(preprocessed_narwhal_data_path,"DataBE2.csv"))
+write.csv(dataBE12,file.path(preprocessed_narwhal_data_path,"DataBE12.csv"))
+write.csv(dataBE24,file.path(preprocessed_narwhal_data_path,"DataBE24.csv"))
 write.csv(dataAE,file.path(preprocessed_narwhal_data_path,"DataAE.csv"))

### 2025-01-19 R/application/CRCVM/fit_crcvm.Rmd
try fit with DistanceShore_translated and theta_translated

@@
 #Fixed measurement error
-sigma_obs=0.05
+sigma_obs=0.03
 H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))
@@
 formulas <- list(tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~s(theta,k=6,bs="cs"),D0=~1)
+                   a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
 
@@
 baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
-                    response = c("x","y"),par0 = c(2,4,0,0.05),other_data=list("H"=H),
-                    fixpar=c("D0"))
+                    response = c("x","y"),par0 = c(1,4,0.1,4,0.2,0.5,0.1,pi/6),other_data=list("H"=H),
+                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"))
 
-baseline_crcvm_cubic$update_lambda(c(exp(3),exp(3),10^-12))
-baseline_crcvm_cubic$update_map(list("log_lambda"=factor(c(1,2,NA))))
 baseline_crcvm_cubic$fit()

### 2025-01-19 R/application/RCVM/fit_rcvm.Rmd
try fit with DistanceShore_translated and theta_translated

@@
 #data before exposure with first 24h removed
-dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")
+dataBE24=read.csv(file.path(narwhal_data_path,"DataBE24.csv"), header = TRUE,dec = ".")
 
 #data before exposure with first 12h removed
-dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")
+dataBE12=read.csv(file.path(narwhal_data_path,"DataBE12.csv"), header = TRUE,dec = ".")
 
@@
 
-D_low=0.073
+D_low=0.05
 D_up=3
@@
 
+```{r}
+
+all_data <- data.frame()
+
+for (id in unique(dataBE12$ID)) {
+
+  sub_dataBE <- dataBE12[dataBE12$ID == id, ]
+  sub_dataAE <- dataAE[dataAE$ID == id, ]
+
+  sub_data <- rbind(sub_dataBE, sub_dataAE)
+
+  all_data <- rbind(all_data, sub_data)
+}
+
+```
+
 # Fit baseline models
@@
 
-The first model is a CTCRW where the covariate AngleNormal influences the persistence of the movement.
+The first model is a CTCRW where the covariate theta influences the persistence of the movement.
 
@@
 #model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))
+formulas <- list(mu1=~1,mu2=~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))
 
@@
                           par0 = par0,
-                          other_data=list("H"=H),
+                          other_data=list("log_sigma_obs0"=log(0.05)),
                           fixpar=c("mu1","mu2"))
@@
 ```{r plot smooth parameters baseline_ctcrw0 }
-baseline_ctcrw0$plot_par(var="AngleNormal",
+baseline_ctcrw0$plot_par(var="theta_translated",
                          par_names=c("tau","nu"),n_post=100)
@@
 formulas <- list(mu1=~1,mu2=~1,
-                 tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),
+                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
                  nu=~s(ID,bs="re"),
-                 omega=~s(AngleNormal,k=10,bs="cs"))
+                 omega=~s(theta_translated,k=10,bs="cs"))
 
@@
 # Fit baseline with measurement error estimated from the data
-baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
+baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
+                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
 
@@
 ```{r plot smooth parameters baseline_racvm0}
-baseline_racvm0$plot_par(var="AngleNormal",
+baseline_racvm0$plot_par(var="theta_translated",
                          par_names=c("omega","tau"),n_post=100)
@@
 
-## Baseline with AngleNormal and Ishore in omega and tau
+## Baseline with AngleNormal and DistanceShore in omega and tau
 
@@
 #parameters formulas
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
-                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))
+formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))
 
 # Fit baseline
-baseline_racvm1_unstable<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
+baseline_racvm1_unstable<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
@@
 
+First, we get good values for the smoothing penalty by approximating the true surface with splines.
+
+```{r}
+
+fomega=function(cov_data,a=2,b=5,D0=0.5,D1=0.8,sigma_D=0.3,sigma_theta=pi/6){
+  Dshore=cov_data$DistanceShore
+  theta=cov_data$theta
+  if (is.null(Dshore)){
+    Dshore=1/cov_data$Ishore
+  }
+
+  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
+      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
+           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
+}
+
+# Splines smooth for omega -------------------- 
+n <- 100000
+SP_DF=5
+
+#sample theta and DistanceShore points in the domain
+theta <- runif(n,-pi,pi);
+DistanceShore <- runif(n,0.02,7)
+samples=data.frame(theta=theta,DistanceShore=DistanceShore)
+
+#define grid of values
+theta_v <- seq(-pi,pi,length=30)
+Dshore_v<- seq(0.02,7,length=30)
+pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
+
+#points on the surface perturbed by gaussian noise
+f <- fomega(samples)
+y <- f+0.1*rnorm(n)
+
+#fit with bivariate splines te
+m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))
+
+```
+
 ```{r fit baseline_racvm1,cache=TRUE}
@@
 par0 <- c(0,0,1,4,0)
-init_lambda=exp(c(3,3,4,-2,-2,-2,-2))
+init_lambda=c(exp(3),exp(3),1,exp(-12),exp(-5))
 
 #mapping for fixed smoothing penalties
-new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,4))))
+new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,2))))
 
@@
 #parameters formulas
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(AngleNormal,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
-                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))
+formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~te(theta_translated,DistanceShore_translated,k=c(5,5),bs="cs"))
 
 # Fit baseline
-baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
+                    response = c("x","y"),par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                     fixpar=c("mu1","mu2"))
@@
 
+```
+
+```{r results baseline_racvm1}
+
+estimates_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Est")
+std_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Std")
 ```
@@
 
-xmin=list("Ishore"=1/D_up)
-xmax=list("Ishore"=1/D_low)
-link=list("Ishore"=(\(x) 1/x))
-xlabel=list("Ishore"="Distance to shore")
+xmin=list("DistanceShore_translated"=0.2)
+xmax=list("DistanceShore_translated"=2)
 
 plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
-      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+      xmin=xmin, xmax=xmax,show_CI="pointwise")
 
@@
 par0 <- c(0,0,2,4,0)
-init_lambda=c(exp(4),exp(4),10^-6,10^-1)
+init_lambda=c(exp(3),exp(3),exp(-12),exp(-5))
 
@@
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~te(AngleNormal,DistanceShore,k=5,bs="cs"))
+                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))
 
@@
 
-xmin=list("Ishore"=1/D_up)
-xmax=list("Ishore"=1/D_low)
-link=list("Ishore"=(\(x) 1/x))
-xlabel=list("Ishore"="Distance to shore")
+xmin=list("DistanceShore_translated"=0.2)
+xmax=list("DistanceShore_translated"=4)
 
 plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
-      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+      xmin=xmin,xmax=xmax,show_CI="pointwise")
 
 plots_bas_racvm2
+
+```
+
+```{r}
+
+estimate_parametric_omega=function(model) {
+
+    data=model$data()
+
+    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),to=quantile(data$DistanceShore_translated,0.95),length.out=30)
+    theta_translated=seq(from=-pi,to=pi,length.out=30)
+    grid<- as.data.frame(expand.grid(Dshore,theta_translated))
+    colnames(grid) <- c("DistanceShore_translated","theta_translated")
+    grid$ID="Asgeir"
+
+    #get model matrices
+    mats=model$make_mat(new_data=grid)
+    X_fe=mats$X_fe
+    X_re=mats$X_re
+
+    par_mat=model$par(new_data=grid,X_fe=X_fe,X_re=X_re)
+
+    #matrix of values for surface plot
+    z=matrix(par_mat[,"omega"],30,30)
+
+    data=grid
+    data$z=as.vector(z)
+
+    # Create the 3D surface plot using plot_ly
+    plot_splines<- plot_ly(x = ~theta_translated, y = ~Dshore, z = ~z) %>%
+      add_surface() %>%
+      layout(title = "Predicted spline surface for omega",
+             scene = list(yaxis = list(title = "Distance to shore"),
+               xaxis = list(title = "theta_translated")))
+
+    fit_optim <- optim(par = c(0.1,2),
+                       fn = function(params) {
+
+                        # Compute predicted z values
+                        z_pred <- fomega(data,1,0.5,params[1],params[2],1,pi/6)
+
+                         # Return sum of squared residuals
+                         sum((data$z - z_pred)^2)
+                       })
+
+    est_par=fit_optim$par
+
+    print(est_par)
+
+    # Compute the predicted values using the starting parameters
+    z_pred_starting <- fomega(grid,1,1,est_par[1],est_par[2],1,
+                                    pi/6)
+
+    # Reshape z_pred_starting to a matrix for plotting
+    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta_translated))
+
+    print(z_pred_matrix)
+
+    # Create the 3D surface plot using plot_ly
+    plot_parametric <- plot_ly(x = ~theta_translated, y = ~Dshore, z = ~z_pred_matrix) %>%
+      add_surface() %>%
+      layout(title = "Estimated parametric surface for omega",
+             scene = list(
+               yaxis = list(title = "Distance to shore"),
+               xaxis = list(title = "theta"),
+               zaxis = list(title = "Predicted z")
+             ))
+
+    return (list("est_par"=est_par,"splines_surface"=plot_splines,"parametric_surface"=plot_parametric))
+
+}
+
+estimate_parametric_omega(baseline_racvm2)
 
@@
 This may be due to several factors : the low frequency at which the data is observed, the measurement error, the small amount of data that we have.
-
-We can try to fit an interpretable parametric surface on the estimated spline surface for the parameter $\omega$ in the models \texttt{baseline\_racvm1} or \texttt{baseline\_racvm2}.
-However the surfaces are so different that it does not work very well.
-
 ```{r}
 
-Ishore=seq(from=1/6,to=1/0.05,length.out=30)
-theta=seq(from=-pi,to=pi,length.out=30)
-grid<- as.data.frame(expand.grid(Ishore,theta))
-colnames(grid) <- c("Ishore","AngleNormal")
-grid$ID="Asgeir"
-
-#get model matrices
-mats=baseline_racvm1$make_mat(new_data=grid)
-X_fe=mats$X_fe
-X_re=mats$X_re
-
-par_mat=baseline_racvm1$par(new_data=grid,X_fe=X_fe,X_re=X_re)
-
-#matrix of values for surface plot
-z=matrix(par_mat[,"omega"],30,30)
-
-# Define the parametric function for omega
-fomega<- function(cov_data, a =5,theta0=pi/2,D0=1) {
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$AngleNormal
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$Ishore
+cores=detectCores()
+cl <- makeCluster(6) #not to overload your computer
+registerDoParallel(cl)
+
+data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+
+  set.seed((seed-1)*N_ID+i)
+
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,
+                land=border,verbose=FALSE,save_omega = TRUE)
+
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore
+
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
 }
-  a*theta*(theta-theta0)*(theta+theta0)*exp(-Dshore/D0)/Dshore
-}
-
-data=grid
-data$z=as.vector(z)
-
-#fix magnitude and slope for omega on the AngleNormal axis
-a_fixed <- 1  
-theta0_fixed <- pi/2
-
-fit_optim <- optim(par = c(log_D0 = log(1)),  # Exclude a from parameters
-                   fn = function(params) {
-                     log_D0 <- params[1]
-
-                     # Use fixed values
-                     a <- a_fixed
-                     theta0 <- theta0_fixed
-
-                      # Compute predicted z values
-                     z_pred <- fomega(data, a,theta0,exp(log_D0))
-
-                     # Return sum of squared residuals
-                     sum((data$z - z_pred)^2)
-                   })
-
-est_par=fit_optim$par
-
-# Compute the predicted values using the starting parameters
-z_pred_starting <- fomega(grid,D0=exp(est_par))
-
-# Reshape z_pred_starting to a matrix for plotting
-z_pred_matrix <- matrix(z_pred_starting, nrow = length(Ishore), ncol = length(theta))
-
-# Create the 3D surface plot using plot_ly
-plot_ly(x = ~theta, y = ~1/Ishore, z = ~z_pred_matrix) %>%
-  add_surface() %>%
-  layout(title = "Parametric surface for omega (starting parameters)",
-         scene = list(
-           yaxis = list(title = "1/Ishore"),
-           xaxis = list(title = "AngleNormal"),
-           zaxis = list(title = "Predicted z")
-         ))
-
-# Create the 3D surface plot using plot_ly
-plot_ly(x = ~theta, y = ~1/Ishore, z = ~z) %>%
-  add_surface() %>%
-  layout(title = "Predicted spline surface for omega",
-         scene = list(
-           yaxis = list(title = "Ishore"),
-           xaxis = list(title = "AngleNormal"),
-         ))
-
+
+#stop cluster
+stopCluster(cl)
 ```
@@
 
-## Try a model similar to baseline racvm2
+## Try a model directly on the whole data
 
@@
 sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),dim=c(2,2,length(all_data$time)))
 
 #model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),
-                 nu=~s(ID,bs="re"),
+formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
+                 nu=~ExpShip+s(ID,bs="re"),
                  omega=~te(AngleNormal,Ishore,k=5,bs="cs"))
 
-response0<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+response0<- SDE$new(formulas = formulas,data = all_data,type = "RACVM_SSM",response = c("x","y"),
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
 response0$update_map(list("log_lambda"=factor(c(1,2,NA,NA))))
-response0$update_lambda(exp(c(4,4,-5,-5)))
+response0$update_lambda(exp(c(3,3,-15,-15)))
 
@@
 ```{r}
+#plot parameters
+xmin=list("AngleNormal"=-pi,"ExpShip"=1/45,"Ishore"=1/3)
+xmax=list("AngleNormal"=pi,"ExpShip"=1/3,"Ishore"=1/0.1)
+link=list("ExpShip"=(\(x) 1/x),"Ishore"=(\(x) 1/x))
+xlabel=list("ExpShip"="Distance to ship","Ishore"="Distance to shore")
 response0$get_all_plots(baseline=NULL,

### 2025-01-19 R/simulation_study_CRCVM/estimation_local/simulation_study_local.Rmd
update local simulation study

--- 2025-01-08 R/simulation_study_CRCVM/spline_estimation_local/simulation_study_spline.Rmd
@@
 library(mgcv)
+library(plotly)
 
@@
 N_ID=6
-TMAX=1*24
-SP_DF=c(7,4)
-TAU_0=1.5
-NU_0=4
+
+TMAX=12
+DELTA=1/60/6
+
+SP_DF=c(5,5)
+
+TAU_0=1.3
+NU_0=4.6
+
 DMIN=0.5
-DELTA=1/360
+DMAX=1
+
 PAR0_RACVM=c(0,0,1,1,0)
-SIGMA_OBS_low=0.01
-SIGMA_OBS_high=0.04
-BY=30
+
+SIGMA_OBS_LOW=0.01
+SIGMA_OBS_HIGH=0.04
+
+BY_LF=5
+BY_HF=1
+
+D_LOW=0.05
+D_UP=3
+
+A=1
+D0=0.3
+D1=1.2
+SIGMA_THETA=pi/6
+SIGMA_D=1
+B=1
 
@@
 
-seed=100
+seed=42
 set.seed(seed)
@@
 
+shrunk_border<-st_read(file.path(greenland_data_path,"shrunk_updated_scoresby_sound_utm.shp"))
+shrunk_border <- st_transform(border, crs = "+init=EPSG:32626 +units=km")
 v0=c(0,0)
@@
     #keep it as initial position if it is at least 50 metres away from the shore
-    if (Dshore>DMIN && Dshore<1) {
+    if (Dshore>DMIN && Dshore<DMAX) {
       x0[i,]=x
@@
 
-fomega_cubic=function(cov_data,a=2,D0=0.5,D1=0.8,sigma_theta=pi/8,sigma_D=0.3,b=5){
+fomega_cubic=function(cov_data,a,b,D0,D1,sigma_D,sigma_theta){
   Dshore=cov_data$DistanceShore
@@
            exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
+}
+
+fomega=function(cov_data) {
+  fomega_cubic(cov_data,A,B,D0,D1,SIGMA_D,SIGMA_THETA)
 }
@@
 
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic,fnu=fnu_constant,
-                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,
+                land=shrunk_border,verbose=FALSE,save_omega = TRUE)
 
@@
   data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$shore
+
+  data_sim=cbind(data_sim,data_shore)
   data_sim
@@
 # add noise
-low_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_low^2,2)))
-high_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_high^2,2)))
-observed_data_le=data_sub
+low_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_LOW^2,2)))
+high_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_HIGH^2,2)))
+observed_data_le=data_sub[,c("y1","y2","time","ID")]
 observed_data_le[,c("y1","y2")]=data_sub[,c("y1","y2")]+low_noise
-observed_data_he=data_sub
+observed_data_he=data_sub[,c("y1","y2","time","ID")]
 observed_data_he[,c("y1","y2")]=data_sub[,c("y1","y2")]+high_noise
@@
 #subsample
-data_lf_he=observed_data_he[seq(1,length(data_sub$time),by=BY),]
-data_lf_le=observed_data_le[seq(1,length(data_sub$time),by=BY),]
-data_hf_le=observed_data_le[seq(1,length(data_sub$time),by=6),]
-data_hf_he=observed_data_he[seq(1,length(data_sub$time),by=6),]
+data_lf_he=observed_data_he[seq(1,length(data_sub$time),by=BY_LF),]
+data_lf_le=observed_data_le[seq(1,length(data_sub$time),by=BY_LF),]
+data_hf_le=observed_data_le[seq(1,length(data_sub$time),by=BY_HF),]
+data_hf_he=observed_data_he[seq(1,length(data_sub$time),by=BY_HF),]
 
@@
   coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(data=data_sub,mapping=aes(y1,y2,col=ID),size=0.1,alpha=0.2,shape=2)+
   geom_point(data=data_lf_he,mapping=aes(y1,y2,col=ID),size=0.1)+
@@
 
-ggsave(filename=paste("plot_","border","_",TMAX,"h_",N_ID,"ID_",DMIN,"km",seed,".png",sep=""),plot=plot,width=10,height=5)
+ggsave(filename="plot_trajectories.pdf",plot=plot,width=10,height=5)
 
@@
 
-add_covs=function(data) {
-  #data has at least columns "time","y1" "y2", "p1","p2"
-  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"
+  # Function to calculate rolling average
+rolling_average <- function(data, window) {
+  n <- nrow(data)
+  result <- matrix(NA, n, ncol(data))  # Initialize result matrix with NA
+  colnames(result) <- colnames(data)
 
-  new_data=data
-  #observed deviation angles
-  n=length(new_data$time)
+  for (i in seq_len(n)) {
+    # Define the range for the current window
+    start <- max(1, i - floor(window / 2))
+    end <- min(n, i + floor(window / 2))
+    result[i, ] <- colMeans(data[start:end, , drop = FALSE])  # Compute column means
+  }
+  return(result)
+}
 
-  new_data$theta=rep(NA,n)
+add_covs_parallel <- function(data,add="all",window=1,n_cores = parallel::detectCores() - 1) {
+  # Setup parallel backend
+  cl <- makeCluster(n_cores)
+  registerDoParallel(cl)
 
-  for (id in unique(new_data$ID)) {
+  # Split data by ID
+  ids <- unique(data$ID)
 
-    #filter data with specific id
-    sub_ind=(new_data$ID==id)
-    sub_data=new_data[sub_ind,]
-    n_sub=length(sub_data$time)
+  # Process each ID in parallel
+  results <- foreach(id = ids, .combine = rbind, .packages = c("sf"),
+                     .export = c("nearest_boundary_points", "is_in_land", "signed_angle","border",
+                                 "D_LOW","D_UP","rolling_average")  ) %dopar% {
+    # Filter data for the current ID
+    sub_data <- data[data$ID == id, ]
+    n_sub <- nrow(sub_data)
 
-    #time steps
-    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]
+    # Time steps
+    dtimes <- sub_data[2:n_sub, "time"] - sub_data[1:(n_sub - 1), "time"]
 
-    #step lengths
-    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
-    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]
+    # Step lengths
+    dx <- sub_data[2:n_sub, "y1"] - sub_data[1:(n_sub - 1), "y1"]
+    dy <- sub_data[2:n_sub, "y2"] - sub_data[1:(n_sub - 1), "y2"]
 
-    #matrix of empirical velocity
-    vexp_df=cbind(dx/dtimes,dy/dtimes)
+    # Empirical velocity
+    vexp_df <- cbind(dx / dtimes, dy / dtimes)
 
-    #nearest points on shore
-    sub_data=cbind(sub_data,nearest_boundary_points(as.matrix(sub_data[,c("y1","y2")]),border))
+    vexp_avg <- rolling_average(vexp_df, window)
 
-    #matrix of normal vectors
-    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])
+    # Nearest points on shore
+    sub_data <- cbind(sub_data, nearest_boundary_points(as.matrix(sub_data[, c("y1", "y2")]), border))
 
-    #angle between velocity and normal vector
-    theta_coast=signed_angle(normal,vexp_df)
+    # Normal vectors
+    normal <- as.matrix(sub_data[2:n_sub, c("y1", "y2")] - sub_data[2:n_sub, c("p1", "p2")])
 
-    #adjust lengths 
-    theta_coast=c(theta_coast,1)
+    # Angle between velocity and normal vector
+    theta_coast <- signed_angle(normal, vexp_avg)
+    theta_coast <- c(theta_coast, 1)  # Adjust length
 
-    new_data[sub_ind,"theta"]=theta_coast
+    # Initialize DistanceShore
+    Dshore <- rep(0, n_sub)
 
-    Dshore=rep(0,n_sub)
     for (i in 1:n_sub) {
-      y1=sub_data[i,"y1"]
-      y2=sub_data[i,"y2"]
-      if (!(is_in_land(st_point(c(y1,y2)),border))) {
-        p1=sub_data[i,"p1"]
-        p2=sub_data[i,"p2"]
-        Dshore[i]=sqrt((y1-p1)^2+(y2-p2)^2)
+      y1 <- sub_data[i, "y1"]
+      y2 <- sub_data[i, "y2"]
+      if (!is_in_land(st_point(c(y1, y2)), border)) {
+        p1 <- sub_data[i, "p1"]
+        p2 <- sub_data[i, "p2"]
+        Dshore[i] <- sqrt((y1 - p1)^2 + (y2 - p2)^2)
       }
     }
-    new_data[sub_ind,"DistanceShore"]=Dshore
 
+    Dshore_avg=as.vector(rolling_average(matrix(Dshore,ncol=1),window))
+
+    # Calculate Ishore
+    Ishore_avg <- ifelse(Dshore_avg < D_LOW, 1 / D_LOW, ifelse(Dshore_avg > D_UP, 0, 1 / Dshore_avg))
+
+    # Add columns to sub_data
+    if (add=="all") {
+      sub_data$theta <- theta_coast
+      sub_data$DistanceShore <- Dshore_avg
+      sub_data$Ishore <- Ishore_avg
+    }
+    else if (add=="theta") {
+      sub_data$theta <- theta_coast
+    }
+    else if (add=="DistanceShore") {
+      sub_data$DistanceShore <- Dshore
+      sub_data$Ishore <- Ishore
+        }
+  else {
+    stop("Options for argument add are \"all\",\"DistanceShore\" and \"theta\"")
   }
 
-  return(new_data)
+    return(sub_data)
+  }
+
+  # Stop the cluster
+  stopCluster(cl)
+
+  return(results)
 }
@@
 ```{r}
-data_lf_he=add_covs(data_lf_he)
-data_hf_le=add_covs(data_hf_le)
-data_lf_le=add_covs(data_lf_le)
-data_hf_he=add_covs(data_hf_he)
+#colnames(data)[8]="DistanceShore"
+data_lf_he=add_covs_parallel(data_lf_he,window=3)
+data_hf_le=add_covs_parallel(data_hf_le,window=3)
+data_lf_le=add_covs_parallel(data_lf_le,window=3)
+data_hf_he=add_covs_parallel(data_hf_he,window=3)
 ```
@@
 ggplot(data_lf_he)+geom_histogram(aes(x=DistanceShore),bins=30)+facet_wrap(~ID)+theme_minimal()
-quantile(data_lf$DistanceShore)
+quantile(data_lf_he$DistanceShore)
 ```
@@
 
-ctcrw_lf_he<- SDE$new(formulas = formulas,data =data_lf_he[data_lf_he$ID %in% c(1,2,3),],type = "CTCRW",
+ctcrw_lf_he<- SDE$new(formulas = formulas,data =data_lf_he[data_lf_he$ID %in% c(1,2,4,5),],type = "CTCRW",
                     response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
@@
 
-formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)
+formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1,omega=~1)
 
-ctcrw_lf_le<- SDE$new(formulas = formulas,data =data_lf_le[data_lf_le$ID %in% c(1,2,3),],type = "CTCRW",
-                    response = c("y1","y2"),par0 = c(0,0,1,1),fixpar=c("mu1","mu2"),
+ctcrw_lf_le<- SDE$new(formulas = formulas,data =data_lf_le[data_lf_le$ID %in% c(1:6),],type = "RACVM_SSM",
+                    response = c("y1","y2"),par0 = c(0,0,1,1,0),fixpar=c("mu1","mu2"),
                     other_data=list("log_sigma_obs0"=log(0.03)))
@@
 #points on the surface perturbed by gaussian noise
-f <- fomega_cubic(samples)
+f <- fomega(samples)
 y <- f+0.1*rnorm(n)
@@
 
-H=array(rep(SIGMA_OBS_high^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
+formulas <- list(mu1=~1,mu2=~1,tau=~1,
+                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs=c("cs")))
 
-formulas <- list(mu1=~1,mu2=~1,tau=~s(theta,k=8,bs="cc"),
-                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs="cs"))
-
-crcvm_splines_lf_he<- SDE$new(formulas = formulas,data = cropped_data,type = "RACVM_SSM",
+crcvm_splines_lf_he<- SDE$new(formulas = formulas,data = data_lf_he,type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
@@
 
-crcvm_splines_lf_he$update_map(list("log_lambda"=factor(c(1,rep(NA,2)))))
-crcvm_splines_lf_he$update_lambda(c(1,10^-7,10^-13))
+crcvm_splines_lf_he$update_map(list("log_lambda"=factor(c(rep(NA,2)))))
+crcvm_splines_lf_he$update_lambda(m1$sp)
 #fit
@@
 
-crcvm_splines_lf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))
+crcvm_splines_lf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
 
@@
 We are well estimating $\nu$ and $\sigma_{obs}$, and we corrected the bias in $\tau$ compared to the estimate we got for the CTCRW on the same data.
+We can try to fit the true parametric function on the estimated spline surface.
+
+```{r}
+
+estimate_parametric_omega=function(model) {
+
+    data=model$data()
+
+    Dshore=seq(from=quantile(data$DistanceShore,0.5),to=quantile(data$DistanceShore,0.95),length.out=30)
+    theta=seq(from=-pi,to=pi,length.out=30)
+    grid<- as.data.frame(expand.grid(Dshore,theta))
+    colnames(grid) <- c("DistanceShore","theta")
+    grid$ID=1
+
+    #get model matrices
+    mats=model$make_mat(new_data=grid)
+    X_fe=mats$X_fe
+    X_re=mats$X_re
+
+    par_mat=model$par(new_data=grid,X_fe=X_fe,X_re=X_re)
+
+    #matrix of values for surface plot
+    z=matrix(par_mat[,"omega"],30,30)
+
+    data=grid
+    data$z=as.vector(z)
+
+    # Create the 3D surface plot using plot_ly
+    plot_splines<- plot_ly(x = ~theta, y = ~Dshore, z = ~z) %>%
+      add_surface() %>%
+      layout(title = "Predicted spline surface for omega",
+             scene = list(yaxis = list(title = "Distance to shore"),
+               xaxis = list(title = "theta")))
+
+    fit_optim <- optim(par = c(0.1,2),
+                       fn = function(params) {
+
+                        # Compute predicted z values
+                        z_pred <- fomega_cubic(data,A,B,params[1],params[2],SIGMA_D,SIGMA_THETA)
+
+                         # Return sum of squared residuals
+                         sum((data$z - z_pred)^2)
+                       })
+
+    est_par=fit_optim$par
+
+    # Compute the predicted values using the starting parameters
+    z_pred_starting <- fomega_cubic(grid,A,B,est_par[1],est_par[1],SIGMA_D,
+                                    SIGMA_THETA)
+
+    # Reshape z_pred_starting to a matrix for plotting
+    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta))
+
+    # Create the 3D surface plot using plot_ly
+    plot_parametric <- plot_ly(x = ~theta, y = ~Dshore, z = ~z_pred_matrix) %>%
+      add_surface() %>%
+      layout(title = "Estimated parametric surface for omega",
+             scene = list(
+               yaxis = list(title = "Distance to shore"),
+               xaxis = list(title = "theta"),
+               zaxis = list(title = "Predicted z")
+             ))
+
+    return (list("est_par"=est_par,"splines_surface"=plot_splines,"parametric_surface"=plot_parametric))
+
+}
+
+estimate_parametric_omega(crcvm_splines_lf_he)
+
+```
 
@@
 
-formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=c(10,5),bs="cs"))
-
-crcvm_splines_lf_le<- SDE$new(formulas = formulas,data = data_lf_le,type = "RACVM_SSM",
+crcvm_splines_lf_le<- SDE$new(formulas = formulas,data = data_lf_le[data_lf_le$ID %in% c(1:6),],type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
-                    other_data=list("log_sigma_obs0"=log(0.03)))
+                    other_data=list("log_sigma_obs0"=log(0.05)))
 
 crcvm_splines_lf_le$update_map(list("log_lambda"=factor(rep(NA,2))))
-crcvm_splines_lf_le$update_lambda(c(10^-12,10^-12))
+crcvm_splines_lf_le$update_lambda(m1$sp)
 #fit
@@
 
-crcvm_splines_lf_le$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))
+crcvm_splines_lf_le$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
 
-exp(as.list(ctcrw_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
+exp(as.list(crcvm_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
 
+```
+
+```{r}
+estimate_parametric_omega(crcvm_splines_lf_le)
 ```
@@
 
-cropped_data=data_hf_he[data_hf_he$ID %in% 2:2,]
-
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
@@
 
-crcvm_splines_hf_he<- SDE$new(formulas = formulas,data = cropped_data,type = "RACVM_SSM",
+crcvm_splines_hf_he<- SDE$new(formulas = formulas,data = data_hf_he[data_hf_he$ID %in% 1:2,],type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
@@
 #initialize smoothing penalties and re variances
-new_lambda=c(10^-6,10^-4)
+new_lambda=c(m1$sp)
 crcvm_splines_hf_he$update_lambda(new_lambda)
@@
 
-plot_results_hf=crcvm_splines_hf$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))
-
-plot_results_hf
+crcvm_splines_hf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
 
@@
 
-For high frequency data, we are able to get $\tau$ and $\nu$ right, but we are still underestimating $\omega$.
-Now we force the parametrization of the surface for $\omega$ with a new SDE model in smoothSDE (called \texttt{CRCVM_SSM}.
-$\omega$ is a cubic function of the angle $\theta$, with an exponential factor.
+```{r}
+
+estimate_parametric_omega(crcvm_splines_hf_he)
+
+```
 
@@
 
-cropped_data=data_hf_he[data_hf_le$ID %in% 1,]
-
-H=array(rep(SIGMA_OBS^2*diag(2),length(data_hf$time)),dim=c(2,2,length(data_hf$time)))
-
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
@@
 
-crcvm_splines_hf<- SDE$new(formulas = formulas,data = data_hf,type = "RACVM_SSM",
+crcvm_splines_hf_le<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% 1:1,],type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
-                    other_data=list("H"=H))
+                    other_data=list("log_sigma_obs0"=log(0.05)))
 
 #initialize smoothing penalties and re variances
-new_lambda=c(exp(-1),exp(-1))
-crcvm_splines_hf$update_lambda(new_lambda)
+new_lambda=c(m1$sp)
+crcvm_splines_hf_le$update_lambda(new_lambda)
 
-crcvm_splines_hf$update_map(list(log_lambda=factor(c(NA,NA))))
+crcvm_splines_hf_le$update_map(list(log_lambda=factor(c(NA,NA))))
 
 #fit
-crcvm_splines_hf$fit(method="BFGS")
+crcvm_splines_hf_le$fit(method="BFGS")
+```
+
+```{r}
+crcvm_splines_hf_le$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
+
 ```
@@
 
-plot_results_hf=crcvm_splines_hf$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega_cubic))
-
-plot_results_hf
-
+estimate_parametric_omega(crcvm_splines_hf_le)
 ```
 
-## Estimation with fixed cubic function of angle
+## Estimation with fixed parametric omega function
 
@@
 ```{r}
-H=array(rep(0.03^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
+H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
 # Estimate from simulated data ------------------
@@
 
-crcvm_cubic_lf_he<- SDE$new(formulas = formulas,data = data_lf_he[data_lf_he$DistanceShore>0.01,],type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1,1,1,5,0.5,0.8,0.3,pi/8),
+crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
+                            data = data_lf_he,type = "CRCVM_SSM",
+                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                     fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("log_sigma_obs0"=log(0.02)))
+                    other_data=list("H"=H))
+
 #fit
@@
 ```{r}
-H=array(rep(0.007^2*diag(2),length(data_lf_le$time)),dim=c(2,2,length(data_lf_le$time)))
+H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_lf_le$time)),dim=c(2,2,length(data_lf_le$time)))
 # Estimate from simulated data ------------------
@@
 
-crcvm_cubic_lf_le<- SDE$new(formulas = formulas,data = data_lf_le,type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1.5,4,1,5,1,0.8,0.2,pi/8),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("log_sigma_obs0"=log(0.03)))
-#fit
+crcvm_cubic_lf_le<- SDE$new(formulas = formulas,data = data_lf_le[data_lf_le$ID %in% c(1:4),],
+                              type = "CRCVM_SSM",
+                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
+                    other_data=list("H"=H))
+
 crcvm_cubic_lf_le$fit(method="BFGS")
@@
 ```{r}
+H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_hf_he$time)),dim=c(2,2,length(data_hf_he$time)))
 
@@
 
-crcvm_cubic_hf_he<- SDE$new(formulas = formulas,data = data_hf_he[data,type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 =c(1,1,1,5,0.5,0.8,0.3,pi/8),
+crcvm_cubic_hf_he<- SDE$new(formulas = formulas,data = data_hf_he[data_hf_he$DistanceShore>0.01 & data_hf_he$ID %in% c(1:6),],type = "CRCVM_SSM",
+                    response = c("y1","y2"),par0 =c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                     fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("log_sigma_obs0"=log(0.05)))
+                    other_data=list("H"=H))
 
-#fit
 crcvm_cubic_hf_he$fit(method="BFGS")
@@
 ```
+
 ### High frequency low error
@@
 
+H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_hf_le$time)),dim=c(2,2,length(data_hf_le$time)))
+
 formulas <- list(tau=~1,
@@
 
-crcvm_cubic_hf_le<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% 4:5,],type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1,1,40,5,0.15,0.8,0.2,pi/8),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("log_sigma_obs0"=log(0.03)))
+crcvm_cubic_hf_le<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% c(1:4),],
+                            type = "CRCVM_SSM",
+                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
+                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
+                    other_data=list("H"=H))
 
@@
 ```{r}
-# Definition of parameters tau and nu ----------------
-tau_re=rnorm(N_ID,mean=0,sd=SIGMA_TAU)
-nu_re=rnorm(N_ID,mean=0,sd=SIGMA_NU)
-true_log_nu=nu_re+log(NU_0)
-
-ftau_gaussian=function(cov_data,sigma_bump=0.5,sigma_dip=0.5,tau0=TAU_0,dtau_bump=1.5,dtau_dip=0) {
-  #cov_data : dataframe of covariates with column for angle theta
-  #sigma_bump : width of the bumps at -pi/2 and pi/2
-  #sigma_dip : width of the dip at 0
-  #tau0 : base value of tau
-  #dtau_bump : height of the bump
-  #dtau_dip : depth of the dip
-
-  theta=cov_data$theta
-  bump1 <- dtau_bump*exp(-((theta - pi/2)^2) / (2 * sigma_bump^2))
-  bump2 <- dtau_bump*exp(-((theta + pi/2)^2) / (2 * sigma_bump^2))
-  dip<- dtau_dip*exp(-((theta)^2) / (2 * sigma_dip^2))
-  tau=tau0+bump1+bump2-dip
-
-  return (tau)
-}
+ggplot()+geom_point(data=data_hf_le,aes(time,theta),col="blue",alpha=0.5,size=0.1)+
+  geom_point(data=data,aes(time,theta),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)
 
 ```
+```{r}
+data_hf_le$omega=fomega(data_hf_le[,c("theta","DistanceShore")])
+
+ggplot()+geom_point(data=data_hf_le,aes(time,omega),col="blue",alpha=0.5,size=0.1)+
+  geom_point(data=data,aes(time,omega),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)
+```

### 2025-01-29 R/application/RCVM/fit_rcvm.Rmd
new results with shrunk land polygons

@@
 library(gganimate) # animated trajectories
+library(doParallel)
+library(foreach)
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R"))) 
 ```
@@
 
+dataBE24[dataBE24$DistanceShore_translated> D_up,"Ishore_translated"]=0
+dataBE24[dataBE24$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low
+
 dataBE12[dataBE12$DistanceShore> D_up,"Ishore"]=0
@@
 
+dataBE12[dataBE12$DistanceShore_translated> D_up,"Ishore_translated"]=0
+dataBE12[dataBE12$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low
+
 dataAE[dataAE$DistanceShore> D_up,"Ishore"]=0
 dataAE[dataAE$DistanceShore< D_low,"Ishore"]=1/D_low
+
+dataAE[dataAE$DistanceShore_translated> D_up,"Ishore_translated"]=0
+dataAE[dataAE$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low
 
@@
 #Fixed measurement error
-sigma_obs=0.03
+sigma_obs=0.05
 H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
@@
                           par0 = par0,
-                          other_data=list("log_sigma_obs0"=log(0.05)),
+                          other_data=list("H"=H),
                           fixpar=c("mu1","mu2"))
@@
 
-```{r results baseline_ctcrw0}
-
-estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
-std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_ctcrw0}
-
+```{r define function to get raw estimates}
 # Function to extract estimate values into a data frame format
@@
 
+get_raw_latex_tab<- function(sde,show=TRUE) {
+
+  estimates=as.list(sde$tmb_rep(),what="Est")
+  std=as.list(sde$tmb_rep(),what="Std")
+
   # Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
-                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe,
-                          std_bas_ctcrw0$coeff_fe, 
-                          rownames(estimates_bas_ctcrw0$coeff_fe))
-
-coeff_re <- combine_to_df(estimates_bas_ctcrw0$coeff_re,
-                          std_bas_ctcrw0$coeff_re, 
-                          rownames(estimates_bas_ctcrw0$coeff_re))
-
-log_lambda <- combine_to_df(estimates_bas_ctcrw0$log_lambda,
-                          std_bas_ctcrw0$log_lambda, 
-                          rownames(estimates_bas_ctcrw0$log_lambda))
+  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
+                                 std$log_sigma_obs, "log_sigma_obs")
+  coeff_fe <- combine_to_df(estimates$coeff_fe,
+                            std$coeff_fe, 
+                            rownames(estimates$coeff_fe))
+
+  coeff_re <- combine_to_df(estimates$coeff_re,
+                            std$coeff_re, 
+                            rownames(estimates$coeff_re))
+
+  log_lambda <- combine_to_df(estimates$log_lambda,
+                            std$log_lambda, 
+                            rownames(estimates$log_lambda))
 
@@
 
-```
-
-```{r, results='asis',message=FALSE}
   latex_table <- xtable(table_data, 
@@
 
-# Print LaTeX table
+  if (show) {
     print(latex_table, type = "latex", include.rownames = FALSE)
-```
+  }
+
+  return (latex_table)
+
+}
+
+```
+
+```{r estimates baseline_ctcrw0}
+raw_tab_baseline_ctcrw0=get_raw_latex_tab(baseline_ctcrw0)
+```
+
 We can get confidence intervals for the intercept of each parameter and the random effects standard deviations.
 
-```{r CI baseline_ctcrw0 parameters}
-
-post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
-post_par_bas_ctcrw0=list(
-  "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
-  "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]),
-  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"tau.s(ID)"])),
-  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"nu.s(ID)"])))
-
-mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
-sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
-quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
+```{r define function to get final results}
+
+get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=TRUE) {
+
+  post_coeff=sde$post_coeff(n_post=10000)
+  post_par=list(
+  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]),
+  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
+  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))
+
+  if (include_sigma_obs) {
+    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
+  }
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
                       quantile, probs = c(0.025, 0.975))
 
-```
-
-```{r CI table baseline_ctcrw0 parameters, message=FALSE}
-
-parameter_names <- names(mean_par_bas_ctcrw0)
-estimates <- mean_par_bas_ctcrw0
+  parameter_names <- names(mean_par)
+  estimates <- mean_par
   conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
-          quant_par_bas_ctcrw0[[param]][2])
+  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
+          quant_par[[param]][2])
     })
 
+  parameter_names=c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$")
+  if (include_sigma_obs) {
+    parameter_names=c(parameter_names,"$\\sigma_{\\obs}$")
+  }
   table_data <- data.frame(
-  Parameter = 
-    c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
+    Parameter = parameter_names,
+  Estimate = sprintf("$%.2f$", mean_par),
   CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-
-```{r,results='asis',echo=FALSE}
+  stringsAsFactors = FALSE)
+
   xtab <- xtable(
   table_data,
-  caption = "CTCRW baseline estimations with random effects",
-  label = "table:ctcrw_baseline_estimations_with_random_effects"
-)
-
+  caption = paste(sde$type(),"baseline estimations with random effects"),
+  label = paste0("table:",sde$type(),"_baseline_estimations_with_random_effects"))
+
+  if (show) {
   print(
@@
   hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
+  table.placement="H")
+  }
+
+  return(xtab)
+
+}
+
+```
+
+```{r final estimates baseline_ctcrw0, message=FALSE}
+
+final_tab_baseline_ctcrw0=get_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)
+
 ```
@@
 We now decide to introduce a rotational parameter $\omega$ in the SDE to allow the velocity to rotate depending on the narwhals position with respect to the shoreline.
-We start by introducing the covariate \texttt{AngleNormal} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.
+We start by introducing the covariate \texttt{theta} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.
 
@@
                  nu=~s(ID,bs="re"),
-                 omega=~s(theta_translated,k=10,bs="cs"))
-
-#Fixed measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
+                 omega=~s(theta_translated,k=5,bs="cs"))
 
@@
 baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 
-```{r results baseline_racvm0}
-
-estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
-std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")
-```
-
-```{r render estimates table baseline_racvm0 }
-
-# Function to extract estimate values into a data frame format
-combine_to_df <- function(estimates, std_dev, param_name) {
-  data.frame(
-    Parameter = param_name,
-    Estimate = unlist(estimates, use.names = FALSE),
-    Std_Error = unlist(std_dev, use.names = FALSE)
-  )
-}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
-                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe, 
-                          std_bas_racvm0$coeff_fe, 
-                          rownames(estimates_bas_racvm0$coeff_fe))
-coeff_re <- combine_to_df(estimates_bas_racvm0$coeff_re, 
-                          std_bas_racvm0$coeff_re, 
-                          rownames(estimates_bas_racvm0$coeff_re))
-log_lambda <- combine_to_df(estimates_bas_racvm0$log_lambda, 
-                          std_bas_racvm0$log_lambda, 
-                          rownames(estimates_bas_racvm0$log_lambda))
-
-# Combine all data frames into one
-table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)
-
-```
-
-```{r, results='asis',message=FALSE}
-latex_table <- xtable(table_data, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE)
-```
-
-```{r CI baseline_racvm0 parameters}
-
-post_coeff_bas_racvm0=baseline_racvm0$post_coeff(n_post=10000)
-post_par_bas_racvm0=list(
-  "tau"=exp(post_coeff_bas_racvm0$coeff_fe[,"tau.(Intercept)"]),
-  "nu"=exp(post_coeff_bas_racvm0$coeff_fe[,"nu.(Intercept)"]),
-  "omega"=post_coeff_bas_racvm0$coeff_fe[,"omega.(Intercept)"],
-  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"tau.s(ID)"])),
-  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"nu.s(ID)"])))
-
-mean_par_bas_racvm0=lapply(post_par_bas_racvm0,mean)
-sd_par_bas_racvm0=lapply(post_par_bas_racvm0,sd)
-quant_par_bas_racvm0 <- lapply(post_par_bas_racvm0, 
-                               quantile, probs = c(0.025, 0.975))
-
-```
-
-```{r CI table baseline_racvm0 parameters, message=FALSE}
-
-parameter_names <- names(mean_par_bas_racvm0)
-estimates <- mean_par_bas_racvm0
-conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm0[[param]][1], 
-          quant_par_bas_racvm0[[param]][2])
-})
-
-table_data <- data.frame(
-  Parameter = 
-    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_racvm0),
-  CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-
-```{r,results='asis',echo=FALSE}
-xtab <- xtable(
-  table_data,
-  caption = "racvm0 baseline estimations",
-  label = "table:racvm0_baseline_estimations_with_random_effects"
-)
-
-print(
-  xtab,type="latex",
-  include.rownames = FALSE,
-  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
-  hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
+```{r estimates baseline_racvm0}
+raw_tab_baseline_racvm0=get_raw_latex_tab(baseline_racvm0)
+```
+
+```{r final estimates baseline_racvm0, message=FALSE}
+
+final_tab_baseline_racvm0=get_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)
+
 ```
@@
 
-We add the effect of the distance to the shore through the covariate \texttt{Ishore}.
+We add the effect of the distance to the shore through the covariate \texttt{Ishore_translated}.
 This model has a lot of parameters and takes around 15 minutes to fit. 
@@
 
-#Fixed measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
-
 #parameters formulas
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))
+                   omega=~te(theta_translated,Ishore_translated,k=5,bs="cs"))
 
@@
 
-```{r get results baseline_racvm1_unstable,message=FALSE}
-
-estimates_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Est")
-std_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Std")
-
-estimates_bas_racvm1_unstable
-
-```
 ```{r std random effects baseline_racvm1_unstable}
@@
 
-We see that the estimate for the standard deviation $\sigma_\tau$ of the random effect for $\tau$ is very high while the estimate for the standard deviation $\sigma_\nu$ of the random effect for $\nu$ is very close to $0$.
-The smoothing penalties estimates are extremely high (the smoothing penalties are the exponential of \texttt{estimates\_bas\_racvm1\_unstable$log\_lambda}).
+The smoothing penalties estimates are extremely low (the smoothing penalties are the inverse of the square root of \texttt{omega.te(theta\_translated,Ishore\_translated)}).
 This suggests that the optimization did not go well and numerical errors occurred.
@@
 
-First, we get good values for the smoothing penalty by approximating the true surface with splines.
+First, we get good values for the smoothing penalty by approximating an expected "reasonable" surface with tensor splines, using the parametric function defined in the paper.
 
@@
 
-fomega=function(cov_data,a=2,b=5,D0=0.5,D1=0.8,sigma_D=0.3,sigma_theta=pi/6){
+fomega=function(cov_data,a=1,b=2,D0=0.3,D1=1,sigma_D=0.3,sigma_theta=pi/6){
   Dshore=cov_data$DistanceShore
@@
 n <- 100000
-SP_DF=5
+SP_DF=c(5,5)
 
@@
 theta <- runif(n,-pi,pi);
-DistanceShore <- runif(n,0.02,7)
+DistanceShore <- runif(n,0.05,3)
 samples=data.frame(theta=theta,DistanceShore=DistanceShore)
@@
 theta_v <- seq(-pi,pi,length=30)
-Dshore_v<- seq(0.02,7,length=30)
+Dshore_v<- seq(0.05,3,length=30)
 pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
@@
 
+Ishore=1/DistanceShore
+
 #fit with bivariate splines te
-m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))
+m1 <- gam(y~te(theta,Ishore,k=SP_DF,bs=c("cs")))
 
@@
 par0 <- c(0,0,1,4,0)
-init_lambda=c(exp(3),exp(3),1,exp(-12),exp(-5))
+init_lambda=c(exp(3),exp(3),1,m1$sp)
 
@@
 
-#Fixed measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
-
 #parameters formulas
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~te(theta_translated,DistanceShore_translated,k=c(5,5),bs="cs"))
+                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
 
@@
 baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
-                    response = c("x","y"),par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                     fixpar=c("mu1","mu2"))
@@
 
-```{r results baseline_racvm1}
-
-estimates_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Est")
-std_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Std")
-```
-
-```{r render estimates table baseline_racvm1 }
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_racvm1$log_sigma_obs, 
-                               std_bas_racvm1$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_racvm1$coeff_fe, 
-                          std_bas_racvm1$coeff_fe, 
-                          rownames(estimates_bas_racvm1$coeff_fe))
-coeff_re <- combine_to_df(estimates_bas_racvm1$coeff_re, 
-                          std_bas_racvm1$coeff_re, 
-                          rownames(estimates_bas_racvm1$coeff_re))
-log_lambda <- combine_to_df(estimates_bas_racvm1$log_lambda, 
-                          std_bas_racvm1$log_lambda, 
-                          rownames(estimates_bas_racvm1$log_lambda))
-
-# Combine all data frames into one
-table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)
-
-```
-
-```{r, results='asis',message=FALSE}
-latex_table <- xtable(table_data, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE)
-```
-
-```{r CI baseline_racvm1 parameters}
-
-post_coeff_bas_racvm1=baseline_racvm1$post_coeff(n_post=10000)
-post_par_bas_racvm1=list(
-  "tau"=exp(post_coeff_bas_racvm1$coeff_fe[,"tau.(Intercept)"]),
-  "nu"=exp(post_coeff_bas_racvm1$coeff_fe[,"nu.(Intercept)"]),
-  "omega"=post_coeff_bas_racvm1$coeff_fe[,"omega.(Intercept)"],
-  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"tau.s(ID)"])),
-  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"nu.s(ID)"])))
-
-mean_par_bas_racvm1=lapply(post_par_bas_racvm1,mean)
-sd_par_bas_racvm1=lapply(post_par_bas_racvm1,sd)
-quant_par_bas_racvm1 <- lapply(post_par_bas_racvm1, 
-                               quantile, probs = c(0.025, 0.975))
-
-```
-
-```{r CI table baseline_racvm1 parameters, message=FALSE}
-
-parameter_names <- names(mean_par_bas_racvm1)
-estimates <- mean_par_bas_racvm1
-conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm1[[param]][1], 
-          quant_par_bas_racvm1[[param]][2])
-})
-
-table_data <- data.frame(
-  Parameter = 
-    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_racvm1),
-  CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-
-```{r,results='asis',echo=FALSE}
-xtab <- xtable(
-  table_data,
-  caption = "racvm1 baseline estimations",
-  label = "table:racvm1_baseline_estimations_with_random_effects"
-)
-
-print(
-  xtab,type="latex",
-  include.rownames = FALSE,
-  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
-  hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
-```
+```{r estimates baseline_racvm1}
+raw_tab_baseline_racvm1=get_raw_latex_tab(baseline_racvm1)
+```
+
+```{r final estimates baseline_racvm1, message=FALSE}
+
+final_tab_baseline_racvm1=get_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)
+
+```
+
 ```{r display plots baseline_racvm1}
@@
 plots=lapply(1:length(Ishore_values),function(i){
-  baseline_racvm1$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
+  baseline_racvm1$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                          par_names=c("omega"),n_post=100)})
@@
 
-xmin=list("DistanceShore_translated"=0.2)
-xmax=list("DistanceShore_translated"=2)
+xmin=list("Ishore_translated"=1/3)
+xmax=list("Ishore_translated"=1/0.3)
+link=list("Ishore_translated"=\(x) 1/x)
+xlabel=list("Ishore_translated"="DistanceShore")
 
 plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
-      xmin=xmin, xmax=xmax,show_CI="pointwise")
+      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
 
@@
 
-Now, we would like to demonstrate how the Distance to shore can influence the smooth parameter $\omega$.
+```{r save plots baseline_racvm1}
+# Create directory
+if (!dir.exists("baseline_racvm1")) {
+    dir.create("baseline_racvm1", recursive = TRUE)
+}
+
+# Use lapply to save each plot
+lapply(seq_along(plots_bas_racvm1), function(i) {
+
+    plot<-plots_bas_racvm1[[i]]
+    plot_name <- names(plots_bas_racvm1)[i]
+    file_path <- file.path("baseline_racvm1", plot_name)
+
+    # Check if the plot is a ggplot or plotly object
+    if (inherits(plot, "ggplot")) {
+      ggsave(file_path, plot, device = "pdf")
+    } else if (inherits(plot, "plotly")) {
+      saveWidget(plot, file_path)
+  }
+    })
+```
+Now, we would like to demonstrate how the Distance to shore influences the smooth parameter $\omega$.
 We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.
@@
 ```{r smooth omega baseline_racvm1}
-plot1=plots_bas_racvm1$fe_omega_AngleNormal_q1_Ishore
-plot2=plots_bas_racvm1$fe_omega_AngleNormal_q2_Ishore
-plot4=plots_bas_racvm1$fe_omega_AngleNormal_q4_Ishore
+
+get_shore_influence_plot<-function(sde) {
+
+  xmin=list("Ishore_translated"=1/2)
+  xmax=list("Ishore_translated"=1/0.2)
+  link=list("Ishore_translated"=\(x) 1/x)
+  xlabel=list("Ishore_translated"="DistanceShore")
+
+  plots=sde$get_all_plots(baseline=NULL,
+      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+  plot1=plots$fe_omega_theta_translated_q1_Ishore_translated
+  plot2=plots$fe_omega_theta_translated_q2_Ishore_translated
+  plot4=plots$fe_omega_theta_translated_q4_Ishore_translated
 
@@
     geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
-  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
-  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
-  theme_minimal() +
+    annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 8) + 
+    annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 8) + 
+    theme_minimal()+
    theme(
-    axis.title.x = element_text(size=14),
-    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
-    axis.text.y = element_text(size=12),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    legend.title = element_text(size=14),
-    legend.text = element_text(size=12))
-
-ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline_racvm1", width=10, height=6, units="in")
-
-combined_plot
+    axis.title.x = element_text(size=22),
+    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=18),
+    axis.text.y = element_text(size=18),
+    legend.title = element_text(size=18),
+    legend.text = element_text(size=18)
+  )
+
+  return(combined_plot)
+}
+
+shore_influence_plot_baseline_racvm1=get_shore_influence_plot(baseline_racvm1)
+
+ggsave("omega_DistanceShore_levels.pdf", plot=shore_influence_plot_baseline_racvm1, path="baseline_racvm1", width=10, height=6, units="in")
 
@@
 
-This model in indeed very complex and involves a large number of parameters given the amount of data points we have for the trajectories before exposure.
-We can try a more parsimonious model to make estimation easier.
-
-## Baseline with AngleNormal and Ishore in omega
-
-We decide to remove the effect of \texttt{AngleNormal} on $\tau$ and only put covariates on $\omega$.
-
-```{r fit baseline_racvm2,cache=TRUE}
-
-#initial parameters
-par0 <- c(0,0,2,4,0)
-init_lambda=c(exp(3),exp(3),exp(-12),exp(-5))
-
-#mapping for fixed smoothing penalties
-new_map=list("log_lambda"=factor(c(1,2,rep(NA,2))))
-
-#Fixed measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
-
-#parameters formulas
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))
-
-# Fit baseline
-baseline_racvm2<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"))
-
-baseline_racvm2$update_lambda(init_lambda)
-baseline_racvm2$update_map(new_map)
-baseline_racvm2$fit()
-
-```
-
-```{r get results baseline_racvm2}
-
-estimates_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Est")
-std_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_racvm2}
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_racvm2$log_sigma_obs, 
-                               std_bas_racvm2$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_racvm2$coeff_fe, 
-                          std_bas_racvm2$coeff_fe, 
-                          rownames(estimates_bas_racvm2$coeff_fe))
-coeff_re <- combine_to_df(estimates_bas_racvm2$coeff_re, 
-                          std_bas_racvm2$coeff_re, 
-                          rownames(estimates_bas_racvm2$coeff_re))
-log_lambda <- combine_to_df(estimates_bas_racvm2$log_lambda, 
-                          std_bas_racvm2$log_lambda, 
-                          rownames(estimates_bas_racvm2$log_lambda))
-
-# Combine all data frames into one
-table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)
-
-```
-
-```{r, results='asis',message=FALSE}
-latex_table <- xtable(table_data, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE)
-```
-```{r CI baseline_racvm2 parameters}
-
-post_coeff_bas_racvm2=baseline_racvm2$post_coeff(n_post=10000)
-post_par_bas_racvm2=list(
-  "tau"=exp(post_coeff_bas_racvm2$coeff_fe[,"tau.(Intercept)"]),
-  "nu"=exp(post_coeff_bas_racvm2$coeff_fe[,"nu.(Intercept)"]),
-  "omega"=post_coeff_bas_racvm2$coeff_fe[,"omega.(Intercept)"],
-  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"tau.s(ID)"])),
-  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"nu.s(ID)"])))
-
-mean_par_bas_racvm2=lapply(post_par_bas_racvm2,mean)
-sd_par_bas_racvm2=lapply(post_par_bas_racvm2,sd)
-quant_par_bas_racvm2 <- lapply(post_par_bas_racvm2, 
-                               quantile, probs = c(0.025, 0.975))
-
-```
-
-```{r CI table baseline_racvm2 parameters, message=FALSE}
-
-parameter_names <- names(mean_par_bas_racvm2)
-estimates <- mean_par_bas_racvm2
-conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm2[[param]][1], 
-          quant_par_bas_racvm2[[param]][2])
-})
-
-table_data <- data.frame(
-  Parameter = 
-    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_racvm2),
-  CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-
-```{r,results='asis',echo=FALSE}
-xtab <- xtable(
-  table_data,
-  caption = "racvm2 baseline estimations",
-  label = "table:racvm2_baseline_estimations_with_random_effects"
-)
-
-print(
-  xtab,type="latex",
-  include.rownames = FALSE,
-  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
-  hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
-```
-```{r display plots baseline_racvm2}
-
-Ishore_values=c(1/0.25,1/2,1/5)
-plots=lapply(1:length(Ishore_values),function(i){
-  baseline_racvm2$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
-                         par_names=c("omega"),n_post=100)})
-plots
-
-```
-
-```{r plot smooth parameters CI baseline_racvm2}
-#plot parameters
-
-xmin=list("DistanceShore_translated"=0.2)
-xmax=list("DistanceShore_translated"=4)
-
-plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
-      xmin=xmin,xmax=xmax,show_CI="pointwise")
-
-plots_bas_racvm2
-
-```
-
 ```{r}
 
-estimate_parametric_omega=function(model) {
+estimate_parametric_omega=function(model,fixed_a=1) {
 
@@
 
-    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),to=quantile(data$DistanceShore_translated,0.95),length.out=30)
+    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),
+               to=quantile(data$DistanceShore_translated,0.95),length.out=30)
+    Ishore=1/Dshore
     theta_translated=seq(from=-pi,to=pi,length.out=30)
-    grid<- as.data.frame(expand.grid(Dshore,theta_translated))
-    colnames(grid) <- c("DistanceShore_translated","theta_translated")
+    grid<- as.data.frame(expand.grid(Ishore,theta_translated))
+    colnames(grid) <- c("Ishore_translated","theta_translated")
     grid$ID="Asgeir"
+    grid$Dshore_translated=1/grid$Ishore_translated
 
@@
 
-    fit_optim <- optim(par = c(0.1,2),
+    data_copy=data
+    colnames(data_copy)=c("Ishore","theta","ID","DistanceShore","z")
+
+    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
                        fn = function(params) {
@@
                         # Compute predicted z values
-                        z_pred <- fomega(data,1,0.5,params[1],params[2],1,pi/6)
+                        z_pred <- fomega(data_copy,fixed_a,exp(params[1]),
+                                         exp(params[2]),exp(params[3]),exp(params[4]),exp(params[5]))
 
                          # Return sum of squared residuals
-                         sum((data$z - z_pred)^2)
+                         sum((data_copy$z - z_pred)^2)
                        })
@@
 
-    print(est_par)
-
-    # Compute the predicted values using the starting parameters
-    z_pred_starting <- fomega(grid,1,1,est_par[1],est_par[2],1,
-                                    pi/6)
+    print(exp(est_par))
+
+    # Compute the predicted values using the estimated parameters
+    z_pred_starting <- fomega(grid,fixed_a,exp(est_par[1]),
+                              exp(est_par[2]),exp(est_par[3]),exp(est_par[4]),exp(est_par[5]))
 
@@
     z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta_translated))
-
-    print(z_pred_matrix)
 
@@
 
-estimate_parametric_omega(baseline_racvm2)
+res_racvm1=estimate_parametric_omega(baseline_racvm1,fixed_a=2)
+
+res_racvm1
+```
+
+This model in indeed very complex and involves a large number of parameters given the amount of data points we have for the trajectories before exposure.
+We can try to fit the exact same model on all the data to check if we get the same kind of results with tighter confidence intervals.
+Here, we fix the measurement error standard deviation because we get a non positive definite Hessian matrix in the optimization otherwise.
+
+```{r fit all_racvm1,cache=TRUE}
+
+#initial parameters
+par0 <- c(0,0,1,4,0)
+init_lambda=c(exp(5),exp(5),1,m1$sp)
+
+#mapping for fixed smoothing penalties
+new_map=list("log_lambda"=factor(c(1,2,3,rep(NA,2))))
+
+#Fixed measurement error
+sigma_obs=0.05
+H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),dim=c(2,2,length(all_data$time)))
+
+#parameters formulas
+formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
+
+# Fit baseline
+all_racvm1<- SDE$new(formulas = formulas,data = all_data,type = "RACVM_SSM",
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+                    fixpar=c("mu1","mu2"))
+
+all_racvm1$update_lambda(init_lambda)
+all_racvm1$update_map(new_map)
+all_racvm1$fit()
+
+```
+
+```{r plot smooth parameters CI baseline_racvm1}
+
+#plot parameters
+
+xmin=list("Ishore_translated"=1/2)
+xmax=list("Ishore_translated"=1/0.2)
+link=list("Ishore_translated"=\(x) 1/x)
+xlabel=list("Ishore_translated"="DistanceShore")
+
+plots_all_racvm1=all_racvm1$get_all_plots(baseline=NULL,
+      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_all_racvm1
+
+```
+
+```{r final estimates all_racvm1, message=FALSE}
+
+final_tab_all_racvm1=get_final_baseline_tab(all_racvm1,include_sigma_obs=FALSE)
+
+```
+
+We now come back to the data before exposure and try a more parsimonious model by removing the dependence of $\tau$ on
+the covariate \texttt{theta\_translated}.
+
+## Baseline with theta and Ishore in omega
+
+```{r fit baseline_racvm2,cache=TRUE}
+
+#initial parameters
+par0 <- c(0,0,1,4,0)
+init_lambda=c(exp(3),exp(3),m1$sp)
+
+#mapping for fixed smoothing penalties
+new_map=list("log_lambda"=factor(c(1,2,rep(NA,2))))
+
+#parameters formulas
+formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs=c("cs")))
+
+# Fit baseline
+baseline_racvm2<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+                    fixpar=c("mu1","mu2"))
+
+baseline_racvm2$update_lambda(init_lambda)
+baseline_racvm2$update_map(new_map)
+baseline_racvm2$fit()
+
+```
+
+```{r final estimates baseline_racvm2, message=FALSE}
+
+final_tab_baseline_racvm2=get_final_baseline_tab(baseline_racvm2,include_sigma_obs=FALSE)
+
+```
+
+```{r display plots baseline_racvm2}
+
+Ishore_values=c(1/0.25,1/2,1/5)
+plots=lapply(1:length(Ishore_values),function(i){
+  baseline_racvm2$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
+                         par_names=c("omega"),n_post=100)})
+plots
+
+```
+
+```{r plot smooth parameters CI baseline_racvm2}
+#plot parameters
+
+xmin=list("Ishore_translated"=1/3)
+xmax=list("Ishore_translated"=1/0.3)
+link=list("Ishore_translated"=\(x) 1/x)
+xlabel=list("Ishore_translated"="Distance to shore")
+
+plots_bas_racvm2=baseline_racvm2$get_all_plots(baseline=NULL,
+      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+
+plots_bas_racvm2
 
@@
 # Use lapply to save each plot
-lapply(seq_along(plots_bas2), function(i) {
-
-    plot<-plots_bas2[[i]]
-    plot_name <- names(plots_bas2)[i]
+lapply(seq_along(plots_bas_racvm2), function(i) {
+
+    plot<-plots_bas_racvm2[[i]]
+    plot_name <- names(plots_bas_racvm2)[i]
     file_path <- file.path("baseline_racvm2", plot_name)
@@
 ```{r smooth omega baseline_racvm2}
-plot1=plots_bas_racvm2$fe_omega_AngleNormal_q1_Ishore
-plot2=plots_bas_racvm2$fe_omega_AngleNormal_q2_Ishore
-plot4=plots_bas_racvm2$fe_omega_AngleNormal_q4_Ishore
-
-omega1 <- ggplot_build(plot1)$data[[1]]
-omega2 <- ggplot_build(plot2)$data[[1]]
-omega4 <- ggplot_build(plot4)$data[[1]]
-
-omega1$Category <- "Far"
-omega2$Category <- "Medium"
-omega4$Category <- "Close"
-
-# Combine the data frames
-combined_omega <- rbind(omega1, omega2, omega4)
-
-ci_omega1 <- ggplot_build(plot1)$data[[2]]
-ci_omega2 <- ggplot_build(plot2)$data[[2]]
-ci_omega4 <- ggplot_build(plot4)$data[[2]]
-
-ci_omega1$Category <- "Far"
-ci_omega2$Category <- "Medium"
-ci_omega4$Category <- "Close"
-
-# Combine the data frames
-combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)
-
-combined_plot <- ggplot() +
-  geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
-  geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
-  labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
-  geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
-  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
-  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
-  theme_minimal() +
-  theme(
-    axis.title.x = element_text(size=14),
-    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
-    axis.text.y = element_text(size=12),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    legend.title = element_text(size=14),
-    legend.text = element_text(size=12))
-
-ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline_racvm2", width=10, height=6, units="in")
-
-combined_plot
-
-```
-
-Again, even though there seem to be a rotation in the movement close to the shore when the angle $\Theta \in [-\pi,-\frac{\pi}{2}] \cup [\frac{\pi}{2},\pi]$, the confidence intervals are too wide to conclude.
-Moreover, the magnitude of the rotation is too low to effectively constrained the movement : it is estimated to be at most around $\pi$ rad/h, while the value we used in the simulation study to obtain constrained trajectories is $\frac{\pi}{2}$ rad/min, that is approximately $30 \pi $ rad/h.
-
-## Simulate trajectories before exposure
-
-```{r simulate baseline trajectories}
+
+shore_influence_plot_baseline_racvm2=get_shore_influence_plot(baseline_racvm2)
+ggsave("omega_DistanceShore_levels.pdf", plot=shore_influence_plot_baseline_racvm2,
+       path="baseline_racvm2", width=10, height=6, units="in")
+
+shore_influence_plot_baseline_racvm2
+
+```
+
+This baseline model seems quite satisfactory.
+We can now try to fit the parametric function defined in the paper on the estimated spline surface.
+We have seen in the simulation study that the splines give good estimates for $\omega$ provided we are not too close to
+the shore. Thus we fit the parametric function on the spline surface cropped close to the shore.
+We fix the value of $a$ to 0.5 to ensure that simulated trajectories will be constrained.
+
+```{r}
+
+res_racvm2=estimate_parametric_omega(baseline_racvm2,fixed_a=2)
+
+res_racvm2
+```
+
+## Simulate trajectories before exposure from fitted models
+
+```{r,cache=TRUE}
+
 #initial narwhals positions (convert to matrix to avoid bugs)
 z0_BE <- as.matrix(dataBE12[!duplicated(dataBE12$ID), c("x", "y")])
-
-sigma_obs=0.03
-
-baseline_ctcrw0_sim=baseline_ctcrw0$simulate(z0 = z0_BE,
-                                            data=baseline_ctcrw0$data(),
-                                            sd_noise=sigma_obs)
-
-baseline_racvm0_sim=baseline_racvm0$simulate(z0 = z0_BE,
-                                            data=baseline_racvm0$data(),
-                                            sd_noise=sigma_obs)
-
-baseline_racvm1_sim=baseline_racvm1$simulate(z0 = z0_BE,
-                                            data=baseline_racvm1$data(),
-                                            sd_noise=sigma_obs)
-
-baseline_racvm2_sim=baseline_racvm2$simulate(z0 = z0_BE,
-                                            data=baseline_racvm2$data(),
-                                            sd_noise=sigma_obs)
-# Add a label to each dataset for plotting
-baseline_ctcrw0_sim <- baseline_ctcrw0_sim %>% 
-  mutate(data_type = "Simulated baseline_ctcrw0")
-baseline_racvm0_sim <- baseline_racvm0_sim %>% 
-  mutate(data_type = "Simulated baseline_racvm0")
-baseline_racvm1_sim <- baseline_racvm1_sim %>% 
-  mutate(data_type = "Simulated baseline_racvm1")
-baseline_racvm2_sim <- baseline_racvm2_sim %>% 
-  mutate(data_type = "Simulated baseline_racvm2")
-dataBE12 <- dataBE12 %>% mutate(data_type = "Observed")
-
-# Combine both datasets
-baseline_trajectories <- bind_rows(
-  baseline_ctcrw0_sim,baseline_racvm0_sim,
-  baseline_racvm1_sim,baseline_racvm2_sim, dataBE12)
-
-# Plot both in a single plot with facets
-ggplot(baseline_trajectories) +
-  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
-  facet_wrap(~ data_type) + 
-  ggtitle("Trajectories before ship exposure") +
-  theme_minimal() +
-  labs(color = "Narwhal ID")
-
-```
-
-As in the preliminary analysis, the simulated trajectories still appear more wiggily or tortuous than the observed trajectories.
-In comparison, look at what type of simulated trajectories we obtain when fixing a higher value for the persistence.
-
-```{r simulate persistent trajectory,cache=TRUE}
-#initial parameters
-par0 <- c(0,0,3,4.6)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1)
-
-#Fixed measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
-
-#Fit baseline with measurement error estimated from the data
-baseline_ctcrw_fixed<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
-                          response = c("x","y"),
-                          par0 = par0,
-                          other_data=list("H"=H),
-                          fixpar=c("mu1","mu2","tau","nu"))
-
-baseline_ctcrw_fixed_sim=baseline_ctcrw_fixed$simulate(z0 = z0_BE,
-                                            data=baseline_ctcrw_fixed$data(),
-                                            sd_noise=sigma_obs)
-
-baseline_ctcrw_fixed_sim <- baseline_ctcrw_fixed_sim %>% 
-  mutate(data_type = "Simulated baseline_ctcrw_fixed")
-
-# Combine both datasets
-baseline_trajectories <- bind_rows(baseline_trajectories,
- baseline_ctcrw_fixed_sim)
-
-# Plot both in a single plot with facets
-ggplot(baseline_trajectories) +
-  geom_point(aes(x = x, y = y, color = ID), size = 0.6) +
-  facet_wrap(~ data_type) + 
-  ggtitle("Trajectories before ship exposure") +
-  theme_minimal() +
-  labs(color = "Narwhal ID")
-
-```
-The simple ctcrw model where we fixed all parameters to some value based on observations seems to produce trajectories that match better the observations than the complicated models that we fitted.
-However, here we simulated with the covariates values that are in the data.
-But the angle and distance to shore should be recomputed as the process evolves.
-We now try to simulate and compute the new value of the covariates \texttt{AngleNormal} and \texttt{Ishore} at each step.
-We use high frequency time step, and not the low frequency times steps in the observed data to have accurate trajectories.
-
-```{r,cache=TRUE}
 
@@
                                  "preprocessed_data","greenland")
+
 land<-st_read(file.path(greenland_data_path,
@@
 land <- st_transform(land, crs = "+init=EPSG:32626 +units=km")
+
+shrunk_land<-st_read(file.path(greenland_data_path,
+                        "shrunk_updated_scoresby_sound_utm.shp"))
+shrunk_land <- st_transform(shrunk_land, crs = "+init=EPSG:32626 +units=km")
+
+# Simulate trajectories from adjusted model
 
@@
   }
-  else if (Dshore<0.02) {
-    return (50)
+  else if (Dshore<0.2) {
+    return (5)
   }
@@
 
-atw=list("AngleNormal"=fangle,"Ishore"=fIshore)
+fDshore=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #distance to shore
+  Dshore=sqrt(normal[1]^2+normal[2]^2)
+  if (Dshore>0.005) {
+      return(Dshore)
+  }
+  else {
+    return(0.005)
+  }
+
+}
+
+atw=list("Ishore"=fIshore,"Ishore_translated"=fIshore,
+         "theta"=fangle,"theta_translated"=fangle,"DistanceShore"=fDshore)
 
 # Define time steps
-delta=1/60
-times=unlist(sapply(unique(dataBE12$ID),function(ID) {seq(min(dataBE12[dataBE12$ID==ID,"time"]),max(dataBE12[dataBE12$ID==ID,"time"]),by=delta)}))
-n=length(times)
-N=length(unique(dataBE12$ID))
-data_reg=data.frame("AngleNormal"=rep(0,n*N),
-                    "Ishore"=rep(1,n*N),time=c(),ID=rep(unique(dataBE12$ID),each=n))
-
-# Simulate trajectories with small time step
-baseline_racvm1_trajectory=baseline_racvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 1)
-baseline_racvm1_trajectory_adjusted=baseline_racvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 40)
-
-baseline_racvm2_trajectory=baseline_racvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 1)
-baseline_racvm2_trajectory_adjusted=baseline_racvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,omega_times = 40)
-
-baseline_racvm1_trajectory <- baseline_racvm1_trajectory %>% 
-  mutate(data_type = "Simulated baseline_racvm1")
-baseline_racvm1_trajectory_adjusted <- baseline_racvm1_trajectory_adjusted %>% 
-  mutate(data_type = "Simulated baseline_racvm1 adjusted")
-baseline_racvm2_trajectory <- baseline_racvm2_trajectory %>% 
-  mutate(data_type = "Simulated baseline_racvm2")
-baseline_racvm2_trajectory_adjusted <- baseline_racvm2_trajectory_adjusted %>% 
-  mutate(data_type = "Simulated baseline_racvm2 adjusted")
-
-baseline_trajectories <- bind_rows(baseline_racvm1_trajectory,
-                                   baseline_racvm1_trajectory_adjusted,
-                                   baseline_racvm2_trajectory,
-                                   baseline_racvm2_trajectory_adjusted,dataBE12)
+delta=1/60/6
+times_list=lapply(unique(dataBE12$ID),function(ID) {
+  min=min(dataBE12[dataBE12$ID==ID,"time"])
+  max=max(dataBE12[dataBE12$ID==ID,"time"])
+  seq(min,max,by=delta)})
+names(times_list)=unique(dataBE12$ID)
+times_df <- data.frame(
+  ID = rep(names(times_list), sapply(times_list, length)),
+  time = unlist(times_list),row.names = NULL
+)
+n=length(times_df$time)
+# Define covariates data (needed in the function simulate)
+data_reg=data.frame("AngleNormal"=rep(0,n),"Ishore"=rep(1,n),"Ishore_translated"=rep(1,n),
+                    "theta"=rep(0,n),"DistanceShore"=rep(1,n),"theta_translated"=rep(0,n))
+
+data_reg=cbind(data_reg,times_df)
 
@@
 ```{r}
+est_par_racvm1=res_racvm1$est_par
+
+formulas=list(tau=~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
+              sigma_D=~1,sigma_theta=~1)
+baseline_crcvm1<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
+               par0=c(baseline_racvm1$par()[1,c("tau","nu")],2,
+                        exp(est_par_racvm1)))
+
+baseline_crcvm1$update_coeff_re(baseline_racvm1$coeff_re()[1:20,])
+baseline_crcvm1$update_lambda(baseline_racvm1$lambda()[1:3,])
+
+baseline_crcvm1_trajectory=baseline_crcvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,verbose=FALSE)
+
+n=length(baseline_crcvm1_trajectory$time)
+
+baseline_crcvm1_trajectory_sub=baseline_crcvm1_trajectory[seq(1,n,by=60),]
+
+plot_baseline_crcvm1_trajectory=ggplot() +
+  geom_sf(data=land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=baseline_crcvm1_trajectory_sub, 
+            aes(x = x, y = y,color=ID),size = 0.5, 
+            lineend = "round",alpha=0.15) +     
+  geom_point(data=baseline_crcvm1_trajectory_sub,
+             aes(x = x, y = y,color=ID),size =0.5 ) +
+  theme_minimal()
+```
+
+```{r,message=FALSE,results=console}
+
+set.seed(1)
+est_par_racvm2=res_racvm2$est_par
+
+formulas=list(tau=~s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
+              sigma_D=~1,sigma_theta=~1)
+baseline_crcvm2<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
+               par0=c(baseline_racvm2$par()[1,c("tau","nu")],2,
+                        exp(est_par_racvm2)))
+
+baseline_crcvm2$update_coeff_re(baseline_racvm2$coeff_re()[1:12,])
+baseline_crcvm2$update_lambda(baseline_racvm2$lambda()[1:2,])
+
+baseline_crcvm2_trajectory=baseline_crcvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=shrunk_land,verbose=FALSE)
+
+n=length(baseline_crcvm2_trajectory$time)
+
+baseline_crcvm2_trajectory_sub=baseline_crcvm2_trajectory[seq(1,n,by=60),]
+
+plot_baseline_crcvm2_trajectory=ggplot() +
+  geom_sf(data=land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=baseline_crcvm2_trajectory_sub, 
+            aes(x = x, y = y,color=ID),size = 0.5, 
+            lineend = "round",alpha=0.5) +     
+  geom_point(data=baseline_crcvm2_trajectory_sub,
+             aes(x = x, y = y,color=ID),size =0.5 ) +
+  theme_minimal()
+
+ggsave("plot_baseline_crcvm2_trajectory.pdf", plot=plot_baseline_crcvm2_trajectory,
+       path="baseline_racvm2", width=10, height=6, units="in")
+
+```
+
+```{r}
+
+baseline_crcvm1_trajectory_sub <- baseline_crcvm1_trajectory_sub %>% 
+  mutate(data_type = "Simulated baseline_crcvm1")
+baseline_crcvm2_trajectory_sub<- baseline_crcvm2_trajectory_sub %>% 
+  mutate(data_type = "Simulated baseline_crcvm2")
+observed_trajectory<- dataBE12%>% 
+  mutate(data_type = "Observed")
+
+baseline_trajectories <- bind_rows(baseline_crcvm1_trajectory_sub,
+                                   baseline_crcvm2_trajectory_sub,observed_trajectory)
 
@@
 
-The adjusted SDE models where the smooth parameters $\omega$ has been multiplied by a factor $40$ produces trajectories that seem to follow the shoreline and look more similar to the actual data.
-However, the trajectories obtained from the non adjusted fitted model does not produce constrained trajectories.
-This suggests that we are indeed underestimating the smooth parameter $\omega$ with the tensor splines.
-This may be due to several factors : the low frequency at which the data is observed, the measurement error, the small amount of data that we have.
+We continue the analysis with the fitted baseline model \texttt{baseline\_racvm2}.
+
+## Check estimation on simulations from fitted model
+
 ```{r}
 
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
+  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
+                                  rows")
+  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
+  ind1 <- which(result > pi)
+  ind2 <- which(result <= -pi)
+  result[ind1] <- result[ind1] - 2*pi
+  result[ind2] <- result[ind2] + 2*pi
+  return(result) 
+}
+
+add_covs_parallel <- function(data, n_cores = parallel::detectCores() - 1) {
+  # Setup parallel backend
+  cl <- makeCluster(n_cores)
   registerDoParallel(cl)
 
-data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N_ID+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
-                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,
-                land=border,verbose=FALSE,save_omega = TRUE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore
-
-  data_sim=cbind(data_sim,data_shore)
-  data_sim
-}
-
-#stop cluster
+  # Split data by ID
+  ids <- unique(data$ID)
+
+  # Process each ID in parallel
+  results <- foreach(id = ids, .combine = rbind, .packages = c("sf"),
+                     .export = c("nearest_boundary_points", "is_in_land", "signed_angle","border",
+                                 "D_low","D_up")  ) %dopar% {
+                                   # Filter data for the current ID
+                                   sub_data <- data[data$ID == id, ]
+                                   n_sub <- nrow(sub_data)
+
+                                   # Time steps
+                                   dtimes <- sub_data[2:n_sub, "time"] - sub_data[1:(n_sub - 1), "time"]
+
+                                   # Step lengths
+                                   dx <- sub_data[2:n_sub, "x"] - sub_data[1:(n_sub - 1), "x"]
+                                   dy <- sub_data[2:n_sub, "y"] - sub_data[1:(n_sub - 1), "y"]
+
+                                   # Empirical velocity
+                                   vexp_df <- cbind(dx / dtimes, dy / dtimes)
+
+                                   # Nearest points on shore
+                                   sub_data <- cbind(sub_data, nearest_boundary_points(as.matrix(sub_data[, c("x", "y")]), border))
+
+                                   # Normal vectors
+                                   normal <- as.matrix(sub_data[2:n_sub, c("x", "y")] - sub_data[2:n_sub, c("p1", "p2")])
+
+                                   # Angle between velocity and normal vector
+                                   theta_coast <- signed_angle(normal, vexp_df)
+                                   theta_coast <- c(theta_coast, 1)  # Adjust length
+
+                                   # Initialize DistanceShore
+                                   Dshore <- rep(0, n_sub)
+
+                                   for (i in 1:n_sub) {
+                                     y1 <- sub_data[i, "x"]
+                                     y2 <- sub_data[i, "y"]
+                                     if (!is_in_land(st_point(c(y1, y2)), border)) {
+                                       p1 <- sub_data[i, "p1"]
+                                       p2 <- sub_data[i, "p2"]
+                                       Dshore[i] <- sqrt((y1 - p1)^2 + (y2 - p2)^2)
+                                     }
+                                   }
+
+                                   # Calculate Ishore
+                                   Ishore <- ifelse(Dshore < D_low, 1 / D_low, ifelse(Dshore > D_up, 0, 1 / Dshore))
+
+                                   # Add columns to sub_data
+                                   sub_data$theta <- theta_coast
+                                   sub_data$DistanceShore <- Dshore
+                                   sub_data$Ishore <- Ishore
+
+                                   return(sub_data)
+                                 }
+
+  # Stop the cluster
   stopCluster(cl)
-```
-
-We continue the analysis with the fitted baseline models \texttt {baseline\_racvm1} and \texttt{baseline\_racvm2}.
+
+  return(results)
+}
+border=shrunk_land
+
+# add noise
+noise=rmvn(length(baseline_crcvm2_trajectory_sub$time),rep(0,2),diag(rep(0.05^2,2)))
+baseline_crcvm2_trajectory_sub_obs=baseline_crcvm2_trajectory[seq(1,n,by=30),c("time","x","y","ID")]
+baseline_crcvm2_trajectory_sub_obs[,c("x","y")]=baseline_crcvm2_trajectory_sub_obs[,c("x","y")]+noise
+baseline_crcvm2_trajectory_sub_obs=add_covs_parallel(baseline_crcvm2_trajectory_sub_obs)
+
+```
+
+```{r}
+
+#initial parameters
+par0 <- c(0,0,2,4,0)
+init_lambda=c(exp(3),exp(3),m1$sp)
+
+#mapping for fixed smoothing penalties
+new_map=list("log_lambda"=factor(c(1,2,rep(NA,2))))
+
+#Fixed measurement error
+sigma_obs=0.05
+H=array(rep(sigma_obs^2*diag(2),length(baseline_crcvm2_trajectory_sub_obs$time)),
+        dim=c(2,2,length(baseline_crcvm2_trajectory_sub_obs$time)))
+
+#parameters formulas
+formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
+                   omega=~te(theta,Ishore,k=c(5,5),bs=c("cs")))
+
+# Fit baseline
+baseline_racvm2_check<- SDE$new(formulas = formulas,data = baseline_crcvm2_trajectory_sub_obs,type = "RACVM_SSM",
+                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
+                    fixpar=c("mu1","mu2"))
+
+baseline_racvm2_check$update_lambda(init_lambda)
+baseline_racvm2_check$update_map(new_map)
+baseline_racvm2_check$fit()
+
+```
+```{r}
+get_final_baseline_tab(baseline_racvm2_check,include_sigma_obs=FALSE)
+```
+
+```{r}
+
+xmin=list("Ishore"=1/3)
+xmax=list("Ishore"=1/0.3)
+link=list("Ishore"=\(x) 1/x)
+xlabel=list("Ishore"="Distance to shore")
+
+baseline_racvm2_check$get_all_plots(baseline=NULL,
+      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
+```
 
@@
                  nu=~ExpShip+s(ID,bs="re"),
-                 omega=~te(AngleNormal,Ishore,k=5,bs="cs"))
+                 omega=~te(theta_translated,Ishore_translated,k=5,bs="cs"))
 
@@
 response0$update_map(list("log_lambda"=factor(c(1,2,NA,NA))))
-response0$update_lambda(exp(c(3,3,-15,-15)))
+response0$update_lambda(exp(c(3,3,m1$sp)))
 
@@
 
-## Response models with splines of ExpShip
-
-### Response models based on baseline_racvm2
+## Response model with splines of ExpShip
 
@@
 
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+sigma_obs=0.05
+H=array(rep(sigma_obs^2*diag(2),length(dataAE$time)),dim=c(2,2,length(dataAE$time)))
 
@@
                  nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(Ishore,k=5,bs="cs")+
-                   ti(AngleNormal,Ishore,k=c(5,5),bs="cs"))
+                 omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
 
@@
                     3,4,rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                    log_lambda=factor(c(5,NA,6,NA,rep(NA,4))))
+                    log_lambda=factor(c(5,NA,6,NA,rep(NA,2))))
 #Fit baseline
-response1_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
+response_racvm_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",response = c("x","y"),
+                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
                baseline_racvm2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline_racvm2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm2$coeff_re()[paste("omega.ti(Ishore).",1:4,sep=""),1],
-               baseline_racvm2$coeff_re()[paste("omega.ti(AngleNormal,Ishore).",1:16,sep=""),1])
+               baseline_racvm2$coeff_re()[paste("omega.te(theta_translated,Ishore_translated).",1:24,sep=""),1])
 new_lambda=c(1,baseline_racvm2$lambda()["tau.s(ID)",1],1,baseline_racvm2$lambda()["nu.s(ID)",1],
-             baseline_racvm2$lambda()[3:6,1])
-
-response1_sp$update_map(new_map)
-response1_sp$update_coeff_re(new_coeff_re)
-response1_sp$update_coeff_fe(baseline_racvm2$coeff_fe()[,1])
-
-response1_sp$update_lambda(new_lambda)
-
-response1_sp$fit()
+             baseline_racvm2$lambda()[3:4,1])
+
+response_racvm_sp$update_map(new_map)
+response_racvm_sp$update_coeff_re(new_coeff_re)
+response_racvm_sp$update_coeff_fe(baseline_racvm2$coeff_fe()[,1])
+
+response_racvm_sp$update_lambda(new_lambda)
+
+response_racvm_sp$fit()
 
@@
 ```{r get results response1}
-
-estimates_res1_sp=as.list(response1_sp$tmb_rep(),what="Est")
-std_res1_sp=as.list(response1_sp$tmb_rep(),what="Std")
-
 #plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShip"="Distance to ship")
-
-plots_res1_sp=response1_sp$get_all_plots(baseline=baseline_racvm2,xmin=xmin,
+xmin=list("Ishore_translated"=1/3,"ExpShip"=1/60)
+xmax=list("Ishore_translated"=1/0.5,"ExpShip"=1/5)
+link=list("Ishore_translated"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
+xlabel=list("Ishore_translated"="Distance to shore","ExpShip"="Distance to ship")
+
+plots_res_racvm_sp=response_racvm_sp$get_all_plots(baseline=baseline_racvm2,xmin=xmin,
                                          xmax=xmax,link=link,xlabel=xlabel,show_CI="simultaneous")
 
-```
-
-```{r display plots response1_sp}
-
-#print plots
-plots_res1_sp
-
-```
-
-```{r estimates on parameter scale response1}
-post_coeff_res1_sp=response1_sp$post_coeff(n_post=1000)
-post_par_res1_sp=list("tau0"=exp(post_coeff_res1_sp$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res1_sp$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res1_sp$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res1_sp$log_lambda[,2])))
-mean_par_res1_sp=lapply(post_par_res1_sp,mean)
-sd_par_res1_sp=lapply(post_par_res1_sp,sd)
-
-```
-
-### Response model with splines of ExpShip based on baseline2
-
-```{r fit response2}
-
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
-
-#mapping for fixed coefficients
-new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",
-        rep(NA,N),"nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),
-        coeff_fe=factor(rep(NA,5)),log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))
-
-#Fit baseline
-response2_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))
-
-new_coeff_re=c(rep(0,2),baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2$lambda()["tau.s(ID)",1],1,baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])
-
-response2_sp$update_map(new_map)
-response2_sp$update_coeff_re(new_coeff_re)
-response2_sp$update_coeff_fe(baseline2$coeff_fe()[,1])
-response2_sp$update_lambda(new_lambda)
-
-response2_sp$fit()
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-
-new_coeff_re=c(rep(0,2),baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_50m$lambda()["tau.s(ID)",1],1,baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-
-response2_sp_50m$update_map(new_map)
-response2_sp_50m$update_coeff_re(new_coeff_re)
-response2_sp_50m$update_coeff_fe(baseline2_50m$coeff_fe()[,1])
-
-response2_sp_50m$update_lambda(new_lambda)
-
-response2_sp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-
-new_coeff_re=c(rep(0,2),baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_75m$lambda()["tau.s(ID)",1],1,baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-
-response2_sp_75m$update_map(new_map)
-response2_sp_75m$update_coeff_re(new_coeff_re)
-response2_sp_75m$update_coeff_fe(baseline2_75m$coeff_fe()[,1])
-
-response2_sp_75m$update_lambda(new_lambda)
-
-response2_sp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
-                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
-
-new_coeff_re=c(rep(0,2),baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline2_100m$lambda()["tau.s(ID)",1],1,baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-
-response2_sp_100m$update_map(new_map)
-response2_sp_100m$update_coeff_re(new_coeff_re)
-response2_sp_100m$update_coeff_fe(baseline2_100m$coeff_fe()[,1])
-
-response2_sp_100m$update_lambda(new_lambda)
-
-response2_sp_100m$fit()
-```
-
-```{r get results response2}
-
-estimates_res2_sp=as.list(response2_sp$tmb_rep(),what="Est")
-std_res2_sp=as.list(response2_sp$tmb_rep(),what="Std")
-
-estimates_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Est")
-std_res2_sp_50m=as.list(response2_sp_50m$tmb_rep(),what="Std")
-
-estimates_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Est")
-std_res2_sp_75m=as.list(response2_sp_75m$tmb_rep(),what="Std")
-
-estimates_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Est")
-std_res2_sp_100m=as.list(response2_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShip"="Distance to ship")
-
-plot_res2_sp=response2_sp$get_all_plots(baseline=baseline2,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res2_sp_50m=response2_sp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res2_sp_75m=response2_sp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res2_sp_100m=response2_sp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-```
-
-```{r display plots response2_sp}
-
-#print plots
-plot_res2_sp
-plot_res2_sp_50m
-plot_res2_sp_75m
-plot_res2_sp_100m
-
-```
-
-```{r}
-
-post_coeff_res2_sp=response2_sp$post_coeff(n_post=1000)
-post_par_res2_sp=list("tau0"=exp(post_coeff_res2_sp$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp$log_lambda[,2])))
-mean_par_res2_sp=lapply(post_par_res2_sp,mean)
-sd_par_res2_sp=lapply(post_par_res2_sp,sd)
-
-post_coeff_res2_sp_50m=response2_sp_50m$post_coeff(n_post=1000)
-post_par_res2_sp_50m=list("tau0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_50m$log_lambda[,2])))
-mean_par_res2_sp_50m=lapply(post_par_res2_sp_50m,mean)
-sd_par_res2_sp_50m=lapply(post_par_res2_sp_50m,sd)
-
-post_coeff_res2_sp_75m=response2_sp_75m$post_coeff(n_post=1000)
-post_par_res2_sp_75m=list("tau0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_75m$log_lambda[,2])))
-mean_par_res2_sp_75m=lapply(post_par_res2_sp_75m,mean)
-sd_par_res2_sp_75m=lapply(post_par_res2_sp_75m,sd)
-
-post_coeff_res2_sp_100m=response2_sp_100m$post_coeff(n_post=1000)
-post_par_res2_sp_100m=list("tau0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res2_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res2_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res2_sp_100m$log_lambda[,2])))
-mean_par_res2_sp_100m=lapply(post_par_res2_sp_100m,mean)
-sd_par_res2_sp_100m=lapply(post_par_res2_sp_100m,sd)
-
-```
-
-### Response model with splines of ExpShip based on baseline_racvm1
-
-```{r fit response3}
-
-#initialize parameters
-par0=c(0,0,1.5,4.5,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-#mapping for fixed coeffs
-new_map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
-                    "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,24))),coeff_fe=factor(rep(NA,5)),
-                    log_lambda=factor(c("tau.s(ExpShip)",NA,"nu.s(ExpShip)",NA,rep(NA,4))))
-
-#Fit response
-response3_sp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("log_sigma_obs0"=log(0.07)))
-
-new_coeff_re=c(rep(0,2),baseline_racvm3$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline_racvm3$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline_racvm3$lambda()["tau.s(ID)",1],1,baseline_racvm3$lambda()["nu.s(ID)",1],baseline_racvm3$lambda()[3:6,1])
-
-response3_sp$update_map(new_map)
-response3_sp$update_coeff_re(new_coeff_re)
-response3_sp$update_coeff_fe(baseline_racvm3$coeff_fe()[,1])
-
-response3_sp$update_lambda(new_lambda)
-response3_sp$fit()
-
-#Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_re=c(rep(0,2),baseline_racvm3_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline_racvm3_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline_racvm3_50m$lambda()["tau.s(ID)",1],1,baseline_racvm3_50m$lambda()["nu.s(ID)",1],baseline_racvm3_50m$lambda()[3:6,1])
-
-response3_sp_50m$update_map(new_map)
-response3_sp_50m$update_coeff_re(new_coeff_re)
-response3_sp_50m$update_coeff_fe(baseline_racvm3_50m$coeff_fe()[,1])
-
-response3_sp_50m$update_lambda(new_lambda)
-response3_sp_50m$fit()
-
-#Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_re=c(rep(0,2),baseline_racvm3_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline_racvm3_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline_racvm3_75m$lambda()["tau.s(ID)",1],1,baseline_racvm3_75m$lambda()["nu.s(ID)",1],baseline_racvm3_75m$lambda()[3:6,1])
-
-response3_sp_75m$update_map(new_map)
-response3_sp_75m$update_coeff_re(new_coeff_re)
-response3_sp_75m$update_coeff_fe(baseline_racvm3_75m$coeff_fe()[,1])
-
-response3_sp_75m$update_lambda(new_lambda)
-response3_sp_75m$fit()
-
-#Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_sp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_re=c(rep(0,2),baseline_racvm3_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
-               rep(0,2),baseline_racvm3_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(1,baseline_racvm3_100m$lambda()["tau.s(ID)",1],1,baseline_racvm3_100m$lambda()["nu.s(ID)",1],baseline_racvm3_100m$lambda()[3:6,1])
-
-response3_sp_100m$update_map(new_map)
-response3_sp_100m$update_coeff_re(new_coeff_re)
-response3_sp_100m$update_coeff_fe(baseline_racvm3_100m$coeff_fe()[,1])
-
-response3_sp_100m$update_lambda(new_lambda)
-response3_sp_100m$fit()
-```
-
-```{r get results response3}
-
-estimates_res3_sp=as.list(response3_sp$tmb_rep(),what="Est")
-std_res3_sp=as.list(response3_sp$tmb_rep(),what="Std")
-
-estimates_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Est")
-std_res3_sp_50m=as.list(response3_sp_50m$tmb_rep(),what="Std")
-
-estimates_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Est")
-std_res3_sp_75m=as.list(response3_sp_75m$tmb_rep(),what="Std")
-
-estimates_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Est")
-std_res3_sp_100m=as.list(response3_sp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/45)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/3)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plot_res3_sp=response3_sp$get_all_plots(baseline=baseline_racvm3,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res3_sp_50m=response3_sp_50m$get_all_plots(baseline=baseline_racvm3_50m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res3_sp_75m=response3_sp_75m$get_all_plots(baseline=baseline_racvm3_75m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plot_res3_sp_100m=response3_sp_100m$get_all_plots(baseline=baseline_racvm3_100m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-```
-
-```{r display plots response3_sp}
-
-#print plots
-plot_res3_sp
-plot_res3_sp_50m
-plot_res3_sp_75m
-plot_res3_sp_100m
-
-```
-
-```{r estimates on parameter scale response3}
-
-post_coeff_res3_sp=response3_sp$post_coeff(n_post=1000)
-post_par_res3_sp=list("tau0"=exp(post_coeff_res3_sp$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp$log_lambda[,2])))
-mean_par_res3_sp=lapply(post_par_res3_sp,mean)
-sd_par_res3_sp=lapply(post_par_res3_sp,sd)
-
-post_coeff_res3_sp_50m=response3_sp_50m$post_coeff(n_post=1000)
-post_par_res3_sp_50m=list("tau0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_50m$log_lambda[,2])))
-mean_par_res3_sp_50m=lapply(post_par_res3_sp_50m,mean)
-sd_par_res3_sp_50m=lapply(post_par_res3_sp_50m,sd)
-
-post_coeff_res3_sp_75m=response3_sp_75m$post_coeff(n_post=1000)
-post_par_res3_sp_75m=list("tau0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_75m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_75m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_75m$log_lambda[,2])))
-mean_par_res3_sp_75m=lapply(post_par_res3_sp_75m,mean)
-sd_par_res3_sp_75m=lapply(post_par_res3_sp_75m,sd)
-
-post_coeff_res3_sp_100m=response3_sp_100m$post_coeff(n_post=1000)
-post_par_res3_sp_100m=list("tau0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_sp_100m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_sp_100m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_sp_100m$log_lambda[,2])))
-mean_par_res3_sp_100m=lapply(post_par_res3_sp_100m,mean)
-sd_par_res3_sp_100m=lapply(post_par_res3_sp_100m,sd)
-
-```
-
-### Response model with exponential of ExpShip based on baseline2
+plots_res_racvm_sp
+```
+
+## Log linear RACVM response model with exponential of ExpShip
 
@@
                  nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")
-                 +ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))
+                 omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
 
 new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))
-
-# Fit response
-response2_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),
-                    other_data=list("log_sigma_obs0"=log(0.05)))
-
-new_coeff_fe=c(baseline2$coeff_fe()[1:3,1],0,baseline2$coeff_fe()[4,1],0,baseline2$coeff_fe()[5,1])
-new_coeff_re=c(baseline2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2$lambda()["tau.s(ID)",1],baseline2$lambda()["nu.s(ID)",1],baseline2$lambda()[3:6,1])
-
-response2_exp$update_map(new_map)
-response2_exp$update_coeff_re(new_coeff_re)
-response2_exp$update_coeff_fe(new_coeff_fe)
-
-response2_exp$update_lambda(new_lambda)
-
-response2_exp$fit()
-
-# Fit response with 50m measurement error
+                             log_lambda=factor(c(NA,NA,NA,NA)))
+
+#sigma_obs=exp(as.list(baseline_racvm2$tmb_rep(),what="Estimate")$log_sigma_obs)
 sigma_obs=0.05
 H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline2_50m$coeff_fe()[1:3,1],0,baseline2_50m$coeff_fe()[4,1],0,baseline2_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_50m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_50m$lambda()["tau.s(ID)",1],baseline2_50m$lambda()["nu.s(ID)",1],baseline2_50m$lambda()[3:6,1])
-
-response2_exp_50m$update_map(new_map)
-response2_exp_50m$update_coeff_re(new_coeff_re)
-response2_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_50m$update_lambda(new_lambda)
-
-response2_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline2_75m$coeff_fe()[1:3,1],0,baseline2_75m$coeff_fe()[4,1],0,baseline2_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_75m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_75m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_75m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_75m$lambda()["tau.s(ID)",1],baseline2_75m$lambda()["nu.s(ID)",1],baseline2_75m$lambda()[3:6,1])
-
-response2_exp_75m$update_map(new_map)
-response2_exp_75m$update_coeff_re(new_coeff_re)
-response2_exp_75m$update_coeff_fe(new_coeff_fe)
-response2_exp_75m$update_lambda(new_lambda)
-
-response2_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response2_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline2_100m$coeff_fe()[1:3,1],0,baseline2_100m$coeff_fe()[4,1],0,baseline2_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline2_100m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],baseline2_100m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(DistanceShore).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline2_100m$coeff_re()[paste("omega.ti(AngleNormal,DistanceShore).",1:16,sep=""),1])
-new_lambda=c(baseline2_100m$lambda()["tau.s(ID)",1],baseline2_100m$lambda()["nu.s(ID)",1],baseline2_100m$lambda()[3:6,1])
-
-response2_exp_100m$update_map(new_map)
-response2_exp_100m$update_coeff_re(new_coeff_re)
-response2_exp_100m$update_coeff_fe(new_coeff_fe)
-
-response2_exp_100m$update_lambda(new_lambda)
-
-response2_exp_100m$fit()
+# Fit response
+response_racvm_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",
+                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),
+                    other_data=list("H"=H))
+
+new_coeff_fe=c(baseline_racvm2$coeff_fe()[1:3,1],0,baseline_racvm2$coeff_fe()[4,1],0,
+               baseline_racvm2$coeff_fe()[5,1])
+new_coeff_re=c(baseline_racvm2$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],
+               baseline_racvm2$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
+               baseline_racvm2$coeff_re()[paste("omega.te(theta_translated,Ishore_translated).",1:24,sep=""),1])
+new_lambda=c(baseline_racvm2$lambda()["tau.s(ID)",1],
+             baseline_racvm2$lambda()["nu.s(ID)",1],baseline_racvm2$lambda()[3:4,1])
+
+response_racvm_exp$update_map(new_map)
+response_racvm_exp$update_coeff_re(new_coeff_re)
+response_racvm_exp$update_coeff_fe(new_coeff_fe)
+
+response_racvm_exp$update_lambda(new_lambda)
+
+response_racvm_exp$fit()
+
 ```
@@
 
-estimates_res2_exp=as.list(response2_exp$tmb_rep(),what="Est")
-std_res2_exp=as.list(response2_exp$tmb_rep(),what="Std")
-
-estimates_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Est")
-std_res2_exp_50m=as.list(response2_exp_50m$tmb_rep(),what="Std")
-
-estimates_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Est")
-std_res2_exp_75m=as.list(response2_exp_75m$tmb_rep(),what="Std")
-
-estimates_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Est")
-std_res2_exp_100m=as.list(response2_exp_100m$tmb_rep(),what="Std")
-
 #plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res2_exp=response2_exp$get_all_plots(baseline=baseline2,xmin=xmin,
+xmin=list("Ishore_translated"=1/3,"ExpShip"=1/60)
+xmax=list("Ishore_translated"=1/0.5,"ExpShip"=1/5)
+link=list("Ishore_translated"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
+xlabel=list("Ishore_translated"="Distance to shore","ExpShip"="Distance to ship")
+
+plots_res_racvm_exp=response_racvm_exp$get_all_plots(baseline=baseline_racvm2,xmin=xmin,
                             xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
 
-plots_res2_exp_50m=response2_exp_50m$get_all_plots(baseline=baseline2_50m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-plots_res2_exp_75m=response2_exp_75m$get_all_plots(baseline=baseline2_75m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-plots_res2_exp_100m=response2_exp_100m$get_all_plots(baseline=baseline2_100m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
 ```
@@
 #print plots
-plots_res2_exp
-plots_res2_exp_50m
-plots_res2_exp_75m
-plots_res2_exp_100m
-
-```
-
-### Response model with exponential of ExpShip based on baseline_racvm3
-
-```{r}
-
-par0=c(0,0,1,4.5,0)
-
-#define model
-formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
-                 nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
-
-#mapping for fixed coeffs
-new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                             log_lambda=factor(c(NA,NA,NA,NA,NA,NA)))
-
-#Fit response
-sigma_obs=exp(estimates_bas3$log_sigma_obs)
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3$coeff_fe()[1:3,1],0,baseline_racvm3$coeff_fe()[4,1],0,baseline_racvm3$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3$lambda()["tau.s(ID)",1],baseline_racvm3$lambda()["nu.s(ID)",1],baseline_racvm3$lambda()[3:6,1])
-
-response3_exp$update_map(new_map)
-response3_exp$update_coeff_re(new_coeff_re)
-response3_exp$update_coeff_fe(new_coeff_fe)
-
-response3_exp$update_lambda(new_lambda)
-response3_exp$fit()
-
-# Fit response with 30m measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_30m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_30m$coeff_fe()[1:3,1],0,baseline_racvm3_30m$coeff_fe()[4,1],0,baseline_racvm3_30m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_30m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_30m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_30m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_30m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_30m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_30m$lambda()["tau.s(ID)",1],baseline_racvm3_30m$lambda()["nu.s(ID)",1],baseline_racvm3_30m$lambda()[3:6,1])
-
-response3_exp_30m$update_map(new_map)
-response3_exp_30m$update_coeff_re(new_coeff_re)
-response3_exp_30m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_30m$update_lambda(new_lambda)
-response3_exp_30m$fit()
-
-# Fit response with 40m measurement error
-sigma_obs=0.04
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_40m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_40m$coeff_fe()[1:3,1],0,baseline_racvm3_40m$coeff_fe()[4,1],0,baseline_racvm3_40m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_40m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_40m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_40m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_40m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_40m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_40m$lambda()["tau.s(ID)",1],baseline_racvm3_40m$lambda()["nu.s(ID)",1],baseline_racvm3_40m$lambda()[3:6,1])
-
-response3_exp_40m$update_map(new_map)
-response3_exp_40m$update_coeff_re(new_coeff_re)
-response3_exp_40m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_40m$update_lambda(new_lambda)
-response3_exp_40m$fit()
-
-# Fit response with 45m measurement error
-sigma_obs=0.045
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_45m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_45m$coeff_fe()[1:3,1],0,baseline_racvm3_45m$coeff_fe()[4,1],0,baseline_racvm3_45m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_45m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_45m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_45m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_45m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_45m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_45m$lambda()["tau.s(ID)",1],baseline_racvm3_45m$lambda()["nu.s(ID)",1],baseline_racvm3_45m$lambda()[3:6,1])
-
-response3_exp_45m$update_map(new_map)
-response3_exp_45m$update_coeff_re(new_coeff_re)
-response3_exp_45m$update_coeff_fe(new_coeff_fe)
-response3_exp_45m$update_lambda(new_lambda)
-
-response3_exp_45m$fit()
-
-# Fit response with 50m measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_50m$coeff_fe()[1:3,1],0,baseline_racvm3_50m$coeff_fe()[4,1],0,baseline_racvm3_50m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_50m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_50m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_50m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_50m$lambda()["tau.s(ID)",1],baseline_racvm3_50m$lambda()["nu.s(ID)",1],baseline_racvm3_50m$lambda()[3:6,1])
-
-response3_exp_50m$update_map(new_map)
-response3_exp_50m$update_coeff_re(new_coeff_re)
-response3_exp_50m$update_coeff_fe(new_coeff_fe)
-
-response3_exp_50m$update_lambda(new_lambda)
-response3_exp_50m$fit()
-
-# Fit response with 75m measurement error
-sigma_obs=0.075
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_75m$coeff_fe()[1:3,1],0,baseline_racvm3_75m$coeff_fe()[4,1],0,baseline_racvm3_75m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_75m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_75m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_75m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_75m$lambda()["tau.s(ID)",1],baseline_racvm3_75m$lambda()["nu.s(ID)",1],baseline_racvm3_75m$lambda()[3:6,1])
-
-response3_exp_75m$update_map(new_map)
-response3_exp_75m$update_coeff_fe(new_coeff_fe)
-response3_exp_75m$update_lambda(new_lambda)
-
-response3_exp_75m$fit()
-
-# Fit response with 100m measurement error
-sigma_obs=0.1
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-response3_exp_100m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                    response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H))
-
-new_coeff_fe=c(baseline_racvm3_100m$coeff_fe()[1:3,1],0,baseline_racvm3_100m$coeff_fe()[4,1],0,baseline_racvm3_100m$coeff_fe()[5,1])
-new_coeff_re=c(baseline_racvm3_100m$coeff_re()[paste("tau.s(ID).",1:6,sep=""),1],baseline_racvm3_100m$coeff_re()[paste("nu.s(ID).",1:6,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal).",1:4,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(ExpShore).",1:4,sep=""),1],
-               baseline_racvm3_100m$coeff_re()[paste("omega.ti(AngleNormal,ExpShore).",1:16,sep=""),1])
-new_lambda=c(baseline_racvm3_100m$lambda()["tau.s(ID)",1],baseline_racvm3_100m$lambda()["nu.s(ID)",1],baseline_racvm3_100m$lambda()[3:6,1])
-
-response3_exp_100m$update_map(new_map)
-response3_exp_100m$update_coeff_re(new_coeff_re)
-response3_exp_100m$update_coeff_fe(new_coeff_fe)
-response3_exp_100m$update_lambda(new_lambda)
-
-response3_exp_100m$fit()
-```
-
-```{r}
-
-estimates_res3_exp=as.list(response3_exp$tmb_rep(),what="Est")
-std_res3_exp=as.list(response3_exp$tmb_rep(),what="Std")
-
-estimates_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Est")
-std_res3_exp_30m=as.list(response3_exp_30m$tmb_rep(),what="Std")
-
-estimates_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Est")
-std_res3_exp_40m=as.list(response3_exp_40m$tmb_rep(),what="Std")
-
-estimates_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Est")
-std_res3_exp_45m=as.list(response3_exp_45m$tmb_rep(),what="Std")
-
-estimates_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Est")
-std_res3_exp_50m=as.list(response3_exp_50m$tmb_rep(),what="Std")
-
-estimates_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Est")
-std_res3_exp_75m=as.list(response3_exp_75m$tmb_rep(),what="Std")
-
-estimates_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Est")
-std_res3_exp_100m=as.list(response3_exp_100m$tmb_rep(),what="Std")
-
-#plot parameters
-xmin=list("ExpShore"=1/D_up,"AngleNormal"=-pi,"ExpShip"=1/60)
-xmax=list("ExpShore"=1/D_low,"AngleNormal"=pi,"ExpShip"=1/5)
-link=list("ExpShore"=(\(x) 1/x),"ExpShip"=(\(x) 1/x))
-xlabel=list("ExpShore"="Distance to shore","ExpShip"="Distance to ship")
-
-plots_res3_exp=response3_exp$get_all_plots(baseline=baseline_racvm3,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-plots_res3_exp_30m=response3_exp_30m$get_all_plots(baseline=baseline_racvm3_30m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-plots_res3_exp_40m=response3_exp_40m$get_all_plots(baseline=baseline_racvm3_40m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-plots_res3_exp_50m=response3_exp_50m$get_all_plots(baseline=baseline_racvm3_50m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plots_res3_exp_75m=response3_exp_75m$get_all_plots(baseline=baseline_racvm3_75m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-plots_res3_exp_100m=response3_exp_100m$get_all_plots(baseline=baseline_racvm3_100m,xmin=xmin,
-                            xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
-
-```
-
-```{r display plots response3_sp}
-
-#print plots
-plots_res3_exp
-plots_res3_exp_50m
-plots_res3_exp_75m
-plots_res3_exp_100m
-
-```
-
-We select the response 3 model with 50m measurement error
-
-```{r}
+plots_res_racvm_exp
+
+```
+
+```{r save plots response_racvm_exp,results="hide",cache=TRUE,message=FALSE}
 # Create directory
-if (!dir.exists("response3_exp_50m")) {
-    dir.create("response3_exp_50m", recursive = TRUE)
+if (!dir.exists("response_racvm_exp")) {
+    dir.create("response_racvm_exp", recursive = TRUE)
 }
@@
 # Use lapply to save each plot
-lapply(seq_along(plots_bas2), function(i) {
-
-    plot<-plots_bas2[[i]]
-    plot_name <- names(plots_bas2)[i]
-    file_path <- file.path("response3_exp_50m", plot_name)
+lapply(seq_along(plots_res_racvm_exp), function(i) {
+
+    plot<-plots_res_racvm_exp[[i]]
+    plot_name <- names(plots_res_racvm_exp)[i]
+    file_path <- file.path("response_racvm_exp", plot_name)
 
@@
 
-### Plots improvements
-
-```{r}
-
-#get angular velocity for different distances to shore
-plot_me_tau <- plots_res3_exp_50m$me_tau_ExpShip
-plot_me_nu <- plots_res3_exp_50m$me_nu_ExpShip
-
-# Update plot_me_tau
-plot_me_tau <- plot_me_tau +
-  labs(y = expression(tau)) +
-  theme_minimal() +
-  theme(
-    axis.title.x = element_text(size=18),
-    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
-    axis.text.x = element_text(size=14),
-    axis.text.y = element_text(size=14),
-    legend.title = element_text(size=14),
-    legend.text = element_text(size=14)
+```{r results response_ctcrw_exp}
+
+get_final_response_tab<-function(sde,show=TRUE) {
+
+  post_coeff=sde$post_coeff(n_post=10000)
+  post_par=list(
+  "alpha_tau"=post_coeff$coeff_fe[,"tau.ExpShip"],
+  "alpha_nu"=post_coeff$coeff_fe[,"nu.ExpShip"])
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
+                      quantile, probs = c(0.025, 0.975))
+
+  parameter_names <- names(mean_par)
+  estimates <- mean_par
+  conf_intervals <- sapply(parameter_names, function(param) {
+    sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
+            quant_par[[param]][2])
+  })
+
+  table_data <- data.frame(
+    Parameter = 
+      c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
+    Estimate = sprintf("$%.2f$", mean_par),
+    CI = conf_intervals,
+    stringsAsFactors = FALSE
   )
-
-# Update plot_me_nu
-plot_me_nu <- plot_me_nu +
-  labs(y = expression(nu)) +
-  theme_minimal() +
-  theme(
-    axis.title.x = element_text(size=18),
-    axis.title.y = element_text(size=18, angle=0, vjust=0.5, margin = margin(r = 10)),
-    axis.text.x = element_text(size=14),
-    axis.text.y = element_text(size=14),
-    legend.title = element_text(size=14),
-    legend.text = element_text(size=14)
+  xtab <- xtable(
+    table_data,
+    caption = "Response log linear estimations",
+    label = "table:response_log_linear_estimations"
   )
 
-# Save the updated plots
-ggsave("plot_me_tau_updated.pdf", plot=plot_me_tau, width=10, height=6, units="in")
-ggsave("plot_me_nu_updated.pdf", plot=plot_me_nu, width=10, height=6, units="in")
-
-```
-
-## Posterior samples
-
-```{r}
-post_coeff_res3_exp_50m=response3_exp_50m$post_coeff(n_post=10000)
-post_par_res3_exp_50m=list(
-              "tau0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"tau.(Intercept)"]),
-              "nu0"=exp(post_coeff_res3_exp_50m$coeff_fe[,"nu.(Intercept)"]),
-              "omega0"=post_coeff_res3_exp_50m$coeff_fe[,"omega.(Intercept)"],
-              "sigma_tau"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,1])),
-              "sigma_nu"=1/sqrt(exp(post_coeff_res3_exp_50m$log_lambda[,2])),
-              "tau.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"tau.ExpShip"],
-              "nu.ExpShip"=post_coeff_res3_exp_50m$coeff_fe[,"nu.ExpShip"])
-mean_par_res3=lapply(post_par_res3_exp_50m,mean)
-sd_par_res3=lapply(post_par_res3_exp_50m,sd)
-quant_par_res3 <- lapply(post_par_res3_exp_50m, quantile, probs = c(0.025, 0.975))
-
-```
-
-#### Thorough uncertainty quantification for response3_exp_50m
-
-```{r}
-
+  if (show) {
+    print(
+    xtab,type="latex",
+    include.rownames = FALSE,
+    sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
+    hline.after = c(-1, 0, nrow(table_data)),
+    table.placement="H")
+  }
+
+  return (xtab)
+
+  }
+
+```
+
+```{r,results='asis',echo=FALSE}
+final_tab_response_exp=get_final_response_tab(response_racvm_exp)
+
+```
+
+These confidence intervals do not include the uncertainty from the baseline estimations.
+
+## Thorough uncertainty quantification
+
+```{r, message=FALSE,results="hide"}
+
+post_coeff_bas_racvm2=baseline_racvm2$post_coeff(n_post=1000)
 #posterior samples of the coefficient of the baseline model
-post_coeff_re=post_coeff_bas3_50m$coeff_re[1:100,]
-post_coeff_fe=post_coeff_bas3_50m$coeff_fe[1:100,]
-post_coeff_lambda=exp(post_coeff_bas3_50m$log_lambda[1:100,])
-
-# Fit response with 50m measurement error
-
-par0=c(0,0,1,4.5,0)
+post_coeff_re=post_coeff_bas_racvm2$coeff_re[1:100,]
+post_coeff_fe=post_coeff_bas_racvm2$coeff_fe[1:100,]
+post_coeff_lambda=exp(post_coeff_bas_racvm2$log_lambda[1:100,])
+
+par0=c(0,0,1,4,0)
+
+#define model
 formulas <- list(mu1=~1,mu2=~1,tau =~ExpShip+s(ID,bs="re"),
                  nu=~ExpShip+s(ID,bs="re"),
-                 omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))
+                 omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
+
+new_map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
+                             log_lambda=factor(c(NA,NA,NA,NA)))
 
@@
 
-estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
-std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+alpha_estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+alpha_std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+
+models_list=list()
 
@@
 
-    response3_exp_50m_temp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",
-                      response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),other_data=list("H"=H),
-                      map=list(coeff_re=factor(c(rep(NA,12),rep(NA,24))),coeff_fe=factor(c(NA,NA,NA,1,NA,2,NA)),
-                               log_lambda=factor(c(NA,NA,NA,NA,NA,NA))))
+    response_racvm_exp_temp<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM_SSM",
+                        response = c("x","y"),par0 = par0,fixpar=c("mu1","mu2"),
+                        other_data=list("H"=H))
 
     new_coeff_fe=c(post_coeff_fe[i,1:3],0,post_coeff_fe[i,4],0,post_coeff_fe[i,5])
+
     new_coeff_re=post_coeff_re[i,]
     new_lambda=post_coeff_lambda[i,]
-    response3_exp_50m_temp$update_coeff_re(new_coeff_re)
-    response3_exp_50m_temp$update_coeff_fe(new_coeff_fe)
-
-    response3_exp_50m_temp$update_lambda(new_lambda)
-    response3_exp_50m_temp$fit()
-
-    est_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
-    std_alpha=as.list(response3_exp_50m_temp$tmb_rep(),what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
-
-    estimates[i,]=est_alpha
-    std[i,]=std_alpha
-}
-
-colnames(estimates)=c("tau.ExpShip","nu.ExpShip")
-colnames(std)=c("tau.ExpShip","nu.ExpShip")
+
+    response_racvm_exp_temp$update_map(new_map)
+    response_racvm_exp_temp$update_coeff_re(new_coeff_re)
+    response_racvm_exp_temp$update_coeff_fe(new_coeff_fe)
+
+    response_racvm_exp_temp$update_lambda(new_lambda)
+
+    response_racvm_exp_temp$fit()
+
+    est_alpha=as.list(response_racvm_exp_temp$tmb_rep(),what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+    std_alpha=as.list(response_racvm_exp_temp$tmb_rep(),what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+
+    alpha_estimates[i,]=est_alpha
+    alpha_std[i,]=std_alpha
+
+    models_list[[i]]=response_racvm_exp_temp
+}
+
+colnames(alpha_estimates)=c("tau.ExpShip","nu.ExpShip")
+colnames(alpha_std)=c("tau.ExpShip","nu.ExpShip")
+
+```
+
+```{r propagate uncertainty coefficient log -linear model}
+
+mean_alpha=list(alpha_tau=mean(alpha_estimates[,1]),
+                alpha_nu=mean(alpha_estimates[,2]))
+quant_alpha<- list(
+  alpha_tau=quantile(alpha_estimates[,1],probs = c(0.025, 0.975)),
+  alpha_nu=quantile(alpha_estimates[,2],probs = c(0.025, 0.975)))
+
+```
+
+```{r propagated CI alpha log linear coefficients, message=FALSE}
+
+parameter_names <- names(mean_alpha)
+estimates <- mean_alpha
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1], 
+          quant_alpha[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = 
+    c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
+  Estimate = sprintf("$%.2f$", mean_alpha),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+
+```{r,results='asis',echo=FALSE}
+xtab <- xtable(
+  table_data,
+  caption = "Response log linear estimations with propagated uncertainty",
+  label = "table:response_log_linear_estimations_with_propagated_uncertainty"
+)
+
+print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  
+  hline.after = c(-1, 0, nrow(table_data)),
+  table.placement="H"
+
+)
+```
+
+```{r all CI plots,cache=TRUE}
+
+CI_plots=lapply(models_list,function(model) { 
+  model$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                      link=link,xlabel=xlabel,show_CI="pointwise")})
+
+```
+
+```{r plot smooth parameters with propagated uncertainty,results='hide'}
+
+# Extract data for each estimated smooth parameter
+tau_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+})
+nu_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+})
+
+# Get confidence intervals as quantiles of estimated parameters
+
+# Combine the x columns from all data frames into a matrix
+x_matrix_tau <- do.call(cbind, lapply(tau_CI_data, function(df) df$y))
+x_matrix_nu <- do.call(cbind, lapply(nu_CI_data, function(df) df$y))
+# Compute the 5% and 95% quantiles row-wise
+quantiles_tau <- apply(x_matrix_tau, 1, 
+                       function(row) quantile(row, probs = c(0.025, 0.975)))
+quantiles_nu <- apply(x_matrix_nu, 1,
+                      function(row) quantile(row, probs = c(0.025, 0.975)))
+
+# Transpose and format the results into a data frame
+quantiles_tau <- as.data.frame(t(quantiles_tau))
+colnames(quantiles_tau) <- c("low", "up")
+quantiles_nu <- as.data.frame(t(quantiles_nu))
+colnames(quantiles_nu) <- c("low", "up")
+
+# Add the "true" estimated smooth
+plots=response_racvm_exp$get_all_plots(
+  baseline=baseline_racvm2,xmin=xmin,xmax=xmax,
+  link=link,xlabel=xlabel,show_CI="pointwise")
+est_tau=ggplot_build(plots$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+est_nu=ggplot_build(plots$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+est_tau=cbind(est_tau,quantiles_tau)
+est_nu=cbind(est_nu,quantiles_nu)
+
+est_tau$baseline=ggplot_build(plots$fe_tau_ExpShip)$data[[2]][,"y"]
+est_nu$baseline=ggplot_build(plots$fe_nu_ExpShip)$data[[2]][,"y"]
+
+plot_tau_response_exp <- ggplot(data = est_tau) +
+  geom_line(aes(x = x, y = y), color = "steelblue", size = 1) +
+  geom_line(aes(x=x,y=baseline),color="steelblue",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "lightblue", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = expression(tau)) +
+  theme_minimal() +theme(
+    axis.title.x = element_text(size=22),
+    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=18),
+    axis.text.y = element_text(size=18),
+    legend.title = element_text(size=18),
+    legend.text = element_text(size=18)
+  )
+
+# Enhanced plot for nu
+plot_nu_response_exp <- ggplot(data = est_nu) +
+  geom_line(aes(x = x, y = y), color = "darkorange", size = 1) +
+   geom_line(aes(x,baseline),color="darkorange",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "navajowhite", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = expression(nu)) +
+  theme_minimal() +theme(
+    axis.title.x = element_text(size=22),
+    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=18),
+    axis.text.y = element_text(size=18),
+    legend.title = element_text(size=18),
+    legend.text = element_text(size=18)
+  )
 
@@
 ```{r}
-#The 9th value in estimates is inconsistent, we drop it
-estimates=estimates[-9,]
-std=std[-9,]
-
-mean_est=apply(X=estimates,MARGIN=2,FUN=mean)
-mean_std=apply(X=na.omit(std),MARGIN=2,FUN=mean)
-```
+
+ggsave("fe_tau_ExpShip_final.pdf", plot=plot_tau_response_exp,
+       path="response_racvm_exp", width=10, height=6, units="in")
+
+ggsave("fe_nu_ExpShip_final.pdf", plot=plot_nu_response_exp,
+       path="response_racvm_exp", width=10, height=6, units="in")
+plot_tau_response_exp
+plot_nu_response_exp
+
+```
+
+## Compute recovery distances
+
+```{r}
+get_recovery_distances_tab<-function(p) {
+
+  parameter_names=names(mean_alpha)
+  scaling=as.list(c(log(1-p),log(1+p)))
+  names(scaling)=parameter_names
+  estimates <- sapply(parameter_names,function(param) {
+    mean_alpha[[param]]/scaling[[param]]})
+
+  conf_intervals <- sapply(parameter_names, function(param) {
+    sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1]/scaling[[param]], 
+            quant_alpha[[param]][2]/scaling[[param]])
+  })
+
+  table_data <- data.frame(
+    Parameter = 
+      c("$D_{\\tau}^{ship}$","$D_{\\nu}^{ship}$"),
+    Estimate = sprintf("$%.2f$", estimates),
+    CI = conf_intervals,
+    stringsAsFactors = FALSE
+  )
+
+  xtab <- xtable(
+  table_data,
+  caption = "Estimated recovery distances of the baseline values",
+  label = paste0("table:recovery_distances_",p)
+  )
+
+  print(
+    xtab,type="latex",
+    include.rownames = FALSE,
+    sanitize.text.function = identity,  
+    hline.after = c(-1, 0, nrow(table_data)),
+    table.placement="H"
+
+  )
+}
+
+get_recovery_distances_tab(0.5)
+
+get_recovery_distances_tab(0.1)
+```

### 2025-02-16 R/simulation_study/examples/illustrative_trajectories.Rmd
update .gitignore

--- 2025-01-12 R/simulation_study_CRCVM/CRCVM_examples/illustrative_trajectories.Rmd
@@
 ```{r}
-fomega_cubic=function(cov_data,a=1,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.2,b=5){
+fomega_cubic=function(cov_data,a=1,b=5,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.3){
   Dshore=cov_data$DistanceShore
@@
 
-Finally, we compare it to Hanks simulation technique for reflected SDE.
+Finally, we compare this to Hanks simulation technique for reflected SDE.
 
@@
 ![Simulated Trajectories](sim_fjords_cubic_attractive_sub.gif)
-```{r}
-
-# Add covariates to simulation -----------
-
-signed_angle <- function(u, v) {
-  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
-  # as in 
-  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
-  u <- matrix(u, ncol = 2)
-  v <- matrix(v, ncol = 2)
-  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
-                                  rows")
-  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
-  ind1 <- which(result > pi)
-  ind2 <- which(result <= -pi)
-  result[ind1] <- result[ind1] - 2*pi
-  result[ind2] <- result[ind2] + 2*pi
-  return(result) 
-}
-
-add_covs=function(data) {
-  #data has at least columns "time","y1" "y2", "p1","p2"
-  #this function new data frame with three more columns "theta", "DistanceShore" and "ExpShore"
-
-  new_data=data
-  #observed deviation angles
-  n=length(new_data$time)
-
-  new_data$theta=rep(NA,n)
-
-  for (id in unique(new_data$ID)) {
-
-    #filter data with specific id
-    sub_ind=(new_data$ID==id)
-    sub_data=new_data[sub_ind,]
-    n_sub=length(sub_data$time)
-
-    #time steps
-    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]
-
-    #step lengths
-    dx=sub_data[2:n_sub,"y1"]-sub_data[1:(n_sub-1),"y1"]
-    dy=sub_data[2:n_sub,"y2"]-sub_data[1:(n_sub-1),"y2"]
-
-    #matrix of empirical velocity
-    vexp_df=cbind(dx/dtimes,dy/dtimes)
-
-    #matrix of normal vectors
-    normal=as.matrix(sub_data[2:n_sub,c("y1","y2")]-sub_data[2:n_sub,c("p1","p2")])
-
-    #angle between velocity and normal vector
-    theta_coast=signed_angle(normal,vexp_df)
-
-    #adjust lengths 
-    theta_coast=c(theta_coast,1)
-
-    new_data[sub_ind,"theta"]=theta_coast
-  }
-
-  #distance to shore
-  Dshore=sqrt((new_data$y1-new_data$p1)^2+(new_data$y2-new_data$p2)^2)
-  new_data$DistanceShore=Dshore
-  new_data$ExpShore=1/Dshore
-
-  return(new_data)
-}
-
-```
-
-```{r}
-data_sim_fjords_cubic_attractive_sub=add_covs(data_sim_fjords_cubic_attractive_sub)
-data_sim_fjords_cubic_attractive_sub$angular_velocity=
-  c(NA,diff(data_sim_fjords_cubic_attractive_sub$theta)/
-      diff(data_sim_fjords_cubic_attractive_sub$time))
-
-ggplot(data=data_sim_fjords_cubic_attractive_sub)+geom_point(aes(x=time,y=DistanceShore,col=ID))+
-  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)
-
-ggplot(data=data_sim_fjords_cubic_attractive_sub)+geom_point(aes(x=time,y=angular_velocity,col=ID))+
-  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)
-```
 
@@
 fomega_cubic_fjords_repulsive=function(cov_data) {
-  fomega_cubic(cov_data,a=40,D0=0.3,D1=0.8,sigma_theta=pi/10,sigma_D=0.3,b=0)
+  fomega_cubic(cov_data,a=1,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.3,b=0)
 }
@@
 ![Simulated Trajectories](sim_fjords_cubic_repulsive_sub.gif)
-```{r}
-data_sim_fjords_cubic_advanced_gaussian_sub=add_covs(data_sim_fjords_cubic_advanced_gaussian_sub)
-data_sim_fjords_cubic_advanced_gaussian_sub$angular_velocity=
-  c(NA,diff(data_sim_fjords_cubic_advanced_gaussian_sub$theta)/
-      diff(data_sim_fjords_cubic_advanced_gaussian_sub$time))
-
-ggplot(data=data_sim_fjords_cubic_advanced_gaussian_sub)+geom_point(aes(x=time,y=DistanceShore,col=ID))+
-  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)
-
-ggplot(data=data_sim_fjords_cubic_advanced_gaussian_sub)+geom_point(aes(x=time,y=angular_velocity,col=ID))+
-  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)
-```
 
@@
 stopCluster(cl)
-```
-
-```{r simulate hanks reflected in rectangle,cache=TRUE}
 ```
@@
     geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
+     scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+
     theme_minimal() + 
@@
   theme(
-    axis.title = element_text(size = 16),
-    axis.text = element_text(size = 12),
-    legend.text = element_text(size = 16),
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
     axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification

### 2025-02-16 R/simulation_study/local/parametric.Rmd
update .gitignore

---
title: "Simulation study parametric"
author: "Alexandre Delporte"
date: "2025-02-12"
output: html_document
---

Estimation with parametric omega function

# Set up for simulation study
```{r, cache=TRUE}
library(here)
library(htmlwidgets)
source(file.path(here("R","simulation_study","local","set_up.R")))  
```

# Low frequency high error

First, we will fix the new parameters that are introduced in the model, and estimate the others, that is $\tau$ and $\nu$.

```{r}

data_lf_he_prep=data_list[["data_lf_he"]]

H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_lf_he_prep$time)),dim=c(2,2,length(data_lf_he_prep$time)))
# Estimate from simulated data ------------------

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

colnames(data_lf_he_prep)[colnames(data_lf_he_prep)==c("observed_BoundaryDistance","observed_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")

crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
                            data = data_lf_he_prep,type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data = list("H"=H))

#fit
crcvm_cubic_lf_he$fit(method="BFGS")
```
```{r}

crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")
exp(as.list(crcvm_cubic_lf_he$tmb_rep(),what="Estimate")$log_sigma_obs)

```

Now we use interpolated distances and angles at high frequency to approximate better the likelihood.
First we compute and plot the interpolation.

```{r}
#prepare data for interpolation
data_lf_he_prep=data_list[["data_lf_he"]][,c("ID","y1","y2","time","observed_BoundaryDistance","observed_BoundaryAngle")]

colnames(data_lf_he_prep)[colnames(data_lf_he_prep) %in% c("observed_BoundaryDistance","observed_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")
n_step=3

#compute spline interpolation
interpolation_data<-interpolate_BoundaryMetrics(data_lf_he_prep,response=c("y1","y2"),
                                                border,n_step,n_cores=6,k=0.5*nrow(data_lf_he_prep))

#create dataframe with interpolated values
interpolated_x_vec=as.vector(t(interpolation_data$x))
interpolated_y_vec=as.vector(t(interpolation_data$y))
interpolated_distance_vec=as.vector(t(interpolation_data$BoundaryDistance))
interpolated_angle_vec=as.vector(t(interpolation_data$BoundaryAngle))
interpolation_time=as.vector(t(interpolation_data$time))
ID=unlist(lapply(unique(data_lf_he_prep$ID), function(id) {
  n=nrow(data_lf_he_prep[data_lf_he_prep$ID==id,])
  rep(id,n*n_step)
}))

#combine dataframes
interpolated_df=data.frame("time"=interpolation_time,"ID"=ID,"y1"=interpolated_x_vec,
                           "y2"=interpolated_y_vec,
                           "BoundaryDistance"=interpolated_distance_vec,
                           "BoundaryAngle"=interpolated_angle_vec)
interpolated_df$type="Interpolated"
data_lf_he_prep$type="Observed"
true_data_lf_he=data_sub[,c("ID","y1","y2","time","true_BoundaryDistance","true_BoundaryAngle")]
colnames(true_data_lf_he)[colnames(true_data_lf_he) %in% c("true_BoundaryDistance","true_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")
true_data_lf_he$type="True"

combined_data<-rbind(interpolated_df,data_lf_he_prep,true_data_lf_he)

# Create plot with legends for shape and linetype
trajectory_plot = ggplot() +geom_sf(data=border$geometry,fill="grey")+
  geom_path(data = combined_data[combined_data$time < 11, ], 
             aes(y1,y2, col = ID, linetype = type), size = 1) +

  # Customize the legend
  labs(color = "Trajectory ID", 
       shape = "Point Type", 
       linetype = "Line Type", 
       x = "Time", y = "Boundary Distance", 
       title = "Interpolated vs. Observed trajectories") +
  scale_shape_manual(values = c(1:3))+
  theme_minimal() +
  theme(legend.position = "right")  # Adjust legend position

distance_plot = ggplot() +
  # Scatter plot for observed BoundaryDistance (Shape Legend)
  geom_line(data = combined_data[combined_data$time < 11, ], 
             aes(time, BoundaryDistance, col = ID, linetype = type), size = 1) +

  # Customize the legend
  labs(color = "Trajectory ID", 
       shape = "Point Type", 
       linetype = "Line Type", 
       x = "Time", y = "Boundary Distance", 
       title = "Interpolated vs. True Boundary Distance") +

  theme_minimal() +
  theme(legend.position = "right")  # Adjust legend position

# Create plot with legends for shape and linetype
angle_plot = ggplot() +
  # Scatter plot for observed BoundaryDistance (Shape Legend)
  geom_line(data = combined_data[combined_data$time < 11 & combined_data$ID==1, ], 
             aes(time, BoundaryAngle, col = ID, linetype = type), size = 1) +

  # Customize the legend
  labs(color = "Trajectory ID", 
       shape = "Point Type", 
       linetype = "Line Type", 
       x = "Time", y = "Boundary Angle", 
       title = "Interpolated vs. True Boundary Angle") +

  theme_minimal() +
  theme(legend.position = "right")  # Adjust legend position

```

```{r}
# Display the plot
dyn_trajectory_plot=ggplotly(trajectory_plot)

dyn_distance_plot=ggplotly(distance_plot)
dyn_angle_plot=ggplotly(angle_plot)

saveWidget(dyn_trajectory_plot,"interpolated_trajectory.html")
saveWidget(dyn_distance_plot,"interpolated_distance.html")
saveWidget(dyn_angle_plot,"interpolated_angle.html")
```

```{r}

H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_lf_he_prep$time)),dim=c(2,2,length(data_lf_he_prep$time)))

crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
                            data = data_lf_he_prep,type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data = list("H"=H,"interpolated_distance"=interpolation_data$BoundaryDistance,
                                              "interpolated_angle"=interpolation_data$BoundaryAngle))

#fit
crcvm_cubic_lf_he$fit(method="BFGS")
```

```{r}

crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")

```

Now we try to estimate the parametric function for omega.

```{r}
H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_lf_he_prep$time)),dim=c(2,2,length(data_lf_he_prep$time)))

crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
                            data = data_lf_he_prep,type = "CRCVM_SSM",
                    response = c("y1","y2"),fixpar=c("sigma_D","sigma_theta"),
                    par0 = c(1,1,1,1,2,1.5,SIGMA_D,SIGMA_THETA),
                    other_data = list("H"=H,"interpolated_distance"=interpolation_data$BoundaryDistance,
                                              "interpolated_angle"=interpolation_data$BoundaryAngle))

#fit
crcvm_cubic_lf_he$fit(method="BFGS")
```

```{r}

crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")

```

# Low frequency low error

```{r}
data_lf_le_prep=data_list[["data_lf_le"]]

colnames(data_lf_le_prep)[colnames(data_lf_le_prep)==c("observed_BoundaryDistance","observed_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")

H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_lf_le$time)),dim=c(2,2,length(data_lf_le$time)))
# Estimate from simulated data ------------------

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_lf_le<- SDE$new(formulas = formulas,data = data_lf_le_prep,
                              type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("H"=H))

crcvm_cubic_lf_le$fit(method="BFGS")
```

```{r}

crcvm_cubic_lf_le$get_all_plots(show_CI="simultaneous")

```

```{r}
#compute spline interpolation
interpolation_data<-interpolate_BoundaryMetrics(data_lf_le_prep,response=c("y1","y2"),
                                                border,n_step=1,n_cores=6,k=100)

H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_lf_le_prep$time)),dim=c(2,2,length(data_lf_le_prep$time)))

crcvm_cubic_lf_le<- SDE$new(formulas = formulas,
                            data = data_lf_le_prep,type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data = list("H"=H,"interpolated_distance"=interpolation_data$BoundaryDistance,
                                              "interpolated_angle"=interpolation_data$BoundaryAngle))

#fit
crcvm_cubic_lf_le$fit(method="BFGS")
```

```{r}

crcvm_cubic_lf_le$get_all_plots(show_CI="simultaneous")

```

Let's now try with high frequency data.

# High frequency high error

```{r}
H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_hf_he$time)),dim=c(2,2,length(data_hf_he$time)))

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_he_fixed<- SDE$new(formulas = formulas,data = data_hf_he[data_hf_he$DistanceShore>0.01 & data_hf_he$ID %in% c(1:6),],type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 =c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("H"=H))

crcvm_cubic_hf_he_fixed$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_he_fixed$get_all_plots()
exp(as.list(crcvm_cubic_hf_he_fixed$tmb_rep(),what="Estimate")$log_sigma_obs)
```

Now we estimate the parameters for the function omega with the true distances and angles.

```{r}

data_hf_he_prep=data_list[["data_hf_he"]]

colnames(data_hf_he_prep)[colnames(data_hf_he_prep)==c("true_BoundaryDistance","true_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")

H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_hf_he_prep$time)),dim=c(2,2,length(data_hf_he_prep$time)))

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_he<- SDE$new(formulas = formulas,data = data_hf_he_prep,
                            type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,1,1,1,1,1,0.2),
                    other_data=list("H"=H))

#fit
crcvm_cubic_hf_he$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_he$get_all_plots()
exp(as.list(crcvm_cubic_hf_he$tmb_rep(),what="Estimate")$log_sigma_obs)
```

# High frequency low error

First fix the parametric function and estimate the other parameters of the SDE.
```{r}

H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_hf_le$time)),dim=c(2,2,length(data_hf_le$time)))

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_le_fixed<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% c(1:6),],
                            type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
                    other_data=list("H"=H))

#fit
crcvm_cubic_hf_le_fixed$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_le_fixed$get_all_plots()
exp(as.list(crcvm_cubic_hf_le_fixed$tmb_rep(),what="Estimate")$log_sigma_obs)
```

Now we try to estimate D0 and a in the parametric function.

```{r}

data_hf_le_prep=data_list[["data_hf_le"]]

colnames(data_hf_le_prep)[colnames(data_hf_le_prep)==c("true_BoundaryDistance","true_BoundaryAngle")]<-
  c("BoundaryDistance","BoundaryAngle")

H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_hf_le$time)),dim=c(2,2,length(data_hf_le$time)))

formulas <- list(tau=~1,
                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)

crcvm_cubic_hf_le<- SDE$new(formulas = formulas,data = data_hf_le_prep,
                            type = "CRCVM_SSM",
                    response = c("y1","y2"),par0 = c(1,1,1,1,1,2,1,0.2),
                    other_data=list("H"=H))

#fit
crcvm_cubic_hf_le$fit(method="BFGS")
```

```{r}
crcvm_cubic_hf_le$get_all_plots()
exp(as.list(crcvm_cubic_hf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
```

```{r}
ggplot()+geom_point(data=data_hf_le,aes(time,theta),col="blue",alpha=0.5,size=0.1)+
  geom_point(data=data,aes(time,theta),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)

```
```{r}
data_hf_le$omega=fomega(data_hf_le[,c("theta","DistanceShore")])

ggplot()+geom_point(data=data_hf_le,aes(time,omega),col="blue",alpha=0.5,size=0.1)+
  geom_point(data=data,aes(time,omega),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)
```

### 2025-02-16 R/simulation_study/local/splines.Rmd
update .gitignore

--- 2025-01-19 R/simulation_study_CRCVM/estimation_local/simulation_study_local.Rmd
@@
 
+# Set up for simulation study
 ```{r}
-library(here)
-library(sf)
-library(smoothSDE)
-library(doParallel)
-library(ggplot2)
-library(dplyr)
-library(mgcv)
-library(plotly)
-
-#get functions to simulate trajectories
-source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))
-
+source(file.path(here("R","simulation_study_CRCVM","estimation_local","set_up.R")))  
 ```
 
-# Simple framework : no random effects
+# Estimate CTCRW
 
-## Define hyperparameters
-
-```{r hyperparameters}
-N_ID=6
-
-TMAX=12
-DELTA=1/60/6
-
-SP_DF=c(5,5)
-
-TAU_0=1.3
-NU_0=4.6
-
-DMIN=0.5
-DMAX=1
-
-PAR0_RACVM=c(0,0,1,1,0)
-
-SIGMA_OBS_LOW=0.01
-SIGMA_OBS_HIGH=0.04
-
-BY_LF=5
-BY_HF=1
-
-D_LOW=0.05
-D_UP=3
-
-A=1
-D0=0.3
-D1=1.2
-SIGMA_THETA=pi/6
-SIGMA_D=1
-B=1
-
-```
-
-```{r initial positions}
-
-seed=42
-set.seed(seed)
-# Set the path to the directory containing the data
-greenland_data_path <- here("Data","preprocessed_data","greenland")
-
-#get the land geometry from shapefile
-border<-st_read(file.path(greenland_data_path,"updated_scoresby_sound_utm.shp"))
-border <- st_transform(border, crs = "+init=EPSG:32626 +units=km")
-
-shrunk_border<-st_read(file.path(greenland_data_path,"shrunk_updated_scoresby_sound_utm.shp"))
-shrunk_border <- st_transform(border, crs = "+init=EPSG:32626 +units=km")
-v0=c(0,0)
-
-#generate random initial points
-x0=matrix(rep(NA,N_ID*2),ncol=2)
-colnames(x0)=c("x1","x2")
-
-i=1
-while (i<=N_ID) {
-  #choose location uniformly in the map
-  x=c(runif(1,min=430,max=500),runif(1,min=7760,max=7900))
-  if (!(is_in_land(st_point(x),border))) {
-    p=nearest_boundary_point(st_point(x),border)
-    Dshore=(x[1]-p[1])^2+(x[2]-p[2])^2
-    #keep it as initial position if it is at least 50 metres away from the shore
-    if (Dshore>DMIN && Dshore<DMAX) {
-      x0[i,]=x
-      i=i+1
-    }
-  }
-}
-
-```
-
-## Define SDE parameters
-
-```{r model parameters}
-# Time steps ----------------
-
-#high frequency time points
-times=seq(0,TMAX,by=DELTA)
-
-n_obs=length(times)-1
-
-# Defintion of smooth parameter omega ----------
-
-fomega_cubic=function(cov_data,a,b,D0,D1,sigma_D,sigma_theta){
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$Ishore
-  }
-
-  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
-      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
-           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
-}
-
-fomega=function(cov_data) {
-  fomega_cubic(cov_data,A,B,D0,D1,SIGMA_D,SIGMA_THETA)
-}
-
-ftau_constant=function(cov_data) {
-  TAU_0
-}
-
-fnu_constant=function(cov_data) {
-  NU_0
-}
-
-```
-
-## Simulate true trajectories
-
-```{r,cache=TRUE}
-
-# Generate samples ---------------
-
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data=foreach (i=1:N_ID,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N_ID+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
-                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,
-                land=shrunk_border,verbose=FALSE,save_omega = TRUE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore
-
-  data_sim=cbind(data_sim,data_shore)
-  data_sim
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-## Remove trajectories that reached land
-
-```{r}
-
-# Points that reached land ---------------
-count=0
-remove_ID=c()
-for (id in unique(data$ID)) {
-  id_data=data[data$ID==id,]
-  if (nrow(id_data) < n_obs) {
-    count=count+1
-    remove_ID=c(remove_ID,id)
-    cat("ID",id,"reached land","\n",sep=" ")
-  }
-}
-
-cat(count/N_ID*100,"percent of the trajectories reached land","\n")
-
-data_sub=data %>%  filter(!(ID %in% remove_ID))
-
-```
-
-## Add noise to get observed trajectories
-
-```{r}
-# add noise
-low_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_LOW^2,2)))
-high_noise=rmvn(n_obs,rep(0,2),diag(rep(SIGMA_OBS_HIGH^2,2)))
-observed_data_le=data_sub[,c("y1","y2","time","ID")]
-observed_data_le[,c("y1","y2")]=data_sub[,c("y1","y2")]+low_noise
-observed_data_he=data_sub[,c("y1","y2","time","ID")]
-observed_data_he[,c("y1","y2")]=data_sub[,c("y1","y2")]+high_noise
-
-#subsample
-data_lf_he=observed_data_he[seq(1,length(data_sub$time),by=BY_LF),]
-data_lf_le=observed_data_le[seq(1,length(data_sub$time),by=BY_LF),]
-data_hf_le=observed_data_le[seq(1,length(data_sub$time),by=BY_HF),]
-data_hf_he=observed_data_he[seq(1,length(data_sub$time),by=BY_HF),]
-
-# Save plot of the trajectories ------------
-
-plot=ggplot()+geom_sf(data=border$geometry,fill="grey")+
-  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
-  geom_point(data=data_sub,mapping=aes(y1,y2,col=ID),size=0.1,alpha=0.2,shape=2)+
-  geom_point(data=data_lf_he,mapping=aes(y1,y2,col=ID),size=0.1)+
-  geom_path(data=data_lf_he,mapping=aes(y1,y2,col=ID),size=0.1)+
-  geom_point(data = data%>% filter(!duplicated(ID)),
-             aes(x = y1, y = y2), shape = 3, size = 4, col = "red")+
-  xlab("x") + ylab("y")
-
-ggsave(filename="plot_trajectories.pdf",plot=plot,width=10,height=5)
-
-```
-
-## Compute covariates from observations
-
-```{r}
-
-# Add covariates to data -----------
-
-signed_angle <- function(u, v) {
-  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
-  # as in 
-  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
-  u <- matrix(u, ncol = 2)
-  v <- matrix(v, ncol = 2)
-  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
-                                  rows")
-  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
-  ind1 <- which(result > pi)
-  ind2 <- which(result <= -pi)
-  result[ind1] <- result[ind1] - 2*pi
-  result[ind2] <- result[ind2] + 2*pi
-  return(result) 
-}
-
-  # Function to calculate rolling average
-rolling_average <- function(data, window) {
-  n <- nrow(data)
-  result <- matrix(NA, n, ncol(data))  # Initialize result matrix with NA
-  colnames(result) <- colnames(data)
-
-  for (i in seq_len(n)) {
-    # Define the range for the current window
-    start <- max(1, i - floor(window / 2))
-    end <- min(n, i + floor(window / 2))
-    result[i, ] <- colMeans(data[start:end, , drop = FALSE])  # Compute column means
-  }
-  return(result)
-}
-
-add_covs_parallel <- function(data,add="all",window=1,n_cores = parallel::detectCores() - 1) {
-  # Setup parallel backend
-  cl <- makeCluster(n_cores)
-  registerDoParallel(cl)
-
-  # Split data by ID
-  ids <- unique(data$ID)
-
-  # Process each ID in parallel
-  results <- foreach(id = ids, .combine = rbind, .packages = c("sf"),
-                     .export = c("nearest_boundary_points", "is_in_land", "signed_angle","border",
-                                 "D_LOW","D_UP","rolling_average")  ) %dopar% {
-    # Filter data for the current ID
-    sub_data <- data[data$ID == id, ]
-    n_sub <- nrow(sub_data)
-
-    # Time steps
-    dtimes <- sub_data[2:n_sub, "time"] - sub_data[1:(n_sub - 1), "time"]
-
-    # Step lengths
-    dx <- sub_data[2:n_sub, "y1"] - sub_data[1:(n_sub - 1), "y1"]
-    dy <- sub_data[2:n_sub, "y2"] - sub_data[1:(n_sub - 1), "y2"]
-
-    # Empirical velocity
-    vexp_df <- cbind(dx / dtimes, dy / dtimes)
-
-    vexp_avg <- rolling_average(vexp_df, window)
-
-    # Nearest points on shore
-    sub_data <- cbind(sub_data, nearest_boundary_points(as.matrix(sub_data[, c("y1", "y2")]), border))
-
-    # Normal vectors
-    normal <- as.matrix(sub_data[2:n_sub, c("y1", "y2")] - sub_data[2:n_sub, c("p1", "p2")])
-
-    # Angle between velocity and normal vector
-    theta_coast <- signed_angle(normal, vexp_avg)
-    theta_coast <- c(theta_coast, 1)  # Adjust length
-
-    # Initialize DistanceShore
-    Dshore <- rep(0, n_sub)
-
-    for (i in 1:n_sub) {
-      y1 <- sub_data[i, "y1"]
-      y2 <- sub_data[i, "y2"]
-      if (!is_in_land(st_point(c(y1, y2)), border)) {
-        p1 <- sub_data[i, "p1"]
-        p2 <- sub_data[i, "p2"]
-        Dshore[i] <- sqrt((y1 - p1)^2 + (y2 - p2)^2)
-      }
-    }
-
-    Dshore_avg=as.vector(rolling_average(matrix(Dshore,ncol=1),window))
-
-    # Calculate Ishore
-    Ishore_avg <- ifelse(Dshore_avg < D_LOW, 1 / D_LOW, ifelse(Dshore_avg > D_UP, 0, 1 / Dshore_avg))
-
-    # Add columns to sub_data
-    if (add=="all") {
-      sub_data$theta <- theta_coast
-      sub_data$DistanceShore <- Dshore_avg
-      sub_data$Ishore <- Ishore_avg
-    }
-    else if (add=="theta") {
-      sub_data$theta <- theta_coast
-    }
-    else if (add=="DistanceShore") {
-      sub_data$DistanceShore <- Dshore
-      sub_data$Ishore <- Ishore
-        }
-  else {
-    stop("Options for argument add are \"all\",\"DistanceShore\" and \"theta\"")
-  }
-
-    return(sub_data)
-  }
-
-  # Stop the cluster
-  stopCluster(cl)
-
-  return(results)
-}
-
-```
-
-```{r}
-#colnames(data)[8]="DistanceShore"
-data_lf_he=add_covs_parallel(data_lf_he,window=3)
-data_hf_le=add_covs_parallel(data_hf_le,window=3)
-data_lf_le=add_covs_parallel(data_lf_le,window=3)
-data_hf_he=add_covs_parallel(data_hf_he,window=3)
-```
-
-```{r}
-
-plot_Dshore=ggplot(data=data_lf_he)+geom_point(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.5)+
-  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)+facet_wrap(~ID)+
-  theme_minimal()
-
-plot_Dshore
-
-```
-
-```{r}
-
-ggplot(data_lf_he)+geom_histogram(aes(x=DistanceShore),bins=30)+facet_wrap(~ID)+theme_minimal()
-quantile(data_lf_he$DistanceShore)
-```
-
-```{r}
-
-data_lf_he$angular_velocity=as.vector(sapply(unique(data_sub$ID),function(id){
-  c(NA,diff(data_lf_he[data_lf_he$ID==id,"theta"])/diff(data_lf_he[data_lf_he$ID==id,"time"]))}))
-
-plot_angular_velocity=ggplot(data=data_lf_he)+geom_point(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.5)+
-  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)+facet_wrap(~ID)+
-  theme_minimal()
-plot_angular_velocity
-
-```
-
-```{r}
-
-ggplot(data_lf_he)+geom_histogram(aes(x=angular_velocity),bins=30)+facet_wrap(~ID)+theme_minimal()
-quantile(data_lf_he$angular_velocity,na.rm=TRUE)
-
-```
-
-```{r}
-data_lf_he$speed=as.vector(sapply(unique(data_sub$ID),function(id) {c(NA,sqrt(diff(data_lf_he[data_lf_he$ID==id,"y1"])^2+diff(data_lf_he[data_lf_he$ID==id,"y2"])^2)
-                                                                      /diff(data_lf_he[data_lf_he$ID==id,"time"]))}))
-
-ggplot(data_lf_he)+geom_histogram(aes(x=speed),bins=30)+facet_wrap(~ID)+theme_minimal()
-quantile(data_lf_he$speed,na.rm=TRUE)
-```
-```{r}
-ggplot(data_lf_he)+geom_histogram(aes(x=theta),bins=30)+facet_wrap(~ID)+theme_minimal()
-```
-
-## Estimate CTCRW
-
-### low frequency high error
+## low frequency high error
 
@@
 ```
+
 ```{r}
@@
 ```{r fit sde model with splines,cache=TRUE}
-
 formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1,omega=~1)
@@
                     other_data=list("log_sigma_obs0"=log(0.03)))
-
 #fit
@@
 ```{r fit sde model with splines,cache=TRUE}
-
 formulas <- list(mu1=~1,mu2=~1,tau=~1,nu=~1)
@@
 DistanceShore <- runif(n,0.1,2)
-samples=data.frame(theta=theta,DistanceShore=DistanceShore)
+samples=data.frame(BoundaryAngle=theta,BoundaryDistance=DistanceShore)
 
@@
 #fit with bivariate splines te
-m1 <- gam(y~te(theta,DistanceShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
-                                                       DistanceShore=seq(0.02,7,len=SP_DF)))
+m1 <- gam(y~te(theta,DistanceShore,k=SP_DF,bs=c("cs","cs")))
 
-ExpShore=1/DistanceShore
-
-#fit with bivariate splines te
-m2 <- gam(y~te(theta,ExpShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
-                                                  ExpShore=seq(1/8,1/0.2,len=SP_DF)))
-
-fomega_cubic_splines=function(cov_data) {
-
-  if (is.null(cov_data$ExpShore)){
-    omega=predict(m1,newdata=cov_data)
-  }
-  else if (is.null(cov_data$DistanceShore)){
-    omega=predict(m2,newdata=cov_data)
-  }
-
-  return(as.numeric(omega))
-}
+#visualize
+vis.gam(m1);title("tensor product")
 
@@
 formulas <- list(mu1=~1,mu2=~1,tau=~1,
-                 nu=~1,omega=~te(theta,DistanceShore,k=SP_DF,bs=c("cs")))
+                 nu=~1,omega=~te(observed_BoundaryAngle,observed_BoundaryDistance,k=SP_DF,bs=c("cs")))
 
-crcvm_splines_lf_he<- SDE$new(formulas = formulas,data = data_lf_he,type = "RACVM_SSM",
+crcvm_splines_lf_he<- SDE$new(formulas = formulas,data = data_list[['data_lf_he']],type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
@@
 crcvm_splines_lf_he$update_lambda(m1$sp)
+
 #fit
@@
 
-crcvm_splines_lf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
+crcvm_splines_lf_he$get_all_plots(show_CI="pointwise")
 
@@
 
-estimate_parametric_omega=function(model) {
+estimate_parametric_omega=function(model,boundary_distance_name,boundary_angle_name,
+                                   fixed_a=NULL,probs=c(0.5,0.95)) {
 
@@
 
-    Dshore=seq(from=quantile(data$DistanceShore,0.5),to=quantile(data$DistanceShore,0.95),length.out=30)
+    Dshore=seq(from=quantile(data$observed_BoundaryDistance,probs[1]),
+               to=quantile(data$observed_BoundaryDistance,probs[2]),
+               length.out=30)
     theta=seq(from=-pi,to=pi,length.out=30)
     grid<- as.data.frame(expand.grid(Dshore,theta))
-    colnames(grid) <- c("DistanceShore","theta")
+    colnames(grid) <- c(boundary_distance_name,boundary_angle_name)
     grid$ID=1
@@
 
-    fit_optim <- optim(par = c(0.1,2),
+    if (is.null(fixed_a)) {
+
+      fit_optim <- optim(par = log(c(1,0.5,0.1,2,0.5,pi/6)),
                        fn = function(params) {
@@
                         # Compute predicted z values
-                        z_pred <- fomega_cubic(data,A,B,params[1],params[2],SIGMA_D,SIGMA_THETA)
+                        z_pred <- fomega_cubic(data,
+                                         boundary_distance_name,
+                                         boundary_angle_name,
+                                         exp(params[1]),
+                                         exp(params[2]),exp(params[3]),
+                                         exp(params[4]),exp(params[5]),
+                                         exp(params[6]))
 
@@
 
-    est_par=fit_optim$par
+    } 
+    else {
+    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
+                       fn = function(params) {
+
+                        # Compute predicted z values
+                        z_pred <- fomega_cubic(data,
+                                         boundary_distance_name,
+                                         boundary_angle_name,
+                                         fixed_a,exp(params[1]),
+                                         exp(params[2]),exp(params[3]),
+                                         exp(params[4]),exp(params[5]))
+
+                         # Return sum of squared residuals
+                         sum((data$z - z_pred)^2)
+                       })
+
+    }
+    est_par=exp(fit_optim$par)
 
     # Compute the predicted values using the starting parameters
-    z_pred_starting <- fomega_cubic(grid,A,B,est_par[1],est_par[1],SIGMA_D,
-                                    SIGMA_THETA)
+    if (is.null(fixed_a)) {
+
+      z_pred <- fomega_cubic(grid,
+                       boundary_distance_name,
+                      boundary_angle_name,
+                      est_par[1],
+                      est_par[2],est_par[3],
+                      est_par[4],est_par[5],
+                      est_par[6])
+
+    } else {
+    z_pred <- fomega_cubic(grid,boundary_distance_name,
+                      boundary_angle_name,
+                     fixed_a,est_par[1],est_par[2],
+                      est_par[3],est_par[4],est_par[5])
+    }
 
     # Reshape z_pred_starting to a matrix for plotting
-    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta))
+    z_pred_matrix <- matrix(z_pred, nrow = length(Dshore), ncol = length(theta))
 
@@
 
-estimate_parametric_omega(crcvm_splines_lf_he)
+estimate_parametric_omega(crcvm_splines_lf_he,probs=c(0.75,0.95),
+                          boundary_distance_name = "observed_BoundaryDistance",
+                          boundary_angle_name = "observed_BoundaryAngle")
 
@@
 
-crcvm_splines_lf_le<- SDE$new(formulas = formulas,data = data_lf_le[data_lf_le$ID %in% c(1:6),],type = "RACVM_SSM",
+crcvm_splines_lf_le<- SDE$new(formulas = formulas,data = data_list[["data_lf_he"]],type = "RACVM_SSM",
                     response = c("y1","y2"),par0 = PAR0_RACVM,fixpar=c("mu1","mu2"),
-                    other_data=list("log_sigma_obs0"=log(0.05)))
+                    other_data=list("log_sigma_obs0"=log(0.01)))
 
@@
 
-crcvm_splines_lf_le$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
+crcvm_splines_lf_le$get_all_plots(show_CI="pointwise")
 
-exp(as.list(crcvm_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
+exp(as.list(crcvm_splines_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
 
@@
 ```{r}
-
 crcvm_splines_hf_he$get_all_plots(show_CI="pointwise",true_smooths=list("omega"=fomega))
@@
 ```{r}
-
 estimate_parametric_omega(crcvm_splines_hf_he)
-
 ```
@@
 ```
-
-## Estimation with fixed parametric omega function
-
-### Low frequency high error
-
-We will fix the new parameters that are introduced in the model, and estimate the others, that is $\tau$ and $\nu$.
-These new parameters might be chosen based on the observed distance to shore and observed angular velocity in the data.
-
-```{r}
-H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_lf_he$time)),dim=c(2,2,length(data_lf_he$time)))
-# Estimate from simulated data ------------------
-
-formulas <- list(tau=~1,
-                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
-
-crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
-                            data = data_lf_he,type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
-                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("H"=H))
-
-#fit
-crcvm_cubic_lf_he$fit(method="BFGS")
-```
-
-```{r}
-
-crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")
-exp(as.list(crcvm_cubic_lf_he$tmb_rep(),what="Estimate")$log_sigma_obs)
-
-```
-
-We are unable to estimate $\tau$ while fixing $\omega$. $\tau$ is totally biased and underestimated, even more than when we neglected the constraints and used a simple CTCRW..
-
-### Low frequency low error
-
-```{r}
-H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_lf_le$time)),dim=c(2,2,length(data_lf_le$time)))
-# Estimate from simulated data ------------------
-
-formulas <- list(tau=~1,
-                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
-
-crcvm_cubic_lf_le<- SDE$new(formulas = formulas,data = data_lf_le[data_lf_le$ID %in% c(1:4),],
-                              type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("H"=H))
-
-crcvm_cubic_lf_le$fit(method="BFGS")
-```
-```{r}
-
-crcvm_cubic_lf_le$get_all_plots(show_CI="simultaneous")
-exp(as.list(crcvm_cubic_lf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
-
-```
-Let's now try with high frequency data.
-
-### High frequency high error
-
-```{r}
-H=array(rep(SIGMA_OBS_HIGH^2*diag(2),length(data_hf_he$time)),dim=c(2,2,length(data_hf_he$time)))
-
-formulas <- list(tau=~1,
-                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
-
-crcvm_cubic_hf_he<- SDE$new(formulas = formulas,data = data_hf_he[data_hf_he$DistanceShore>0.01 & data_hf_he$ID %in% c(1:6),],type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 =c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
-                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("H"=H))
-
-crcvm_cubic_hf_he$fit(method="BFGS")
-```
-
-```{r}
-crcvm_cubic_hf_he$get_all_plots()
-exp(as.list(crcvm_cubic_hf_he$tmb_rep(),what="Estimate")$log_sigma_obs)
-```
-
-### High frequency low error
-
-```{r}
-
-H=array(rep(SIGMA_OBS_LOW^2*diag(2),length(data_hf_le$time)),dim=c(2,2,length(data_hf_le$time)))
-
-formulas <- list(tau=~1,
-                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
-
-crcvm_cubic_hf_le<- SDE$new(formulas = formulas,data = data_hf_le[data_hf_le$ID %in% c(1:4),],
-                            type = "CRCVM_SSM",
-                    response = c("y1","y2"),par0 = c(1,1,A,B,D0,D1,SIGMA_D,SIGMA_THETA),
-                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
-                    other_data=list("H"=H))
-
-#fit
-crcvm_cubic_hf_le$fit(method="BFGS")
-```
-
-```{r}
-crcvm_cubic_hf_le$get_all_plots()
-exp(as.list(crcvm_cubic_hf_le$tmb_rep(),what="Estimate")$log_sigma_obs)
-```
-
-```{r}
-ggplot()+geom_point(data=data_hf_le,aes(time,theta),col="blue",alpha=0.5,size=0.1)+
-  geom_point(data=data,aes(time,theta),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)
-
-```
-```{r}
-data_hf_le$omega=fomega(data_hf_le[,c("theta","DistanceShore")])
-
-ggplot()+geom_point(data=data_hf_le,aes(time,omega),col="blue",alpha=0.5,size=0.1)+
-  geom_point(data=data,aes(time,omega),col="red",alpha=0.2,size=0.1)+facet_wrap(~ID)
-```

### 2025-02-16 R/simulation_study/cluster/CRCVM/circ/Rscripts/circ_Ishore13.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 13
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/circ/Rscripts/circ_Ishore51.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 51
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/circ/Rscripts/circ_Ishore79.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="circ"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 79
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/fjords/Rscripts/fjords_parametric54.R

@@ -1,0 +1,325 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 54
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/fjords/Rscripts/fjords_parametric88.R

@@ -1,0 +1,325 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 88
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/fjords/Rscripts/fjords_parametric97.R

@@ -1,0 +1,325 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 97
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/rect/Rscripts/rect_Ishore24.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 24
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/rect/Rscripts/rect_Ishore27.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 27
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/rect/Rscripts/rect_Ishore29.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 29
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-02-16 R/simulation_study/cluster/CRCVM/rect/Rscripts/rect_Ishore9.R

@@ -1,0 +1,565 @@
+# SETUP --------------------------------------------------------------------
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+domain_name="rect"
+par_dir=here("R","simulation_study_CRCVM","spline_estimation_cluster",domain_name)
+set_up_file=paste("set_up_",domain_name,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 9
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                      log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
[…]

### 2025-03-17 R/application/CRCVM/fit_crcvm.Rmd
update application to narwhal data

@@
 ---
-To circumvent this issue, we can fix a smooth parametric function for $\omega$, based on the observed distances to shore and the observed angular velocities.
+
+We fit the CRCVM defined in the paper on the narwhal data.
 
@@
 ```{r packages,message=FALSE}
-set.seed(42)
 
@@
 library(sf)
-library(gganimate) # animated trajectories
+library(gganimate)
+library(doParallel)
+library(foreach)
+library(ggpubr)
+
 ```
 
-```{r get data,message=FALSE}
+# Read data
+
+```{r get data,message=FALSE,}
 
@@
 par_dir=here() #parent directory
+
 narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")
 
-#data before exposure with first 24h removed
-dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")
-
 #data before exposure with first 12h removed
-dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")
+dataBE12=read.csv(file.path(narwhal_data_path,"DataBE12.csv"), header = TRUE,dec = ".")
 
@@
 
-```{r}
+```{r get border,message=FALSE}
+# Set the path to the directory containing the greenland data
+greenland_data_path <- here("Data","preprocessed_data","greenland")
 
-N=length(unique(dataBE12$ID))
-
-n_pre12=length(dataBE12$time) #number of observations before exposure
-n_pre24=length(dataBE24$time) #number of observations before exposure
-n_post=length(dataAE$time) #number of observations after exposure
-```
-
-```{r}
-
-plot_Dshore=ggplot(data=dataBE12)+geom_point(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.5)+
-  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)+
-  theme_minimal()
-
-ggsave("observed_Dshore.pdf",plot=plot_Dshore)
-
-quantile(dataBE12$DistanceShore,probs=0.05)
-
-```
-
-```{r}
-dataBE12$angular_velocity=c(NA,diff(dataBE12$AngleNormal)/diff(dataBE12$time))
-plot_angular_velocity=ggplot(data=dataBE12)+geom_point(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.5)+
-  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)+
-  theme_minimal()
-ggsave("observed_angular_velocity.pdf",plot=plot_angular_velocity)
-max(dataBE12$angular_velocity,na.rm=TRUE)/(pi*(pi-pi/2)*(pi+pi/2)*exp(-1))
+border200<-st_read(file.path(greenland_data_path,"shrunk200_scoresby_sound_utm.shp"))
+border200 <- st_transform(border200, crs = "+init=EPSG:32626 +units=km")
 
@@
 
-renamed_dataBE12=dataBE12
-colnames(renamed_dataBE12)[48]="theta"
-renamed_dataBE12[renamed_dataBE12$DistanceShore==0,]=0.02
+N=length(unique(dataBE12$ID)) #number of ID
+n_pre=nrow(dataBE12) #number of observations before exposure
+n_post=nrow(dataAE) #number of observations after exposure
+
+```
+
+# Utility functions
+
+```{r define function to plot interpolation}
+
+#' Plot interpolated trajectories, angles and distances
+#' @param data Dataframe with the columns time, ID, BoundaryDistance, and BoundaryAngle
+#' @param interpolation_data 
+#' @param border Boundary of the domain as a sf object with geometry 
+#' @param response name of response variables
+#' @return List of plots
+get_interpolation_plots<-function(data,interpolation_data,border,response=c("x","y")) {
+
+  #create dataframe with interpolated values
+  end_indices <- c(which(diff(as.numeric(factor(data$ID))) != 0),nrow(data))
+  n_step<-ncol(interpolation_data$BoundaryDistance)
+  interpolated_x_vec=as.vector(t(interpolation_data[[response[1]]][-end_indices,]))
+  interpolated_y_vec=as.vector(t(interpolation_data[[response[2]]][-end_indices,]))
+  interpolated_distance_vec=as.vector(t(interpolation_data$BoundaryDistance[-end_indices,]))
+  interpolated_angle_vec=as.vector(t(interpolation_data$BoundaryAngle[-end_indices,]))
+  interpolation_time=as.vector(t(interpolation_data$time[-end_indices,]))
+  ID=unlist(lapply(unique(data$ID), function(id) {
+    n=nrow(data[data$ID==id,])
+    rep(id,(n-1)*n_step)
+  }))
+
+  #combine dataframes
+  interpolated_df=data.frame("time"=interpolation_time,"ID"=ID,"x"=interpolated_x_vec,
+                             "y"=interpolated_y_vec,
+                             "BoundaryDistance"=interpolated_distance_vec,
+                             "BoundaryAngle"=interpolated_angle_vec)
+  interpolated_df$type="Interpolated"
+  data$type="Observed"
+  colnames(data)[colnames(data)==response[1]]<-"x"
+  colnames(data)[colnames(data)==response[2]]<-"y"
+
+  combined_data<-rbind(interpolated_df,data)
+
+  # Create plot with legends for shape and linetype
+  trajectory_plot = ggplot() +geom_sf(data=border$geometry,fill="grey")+
+    geom_point(data = combined_data, 
+               aes(x,y, col = ID, shape = type), size = 0.2,alpha=1) +
+
+    # Customize the legend
+    labs(color = "Trajectory ID", 
+         shape = "Point Type", 
+         linetype = "Line Type", 
+         title = "Interpolated vs. Observed trajectories") +
+    scale_shape_manual(values = c(10:12))+
+    theme_minimal() +
+    theme(legend.position = "right")  # Adjust legend position
+
+  distance_plot = ggplot() +
+    # Scatter plot for observed BoundaryDistance (Shape Legend)
+    geom_line(data = combined_data, 
+               aes(time, BoundaryDistance, col = ID, linetype = type), size = 0.5) +
+
+    # Customize the legend
+    labs(color = "Trajectory ID", 
+         shape = "Point Type", 
+         linetype = "Line Type", 
+         x = "Time", y = "Boundary Distance", 
+         title = "Interpolated vs. True Boundary Distance") +
+
+    theme_minimal() +
+    theme(legend.position = "right")  # Adjust legend position
+
+  # Create plot with legends for shape and linetype
+  angle_plot = ggplot() +
+    # Scatter plot for observed BoundaryDistance (Shape Legend)
+    geom_line(data = combined_data, 
+               aes(time, BoundaryAngle, col = ID, linetype = type), size = 0.5) +
+
+    # Customize the legend
+    labs(color = "Trajectory ID", 
+         shape = "Point Type", 
+         linetype = "Line Type", 
+         x = "Time", y = "Boundary Angle", 
+         title = "Interpolated vs. True Boundary Angle") +
+
+    theme_minimal() +
+    theme(legend.position = "right")  # Adjust legend position
+
+  return(list("angle_plot"=angle_plot,"distance_plot"=distance_plot,"trajectory_plot"=trajectory_plot))
+}
+
+```
+
+```{r define function to fit the baseline model}
+
+#' Fit baseline on data
+#' @param data data before exposure 
+#' @param interpolation_data 
+#' @param sigma_obs GPS measurement error std
+#' @param a,b,D0,D1,sigma_D,sigma_theta parameters in the function omega
+#' @param fixpar parameters to fix
+#' @param response name of response variables
+#' @return Fitted baseline SDE object
+fit_baseline<-function(data,interpolation_data,sigma_obs=0.05,a=2,b=1,D0=0.3,D1=0.5,
+                       sigma_D=1,sigma_theta=0.6,fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"),
+                       response=c("x","y")) {
 
   #Fixed measurement error
-sigma_obs=0.03
-H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))
+  H=array(rep(sigma_obs^2*diag(2),length(data$time)),dim=c(2,2,length(data$time)))
+
+  interpolated_dist<- interpolation_data$BoundaryDistance
+  interpolated_dist[interpolated_dist == 0] <- 0.1
 
@@
   # Fit baseline
-baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
-                    response = c("x","y"),par0 = c(1,4,0.1,4,0.2,0.5,0.1,pi/6),other_data=list("H"=H),
-                    fixpar=c("a","b","D0","D1","sigma_D","sigma_theta"))
+  baseline <- SDE$new(formulas = formulas,
+                              data = data,type = "CRCVM_SSM",
+                              response = response,par0 = c(2,1,a,b,D0,D1,sigma_D,sigma_theta),
+                              fixpar=fixpar,
+                              other_data=
+                                list("H"=H,"interpolated_distance"=interpolated_dist,
+                                     "interpolated_angle"=interpolation_data$BoundaryAngle))
+  baseline$update_lambda(c(1,1))
+  baseline$fit()
 
-baseline_crcvm_cubic$fit()
+  return (baseline)
+}
+
+```
+
+```{r define function to get the estimates of the fitted model}
+
+#' Get estimated from fitted baseline SDE model
+#' @param sde fitted sde
+#' @param include_sigma_obs boolean
+#' @param n_post number of draws from the posterior distribution used to give estimates
+#' and confidence intervals
+#' @param resp Logical (default: TRUE). Should the output be on 
+#' the response scale? If FALSE, the output is on the linear 
+#' predictor scale
+#' @return List of estimates, standard error and quantiles for each parameter
+get_baseline_estimates<- function(sde,include_sigma_obs=FALSE,n_post=100,
+                                  resp=FALSE) {
+
+  post_coeff=sde$post_coeff(n_post=n_post)
+  link=ifelse(resp,exp,identity)
+  post_par=list(
+  "tau"=link(post_coeff$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=link(post_coeff$coeff_fe[,"nu.(Intercept)"]),
+  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
+  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))
+
+  if (include_sigma_obs) {
+    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
+  }
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
+                      quantile, probs = c(0.025, 0.975))
+
+  return(list("estimates"=mean_par,"std.error"=sd_par,"quantiles"=quant_par))
+
+}
+
+#' Make latex table to show the estimates and confidence intervals
+#' @param sde fitted sde
+#' @param include_sigma_obs boolean
+#' @param n_post number of draws from the posterior distribution used to give estimates
+#' and confidence intervals
+#' @return NULL (just print latex table)
+make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,n_post=100) {
+
+  est=get_baseline_estimates(sde,include_sigma_obs,n_post)
+  mean_par=est$estimates
+  sd_par=est$std.error
+  quant_par=est$quantiles
+  parameter_names <- names(mean_par)
+  estimates <- mean_par
+  conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
+          quant_par[[param]][2])
+    })
+
+  parameter_names=c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$")
+  if (include_sigma_obs) {
+    parameter_names=c(parameter_names,"$\\sigma_{\\obs}$")
+  }
+  table_data <- data.frame(
+    Parameter = parameter_names,
+  Estimate = sprintf("$%.2f$", mean_par),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE)
+
+  xtab <- xtable(
+  table_data,
+  caption = paste(sde$type(),"baseline estimations with random effects"),
+  label = paste0("table:",sde$type(),"_baseline_estimations_with_random_effects"))
+
+  print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
+  hline.after = c(-1, 0, nrow(table_data)),  
+  table.placement="H")
+
+}
+
+```
+
+```{r define functions to compare simulation and observations}
+
+#' Make plot to compare BoundaryDistance in simulation and observation
+#' @param simulated_data list of data frames simulated from an SDE model, each containing a column "BoundaryDistance".
+#' @param observed_data real GPS data (data frame).
+#' @return ggplot of boundary distances distributions
+make_distance_density_plot <- function(simulated_data, observed_data) {
+
+  # Ensure simulated_data is a list of data frames
+  if (!is.list(simulated_data) || !all(sapply(simulated_data, is.data.frame))) {
+    stop("simulated_data must be a list of data frames")
+  }
+
+  simulated_combined <- bind_rows(
+    lapply(seq_along(simulated_data), function(i) {
+      simulated_data[[i]] %>% mutate(SimulationID = as.factor(i), Source = "Simulated")
+    })
+  )
+
+  observed_combined <- observed_data %>% mutate(Source = "Observed")
+
+  combined_data <- bind_rows(simulated_combined, observed_combined)
+
+  # Plot density distribution
+  plot_combined <- ggplot(combined_data, aes(x = BoundaryDistance, color = Source, fill = Source)) +
+    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
+    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
+    xlab("Distance to Shore") +
+    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
+    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
+    theme_minimal() +
+    theme(legend.title = element_blank())
+
+  return(plot_combined)
+}
+
+#' Make plot to compare BoundaryAngle in simulation and observation
+#' @param simulated_data data simulated from a sde model containing a column "BoundaryAngle".
+#' @param observed real GPS data
+#' @return ggplot of boundary angle distributions
+make_angle_density_plot <- function(simulated_data, observed_data) {
+
+  # Ensure simulated_data is a list of data frames
+  if (!is.list(simulated_data) || !all(sapply(simulated_data, is.data.frame))) {
+    stop("simulated_data must be a list of data frames")
+  }
+
+  simulated_combined <- bind_rows(
+    lapply(seq_along(simulated_data), function(i) {
+      simulated_data[[i]] %>% mutate(SimulationID = as.factor(i), Source = "Simulated")
+    })
+  )
+
+  observed_combined <- observed_data %>% mutate(Source = "Observed")
+
+  simulated_combined$BoundaryAngle<-abs(simulated_combined$BoundaryAngle)
+  observed_combined$BoundaryAngle<-abs(observed_combined$BoundaryAngle)
+  combined_data <- bind_rows(simulated_combined, observed_combined)
+
+  # Plot density distribution
+  plot_combined <- ggplot(combined_data, aes(x = BoundaryAngle, color = Source, fill = Source)) +
+    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
+    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
+    xlab(expression(Theta)) +
+    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
+    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
+    theme_minimal() +
+    theme(legend.title = element_blank())
+
+  return(plot_combined)
+
+}
+
+#' Compute empirical speed from finite differences
+#' @param data data with column time
+#' @param response name of response variables
+#' @return vector of empirical speeds
+compute_speed<-function(data,response=c("x","y")) {
+
+  speed_vector<-rep(0,nrow(data))
+
+  for (id in unique(data$ID)){
+
+    #filter data with specific id
+    sub_ind=(data$ID==id)
+    sub_data=data[sub_ind,]
+    n_sub=length(sub_data$time)
+
+    #time steps
+    dtimes=sub_data[2:n_sub,"time"]-sub_data[1:(n_sub-1),"time"]
+
+    #step lengths
+    dx=sub_data[2:n_sub,response[1]]-sub_data[1:(n_sub-1),response[1]]
+    dy=sub_data[2:n_sub,response[2]]-sub_data[1:(n_sub-1),response[2]]
+
+    speed=c(NA,sqrt((dx/dtimes)^2+(dy/dtimes)^2))
+
+    speed_vector[sub_ind]<-speed
+  }
+
+  return(speed_vector)
+}
+
+#' Make plot to compare speed in simulation and observation
+#' @param simulated_data data simulated from a sde model containing a column "speed".
+#' @param observed real GPS data
+#' @return ggplot of boundary angle distributions
+make_speed_density_plot <- function(simulated_data, observed_data,response) {
+
+    # Ensure simulated_data is a list of data frames
+  if (!is.list(simulated_data) || !all(sapply(simulated_data, is.data.frame))) {
+    stop("simulated_data must be a list of data frames")
+  }
+
+  simulated_combined <- bind_rows(
+    lapply(seq_along(simulated_data), function(i) {
+      simulated_data[[i]] %>% mutate(SimulationID = as.factor(i), Source = "Simulated")
+    })
+  )
+
+  observed_combined <- observed_data %>% mutate(Source = "Observed")
+
+  combined_data <- bind_rows(simulated_combined, observed_combined)
+
+  # Plot density distribution
+  plot_combined <- ggplot(combined_data, aes(x = speed, color = Source, fill = Source)) +
+    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
+    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
+    xlab("Speed") +
+    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
+    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
+    theme_minimal() +
+    theme(legend.title = element_blank())
+
+  return(plot_combined)
+}
+```
+
+# Fit baseline on data before exposure
+
+```{r interpolate positions befor exposure}
+
+#prepare data for interpolation
+dataBE12_prep200=dataBE12[,c("ID","x","y","time","theta200","DistanceShore200")]
+
+colnames(dataBE12_prep200)[colnames(dataBE12_prep200)=="DistanceShore200"]<-"BoundaryDistance"
+colnames(dataBE12_prep200)[colnames(dataBE12_prep200)=="theta200"]<-"BoundaryAngle"
+
+#remove points on land
+dataBE12_prep200<-dataBE12_prep200[dataBE12_prep200$BoundaryDistance>0,]
+
+n_step=15
+
+#compute spline interpolation
+interpolation_bas_200<-interpolate_BoundaryMetrics(dataBE12_prep200,response=c("x","y"),
+                                                border200,n_step=n_step,n_cores=6,df_ratio=0.95)
+```
+
+```{r}
+interpolation_plots<-get_interpolation_plots(dataBE12_prep200,interpolation_bas_200,border200)
+
+interpolation_plots
+```
+
+```{r}
+D0=quantile(dataBE12_prep200$BoundaryDistance,0.8)
+D1=quantile(dataBE12_prep200$BoundaryDistance,0.5)
+a=0.3*D0
+
+baseline_crcvm200<- fit_baseline(dataBE12_prep200,interpolation_bas_200,
+                    b=0.5,a=a,sigma_D=1,D0=D0,D1=D1,sigma_theta=pi/6,
+                    sigma_obs=0.05)
 
@@
 ```{r}
-plots_bas_crcvm=baseline_crcvm_cubic$get_all_plots(baseline=NULL,show_CI="simultaneous")
+baseline_crcvm200$get_all_plots(baseline=NULL,show_CI="simultaneous")
+```
 
-plots_bas_crcvm
+```{r}
+make_final_baseline_tab(baseline_crcvm200,include_sigma_obs=FALSE,n_post=10000)
 ```
+
+# Check baseline models
+
+## Simulate trajectories
+
+```{r set up for simulation}
+
+#initial narwhals positions (convert to matrix to avoid bugs)
+z0_BE <- as.matrix(dataBE12_prep200[!duplicated(dataBE12_prep200$ID), c("x", "y")])
+
+# Define functions to compute covariates along the way
+fangle=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #return angle
+  return (signed_angle(normal,v))
+}
+
+fDshore=function(z,v,p) {
+
+  #normal vector
+  normal=c(z[1]-p[1],z[2]-p[2])
+
+  #distance to shore
+  Dshore=sqrt(normal[1]^2+normal[2]^2)
+
+  #fix a threshold to avoid error when we hit the boundary
+  if (Dshore>0.005) {
+      return(Dshore)
+  }
+  else {
+    return(0.005)
+  }
+
+}
+
+atw=list("BoundaryAngle"=fangle,"BoundaryDistance"=fDshore)
+
+# Define time steps
+delta=1/60/6
+times_list=lapply(unique(dataBE12_prep200$ID),function(ID) {
+  min=min(dataBE12[dataBE12_prep200$ID==ID,"time"])
+  max=max(dataBE12[dataBE12_prep200$ID==ID,"time"])
+  seq(min,max,by=delta)})
+names(times_list)=unique(dataBE12_prep200$ID)
+times_df <- data.frame(
+  ID = rep(names(times_list), sapply(times_list, length)),
+  time = unlist(times_list),row.names = NULL
+)
+n=length(times_df$time)
+# Define covariates data (needed in the function simulate)
+data_reg=data.frame("BoundaryDistance"=rep(1,n),
+                    "BoundaryAngle"=rep(0,n))
+
+data_reg=cbind(data_reg,times_df)
+
+```
+
+```{r simulate trajectory from baseline model}
+all_sim=list()
+n_sim=4
+
+for (i in 1:n_sim) {
+  set.seed(i)
+  baseline_crcvm200_trajectory=baseline_crcvm200$simulate(z0=z0_BE,
+                     data=data_reg,atw=atw,land=border200,verbose=FALSE,n_cores=6)
+  colnames(baseline_crcvm200_trajectory)[colnames(baseline_crcvm200_trajectory)=="z1"]<-"x"
+  colnames(baseline_crcvm200_trajectory)[colnames(baseline_crcvm200_trajectory)=="z2"]<-"y"
+
+  all_sim[[i]]<-baseline_crcvm200_trajectory
+
+}
+
+```
+
+```{r subsample simulations}
+#subsample to 5 minutes and add measurement noise
+all_sim_sub<-lapply(all_sim,function(data) {
+
+  data_sub<-data[seq(1,nrow(data),by=60),]
+  noise<-rmvn(nrow(data_sub),rep(0,2),diag(rep(0.05^2,2)))
+
+  data_sub[,c("x","y")]=data_sub[,c("x","y")]+noise
+  return (data_sub)
+})
+
+```
+
+```{r plot one simulated trajectory}
+
+plot_sim_trajectory=ggplot() +
+  geom_sf(data=border200$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=all_sim_sub[[2]], 
+            aes(x = x, y = y,color=ID),size = 0.2, 
+            lineend = "round",alpha=0.5) +     
+  geom_point(data=all_sim_sub[[2]],
+             aes(x = x, y = y,color=ID),size =0.2 ) +
+  theme_minimal()+
+  scale_color_viridis_d(name="ID",labels=c("A1","A2","A3","A4","A5","A6"))+theme(
+    axis.title = element_text(size = 14),
+    axis.text = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank())
+
+plot_obs_trajectory=ggplot() +
+  geom_sf(data=border200$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=dataBE12, 
+            aes(x = x, y = y,color=ID),size = 0.2, 
+            lineend = "round",alpha=0.5) +     
+  geom_point(data=dataBE12,
+             aes(x = x, y = y,color=ID),size =0.2 ) +
+  theme_minimal()+
+  scale_color_viridis_d(name="ID",labels=c("A1","A2","A3","A4","A5","A6"))+theme(
+    axis.title = element_text(size = 14),
+    axis.text = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank())
+
+final_plot=ggarrange(plotlist=list(plot_obs_trajectory,plot_sim_trajectory),
+                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom",
+                     labels= c("Observed","Simulated"), vjust=2)
+
+ggsave(plot = final_plot, "sim_crcvm_trajectory.pdf", 
+       width = 10, height = 6, units = "in", limitsize = FALSE)
+```
+
+```{r animate observed vs simulated trajectories}
+dataBE12_prep200$type <- "Observed"
+all_sim_sub[[1]]$type <- "Simulated"
+
+combined_data <- rbind(dataBE12_prep200[,c("BoundaryDistance","BoundaryAngle","ID","time","x","y","type")], all_sim_sub[[1]])
+
+combined_plot <- ggplot() +
+  geom_sf(data=border200$geometry, fill="grey") +
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km")) +
+  geom_path(data=combined_data, aes(x = x, y = y, color=ID, group=interaction(ID, type)), 
+            size = 0.5, lineend = "round", alpha=0.5) +     
+  geom_point(data=combined_data, aes(x = x, y = y, color=ID, shape=ID,group=interaction(ID, type)), 
+             size = 0.5) +
+  facet_wrap(~type) + 
+  theme_minimal() +
+  scale_color_viridis_d() +
+  theme(axis.title = element_text(size = 18),
+        axis.text = element_text(size = 14),
+        legend.text = element_text(size = 18),
+        axis.title.y = element_blank(),
+        axis.title.x = element_blank()) +
+  transition_reveal(time)
+
+animate(combined_plot, fps = 10, width = 800, height = 400)
+anim_save("sim_crcvm_trajectory.gif")
+
+```
+
+```{r}
+all_boundary_data<- lapply(all_sim_sub,function(data) {
+  get_BoundaryMetrics(data,
+                      response=c("x","y"),
+                      border=border200)
+})
+
+for (i in seq_along(all_sim_sub)) {
+
+  all_sim_sub[[i]][,c("BoundaryDistance","BoundaryAngle")]<- all_boundary_data[[i]]
+  all_sim_sub[[i]]$speed<- compute_speed(data=all_sim_sub[[i]])
+}
+
+dataBE12_prep200$speed<-compute_speed(dataBE12_prep200)
+
+plot_distance<-make_distance_density_plot(all_sim_sub,dataBE12_prep200) 
+plot_angle<-make_angle_density_plot(all_sim_sub,dataBE12_prep200)
+plot_speed<-make_speed_density_plot(all_sim_sub,dataBE12_prep200)
+
+final_densities_plots<-ggarrange(plotlist=list(plot_distance,plot_angle),
+                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom", vjust=2)
+ggsave("BoundaryDistance_density.pdf",path="baseline_crcvm_200",plot=plot_distance)
+ggsave("BoundaryAngle_density.pdf",path="baseline_crcvm_200",plot=plot_angle)
+ggsave("speed_density.pdf",path="baseline_crcvm_200",plot=plot_speed)
+```
+
+# Fit response
+
+```{r}
+
+#prepare data for interpolation
+dataAE_prep200=dataAE[,c("ID","x","y","time","theta200","DistanceShore200","ExpShip")]
+
+colnames(dataAE_prep200)[colnames(dataAE_prep100)=="DistanceShore200"]<-"BoundaryDistance"
+colnames(dataAE_prep200)[colnames(dataAE_prep100)=="theta200"]<-"BoundaryAngle"
+
+#remove points on land
+dataAE_prep200<-dataAE_prep200[dataAE_prep200$BoundaryDistance>0,]
+
+n_step=15
+
+#compute spline interpolation
+interpolation_resp_200<-interpolate_BoundaryMetrics(dataAE_prep200,response=c("x","y"),
+                                                border200,n_step=n_step,n_cores=6,df_ratio=0.95)
+
+```
+
+```{r}
+
+fit_response<-function(baseline,data,interpolation_data,sigma_obs=0.05)  {
+
+  #Fixed measurement error
+  H=array(rep(sigma_obs^2*diag(2),nrow(data)),dim=c(2,2,nrow(data)))
+
+  interpolated_distance<- interpolation_data$BoundaryDistance
+  interpolated_distance[interpolated_distance == 0] <- 0.1
+
+  #parameters formulas
+  formulas <- list(tau =~ExpShip+s(ID,bs="re"),nu=~ExpShip+s(ID,bs="re"),
+                     a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
+
+  # Create SDE object
+  response_crcvm<- SDE$new(formulas = formulas,
+                      data = data,type = "CRCVM_SSM",
+                      response = c("x","y"),par0 = baseline$par(),
+                      fixpar=c("a","b","D0","D1","sigma_theta","sigma_D"),
+                      other_data=
+                        list("H"=H,"interpolated_distance"=interpolated_distance,
+                        "interpolated_angle"=interpolation_data$BoundaryAngle))
+
+  new_map=list(coeff_re=factor(rep(NA,12)),coeff_fe=factor(c(NA,1,NA,2,rep(NA,6))),
+                               log_lambda=factor(c(NA,NA)))
+
+  new_coeff_fe=c(baseline$coeff_fe()["tau.(Intercept)",1],0,baseline$coeff_fe()["nu.(Intercept)",1],0,
+                 baseline$coeff_fe()[3:8,1])
+  new_coeff_re=baseline$coeff_re()
+  new_lambda=c(baseline$lambda())
+
+  response_crcvm$update_map(new_map)
+  response_crcvm$update_coeff_re(new_coeff_re)
+  response_crcvm$update_coeff_fe(new_coeff_fe)
+  response_crcvm$update_lambda(new_lambda)
+
+  response_crcvm$fit()
+
+  return (response_crcvm)
+}
+
+```
+
+```{r}
+response_crcvm<-fit_response(baseline_crcvm200,dataAE_prep200,interpolation_resp_200)
+
+```
+
+```{r}
+
+#plot parameters
+xmin=list("ExpShip"=1/60)
+xmax=list("ExpShip"=1/5)
+link=list("ExpShip"=(\(x) 1/x))
+xlabel=list("ExpShip"="Distance to ship")
+
+plots_resp_crcvm200=response_crcvm100$get_all_plots(baseline=baseline_crcvm200,show_CI="pointwise",xmin=xmin,xmax=xmax,link=link,xlabel=xlabel)
+
+plots_resp_crcvm200
+```
+
+# Propagate 200m shrunk border baseline uncertainty
+
+```{r, message=FALSE,results="hide"}
+
+propagate_baseline_uncertainty<-function(baseline,data,interpolation_data,n_sim=100,sigma_obs=0.05) {
+
+  #posterior samples of the coefficient of the baseline model
+  post_coeff_bas=baseline$post_coeff(n_post=n_sim)
+  post_coeff_re=post_coeff_bas$coeff_re
+  post_coeff_fe=post_coeff_bas$coeff_fe
+  post_coeff_lambda=exp(post_coeff_bas$log_lambda)
+
+  #define model formulas
+  formulas <- list(tau =~ExpShip+s(ID,bs="re"),
+                   nu=~ExpShip+s(ID,bs="re"),a=~1,b=~1,D0=~1,
+                   D1=~1,sigma_D=~1,sigma_theta=~1)
+
+  #Fixed measurement error
+  H=array(rep(sigma_obs^2*diag(2),nrow(data)),dim=c(2,2,nrow(data)))
+
+  interpolated_distance<- interpolation_data$BoundaryDistance
+  interpolated_distance[interpolated_distance == 0] <- 0.1
+
+  # vector to store estimates and std
+  alpha_estimates=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+  alpha_std=matrix(rep(0,2*nrow(post_coeff_fe)),ncol=2)
+
+  #list to store models
+  models_list=list()
+
+  for (i in 1:nrow(post_coeff_fe)) {
+
+      response_crcvm_temp<- SDE$new(formulas = formulas,
+                          data = data,type = "CRCVM_SSM",
+                          response = c("x","y"),par0 = baseline$par(),
+                          fixpar=c("a","b","D0","D1","sigma_theta","sigma_D"),
+                          other_data=
+                            list("H"=H,"interpolated_distance"=interpolated_distance,
+                            "interpolated_angle"=interpolation_data$BoundaryAngle))
+
+      new_map=list(coeff_re=factor(rep(NA,12)),coeff_fe=factor(c(NA,1,NA,2,rep(NA,6))),
+                                   log_lambda=factor(c(NA,NA)))
+
+      new_coeff_fe=c(post_coeff_fe[i,"tau.(Intercept)"],0,post_coeff_fe[i,"nu.(Intercept)"],0,
+                     post_coeff_fe[i,3:8])
+      new_coeff_re=post_coeff_re[i,]
+      new_lambda=post_coeff_lambda[i,]
+
+      response_crcvm_temp$update_map(new_map)
+      response_crcvm_temp$update_coeff_re(new_coeff_re)
+      response_crcvm_temp$update_coeff_fe(new_coeff_fe)
+      response_crcvm_temp$update_lambda(new_lambda)
+
+      try(response_crcvm_temp$fit())
+
+      est_alpha=as.list(response_crcvm_temp$tmb_rep(),what="Est")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+      std_alpha=as.list(response_crcvm_temp$tmb_rep(),what="Std")$coeff_fe[c("tau.ExpShip","nu.ExpShip"),1]
+
+      alpha_estimates[i,]=est_alpha
+      alpha_std[i,]=std_alpha
+
+      models_list[[i]]=response_crcvm_temp
+  }
+
+  colnames(alpha_estimates)=c("tau.ExpShip","nu.ExpShip")
+  colnames(alpha_std)=c("tau.ExpShip","nu.ExpShip")
+
+  return(list("models"=models_list,"estimates"=alpha_estimates,"std"=alpha_std))
+}
+
+results_200_batch1<-propagate_baseline_uncertainty(baseline_crcvm200,dataAE_prep200,
+                                        interpolation_resp_200,n_sim=50,sigma_obs=0.05)
+
+```
+
+```{r propagate uncertainty coefficient log -linear model}
+alpha_estimates<- results_200_batch1$estimates
+
+index<- rowSums(alpha_estimates == 0) < 2
+alpha_estimates_filtered<-alpha_estimates[index, ]
+mean_alpha=list(alpha_tau=mean(alpha_estimates_filtered[,1]),
+                alpha_nu=mean(alpha_estimates_filtered[,2]))
+quant_alpha<- list(
+  alpha_tau=quantile(alpha_estimates_filtered[,1],probs = c(0.025, 0.975)),
+  alpha_nu=quantile(alpha_estimates_filtered[,2],probs = c(0.025, 0.975)))
+
+```
+
+```{r propagated CI alpha log linear coefficients, message=FALSE}
+
+parameter_names <- names(mean_alpha)
+estimates <- mean_alpha
+conf_intervals <- sapply(parameter_names, function(param) {
+  sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1], 
+          quant_alpha[[param]][2])
+})
+
+table_data <- data.frame(
+  Parameter = 
+    c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
+  Estimate = sprintf("$%.2f$", mean_alpha),
+  CI = conf_intervals,
+  stringsAsFactors = FALSE
+)
+
+```
+
+```{r,results='asis',echo=FALSE}
+xtab <- xtable(
+  table_data,
+  caption = "Response log linear estimations with propagated uncertainty",
+  label = "table:response_log_linear_estimations_with_propagated_uncertainty"
+)
+
+print(
+  xtab,type="latex",
+  include.rownames = FALSE,
+  sanitize.text.function = identity,  
+  hline.after = c(-1, 0, nrow(table_data)),
+  table.placement="H"
+
+)
+```
+
+```{r all CI plots,cache=TRUE}
+models_list_filtered<- results_200_batch1$models[index]
+CI_plots=lapply(models_list_filtered,function(model) { 
+  model$get_all_plots(baseline=NULL,xmin=xmin,xmax=xmax,
+                      link=link,xlabel=xlabel,show_CI="pointwise")})
+
+```
+
+```{r plot smooth parameters with propagated uncertainty,results='hide'}
+
+# Extract data for each estimated smooth parameter
+tau_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+})
+nu_CI_data <- lapply(CI_plots, function(plot) {
+  ggplot_build(plot$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+})
+
+# Get confidence intervals as quantiles of estimated parameters
+
+# Combine the x columns from all data frames into a matrix
+x_matrix_tau <- do.call(cbind, lapply(tau_CI_data, function(df) df$y))
+x_matrix_nu <- do.call(cbind, lapply(nu_CI_data, function(df) df$y))
+# Compute the 5% and 95% quantiles row-wise
+quantiles_tau <- apply(x_matrix_tau, 1, 
+                       function(row) quantile(row, probs = c(0.025, 0.975)))
+quantiles_nu <- apply(x_matrix_nu, 1,
+                      function(row) quantile(row, probs = c(0.025, 0.975)))
+
+# Transpose and format the results into a data frame
+quantiles_tau <- as.data.frame(t(quantiles_tau))
+colnames(quantiles_tau) <- c("low", "up")
+quantiles_nu <- as.data.frame(t(quantiles_nu))
+colnames(quantiles_nu) <- c("low", "up")
+
+# Add the "true" estimated smooth
+plots=response_crcvm200$get_all_plots(
+  baseline=baseline_crcvm200,xmin=xmin,xmax=xmax,
+  link=link,xlabel=xlabel,show_CI="pointwise")
+est_tau=ggplot_build(plots$fe_tau_ExpShip)$data[[1]][,c("x","y")]
+est_nu=ggplot_build(plots$fe_nu_ExpShip)$data[[1]][,c("x","y")]
+est_tau=cbind(est_tau,quantiles_tau)
+est_nu=cbind(est_nu,quantiles_nu)
+
+est_tau$baseline=ggplot_build(plots$fe_tau_ExpShip)$data[[2]][,"y"]
+est_nu$baseline=ggplot_build(plots$fe_nu_ExpShip)$data[[2]][,"y"]
+
+plot_tau_response_crcvm200<- ggplot(data = est_tau) +
+  geom_line(aes(x = x, y = y), color = "steelblue", size = 1) +
+  geom_line(aes(x=x,y=baseline),color="steelblue",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "lightblue", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = expression(tau)) +
+  theme_minimal() +theme(
+    axis.title.x = element_text(size=22),
+    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=18),
+    axis.text.y = element_text(size=18),
+    legend.title = element_text(size=18),
+    legend.text = element_text(size=18)
+  )
+
+# Enhanced plot for nu
+plot_nu_response_crcvm200 <- ggplot(data = est_nu) +
+  geom_line(aes(x = x, y = y), color = "darkorange", size = 1) +
+   geom_line(aes(x,baseline),color="darkorange",size=1,linetype="dashed")+
+  geom_ribbon(aes(x = x, ymin = low, ymax = up), 
+              fill = "navajowhite", alpha = 0.3) +
+  labs(
+    x = "Distance to Ship",
+    y = expression(nu)) +
+  theme_minimal() +theme(
+    axis.title.x = element_text(size=22),
+    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
+    axis.text.x = element_text(size=18),
+    axis.text.y = element_text(size=18),
+    legend.title = element_text(size=18),
+    legend.text = element_text(size=18)
+  )
+
+```
+
+```{r}
+
+ggsave("fe_tau_ExpShip_final.pdf", plot=plot_tau_response_crcvm200,
+       path="response_crcvm200", width=10, height=6, units="in")
+
+ggsave("fe_nu_ExpShip_final.pdf", plot=plot_nu_response_crcvm200,
+       path="response_crcvm200", width=10, height=6, units="in")
+plot_tau_response_crcvm200
+plot_nu_response_crcvm200
+
+```
+
+## Compute recovery distances
+
+```{r}
+make_recovery_distances_tab<-function(p) {
+
+  parameter_names=names(mean_alpha)
+  scaling=as.list(c(log(1-p),log(1+p)))
+  names(scaling)=parameter_names
+  estimates <- sapply(parameter_names,function(param) {
+    mean_alpha[[param]]/scaling[[param]]})
+
+  conf_intervals <- sapply(parameter_names, function(param) {
+    sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1]/scaling[[param]], 
+            quant_alpha[[param]][2]/scaling[[param]])
+  })
+
+  table_data <- data.frame(
+    Parameter = 
+      c("$D_{\\tau}^{ship}$","$D_{\\nu}^{ship}$"),
+    Estimate = sprintf("$%.2f$", estimates),
+    CI = conf_intervals,
+    stringsAsFactors = FALSE
+  )
+
+  xtab <- xtable(
+  table_data,
+  caption = "Estimated recovery distances of the baseline values",
+  label = paste0("table:recovery_distances_",p)
+  )
+
+  print(
+    xtab,type="latex",
+    include.rownames = FALSE,
+    sanitize.text.function = identity,  
+    hline.after = c(-1, 0, nrow(table_data)),
+    table.placement="H"
+
+  )
+}
+
+make_recovery_distances_tab(0.5)
+
+make_recovery_distances_tab(0.1)
+```

### 2025-03-17 R/application/CVM/fit_cvm.Rmd
update application to narwhal data

@@
 
-estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
-std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")
-
-```
-
-We process the results and show a table with the estimates and their standard deviations.
-
-```{r render coefficient estimates table baseline_ctcrw0 }
-
 # Function to extract estimate values into a data frame format
@@
 
+get_raw_latex_tab<- function(sde,show=TRUE) {
+
+  estimates=as.list(sde$tmb_rep(),what="Est")
+  std=as.list(sde$tmb_rep(),what="Std")
+
   # Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
-                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe, 
-                          std_bas_ctcrw0$coeff_fe, 
-                          rownames(estimates_bas_ctcrw0$coeff_fe))
+  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
+                                 std$log_sigma_obs, "log_sigma_obs")
+  coeff_fe <- combine_to_df(estimates$coeff_fe,
+                            std$coeff_fe, 
+                            rownames(estimates$coeff_fe))
+  if (!(is.null(rownames(estimates$coeff_re)))) {
+
+    coeff_re <- combine_to_df(estimates$coeff_re,
+                            std$coeff_re, 
+                            rownames(estimates$coeff_re))
+
+    log_lambda <- combine_to_df(estimates$log_lambda,
+                            std$log_lambda, 
+                            rownames(estimates$log_lambda))
 
      # Combine all data frames into one
-table_data_baseline_ctcrw0 <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis',message=FALSE}
-latex_table <- xtable(table_data_baseline_ctcrw0, 
+    table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)
+  } else {
+    table_data <- bind_rows(log_sigma_obs, coeff_fe)
+  }
+
+  latex_table <- xtable(table_data, 
                         caption = "Parameter Estimates and Standard Errors")
 
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
-```
-
-From this we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$.
-
-```{r CI baseline_ctcrw0 parameters}
-
-post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
-post_par_bas_ctcrw0=
-  list("sigma_obs"=1000*exp(post_coeff_bas_ctrcw0$log_sigma_obs),
-              "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
-              "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]))
-mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
-sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
-quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
+  if (show) {
+    print(
+    latex_table,type="latex",
+    include.rownames = FALSE,
+    sanitize.text.function = identity,
+    hline.after = c(-1, 0, nrow(table_data)-1),  
+    table.placement="H")
+  }
+
+  return (latex_table)
+
+}
+
+raw_ctcrw0_tab=get_raw_latex_tab(baseline_ctcrw0)
+
+```
+
+```{r}
+
+get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=FALSE) {
+
+  post_coeff=sde$post_coeff(n_post=100000)
+  if (ncol(post_coeff$log_lambda)==0) {
+
+  post_par=list(
+  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]))
+
+  parameter_names=c("$\\tau$", "$\\nu$")
+
+  } else {
+  post_par=list(
+  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]),
+  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
+  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))
+
+  parameter_names=c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$")
+  }
+
+  if (include_sigma_obs) {
+    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
+  }
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
                       quantile, probs = c(0.025, 0.975))
 
-```
-
-We include the results in a latex table.
-
-```{r CI table baseline_ctcrw0 parameters,results='asis',message=FALSE}
-
-parameter_names <- names(mean_par_bas_ctcrw0)
-estimates <- mean_par_bas_ctcrw0
+  parameter_names <- names(mean_par)
+  estimates <- mean_par
   conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
-          quant_par_bas_ctcrw0[[param]][2])
+  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
+          quant_par[[param]][2])
     })
 
+  if (include_sigma_obs) {
+    parameter_names=c(parameter_names,"$\\sigma_{\\obs}$")
+  }
   table_data <- data.frame(
-  Parameter = c("$\\sigma_{obs}$","$\\tau$", "$\\nu$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
+    Parameter = parameter_names,
+  Estimate = sprintf("$%.2f$", mean_par),
   CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-
-```{r,results='asis',echo=FALSE}
-# Create an xtable object
+  stringsAsFactors = FALSE)
+
   xtab <- xtable(
   table_data,
-  caption = "Baseline estimations",
-  label = "table:baseline_estimations"
-)
-
-# Customize xtable options
+  caption = paste(sde$type(),"baseline estimations"),
+  label = paste0("table:",sde$type(),"_baseline_estimations"))
+
+  if (show) {
   print(
-  xtab,
+  xtab,type="latex",
   include.rownames = FALSE,
@@
   hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
+  table.placement="H")
+  }
+
+  return(xtab)
+
+}
+
+final_ctcrw0_tab=get_final_baseline_tab(baseline_ctcrw0,show=TRUE)
 ```
@@
   list_baseline_ctcrw0_std[[i]]=std
-}
-
-```
-
-```{r render estimates table baselines with fixed mesaurement error,results='asis', message=FALSE}
-
-for (i in seq_along(list_baseline_ctcrw0)) {
-  estimates=list_baseline_ctcrw0_est[[i]]
-  std=list_baseline_ctcrw0_std[[i]]
-
-  # Create data frames for each set
-  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
-                               std$log_sigma_obs, "log_sigma_obs")
-  coeff_fe <- combine_to_df(estimates$coeff_fe, 
-                          std$coeff_fe, 
-                          rownames(estimates$coeff_fe))
-
-  # Combine all data frames into one
-  table_data <- bind_rows(log_sigma_obs, coeff_fe)
-
-  latex_table <- xtable(
-    table_data, 
-    caption = paste(
-      "Parameter Estimates and Standard Errors for fixed \\(\\sigma_{obs}=\\) ",
-      sigma_obs[i]," m"),
-    xtable.table.placement="H")
-
-  # Print LaTeX table
-  print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
-
 }
@@
 We see a clear effect of the fixed measurement error on the estimates of $\tau$, and a slight influence of $\nu$.
-We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ increases.
+We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ incre.ases.
 Overall, knowing or estimating precisely the measurement error will be crucial in the analysis.
+On the whole data, it seems more likely to have an error about 50 metres.
+We now fix the measurement error to 50m
 
@@
 formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))
+
+H=array(rep(0.05^2*diag(2),n_pre12),dim=c(2,2,n_pre12))
 
@@
                           response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
+                    par0 = par0,other_data=list("H"=H),
                     fixpar=c("mu1","mu2"))
@@
 ```
-
-```{r results baseline_ctcrw1}
-
-estimates_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Est")
-std_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_ctcrw1}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_ctcrw1$log_sigma_obs,
-                               std_bas_ctcrw1$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_ctcrw1$coeff_fe,
-                          std_bas_ctcrw1$coeff_fe,
-                          rownames(estimates_bas_ctcrw1$coeff_fe))
-log_lambda <- combine_to_df(estimates_bas_ctcrw1$log_lambda, 
-                            std_bas_ctcrw1$log_lambda, 
-                            rownames(estimates_bas_ctcrw1$log_lambda))
-coeff_re <- combine_to_df(estimates_bas_ctcrw1$coeff_re, 
-                          std_bas_ctcrw1$coeff_re, 
-                          rownames(estimates_bas_ctcrw1$coeff_re))
-
-# Combine all data frames into one
-table_data_baseline_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe, log_lambda, coeff_re)
-
-```
-
-```{r, results='asis', echo=FALSE}
-latex_table <- xtable(table_data_baseline_ctcrw1, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
-```
-
-As previously, we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$, and include the results in a latex table.
-
-```{r CI baseline_ctcrw1 parameters}
-
-post_coeff_bas_ctrcw1=baseline_ctcrw1$post_coeff(n_post=10000)
-post_par_bas_ctcrw1=list(
-  "sigma_obs"=1000*exp(post_coeff_bas_ctrcw1$log_sigma_obs),
-  "tau"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"tau.(Intercept)"]),
-  "nu"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"nu.(Intercept)"]),
-  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"tau.s(ID)"])),
-  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"nu.s(ID)"])))
-
-mean_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,mean)
-sd_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,sd)
-quant_par_bas_ctcrw1 <- lapply(post_par_bas_ctcrw1, 
-                               quantile, probs = c(0.025, 0.975))
-
-```
-
-```{r CI table baseline_ctcrw1 parameters, message=FALSE}
-
-parameter_names <- names(mean_par_bas_ctcrw1)
-estimates <- mean_par_bas_ctcrw1
-conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw1[[param]][1], 
-          quant_par_bas_ctcrw1[[param]][2])
-})
-
-table_data <- data.frame(
-  Parameter = 
-    c("$\\sigma_{obs}$","$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw1),
-  CI = conf_intervals,
-  stringsAsFactors = FALSE
-)
-
-```
-```{r,results='asis',echo=FALSE}
-xtab <- xtable(
-  table_data,
-  caption = "Baseline estimations with random effects",
-  label = "table:baseline_estimations_with_random_effects"
-)
-
-print(
-  xtab,type="latex",
-  include.rownames = FALSE,
-  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
-  hline.after = c(-1, 0, nrow(table_data)),  
-  table.placement="H"
-)
+```{r}
+final_tab_ctcrw1=get_final_baseline_tab(baseline_ctcrw1,show=TRUE)
 ```
@@
 ```
-
-### RACVM without random individual effects
-
-We can also test whether adding a constant rotational $\omega$ in the SDE model would make sense. For this, we fit a RACVM model.
-
-```{r fit baseline racvm,message=FALSE}
-
-#initial parameters
-par0 <- c(0,0,1,4,0)
-
-#model formula
-formulas <- list(mu1=~1,mu2=~1,
-                 tau =~1,
-                 nu=~1,
-                 omega=~1)
-
-#Fit baseline with measurement error estimated from the data
-baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
-                          response = c("x","y"),
-                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
-                    fixpar=c("mu1","mu2"))
-baseline_racvm0$fit()
-
-```
-
-```{r results baseline_racvm0}
-
-estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
-std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_racvm0}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
-                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe,
-                          std_bas_racvm0$coeff_fe, 
-                          rownames(estimates_bas_racvm0$coeff_fe))
-
-# Combine all data frames into one
-table_data_baseline_racvm0 <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis', echo=FALSE}
-latex_table <- xtable(table_data_baseline_racvm0, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
-```
-
-The estimate of $\omega$ is $0$.
-This indicates that a \textbf{constant} parameter $\omega$ is irrelevant in the model.
-As a consequence, the final baseline model we keep is \texttt{baseline\_ctcrw1} which corresponds to a CTCRW with constant parameters $\tau$ and $\nu$ and mean velocity fixed to zero.
 
@@
   response = c("x","y"),par0 = par0,
-  other_data=list("log_sigma_obs0"=log(0.05)),
-  fixpar=c("mu1","mu2"))
-
-baseline_ctcrw1_24h$fit()
-
-```
-
-```{r results baseline_ctcrw1_24h}
-
-estimates_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Est")
-std_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_ctcrw1_24h}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(
-  estimates_bas_ctcrw1_24h$log_sigma_obs, 
-  std_bas_ctcrw1_24h$log_sigma_obs, "log_sigma_obs")
-
-coeff_fe <- combine_to_df(estimates_bas_ctcrw1_24h$coeff_fe, 
-                          std_bas_ctcrw1_24h$coeff_fe, 
-                          rownames(estimates_bas_ctcrw1_24h$coeff_fe))
-
-# Combine all data frames into one
-table_data_baseline_ctcrw1_24h <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis',echo=FALSE}
-latex_table <- xtable(table_data_baseline_ctcrw1_24h, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,
-      table.placement="H")
-```
-
-The standard deviations are wider due to the smaller amount of data, but also due to the fact that the measurement error is estimated higher than before.
-$\tau$ seems to be estimated slightly below the value we obtained for the dataset \texttt{dataBE12}, but the difference seems reasonable as the CI of the two estimates overlap.
-
-```{r fit baseline sensitivity analysis tagging effect fixed error,message=FALSE}
-
-sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
-
-#Fit baseline with measurement error 
-baseline_ctcrw1_24h_fixed_error<- SDE$new(formulas = formulas,
-                                          data = dataBE24,type = "CTCRW",
-                                          response = c("x","y"),
-                                          par0 = par0,
   other_data=list("H"=H),
   fixpar=c("mu1","mu2"))
-baseline_ctcrw1_24h_fixed_error$fit()
-
-```
-
-```{r results baseline_ctcrw1_24h_fixed_error}
-
-estimates_bas_ctcrw1_24h_fixed_error=as.list(
-  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Est")
-std_bas_ctcrw1_24h_fixed_error=as.list(
-  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table baseline_ctcrw1_24h_fixed_error }
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(
-  estimates_bas_ctcrw1_24h_fixed_error$log_sigma_obs,
-  std_bas_ctcrw1_24h_fixed_error$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(
-  estimates_bas_ctcrw1_24h_fixed_error$coeff_fe, 
-  std_bas_ctcrw1_24h_fixed_error$coeff_fe, 
-  rownames(estimates_bas_ctcrw1_24h_fixed_error$coeff_fe))
-
-# Combine all data frames into one
-table_data_baseline_ctcrw1_24h_fixed_error <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis',echo=FALSE}
-latex_table <- xtable(table_data_baseline_ctcrw1_24h_fixed_error, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,
-      table.placement="H")
-```
-
-When the measurement error is fixed to the value we initially estimated from the whole dataset \texttt{dateBE12}, we observe a big decrease in the estimate of $\tau$.
-This indeed suggest a shift in the behavior of the whales within the time interval [12h,24h] after tagging, which might invalidate our results.
+
+baseline_ctcrw1_24h$update_lambda(c(exp(3),exp(5)))
+baseline_ctcrw1_24h$update_map(list("log_lambda"=factor(c(NA,NA))))
+
+baseline_ctcrw1_24h$fit()
+
+```
+```{r}
+final_tab_ctcrw1_24h=get_final_baseline_tab(baseline_ctcrw1_24h,show=TRUE)
+```
+
+The standard deviations are wider due to the smaller amount of data, but also due to the fact that the measurement error is estimated higher than before.
+$\tau$ seems to be estimated below the value we obtained for the dataset \texttt{dataBE12}, but the CIs of the two estimates overlap.
 
@@
 
-sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))
 
@@
 
-```{r results response_ctcrw1}
-
-estimates_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Est")
-std_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table response_ctcrw1}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_res_ctcrw1$log_sigma_obs, 
-                               std_res_ctcrw1$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_res_ctcrw1$coeff_fe,
-                          std_res_ctcrw1$coeff_fe, 
-                          rownames(estimates_res_ctcrw1$coeff_fe))
-
-# Combine all data frames into one
-table_data_response_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis', echo=FALSE}
-latex_table <- xtable(table_data_response_ctcrw1, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,
-      table.placement="H")
-```
-
-The estimated standard deviation of the random effects can be computed with \texttt{sdev} method in \texttt{smoothSDE}.
-
 ```{r}
-response_ctcrw1$sdev()
-```
-
-The estimates are lower than those obtained from the data before exposure.
-This may happen because we have short trajectories before exposure, that may not represent the whole range of possible movement for each narwhal, and therefore there may be more varieties of movements between the different individuals.
-On the contrary, the trajectories during exposure are longer so we can reasonably say that they represent better the range of possible movement.
-The lower standard deviation of the random effects might then indicate that on the long run, the narwhals in the study tend to have similar movement characteristics.
+final_tab_response_ctcrw1=get_final_baseline_tab(response_ctcrw1,show=TRUE)
+```
+
+The estimates are lower than those obtained from the data before exposure with first 12h removed, but quite similar to the data before exposure with first 24h removed.
 
@@
 The intercept of the parameters $\tau$ and $\nu$ are considered as offset in the statistical model, and fixed using the estimates we get from the baseline model \texttt{baseline\_ctcrw1}.
-The measurement error is an offset 7that is fixed to the value that has been estimated from the data before exposure.
+The measurement error is an offset that is fixed to the value that has been estimated from the data before exposure.
 
@@
 
-# Fit response with measurement error fixed to baseline value
-sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+# Fit response 
+H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))
 response_ctcrw_sp<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
@@
 #to keep the coefficients in the right order
-#otherwise we may have issues in function post_coeff
+#otherwise we may have trouble in function post_coeff
 response_ctcrw_sp$update_map(list(coeff_re=factor(
@@
 
-```
-
-```{r results response spline}
-
-estimates_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Est")
-std_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table response_ctcrw_sp}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_res_ctcrw_sp$log_sigma_obs, 
-                               std_res_ctcrw_sp$log_sigma_obs, "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_res_ctcrw_sp$coeff_fe,
-                          std_res_ctcrw_sp$coeff_fe,
-                          rownames(estimates_res_ctcrw_sp$coeff_fe))
-coeff_re <- combine_to_df(estimates_res_ctcrw_sp$coeff_re, 
-                          std_res_ctcrw_sp$coeff_re, 
-                          rownames(estimates_res_ctcrw_sp$coeff_re))
-log_lambda <- combine_to_df(estimates_res_ctcrw_sp$log_lambda, 
-                            std_res_ctcrw_sp$log_lambda, 
-                            rownames(estimates_res_ctcrw_sp$log_lambda))
-
-# Combine all data frames into one
-table_data_response_ctcrw_sp <- bind_rows(
-  log_sigma_obs, coeff_fe,coeff_re,log_lambda)
-
-```
-
-```{r,results='asis',echo=FALSE}
-latex_table <- xtable(table_data_response_ctcrw_sp, 
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,
-      table.placement="H")
 ```
@@
 This is likely connected to figure 7 in [Heide-Jorgensen], where depending on how the narwhals were first exposed to the ship, they either change direction (lower persistence) or persist in their heading to flee away (higher persistence).
+This may be also linked to the complex fjords boundaries in which the narwhals move that can trigger very different reaction depending on where the narwhals has been exposed to the ship.
 
@@
 # Fit response with measurement error fixed to baseline value
-sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))
 
@@
 
-```{r results response_ctcrw_exp}
-
-estimates_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Est")
-std_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Std")
-
-```
-
-```{r render estimates table response_ctcrw_exp}
-
-# Create data frames for each set
-log_sigma_obs <- combine_to_df(estimates_res_ctcrw_exp$log_sigma_obs, 
-                               std_res_ctcrw_exp$log_sigma_obs, 
-                               "log_sigma_obs")
-coeff_fe <- combine_to_df(estimates_res_ctcrw_exp$coeff_fe, 
-                          std_res_ctcrw_exp$coeff_fe,
-                          rownames(estimates_res_ctcrw_exp$coeff_fe))
-
-# Combine all data frames into one
-table_data_response_ctcrw_exp <- bind_rows(log_sigma_obs, coeff_fe)
-
-```
-
-```{r, results='asis', echo=FALSE}
-latex_table <- xtable(table_data_response_ctcrw_exp,
-                      caption = "Parameter Estimates and Standard Errors")
-
-# Print LaTeX table
-print(latex_table, type = "latex", include.rownames = FALSE,
-      table.placement="H")
-```
-
-We can also compute $95 \%$ CI for the coefficients in the log-linear model.
-
-```{r CI response_ctcrw_exp parameters}
-
-post_coeff_res_exp=response_ctcrw_exp$post_coeff(n_post=10000)
-post_par_res_exp=list(
-  "alpha_tau"=post_coeff_res_exp$coeff_fe[,"tau.ExpShip"],
-  "alpha_nu"=post_coeff_res_exp$coeff_fe[,"nu.ExpShip"])
-
-mean_par_res_exp=lapply(post_par_res_exp,mean)
-sd_par_res_exp=lapply(post_par_res_exp,sd)
-quant_par_res_exp <- lapply(post_par_res_exp, 
+```{r}
+get_final_response_tab<-function(sde,show=TRUE) {
+
+  post_coeff=sde$post_coeff(n_post=10000)
+  post_par=list(
+  "alpha_tau"=post_coeff$coeff_fe[,"tau.ExpShip"],
+  "alpha_nu"=post_coeff$coeff_fe[,"nu.ExpShip"])
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
                       quantile, probs = c(0.025, 0.975))
 
-```
-
-```{r CI table response_ctcrw_exp parameters, message=FALSE}
-
-parameter_names <- names(mean_par_res_exp)
-estimates <- mean_par_res_exp
+  parameter_names <- names(mean_par)
+  estimates <- mean_par
   conf_intervals <- sapply(parameter_names, function(param) {
-  sprintf("$[%.2f; %.2f]$", quant_par_res_exp[[param]][1], 
-          quant_par_res_exp[[param]][2])
+    sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
+            quant_par[[param]][2])
   })
@@
       c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
-  Estimate = sprintf("$%.2f$", mean_par_res_exp),
+    Estimate = sprintf("$%.2f$", mean_par),
     CI = conf_intervals,
@@
   )
-
-```
-
-```{r,results='asis',echo=FALSE}
   xtab <- xtable(
@@
 
+  if (show) {
     print(
@@
     hline.after = c(-1, 0, nrow(table_data)),
-  table.placement="H"
-)
+    table.placement="H")
+  }
+
+  return (xtab)
+
+}
+
+final_tab_response_ctcrw_exp=get_final_response_tab(response_ctcrw_exp)
 ```
@@
 
-The results are indeed similar to what we obtain using cubic splines with a low degree of freedom.
+The results are indeed similar to what we obtain using cubic splines with a low degree of freedom on both $\nu$ and $\tau$.
 
@@
 post_coeff_lambda=exp(post_coeff$log_lambda)
-post_coeff_sigma_obs=exp(post_coeff$log_sigma_obs)
 
@@
 
-    sigma_obs=post_coeff_sigma_obs[i,1]
-    H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
+    H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))
 
@@
 ```
+
 ```{r propagated CI alpha log linear coefficients, message=FALSE}
@@
 
+## Compute recovery distances
+
+```{r}
+make_recovery_distances_tab<-function(p) {
+
+  parameter_names=names(mean_alpha)
+  scaling=as.list(c(log(1-p),log(1+p)))
+  names(scaling)=parameter_names
+  estimates <- sapply(parameter_names,function(param) {
+    mean_alpha[[param]]/scaling[[param]]})
+
+  conf_intervals <- sapply(parameter_names, function(param) {
+    sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1]/scaling[[param]], 
+            quant_alpha[[param]][2]/scaling[[param]])
+  })
+
+  table_data <- data.frame(
+    Parameter = 
+      c("$D_{\\tau}^{ship}$","$D_{\\nu}^{ship}$"),
+    Estimate = sprintf("$%.2f$", estimates),
+    CI = conf_intervals,
+    stringsAsFactors = FALSE
+  )
+
+  xtab <- xtable(
+  table_data,
+  caption = "Estimated recovery distances of the baseline values",
+  label = paste0("table:recovery_distances_",p)
+  )
+
+  print(
+    xtab,type="latex",
+    include.rownames = FALSE,
+    sanitize.text.function = identity,  
+    hline.after = c(-1, 0, nrow(table_data)),
+    table.placement="H"
+
+  )
+}
+
+make_recovery_distances_tab(0.5)
+
+make_recovery_distances_tab(0.1)
+```
+
 # Simulation from the baseline and response models
@@
 We know want to investigate how the fitted model match the actual observations.
-To fo this, we simulate trajectories from the fitted model before and during exposure.
-First, we ignore that the narwhals move in a restricted area and only look at the shape of the trajectories.
+To do this, we simulate trajectories from the fitted model before and during exposure.
+We ignore that the narwhals move in a restricted area and only look at the shape of the trajectories.
 
@@
                                             data=baseline_ctcrw1$data(),
-                                            sd_noise=sigma_obs,land=land)
+                                            sd_noise=sigma_obs,land=land,reflect=TRUE)
 # Add a label to each dataset for plotting
@@
                                             data=response_ctcrw_exp$data(),
-                                            sd_noise=sigma_obs,land=land)
+                                            sd_noise=sigma_obs,land=land,reflect=TRUE)
 # Add a label to each dataset for plotting

### 2025-03-17 R/application/RCVM/fit_rcvm.Rmd
update application to narwhal data

@@
 
-In this analysis, we introduce covariates \texttt{AngleNormal} and \texttt{Ishore} ($\Theta$ and $I_{shore}$) in the SDE models for the narwhal data.
-This allows to better understand how the fjords shoreline influences the narwhals movement in the baseline model.
+In this analysis, we introduce covariates \texttt{AngleNormal} and 
+\texttt{Ishore} ($\Theta$ and $I_{shore}$) in the SDE models for 
+the narwhal data.
+This allows to better understand how the fjords shoreline influences 
+the narwhals movement in the baseline model.
 
@@
 library(ggplot2)
+library(ggpubr)
+library(patchwork)
 library(dplyr)
@@
 library(sf)
+library(purrr)
 library(gganimate) # animated trajectories
@@
 par_dir=here() #parent directory 
-narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")
+narwhal_data_path <- file.path(par_dir,
+                               "Data","preprocessed_data","narwhals")
 
 #data before exposure with first 24h removed
-dataBE24=read.csv(file.path(narwhal_data_path,"DataBE24.csv"), header = TRUE,dec = ".")
+dataBE24=read.csv(file.path(narwhal_data_path,"DataBE24.csv"),
+                  header = TRUE,dec = ".")
 
 #data before exposure with first 12h removed
-dataBE12=read.csv(file.path(narwhal_data_path,"DataBE12.csv"), header = TRUE,dec = ".")
+dataBE12=read.csv(file.path(narwhal_data_path,"DataBE12.csv"),
+                  header = TRUE,dec = ".")
 
 #data after exposure
-dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"), header = TRUE,dec = ".")
-
-```
-
-We choose thresholds on the distance to the shore to define the Ishore covariate.
+dataAE=read.csv(file.path(narwhal_data_path,"DataAE.csv"),
+                header = TRUE,dec = ".")
+
+```
+
+We choose thresholds on the distance to the shore
+to define the Ishore covariate.
 
@@
 
-```{r}
+We bind data before exposure and after exposure in a single dataframe.
+
+```{r bind data before and after exposure}
 
@@
 
+# Interpolate every minute
+
+```{r smooth and interpolate trajectories}
+smooth_interpolate <- function(df) {
+
+  # Create a sequence of regular time points (every minute)
+  time_seq <- seq(min(df$time), max(df$time), by = 1/30)
+
+  # Fit smoothing splines
+  smooth_x <- smooth.spline(df$time, df$x)
+  smooth_y <- smooth.spline(df$time, df$y)
+
+  # Predict at regular intervals
+  interp_x <- predict(smooth_x, as.numeric(time_seq))$y
+  interp_y <- predict(smooth_y, as.numeric(time_seq))$y
+
+  # Return the interpolated data
+  data.frame(time = time_seq, x = interp_x, y = interp_y, ID = unique(df$ID))
+}
+
+# Apply the function for each ID
+interpolated_dataBE12 <- dataBE12 %>%
+  group_by(ID) %>%
+  group_split() %>%
+  map_dfr(smooth_interpolate)
+
+# Plot the original and smoothed trajectories by ID
+plot_interpolation=ggplot() +
+  geom_point(data = dataBE12, aes(x = x, y = y, color = factor(ID)), size = 2) +
+  geom_path(data = interpolated_dataBE12, aes(x = x, y = y, color = factor(ID))) +
+  theme_minimal() +
+  labs(color = "Trajectory ID") +
+  ggtitle("GPS points and splines interpolation")
+```
+
+# Compute Dshore and theta from interpolation
+
+```{r}
+
+signed_angle <- function(u, v) {
+  #Compute signed angle in [-pi,pi] that rotates first vector into second vector 
+  # as in 
+  # https://math.stackexchange.com/questions/529555/signed-angle-between-2-vectors
+  u <- matrix(u, ncol = 2)
+  v <- matrix(v, ncol = 2)
+  if (nrow(u) != nrow(v)) stop("u and v must have the same number of 
+                                  rows")
+  result <- as.numeric(atan2(v[,2], v[,1]) - atan2(u[,2], u[,1]))
+  ind1 <- which(result > pi)
+  ind2 <- which(result <= -pi)
+  result[ind1] <- result[ind1] - 2*pi
+  result[ind2] <- result[ind2] + 2*pi
+  return(result) 
+}
+
+add_covs_parallel <- function(data, n_cores = parallel::detectCores() - 1) {
+  # Setup parallel backend
+  cl <- makeCluster(n_cores)
+  registerDoParallel(cl)
+
+  # Split data by ID
+  ids <- unique(data$ID)
+
+  # Process each ID in parallel
+  results <- foreach(id = ids, .combine = rbind, .packages = c("sf"),
+                     .export = c("nearest_boundary_points", "is_in_land", "signed_angle","border",
+                                 "D_low","D_up")  ) %dopar% {
+                                   # Filter data for the current ID
+                                   sub_data <- data[data$ID == id, ]
+                                   n_sub <- nrow(sub_data)
+
+                                   # Time steps
+                                   dtimes <- sub_data[2:n_sub, "time"] -sub_data[1:(n_sub - 1), "time"]
+
+                                   # Step lengths
+                                   dx <- sub_data[2:n_sub, "x"] - sub_data[1:(n_sub - 1), "x"]
+                                   dy <- sub_data[2:n_sub, "y"] - sub_data[1:(n_sub - 1), "y"]
+
+                                   # Empirical velocity
+                                   vexp_df <- cbind(dx / dtimes, dy / dtimes)
+
+                                   # Nearest points on shore
+                                   sub_data <- cbind(sub_data, nearest_boundary_points(as.matrix(sub_data[, c("x", "y")]), border))
+
+                                   # Normal vectors
+                                   normal <- as.matrix(sub_data[2:n_sub, c("x", "y")] - sub_data[2:n_sub, c("p1", "p2")])
+
+                                   # Angle between velocity and normal vector
+                                   theta_coast <- signed_angle(normal, vexp_df)
+                                   theta_coast <- c(theta_coast, 1)  # Adjust length
+
+                                   # Initialize DistanceShore
+                                   Dshore <- rep(0, n_sub)
+
+                                   for (i in 1:n_sub) {
+                                     y1 <- sub_data[i, "x"]
+                                     y2 <- sub_data[i, "y"]
+                                     if (!is_in_land(st_point(c(y1, y2)), border)) {
+                                       p1 <- sub_data[i, "p1"]
+                                       p2 <- sub_data[i, "p2"]
+                                       Dshore[i] <- sqrt((y1 - p1)^2 + (y2 - p2)^2)
+                                     }
+                                   }
+
+                                   # Calculate Ishore
+                                   Ishore <- ifelse(Dshore < D_low, 1 / D_low, ifelse(Dshore > D_up, 0, 1 / Dshore))
+
+                                   # Add columns to sub_data
+                                   sub_data$theta <- theta_coast
+                                   sub_data$DistanceShore <- Dshore
+                                   sub_data$Ishore <- Ishore
+
+                                   return(sub_data)
+                                 }
+
+  # Stop the cluster
+  stopCluster(cl)
+
+  return(results)
+}
+
+interpolated_dataBE12=add_covs_parallel(interpolated_dataBE12)
+
+```
+
 # Fit baseline models
 
-We try sde models where the parameters are smooth function of the covariates AngleNormal and Ishore.
-Here, for simplicity we will always fix the measurement error to $30$ m. This choice is based on the paper of Weensven.
+We try sde models where the parameters are smooth function 
+of the covariates theta and Ishore.
+Here, for simplicity we will always fix the measurement error to $50$ m. 
+This choice is based on the paper of Weensven.
 We will assess the sensitivity of our results to the hyperparameter later.
@@
 
-The first model is a CTCRW where the covariate theta influences the persistence of the movement.
+The first model is a CTCRW where the covariate theta influences 
+the persistence of the movement.
 
@@
 #model formula
-formulas <- list(mu1=~1,mu2=~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))
+formulas <- list(mu1=~1,mu2=~1,tau =~s(theta,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))
 
 #Fixed measurement error
-sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))
+sigma_obs=0.01
+H=array(rep(sigma_obs^2*diag(2),length(interpolated_dataBE12$time)),dim=c(2,2,length(interpolated_dataBE12$time)))
 
 #Fit baseline with measurement error estimated from the data
-baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
+baseline_ctcrw0<- SDE$new(formulas = formulas,data = interpolated_dataBE12,type = "CTCRW",
                           response = c("x","y"),
@@
 
-get_raw_latex_tab<- function(sde,show=TRUE) {
+make_raw_latex_tab<- function(sde) {
 
@@
 
-  if (show) {
   print(latex_table, type = "latex", include.rownames = FALSE)
-  }
-
-  return (latex_table)
-
-}
-
-```
-
-```{r estimates baseline_ctcrw0}
-raw_tab_baseline_ctcrw0=get_raw_latex_tab(baseline_ctcrw0)
-```
-
-We can get confidence intervals for the intercept of each parameter and the random effects standard deviations.
+
+}
+
+```
+
+```{r raw estimates baseline_ctcrw0}
+make_raw_latex_tab(baseline_ctcrw0)
+```
+
+We can get confidence intervals for the intercept 
+of each parameter and the random effects standard deviations.
 
@@
 
-get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=TRUE) {
+get_baseline_estimates<- function(sde,include_sigma_obs=FALSE) {
 
@@
 
+  return(list("estimates"=mean_par,"std.error"=sd_par,"quantiles"=quant_par))
+
+}
+
+make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE) {
+
+  est=get_baseline_estimates(sde,include_sigma_obs)
+  mean_par=est$estimates
+  sd_par=est$std.error
+  quant_par=est$quantiles
   parameter_names <- names(mean_par)
@@
 
-  if (show) {
   print(
@@
   table.placement="H")
-  }
-
-  return(xtab)
 
@@
 
-final_tab_baseline_ctcrw0=get_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)
-
-```
-
-We can now plot the smooth parameter $\tau$ as a function of the covariate \texttt{AngleNormal}.
+make_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)
+
+```
+
+We can now plot the smooth parameter $\tau$ as a 
+function of the covariate \texttt{AngleNormal}.
 
 ```{r plot smooth parameters baseline_ctcrw0 }
-baseline_ctcrw0$plot_par(var="theta_translated",
+baseline_ctcrw0$plot_par(var="theta",
                          par_names=c("tau","nu"),n_post=100)
@@
 
-The movement is clearly more persistent along the shoreline, meaning that $\tau$ increases when \texttt{AngleNormal} is close to $\pm \frac{\pi}{2}$).
-
-## Baseline racvm with AngleNormal covariate in tau and omega
-
-We now decide to introduce a rotational parameter $\omega$ in the SDE to allow the velocity to rotate depending on the narwhals position with respect to the shoreline.
-We start by introducing the covariate \texttt{theta} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.
+The movement is clearly more persistent along the shoreline, meaning
+that $\tau$ increases when \texttt{AngleNormal} is close to $\pm \frac{\pi}{2}$).
+
+## Baseline racvm with $\Theta$ covariate in tau and omega
+
+We now decide to introduce a rotational parameter $\omega$ in the SDE to
+allow the velocity to rotate depending on the narwhals position with 
+respect to the shoreline.
+We start by introducing the covariate \texttt{theta} in the new smooth 
+parameter $\omega$ to understand how the animals may rotate depending on 
+the direction of their movement with respect to the shoreline.
 
@@
 # Fit baseline with measurement error estimated from the data
-baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
+baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,
+                          type = "RACVM_SSM",response = c("x","y"),
                     par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
@@
 
-```{r estimates baseline_racvm0}
-raw_tab_baseline_racvm0=get_raw_latex_tab(baseline_racvm0)
+```{r raw estimates baseline_racvm0}
+make_raw_latex_tab(baseline_racvm0)
 ```
@@
 
-final_tab_baseline_racvm0=get_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)
+make_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)
 
@@
 $\omega$ close to $\pm \pi$ indicates movement toward the shoreline.
-Here, it seems that the narwhals have a slight tendency to rotate, presumably to move along the shoreline or change direction, when they are heading toward the shore.
+Here, it seems that the narwhals have a slight tendency to rotate,
+presumably to move along the shoreline or change direction, 
+when they are heading toward the shore.
 However, it is not very clear since the confidence intervals are wide.
@@
 
-## Baseline with AngleNormal and DistanceShore in omega and tau
-
-We add the effect of the distance to the shore through the covariate \texttt{Ishore_translated}.
+## Baseline with theta and DistanceShore in $\omega$ and $\tau$
+
+We add the effect of the distance to the shore 
+through the covariate \texttt{Ishore_translated}.
 This model has a lot of parameters and takes around 15 minutes to fit. 
-Indeed, we will need to fix the penalties in the tensor splines in advance to ensure convergence of the optimizer.
-The standard deviation of the random effects and the smoothing penalties for the splines in $\tau$ will still be estimated from the data and not fixed.
-To illustrate this we first fit a model where the smoothing penalties are all estimated from the data.
+Indeed, we will need to fix the penalties in the tensor splines in 
+advance to ensure convergence of the optimizer. 
+The standard deviation of the random effects and the smoothing penalties 
+for the splines in $\tau$ will still be estimated from the data and not fixed.
+To illustrate this we first fit a model where the smoothing penalties 
+are all estimated from the data.
 
@@
 
-The smoothing penalties estimates are extremely low (the smoothing penalties are the inverse of the square root of \texttt{omega.te(theta\_translated,Ishore\_translated)}).
-This suggests that the optimization did not go well and numerical errors occurred.
-For this reason, we decide to fix the smoothing penalties to "reasonable" values in the tensor splines.
-
-First, we get good values for the smoothing penalty by approximating an expected "reasonable" surface with tensor splines, using the parametric function defined in the paper.
-
-```{r}
-
-fomega=function(cov_data,a=1,b=2,D0=0.3,D1=1,sigma_D=0.3,sigma_theta=pi/6){
+The smoothing penalties estimates are extremely low (the smoothing penalties are
+the inverse of the square root
+of \texttt{omega.te(theta\_translated,Ishore\_translated)}).
+This suggests that the optimization did not go well and numerical errors
+occurred. For this reason, we decide to fix the smoothing penalties to 
+"reasonable" values in the tensor splines.
+
+First, we get good values for the smoothing penalty by approximating an 
+expected "reasonable" surface with tensor splines, using the 
+parametric function defined in the paper.
+
+```{r get smooth penalty values}
+
+fomega=function(cov_data,a=3,b=2,D0=0.4,D1=1,sigma_D=0.5,sigma_theta=pi/6){
   Dshore=cov_data$DistanceShore
@@
   a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
-      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
-           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
+      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+
+                     ((Dshore-D1)/sigma_D)^2))-
+           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+
+                       ((Dshore-D1)/sigma_D)^2)))
 }
@@
 #parameters formulas
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
+formulas <- list(mu1 = ~1 ,mu2 =~1,
+                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),
                    omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
@@
 
-```{r estimates baseline_racvm1}
-raw_tab_baseline_racvm1=get_raw_latex_tab(baseline_racvm1)
+```{r raw estimates baseline_racvm1}
+make_raw_latex_tab(baseline_racvm1)
 ```
@@
 
-final_tab_baseline_racvm1=get_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)
+make_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)
 
@@
 plots=lapply(1:length(Ishore_values),function(i){
-  baseline_racvm1$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
+  baseline_racvm1$plot_par(
+    var="theta_translated",
+    covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
     par_names=c("omega"),n_post=100)})
+
 plots
@@
 ```
-Now, we would like to demonstrate how the Distance to shore influences the smooth parameter $\omega$.
-We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.
+
+Now, we would like to demonstrate how the Distance to shore 
+influences the smooth parameter $\omega$.
+We plot the estimated smooth parameter $\omega$ as a function 
+of the angle for different values of the Distance to shore.
 
@@
 
-ggsave("omega_DistanceShore_levels.pdf", plot=shore_influence_plot_baseline_racvm1, path="baseline_racvm1", width=10, height=6, units="in")
-
-```
-
-There seems to be a rotation close to the shore when the angle $\Theta \in [-\pi,-\frac{\pi}{2}] \cup [\frac{\pi}{2},\pi]$, which means movement toward the shore.
-However, again the confidence intervals are too wide to really conclude.
+ggsave("omega_DistanceShore_levels.pdf",
+       plot=shore_influence_plot_baseline_racvm1,
+       path="baseline_racvm1", width=10, height=6, units="in")
+
+```
+
+There seems to be a rotation close to the shore when the angle
+$\Theta \in [-\pi,-\frac{\pi}{2}] \cup [\frac{\pi}{2},\pi]$, which means
+movement toward the shore. However, again the confidence intervals are 
+too wide to really conclude.
 Moreover, we fixed the smoothing penalties in the tensor splines. 
-Changing their values can influence the shape of the estimated functions : lower values will produce functions that are more wiggily in each direction.
-
-```{r}
-
-estimate_parametric_omega=function(model,fixed_a=1) {
+Changing their values can influence the shape of the estimated functions :
+lower values will produce functions that are more wiggily in each direction.
+
+```{r define smooth parametric estimate baseline_racvm1}
+
+estimate_parametric_omega=function(model,fixed_a=NULL,probs=c(0.5,0.95)) {
 
@@
 
-    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),
-               to=quantile(data$DistanceShore_translated,0.95),length.out=30)
+    Dshore=seq(from=quantile(data$DistanceShore_translated,probs[1]),
+               to=quantile(data$DistanceShore_translated,probs[2]),length.out=30)
     Ishore=1/Dshore
@@
 
-    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
+    if (is.null(fixed_a)) {
+
+      fit_optim <- optim(par = log(c(1,0.5,0.1,2,0.5,pi/6)),
                        fn = function(params) {
@@
                         # Compute predicted z values
-                        z_pred <- fomega(data_copy,fixed_a,exp(params[1]),
-                                         exp(params[2]),exp(params[3]),exp(params[4]),exp(params[5]))
+                        z_pred <- fomega(data_copy,exp(params[1]),
+                                         exp(params[2]),exp(params[3]),
+                                         exp(params[4]),exp(params[5]),
+                                         exp(params[6]))
 
@@
 
+    } 
+    else {
+    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
+                       fn = function(params) {
+
+                        # Compute predicted z values
+                        z_pred <- fomega(data_copy,fixed_a,exp(params[1]),
+                                         exp(params[2]),exp(params[3]),
+                                         exp(params[4]),exp(params[5]))
+
+                         # Return sum of squared residuals
+                         sum((data_copy$z - z_pred)^2)
+                       })
+
+    }
     est_par=fit_optim$par
 
-    print(exp(est_par))
-
     # Compute the predicted values using the estimated parameters
-    z_pred_starting <- fomega(grid,fixed_a,exp(est_par[1]),
-                              exp(est_par[2]),exp(est_par[3]),exp(est_par[4]),exp(est_par[5]))
+    if (is.null(fixed_a)) {
+
+      z_pred_starting <- fomega(grid,exp(est_par[1]),
+                              exp(est_par[2]),exp(est_par[3]),
+                              exp(est_par[4]),exp(est_par[5]),
+                              exp(est_par[6]))
+
+    } else {
+    z_pred_starting <- fomega(grid,fixed_a,
+                              exp(est_par[1]),exp(est_par[2]),
+                              exp(est_par[3]),exp(est_par[4]),
+                              exp(est_par[5]))
+    }
 
     # Reshape z_pred_starting to a matrix for plotting
-    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore), ncol = length(theta_translated))
+    z_pred_matrix <- matrix(z_pred_starting, nrow = length(Dshore),
+                            ncol = length(theta_translated))
 
     # Create the 3D surface plot using plot_ly
-    plot_parametric <- plot_ly(x = ~theta_translated, y = ~Dshore, z = ~z_pred_matrix) %>%
+    plot_parametric <- plot_ly(x = ~theta_translated, y = ~Dshore,
+                               z = ~z_pred_matrix) %>%
       add_surface() %>%
@@
 
-    return (list("est_par"=est_par,"splines_surface"=plot_splines,"parametric_surface"=plot_parametric))
-
-}
-
-res_racvm1=estimate_parametric_omega(baseline_racvm1,fixed_a=2)
+    return (list("est_par"=exp(est_par),"splines_surface"=plot_splines,
+                 "parametric_surface"=plot_parametric))
+
+}
+
+res_racvm1=estimate_parametric_omega(baseline_racvm1,fixed_a=0.3)
 
@@
 
-This model in indeed very complex and involves a large number of parameters given the amount of data points we have for the trajectories before exposure.
-We can try to fit the exact same model on all the data to check if we get the same kind of results with tighter confidence intervals.
-Here, we fix the measurement error standard deviation because we get a non positive definite Hessian matrix in the optimization otherwise.
+This model in indeed very complex and involves a large number of parameters 
+given the amount of data points we have for the trajectories before exposure.
+We can try to fit the exact same model on all the data to check if
+we get the same kind of results with tighter confidence intervals.
+Here, we fix the measurement error standard deviation because 
+we get a non positive definite Hessian matrix in the optimization otherwise.
 
@@
 sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),dim=c(2,2,length(all_data$time)))
+H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),
+        dim=c(2,2,length(all_data$time)))
 
 #parameters formulas
-formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
+formulas <- list(mu1 = ~1 ,mu2 =~1,
+                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                    omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))
@@
 
-final_tab_all_racvm1=get_final_baseline_tab(all_racvm1,include_sigma_obs=FALSE)
-
-```
-
-We now come back to the data before exposure and try a more parsimonious model by removing the dependence of $\tau$ on
+make_final_baseline_tab(all_racvm1,include_sigma_obs=FALSE)
+
+```
+
+We now come back to the data before exposure and try a more parsimonious 
+model by removing the dependence of $\tau$ on
 the covariate \texttt{theta\_translated}.
@@
 
+#Fixed measurement error
+sigma_obs=0.01
+H=array(rep(sigma_obs^2*diag(2),length(interpolated_dataBE12$time)),dim=c(2,2,length(interpolated_dataBE12$time)))
+
 #parameters formulas
 formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
-                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs=c("cs")))
+                   omega=~te(theta,Ishore,k=c(5,5),bs=c("cs")))
 
 # Fit baseline
-baseline_racvm2<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
-                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
-                    fixpar=c("mu1","mu2"))
+baseline_racvm2<- SDE$new(formulas = formulas,data = interpolated_dataBE12,
+                    type = "RACVM_SSM",response = c("x","y"),par0 = par0,
+                    other_data=list("H"=H),fixpar=c("mu1","mu2"))
 
@@
 
-final_tab_baseline_racvm2=get_final_baseline_tab(baseline_racvm2,include_sigma_obs=FALSE)
+make_final_baseline_tab(baseline_racvm2,include_sigma_obs=FALSE)
 
@@
 plots=lapply(1:length(Ishore_values),function(i){
-  baseline_racvm2$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
+  baseline_racvm2$plot_par(
+    var="theta_translated",
+    covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                          par_names=c("omega"),n_post=100)})
@@
 
-xmin=list("Ishore_translated"=1/3)
-xmax=list("Ishore_translated"=1/0.3)
-link=list("Ishore_translated"=\(x) 1/x)
-xlabel=list("Ishore_translated"="Distance to shore")
+xmin=list("Ishore"=1/3)
+xmax=list("Ishore"=1/0.1)
+link=list("Ishore"=\(x) 1/x)
+xlabel=list("Ishore"="Distance to shore")
 
@@
 
-```{r}
-
-res_racvm2=estimate_parametric_omega(baseline_racvm2,fixed_a=2)
+```{r get smooth parametric estimates baseline_racvm2}
+
+res_racvm2=estimate_parametric_omega(baseline_racvm2,
+                                     fixed_a=4,probs=c(0.75,0.95))
 
@@
 
-## Simulate trajectories before exposure from fitted models
-
-```{r,cache=TRUE}
+To check the baseline model, we simulate trajectory using the estimated 
+parameters, compare it to the observed data and use the simulated trajectory to 
+estimate again the parameters. We check that the values estimated 
+from the simulated trajectories are consistent with the values 
+estimated from the data.
+
+# Check baseline models
+
+## Simulate trajectories using estimated parameters
+
+```{r set up for simulation of trajectories,cache=TRUE}
 
@@
 # Define covariates data (needed in the function simulate)
-data_reg=data.frame("AngleNormal"=rep(0,n),"Ishore"=rep(1,n),"Ishore_translated"=rep(1,n),
-                    "theta"=rep(0,n),"DistanceShore"=rep(1,n),"theta_translated"=rep(0,n))
+data_reg=data.frame("Ishore"=rep(1,n),"Ishore_translated"=rep(1,n),
+                    "theta"=rep(0,n),"DistanceShore"=rep(1,n),
+                    "theta_translated"=rep(0,n))
 
@@
 
-```{r}
-est_par_racvm1=res_racvm1$est_par
-
-formulas=list(tau=~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
-              sigma_D=~1,sigma_theta=~1)
-baseline_crcvm1<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
-               par0=c(baseline_racvm1$par()[1,c("tau","nu")],2,
-                        exp(est_par_racvm1)))
-
-baseline_crcvm1$update_coeff_re(baseline_racvm1$coeff_re()[1:20,])
-baseline_crcvm1$update_lambda(baseline_racvm1$lambda()[1:3,])
-
-baseline_crcvm1_trajectory=baseline_crcvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,verbose=FALSE)
-
-n=length(baseline_crcvm1_trajectory$time)
-
-baseline_crcvm1_trajectory_sub=baseline_crcvm1_trajectory[seq(1,n,by=60),]
-
-plot_baseline_crcvm1_trajectory=ggplot() +
-  geom_sf(data=land$geometry,fill="grey")+
-  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
-  geom_path(data=baseline_crcvm1_trajectory_sub, 
-            aes(x = x, y = y,color=ID),size = 0.5, 
-            lineend = "round",alpha=0.15) +     
-  geom_point(data=baseline_crcvm1_trajectory_sub,
-             aes(x = x, y = y,color=ID),size =0.5 ) +
-  theme_minimal()
-```
-
 ```{r,message=FALSE,results=console}
 
-set.seed(1)
+#set.seed(51)
+set.seed(52)
 est_par_racvm2=res_racvm2$est_par
+fixed_a=4
 
@@
               sigma_D=~1,sigma_theta=~1)
-baseline_crcvm2<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
-               par0=c(baseline_racvm2$par()[1,c("tau","nu")],2,
-                        exp(est_par_racvm2)))
+baseline_crcvm2<-SDE$new(formulas=formulas,data=dataBE12,
+                         type="CRCVM_SSM",response=c("x","y"),
+               par0=c(baseline_racvm2$par()[1,c("tau","nu")],fixed_a,
+                        est_par_racvm2))
 
@@
 
-baseline_crcvm2_trajectory=baseline_crcvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=shrunk_land,verbose=FALSE)
+baseline_crcvm2_trajectory=baseline_crcvm2$simulate(
+  z0=z0_BE,data=data_reg,atw=atw,land=shrunk_land,verbose=FALSE)
 
@@
 plot_baseline_crcvm2_trajectory=ggplot() +
-  geom_sf(data=land$geometry,fill="grey")+
+  geom_sf(data=shrunk_land$geometry,fill="grey")+
   coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
@@
              aes(x = x, y = y,color=ID),size =0.5 ) +
-  theme_minimal()
-
-ggsave("plot_baseline_crcvm2_trajectory.pdf", plot=plot_baseline_crcvm2_trajectory,
-       path="baseline_racvm2", width=10, height=6, units="in")
+  theme_minimal()+
+  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
 
@@
 
-baseline_crcvm1_trajectory_sub <- baseline_crcvm1_trajectory_sub %>% 
-  mutate(data_type = "Simulated baseline_crcvm1")
 baseline_crcvm2_trajectory_sub<- baseline_crcvm2_trajectory_sub %>% 
@@
 
-baseline_trajectories <- bind_rows(baseline_crcvm1_trajectory_sub,
-                                   baseline_crcvm2_trajectory_sub,observed_trajectory)
+baseline_trajectories <- bind_rows(baseline_crcvm2_trajectory_sub,
+                                   observed_trajectory)
 
@@
 
-We continue the analysis with the fitted baseline model \texttt{baseline\_racvm2}.
-
 ## Check estimation on simulations from fitted model
@@
                                    # Nearest points on shore
-                                   sub_data <- cbind(sub_data, nearest_boundary_points(as.matrix(sub_data[, c("x", "y")]), border))
+                                   sub_data <-cbind(sub_data,
+                                                    nearest_boundary_points(as.matrix(sub_data[, c("x", "y")]), border))
 
@@
 # add noise
-noise=rmvn(length(baseline_crcvm2_trajectory_sub$time),rep(0,2),diag(rep(0.05^2,2)))
-baseline_crcvm2_trajectory_sub_obs=baseline_crcvm2_trajectory[seq(1,n,by=30),c("time","x","y","ID")]
-baseline_crcvm2_trajectory_sub_obs[,c("x","y")]=baseline_crcvm2_trajectory_sub_obs[,c("x","y")]+noise
-baseline_crcvm2_trajectory_sub_obs=add_covs_parallel(baseline_crcvm2_trajectory_sub_obs)
+baseline_crcvm2_trajectory_lf_obs=baseline_crcvm2_trajectory_sub[,c("time","x","y","ID")]
+noise=rmvn(length(baseline_crcvm2_trajectory_lf_obs$time),rep(0,2),diag(rep(0.05^2,2)))
+
+baseline_crcvm2_trajectory_lf_obs[,c("x","y")]=baseline_crcvm2_trajectory_lf_obs[,c("x","y")]+noise
+
+baseline_crcvm2_trajectory_hf_obs=baseline_crcvm2_trajectory[seq(1,nrow(baseline_crcvm2_trajectory),by=30),c("time","x","y","ID")]
+noise=rmvn(length(baseline_crcvm2_trajectory_hf_obs$time),rep(0,2),diag(rep(0.05^2,2)))
+baseline_crcvm2_trajectory_hf_obs[,c("x","y")]=baseline_crcvm2_trajectory_hf_obs[,c("x","y")]+noise
+
+baseline_crcvm2_trajectory_lf_obs=add_covs_parallel(baseline_crcvm2_trajectory_lf_obs)
+baseline_crcvm2_trajectory_hf_obs=add_covs_parallel(baseline_crcvm2_trajectory_hf_obs)
 
@@
 
+plot_baseline_crcvm2_trajectory_obs=ggplot() +
+  geom_sf(data=shrunk_land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=baseline_crcvm2_trajectory_lf_obs, 
+            aes(x = x, y = y,color=ID),size = 0.2, 
+            lineend = "round",alpha=0.5) +     
+  geom_point(data=baseline_crcvm2_trajectory_lf_obs,
+             aes(x = x, y = y,color=ID),size =0.2 ) +
+  theme_minimal()+ scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
+
+plot_dataBE12=ggplot() +
+  geom_sf(data=shrunk_land$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=dataBE12, 
+            aes(x = x, y = y,color=ID),size = 0.2, 
+            lineend = "round",alpha=0.5) +     
+  geom_point(data=dataBE12,
+             aes(x = x, y = y,color=ID),size =0.2 ) +
+  theme_minimal()+ scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
+
+final_plot=ggarrange(plotlist=list(plot_dataBE12,plot_baseline_crcvm2_trajectory_obs),
+                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom",
+                     labels= c("Observed","Simulated"))
+
+ggsave("plot_baseline_crcvm2_trajectory.pdf", plot = final_plot,path="baseline_racvm2",width = 14, height = 8)
+```
+
+```{r}
+
+# Determine the common x-axis limits
+x_min <- min(baseline_crcvm2_trajectory_hf_obs$DistanceShore, 
+              dataBE12$DistanceShore_translated, na.rm = TRUE)
+x_max <- max(baseline_crcvm2_trajectory_hf_obs$DistanceShore, 
+              dataBE12$DistanceShore_translated, na.rm = TRUE)
+
+# Plot the two histograms with consistent x-axis scales
+Dshore_hist_crcvm2 <- ggplot(baseline_crcvm2_trajectory_hf_obs) +
+  geom_histogram(aes(x = DistanceShore), color = "red", fill = "white", bins = 30) +
+  theme_minimal() +
+  xlim(x_min, x_max) +
+  xlab("Distance to Shore") +
+  ggtitle("Simulated data") +theme(
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
+
+Dshore_hist_dataBE12 <- ggplot(dataBE12) +
+  geom_histogram(aes(x = DistanceShore_translated), color = "red", fill = "white", bins = 30) +
+  theme_minimal() +
+  xlim(x_min, x_max) +
+  xlab("Distance to Shore ") +
+  ggtitle("Observed data")+theme(
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 14),
+    legend.text = element_text(size = 18),
+    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
+  )
+
+# Combine the plots
+Dshore_histograms <- Dshore_hist_crcvm2 + Dshore_hist_dataBE12
+
+Dshore_histograms
+
+# Save the plot as a PDF
+ggsave("Dshore_histograms.pdf", plot = Dshore_histograms,path="baseline_racvm2", width = 10, height = 5)
+
+```
+
+```{r}
+
 #initial parameters
-par0 <- c(0,0,2,4,0)
-init_lambda=c(exp(3),exp(3),m1$sp)
+par0 <- c(0,0,1,1,0)
+init_lambda=c(1/0.5^2,1/0.5^2,m1$sp)
 
@@
 sigma_obs=0.05
-H=array(rep(sigma_obs^2*diag(2),length(baseline_crcvm2_trajectory_sub_obs$time)),
-        dim=c(2,2,length(baseline_crcvm2_trajectory_sub_obs$time)))
+H=array(rep(sigma_obs^2*diag(2),length(baseline_crcvm2_trajectory_hf_obs$time)),
+        dim=c(2,2,length(baseline_crcvm2_trajectory_hf_obs$time)))
 
@@
 # Fit baseline
-baseline_racvm2_check<- SDE$new(formulas = formulas,data = baseline_crcvm2_trajectory_sub_obs,type = "RACVM_SSM",
+baseline_racvm2_check<- SDE$new(formulas = formulas,
+                                data = baseline_crcvm2_trajectory_hf_obs[baseline_crcvm2_trajectory_hf_obs$ID %in% c("Asgeir","Frederik","Kyrri","Nemo"),],type = "RACVM_SSM",
                     response = c("x","y"),par0 = par0,other_data=list("H"=H),
@@
 ```
+
 ```{r}
-get_final_baseline_tab(baseline_racvm2_check,include_sigma_obs=FALSE)
+make_final_baseline_tab(baseline_racvm2_check,include_sigma_obs=FALSE)
 ```
@@
 
+We continue the analysis with the fitted baseline model \texttt{baseline\_racvm2}.
+
 # Fit response models
@@
 
-```{r}
+```{r fit response all data}
 
@@
 ```
-```{r}
+
+```{r plot estimates response all data}
 #plot parameters
@@
 
-```{r fit response1}
+```{r fit response splines}
 
@@
 
-```{r get results response1}
+```{r get results response splines}
 #plot parameters
@@
 
-```{r}
+```{r fit response log linear}
 
@@
 
-```{r}
+```{r plot estimates response log linear}
 
@@
 
-```{r display plots response3_sp}
+```{r display plots response log linear}
 
@@
 
-```{r results response_ctcrw_exp}
-
-get_final_response_tab<-function(sde,show=TRUE) {
+```{r get final estimates response_racvm_exp}
+
+make_final_response_tab<-function(sde,show=TRUE) {
 
@@
     table.placement="H")
-  }
-
-  return (xtab)
-
-  }
-
-```
-
-```{r,results='asis',echo=FALSE}
-final_tab_response_exp=get_final_response_tab(response_racvm_exp)
-
+
+  }
+
+make_final_response_tab(response_racvm_exp)
 ```
@@
 ```{r}
-get_recovery_distances_tab<-function(p) {
+make_recovery_distances_tab<-function(p) {
 
@@
 
-get_recovery_distances_tab(0.5)
-
-get_recovery_distances_tab(0.1)
-```
+make_recovery_distances_tab(0.5)
+
+make_recovery_distances_tab(0.1)
+```
+
+The main difference with the CVM without omega, that is without influence of the shore, is that we now get much wider confidence intervals for the recovery distances.
+However, the mean values are somehow similar.

### 2025-03-17 R/preprocessing/narwhals/data_preprocessing.Rmd
update preprocessing of borders and exploratory plots

@@
 
-## Import necessary packages
+## Imports
 
@@
 
-## Choose directory where the plots should be saved
+## Create plots directory
 
@@
 #add time since tagging column
-
 TimeFunction <- function(n, dt = 1){
@@
 #animated map with gganimate
-animate_map=milne_land_map + 
-  geom_line(dataprep[,c("ID","Long","Lat","time")], 
-            mapping=aes(Long, Lat, col = ID),size= 1,alpha=0.5) +
-  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+transition_reveal(time)
-
-animation=animate(animate_map,renderer = gifski_renderer())
+animate_map <- milne_land_map + 
+  geom_point(data = dataprep[,c("ID","Long","Lat","time")], 
+             mapping = aes(Long, Lat, shape = ID, col = ID), 
+             size = 4, alpha = 1) +
+  geom_path(data = dataprep[,c("ID","Long","Lat","time")], 
+            mapping = aes(Long, Lat, col = ID), 
+            size = 1, alpha = 0.5) +
+  scale_color_viridis_d(name = "ID",  # Give a common legend title
+                        labels = c("A1","A2","A3","A4","A5","A6")) + 
+  scale_shape_manual(name = "ID",  # Same name as color to merge legends
+                     values = c(0, 1, 2, 3, 4, 5),  
+                     labels = c("A1","A2","A3","A4","A5","A6")) +  
+  guides(color = "legend", shape = "legend") +  # Ensure single legend
+  transition_reveal(time) +
+  labs(subtitle = "Time: {round(frame_along,1)} hours")
+
+animate_map
+
+animation=animate(animate_map,nframes = 100, fps = 5,renderer = gifski_renderer())
 anim_save(path=plots_folder,filename="all_narwhals_tracks.gif", 
@@
 
-scoresby_sound_utm=st_transform(scoresby_sound_utm,crs= "+init=EPSG:32626 +units=km")
+scoresby_sound_utm<-st_transform(scoresby_sound_utm,crs= "+init=EPSG:32626 +units=km")
 ```
@@
 plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")
-```
-
-```{r display land , fig.cap="Land polygons"}
 ggplotly(plot_land)
@@
 ```
-
-## Points on land
 
@@
 ```{=tex}
-A lot of points are indeed still very close to land.
-We think that this means the shoreline is still not accurate.
-To solve the problem, we translate the polygon vertices 200m inward.
+A lot of points are indeed still very close to land. We think that this means the shoreline is still not accurate.
+To solve the problem, we translate the polygon vertices 100m, 200m and 300m inward. 
 We compute new distances to shore based on the translated polygon.
@@
 
+## Compute distances from shrunk land polygons
+
 ```{r get shrunk polygons}
 
-#get updated land polygons
-shrunk_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk_updated_scoresby_sound_utm.shp"))
-
-shrunk_scoresby_sound_utm <- st_transform(shrunk_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
+#get shrunk land polygons
+shrunk100_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk100_scoresby_sound_utm.shp"))
+shrunk100_scoresby_sound_utm <- st_transform(shrunk100_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
+
+shrunk200_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk200_scoresby_sound_utm.shp"))
+shrunk200_scoresby_sound_utm <- st_transform(shrunk200_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
+
+shrunk300_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk300_scoresby_sound_utm.shp"))
+shrunk300_scoresby_sound_utm <- st_transform(shrunk300_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")
 
@@
 ```{r compute updated translated distance to shore}
-res_shrunk_land=nearest_points(positions=narwhal_positions,border=shrunk_scoresby_sound_utm)
-
-cat("Percentage of points on land with distance from shrunk updated osm land polygons : \n",
-    length(which(res_shrunk_land$min_distances==0))/length(dataprep$time)*100,"\n")
-
-```
-
-```{r distribution of points close to the shore}
-
-df=data.frame(DistanceShore=res_shrunk_land$min_distances[res_shrunk_land$min_distances<1 & res_shrunk_land$min_distances>0  ])
-histo=ggplot(df, aes(x =DistanceShore)) +
-  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")
+res_shrunk100_land=nearest_points(positions=narwhal_positions,border=shrunk100_scoresby_sound_utm)
+res_shrunk200_land=nearest_points(positions=narwhal_positions,border=shrunk200_scoresby_sound_utm)
+res_shrunk300_land=nearest_points(positions=narwhal_positions,border=shrunk300_scoresby_sound_utm)
+
+cat("Percentage of points on land with distance from 100m shrunk land polygons : \n",
+    length(which(res_shrunk100_land$min_distances==0))/length(dataprep$time)*100,"\n")
+
+cat("Percentage of points on land with distance from 200m shrunk land polygons : \n",
+    length(which(res_shrunk200_land$min_distances==0))/length(dataprep$time)*100,"\n")
+
+cat("Percentage of points on land with distance from 300m shrunk land polygons : \n",
+    length(which(res_shrunk300_land$min_distances==0))/length(dataprep$time)*100,"\n")
+
+```
+
+```{r distribution of points close to the shore for shrunk polygons}
+
+df100=data.frame(DistanceShore=res_shrunk100_land$min_distances[res_shrunk100_land$min_distances<1.5])
+df200=data.frame(DistanceShore=res_shrunk200_land$min_distances[res_shrunk200_land$min_distances<1.5])
+df300=data.frame(DistanceShore=res_shrunk300_land$min_distances[res_shrunk300_land$min_distances<1.5])
+
+df100$Category="100m"
+df200$Category="200m"
+df300$Category="300m"
+
+df<- rbind(df100,df200,df300)
+histo=ggplot(df, aes(x =DistanceShore,fill=Category)) +
+  geom_histogram(bins=100,color="black")+xlab("Computed distance to shore (km)")
 
@@
 
-#keep points calculated from land polygons
-nearest_points=res_updated_land$nearest_points
-min_distances=res_updated_land$min_distances
+add_boundary_data<-function(
+    data,res,names=list(px="px",py="py",DistanceShore="DistanceShore",nx="nx",ny="ny"))
+
+  {
+
+  #data : dataframe to which wa add columns 
+  # res : list as returned by nearest points, with elements
+  # nearest points and min_distances
+  # names for the distance, nearest points coordinates and normal vectors
+
+  nearest_points=res$nearest_points
+  min_distances=res$min_distances
 
   #add columns px and py with coordinates of the nearest point on the shore line
-dataprep$px=nearest_points[,1]
-dataprep$py=nearest_points[,2]
+  data[[names$px]]=nearest_points[,1]
+  data[[names$py]]=nearest_points[,2]
 
   # add columns with difference between narwhal position and nearest point on the shoreline
-dataprep$nx=dataprep$x-dataprep$px
-dataprep$ny=dataprep$y-dataprep$py
+  data[[names$nx]]=data$x- data[[names$px]]
+  data[[names$ny]]=data$y-data[[names$py]]
 
   #replace distance to shore with distances calculated from land polygons (in km)
-dataprep$DistanceShore=min_distances
-
-#same for shrunk polygons
-nearest_points_shrunk=res_shrunk_land$nearest_points
-min_distances_shrunk=res_shrunk_land$min_distances
-
-#add columns px and py with coordinates of the nearest point on the shore line
-dataprep$px_translated=nearest_points_shrunk[,1]
-dataprep$py_translated=nearest_points_shrunk[,2]
-
-# add columns with difference between narwhal position and nearest point on the shoreline
-dataprep$nx_translated=dataprep$x-dataprep$px_translated
-dataprep$ny_translated=dataprep$y-dataprep$py_translated
-dataprep$DistanceShore_translated=min_distances_shrunk
+  data[[names$DistanceShore]]=min_distances
+
+  return(data)
+}
+
+dataprep<- add_boundary_data(dataprep,res_shrunk100_land,
+                  names=list(px="px100",py="py100",DistanceShore="DistanceShore100",nx="nx100",ny="ny100"))
+
+dataprep <- add_boundary_data(dataprep,res_shrunk200_land,
+                  names=list(px="px200",py="py200",DistanceShore="DistanceShore200",nx="nx200",ny="ny200"))
+
+dataprep <- add_boundary_data(dataprep,res_shrunk300_land,
+                  names=list(px="px300",py="py300",DistanceShore="DistanceShore300",nx="nx300",ny="ny300"))
+
 ```
@@
 #add Ishore covariate based on 1/distance in km
-dataprep$Ishore <- ExposureFunction(X = dataprep$DistanceShore)
-
-#add Ishore covariate based on 1/distance in km
-dataprep$Ishore_translated <- ExposureFunction(X = dataprep$DistanceShore_translated)
+dataprep$Ishore100 <- ExposureFunction(X = dataprep$DistanceShore100) 
+dataprep$Ishore200 <- ExposureFunction(X = dataprep$DistanceShore200) 
+dataprep$Ishore300 <- ExposureFunction(X = dataprep$DistanceShore300)
 
@@
 
-dataprep$AngleShore=rep(NA,n)
 dataprep$AngleVelocities=rep(NA,n)
-dataprep$theta=rep(NA,n)
-dataprep$theta_translated=rep(NA,n)
+dataprep$theta100=rep(NA,n)
+dataprep$theta200=rep(NA,n)
+dataprep$theta300=rep(NA,n)
 dataprep$VelocityPhase=rep(NA,n)
@@
   #matrix of normal vectors
-  normal=as.matrix(sub_data[2:n_sub,c("nx","ny")])
-  normal_translated=as.matrix(sub_data[2:n_sub,c("nx_translated","ny_translated")])
+  normal100=as.matrix(sub_data[2:n_sub,c("nx100","ny100")])
+  normal200=as.matrix(sub_data[2:n_sub,c("nx200","ny200")])
+  normal300=as.matrix(sub_data[2:n_sub,c("nx300","ny300")])
 
   #angle between velocity and normal vector
-  theta=signed_angle(vexp_df,normal)
-  theta_translated=signed_angle(vexp_df,normal_translated)
+  theta100=signed_angle(vexp_df,normal100)
+  theta200=signed_angle(vexp_df,normal200)
+  theta300=signed_angle(vexp_df,normal300)
 
@@
 
-  #deviation from trajectory parallel to coastline
-  phi=ifelse(theta > 0, theta - pi/2, theta + pi/2)
-
   #adjust lengths by adding NA when necessary
   angle_vel=c(NA,angle_vel,NA)
-  phi=c(NA,phi)
-  theta=c(NA,theta)
-  theta_translated=c(NA,theta_translated)
+  theta100=c(NA,theta100)
+  theta200=c(NA,theta200)
+  theta300=c(NA,theta300)
   vel_phase=c(NA,vel_phase,NA)
 
-  dataprep[sub_ind,"AngleShore"]=phi
   dataprep[sub_ind,"AngleVelocities"]=angle_vel
-  dataprep[sub_ind,"theta"]=-theta
-  dataprep[sub_ind,"theta_translated"]=-theta_translated
+  dataprep[sub_ind,"theta100"]=-theta100
+  dataprep[sub_ind,"theta200"]=-theta200
+  dataprep[sub_ind,"theta300"]=-theta300
   dataprep[sub_ind,"VelocityPhase"]=vel_phase
@@
 
-plot_AngleShore=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleShore,col=ID))+
-  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
-  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")+
-  theme_minimal()
-
 plot_AngleVelocities=ggplot()+geom_line(data=dataprep,aes(time,AngleVelocities,col=ID))+
@@
 
-plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,theta,col=ID))+
+plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,theta100,col=ID))+
   geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
@@
 ggsave(path=plots_folder,"plot_horizontal_speed.png", plot = plot_speed)
-ggsave(path=plots_folder,"plot_AngleShore.png", plot = plot_AngleShore)
 ggsave(path=plots_folder,"plot_AngleVelocities.png", plot = plot_AngleVelocities)
@@
 
-```{r display plot_AngleShore , fig.cap="",warning=FALSE}
-plot_AngleShore
-plot_AngleVelocities
-```
-
 ```{r display plot_AngleVelocities , fig.cap="",warning=FALSE}
@@
 ```{r histogram deviation angles}
-
-AngleShore_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleShore,col=ID)) +
-  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+
-  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")+ theme_minimal()
 
@@
 
-AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = theta,col=ID)) +
+AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = theta100,col=ID)) +
   geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
@@
 
-ggsave(path=plots_folder,"AngleShore_histo.png", plot = AngleShore_histo)
 ggsave(path=plots_folder,"AngleVelocities.png", plot = AngleVelocities_histo)
@@
 
-```{r display AngleShore_histo, fig.cap="",warning=FALSE}
-AngleShore_histo
-```
-
 ```{r display AngleVelocities_histo, fig.cap="",warning=FALSE}
@@
 
-plot_AngleShoreDistance=ggplot()+geom_point(data=dataprep,aes(DistanceShore,AngleShore,col=ID))+
-  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation from coastline")
-
-ggsave(path=plots_folder,"plot_AngleShore_DistanceShore.png", plot = plot_AngleShoreDistance)
-
 plot_AngleVelocitiesDistance=ggplot()+geom_point(data=dataprep,aes(DistanceShore,AngleVelocities,col=ID))+
@@
 ```{r deviation angles and velocity as a smooth function of DistanceShore}
-
-# Plot smoothed AngleSHore
-plot_AngleShoreDistanceSmoothed <-  ggplot()+ geom_smooth(data=dataprep,aes(DistanceShore,abs(AngleShore),col=ID))+
-  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Distance to shore (km)")+ylab("Empirical deviation from coastline")
 
@@
 
-```{r display plot_AngleShoreDistanceSmoothed, fig.cap=""}
-plot_AngleShoreDistanceSmoothed
-```
-
 ```{r display plot_AngleVelocitiesDistanceSmoothed, fig.cap=""}
@@
 ```{r plot empirical velocities and deviation angles without points on land}
-# Plot smoothed AngleSHore
-plot_AngleShoreDistanceSmoothed <-  ggplot()+
-  geom_smooth(data=dataprep[dataprep$DistanceShore>0,],aes(DistanceShore,abs(AngleShore),col=ID))+
-  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
-  scale_color_viridis_d(guide="none")+
-  theme(legend.position="none")+xlab("Distance to shore (km)")+
-  ylab("Empirical deviation from coastline")
 
@@
 ```{r distance shore histogram}
-Dshore_histo=ggplot(dataprep, aes(x = DistanceShore)) +
+Dshore100_histo=ggplot(dataprep, aes(x = DistanceShore100)) +
   geom_histogram(fill="white",color="black",bins=50)+
@@
 
-ggsave(path=plots_folder,"Dshore_histo.png", plot = Dshore_histo)
+Dshore200_histo=ggplot(dataprep, aes(x = DistanceShore200)) +
+  geom_histogram(fill="white",color="black",bins=50)+
+  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+  theme(legend.position="none")+xlab("Distance to shore (km)")
+
+Dshore300_histo=ggplot(dataprep, aes(x = DistanceShore300)) +
+  geom_histogram(fill="white",color="black",bins=50)+
+  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+  theme(legend.position="none")+xlab("Distance to shore (km)")
+
+ggsave(path=plots_folder,"Dshore100_histo.png", plot = Dshore100_histo)
+ggsave(path=plots_folder,"Dshore200_histo.png", plot = Dshore200_histo)
+ggsave(path=plots_folder,"Dshore300_histo.png", plot = Dshore300_histo)
 ```
@@
 ```{r display Dshore_histo, fig.cap="Histograms of distance to shore"}
-Dshore_histo
+Dshore100_histo
+Dshore200_histo
+Dshore300_histo
 ```
@@
 Dship_histo=ggplot(dataprep, aes(x = Dist_Ship)) +
-  geom_histogram(fill="white",col="black",bins=30)+
+  geom_histogram(fill="white",col="black",bins=50)+
   facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
@@
 df=cropped_data
-df[df$DistanceShore<D_low,"Ishore"]=1/D_low
-df[df$DistanceShore>D_up,"Ishore"]=0
-
-pIshore=ggplot(df,aes(time,Ishore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
+df[df$DistanceShore<D_low,"Ishore100"]=1/D_low
+df[df$DistanceShore>D_up,"Ishore100"]=0
+
+pIshore=ggplot(df,aes(time,Ishore100,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
   theme(legend.position="none")+xlab("Time (h)")+ylab("Inverse distance to shore")
 
-ggsave(path=plots_folder,"realIshore_through_time.png", plot=pIshore)
+ggsave(path=plots_folder,"realIshore100_through_time.png", plot=pIshore)
 ```
@@
 ```
+
+# Animate trajectories with Ship Exposure
+
+```{r simulate response trajectories with land reflection,cache=TRUE,message=FALSE}
+
+# Plot both in a single plot with facets
+ggplot(dataprep) +geom_sf(data=shrunk100_scoresby_sound_utm$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_point(aes(x = x, y = y, color = ID), size = 0.6) + 
+  ggtitle("Trajectories during ship exposure") +
+  theme_minimal() +
+  labs(color = "Narwhal ID")
+
+```
+
+```{r animate response trajectories with land reflection, results=FALSE,eval=interactive()}
+
+# Basic ggplot setup
+plot_trajectories_land <- ggplot() +
+  geom_sf(data=shrunk100_scoresby_sound_utm$geometry,fill="grey")+
+  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
+  geom_path(data=dataprep, 
+            aes(x = x, y = y, color =ExpShip,group=ID),size = 1, 
+            lineend = "round",alpha=0.15) +     
+  geom_point(data=dataprep,
+             aes(x = x, y = y, color =ExpShip,group=ID,shape=ID),size =4 ) +  
+  scale_color_gradient(low = "blue", high = "red") +
+  labs(title = "Narwhal trajectories with ship exposure", 
+       subtitle = "Time : {frame_along} h", 
+       x = "X Position", 
+       y = "Y Position") +
+  theme_minimal()
+
+anim <- plot_trajectories_land + 
+  transition_reveal(along = Posixct) + 
+  labs(subtitle = "Time: {round(frame_along,1)} hours")
+
+# Save the animation
+animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
+        renderer = gifski_renderer("trajectories_ExpShip.gif"))
+
+```

### 2025-03-17 R/simulation_study/examples/illustrative_trajectories.Rmd
try some new smooth functions omega

@@
 #get functions to simulate trajectories
-source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R")))
+source(file.path(here("R","simulation_study","CVM_functions.R")))
 
@@
 
-This type of behavior can be obtained with hyperbolic tangent function.
+This type of behavior can be obtained with several function.
+We start by using the hyperbolic tangent function.
 
@@
 
-fomega_tanh=function(cov_data,D0=0.5,omega0=60*pi/2,lambda=2,kappa=0.2) {
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$ExpShore
-  }
+fomega_tanh=function(cov_data,D0=0.5,omega0=30*pi/2,lambda=2,kappa=0.2) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
   coeff=exp(-kappa*(Dshore/D0)^2)
@@
 }
-
 ```
@@
 
-#grid 
-Dshore=seq(from=0.05,to=4,length.out=30)
-theta=seq(from=-pi+0.1,to=pi-0.1,length.out=30)
+get_omega_plot<-function(fomega,Dshore,theta) {
+
+  #grid of values
   grid<- as.data.frame(expand.grid(Dshore,theta))
-colnames(grid)=c("DistanceShore","theta")
+  colnames(grid)=c("BoundaryDistance","BoundaryAngle")
 
   #smooth omega
-omega_tanh=matrix(fomega_tanh(grid),30,30)
+  npoints=length(Dshore)
+  omega=matrix(fomega(grid),npoints,npoints)
 
   # dynamic plots for omega 
-
-plot_omega_tanh <- plot_ly(type = "surface",
-                           contours = list(z = list(show = TRUE, start = min(omega_tanh), end = max(omega_tanh), size = 5,color="black")),
-                           x = ~theta,y = ~Dshore,z=~omega_tanh,
+  plot_omega <- plot_ly(type = "surface",
+                           contours = list(z = list(show = TRUE, start = min(omega), end = max(omega), size = 5,color="black")),
+                             x = ~theta,y = ~Dshore,z=~omega,
                            colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
@@
 
-plot_omega_tanh
+  return (plot_omega)
+}
+
+Dshore=seq(from=0.05,to=4,length.out=30)
+theta=seq(from=-pi+0.1,to=pi-0.1,length.out=30)
+
+get_omega_plot(fomega_tanh,Dshore,theta)
+
 ```
@@
 
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$Ishore
-  }
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
 
@@
 
-```
-
-```{r plot omega sqrt}
-#smooth omega
-omega_sqrt=matrix(fomega_sqrt(grid),30,30)
-
-# dynamic plots for omega
-
-plot_omega_sqrt <- plot_ly(type = "surface",
-                           contours = list(z = list(show = TRUE, start = min(omega_sqrt), end = max(omega_sqrt), size = 5,color="black")),
-                           x = ~theta,y = ~Dshore,z=~omega_sqrt,
-                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-  layout(title=paste("Omega"),
-         scene=list(xaxis = list(title = "Theta",showgrid = F),
-                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
-
-plot_omega_sqrt
-```
-
-These smooth functions $\omega$ can produce a SDE model that is constrained within a restricted domain of $\R^2$.
-The trajectories avoid the shoreline, but are not attracted to it.
-To model a tendecy to follow the shoreline, we need to modify a bit the smooth function to get a rotation of the velocity toward the shore when it is on the verge of leaving it.
-We can do this in several ways.
-First, we use a sin function.
+get_omega_plot(fomega_sqrt,Dshore,theta)
+
+```
+
+These smooth functions $\omega$ will produce a SDE model that is constrained within a restricted domain of $\R^2$.
+The trajectories avoid the shoreline, but are not attracted to it. To model a tendency to follow the shoreline, we need to modify 
+a bit the smooth function to get a rotation of the velocity toward the shore when it is on the verge of leaving it. 
+We can do this in several ways. First, we use a sin function.
 
@@
 # Define the extended function f(theta)
-fomega_sin<- function(cov_data, a =4*pi,b=80*pi,D0=0.8) {
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$Ishore
-  }
+fomega_sin<- function(cov_data, a =4*pi,b=10*pi,D0=0.8) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
   ifelse(abs(theta) < pi/2, 
@@
 
-#smooth omega
-omega_sin=matrix(fomega_sin(grid),30,30)
-
-# dynamic plots for omega
-
-plot_omega_sin <- plot_ly(type = "surface",
-                           contours = list(z = list(show = TRUE, start = min(omega_sin), end = max(omega_sin), size = 5,color="black")),
-                           x = ~theta,y = ~Dshore,z=~omega_sin,
-                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-  layout(title=paste("Omega"),
-         scene=list(xaxis = list(title = "Theta",showgrid = F),
-                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
-
-plot_omega_sin
-
-```
-
-Now we use a cubic polynomial with an additional gaussian kernel.
-
-```{r}
-fomega_cubic=function(cov_data,a=1,b=5,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.3){
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
-  if (is.null(Dshore)){
-    Dshore=1/cov_data$Ishore
-  }
-
-  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
-      b*(exp(-1/2*(theta+pi/2/sqrt(3))^2/sigma_theta^2)-
-           exp(-1/2*(theta-pi/2/sqrt(3))^2/(sigma_theta)^2))*exp(-1/2*(Dshore-D1)^2/sigma_D^2)
-}
-
-#smooth omega
-omega_cubic=matrix(fomega_cubic(grid),30,30)
-
-# dynamic plots for omega
-
-plot_omega_cubic <- plot_ly(type = "surface",
-                           contours = list(z = list(show = TRUE, start = min(omega_cubic), end = max(omega_cubic), size = 5,color="black")),
-                           x = ~theta,y = ~Dshore,z=~omega_cubic,
-                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-  layout(title=paste("Omega"),
-         scene=list(xaxis = list(title = "Theta",showgrid = F),
-                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
-
-plot_omega_cubic
+get_omega_plot(fomega_sin,Dshore,theta)
+
+```
+
+In these functions, there is no mid range effect of the shore on the drift of the process, which means
+that when the process is far enough from the shore, it cannot be attracted back.
+To resolve this, we add a gaussian kernel in the functions.
+
+```{r}
+
+fomega_gaussian= function(cov_data,b=1,D1=1.5,theta1=pi/2/sqrt(3),sigma_D=1,sigma_theta=0.6) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
+  b*(exp(-1/2*(theta+theta1)^2/sigma_theta^2)-
+           exp(-1/2*(theta-theta1)^2/(sigma_theta)^2))*exp(-1/2*(Dshore-D1)^2/sigma_D^2)
+}
+
+get_omega_plot(fomega_gaussian,Dshore,theta)
+
+```
+
+We define a product of sine and cosine with an additional gaussian attractive term
+
+```{r define and plot omega sin}
+# Define the extended function f(theta)
+fomega_sin_cos<- function(cov_data,omega_0=10,D0=0.8,p=1,b=1,D1=1,theta1=pi/4,sigma_D=1,sigma_theta=0.5) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
+
+  attraction <- fomega_gaussian(cov_data,b,D1,theta1,sigma_D,sigma_theta)
+
+  if (p%%2==0) {
+    repulsion <- -omega_0*sign(theta)*sin(theta/2)^p*cos(theta)*D0/Dshore*exp(-Dshore/D0)
+  }
+  else {
+    repulsion <- -omega_0*sin(theta/2)^p*cos(theta)*D0/Dshore*exp(-Dshore/D0)
+  }
+  attraction+repulsion
+}
+
+get_omega_plot(fomega_sin_cos,Dshore,theta)
+
+```
+
+Now we use a polynomial with an additional gaussian kernel to model attraction toward the shore.
+
+```{r}
+fomega_poly=function(cov_data,a=0.2,b=5,D0=0.3,D1=1.5,sigma_D=1,sigma_theta=pi/8,
+                     theta1=pi/2/sqrt(3),p=3) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
+
+  if (p%%2==0) {
+    repulsion<- a*sign(theta)*theta^p*(theta-pi/2)*(theta+pi/2)*D0/Dshore*exp(-Dshore/D0)
+  }
+  else {
+    repulsion<- a*theta^p*(theta-pi/2)*(theta+pi/2)*D0/Dshore*exp(-Dshore/D0)
+  }
+
+  attraction<- fomega_gaussian(cov_data,b,D1,theta1,sigma_D,sigma_theta)
+
+  return (repulsion+attraction)
+}
+
+get_omega_plot(fomega_poly,Dshore,theta)
+```
+```{r}
+fomega_quadratic=function(cov_data,a=0.2,D0=0.3,D1=1.5) {
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
+
+    return (a*(theta-pi/2)*(theta+pi/2)*D0/Dshore*exp(-Dshore/D0))
+}
+
+get_omega_plot(fomega_quadratic,Dshore,theta)
 ```
@@
 fomega_minimal=function(cov_data,nu=4,D0=0.5,D1=1.5,a=1,b=10,sigma_D=0.5,sigma_theta=pi/8) {
-  Dshore=cov_data$DistanceShore
-  theta=cov_data$theta
+  Dshore=cov_data$BoundaryDistance
+  theta=cov_data$BoundaryAngle
   ifelse(abs(theta)>pi/2,a*sign(theta)*(pi/2-abs(theta))*nu*cos(theta)/Dshore*exp(-Dshore/D0),-sign(theta)*b*
@@
 
-#smooth omega
-omega_minimal=matrix(fomega_minimal(grid),30,30)
-
-# dynamic plots for omega
-
-plot_omega_minimal <- plot_ly(type = "surface",
-                           contours = list(z = list(show = TRUE, start = min(omega_minimal), end = max(omega_minimal), size = 5,color="black")),
-                           x = ~theta,y = ~Dshore,z=~omega_minimal,
-                           colors=colorRamp(c("blue", "lightblue", "chartreuse3", "yellow", "red")))%>%
-  layout(title=paste("Omega"),
-         scene=list(xaxis = list(title = "Theta",showgrid = F),
-                    yaxis = list(title = "Distance to shore",showgrid = F),zaxis = list(title = "omega")))
-
-plot_omega_minimal
-```
-
-We can approximate these surfaces using tensor splines
-
-```{r approximate tanh omega with splines}
-# Splines smooth for omega -------------------- 
-n <- 100000
-SP_DF=5
-
-#sample theta and DistanceShore points in the domain
-theta <- runif(n,-pi,pi);
-DistanceShore <- runif(n,0.2,8);
-samples=data.frame(theta=theta,DistanceShore=DistanceShore)
-
-#define grid of values
-theta_v <- seq(-pi,pi,length=30)
-Dshore_v<- seq(0.2,8,length=30)
-pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))
-
-#points on the surface perturbed by gaussian noise
-f <- fomega_tanh(samples)
-y <- f+0.1*rnorm(n)
-
-#fit with bivariate splines te
-m1 <- gam(y~te(theta,DistanceShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
-                                                       DistanceShore=seq(0.2,8,len=SP_DF)))
-
-ExpShore=1/DistanceShore
-
-#fit with bivariate splines te
-m2 <- gam(y~te(theta,ExpShore,k=SP_DF),knots=list(theta=seq(-pi,pi,len=SP_DF),
-                                                  ExpShore=seq(1/8,1/0.2,len=SP_DF)))
-
-fomega_tanh_splines=function(cov_data) {
-
-  if (is.null(cov_data$ExpShore)){
-    omega=predict(m1,newdata=cov_data)
-  }
-  else if (is.null(cov_data$DistanceShore)){
-    omega=predict(m2,newdata=cov_data)
-  }
-
-  return(as.numeric(omega))
-}
-
-```
-
-```{r plot tanh parametric and spline omega}
+get_omega_plot(fomega_minimal,Dshore,theta)
+```
+
+We create and save the perpective plot for the paper.
+
+```{r persp plot cubic parametric omega}
+
+#Define function omega with adequate parameters
+fomega_cubic<-function(cov_data) {
+  fomega_poly(cov_data,a=2,b=5,D0=0.3,D1=0.8,sigma_D=0.3,sigma_theta=pi/8,
+                     theta1=pi/2/sqrt(3),p=1)
+}
 
@@
 # Grid 
-Dshore = seq(from = 0, to = 2, length.out = 30)
+Dshore = seq(from = 0.05, to = 2, length.out = 30)
 theta = seq(from = -pi + 0.1, to = pi - 0.1, length.out = 30)
 grid <- as.data.frame(expand.grid(Dshore, theta))
-colnames(grid) = c("DistanceShore", "theta")
+colnames(grid) = c("BoundaryDistance", "BoundaryAngle")
 
 # Smooth omega
-z = matrix(fomega_cubic_advanced(grid), 30, 30)
+z = matrix(fomega_cubic(grid), 30, 30)
 
@@
 
+simulate_paralell<-function(ftau,fnu,fomega,v0=v0,x0=x0,times,border,n_cores=6,seed=42) {
+
+  #start cluster
+  cl <- makeCluster(n_cores) 
+  registerDoParallel(cl)
+
+  N<-nrow(x0)
+
+  #paralell loop
+  data=foreach (i=1:N,.combine='rbind',.export=c("sim_CRCVM","fomega_tanh","fomega_poly",
+                "fomega_sin","fomega_sin_cos","fomega_sqrt","fomega_gaussian","fomega_minimal","fomega_quadratic"),
+                .packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+
+    set.seed((seed-1)*N+i)
+
+    res=sim_CRCVM(ftau=ftau,fomega=fomega,fnu=fnu,v0=v0,x0=x0[i,],
+                      times=times,land=border,verbose=FALSE)
+
+    data_sim=res$sim
+    data_sim$ID=factor(rep(i,length(data_sim$y1)))
+    data_shore=res$boundary[,c("p1","p2")]
+    cbind(data_sim,data_shore)
+  }
+
+  #stop cluster
+  stopCluster(cl)
+
+  return (data)
+}
+
+data_rect_tanh<-simulate_paralell(ftau_constant,fnu_constant,fomega_tanh,v0,x0,times,rect_border)
+
+```
+
+```{r plot rect_tanh, message=FALSE}
+
+get_trajectory_plot<-function(data,border,by=60) {
+
+  #subsample to 10 minutes
+  data_sub=data[seq(1,nrow(data),by=by),]
+
+  #create plot
+  plot<- ggplot() +
+    geom_sf(data = border$geometry, fill = "grey", inherit.aes = FALSE) +
+    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
+    geom_path(data = data_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
+    geom_point(data = data_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
+    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
+    theme_minimal() +
+    xlab("x") + ylab("y") +
+    theme(legend.position = "none")
+
+  return (plot)
+}
+
+get_trajectory_plot(data_rect_tanh,rect_border)
+```
+
+The process is avoiding land by rotating when it approaches the boundary. 
+Now, to have trajectories that tends to remain close to the boundary and move along it we can use the function \texttt{fomega_poly} with 
+a non zero attractive gaussian term instead.
+
+```{r simulate with omega cubic attractive in rectangle,cache=TRUE}
+
+fomega_cubic_rect=function(cov_data) {
+
+  fomega_poly(cov_data,a=1,b=5,D0=0.5,D1=1.5,sigma_D=0.5,sigma_theta=pi/8,p=1)
+
+}
+
+data_rect_cubic<-simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_rect,v0,x0,times,rect_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_rect_cubic,rect_border)
+```
+
+We remove the attractive term in the function by taking parameter $b=0$.
+
+```{r simulate with omega cubic repulsive in rectangle,cache=TRUE,message=FALSE}
+
+fomega_cubic_repulsive_rect=function(cov_data) {
+
+  fomega_poly(cov_data,a=1,b=0,D0=0.5,D1=1.5,sigma_D=0.5,sigma_theta=pi/8,p=1)
+
+}
+
+data_rect_cubic_repulsive<-simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_repulsive_rect,
+                                             v0,x0,times,rect_border)
+
+```
+
+```{r}
+get_trajectory_plot(data_rect_cubic_repulsive,rect_border)
+```
+
+We can also increase the parameter $p$ in the polynomial omega function to get a steeper function
+at the extremities.
+
+```{r simulate with omega cubic attractive in rectangle,cache=TRUE,message=FALSE}
+
+fomega_poly_rect=function(cov_data) {
+
+  fomega_poly(cov_data,a=1,b=0,D0=0.5,D1=1.5,sigma_D=0.5,sigma_theta=pi/8,p=2)
+
+}
+
+data_rect_poly<-simulate_paralell(ftau_constant,fnu_constant,fomega_poly_rect,v0,x0,times,rect_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_rect_poly,rect_border)
+```
+
+We now try our \texttt{fome_sin_cos} function defined with a product of sin and cosine.
+
+```{r simulate with omega cubic attractive in rectangle,cache=TRUE,message=FALSE}
+
+fomega_sin_cos_rect=function(cov_data) {
+
+  fomega_sin_cos(cov_data,omega_0=10,D0=0.8,p=1,b=0,D1=1,theta1=pi/4,sigma_D=1,sigma_theta=0.5)
+
+}
+
+data_rect_sin_cos<-simulate_paralell(ftau_constant,fnu_constant,fomega_sin_cos_rect,v0,x0,times,rect_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_rect_sin_cos,rect_border)
+```
+
+```{r simulate with omega minimal in rectangle,cache=TRUE,message=FALSE}
+
+fomega_minimal_rect=function(cov_data) {
+
+  fomega_minimal(cov_data,nu=4,D0=0.5,D1=1.5,a=3,b=10,sigma_D=0.5,sigma_theta=pi/8)
+}
+
+data_rect_minimal<-simulate_paralell(ftau_constant,fnu_constant,fomega_minimal_rect,v0,x0,times,rect_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_rect_minimal,rect_border)
+```
+
+Until now, we kept the parameter $\tau$ constant.
+We could emphasize the tendency to move along the boundary by using a smooth $\tau$ given by the functions \texttt{ftau_tanh} or \texttt{ftau_gaussian}.
+
+Finally, we compare this to Hanks simulation technique for reflected SDE.
+
+```{r simulate hanks reflected in rectangle,cache=TRUE}
+
 # Generate samples ---------------
@@
 
-data_sim_rect_tanh=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+data_rect_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
 
@@
 
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_tanh,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=rect_border,verbose=FALSE)
-
-  data_sim=res$sim
+  data_sim=sim_HANKSreflected(mu1=0,mu2=0,beta=0.5,sigma=8/sqrt(pi*2),x0[i,],v0,dt,TMAX,land=rect_border)
   data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
+  data_sim
 }
@@
 
-#subsample to 10 minutes
-data_sim_rect_tanh_sub=data_sim_rect_tanh[seq(1,length(data_sim_rect_tanh$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-The process is avoiding land by rotating when it approaches the boundary. 
-Now, to have trajectories that tends to remain close to the boundary and move along it we can use the function \texttt{fomega_cubic} instead.
-
-```{r simulate with omega cubic attractive in rectangle,cache=TRUE}
-
-fomega_cubic_rect_attractive=function(cov_data) {
-
-  fomega_cubic(cov_data,a=3,D0=0.5,D1=1.5,sigma_theta=pi/8,sigma_D=0.5,b=5)
-
-}
+get_trajectory_plot(data_rect_hanks,rect_border)
+```
+
+# Trajectories in a circular domain.
+
+The values for the parameter of the smooth function $\omega$ and $\tau$ that produce constrained trajectories in a domain may not work for another domain.
+
+```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}
+
+#change parameter values 
+fomega_cubic_attractive_circ=function(cov_data) {
+
+   fomega_poly(cov_data,a=30,D0=0.5,D1=3,sigma_theta=pi/10,sigma_D=1,b=5,p=1,theta1=pi/2)
+}
+
+data_circ_cubic_attractive <-simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_attractive_circ,v0,x0,times,circ_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_circ_cubic_attractive,circ_border)
+```
+
+```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}
+
+#change parameter values 
+fomega_cubic_repulsive_circ=function(cov_data) {
+
+   fomega_poly(cov_data,a=3,D0=0.5,D1=1.5,sigma_theta=pi/10,sigma_D=0.5,b=0,theta=pi/2,p=1)
+}
+
+data_circ_cubic_repulsive <-simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_repulsive_circ,v0,x0,times,circ_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_circ_cubic_repulsive,circ_border)
+```
+
+Finally, we compare it to Hanks simulation technique for reflected SDE.
+
+```{r simulate hanks reflected in rectangle,cache=TRUE}
 
@@
 
-data_sim_rect_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=rect_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_rect_cubic_attractive_sub=data_sim_rect_cubic_attractive[seq(1,length(data_sim_rect_cubic_attractive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-We remove the attractive term in the function by taking parameter $b=0$.
-
-```{r simulate with omega cubic repulsive in rectangle,cache=TRUE}
-
-fomega_cubic_rect_repulsive=function(cov_data) {
-
-  fomega_cubic(cov_data,a=2,D0=0.5,D1=1.5,sigma_theta=pi/10,sigma_D=0.5,b=0)
-
-}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_rect_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_rect_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=rect_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_rect_cubic_repulsive_sub=data_sim_rect_cubic_repulsive[seq(1,length(data_sim_rect_cubic_repulsive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-```{r simulate with omega minimal in rectangle,cache=TRUE}
-
-fomega_minimal_rect_attractive=function(cov_data) {
-
-  fomega_minimal(cov_data,nu=4,D0=0.5,D1=1.5,a=3,b=10,sigma_D=0.5,sigma_theta=pi/8)
-}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_rect_minimal_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_minimal_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=rect_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_rect_minimal_attractive_sub=data_sim_rect_minimal_attractive[seq(1,length(data_sim_rect_minimal_attractive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_minimal_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_minimal_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-Until now, we kept the parameter $\tau$ constant.
-We can emphasize the tendency to move along the boundary by using a smooth $\tau$ given by the functions \texttt{ftau_tanh} or \texttt{ftau_gaussian}.
-Here, we plot trajectories for the SDE model where the smoth $\omega$ parameter is given by \texttt{fomega_cubic} and the smooth $\tau$ parameter is given by \texttt{ftau\_gaussian}.
-
-```{r simulate with omega cubic tau gaussian in rectangle,cache=TRUE}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_rect_cubic_gaussian_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_gaussian,fomega=fomega_cubic_rect_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=rect_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-#subsample to 10 minutes
-data_sim_rect_cubic_gaussian_attractive_sub=data_sim_rect_cubic_gaussian_attractive[seq(1,length(data_sim_rect_cubic_gaussian_attractive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_cubic_gaussian_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_cubic_gaussian_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-Finally, we compare this to Hanks simulation technique for reflected SDE.
-
-```{r simulate hanks reflected in rectangle,cache=TRUE}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_rect_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  data_sim=sim_HANKSreflected(mu1=0,mu2=0,beta=0.5,sigma=8/sqrt(pi*2),x0[i,],v0,dt,TMAX,land=rect_border)
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_sim
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-#subsample to 10 minutes
-data_sim_rect_hanks_sub=data_sim_rect_hanks[seq(1,length(data_sim_rect_hanks$time),by=60),]
-
-ggplot() +
-    geom_sf(data = rect_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_rect_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_rect_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-# Trajectories in a circular domain.
-
-The values for the parameter of the smooth function $\omega$ and $\tau$ that produce constrained trajectories in a domain may not work for another domain.
-
-```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}
-
-#change parameter values 
-fomega_cubic_circ_attractive=function(cov_data) {
-
-   fomega_cubic(cov_data,a=30,D0=0.5,D1=3,sigma_theta=pi/10,sigma_D=1,b=5)
-}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_circ_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_circ_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=circ_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_circ_cubic_attractive_sub=data_sim_circ_cubic_attractive[seq(1,length(data_sim_circ_cubic_attractive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_circ_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_circ_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-```{r simulate with omega cubic tau gaussian in circle,cache=TRUE}
-
-#change parameter values 
-fomega_cubic_circ_repulsive=function(cov_data) {
-
-   fomega_cubic(cov_data,a=30,D0=0.5,D1=1.5,sigma_theta=pi/10,sigma_D=0.5,b=0)
-}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_circ_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_circ_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=circ_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_circ_cubic_repulsive_sub=data_sim_circ_cubic_repulsive[seq(1,length(data_sim_circ_cubic_repulsive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_circ_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_circ_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-Finally, we compare it to Hanks simulation technique for reflected SDE.
-
-```{r simulate hanks reflected in rectangle,cache=TRUE}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_circ_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+data_circ_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
 
@@
 ```{r}
-#subsample to 10 minutes
-data_sim_circ_hanks_sub=data_sim_circ_hanks[seq(1,length(data_sim_circ_hanks$time),by=60),]
-
-ggplot() +
-    geom_sf(data = circ_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_circ_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_circ_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
+get_trajectory_plot(data_circ_hanks,circ_border)
+```
+
 # Trajectories in a circular domain with islands
@@
 
-set.seed(42)
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_circ_island_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_gaussian_circ_island,fomega=fomega_cubic_circ_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=circ_island_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-data_sim_circ_island_cubic_attractive_sub=data_sim_circ_island_cubic_attractive[seq(1,length(data_sim_circ_island_cubic_attractive$y1),by=10),]
-
-ggplot() +
-    geom_sf(data = circ_island_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_circ_island_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_circ_island_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
+data_circ_island_cubic_attractive=simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_attractive_circ,v0,x0,times,circ_island_border)
+
+```
+
+```{r}
+
+get_trajectory_plot(data_circ_island_cubic_attractive,circ_island_border)
 ```
@@
 ```{r simulate with omega cubic and tau gaussian in circle with islands,cache=TRUE}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_circ_island_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_gaussian_circ_island,fomega=fomega_cubic_circ_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=circ_island_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-data_sim_circ_island_cubic_repulsive_sub=data_sim_circ_island_cubic_repulsive[seq(1,length(data_sim_circ_island_cubic_repulsive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = circ_island_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_circ_island_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_circ_island_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
+data_circ_island_cubic_repulsive<- simulate_paralell(ftau_constant,fnu_constant,fomega_cubic_repulsive_circ,v0,x0,times,circ_island_border)
+
+```
+
+```{r}
+get_trajectory_plot(data_circ_island_cubic_repulsive,circ_island_border)
 ```
@@
 
-```{r simulate with omega tanh in fjords,cache=TRUE}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_fjords_tanh=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_tanh,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=land_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_fjords_tanh_sub=data_sim_fjords_tanh[seq(1,length(data_sim_fjords_tanh$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_fjords_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_fjords_tanh_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none")
-
-```
-
-```{r simulate with omega cubic and tau gaussian in fjords,cache=TRUE}
-
-#set parameters values
-fomega_cubic_fjords_attractive<- function(cov_data){
- fomega_cubic(cov_data,a=1,D0=0.3,D1=0.8,sigma_theta=pi/8,sigma_D=0.5,b=5)
-}
-
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) 
-registerDoParallel(cl)
-
-data_sim_fjords_cubic_attractive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_fjords_attractive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=land_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-#subsample to 10 minutes
-data_sim_fjords_cubic_attractive_sub=data_sim_fjords_cubic_attractive[seq(1,length(data_sim_fjords_cubic_attractive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_fjords_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_fjords_cubic_attractive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
-
-```{r animate trajectory sim_fjords_cubic_gaussian_sub,eval=FALSE}
+Now we simulate trajectories in the fjords.
+
+```{r simulate with omega cubic and tau gaussian in circle with islands,cache=TRUE}
+data_fjords_tanh<- simulate_paralell(ftau_constant,fnu_constant,fomega_tanh,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_tanh,land_border)
+```
+
+```{r define function to get animated trajectories,eval=FALSE}
+
+animate_trajectory_plot<- function(data,border,by=60,anim_name="sim.gif") {
+
+  #subsample 
+  data_sub=data[seq(1,nrow(data),by=by),]
 
   # Basic ggplot setup
-plot_data_sim_fjords_cubic_attractive_sub <- ggplot() +
-  geom_sf(data=land_border$geometry,fill="grey")+
+  plot <- ggplot() +
+    geom_sf(data=border$geometry,fill="grey")+
     coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
-  geom_path(data=data_sim_fjords_cubic_attractive_sub, 
+    geom_path(data=data_sub, 
               aes(x = y1, y = y2,group=ID),size = 1, 
               lineend = "round",alpha=0.15) +     
-  geom_point(data=data_sim_fjords_cubic_attractive_sub,
+    geom_point(data=data_sub,
                aes(x = y1, y = y2,shape=ID),size =4 ) +  
@@
 
-anim <- plot_data_sim_fjords_cubic_attractive_sub + 
+  anim <- plot + 
     transition_reveal(along = time) + 
@@
   animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
-        renderer = gifski_renderer("sim_fjords_cubic_attractive_sub.gif"))
-
-```
-
-![Simulated Trajectories](sim_fjords_cubic_attractive_sub.gif)
+          renderer = gifski_renderer(anim_name))
+  }
+
+```
+
+```{r}
+animate_trajectory_plot(data_fjords_tanh,land_border,by=60,anim_name="sim_fjords_tanh.gif")
+```
+
+```{r simulate with omega cubic in fjords,cache=TRUE}
+
+#set parameters values
+fomega_fjords_cubic_attractive<- function(cov_data){
+ fomega_poly(cov_data,a=3,D0=0.5,D1=1,sigma_theta=pi/8,sigma_D=0.5,b=4,theta1=pi/2,p=1)
+}
+
+data_fjords_cubic_attractive<- simulate_paralell(ftau_constant,fnu_constant,fomega_fjords_cubic_attractive,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_cubic_attractive,land_border)
+```
+
+```{r}
+animate_trajectory_plot(data_fjords_cubic_attractive,land_border,anim_name="sim_fjords_cubic_attractive.gif")
+```
+
+![Simulated Trajectories](sim_fjords_cubic_attractive.gif)
 
@@
 
-```{r simulate with omega cubic advanced and tau gaussian in fjords,cache=TRUE}
+```{r simulate with omega cubic repulsive in fjords,cache=TRUE}
 
@@
 
-# Generate samples ---------------
-cores=detectCores()
-cl <- makeCluster(6) #not to overload your computer
-registerDoParallel(cl)
-
-data_sim_fjords_cubic_repulsive=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
-
-  set.seed((seed-1)*N+i)
-
-  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega_cubic_fjords_repulsive,fnu=fnu_constant,v0=v0,x0=x0[i,],
-                    times=times,land=land_border,verbose=FALSE)
-
-  data_sim=res$sim
-  data_sim$ID=factor(rep(i,length(data_sim$y1)))
-  data_shore=res$shore[,c("p1","p2")]
-  cbind(data_sim,data_shore)
-}
-#stop cluster
-stopCluster(cl)
-
-```
-
-```{r}
-
-#subsample to 10 minutes
-data_sim_fjords_cubic_repulsive_sub=data_sim_fjords_cubic_repulsive[
-  seq(1,length(data_sim_fjords_cubic_repulsive$y1),by=60),]
-
-ggplot() +
-    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_fjords_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_fjords_cubic_repulsive_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none")
-
-```
-
-```{r animate trajectory sim_fjords_cubic_advanced_gaussian_sub,eval=FALSE}
-
-# Basic ggplot setup
-plot_data_sim_fjords_cubic_repulsive_sub <- ggplot() +
-  geom_sf(data=land_border$geometry,fill="grey")+
-  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
-  geom_path(data=data_sim_fjords_cubic_repulsive_sub, 
-            aes(x = y1, y = y2,group=ID),size = 1, 
-            lineend = "round",alpha=0.15) +     
-  geom_point(data=data_sim_fjords_cubic_repulsive,
-             aes(x = y1, y = y2,shape=ID),size =4 ) +  
-  labs(title = "Simulated trajectory in Scoresby Sound ", 
-       subtitle = "Time : {frame_along} h", 
-       x = "X Position", 
-       y = "Y Position") +
-  theme_minimal()
-
-anim <- plot_data_sim_fjords_cubic_repulsive_sub + 
-  transition_reveal(along = time) + 
-  labs(subtitle = "Time: {round(frame_along,1)} hours")
-
-# Save the animation
-animate(anim, nframes = 100, fps = 5, width = 800, height = 600, 
-        renderer = gifski_renderer("sim_fjords_cubic_repulsive_sub.gif"))
-
-```
-
-![Simulated Trajectories](sim_fjords_cubic_repulsive_sub.gif)
+data_fjords_cubic_repulsive<- simulate_paralell(ftau_constant,fnu_constant,fomega_fjords_cubic_repulsive,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_cubic_repulsive,land_border)
+```
+
+```{r}
+animate_trajectory_plot(data_fjords_cubic_repulsive,land_border,anim_name="sim_fjords_cubic_repulsive.gif")
+```
+
+![Simulated Trajectories](sim_fjords_cubic_repulsive.gif)
+
+```{r simulate with omega quadratic in fjords,cache=TRUE}
+
+fomega_fjords_quadratic=function(cov_data) {
+  fomega_quadratic(cov_data,a=0.5,D0=3)
+}
+
+data_fjords_quadratic<- simulate_paralell(ftau_constant,fnu_constant,fomega_fjords_quadratic,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_quadratic,land_border)
+```
+```{r simulate with omega sin cos in fjords,cache=TRUE}
+
+fomega_fjords_sin_cos=function(cov_data) {
+  fomega_sin_cos(cov_data,omega_0=10,D0=0.8,p=1,b=0)
+}
+
+data_fjords_sin_cos<- simulate_paralell(ftau_constant,fnu_constant,fomega_fjords_sin_cos,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_sin_cos,land_border)
+```
 
@@
 
-data_sim_fjords_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
+data_fjords_hanks=foreach (i=1:N,.combine='rbind',.packages=c("progress","MASS","sf","mgcv")) %dopar% {
 
@@
 ```{r}
-#subsample to 10 minutes
-data_sim_fjords_hanks_sub=data_sim_fjords_hanks[seq(1,length(data_sim_fjords_hanks$time),by=60),]
-
-ggplot() +
-    geom_sf(data = land_border$geometry, fill = "grey", inherit.aes = FALSE) +
-    coord_sf(datum = st_crs("+init=EPSG:32626 +units=km")) +
-    geom_path(data = data_sim_fjords_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = data_sim_fjords_hanks_sub, aes(x = y1, y = y2, col = ID), size = 0.1) +
-    geom_point(data = x0, aes(x = x1, y = x2), shape = 3, size = 2, col = "red") +
-    theme_minimal() +
-    xlab("x") + ylab("y") +
-    theme(legend.position = "none") 
-```
+get_trajectory_plot(data_fjords_hanks,land_border)
+```
+
+```{r}
+animate_trajectory_plot(data_fjords_hanks,land_border,anim_name="sim_fjords_hanks.gif")
+```
+
+```{r simulate with omega sin cos in fjords,cache=TRUE}
+
+#set parameters values
+fomega_fjords_sin_cos_attractive<- function(cov_data){
+ fomega_sin_cos(cov_data,omega_0=20,D0=0.5,p=2,b=3,D1=1,theta1=pi/2/sqrt(3),sigma_D=1,sigma_theta=0.5)
+}
+
+data_fjords_sin_cos_attractive<- simulate_paralell(ftau_constant,fnu_constant,
+                                                   fomega_fjords_sin_cos_attractive,v0,x0,times,land_border)
+```
+
+```{r}
+get_trajectory_plot(data_fjords_sin_cos_attractive,land_border)
+```
+We try a higher order polynomial.
+
+```{r simulate with omega cubic repulsive in fjords,cache=TRUE}
+
+fomega_fjords_poly_attractive=function(cov_data) {
+  fomega_poly(cov_data,a=1,D0=0.3,D1=1,sigma_theta=pi/8,sigma_D=1,b=3,p=3)
+}
+
+data_fjords_poly_attractive<- simulate_paralell(ftau_constant,fnu_constant,
+                                                fomega_fjords_poly_attractive,v0,x0,times,land_border)
+```
+
+```{r, message=FALSE}
+plot=get_trajectory_plot(data_fjords_poly_attractive,land_border)
+```
+
+```{r}
+animate_trajectory_plot(data_fjords_cubic_repulsive,land_border,anim_name="sim_fjords_cubic_repulsive.gif")
+```
+# Create final plots for the paper
 

### 2025-03-17 R/simulation_study/local/parametric.Rmd
reorganize simulation study

@@
 
-formulas <- list(tau=~1,
-                 nu=~1,a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
+formulas <- list(tau=~s(ID,bs="re"),
+                 nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,sigma_D=~1,sigma_theta=~1)
 
@@
   c("BoundaryDistance","BoundaryAngle")
-n_step=3
+n_step=5
 
@@
 interpolation_data<-interpolate_BoundaryMetrics(data_lf_he_prep,response=c("y1","y2"),
-                                                border,n_step,n_cores=6,k=0.5*nrow(data_lf_he_prep))
+                                                border,n_step,n_cores=6,df_ratio=0.95)
 
 #create dataframe with interpolated values
-interpolated_x_vec=as.vector(t(interpolation_data$x))
-interpolated_y_vec=as.vector(t(interpolation_data$y))
+interpolated_x_vec=as.vector(t(interpolation_data$y1))
+interpolated_y_vec=as.vector(t(interpolation_data$y2))
 interpolated_distance_vec=as.vector(t(interpolation_data$BoundaryDistance))
@@
 
+#interpolated_distance[interpolated_distance==0] <- 0.05
 crcvm_cubic_lf_he<- SDE$new(formulas = formulas,
@@
                                               "interpolated_angle"=interpolation_data$BoundaryAngle))
+crcvm_cubic_lf_he$update_lambda(c(1/0.5^2,1/0.5^2))
 
@@
 crcvm_cubic_lf_he$get_all_plots(show_CI="simultaneous")
+
+```
+```{r}
+
+get_estimates<- function(sde,include_sigma_obs=FALSE) {
+
+  post_coeff=sde$post_coeff(n_post=10000)
+  post_par=list(
+  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
+  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]),
+  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
+  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))
+
+  if (include_sigma_obs) {
+    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
+  }
+
+  mean_par=lapply(post_par,mean)
+  sd_par=lapply(post_par,sd)
+  quant_par <- lapply(post_par,
+                      quantile, probs = c(0.025, 0.975))
+
+  return(list("estimates"=mean_par,"std.error"=sd_par,"quantiles"=quant_par))
+
+}
+
+get_estimates(crcvm_cubic_lf_he)
 

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric36.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 36
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric75.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 75
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric9.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 9
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric92.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 92
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric94.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 94
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 R/simulation_study/cluster/CTCRW/fjords/Rscripts/fjords_parametric98.R

@@ -1,0 +1,350 @@
+cat("\014")                 # Clears the console
+rm(list = ls())             # Remove all variables of the work space
+
+library(here)               # To get root of git repo
+library(smoothSDE)          #to compute nearest shore point
+library(foreach)            #foreach loop
+library(doParallel)         #parallel computing
+
+
+
+domain_name="fjords"
+type="parametric"
+par_dir=here("R","simulation_study","cluster","CRCVM",domain_name)
+set_up_file=paste("set_up_",domain_name,"_",type,".R",sep="")
+source(file.path(par_dir,set_up_file))     #get set up for simulation study
+source(file.path(here("R","simulation_study","CVM_functions.R")))  #get functions to simulate trajectories
+
+
+seed= 98
+set.seed(seed)
+
+
+
+
+# Generate samples -----------------------------
+
+cores=detectCores()
+cl <- makeCluster(cores[1]-1) #not to overload your computer
+registerDoParallel(cl)
+
+
+data=foreach (i=1:N_ID_HIGH,.combine='rbind',.packages=c("progress","MASS","sf","mgcv","smoothSDE")) %dopar% {
+  
+  set.seed((seed-1)*N_ID_HIGH+i)
+  
+  #constant nu
+  fnu_constant=function(cov_data) {
+    return (exp(true_log_nu[i]))
+  }
+  
+  ftau_constant=function(cov_data) {
+    return (exp(true_log_tau[i]))
+  }
+  
+  res=sim_CRCVM(ftau=ftau_constant,fomega=fomega,fnu=fnu_constant,
+                log_sigma_obs=NULL,v0=v0,x0=x0[i,],times=times,land=border,verbose=FALSE)
+  
+  data_sim=res$sim
+  data_sim$ID=factor(rep(i,length(data_sim$y1)))
+  data_shore=res$boundary
+  data_sim=cbind(data_sim,data_shore)
+  data_sim
+}
+
+#stop cluster
+stopCluster(cl)
+
+# Points that reached land 
+count=0
+for (id in unique(data$ID)) {
+  sub_data=data[data$ID==id,]
+  if (nrow(sub_data) < n_obs) {
+    count=count+1
+    cat("ID",id,"reached land","\n",sep=" ")
+  }
+}
+
+cat(count/N_ID_HIGH*100,"percent of the samples reached land \n")
+
+if (count>0) {
+  stop("At least one trajectory reached land.")
+}
+
+
+# add noise
[…]

### 2025-03-17 Latex/AoAS - Revision 2024. Alexandre, Adeline, Susanne/response.tex
add paper revision

\documentclass{amsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage[colorlinks,allcolors=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures
\RequirePackage[numbers]{natbib}

%\usepackage[utf8]{inputenc}
\usepackage[dvipsnames]{xcolor}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{breqn}
\usepackage{hyperref}
%\usepackage{biblatex}
\usepackage[foot]{amsaddr}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{array}
\usepackage{tabularx}

\usepackage[a4paper, total={7in, 10in}]{geometry}
\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-2.6in}
\calclayout

\newcolumntype{N}{>{\centering\arraybackslash}m{.4in}}
\newcolumntype{G}{>{\centering\arraybackslash}m{2in}}

\theoremstyle{plain}
\newtheorem{lemma}{Lemma}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{claim}{Claim}
\newtheorem{corollary}{Corollary}[section]

\theoremstyle{remark}
\newtheorem{assumption}{Assumption}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{remark}{Remark}[section]
% Alternative Assumption!
\newtheorem{assumptionalt}{Assumption}[assumption]
% New environment
\newenvironment{assumptionp}[1]{
  \renewcommand\theassumptionalt{#1}
  \assumptionalt
}{\endassumptionalt}
\newtheorem{innerass}{\innerthmname}
\newcommand{\innerthmname}{}% initialize
\newenvironment{upstatement}[1]
 {\renewcommand{\innerthmname}{#1}\innerass}
 {\endinnerass}

\newtheorem{manualtheoreminner}{Assumption}
\newenvironment{manualtheorem}[1]{%
  \renewcommand\themanualtheoreminner{#1}%
  \manualtheoreminner
}{\endmanualtheoreminner}

% Custom commands
\newcommand{\peq}{\stackrel{\mathcal{L}}{=}}
\newcommand{\tr}{\text{tr}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand {\R}{\mathbb{R}}
\newcommand {\E}{\mathbb{E}}
\newcommand {\prob}{\mathbb{P}}
\newcommand {\N}{\mathbb{N}}
\newcommand {\Z}{\mathbb{Z}}
\newcommand {\1}{\mathbb{1}}

%\addbibresource{bibliography.bib}

\title[]{Response}

\begin{document}
\maketitle

We thank the associate editor as well as the two reviewers for careful reading of our manuscript. We have done our best to accommodate all suggestions, which we believe have improved the paper. There are two versions of the revised manuscript: a clean version, and a version where new text is indicated in blue, and deleted parts are in red and striked over. Below we respond point by point to the comments.

\section{Response to Reviewer 1}

\begin{itemize}
\item \textcolor{blue}{My main concern with your method is that you do not propagate the uncertainty from the baseline model to the response model. Underestimating the uncertainty could lead someone to deduce that there is a clear deviation from baseline (i.e., the animals' behaviour was disturbed) when this is not the case. This goes against your claim that you are giving "a conservative estimate of the effect of the noise exposure" on page 5.}
We agree that the uncertainty on the baseline can greatly invalidate our conclusions, and it is therefore crucial to asses it precisely. 
We tried to better capture the uncertainty from the baseline as follows : we draw 100 samples from the estimated distribution of the baseline parameters $\tau_0$ and $\nu_0$ using the joint Fisher information matrix.  We estimate the response model with these fixed value and get an estimate of $\alpha_\tau$ and $\alpha_\nu$. We compute the $2.5$ and $97.5 \%$ quantiles of the 100 estimates of $\alpha_\tau$ and $\alpha_\nu$ thus obtained to get a wider confidence interval for $\alpha_\tau$ and  $\alpha_\nu$. Results are shown in table 4. We know show mean estimates of the recovery distances and $95\%$ CI in table 5 including the uncertainty from the baseline estimates.

\item \textcolor{blue}{ p4: "The land geometry for the specific region..." If I understand correctly, you first tried using land maps from OpenStreetMap, but finally decided to use the Google Earth Engine API for higher accuracy. If that's the case, I would recommend presenting the Google Earth approach directly at the start of the paragraph.}
We finally used another approach that is explained in the beginning of the second paragraph of section 2.2.

\item \textcolor{blue}{- p5: Just above Figure 1, shouldn't the interval be open either at pi or -pi (depending on whether you work on (-pi, pi] or [-pi, pi))? Same comment under Equation 7 on page 8.}
The interval has been changed to $]-\pi,\pi]$.

\item \textcolor{blue}{- Fig 2: The figure labels are small and hard to read.} The size of the labels and the figure are now increased.

\item \textcolor{blue}{- p9 and Fig 4: I think you should mention more explicitly in text and in the caption that Figure 4 is showing simulated data. Please consider replacing "sampled trajectories" by "simulated trajectories" in the text for clarity.} We replaced "sampled trajectories" by "simulated trajectories" accordingly.

\item \textcolor{blue}{- Figure 4: Please make the labels a little larger for readability, though.}

\item \textcolor{blue}{- p 12, under Equation 12: It seems that you don't explicitly account for the cyclical nature of the angular covariate. Shouldn't you be using a cyclical spline (e.g., "cc" in mgcv syntax) for this?} I don't think cyclic splines are of any interest there since the values of the smooth functions at $-\pi$ and $\pi$ should not be equal but opposite. A basis of odd spline functions might be used though. We did not investigate this any further as this shoud be equivalent to fixing some of the splines weights to $0$.

\item \textcolor{blue} { - p12: The text under Equation 14 seems to imply that you first estimate the baseline parameters (e.g., tau0, nu0, etc), and then estimate deviation parameters (alpha_tau and alpha_nu) while treating baseline parameters as known. This doesn't seem to properly propagate uncertainty about the baseline parameters onto inferences about the deviation parameters, which could affect your conclusions (e.g., whether there seems to be a clear deviation from baseline behaviour, as you claim in Section 6.2 based on confidence intervals). I think you should consider estimating all parameters jointly to circumvent this problem, or at the very least discuss the implications of your approach. You briefly mention this on page 17, but it should be discussed earlier and in more detail. } See first remark. We now propagate uncertainty and obtain wider confidence intervals in tables 4 and 5. Using a reference baseline model allows us to estimate recovery distances with respect to the baseline values, which gives a clear criteria for mitigation of noise exposure effect. This is why we eventually chose this method.

\item \textcolor{blue}{- Table 1: I am not sure that it is necessary/helpful to report estimates for the spline coefficients, as they aren't meaningful when taken separately. It might be more useful to report some summary of the discrepancy between the true smooth function and the estimated smooth function.} We now show the estimated smooth surfaces $\omega$ instead of the values of the spline coefficients.

\item \textcolor{blue}{- p16: Can you please spell out why setting "the degree of freedom of the marginal splines to 5" leads to "a total number of 25 degrees of freedom?}
For tensor spline the model matrix is the tensorial product of the individual model matrices for each single variable. Its size is thus the product of the degree of freedoms in each single variable.

\item \textcolor{blue}{- p16: You fixed the smoothing penalties to 1 for the 2d tensor spline modelling the relationship between the angular velocity and covariates. This seems like a strong assumption, can you please clarify why you did this rather than estimating the smoothing penalty, and how you chose this specific value of 1?}
When estimating the smoothing penalty from the data, we encounter numerical errors that are solved when they are fixed. Typically, the smoothing penalties are estimated to extremely high values. Fixing the value to $1$ for the smoothing penalty was indeed not the best choice. For the simulation study, we fixed the smoothing penalties based on the values we obtain by estimating the true surface with splines. However in the case study we can only guess reasonable values based on the surface we expect to obtain. Here, fixing its value close to zero, for instance $10^-6$, is a better choice.

\item \textcolor{blue}{- p16, above Table 2: You should probably remind the readers here that tau0 is the population mean for the persistence parameter. } Done

\end{itemize}

\section{Response to Reviewer 2}

\begin{itemize}
    \item \textcolor{blue}{1.	I’m honestly shocked to see a paper submitted to AoAS on animal movement without a single plot of the animal tracking data in the main manuscript.  I think it is critical that you show readers the data you are modeling.  You show multiple sets of simulated data tracks in the main manuscript, but never the actual movement data.  On a more general note, while I think the modeling approach to  modeling constrained movement is very interesting, it isn’t clear to me what scientific impact this modeling approach has.  Not even showing the data clearly is indicative of a more general lack of serious applied scientific analysis in this manuscript, in my opinion.  I think AoAS readers will really appreciate understanding the scientific advances that can be made now that you have made this modeling advancement.    }

    The plot of the actual data has been moved to section 2 and now appears on page 4. The simulation study now shows the result of the estimation with a CTCRW model where the constraints on the movement are not considered. The estimate of the persistence parameter $\tau$ appears to be biaised. That is one of the reason why constraints should not be neglected in general. This biais was also discussed in Hanks paper that is cited in the text.
    We hope this will make the scientific advances in the paper clearer.
    \item \textcolor{blue}{Your data analysis focuses on the effect of a disturbance on movement.  This seems to be a relatively minor scientific result, and one that possibly could have been obtained without modeling the shoreline at all.  Can you somehow convince readers that this data and this question are interesting? }
    The question of assessing behavioral disturbances in marine mammals movement have long been a subject of scientific interest.

    \item \textcolor{blue}{It has been analyzed before.  Are your results novel and different from those obtained with simpler analyses?  Could your results not have been achieved without your new modeling approach?} 
    We now go in more detail about the results of past studies in the introduction and explain how our approach differs from it.

    \item \textcolor{blue}{ Why is it important to know that marine mammals respond to loud sounds?  What implications does it have for them? }
    This is not the main subject of this paper, however we refer to the articles of Heide-Jorgensen et al. as well as the very interesting textbook of Southall et al. Essentially, we mentioned in the paper that narwhals rely on sound perception for orientation and communication, which makes it clear that loud underwater sounds may alter there ability to forage or mate for instance.

    \item \textcolor{blue}{ Why is it critical to model in such a way that the constrained waterways are seen as hard barriers?  There are multiple possible approaches here, but I really think readers will appreciate additional scientific motivation and results for this work.}
    Waterways typically are hard barriers for the narwhals so we think that treating them as such makes some sense.

    \item \textcolor{blue}{ 2.	I think your novel modeling approach for modeling constrained movement is very interesting.  However, I think readers will very much appreciate a better understanding of where this approach might break down, and a more serious comparison with existing approaches.  For example, coarsening the time step of your observations seems like it would result in simulations that do not respect the firm boundaries, but this is never discussed.  Is there any issue to modeling coarse resolution data near shorelines?  Can you show that your movement data are close enough together in time that this is not a problem for this data set?  Or can you compare with an existing approach for modeling constrained movement (you cite some in your paper) and clearly illustrate how your approach improves on existing approaches?  You make some statements about how your approach has some nice properties relative to other approaches, but without a direct comparison it isn’t clear how your new approach improves on old approaches. }

    Simulated trajectories from a reflected CVM (the model used in Hanks paper) are compared to simulated trajectories of our model in figure 5. In a reflected CVM, there is no reason for the process to remain close and follow the shoreline. This is where our model is interesting : it can model different behaviors close to the boundary of the domain.

    \item \textcolor{blue}{ 3.	Your simulation study shows that you can simulate data constrained to be within very interesting boundaries, and also shows that if your data are high resolution, you can recover parameters when you fit simulated data.  However, it isn’t clear how well your model actually fits the data you are using.  Can you show that data simulated from your model is very similar to observed movement?  Can you show that you can recover parameters well if you simulate movement data from the parameters you estimated from your tracking data?  Doing this will help readers better understand what your model captures well in the data and what it does not.}
    We added a new section 6.3 in which we show data simulated from the fitted model. We also show parameters estimated from the simulated data.

    \item \textcolor{blue}{Page 2, 3 lines from bottom.  You say that reflected SDEs do not have a known transition density, but I don’t think that is true, or perhaps you need to clarify.  For example, your model doesn’t have an analytically tractable transition density, but once you discretize in time and assume that some things (like relative position to shore) are constant over that time interval, you can then derive a transition density.  I think this is very similar to what happens in reflected SDEs.  Perhaps clarify here.}
    We tried to emphasize that the use of a reflected SDE (as in Hanks2017) generally prevents from using a state space model for estimation since the next position is conditioned by whether the previous has reached land or not, which relies on a projection operator that is not explicitly known and of which we don't know the distribution. Therefore it is not straightforward to get an approximate linear gaussian state space model in this case, and Monte-Carlo technique are much more popular. In our case, the projection operator is hidden in the covariates $\Theta$ and $D^{shore}$.
    To be exhaustive, reflected Brownian motion and reflected Ornstein-Uhlenbeck indeed have a known transition density in one dimension (reflection at $0$ or any threshold $a$). However, in multiple dimensions, we can have an explicit transition density only for simple domains as the half space ($\R \times \R^+$). For a general polygon as it is the case in the paper, we cannot simulate explicitly a reflected RACVM but we can only use numerical schemes as the projection method in Hanks's paper (or the penalization scheme in Brillinger's paper).

    \item \textcolor{blue}{ Section 2.1.  As stated above, I think readers will really appreciate more scientific analysis here – in particular a better understanding of why this dataset should be analyzed again with your methods.  If possible, clearly state what sorts of results have been obtained before (you allude to these), and clearly show if possible how your results improve on these, or make management actions clear that were not clear before, or something similar.}

    \item \textcolor{blue}{Page 4, paragraph 2, sentence starting “Measurements” is not clear to me – please clarify.} The sentence is changed to "Additionally, a velocity filter was applied to retain only positions with an empirical velocity below 20 km/h. This filter removed two data points from the dataset."

    \item \textcolor{blue}{Page 4, sentence starting “About 9 \% of the GPS measurements…” – can you give more details here?  It might make more sense to not mentione using OpenStreetMap, and instead just focus on the Google Earth data?} This sentence was removed since we changed our approach. The explanation is found in paragraph 2 of section 2.2.

    \item \textcolor{blue}{Page 4, last line.  You assume that narwhals are “influenced by the shore within a 3km range.  This seems like a very important assumption – can you justify it?  Or do some sort of sensitivity analysis to this assumption?}

    Figure 6 shows that at 1.45 km the estimated smooth function $\omega$ is already at 0 (less than $1$ rad / h) meaning that at this distance there is already no effect of the shore on the parameter $\omega$, which clearly indicates that our assumption holds.

    \item \textcolor{blue}{Page 7, paragraph starting with “Equation (5) shows that…”  At this point in the manuscript, it isn’t clear that you are going to let “omega” vary over time, and my thought at this point was that this model would require the animal to have a constant mean turning angle of clockwise or counterclockwise.  I suggest you tell readers that you will later model “omega” as varying over time and as a function of distance to the shore.   }
    We made further comment about this in the first paragraph of section 3 : "We will later model $\omega$ as varying over time as a function of the distance to shore and the angle $\Theta$."

    \item \textcolor{blue}{
    Page 9, first paragraph.  Is this persistent movement along the shoreline something that animals will only do when they are close to shore?  In your model, could animals follow the shoreline when they are very far from it?  For example moving north/south a long ways from a north/south shoreline?  If your model only has this as a possibility when animals are close to shore, I suggest mentioning that here.}
    We changed our approach and now kept $\tau$ constant over time in the illustrative trajectories from our model, though it is completely doable to model $\tau$ as a function of $\Theta$ and $D^{shore}$. The reason is that we only estimate a constant $\tau$ in the application to real data, in an effort to have a model as parsimonious as possible.

    \item \textcolor{blue}{Page 10, last line.  You say “to prevent the animal from hitting the boundary, rho should remain smaller than D^shore.”  This seems to me like an incredibly important result, and is not given any justification.  Can you provide some details, especially in the context of your numerical approximation you present later (a time discretization)?}
    This is a rule of thumb that we stated based on our simulations and on common sense. It would obviously depend on the sharpness of the domain and the size of the time steps used in the simulation. Maybe a proof in a simple domain as the half space $\R^+ \times \R$ could work but we did not investigate this any further.

    \item \textcolor{blue}{Page 11, line before Equation (10).  I suggest changing “solve” to “are the solutions to”} Done

    \item \textcolor{blue}{Page 12, paragraph after Equation (12).  I believe these are also called “product B-splines” } Thank you. The main reference we have is section 4.1.8 of S.Wood 2017.
\end{itemize}

\section{Response to the associate editor}

\begin{itemize}
    \item \textcolor{blue}{I understand that the contribution of this work is in employing the CTCS model to analyze
GPS positioning of the narwals, which in turn helps to quantify the effects of sound exposure
as a function of distance from shore. What isn’t clear to me is what are the shortcomings
of the previous work to this aim, and, more specifically, how can the benefits of the newly
proposed approach be demonstrated.} This is slightly discussed in the introduction.

\item \textcolor{blue}{In this particular application, is the main challenge the irregular boundaries of fjords? I think
some significant revisions are required for the introduction and/or the data section to depict
the application in this analysis that is motivating this work. As one of the other reviewers
mentions, it is critical to the readership of AOAS to show the data and clearly demonstrate
why this application motivated the need for advanced statistical methods and models.}

\item \textcolor{blue}{It seems that the new methods for constrained animal movement models could/should be
more broadly useful beyond just investigating sound exposure impacts on narwals. If the
main methodological contribution is the developing of methods for constrained movement,
this should be expanded upon to various application areas. I’m not saying fit the models
to additional data but rather paint a bigger picture to the contributions here. For example,
surely other animals have movement constraints. But what about fire? There has been quite
a bit of work as of late for movement of fires so perhaps you could discuss this as well? Or
others?}
\end{itemize}

\end{document}

### 2025-03-17 Latex/AoAS - Revision 2024. Alexandre, Adeline, Susanne/main_revised.tex
add paper revision

--- 2024-11-08 Latex/main/aoas-template.tex
@@
 The fjord system has a complex shore line, restricting the behavioral response options for the narwhals to escape the threats.
-We propose a new continuous-time correlated velocity model with varying coefficients that includes spatial constraints on movement.
+We propose a new continuous-time correlated velocity model with varying coefficients that includes spatial constraints by adjusting the direction of the movement to the boundary of the domain.
 To assess the sound exposure effect we compare a baseline model for the movement before exposure to a response model for the movement during exposure.
-Our model, applied to the narwhal data, suggests increased tortuosity of the trajectories as a consequence of the spatial constraints, and further indicates that sound exposure can disturb narwhal motion up to a couple of tens of kilometers.
-Specifically, we found an increase in velocity and a decrease in the movement persistence.
+Our model, applied to the narwhal data, indicates that sound exposure can disturb motion up to a couple of tens of kilometers.
+Specifically, we found an increase in velocity and a decrease in persistence.
+We give estimates and confidence intervals of threshold distances to the sound source for medium and strong shifts in the whales motion.
 \end{abstract}
@@
 
+%IN THIS DOCUMENT YOU SHOULD DO ALL CHANGES.
+YOU CAN JUST DO THEM DIRECTLY.
+IF YOU REMOVE THE \% SIGN FROM THE FILE latexmkrc YOU WILL SEE THE CLEAN DOCUMENT.
+WHAT IS ADDED IS COLORED BLUE, WHAT IS REMOVED IS RED AND STRIKED OVER.
+
 %general context
+
+%the models we consider
+In ecology, animal movement is often modeled using continuous-time processes defined as solutions to stochastic differential equations (SDEs).
+One well-known example is the continuous-time correlated random walk, also called the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \citep{johnson_continuous_2008}.
+It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \citep{preisler_analyzing_2013}.
+Varieties of CVMs have been applied to birds \citep{janaswamy_state_2018}, marine algae \citep{gurarie_estimating_2011}, ants \citep{russell_spatially_2018}, seals \citep{johnson_continuous_2008,albertsen_generalizing_2018}, bowhead whales \citep{gurarie_correlated_2017} and sea lion motion \citep{hanks_reflected_2017}.
+In many cases, animal movement is assumed to occur without spatial constraints \citep{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}, though most animals encounter physical barriers during their locomotion.
+Ignoring such constraints can lead to biased parameter estimates \citep{hanks_reflected_2017}.
+These challenges have been discussed in the past by \cite{brillinger_simulating_2003}, who judiciously pointed out that the way to model motion near the boundary of the domain should be species-specific and account for the particular behaviors and spatial use of the studied animal.
+Example models for constrained motion are reflected CVMs that have been considered to study sea lion telemetry data \citep{hanks_reflected_2017} and CVMs with drift described by a repulsive potential function to constrain ants movement within a box \citep{russell_spatially_2018}.
+While these approaches effectively introduce movement constraints, they fail to capture certain behaviors : what if the animals have a tendency to move along the boundary, or remain close to it ?
+Neither a reflected SDE nor SDE with exponential repulsive potential in \cite{hanks_reflected_2017} and \cite{russell_spatially_2018} would allow for this type of behavior.
+As a consequence, there is a need to develop new models adapted to such spatial constraints and behaviors.\\
+
+%new contributions
+In this paper, we extend the rotational CVM defined in \citep{gurarie_correlated_2017} by considering a rotation parameter $\omega$ expressed as a smooth function of the distance to the boundary and the angle between the animal's heading and the boundary normal vector.
+This makes the drift in the velocity equation dependent on the location process and the boundary of the domain, and allows the velocity to rotate as the animal approaches the boundary of the domain.
+We show that parameter estimation for this SDE model can be performed in the framework of the \texttt{R} package \texttt{smoothSDE} \citep{michelot_varying-coefficient_2021}.
+To make this possible, we derive explicit formulas for the transition density of the location and velocity process and formulate an approximate linear Gaussian state-space model from the discretisation of these formulas.
+Maximum likelihood estimation based on the Kalman filter is then performed as for any other model in \texttt{smoothSDE}.
+The estimation procedure exploits the capabilities of the \texttt{R} package \texttt{TMB} to get approximate likelihood using Laplace's approximation \citep{kristensen_tmb_2016,albertsen_fast_2015}.
+It is considerably less computationally intensive than an estimation procedure based on Markov Chain Monte Carlo posterior samples as done in \citep{hanks_reflected_2017,russell_spatially_2018}.\\
+
+%general purpose of the paper
+Throughout the paper, we will apply our SDE model to the horizontal motion of narwhals.
+In our case study, the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland), the largest fjord system in the world, known to be the summer residence for an isolated population of narwhals.
+These narwhals have a proclivity to rotate to avoid or align with the shoreline without reaching the boundary, which motivates our new modeling approach.
+Our aim is to be able to detect and quantify any potential behavioral response to external human disturbances from the observed narwhals tracks.
+We base our analysis on data collected in 2018 \citep{heide-jorgensen_behavioral_2021} from animal-borne tags during one week of controlled exposure to seismic airguns.
+\cite{heide-jorgensen_behavioral_2021} observed that some narwhals react to sound exposure by changing their swimming speed and directions at distances between $5$ and $24$ km from the ship.
+For some individuals an increase up to  $30\%$ in the horizontal travel speed could be detected.
+In our approach, the mean velocity and the tendency of the narwhals to change direction are captured within the SDE parameters, and we are therefore able to quantify precisely the effects of sound exposure on these parameters at the population level.
+More precisely, sound exposure is introduced in the model through a covariate defined as the inverse of the distance to the sound source.
+This choice is motivated by \citep{heide-jorgensen_behavioral_2021, tervo_narwhals_2021,tervo_stuck_2023}.
+Following \cite{michelot_continuous-time_2022}, a response model is compared to a baseline model for the narwhal movement before exposure to the disturbance, that is, under normal conditions.
+The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.
+We eventually give estimates and confidence intervals of threshold distances to the sound source for medium and strong shifts in the whales movement.
+These estimates can serve as a guideline for mitigation measures and conservation policies.
+\\
+
+%summary
+To summarize, the main contributions of this paper are:
+\begin{itemize}
+	\item Definition of a constrained version of the CVM, where deviation angles from shoreline and distance to shore are used to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
+	\item Derivation of explicit formulas for the transition density of the rotational CVM defined in \citep{gurarie_correlated_2017}, and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
+	\item Statistical analysis of narwhal trajectories based on our SDE model, using the exposure covariate defined as the inverse of the distance to the sound source, and estimation of threshold distances from the sound source for strong and medium shift in the behaviour of the whales.
+\end{itemize}
+
+In Section 2, we give the general context of controlled exposure experiments as well as an overview of the narwhal data available for the analysis.
+The diffusion models are then discussed in Section 3.
+Section 4 is dedicated to the statistical model used to infer the shore constraint and the sound exposure effects.
+Results of a simulation study are shown in Section 5.
+Finally, our models are applied to analyse the behavioral response of the narwhals in Section 6.
+
+\section{Controlled exposure experiments data}
+\label{section: movement data}
+
+\subsection{Motivation}
+\label{subsection: controlled exposure experiments}
+
 There is a large variety of sources of anthropogenic underwater noise in the Arctic region, including sonars, ice breakers, vessel traffic, drilling  and seismic airguns used for oil and gas exploration \citep{halliday_underwater_2020}.
@@
 Besides causing physical harm, anthropogenic noise can also disturb their behavior and interrupt their natural foraging habits.
-However, there is still no clear criteria for measuring behavioral disturbances due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions \citep{southall_marine_2008,southall_marine_2019}.
+However, there is still no clear criteria for measuring behavioral disturbances, due to the multiplicity of contextual variables that can influence a behavioral reaction, and the variety of these reactions \citep{southall_marine_2008,southall_marine_2019}.
 Comprehensive behavioral studies are thus essential.\\
@@
 Our study is based on these data collected in 2018.
-\\ %%% link sentence missing
-
-%general purpose of the paper
-This paper concentrates on the horizontal motion of narwhals.
-Previous analysis showed that the probabilities of the narwhals being close to the shore or moving towards the shore increase with exposure to the sound.
-However, these analyses were based on discrete-time discrete-space Markov chains, despite the continuous nature of narwhals movement in both time and space.
-To address this, we employ a continuous-time continuous-state model to analyze GPS positions, aiming to quantify the effects of sound exposure on the  speed and heading of the narwhals.
-This modeling approach offers multiple advantages: it avoids the need for positional interpolation at regular intervals, eliminates manual state labeling, incorporates GPS measurement errors, and allows to precisely quantify the magnitude of the changes in the movement depending on the distance to the sound source.\\
-
-%the models we consider
-We specifically focus on continuous-time models defined as a solution of a stochastic differential equation (SDE) with time-varying coefficients.
-One widely known model in this context is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \citep{johnson_continuous_2008}.
-It can be viewed as a special case of velocity potential models, where the potential is a second degree polynomial \citep{preisler_analyzing_2013}.
-Varieties of CVMs have been applied to birds \citep{janaswamy_state_2018}, marine algae \citep{gurarie_estimating_2011}, ants \citep{russell_spatially_2018}, seals \citep{johnson_continuous_2008,albertsen_generalizing_2018}, bowhead whales \citep{gurarie_estimating_2011} and sea lion motion \citep{hanks_reflected_2017}.\\
-
-% specificity of our data
-In our case study, the narwhals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland), the largest fjord system in the world, known to be the summer residence for an isolated population of narwhals.
-Thus, the modeling process must include shoreline boundaries through movement constraints.
-Most existing CVMs for animal movement do not account for such constraints \citep{johnson_continuous_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}.
-Example models including landscape boundaries are constrained SDEs  \citep{hanks_reflected_2017} in which a reflected version of the CVM is considered to study sea lion telemetry data, and SDEs with drift described by a potential function \citep{russell_spatially_2018} where a repulsive potential function  constrains ants movement within a box.
-\citep{brillinger_simulating_2003} also illustrates different methods of simulation for constrained animal movement, and emphasizes that the way to model motion near the boundary of the domain should be species-specific and account for the particular behaviors and spatial use of the studied animal.
-Narwhals have a proclivity to rotate to avoid or align with the shoreline without reaching the boundary.
-A reflected SDE as in \citep{hanks_reflected_2017} does not allow this type of behavior.
-In addition, a simple exponential potential function is suitable only with basic boundaries, such as the rectangular box in \citep{russell_spatially_2018}, but would break down for a complex boundary such as Scoresby Sound fjords.
-As a consequence, there is a need to develop new models adapted to such spatial constraints and behaviors.
-Reflected SDEs have certain limitations.
-On the one hand, it is not a modelling of shoreline avoidance behaviour.
-The model only uses a projection operator to ensure that the process remains within the domain.
-On the other hand, the transition density of these models is not known, which complicates the statistical inference of the parameters.
-We therefore propose to work with the second class of model, SDEs whose drift constraints movement within the system of fjords.
-\\
-
-%new contributions
-The novelty of this paper consists in extending the rotational CVM defined in \citep{gurarie_correlated_2017} with a rotation parameter expressed as a smooth function of the distance to the boundary and the angle between the animal's heading and the boundary normal vector.
-This makes the drift in the velocity equation dependent on the location process and the boundary of the domain, and allows the velocity to rotate as the narwhal approaches the shore.
-We show that the rotational parameter can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \citep{michelot_varying-coefficient_2021}.
-To make this possible, we derive explicit formulas for the transition density of the location and velocity process and formulate a linear Gaussian state-space model from the discretisation of these formulas.
-Maximum likelihood estimation based on the Kalman filter is then performed as for any other model in \texttt{smoothSDE}.
-The estimation procedure exploits the capabilities of the \texttt{R} package \texttt{TMB} to get approximate likelihood using Laplace's approximation \citep{kristensen_tmb_2016,albertsen_fast_2015}.
-It is considerably less computationally intensive than an estimation procedure based on Markov Chain Monte Carlo posterior samples as done in \citep{hanks_reflected_2017,russell_spatially_2018}.\\
-
-In this general framework, sound exposure is introduced in the model through smooth functions of a covariate defined as the inverse of the distance to the sound source.
-This choice is motivated by \citep{heide-jorgensen_behavioral_2021, tervo_narwhals_2021,tervo_stuck_2023}.
-We estimate the effect of the exposure variables on both the velocity and the persistence parameters of the diffusion process.
-A response model is compared to a baseline model for the narwhal movement before exposure to the disturbance, that is, under normal conditions \citep{michelot_continuous-time_2022}.
-The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.
-\\
-
-%summary
-To summarize, the main contributions of this paper are:
-\begin{itemize}
-	\item Definition of a constrained version of the CVM, where deviation angles from shoreline and distance to shore are used to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
-	\item Derivation of explicit formulas for the transition density of the rotational CVM defined in \citep{gurarie_correlated_2017}, and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
-	\item Analysis of narwhal data to assess a behavioural response using the exposure covariate defined as the inverse of the distance to the sound source.
-\end{itemize}
-
-In Section 2, we give an overview of the narwhal data available for the analysis.
-The diffusion models are then discussed in Section 3.
-Section 4 is dedicated to the statistical model used to infer the shore constraint and the sound exposure effects.
-Results of a simulation study are shown in Section 5.
-Finally, our models are applied to analyse the behavioral response of the narwhals in Section 6.
-
-\section{Movement data of six narwals in South-East Greenland}
-\label{section: movement data}
-
-\subsection{Description of the controlled exposure experiments}
-\label{subsection: data description}
-The dataset analysed in this paper has been subject of several studies, focusing mainly on vocalizations and avoidance reactions \citep{heide-jorgensen_behavioral_2021,tervo_narwhals_2021,tervo_stuck_2023}.
-Here we briefly describe the data, for more details see \citep{heide-jorgensen_behavioral_2021}. 
+ \\
+
+\subsection{Description of the data}
+We briefly recall the context for these controlled exposure experiments.
+For more details, the reader can refer to \cite{heide-jorgensen_behavioral_2021}.
 Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound in South-East Greenland by biologists from the Greenland Institute of Natural Ressources, with the help of local Inuit hunters.
@@
 The guns were fired synchronously every $80$ s, while the GPS navigation system recorded the location of every shot.
-The data include latitude and longitude of the narwhal positions, distance relative to the ship in metres, GPS time, and distance to the shore in metres.
+The data includes latitude and longitude of the narwhal positions, distance relative to the ship in metres, GPS time, and distance to the shore in metres.
 GPS positions are known only at times when the narwhals are at the surface.
 The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours.
-While the statistical analysis in \citep{heide-jorgensen_behavioral_2021} relied on augmented data with positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.\\
+While the statistical analysis in \citep{heide-jorgensen_behavioral_2021} relied on augmented data with positions linearly interpolated at each second, here we will only consider the actual GPS measurements with irregular time steps.
+The GPS tracks are shown in Figure~\ref{fig: observed GPS data}.\\
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=0.6]{images/data_exploration/all_narwhals_tracks.pdf}
+\caption{Observed narwhals GPS tracks. Red crosses indicate the starting points of the trajectories.}
+\label{fig: observed GPS data}
+\end{figure}
 
@@
 We discard the GPS measurements from the first 12 hours after tagging, to avoid any tagging effects on the behavior.
-Measurements that result in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
+We applied a velocity filter to keep only positions that have empirical velocity lower than 20 km/h.
+Two data points were removed with this filter.
 Overall, $4815$ GPS positions are kept for the analysis. The splitting between data before and after exposure results in $1558$ measurements before exposure and $3257$ measurements after exposure.
 
-\subsection{Notation and definition of covariates}
+\subsection{Notations and covariates}
 \label{subsection: covariates}
@@
 
-The land geometry for the specific region in South-East Greenland is gathered from the \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database.
 The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move.
-For each narwhal $i\in\{1, \ldots N\}$ and each time $j\in\{1, \ldots, n_i\}$, the closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$.
+It is critical to accurately represent the shoreline geometry to ensure realistic movement constraints within the fjord system.
+Initially, we used land polygons obtained from OpenStreetMap, but these led to approximately $9\%$ of recorded narwhal positions being incorrectly classified as on land, likely due to inaccuracies in the shoreline representation.
+To address this, we manually refined the polygons using satellite raster data in QGIS, improving accuracy but still observing some positions classified as on land or unrealistically close to the shoreline (<10 m).
+To account for both shoreline inaccuracies and potential GPS error, we applied a geometric buffer using the \texttt{st\_buffer} function in \texttt{R} to shrink the land polygons by 200 m.
+This adjustment significantly reduced the proportion of positions classified as on land to $0.12\%$ and yielded results more consistent with expected narwhal behavior.
+The shapefile defining the shoreline is made available in Supplementary materials.
+For each narwhal $i \in\{1, \ldots N\}$ and each time $j\in\{1, \ldots, n_i\}$, the closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$.
 Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary.
-About $9\%$ of the GPS measurements in the dataset turned out to be on land.
-This was due  to inaccuracies in the shoreline maps rather than errors in GPS measurements.
-We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons.
-After adjustement of the polygon data using \texttt{qgis} software,  approximately $1 \%$ of the observations remain on land.
-We kept these positions for the analysis.
 Distance to shore values range from $0$ to $7.6$ km.
-The shapefile defining the adjusted shoreline is made available in Supplementary materials.
-We define a new covariate $I^{shore}$ based on the inverse of the distance to the shore in kilometres with the following formula
-\begin{equation}
-	\left\{
-	\begin{array}{lll}
-		I^{shore}_{ij}=0 && \mbox{if } D^{shore}_{ij}>D_{up} \\
-		I^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} && \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
-		I^{shore}_{ij}=\frac{1}{D_{low}} && \mbox{ if } D^{shore}_{ij} \leq D_{low}
-	\end{array}
-	\right.
-	\label{eq: exp shore definition}
-\end{equation}
-where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of the observed distances, and $D_{up}=3$ km, assuming narwhals are influenced by the shore within a $3$ km range. 
 The angle $\Theta_{ij}$ between the narwhal's heading and the shoreline is defined as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{ij+1}-y_{ij}}{\Delta_{ij}}$, where $\Delta_{ij}=t_{ij+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$, as illustrated in Figure~\ref{fig: illustration nearest shore points}.
-A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates a movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement towards the shore.
+A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates a movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left]-\pi,-\frac{\pi}{2}\right[$ indicates movement towards the shore.
 
@@
 The distance to the ship $D^{ship}$ (km) is defined as the distance between observed GPS locations of the ship and the narwhal. The values are comprised between $2.68$ and $63.8$ km.
-Exposure to the ship for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \citep{heide-jorgensen_behavioral_2021}.
+Exposure to the ship for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km).
 The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals.
@@
 This provides a conservative estimate of the effect of the noise exposure.
-Levels of the covariates $I^{shore}$ and $E^{ship}$ for each narwhal are displayed in Figure~\ref{fig: realexpthroughtime}.
+Levels of the covariate $E^{ship}$ for each narwhal is displayed in Figure~\ref{fig: realexpthroughtime}.
 
@@
 		\centering
-	\begin{subfigure}{0.49\textwidth}
-		\centering
-		\includegraphics[scale=0.40]{images/data_exploration/realExpShip_through_time.png}
-		\caption{}
-	\end{subfigure}
-	\begin{subfigure}{0.49\textwidth}
-		\centering
-		\includegraphics[scale=0.40]{images/data_exploration/realExpShore_through_time.png}
-		\caption{}
-	\end{subfigure}
-	\caption{Covariates of exposure and distance to shore over time for each narwhal.
-Time 0 corresponds to 12 hours after tagging for each individual whale.
-(a) Exposure to the ship (in $km^{-1}$).
-(b) Inverse of the distance to the shore (in $km^{-1}$).}
+		\includegraphics[scale=0.50]{images/data_exploration/realExpShip_through_time.pdf}
+
+	\caption{Covariate $E^{ship}$ (in $km^{-1}$) over time for each narwhal. Time 0 corresponds to 12 hours after tagging for each individual whale.}  
 	\label{fig: realexpthroughtime}
@@
 
-The dynamics of the true positions of the narwhals is modelled by an SDE.
-First, we recall the standard rotational CVM, and then we extend it to include the effect of the distance to shore on the movement and constrain the position within a polygon.
+The dynamics of the true positions of the narwhals is modelled by a SDE.
+First, we recall the standard rotational CVM \citep{gurarie_correlated_2017}, and then we extend it to include the effect of the distance to shore on the movement and constrain the position within a polygon.
 
@@
 In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \citep{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.
-Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
+We will later model $\omega$ as varying over time as a function of the distance to shore and the angle $\Theta$.
+Whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study.
 Typically, it makes sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source.
-We refer to~\eqref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) when $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) when $\mu \neq (0,0)$.\\
+We will later consider $\mu=0$ in our model.\\
 
@@
 
-Here, we show that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in~\eqref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula.
-This is stated in the following proposition.\\
+Here, we derive a closed form formula for the transition density of the Markov process $(X ,V)$ under the hypothesis of isotropic movement, with diagonal diffusion matrix as in~\eqref{eq: RACVM equation}.\\
 
@@
 \end{equation}
-where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$ with  $f_{\omega}$ a smooth function from $\R^{+} \times [-\pi,\pi]$ to $\R$.
+where $\omega(X(t),V(t))=f(D^{shore}(t),\Theta(t))$ with  $f_{\omega}$ a smooth function from $\R^{+} \times ]-\pi,\pi]$ to $\R$.
 In the sequel, such models will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).\\
@@
 To induce a rotation when the animal is close to the boundary and heading in the direction of the shore,
-the following assumptions must guide the shape of $f_{\omega}$:
+the following assumptions must guide the shape of $f$:
 \begin{itemize}
-	\item[1)] for a distance to the shore lower than some threshold, $f_{\omega}(D^{shore},\cdot)$ should be positive and increasing on $]\frac{\pi}{2},\pi[$, 
-	\item[2)] for a distance to shore  lower than some threshold, $f_{\omega}(D^{shore},\cdot)$ should be negative and decreasing on $]-\pi,-\frac{\pi}{2}[$,
-	\item[3)] as distance to shore decreases, the magnitude of the functions  $f_{\omega}(D^{shore},\cdot)$ on $]\frac{\pi}{2},\pi[$  and $]-\pi,-\frac{\pi}{2}[$ should increase.
+	\item[1)] for a distance to the shore lower than some threshold, $f(D^{shore},\cdot)$ should be positive and increasing on $]\frac{\pi}{2},\pi]$, 
+	\item[2)] for a distance to shore  lower than some threshold, $f(D^{shore},\cdot)$ should be negative and decreasing on $]-\pi,-\frac{\pi}{2}[$,
+	\item[3)] as distance to shore decreases, the magnitude of the functions  $f(D^{shore},\cdot)$ on $]\frac{\pi}{2},\pi[$  and $]-\pi,-\frac{\pi}{2}[$ should increase.
 \end{itemize}
 
-We start by proposing a parametric model for the function $f_{\omega}$ and then discuss a non-parametric representation.
-For $\Theta \in [-\pi,\pi]$ and $D^{shore} \geq0$, we define $f_{\omega}$ as follows
+We propose a parametric model for the function $f$. For $\Theta \in ]-\pi,\pi]$ and $D^{shore} \geq 0$, we define $f$ as the sum of a repulsive term and an attractive term :
 
@@
 	\label{eq: smooth omega}
-	f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\Theta+\theta^{*}))+\tanh(\lambda(\Theta-\theta^{*})))\right)\times \exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)
-\end{equation}
-for some $D_0>0$, $\kappa>0$, $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,\pi]$.
-The parameter $\omega_0$ controls the magnitude of the angular velocity, $D_0$ controls the distance at which the velocity starts to rotate to avoid the boundary, $\lambda$ and $\kappa$ control the steepness of the curves in each dimension, and $\theta^*$ is chosen so that the rotation starts when the angle $\Theta$ gets close to $\frac{\pi}{2}$ or $-\frac{\pi}{2}$.
-\\
-
-This parametrization of $\omega$ is motivated by the observations above, and do not pretend to be optimal. 
-Any smooth function, including~\eqref{eq: smooth omega}, could be approximated by a finite number of weighted spline basis functions of the form
-
-\[f_\omega(\Theta,D^{shore})=\sum_{k=1}^L \omega_k \psi_k(\Theta,D^{shore})\]
-The $\psi_k, k=1, \ldots, L$ can be constructed in several ways based on thin plate regression splines, or on the tensor product of any univariate spline basis.
-The coefficients $\omega_k$ control the contribution of each basis function to the final shape of the smooth function \citep{wood_generalized_2017}.
-Figure~\ref{fig: combinedperspplots}  shows function~\eqref{eq: smooth omega} in the top, along with a smooth function $f_{\omega}$ obtained as a combination of $L=9$ cubic spline basis functions approximating~\eqref{eq: smooth omega} in the bottom plots.
+	f(\Theta,D^{shore})=g(\Theta,D^{shore})+h(\Theta,D^{shore}))
+\end{equation}
+where 
+\[g(\Theta,D^{shore})=a\theta\left(\theta-\frac{\pi}{2}\right)\left(\theta+\frac{\pi}{2}\right)\times \frac{1}{D_{shore}}\exp\left(-\frac{D_{shore}}{D_0}\right)\]is the repulsive term and \[h(\Theta,D^{shore})=b\exp\left(-\left(\begin{pmatrix} \Theta & D^{shore} \end{pmatrix}-m\right) \Sigma^{-1} \left(\begin{pmatrix} \Theta & D^{shore}\end{pmatrix}-m)^\top \right)\right)\]
+is the attractive term.
+Here, $ m=(\frac{\pi}{2\sqrt{3}},D_1)$, $\Sigma=\begin{pmatrix} \sigma_D^2 & 0 \\ 0 & \sigma_\Theta^2\end{pmatrix}$ and $D_a,D_b>0$, $a,b>0$ , $\sigma_D,\sigma_\Theta >0$ are parameters.
+The parameter $a$ controls the magnitude of the angular velocity, $D_a$ controls the distance at which the velocity starts to rotate to model repulsion from the boundary and $D_b$ controls the distance at which the velocity starts to rotate to model attraction to the boundary.
+Then, $ \sigma_D$ and $\sigma_\theta$ are the standard deviations in the gaussian kernels.
+Figure~\ref{fig: perspectiveplots}  shows function~\eqref{eq: smooth omega}.
 
@@
 	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\includegraphics[scale=0.25]{images/crcvm/combined_persp_plots.png}
-		\caption{}
-		\label{fig: combinedperspplots}
-	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\raisebox{0.5cm}{
-			\includegraphics[scale=0.25]{images/crcvm/smooth_tau_bump.pdf}
-		}
-		\caption{}
-		\label{fig: smoothtau}
-	\end{subfigure}
-\caption{(a) Examples of smooth functions $f_{\omega}$.
-Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.
-Top: Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min.
-Bottom: Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions approximating~\eqref{eq: smooth omega}.
-(b) Example of a smooth parameter $\tau$ to get persistent trajectories along the shore.
-The function used is $\tau_0+\frac{\tau_1-\tau_0}{2}\times \left(\tanh(\frac{\Theta+\pi/2+\varepsilon}{\alpha})-\tanh(\frac{\Theta+\pi/2-\varepsilon}{\alpha})+\tanh(\frac{\Theta-\pi/2+\varepsilon}{\alpha})-\tanh(\frac{\Theta-\pi/2-\varepsilon}{\alpha})\right)$ for some parameters $\tau_0$, $\tau_1$, $\varepsilon$, $\alpha$.}
+		\includegraphics[scale=0.4]{images/crcvm/perspective_plots.png}
+
+\caption{Example of smooth functions $f_{\omega}$ defined by~\eqref{eq: smooth omega}.
+Angular velocity $\omega$ increases in absolute value as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore.
+Here, angle $\Theta$ is in $rad$ and $D^{shore}$ in $km$.
+Parameter values are $a=1$, $b=5$, $D_0=0.3$ km,$D_1=0.8$ km, $\sigma_{\theta}=\pi/8$ rad , $\sigma_D=0.2$ km.}
+\label{fig: perspectiveplots}
 
@@
 
-Model~\eqref{eq: CRCVM equation} is flexible, and different species-specific behaviours can be described.
-Function $f_{\omega}$ controls at which distance the animal will turn away from the shore, and how fast it will turn.
-Specific constraints on the smooth function $f_{\omega}$ can be set from biological knowledge about the species in question.
-
-The tendency of an animal to move along the boundary of its spatial domain can be captured by the parameter $\tau$, as more persistent movement when the angle $\theta$ is close to $\pm\frac{\pi}{2}$ would induce such behavior.
-More precisely, to model persistent movement along the shoreline, we modify~\eqref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function:
-\[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] with $f_{\tau}:[-\pi,\pi] \longrightarrow \R$.
-The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain.
-An example of such a function is given in Figure~\ref{fig: smoothtau}.
-However, for simplicity we will assume $\tau$ constant in our analysis of the data.\\
+Compared to the reflected CVM in \cite{hanks_reflected_2017}, model~\eqref{eq: CRCVM equation} is very flexible, and different species-specific behaviours can be described.
+The function $f$ controls at which distance the animal will turn away from the shore, and how fast it will turn.
+Specific constraints on $f$ can be set from biological knowledge about the species in question.
 
@@
 But it is possible to get an approximate solution.
-We choose a time step $\Delta$ and approximate $D^{shore}$ and $\Theta$ by piecewise constant functions on each time step.
+We choose a small time step $\Delta$ and approximate $D^{shore}$ and $\Theta$ by piecewise constant functions on each time step.
 We then use the formulas of Section~\ref{section: RACVM} to simulate the process and get the next position from the approximate transition densities.
-Sampled trajectories obtained within circular, rectangular and fjord domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown in Figure~\ref{fig: crcvm examples}.
-Depending on the smooth functions $f_{\omega}$ and $f_{\tau}$, the trajectories near the boundary have different characteristics.
-The trajectories labelled "persistent" tend to keep the same direction along the boundary, while the trajectories labelled standard turn whenever they are about to hit the boundary, and behave like an unconstrained CVM with $\omega=0$ otherwise.
-The trajectories labelled "tortuous" also turn when going toward the shore, but keep turning when leaving the shore.\\
+Simulated trajectories obtained within the fjords domain are shown in Figure~\ref{fig: crcvm examples} and compared to trajectories simulated from a reflected CVM \citep{hanks_reflected_2017}.
+Trajectories of our model are effectively constrained and show different features from the reflected CVM : movement along and close to the boundary is favoured.
+\\
 
 \begin{figure}[ht!]
-	\includegraphics[scale=0.4]{images/crcvm/final_combined_plot.png}
-
-	\caption{Five days trajectories with time step $\Delta=5$ min within circular (bottom), rectangular (middle) and fjords (top) domains.
-Plots labelled "Persistent CRCVM" and "Standard CRCVM" have the same smooth parameter $\omega$ defined by~\eqref{eq: smooth omega}, whereas the persistent trajectory has a smooth varying parameter $\tau$ as in Figure~\ref{fig: smoothtau}.
-The standard trajectory has a constant $\tau =1$h.
-Trajectories labelled "tortuous" are obtained from a constant $\tau=1$h and a smooth parameter $\omega$ defined with spline basis functions as in Figures~\ref{fig: combinedperspplots}.
-All trajectories have constant velocity parameter $\nu=4$km/h.}
+	\includegraphics[scale=0.42]{images/crcvm/illustrative_trajectories_fjords.pdf}
+
+	\caption{Three days long simulated trajectories for $6$ different initial positions within the fjords domain for the CRCVM defined by equation~\eqref{eq: CRCVM equation} (top) and the reflected CVM in \cite{hanks_reflected_2017} (bottom).
+Trajectories are simulated with fixed time step $\Delta= 10 $s and then subsampled to $10 $ min.
+All trajectories have constant velocity parameter $\nu=4$ km/h and constant persistence $\tau=2$ h.}
  \label{fig: crcvm examples}
@@
 A critical parameter in this model is the spatial scale of the rotational movement, defined as $\rho=\frac{\nu}{\vert \omega \vert}$, representing the diameter of rotation.
-To prevent the animal from hitting the boundary, $\rho$ should remain smaller than $D^{shore}$.\\
-
-We will use the CRCVM \eqref{eq: CRCVM equation} allowing $\omega$ to vary to analyze narwhals trajectories while considering spatial constraints, whereas we keep $\tau$ fixed.
-In section~\ref{section: simulation constrained motion}, we show that a smooth function $f_{\omega}$ depending on $D^{shore}$ and $\Theta$ can be estimated from discrete time obsevations of the process $X$.
+We hint that $\rho$ should remain smaller than $D^{shore}$ in order to reduce the probability of the process hitting the boundary, .\\
+
+We will use the CRCVM \eqref{eq: CRCVM equation} allowing $\omega$ to vary to analyze narwhals trajectories while considering spatial constraints, whereas we keep $\tau$ and $\nu$ fixed.
+We detail the statistical model in the next section.
 
@@
 
-\subsection{Mixed effect CRCVM model for  shore influence and sound exposure}
+\subsection{Mixed effect CRCVM for  shore influence and sound exposure}
 \label{subsection: ship exposure effect}
@@
 
-For the baseline model, only the effect of the shore is included through the covariates $I^{shore}$ and $\Theta$.
+For the baseline model, only the effect of the shore is included through the covariates $D^{shore}$ and $\Theta$.
 For each narwhal $i \in \{1, \cdots, N\}$ and each time $j\in\{1, \cdots, n_{i,pre}\}$, $y_{ij}$ is an observation with measurement error of position $X_{ij}$:
@@
 \end{equation}
-%The process  $V_{ij}$ is  unobserved. 
+
 The standard deviation $\sigma_{obs}$ represents GPS measurement errors variability.
@@
 Furthermore, the error can differ between the $x$ and $y$ directions.
-For a comprehensive study on Fastloc-GPS measurement errors, we refer to \citep{wensveen_path_2015}.
+For a comprehensive study on Fastloc-GPS measurement errors, we refer to \cite{wensveen_path_2015}.
 It is often more appropriate to use a distribution with heavier tails, such as the Student's $t$-distribution, to model measurement error.
-However, this introduces additional complexity to the inference process, as a standard linear Kalman filter cannot be directly applied in such cases.
+However, this introduces additional complexity to the inference process, as a standard linear Kalman filter introduced by \cite{johnson_continuous_2008}
+cannot be directly applied in such cases.
 Since the Student's $t$-distribution approaches a Gaussian distribution as its degrees of freedom increase, the Gaussian distribution is considered a reasonable trade-off between simplicity and suitability.\\
@@
 %We denote $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the CRCVM for narwhal $i$. 
-The latent processes $X_i$ and $V_i$ solve
+The latent processes $X_i$ and $V_i$ are solutions do
 
@@
 
-Finally, as discussed in Section \ref{section: CRCVM}, the angular velocity function of individual $i$, $\omega_i$, is expressed as a smooth function of $\Theta$ and $I^{shore}$ through splines: 
+Finally, as discussed in Section \ref{section: CRCVM}, the angular velocity $\omega_i$ for individual $i$ is expressed as a smooth function of $\Theta$ and $D^{shore}$ : 
 %
 \begin{equation} 
-	\omega_i(t)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(I_i^{shore}(t),\Theta_i(t))
-	\label{eq: baseline splines}
-\end{equation}
-
-Since the two covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines are more appropriate than thin plate regression splines.
-The functions $\psi_k$ in equation~\eqref{eq: baseline splines} are the bivariate basis functions constructed from univariate cubic spline basis functions.
-We refer to \citep{wood_generalized_2017} section $4.
-1.8$ for the construction of tensor splines basis functions.
-The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the number of knots in the splines are $q_E-1$ and $q_{\Theta}-1$ for the covariates $I_{shore}$ and $\Theta$, respectively.\\
+	\omega_i(t)=f(D^{shore}(t),\Theta(t);a,b,D_a,D_b,\sigma_D,\sigma_\Theta)
+\end{equation}
+where the function $f$ is defined in equation \eqref{eq: smooth omega}. The parameters $a$,$b$,$D_a$,$D_b$, $\sigma_D$ and $\sigma_\Theta$ will be considered as fixed and known.
+\\
 
@@
 
-\subsection{Linear Gaussian state-space model}
+\subsection{Approximate linear Gaussian state-space model}
 \label{section: state space model}
-In the baseline statistical model \eqref{eq:baseline latent processes}, we make the approximation suggested in \citep{michelot_varying-coefficient_2021} that the function $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is
-\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(I_{ij}^{shore},\Theta_{ij}).\]
-We write %$\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, 
-$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
-	\frac{1}{\tau_{i}} && -\omega_{ij} \\
-	\omega_{ij} && \frac{1}{\tau_{i}}
-\end{pmatrix}$ and $U_{ij}=\begin{pmatrix} X_{ij,1}  && X_{ij,2} && V_{ij,1} && V_{ij,2}\end{pmatrix}^\top$.
-% where $\log(\tau_i)=\log(\tau_0)+b_{\tau}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu}^{(i)}$.
-%
-We use the exact formulas from Proposition \ref{prop: transition density} to obtain the state-space matrix equations for $i\in\{1,\ldots, N\}$, and $j\in\{1, \ldots, n_{i,pre}\}$:%.
-%Discretisation of~\eqref{eq: X V distribution} gives the following system:
-
+
+In the baseline statistical model \eqref{eq:baseline latent processes}, we make the approximation suggested in \citep{michelot_varying-coefficient_2021} that the time-varying parameter $\omega_i$ is piecewise constant.
+On each time step $[t_{ij},t_{i,j}+\Delta_{ij}]$ between two observations, we interpolate valus of $D^{shore}_i$ and $\Theta_i$ at times $t_{ij}^{(0)} = t_{ij} < t_{ij}^{(1)} < \cdots < t_{ij}^{(k)}=t_{ij}+\Delta_{ij}$ by fitting splines on the observed trajectories.
+
+We write $U_{ij}=\begin{pmatrix} X_{ij,1}  && X_{ij,2} && V_{ij,1} && V_{ij,2}\end{pmatrix}^\top$ and for $l \in \{0,\cdots,k\}$,
+\[\omega_{ij}^{(l)}=\omega(t_{ij}^{(l)}), \quad   A_{ij}^{(l)}=\begin{pmatrix} 
+	\frac{1}{\tau_{i}} && -\omega_{ij}^{(l)} \\
+	\omega_{ij}^{(l)} && \frac{1}{\tau_{i}}
+\end{pmatrix} \]
+
+We use the formulas from Proposition \ref{prop: transition density} to obtain an approximation of the state-space matrix equations for each individual $i\in\{1,\ldots, N\}$, and each observation $j\in\{1, \ldots, n_{i,pre}\}$ when $\omega$ varies with time.
+We write $T_{ij}^{(l)}=T(\Delta_{ij}^{(l)})$ and $Q_{ij}^{(l)}=Q(\Delta_{ij}^{(l)})$ as defined in \eqref{eq: link matrices} and \eqref{eq: covariance matrix}.
+Then the approximate transition density is 
+	\begin{equation}
+	U_{i,j+1} \vert U_{ij}=u \sim \mathcal{N}\left( \tilde{T}_k(\Delta_{ij}) u, \tilde{Q}_k(\Delta_{ij})\right)
+	\label{eq: approximate transition density}
+\end{equation}
+where 
+    \begin{equation}
+    \tilde{T}_k(\Delta) = T_{k-1} \times \cdots \times T_1 \times T_0 
+    \end{equation}
+and 
+    \begin{equation}
+        \tilde{Q}_k(\Delta) =\sum_{l=1}^{k} T_{k-1}^\top \cdots T_{l+1}^\top Q_l T_{l+1} \cdots T_{k-1}
+    \end{equation}
+
+Thus we obtain the following approximate linear state-space model 
 \begin{equation}
@@
 		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
-		U_{i,j+1}=T_{ij}\,U_{ij} + \eta_{ij}, \quad 
-		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
+		U_{i,j+1}=\tilde{T}_{ij}\,U_{ij} + \eta_{ij}, \quad 
+		\eta_{ij} \sim \mathcal{N}(0,\tilde{Q}_{ij})
 	\end{array}
@@
 \end{equation}
-where $Z=\begin{pmatrix} I_2 && 0_{2,2}\end{pmatrix}$, the link matrix $T_{ij}$ is computed according to~\eqref{eq: link matrices} and the covariance matrix $Q_{ij}$ is computed according to~\eqref{eq: covariance matrix}.
-\\
-A linear Gaussian state-space model is obtained similarly for the response model. Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation.
+where $Z=\begin{pmatrix} I_2 && 0_{2,2}\end{pmatrix}$, $\tilde{T}_{ij}=\tilde{T}_k(\Delta_{ij})$ and $\tilde{Q}_{ij}=\tilde{Q}_k(\Delta_{ij})$.
+\\
+An approximate linear Gaussian state-space model is obtained similarly for the response model. Estimation with $\texttt{smoothSDE}$ relies on this state-space formulation.
 The complete likelihood is computed from the observations $y_{ij}$ as a by-product of the Kalman filter algorithm \citep{michelot_varying-coefficient_2021}.
@@
 We refer to \citep{kristensen_tmb_2016} for more details about the derivation of the multidimensional Laplace's approximation and the gradient computations.
-We refer to \cite[section 6.6]{wood_generalized_2017} for the specification of the mixed effect model and the regularization terms that are based on a wiggliness measure involving Gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$.
-\\
 This method is already implemented for the CVM ($\omega=0$) in \texttt{smoothSDE}.
@@
 \label{section: simulation constrained motion}
-To test the estimator of the constrained model for our application, we simulate CRCVM trajectories in the Scoresby Sound fjord system.
+To test the estimator of the constrained model for our application, we simulate CRCVM trajectories over one day in the Scoresby Sound fjord system.
 We consider observations of a baseline model as described in Section~\ref{subsection: ship exposure effect}.
-We set $L=9$ degrees of freedom for the bivariate splines in the parameter $\omega$ and  we place the knots evenly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $1$ and $5$ km for $D^{shore}$.
-The values of the spline coefficients are chosen to get a process that turns fast enough to meet the spatial constraints.
 We fix $\tau_0=1.5$ h and $\nu_0=4$ km/h.
 The random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$.
-Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in the fjord system.
-\\
-
-Two setups are considered: high frequency data with short trajectories and low frequency data with longer trajectories.
+The smooth function $\omega$ is defined as in equation \eqref{eq: smooth omega} with parameter values $D_0=0.5$,$D_1=1$,$a=2$,$b=4$,$\sigma_\theta=\frac{\pi}{8}$, $\sigma_D=0.3$.
+Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled between $0.1$ and $1$ km from the shore in the fjord system.
+$1.02\%$ of the simulated trajectories reached land.
+The batches containing at least one trajectory that reached land were not used for the estimation.\\
+
+Two scenarios are considered: an ideal scenario where we have high frequency data with either high or low measurement error, and a practical scenario where the data is low frequency and measurement errors are high.
 For each setup, we simulate $M=100$ batches of $N$ individual trajectories with $N \in \{6,12\}$.
-In the high frequency framework, the time step between consecutive observations is $\Delta=1$ min, the overall observation time is $12$h and measurement error is Gaussian with standard deviation $\sigma_{obs}=5$ m.
-This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.
-In the low frequency framework, the trajectories are simulated with one minute time steps and then subsampled to a time step $\Delta=5$ min.
-The overall observation time is $2.5$ days and measurement error is Gaussian with standard deviation $\sigma_{obs}=25$ m.
-This also amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$.\\
-
-For the estimation, we only keep the trajectories whose observed distances to the shore in km include the interval $[1,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$.
-The reason is that if a trajectory happens to be confined at a few hundreds metres from the shore in a corner of the fjords there is no sense in trying to estimate its movement $5$ km from the shore, and conversely, if the trajectories occur in open sea far from the shore, there is no sense in trying to estimate the movement characteristics $1$ km from the shore.\\
-
-For the smooth function $\omega$, $9$ coefficients, including the intercept, are to be estimated, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the random effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$.
-The measurement error $\sigma_{obs}$ is supposed to be known, usually given by the known GPS precision.
-We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$ in the optimization. 
-For each batch of $N$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard error of the estimates.
-Simulation and optimization of the log-likelihood for one single batch of trajectories are performed on $16$ CPU cores within less than $2$ h.
-Parallelization within \texttt{R} is used to simulate the trajectories since it can be time consuming due to the calculation of the nearest shore point for each new sampled position.
-\\
-
-The results are summarized in Table~\ref{table: simulation study results}. We show the mean and the standard error of the estimates for each coefficient in the high frequency and low frequency cases.
-
-Estimation of the spline coefficients has overall less than $20 \%$ bias in the high frequency framework, and becomes more precise as $N$ increases.
-The variances of the random effects are the most challenging parameters to estimate.
-At least one of the standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$ is often estimated close to $0$.
-This tendency is clear for $N=6$ and $N=12$ individuals, though it seems to attenuate as $N$ increases.
-This is expected, since $N$ is the number of independent observations used for estimating these standard deviations.
-Regarding the intercepts, we obtain better estimates for the velocity parameter $\log(\nu_0)$ than for the persistence parameter $\log(\tau_0)$.
-The biases are about $4 \%$ and $15\%$, respectively,  and the standard error is lower for $\log(\nu_0)$.
-
-Estimation is more difficult in the low frequency framework with higher measurement errors.
-Indeed, as the measurement error grows, the covariates are becoming less accurate, which clearly makes estimation less precise.
-In particular, we are unable to estimate the intercept $\log(\tau_0)$, as it remains close to the initial value.
-However, in practice this can be solved step by step by trying several initial values to find the coefficients giving the highest log-likelihood value.
-Some of the spline coefficients cannot be estimated as well.
-Perhaps surprisingly, the estimates of the random effects standard deviations seem slightly better in the low-frequency framework.\\
-
-In practice, confidence intervals are often derived using the observed Fisher information matrix, which is done in Section~\ref{section: application}. 
-It requires the Hessian matrix of the log-likelihood at the estimated coefficients to be positive definite. If this is not the case, this may indicate numerical errors or improper convergence.
-
-\begin{landscape}
-
-	\begin{table}
+The time steps between consecutive observations are set $\Delta_{high}=1$ min for high frequency and $\Delta_{low}=5$ min for low frequency.
+The trajectories are simulated with time step $\Delta =1$ min and then subsampled to $\Delta_{high}$ for low frequency observations.
+Measurement noise is Gaussian with standard deviation $\sigma_{obs,high}=40$ m and $\sigma_{obs,low}=10$ m for high and low measurement errors.
+They are supposed to be fixed and known in this simulation study.
+
+For the estimation, we fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=0.5$, $\sigma_{\nu}=0.5$. 
+For each batch of $N$ trajectories, we obtain estimates $\widehat{\tau_{0}}^{(k)}$, $\widehat{\nu_{0}}^{(k)}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard error of the estimates.
+Simulation and optimization of the log-likelihood for one single batch of trajectories in all scenarios are performed on $16$ CPU cores in about $5$ h.
+Parallelization within \texttt{R} is used to simulate the trajectories, compute the angle $\theta$ and the distance $D^{shore}$, and fit the models in the different scenarios.
+\\
+
+We estimated all the parameters while considering the smooth function $\omega$ to be known.
+We used the true $D^{shore}$ and $\theta$, the observed noisy $D^{shore}$ and $\theta$ and smoothed $D^{shore}$ and $\theta$ obtained after fitting splines on the trajectories with \texttt{gam} function in \texttt{R} software.
+The results are shown in Table~\ref{table: simulation study hf parametric}.
+This simulation study shows that the main difficulty in the estimation is the parameter $\tau_0$.
+While we are able to get an unbiaised estimate with the true $D^{shore}$ and $\theta$ from high frequency observations, both the observed and smoothed $D^{shore}$ and $\theta$ yield biaised estimates of $\tau_0$.
+We also show the estimates obtained for the SDE model with a constant $\omega=0$, which corresponds to the CTCRW in \cite{johnson_continuous_2008}.
+
+	\begin{table}[ht!]
 		\centering
-		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-			\hline
-			&\multicolumn{3}{|c|}{Intercepts} & \multicolumn{8}{|c|}{Spline coefficients} & \multicolumn{2}{|c|}{SD} \\
-			\hline
-			&$\log(\tau_0)$ & $\log(\nu_0)$ & $\omega_0$ & $\omega_1$ & $\omega_2$ & $\omega_3$ & $\omega_4$  & $\omega_5$ & $\omega_6$ & $\omega_7$ & $\omega_8$ & $\sigma_{\tau}$ & $\sigma_{\nu}$ \\
-			\hline 
-			True & $0.41$ & $1.39$ & $0.02$ & $8.87$& $-2.54$& $6.72$&
-			$22.50$& $6.73$& $30.88$& $4.78$& $6.62$& $0.2$ & $0.1$ \\
-			\hline
-			\multicolumn{14}{|c|}{High frequency} \\
-			\hline
-			\multirow{2}{4em}{$N=6$} & $0.35$ &
-			$1.34$ &
-			$-0.04$ &
-			$7.68$&
-			$-2.23$&
-			$6.89$&
-			$20.38$&
-			$4.57$&
-			$28.11$&
-			$4.24$&
-			$5.04$&
-
-			$0.20$&
+		\begin{tabular}{|c|c|cccc|}
+			\hline
+			     & &\multicolumn{2}{|c}{\textbf{Intercepts}}  & \multicolumn{2}{c|}{\textbf{Random effects}} \\
+			\cline{3-6}
+			     \textbf{Model type} & &$ \tau_0 $ (h) & $\nu_0 $ (km/h)& $\sigma_{\tau}$ & $\sigma_{\nu}$  \\
+			\cline{2-6} 
+			& True & $1.50$ & $4.00$ & $0.2$ & $0.1$ \\
+			\hline
+			\multicolumn{6}{|c|}{\textbf{Low measurement error} ($\sigma_{obs} =10$ m)} \\
+			\hline
+            \multirow{4}{6em}{CTCRW} & \multirow{2}{4em}{$N=6$}  & $1.05$ &
+			$4.07$ &
+			$0.20$ & $0.08$ 
+			\\
+			& & $\pm 0.11$ &
+			$\pm 0.21$ &
+			$\pm 0.11$ & 
+			$\pm 0.05$
+
+			\\
+			%\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $1.04$ &
+			$4.07$ &
+			$0.23$ & $0.09$ 
+			\\
+			& & $\pm 0.08$  &
+			$ \pm 0.11$ &
+			$ \pm 0.10 $ & $ \pm 0.03$ 
+            \\
+            \hline
+			\multirow{4}{6em}{CRCVM \hspace{0.1cm} True} & \multirow{2}{4em}{$N=6$} & $1.42$ &
+			$4.09$ &
+			$0.23$ &
+			$0.12$
+
+			\\
+			& &	$[1.20,1.74]$ &
+			$ [3.77,4.49] $&
+			$[0.17,0.28]$ & $ [0.09,0.28]$ 
+			\\
+            %\cline{2-6} 
+			& \multirow{2}{4em}{$N=12$} & $1.48$ &
+			$4.03$ & $0.25$ & $0.10$
+
+			\\
+			& & $[1.29,1.72]$ &
+			$  [3.79,4.33]$ &
+			$ [0.18,0.30]$ &
+			$ [0.08,0.17]$ 
+			\\
+            \hline
+            \multirow{4}{6em}{CRCVM Observed} & \multirow{2}{4em}{$N=6$} & $1.16$ &
+			$4.12$ &
+			$0.24$ &
 			$0.13$
@@
 			\\
-			&	$\pm 0.16$ &
-			$\pm 0.09$&
-			$\pm 0.23$ & $\pm 1.37$ &
-			$\pm 1.59$ &
-			$\pm 1.74$ &
-			$\pm 2.56$ &
-			$\pm 3.67$ &
-			$\pm 1.51$ &
-			$\pm 1.27$ &
-			$\pm 1.59$ 
-			& $\pm 0.16$ &
-			$\pm 0.08$
-			\\
-			\hline
-			\multirow{2}{4em}{$N=12$} & $0.35$ &
-			$1.42$ & $0.00$ & $7.54$&
-			$-2.04$ &
-			$7.02$ &
-			$20.91$ &
-			$5.33$ &
-			$28.59$ &
-			$4.74$ &
-			$5.48$
-
-			& $0.14$ &
-			$0.09$
-
-			\\
-			& $ \pm 0.10$ &
-			$ \pm0.05$ &
-			$\pm 0.22$ &
-			$\pm 0.94$ &
-			$\pm 1.25$ &
-			$\pm 1.10$ &
-			$\pm 1.59$ &
-			$\pm 2.70$ &
-			$\pm 1.34$ &
-			$\pm 0.86$ &
-			$\pm 1.09$ &
-
-			$\pm 0.09$ &
-			$\pm 0.05$ 
-			\\
-			\hline	
-			\multicolumn{14}{|c|}{Low frequency} \\
-			\hline
-			\multirow{2}{4em}{$N=6$}  & $0.06$ &
-			$1.36$ &
-			$0.00$ & $5.67$ &
-			$0.78$&
-			$6.20$ &
-			$16.31$ &
-			$6.26$ &
-			$25.19$ &
-			$4.52$ &
-			$3.92$ &
-			$0.22$ &
-			$0.14$
-			\\
-			& $\pm 0.11$ &
-			$\pm 0.03$ &
-			$\pm 0.09$ & 
-			$\pm 2.11$ &
-			$\pm 2.45$ &
-			$\pm 1.36$ &
-			$\pm 6.33$ &
-			$\pm 3.90$ &
-			$\pm 2.96$ &
-			$\pm 1.61$ &
-			$\pm 2.03$ &
+			& &	$[0.93,1.44]$ &
+			$[3.76,4.50]$&
+			$[0.16,0.36]$ & $[0.07,0.30]$ 
+			\\
+            %\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $1.15$ &
+			$4.01$ & $0.25$ & $0.10$
+
+			\\
+			& & $ [1.01,1.30]$ &
+			$ [3.75,4.25]$ &
+			$[0.17,0.34]$ &
+			$[0.07,0.16]$ 
+			\\
+			\hline	
+			\multirow{4}{6em}{CRCVM Smoothed} & \multirow{2}{4em}{$N=6$}  & $1.27$ &
+			$4.10 $ &
+			$0.24$ & $0.14$ 
+			\\
+			& & $[1.02,1.51]$ &
+			$[3.74,4.47]$ &
+			$[0.15,0.36]$ & 
+			$[0.08,0.40]$
+
+			\\
+			%\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $1.29$ &
+			$4.03$ &
+			$0.28$ & $0.12$
+
+			\\
+			& & $[1.11,1.42]$  &
+			$ [3.85,4.20]$ &
+			$ [0.18,0.39]$& $[0.07,0.18] $  \\
+            \hline
+			\multicolumn{6}{|c|}{\textbf{High measurement error} ($\sigma_{obs} =40$ m)} \\
+			\hline
+            \multirow{4}{6em}{CTCRW} & \multirow{2}{4em}{$N=6$}  & $1.05$ &
+			$4.06$ &
+			$0.17$ & $0.09$ 
+			\\
+			& & $\pm 0.10$ &
+			$\pm 0.21$ &
 			$\pm 0.13$ & 
 			$\pm 0.05$ 
-
-			\\
-			\hline
-			\multirow{2}{4em}{$N=12$} & $0.04$ &
-			$1.39$ &
-			$-0.02$ & $6.05$ &
-			$-0.11$ &
-			$6.19$ &
-			$17.39$ &
-			$4.03$ &
-			$24.98$ &
-			$5.01$ &
-			$3.40$ & $0.25$ &
-			$0.08 $
-			\\
-			& $\pm 0.06$  &
-			$ \pm 0.03$ &
-			$ \pm 0.11 $ & $ \pm 2.04$ &
-			$ \pm 2.24$&
-			$\pm 0.78$&
-			$\pm 6.12$&
-			$\pm 4.11$&
-			$\pm 1.40$&
-			$\pm 1.74$&
-			$\pm 1.01$& $\pm 0.08$ &
-			$\pm 0.03$
-
-			\\
-			\hline	
+			\\
+			%\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $1.04$ &
+			$4.04$ &
+			$0.17$ & $0.09$ 
+			\\
+			& & $\pm 0.07$  &
+			$ \pm 0.11$ &
+			$ \pm 0.12 $ & $ \pm 0.03$ 
+            \\
+            \hline
+            \multirow{4}{6em}{CRCVM \hspace{0.1cm} True} & \multirow{2}{4em}{$N=6$} & $1.40$ &
+			$4.07$ &
+			$0.24$ &
+			$0.12$
+			\\
+			& &	$[1.15,1.76]$ &
+			$ [3.71,4.49] $&
+			$[0.17,0.60]$ & $ [0.08,0.34]$ 
+			\\
+            %\cline{2-6} 
+			& \multirow{2}{4em}{$N=12$} & $1.43$ &
+			$3.98$ & $0.27$ & $0.10$
+
+			\\
+			& & $[1.27,1.60]$ &
+			$  [3.78,4.20]$ &
+			$ [0.19,0.33]$ &
+			$ [0.07,0.17]$ 
+			\\
+            \hline
+            \multirow{4}{6em}{CRCVM Observed} & \multirow{2}{4em}{$N=6$} & $0.41$ &
+			$4.14$ &
+			$0.46$ &
+			$0.13$
+
+			\\
+			& &	$[0.27,0.61]$ &
+			$[3.71,4.57]$&
+			$[0.17,0.75]$ & $[0.06,0.24]$ 
+			\\
+            %\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $0.40$ &
+			$4.06$ & $0.46$ & $0.13$
+
+			\\
+			& & $ [0.30,0.48]$ &
+			$ [3.72,4.28]$ &
+			$[0.28,0.64]$ &
+			$[0.07,0.21]$ 
+			\\
+			\hline	
+			\multirow{4}{6em}{CRCVM Smoothed} & \multirow{2}{4em}{$N=6$}  & $1.21$ &
+			$4.07 $ &
+			$0.24$ & $0.14$ 
+			\\
+			& & $[1.00,1.42]$ &
+			$[3.71,4.45]$ &
+			$[0.17,0.36]$ & 
+			$[0.08,0.40]$
+
+			\\
+			%\cline{2-6}
+			& \multirow{2}{4em}{$N=12$} & $1.23$ &
+			$4.01$ &
+			$0.27$ & $0.11$
+
+			\\
+			& & $[1.09,1.43]$  &
+			$ [3.82,4.21]$ &
+			$ [0.19,0.36]$& $[0.07,0.18] $  \\
+            \hline
+
 		\end{tabular}
-		\caption{Results from the simulation study.
-Numbers are average $\pm$ standard error of the estimates obtained from $M=100$ simulated data sets in four scenarios: high and low frequency, and $N=6$ or $12$ individuals.
-High frequency are sampled at $\Delta = 1$ min, low frequency is sampled at $\Delta = 5$ min.
-There are 720 observations per individual in each simulated data set.
+        \vspace{0.1cm}
+		\caption{Results for the simulation study in the ideal scenario.
+Numbers are average and $95 \%$ quantiles of the estimates obtained from $M=100$ simulated data sets with high frequency ($\Delta = 1$ min) and with $N=6$ or $12$ individuals.
 The true values used in the simulations are given in the upper row.}
-		\label{table: simulation study results}
+		\label{table: simulation study hf parametric}
 	\end{table}
 
-\end{landscape}
+In the low frequency case, we smoothed the trajectories with splines and computed values of $D^{shore}$ and $\theta$ based on the smooth trajectories to obtain a better approximation of the log-likelihood function.
+
+\begin{table}[ht!]
+		\centering
+		\begin{tabular}{|c|c|cccc|}
+			\hline
+			     & &\multicolumn{2}{|c}{\textbf{Intercepts}}  & \multicolumn{2}{c|}{\textbf{Random effects}} \\
+			\cline{3-6}
+			     \textbf{Model type} & &$ \tau_0 $ (h) & $\nu_0 $ (km/h)& $\sigma_{\tau}$ & $\sigma_{\nu}$  \\
+			\cline{2-6} 
+			& True & $1.50$ & $4.00$ & $0.2$ & $0.1$ \\
+            \hline
+            \multirow{4}{6em}{CTCRW} & \multirow{2}{4em}{$N=6$}  & $1.05$ &
+			$4.06$ &
+			$0.17$ & $0.09$ 
+			\\
+			& & $\pm 0.10$ &
+			$\pm 0.21$ &
+			$\pm 0.13$ & 
+			$\pm 0.05$ 
+			\\
+			& \multirow{2}{4em}{$N=12$} & $1.04$ &
+			$4.04$ &
+			$0.17$ & $0.09$ 
+			\\
+			& & $\pm 0.07$  &
+			$ \pm 0.11$ &
+			$ \pm 0.12 $ & $ \pm 0.03$ 
+            \\
+            \hline
+            \multirow{4}{6em}{CRCVM \hspace{0.1cm} True} & \multirow{2}{4em}{$N=6$} & $0.90$ &
+			$4.12$ &
+			$0.31$ &
+			$0.14$
+			\\
+			& &	$[0.70,1.14]$ &
+			$ [3.76,4.53] $&
+			$[0.19,0.48]$ & $ [0.09,0.23]$ 
+			\\
+			& \multirow{2}{4em}{$N=12$} & $0.90$ &
+			$4.03$ & $0.31$ & $0.12$
+
+			\\
+			& & $[0.75,1.10]$ &
+			$  [3.78,4.27]$ &
+			$ [0.18,0.44]$ &
+			$ [0.08,0.20]$ 
+			\\
+            \hline
+            \multirow{4}{6em}{CRCVM Observed} & \multirow{2}{4em}{$N=6$} & $0.55$ &
+			$4.20$ &
+			$0.37$ &
+			$0.15$
+
+			\\
+			& &	$[0.39,0.72]$ &
+			$[3.77,4.68]$&
+			$[0.16,0.68]$ & $[0.07,0.25]$ 
+			\\
+			& \multirow{2}{4em}{$N=12$} & $0.57$ &
+			$4.09$ & $0.36$ & $0.14$
+
+			\\
+			& & $ [0.45,0.70]$ &
+			$ [3.76,4.35]$ &
+			$[0.18,0.58]$ &
+			$[0.06,0.22]$ 
+			\\
+			\hline	
+			\multirow{4}{6em}{CRCVM Smoothed} & \multirow{2}{4em}{$N=6$}  & $1.14$ &
+			$4.12 $ &
+			$0.27$ & $0.13$ 
+			\\
+			& & $[0.87,1.43]$ &
+			$[3.76,4.61]$ &
+			$[0.17,0.47]$ & 
+			$[0.08,0.21]$
+
+			\\
+			& \multirow{2}{4em}{$N=12$} & $1.14$ &
+			$4.00$ &
+			$0.26$ & $0.12$
+
+			\\
+			& & $[0.97,1.36]$  &
+			$ [3.73,4.27]$ &
+			$ [0.17,0.39]$& $[0.07,0.20] $  \\
+            \hline
+
+		\end{tabular}
+        \vspace{0.1cm}
+		\caption{Results for the simulation study in the practical scenario.
+Numbers are average and $95 \%$ quantiles of the estimates obtained from $M=100$ simulated data sets, with $N=6$ or $12$ individuals.
+The true values used in the simulations are given in the upper row.}
+		\label{table: simulation study lf parametric}
+	\end{table}
 
@@
 
-In this section, we apply our CRCVM diffusion model to analyse the  behavioral response of the narwhals to ship and seismic airgun exposure.
+In this section, we apply our CRCVM to analyse the  behavioral response of the narwhals to ship and seismic airgun exposure.
 The baseline model is fitted on the tracks before exposure to capture normal behavior.
@@
 We estimate the baseline model as described in Section~\ref{subsection: ship exposure effect} on the data before exposure.
-We set the degree of freedom of the marginal splines to $5$ leading to a total number of $25$ degrees of freedom, fix the smoothing penalties of the tensor splines to $1$, and initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=1$, and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ to 0.
-The measurement error standard deviation is fixed to $\sigma_{obs}=50$ m based on the measurement errors of Fastloc-GPS found in \citep{wensveen_path_2015}.\\
+We initialize the random effects standard deviations to $\sigma_{\nu}=\sigma_{\tau}=0.5$, and fix the parameters for the smooth function $\omega$ \eqref{eq: smooth omega} to $D_0= $ km, $D_1 = $ km, which are respectively the $25 \%$ and $50\%$ quantiles of the observed distance to shore, $a=4D_0$, $b=2$, $\sigma_D=2$ and $\sigma_\Theta =0.5$.
+The measurement error standard deviation is also fixed to $\sigma_{obs}=50$ m based on the measurement errors of Fastloc-GPS found in \citep{wensveen_path_2015}.
+\\
 
 Estimation and confidence intervals on the parameter scale are shown in Table~\ref{table: baseline estimations}. 
-The persistence is estimated to $\hat{\tau}_{0}=1.35 $ h.
+The population mean for the persistence parameter is estimated to $\hat{\tau}_{0}=1.34 $ h.
 In comparison, harbour seal in Alaska were shown to exhibit slightly more persistent motion $\hat{\tau}=1.51$ h with $95\%$ CI ; $1.30 -1.75$  \citep{johnson_continuous_2008} while bowhead whales in Greenland showed much less persistence $\hat{\tau}=0.17$ h with $95\%$ CI ; $0.14 -0.20$  \citep{gurarie_correlated_2017}.
-The intercept value for $\nu$ is estimated to $\hat{\nu}_{0}=4.60$ km/h which is of the order of magnitude of the average of the observed velocities. 
+The population mean for the velocity parameter is estimated to $\hat{\nu}_{0}=4.57$ km/h which is of the order of magnitude of the average of the observed velocities.
+
 \begin{table}[ht!]
 	\centering
-	\begin{tabular}{|c|c|c|}
-		\hline
-		Coefficient   & Estimate  & $95\%$ CI \\
-		\hline
-		$\tau_0$ (h)   & $1.36$    & $[1.06; 1.71]$  \\
-		$\nu_0$ (km/h)  & $4.60$   &  $[4.12; 5.12]$\\
-		$\omega_0$ (rad/h)    & $0.00$    &  $[-0.14; 0.13]$ \\
-		$\sigma_{\tau,pre}$    & 0.21   &  $[0.05; 0.56]$ \\
-		$\sigma_{\nu,pre}$    & 0.09   &  $[0.02; 0.24]$ \\
+	\begin{tabular}{lll}
+		\hline
+		Parameter & Estimate & CI \\ 
+		\hline
+		$\tau$ & $1.16$ & $[0.87; 1.51]$ \\ 
+		$\nu$ & $4.74$ & $[4.19; 5.33]$ \\ 
+		$\sigma_{\tau}$ & $0.28$ & $[0.10; 0.63]$ \\ 
+		$\sigma_{\nu}$ & $0.12$ & $[0.04; 0.25]$ \\ 
 		\hline
@@
 
-Figure \ref{fig:febaseline3omegaanglenormalq0} shows the estimated smooth $\omega$ as a function of $\Theta$ for different values of the distance to shore.
-Movement towards the shoreline matches values of $\Theta$ such that $\vert \Theta \vert > \frac{\pi}{2}$.
-At $250$ m away from the shore, there is a clear rotation of the velocity estimated to be $\pm 2.5$ rad/h as $\Theta$ approaches $\pm\pi$.
-As distance increases, the magnitude of the angular velocity decreases, which means that the movement becomes less tortuous.
-This indicates that the spatial constraints influence the tortuosity of the movement by forcing the narwhal's velocity to rotate when they are close to the shoreline.
-%However, the magnitude of the estimated $\omega$ is not sufficient to effectively obtain constrained trajectories similar to Figure~\ref{fig: crcvm examples} when simulating from the fitted model.
-\begin{figure}[ht!]
-	\centering
-	\includegraphics[width=0.7\linewidth]{"images/application/baseline/omega_DistanceShore_levels.pdf"}
-	\caption{Estimated smooth $\omega$ as a function of the angle $\Theta$ for fixed distances to the shore: $250$ metres for close, $1.41$ km for medium, and $3$ km or more for far (no effect of the shore).}
-
-	\label{fig:febaseline3omegaanglenormalq0}
-\end{figure}
-
-Fixing a high standard deviation for the measurement error tends to increase the estimate of the parameter $\tau$.
-Intuitively, for a low value of the measurement error, a tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, and therefore a higher estimate of the persistence parameter.
-We fixed the measurement errors $\sigma_{obs}$ at different values and found that the highest log-likelihood value was obtained for a standard deviation $\sigma_{obs}$ close to $50$m.
-More details can be found in Appendix.
-
 \subsection{Response estimation}
 
-The baseline estimates of the spline coefficients $\omega_k$, the intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the standard deviations $\sigma_{\tau}$, $\sigma_{\nu}$ are used as offsets in the response model.
+The baseline estimates of the the intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the standard deviations $\sigma_{\tau}$, $\sigma_{\nu}$ are used as offsets in the response model.
 Only the sound exposure effects on $\tau$ and $\nu$ are estimated from the data after exposure.
-Estimates and $95\%$ confidence intervals are shown in Table~\ref{table: response estimations}.
-As the coefficients $\alpha_\tau$ and $\alpha_\nu$ significatively deviate from zero, this suggests that sound exposure influences both the persistence and the velocity of the movement.\\
-%However, we miss the uncertainty coming from the baseline estimates here.
-To better quantify this uncertainty, we draw $100$ "posterior" samples of the coefficients of the baseline using the Hessian matrix of the log-likelihood to approximate the covariance, and we use the drawn coefficients as offsets for the response model.
-The mean of the estimates and the standard error over the $100$ estimations are $mean(\hat{\alpha}_\tau)=-3.26$, $mean(\hat{\alpha}_\nu)=0.86$, $mean(\hat{\sigma}_{\alpha_\tau})=0.57$ and $mean(\hat{\sigma}_{\alpha_\nu})=0.27$.
-\\
+The uncertainty stemming from the baseline estimates is considered by drawing $100$ "posterior" samples of the coefficients of the baseline using the Hessian matrix of the log-likelihood to approximate the covariance, and using the drawn coefficients for the estimations of the response model.
+The mean of the estimates and the confidence intervals obtained over the $100$ estimations are shown in Table~\ref{table: response estimations}.
+As the coefficients $\alpha_\tau$ and $\alpha_\nu$ significantly deviate from zero, this suggests that sound exposure influences both the persistence and the velocity of movement.\\
 
@@
 	\centering
-	\begin{tabular}{|c|c|c|}
-		\hline
-		Coefficient   & Estimate  & $95 \%$ CI \\
-		\hline
-		$\alpha_\tau$    & $-3.43$    &  $[-4.77; -2.04]$ \\
-		$\alpha_\nu$  & $0.74$   &  $[0.23; 1.25]$\\
+	\begin{tabular}{lll}
+		\hline
+		Parameter & Estimate & CI \\ 
+		\hline
+		$\alpha_{\tau}$ & $-2.97$ & $[-4.94; -1.39]$ \\ 
+		$\alpha_{\nu}$ & $1.07$ & $[0.49; 1.72]$ \\ 
 		\hline
@@
 
-The estimated $\tau$ and $\nu$ are shown in Figure~\ref{fig: me response} as a function of the distance to the sound source.
+The estimated $\tau$ and $\nu$ are shown in Figure~\ref{fig: population parameters response} as a function of the distance to the sound source.
 The value of $\nu$ increases with increased exposure to the sound.
@@
 This can be interpreted as a drop in persistence due to the appearance of the ship.
-Our analysis suggests that the magnitude of this drop might be alarming: for most of the narwhals, at distance less than $5$ km from the ship, the persistence is half the baseline value.
+For most of the narwhals, at distance less than $5$ km from the ship, the persistence is half the baseline value.
 This is evidence of a strong shift in the behavior of the whales, which might have consequences on their capacity to rest and forage at short term.
@@
 	\begin{subfigure}{0.98\textwidth}\centering
-		\includegraphics[width=0.7\linewidth]{"images/application/response/plot_me_tau_updated.pdf"}
+		\includegraphics[width=0.7\linewidth]{"images/application/response/fe_tau_ExpShip_final.pdf"}
 		\caption{}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth]{"images/application/response/plot_me_nu_updated.pdf"}
+		\includegraphics[width=0.7\linewidth]{"images/application/response/fe_nu_ExpShip_final.pdf"}
 		\caption{}
 	\end{subfigure}
-	\caption{Estimated effect of ship exposure on the parameters.
-The horizontal dashed lines represent the baseline values for each individual.
+	\caption{Estimated effect of ship exposure on the parameters with $95\%$ confidence intervals.
+The horizontal dashed lines represent the baseline values.
 The values on the $x$-axis are in kilometres.
-(a)  Persistence parameter $\tau$.
+(a)  Population persistence parameter $\tau$.
 Values on the $y$-axis are in hour.
-(b) Velocity parameter $\nu$.
+(b) Population velocity parameter $\nu$.
 Values on the $y$-axis are in km/h.}
-	\label{fig: me response}
+	\label{fig: population parameters response}
 \end{figure}
@@
 \end{equation}
-Table~\ref{table: baseline vs response parameters comparison} shows these values for different proportions $p$.
+Table~\ref{table: recovery_distances} shows these values for different proportions $p$.
 For $\tau$, $90\%$ of the baseline value is recovered at a distance about $32$ km.
-For $\nu$, $110\%$ of the baseline value is reached at a distance about $7$ km.
+For $\nu$, $110\%$ of the baseline value is reached at a distance about $10$ km.
 This indicates that sound exposure can be perceived and can disturb the narwhals motion up to tens of kilometres.
 
-We want to point out that the standard errors for these recovery distances are likely to be underestimated, due to the uncertainty in the baseline estimates, and the dependence between the measurement error and the estimates of the persistence parameters.
-However, the general take-away remains valid: there is evidence of a decrease in persistence and an increase in velocity up to a couple of tens of kilometers from the sound source.
 Along with other studies, we believe it can serve as a guideline for mitigation measures towards the effects of anthropogenic noise on the narwhals behavior.
 
-\renewcommand{\arraystretch}{1.5} % Increase row height by 1.5 times
+\renewcommand{\arraystretch}{1.75} % Increase row height by 1.5 times
 
@@
 	\centering
-	\begin{tabular}{| c | *{2}{w{c}{4em}|}}
+	\begin{tabular}{ |c| *{2}{w{c}{10em}|}}
 		\hline
@@
 		\hline
-		$50 $ & $4.9 \pm 1.0$ &  $1.8 \pm 0.4$ \\
-		$10$ & $32.5 \pm 6.7 $ & $7.8 \pm 2.9$ \\
+		$50 $ & $4.28 - [2.00,7.13] $&  $2.64 - [1.20,4.25]$ \\
+		$10$ & $28.16 -[46.88; 13.18]$ & $11.24 - [5.09; 18.09]$  \\
 		\hline
@@
 	\end{tabular}
-
-	\caption{Estimated recovery distances of the baseline values within a $50 \%$ and $10\%$ range with standard errors.}
-	\label{table: baseline vs response parameters comparison}
+	\vspace{0.1cm}
+	\caption{Estimated recovery distances of the baseline values within a $50 \%$ and $10\%$ range with $95\%$ confidence intervals.}
+	\label{table: recovery_distances}
 \end{table}
+
+\subsection{ Baseline model checking}
+
+We use the fitted model to simulate constrained trajectories in the fjords domain with $10$ second time steps.
+We then fitted the baseline model on this simulated data.
+The simulated trajectories from the fitted model before exposure subsampled to $10$ minutes are shown in ~\ref{fig: simulated data} along with the observed data before exposure in figure~\ref{fig: simulated data}.
+Table ~\ref{table:check_baseline_estimations} shows that we are able to recover pretty well the parameters of the SDE.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+  \hline
+Parameter & Estimate & CI \\ 
+  \hline
+$\tau$ & $1.23$ & $[0.92; 1.61]$ \\ 
+  $\nu$ & $4.46$ & $[3.82; 5.18]$ \\ 
+  $\sigma_{\tau}$ & $0.22$ & $[0.04; 0.74]$ \\ 
+  $\sigma_{\nu}$ & $0.13$ & $[0.04; 0.30]$ \\ 
+   \hline
+\end{tabular}
+\caption{Estimates from simulated trajectories} 
+\label{table:check_baseline_estimations}
+\end{table}
+
+\begin{figure}
+\centering
+    \includegraphics[scale=0.4]{images/application/baseline/plot_baseline_crcvm2_trajectory.pdf}
+     \caption{Observed VS simulated trajectories}
+    \label{fig: simulated data}
+
+\end{figure}
+
+In terms of distance to the shore, observed and simulated data have similar characteristics, which illustrates that our model captures well the space use of the narwhals.
+We show the histograms of distances to shore for both observed and simulated data in Figure~\ref{fig: Dshore histograms}.
+
+\begin{figure}
+\centering
+    \includegraphics[scale=0.4]{images/application/baseline/Dshore_histograms.pdf}
+     \caption{Histograms of distances to shore from observed and simulated data}
+    \label{fig: Dshore histograms}
+
+\end{figure}
+
+One major difference between the observed trajectories and the simulated ones is that the narwhals have a tendency to move together and follow each other that is not captured in our model.
+In the data, A5 and A6 as well as A2 and A3 seem to move together for some period of time.
+We did not include such a behavior in our statistical model.
 
@@
 More importantly, we showed that noise exposure has a significant effect on some parameters that drive the motion by comparing a baseline and a response model.
-We believe this method can be used as a basis for the assessment of behavioral responses in many contexts, and hope it will help understanding better the effects of anthropogenic noise on marine mammals movement.\\
+We believe this method can be used as a basis for the assessment of behavioral responses in many contexts, and hope it will help understanding better the effects of anthropogenic noise on marine mammals movement.
 
@@
 \stitle{Shapefile for land geometry}
-\sdescription{Shapefile based on OpenStreetMap data for the land polygons in Scoresby Sound fjords system. We used high resolution satellite images to improve it with \texttt{qgis} software}.
+\sdescription{Shapefile based on OpenStreetMap data for the land polygons in Scoresby Sound fjords system.
+We used high resolution satellite images to improve it with \texttt{qgis} software and applied a geometric buffer of $100$ metres.}
 \end{supplement}

### 2025-03-17 Latex/AoAS - Revision 2024. Alexandre, Adeline, Susanne/supp_revised.tex
add paper revision

--- 2024-11-08 Latex/appendix/supp-aoas-template.tex
@@
 Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals.
-Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure with a base map of Scoresby Sound fjords system.
-
 \begin{table}[ht!]
@@
 
-\begin{figure}[ht!]
-	\centering
-	\begin{subfigure}{0.8\linewidth}
-		\centering
-		\includegraphics[scale=0.5]{images/data_exploration/tracksBE2.png}
-		\caption{Tracks before exposure experiments}
-	\end{subfigure}
-
-	\begin{subfigure}{0.8\linewidth}
-		\centering
-		\includegraphics[scale=0.5]{images/data_exploration/tracksAE.png}
-		\caption{Tracks during exposure experiments}
-	\end{subfigure}
-	\caption{Movement data of East Greenland narwhals. The red crosses indicate initial positions.}
-	\label{fig: tracks before and after exposure}
-\end{figure}
-
-All the relevant covariates   used for the analysis of narwhals movement are summarized in Table \ref{tab: summary covariates}.
+The relevant covariates   used for the analysis of narwhals movement are summarized in Table \ref{tab: summary covariates}.
 
@@
 		$D^{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
-		\hline
-		$I^{shore}(t)=\frac{1}{D^{shore}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
 		\hline
@@
 	\end{align*}
-	Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$ where $R_{-\omega u}$ is the rotation matrix with angle $-\omega u$, the matrix product  $\exp(-Au) \exp(-Au)^{\top}$ is simply $\exp(-\frac{2u}{\tau}) I_2$.
-We deduce that the two components of the velocity are independent and have the same variance, denoted  $q_2(\Delta)$.
-The variance is
+	An integration by part gives 
+	\[\int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du =A^{-1}(I_2-\exp(-A\Delta))(I_2-\exp(-A\Delta))^\top A^{-\top}\]. In the sequel, we denote $S(\Delta)=A^{-1}(I_2-\exp(-A\Delta))$.
+	Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$ where $R_{-\omega u}$ is the rotation matrix with angle $-\omega u$, we can deduce that the two components of the velocity are independent and have the same variance, denoted  $q_2(\Delta)$.
+This variance is
 	\begin{equation}
@@
     X(t+\Delta) &= X(t) + \int_t^{t+\Delta} V(s) \, ds \\
-    &= X(t) + \mu \Delta + \int_t^{t+\Delta} \exp(-A(s-t))(X(t) - \mu) \, ds \\
+    &= X(t) + \mu \Delta + \int_t^{t+\Delta} \exp(-A(s-t))(V(t) - \mu) \, ds \\
     &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
@@
 \begin{align*}
-    g(r) &= \frac{1}{\tau} \left(\cos(\omega r) + \exp\left( -\frac{r}{\tau} \right)\right) 
-    - \omega \sin(\omega r), \\
+    g(r) &= \frac{1}{\tau} \left(\cos(\omega r) - \exp\left( -\frac{r}{\tau} \right)\right) 
+    + \omega \sin(\omega r), \\
     h(r) &= -\frac{1}{\tau} \sin(\omega r) 
@@
 	\[
-	\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right),
+	\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right),
 	\]
@@
 Similarly, in the case $\omega=0$, we get $\gamma_2=0$ and the expression for $\gamma_1$ match equation $(7)$ in \citep{johnson_continuous_2008}.
+
+\end{proof}
+
+\section{Approximate transition density}
+
+We now suppose that the angular velocity $\omega$ varies with time. Denote $\mathcal{A}(0,t)=\int_0^t A(s) ds$ the time integral of the drift matrix.
+A similar calculation as before gives
+\[V(t+\Delta)=\exp(-\mathcal{A}(t,t+\Delta)) V(t)+\sigma \int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) dW(s)\]
+
+Therefore, $V(t+\Delta) \vert V(t)$ is normally distributed with mean
+\[\E(V(t+\Delta) \vert V(t))=\exp(-\mathcal{A}(t,t+\Delta)) V(t)\]
+and covariance matrix 
+\[\mbox{Var}(V(t+\Delta) \vert V(t))=\sigma^2\int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) \exp(-\mathcal{A}(s,t+\Delta))^\top ds\]
+
+The exact formula for the process $X$ is 
+\[X(t+\Delta)=X(t)+\int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) V(t) ds +\sigma \int_t^{t+\Delta} \int_t^s \exp(-\mathcal{A}(u,s)) dW(u) ds \]
+Consequently, the process is gaussian with mean given by
+\[\E(X(t+\Delta) \vert X(t))=X(t)+\int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) V(t) ds\]
+and covariance matrix 
+\[\mbox{Var}(X(t+\Delta) \vert X(t))=\sigma^2 \int_t^{t+\Delta} \lbrace \int_u^{t+\Delta} \exp(-\mathcal{A}(u,s))ds \rbrace \lbrace \int_u^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \rbrace^\top du\]
+
+Denote $U(t)=(X(t), V(t))$. It is a Markov process with transition density
+\[U(t+\Delta) \vert U(t)=u \sim \mathcal{N}(T(\Delta) u, Q(\Delta))\]
+where 
+\begin{equation}
+	T(\Delta) = \begin{pmatrix}
+		I_2 & \int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) ds \\
+		0_2 & \exp(-\mathcal{A}(t,t+\Delta))
+	\end{pmatrix}
+\end{equation}
+
+\begin{equation}
+	Q(\Delta) = \begin{pmatrix}
+
+		Q_{11}(\Delta) & Q_{12}(\Delta)  \\
+		Q_{21}(\Delta) &  Q_{22}(\Delta)
+	\end{pmatrix}
+\end{equation}
+where 
+\[Q_{11}(\Delta)=\sigma^2 \int_{t}^{t+\Delta} \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right) \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right)^\top du \]
+\[Q_12(\Delta)=\sigma^2 \int_{t}^{t+\Delta} \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right) \exp(-\mathcal{A}(u,t+\Delta))^\top du\]
+\[Q_{21}(\Delta)=Q_{12}(\Delta)^\top\]
+\[Q_{22}(\Delta)=\sigma^2 \int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) \exp(-\mathcal{A}(s,t+\Delta))^\top ds\]
+
+Suppose we have approximations of the time varying matrix $A$ at times $t_0=t < t_1 < \cdots < t_n=t+\Delta$ and denote $\Delta_i=t_{i+1}-t_i$ for $i \in \{0,\cdots,n-1\}$.
+Then we can approximate
+\[\exp(-\mathcal{A}(t,t+\Delta))=\prod_{i=0}^{n-1} \exp(-A_i\Delta_i)\]
+where $A_i=A(t_i)$. Each matrix $\exp(-A_i\Delta_i)$ is a weighted rotation matrix with angle $-\omega_i\Delta_i$.
+For conciseness, in the sequel we denote $E_i=\exp(-A_i\Delta_i)$, $P_{jk}= \prod_{i=j}^{k-1} E_i$ and $S_j=S(\Delta_j)$.
+
+The covariance matrix of $V$ is approximated by 
+\[Q_{22}(\Delta)\simeq\sum_{j=0}^{n-1} \sigma^2 P_{j+1,n} P_{j+1,n}^\top S_j S_j^\top\]
+The mean for the process $X$ can be estimated with the following formula
+\[\E(X(t+\Delta) \vert X(t)) \simeq X(t)+\sum_{k=0}^{n-1} P_{0,k} S_k  V(t) \]
+The covariance can also be approximated. The first step is to approximate $\mathcal{A}(u,s)$ for $u \in [t,t+\Delta]$ and $s \in [u,t+\Delta]$. For a fixed $u \in [t_{j-1},t_j]$,
+
+\begin{align*}
+	& \exp(-\mathcal{A}(u,s)) \simeq 
+	\exp(-A_{j-1}(t_j-u)) \1_{[u,t_j]}(s) +\\
+	& \sum_{k=j}^{n-1} \exp(-A_{j-1}(t_j-u)) \times \prod_{i=j}^{k-1} \exp(-A_i \Delta_i) \times \exp(-A_k (s-t_k)) \1_{[t_k,t_{k+1}]}(s)
+\end{align*}
+
+This gives the following approximation for any $u \in [t_{j-1},t_j]$
+\begin{align*}
+	&\int_u ^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \simeq A_{j-1}^{-1} (I_2-\exp(-A_{j-1} (t_j-u)))+ \\
+	&\sum_{k=j}^{n-1} \exp(-A_{j-1} (t_j-u)) \prod_{i=j}^{k-1} \exp(-A_i \Delta_i) A_k^{-1} (I_2- \exp(A_k \Delta_k)) 
+\end{align*}
+We then deduce an approximation of the covariance matrix.
+
+\begin{align*}
+	\mbox{Var}(X(t+\Delta) \vert X(t)) \simeq &\sum_{j=1}^n \sigma^2 A_{j-1}^{-1} A_{j-1}^{-\top}(\Delta_{j-1} I_2-S_{j-1}-S_{j-1}^\top + S_{j-1}S_{j-1}^\top) \\
+	&+ \sigma^2  M_{jn}^\top A_{j-1}^{-1} (S_{j-1}^\top -S_{j-1} S_{j-1}^\top)M_j M_j^\top + \sigma^2  M_{jn} A_{j-1}^{-\top} (S_{j-1} -S_{j-1} S_{j-1}^\top) \\
+	&+M_{jn} M_{jn}^\top S_{j-1} S_{j-1}^\top 
+\end{align*}
+
+where $M_{jn} = \sum_{k=j}^{n-1} P_{jk} S_k $.
+
+Finally, we have the following formula to approximate the covariance matrix between $X$ and $V$.
+
+\begin{align*}
+	\mbox{Var}(X(t+\Delta),V(t+\Delta) \vert X(t),V(t))=& \sum_{j=1}^n \sigma^2 P_{j,n}^\top A_{j-1}^{-1}(S_{j-1}^\top-S_{j-1} S_{j-1}^\top)  
+	+\sigma^2 M_{jn} P_{j,n}^\top S_{j-1} S_{j-1}^\top 
+\end{align*}
+
+We denote
+\[\tilde{T}_n=\begin{pmatrix} I_2 & \sum_{k=0}^{n-1} P_{0,k} S_k \\
+	0_2  &  \prod_{i=0}^{n-1}  E_i \end{pmatrix}
+\]
+and \[\tilde{Q}_n = \begin{pmatrix}  \tilde{Q}_n^{11}& \tilde{Q}_n^{12} \\
+	\tilde{Q}_n^{21}& \tilde{Q}_n^{22} \end{pmatrix}\]
+where 
+\begin{align*}
+	\tilde{Q}_n ^{11}=&\sum_{j=1}^n \sigma^2 A_{j-1}^{-1} A_{j-1}^{-\top}(\Delta_{j-1} I_2-S_{j-1}-S_{j-1}^\top + S_{j-1}S_{j-1}^\top) \\
+	&+ \sigma^2  M_{jn}^\top A_{j-1}^{-1} (S_{j-1}^\top -S_{j-1} S_{j-1}^\top) + \sigma^2  M_{jn} A_{j-1}^{-\top} (S_{j-1} -S_{j-1} S_{j-1}^\top) \\
+	&+\sigma^2M_{jn} M_{jn}^\top S_{j-1} S_{j-1}^\top \\
+	\tilde{Q}_n ^{12}=&\sum_{j=1}^n \sigma^2 P_{j,n}^\top A_{j-1}^{-1}(S_{j-1}^\top-S_{j-1} S_{j-1}^\top)  
+	+\sigma^2 M_{jn} P_{j,n}^\top S_{j-1} S_{j-1}^\top \\
+	\tilde{Q}_n ^{21}=&(\tilde{Q}_n ^{12})^\top \\
+	\tilde{Q}_n^{22} = & \sum_{j=0}^{n-1} \sigma^2 P_{j+1,n} P_{j+1,n}^\top S_j S_j^\top
+\end{align*}
+We can express $\tilde{T}_n$ and $\tilde{Q}_n$ in terms of the elementary matrices 
+\[T_i=\begin{pmatrix} I_2 & S_i \\ 0_2 & E_i \end{pmatrix}  \mbox {and } 
+Q_i=\begin{pmatrix} \sigma^2 A_i^{-1}A_i^{-\top} ( \Delta_iI_2-S_i-S_i^\top+S_iS_i^\top) & \sigma^2 A_i^{-1} (S_i^\top-S_iS_i^\top) \\
+	\sigma^2 A_i^{-\top} (S_i-S_iS_i^\top) & \sigma^2 S_i S_i^\top
+\end{pmatrix}
+\]
+
+\begin{equation}
+	\tilde{T}_n = T_{n-1} \times \cdots \times T_1 \times T_0
+	\label{eq: approximate link matrix}
+\end{equation}
+\begin{equation}
+	\tilde{Q}_n= \sum_{k=1}^{n-1} T_{n-1}^\top \cdots T_{k+1}^\top Q_k T_{k+1} \cdots T_{n-1}
+\end{equation}
+
+\begin{proof}
+	$\tilde{T}_n=T_0$ and $\tilde{Q}_1=Q_0$. Then, for $n \geq 2$, 
+	\begin{align*}
+		\tilde{T}_{n}=&\begin{pmatrix} 
+			I_2 & P_{0,n-1}S_{n-1}+\sum_{j=0}^{n-2} P_{0,j} S_j \\
+			0_2 & P_{0,n} E_{n-1}
+		\end{pmatrix}
+	=\begin{pmatrix}
+		I_2 & S_{n-1} \\
+		0_2 & E_{n-1}
+	\end{pmatrix}
+	\begin{pmatrix} 
+	I_2 & \sum_{j=0}^{n-2} P_{0,j} S_j \\
+	0_2 & P_{0,n-1}
+	\end{pmatrix}\\
+	&= T_{n-1} \tilde{T}_{n-1}
+	\end{align*}
+Then, using that $M_{n,n}= 0_2 $ and $P_{n,n}=I_2$ we get
+\[ \tilde{Q}_n=Z_{n-1}+Q_{n-1} \]
+where 
+\[Z_{n-1}=\begin{pmatrix} \sum_{j=1}^{n-1} Q_{j-1}^{11} +M_{jn}^\top Q_{j-1}^{12}+M_{jn} Q_{j-1}^{21}+M_{j,n}M_{j,n}^\top Q_{j-1}^{22} & \sum_{j=1}^{n-1}  P_{j,n}^\top Q_{j-1}^{12} +M_{j,n} P_{j,n}^\top Q_{j-1}^{22} \\
+	\sum_{j=1}^{n-1} P_{j,n} Q_{j-1} ^{21} + M_{jn}^\top P_{jn} Q_{j-1}^{22} & \sum_{j=1}^{n-1} P_{jn}^\top P_{jn} Q_{j-1}^{22}
+	\end{pmatrix}
+\]
+Then, since $M_{j,n}=M_{j,n-1}+P_{j,n-1} S_{n-1}$ we get
+\begin{align*} Z_{n-1}=\begin{pmatrix} \tilde{Q}_{n-1}^{11}+S_{n-1}^\top Q_{n-1}^{12}+S_{n-1} \tilde{Q}_{n-1}^{21}+S_{n-1}S_{n-1}^\top \tilde{Q}_{n-1}^{22} &  E_{n-1}^\top \tilde{Q}_{n-1}^{12}+E_{n-1}^\top S_{n-1} \tilde{Q}_{n-1}^{12}\\
+		E_{n-1} \tilde{Q}_{n-1}^{21}+E_{n-1} S_{n-1}^\top \tilde{Q}_{n-1}^{22} & E_{n-1}E_{n-1}^\top
+	\end{pmatrix}
+\end{align*}
+And this is equal to $T_{n-1}^\top \tilde{Q}_{n-1} T_{n-1}$.
 

### 2025-03-21 R/application/local/CRCVM/fit_crcvm.Rmd
update code for application to narwhal data

--- 2025-03-17 R/application/CRCVM/fit_crcvm.Rmd
@@
 library(ggpubr)
-
+library(fitdistrplus)
 ```
@@
                               data = data,type = "CRCVM_SSM",
-                              response = response,par0 = c(2,1,a,b,D0,D1,sigma_D,sigma_theta),
+                              response = response,par0 = c(1.2,4.5,a,b,D0,D1,sigma_D,sigma_theta),
                               fixpar=fixpar,
@@
 #' @return NULL (just print latex table)
-make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,n_post=100) {
-
-  est=get_baseline_estimates(sde,include_sigma_obs,n_post)
+make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,n_post=100,resp=FALSE) {
+
+  est=get_baseline_estimates(sde,include_sigma_obs,n_post,resp)
   mean_par=est$estimates
@@
 
-#' Make plot to compare BoundaryDistance in simulation and observation
-#' @param simulated_data list of data frames simulated from an SDE model, each containing a column "BoundaryDistance".
+#' Make plot to compare distributions in simulation and observation
+#' @param simulated_data list of data frames simulated from an SDE model, each containing columns "BoundaryDistance", "BoundaryAngle" and "speed"
 #' @param observed_data real GPS data (data frame).
 #' @return ggplot of boundary distances distributions
-make_distance_density_plot <- function(simulated_data, observed_data) {
+make_density_plots <- function(simulated_data, observed_data) {
 
@@
   }
+
+  # drop trajectories that reached land
 
@@
   # Plot density distribution
-  plot_combined <- ggplot(combined_data, aes(x = BoundaryDistance, color = Source, fill = Source)) +
+  plot_distance <- ggplot(combined_data, aes(x = BoundaryDistance, color = Source, fill = Source)) +
     geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
-    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
+    geom_density(data = observed_combined, alpha = 0.2) +
     xlab("Distance to Shore") +
@@
     scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
-    theme_minimal() +
+    theme(
+    axis.title = element_text(size = 16),
+    axis.text = element_text(size = 16),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 20),
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank())
     theme(legend.title = element_blank())
 
-  return(plot_combined)
-}
-
-#' Make plot to compare BoundaryAngle in simulation and observation
-#' @param simulated_data data simulated from a sde model containing a column "BoundaryAngle".
-#' @param observed real GPS data
-#' @return ggplot of boundary angle distributions
-make_angle_density_plot <- function(simulated_data, observed_data) {
-
-  # Ensure simulated_data is a list of data frames
-  if (!is.list(simulated_data) || !all(sapply(simulated_data, is.data.frame))) {
-    stop("simulated_data must be a list of data frames")
-  }
-
-  simulated_combined <- bind_rows(
-    lapply(seq_along(simulated_data), function(i) {
-      simulated_data[[i]] %>% mutate(SimulationID = as.factor(i), Source = "Simulated")
-    })
-  )
-
-  observed_combined <- observed_data %>% mutate(Source = "Observed")
-
-  simulated_combined$BoundaryAngle<-abs(simulated_combined$BoundaryAngle)
-  observed_combined$BoundaryAngle<-abs(observed_combined$BoundaryAngle)
-  combined_data <- bind_rows(simulated_combined, observed_combined)
-
-  # Plot density distribution
-  plot_combined <- ggplot(combined_data, aes(x = BoundaryAngle, color = Source, fill = Source)) +
-    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
-    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
+    plot_angle <- ggplot(combined_data, aes(x = abs(BoundaryAngle), color = Source, 
+                                            fill = Source)) +
+    geom_density(data = simulated_combined, aes(group = SimulationID), 
+                 alpha = 0.1) + 
+    geom_density(data = observed_combined, alpha = 0.2) + 
     xlab(expression(Theta)) +
@@
     scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
-    theme_minimal() +
+    theme(
+    axis.title = element_text(size = 16),
+    axis.text = element_text(size = 16),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 20),
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank())
     theme(legend.title = element_blank())
 
-  return(plot_combined)
-
+    plot_speed <- ggplot(combined_data, aes(x = speed, color = Source, 
+                                            fill = Source)) +
+    geom_density(data = simulated_combined, aes(group = SimulationID), 
+                 alpha = 0.1) + 
+    geom_density(data = observed_combined, alpha = 0.2) + 
+    xlab("Speed") +
+    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
+    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
+  theme(
+    axis.title = element_text(size = 16),
+    axis.text = element_text(size = 16),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 20),
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank())
+    theme(legend.title = element_blank())
+
+  return(list("distance"=plot_distance,"speed"=plot_speed,"angle"=plot_angle))
 }
@@
 
-#' Make plot to compare speed in simulation and observation
-#' @param simulated_data data simulated from a sde model containing a column "speed".
-#' @param observed real GPS data
-#' @return ggplot of boundary angle distributions
-make_speed_density_plot <- function(simulated_data, observed_data,response) {
-
-    # Ensure simulated_data is a list of data frames
-  if (!is.list(simulated_data) || !all(sapply(simulated_data, is.data.frame))) {
-    stop("simulated_data must be a list of data frames")
-  }
-
-  simulated_combined <- bind_rows(
-    lapply(seq_along(simulated_data), function(i) {
-      simulated_data[[i]] %>% mutate(SimulationID = as.factor(i), Source = "Simulated")
-    })
-  )
-
-  observed_combined <- observed_data %>% mutate(Source = "Observed")
-
-  combined_data <- bind_rows(simulated_combined, observed_combined)
-
-  # Plot density distribution
-  plot_combined <- ggplot(combined_data, aes(x = speed, color = Source, fill = Source)) +
-    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
-    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
-    xlab("Speed") +
-    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
-    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
-    theme_minimal() +
-    theme(legend.title = element_blank())
-
-  return(plot_combined)
-}
 ```
@@
 ```{r}
-D0=quantile(dataBE12_prep200$BoundaryDistance,0.8)
-D1=quantile(dataBE12_prep200$BoundaryDistance,0.5)
-a=0.3*D0
+D0=quantile(dataBE12_prep200$BoundaryDistance,0.9)
+D1=quantile(dataBE12_prep200$BoundaryDistance,0.75)
+a=0.22*D0
 
 baseline_crcvm200<- fit_baseline(dataBE12_prep200,interpolation_bas_200,
-                    b=0.5,a=a,sigma_D=1,D0=D0,D1=D1,sigma_theta=pi/6,
+                    b=2,a=a,sigma_D=1,D0=D0,D1=D1,sigma_theta=pi/6,
                     sigma_obs=0.05)
@@
 ```{r}
-make_final_baseline_tab(baseline_crcvm200,include_sigma_obs=FALSE,n_post=10000)
-```
-
-# Check baseline models
+make_final_baseline_tab(baseline_crcvm200,include_sigma_obs=FALSE,n_post=100)
+```
+
+# Check baseline model
 
@@
 
+## Simulate time steps
+
+```{r reproduce realistic time steps,warning=FALSE,message=FALSE}
+
+dtimes<-unlist(lapply(unique(dataBE12_prep200$ID),function(id) {
+    sub_data<-dataBE12_prep200[dataBE12_prep200$ID==id,]
+    dt=diff(sub_data$time)
+    dt
+}))
+
+fit_lognorm <- fitdist(dtimes * 60, "lnorm")
+fit_gamma <- fitdist(dtimes * 60, "gamma")
+gofstat(list(fit_lognorm,fit_gamma))
+
+```
+
+```{r}
+# Extract estimated parameters
+meanlog <- fit_lognorm$estimate["meanlog"]
+sdlog <- fit_lognorm$estimate["sdlog"]
+
+# Generate new time steps
+dtimes_sim <- rlnorm(nrow(dataBE12_prep200)-6, meanlog, sdlog)
+
+times_sim<-lapply(unique(dataBE12_prep200$ID),function(ID) {
+  sub_time<- dataBE12_prep200[dataBE12_prep200$ID==ID,"time"]
+  sub_n<-length(sub_time)
+  min_time=min(sub_time)
+  times<-rep(min_time,sub_n)+c(0,cumsum(dtimes_sim[1:(sub_n-1)]/60))
+  return (times)})
+names(times_sim)<-unique(dataBE12_prep200$ID)
+```
+
+## Subsample and interpolate simulated trajectory
+
+```{r}
+all_sim_sub=list()
+
+for (i in seq_along(all_sim)) {
+  sim_data<-all_sim[[i]]
+  interp_sim_data<-dataBE12_prep200[,c("time","ID","x","y")]
+  interp_sim_data$time<-unlist(times_sim)
+
+  for (id in unique(sim_data$ID)) {
+    sub_sim_data<-sim_data[sim_data$ID==id,]
+    sim_x_interp <- approx(sub_sim_data$time, sub_sim_data$x, xout = times_sim[[id]])$y
+    sim_y_interp <- approx(sub_sim_data$time, sub_sim_data$y, xout = times_sim[[id]])$y
+    sub_ind=interp_sim_data$ID==id
+    interp_sim_data[sub_ind,"x"]<-sim_x_interp
+    interp_sim_data[sub_ind,"y"]<-sim_y_interp
+  }
+
+  all_sim_sub[[i]]<-interp_sim_data
+}
+```
+
 ```{r subsample simulations}
 #subsample to 5 minutes and add measurement noise
-all_sim_sub<-lapply(all_sim,function(data) {
-
-  data_sub<-data[seq(1,nrow(data),by=60),]
-  noise<-rmvn(nrow(data_sub),rep(0,2),diag(rep(0.05^2,2)))
-
-  data_sub[,c("x","y")]=data_sub[,c("x","y")]+noise
-  return (data_sub)
+all_sim_sub<-lapply(all_sim_sub,function(data) {
+
+  data<-data[!(is.na(data$x)),]
+  noise<-rmvn(nrow(data),rep(0,2),diag(rep(0.02^2,2)))
+
+  data[,c("x","y")]=data[,c("x","y")]+noise
+  return (data)
 })
@@
 dataBE12_prep200$type <- "Observed"
-all_sim_sub[[1]]$type <- "Simulated"
-
-combined_data <- rbind(dataBE12_prep200[,c("BoundaryDistance","BoundaryAngle","ID","time","x","y","type")], all_sim_sub[[1]])
+all_sim_sub[[2]]$type <- "Simulated"
+
+combined_data <- rbind(dataBE12_prep200[,c("ID","time","x","y","type")], all_sim_sub[[2]])
 
@@
 
-plot_distance<-make_distance_density_plot(all_sim_sub,dataBE12_prep200) 
-plot_angle<-make_angle_density_plot(all_sim_sub,dataBE12_prep200)
-plot_speed<-make_speed_density_plot(all_sim_sub,dataBE12_prep200)
-
-final_densities_plots<-ggarrange(plotlist=list(plot_distance,plot_angle),
+plots<-make_density_plots(all_sim_sub,dataBE12_prep200)
+
+final_densities_plots<-ggarrange(plotlist=list(plots$distance,plots$angle),
                      ncol=2, nrow=1, common.legend = TRUE, legend="bottom", vjust=2)
-ggsave("BoundaryDistance_density.pdf",path="baseline_crcvm_200",plot=plot_distance)
-ggsave("BoundaryAngle_density.pdf",path="baseline_crcvm_200",plot=plot_angle)
-ggsave("speed_density.pdf",path="baseline_crcvm_200",plot=plot_speed)
+ggsave("BoundaryDistance_density.pdf",path="baseline_crcvm_200",plot=plots$distance)
+ggsave("BoundaryAngle_density.pdf",path="baseline_crcvm_200",plot=plots$angle)
+ggsave("speed_density.pdf",path="baseline_crcvm_200",plot=plots$speed)
 ```
@@
 
-colnames(dataAE_prep200)[colnames(dataAE_prep100)=="DistanceShore200"]<-"BoundaryDistance"
-colnames(dataAE_prep200)[colnames(dataAE_prep100)=="theta200"]<-"BoundaryAngle"
-
+colnames(dataAE_prep200)[colnames(dataAE_prep200)=="DistanceShore200"]<-"BoundaryDistance"
+colnames(dataAE_prep200)[colnames(dataAE_prep200)=="theta200"]<-"BoundaryAngle"
+
+dataAE_prep200$speed<-compute_speed(dataAE_prep200)
 #remove points on land
